### 2023-12-12 paper/spec.tex
Add spec file for the rules

\documentclass{article}

\usepackage{ottalt}
\usepackage{mathpartir}
\usepackage{supertabular}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}

\inputott{rules}

\title{System Specification}
\author{Yiyun Liu}

\begin{document}
\maketitle

\section{Grammar}
\ottgrammar

\section{Dynamics}
\label{sec:dynamics}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse, Bool, Eq, Refl, J, JRefl}

\section{Statics}
\label{sec:statics}
\drules[T]{$[[G |- a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Refl, J, Bool, True, False, If, Void}

\end{document}

### 2023-12-15 paper/paper.tex
Add abstract

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[nonacm]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A Short and Mechanized Consistency Proof for a Dependent Type Theory}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Yiyun Liu}
\orcid{0009-0006-8717-2498}
\affiliation{%
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \country{USA}
}
\email{liuyiyun@seas.upenn.edu}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
% \renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed lambda calculus and System F, a streamlined
presentation of proof by logical relation for a dependently language
is lacking in comparison. In this paper, I present a short
consistency proof for a dependently typed language that contains a
rich set of features, including an infinite and cumulative universe
hierarchy, natural numbers, and an intensional identity type. We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code. 
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Logical Relation, Dependent Types, Logical Consistency, Coq}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
NbE in Coq~\citep{nbeincoq}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

%%
%% If your work has an appendix, this is the place to put it.

\end{document}
\endinput
%%
%% End of file `sample-acmsmall.tex'.

### 2023-12-18 paper/paper.tex
Add intro

@@
 \section{Introduction}
-NbE in Coq~\citep{nbeincoq}
+% Depending on its application, we care about certain metatheoretic
+% properties about a type system. As a programming language, we may care
+% about type soundness, which states that a well-typed never gets stuck
+% during evaluation.
+When a type system is used as a program logic where terms encode proofs, we
+want our type system to be logically consistent, meaning
+that the empty type is not inhabited.
+This paper is specifically
+about establishing logical consistency for a fully dependently typed
+system with an infinite universe hierarchy and support for large
+elimination. % The type system, presented in Section~\ref{sec:spec}, is
+% most similar to Martin-Lof's predicate type theory with the minor
+% difference that type conversion is based on untyped equality.
+
+The proof of various dependently typed systems, including Martin-Lof's
+type theory, has long been
+established in the literature and therefore the consistency result of
+this paper is not novel. Rather,
+our goal is to present the proof in a form that is digestable by a
+working type theorist and can be more readily mechanized in a proof
+assistant. Compared to existing efforts at mechanizing logical
+consistency or stronger properties such as existence of normal
+form~\citep{nbeincoq},
+decidable type checking~\citep{decagda}, our work is minimal since it requires very
+little scaffolding and therefore results in an extremely succinct
+proof of under 1000 lines of manually written Coq code for a dependent
+type theory that is reasonably complete in its features.
+
+The key technique that underlies our consistency proof is proof by
+logical relation. In Section~\ref{sec:logrelstlc}, we review the
+technique in the context of the simply typed lambda calculus. In
+Section~\ref{sec:spec}, we present the dependent type theory of
+interest. In Section~\ref{sec:logreldep}, we give the definition
+of the logical relation for the dependent type theory. Rather than
+presenting the logical relation as an inductive-recursive definition,
+we use the more elementary concept of a partial function to capture
+the interpretation of types. The alternative representation requires us
+to show that the set of equations indeed defines a partial function;
+that is, for each input, there should always be a unique
+output.
+From the interpretation function, we can define the semantic
+typing judgment for the set of lambda terms.
+In Section~\ref{sec:logrelproof}, we prove the fundamental lemma,
+which states that syntactic typing implies semantic typing. Once the
+fundamental lemma is established, logical consistency follows as a
+trivial corollary.
+
+The downside of a pen and paper proof is that the validity of the
+definitions themselves can be dubious at times. When working with a
+dependent type theory that has a universe hierarchy, an intuitive
+understanding of naive set theory no longer suffices as the size of
+sets matters. For example, the partial function representing the
+interpretation of types does not immediately correspond to a valid
+definition in a proof assistant. We cannot easily define it as a
+function as proof assistants such as Coq and Agda only accept total
+functions. We cannot naively encode the interpretation function as an
+inductive relation either since we violate the strict positivity
+constraint. A detailed discussion of how we work around such technical
+issues can be distracting. However, what can be equally distracting is
+following along a proof where even the validity of the statement is
+not trusted. As a result, we clearly mark the ``suspicious''
+definitions, and defer the discussion of these definitions
+until Section~\ref{sec:logrelmech}, where we give an overview of our
+short Coq mechanization proof and the various proof engineering
+techniques we adopt to make the mechanization for tractable.
+
+Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
+existing literature related to logical consistency about dependent
+type theory.
+
+\section{Review of Proof by Logical Relation}
+\label{sec:logrelstlc}
+
+\section{Specification of a Dependent Type Theory}
+\label{sec:spec}
+
+\section{Semantic Typing}
+\label{sec:logreldep}
+
+\section{Consistency}
+\label{sec:logreldep}
+
+\section{Mechanization}
+\label{sec:logrelmech}
+
+\section{Related Work}
+\label{sec:relatedwork}
+
+% Type soundness can be proven through a syntactic
+% approach~\citep{syntacticsoundness} as a corollary of two properties:
+% progress and preservation. % The syntactic type soundness proof
+% % varies in complexity depending on the underlying type
+% % system. For example, a type system that tracks information flow would
+% % require additional structural rules related to security levels. In
+% % this paper, we focus on one specific type of complexity: the
+% In Figure~\ref{fig:stlcsoundness}, we summarize the structure of the
+% syntactic type soundness proof for the simply typed lambda
+% calculus. Each lemma can be proven by structural induction over the
+% typing derivation, while using the previous established results as
+% lemmas for specific cases that do not immediately follow from the
+% induction hypothesis. If we make our language more complex by adding
+% full dependent type support, the overall structure remains almost
+% identical.
+
+% NbE in Coq
 

### 2023-12-18 paper/paper.tex
Add parallel reduction to text

@@
 \documentclass[nonacm]{acmart}
-
+\usepackage{ottalt}
+\usepackage{xspace}
+\newcommand{\lang}{$\lambda^H$\xspace}
+\inputott{rules}
 %%
@@
 \label{sec:spec}
+In this section, we present the dynamics and statics of the
+dependent type theory whose logical consistency will be proven in
+Section~\ref{sec:logrelproof}. For concision, we refer to this system
+as \lang.
+
+The syntax of \lang can be found in Figure~\ref{fig:syntax}. As a
+dependent type theory, terms and types are collapsed into the same
+syntactic category. Dependent functions take the form $[[Pi A B]]$ and
+we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
+dependent on the input variable. $[[Set i]]$ represents the universe
+type where $[[i]]$ ranges over the set of natural numbers. Finally,
+\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
+proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
+$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
+is the term whose type is to be casted.
+
+\lang is expressive enough to support large
+elimination, the ability to compute a type using a term as input. For
+example, the function $[[\ Bool if 0 Bool Bool -> Bool]]$ returns
+either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
+is $[[true]]$ or $[[false]]$.
+
+Figure~\ref{fig:par} shows the definition of the parallel reduction
+relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
+represent its transitive closure, which in turn allows us to define
+the coherence relation $[[a <=> b]]$. % The $*$ symbol is more commonly
+% used for the transitive and reflexive closure of a relation. However,
+% since we can show that parallel reduction is reflexive, the transitive
+% closure of parallel reduction must also be reflexive.
+We say that two terms $[[a]]$
+and $[[b]]$ are coherent if they can eventually reduce to some common
+term $[[c]]$ through parallel reduction. The symmetric notation of
+coherence suggests that it is an equivalence relation, and in fact, we
+will use coherence as the equational theory for our type conversion rule.
+
+The following lemma shows us that parallel reduction is
+reflexive. As a result, we only need to take its transitive closure
+for the resulting relation to also be reflexive.
+\begin{lemma}[Par Refl]
+  \label{lemma:parrefl}
+  For all terms $[[a]]$, $[[a => a]]$.
+\end{lemma}
+
+\begin{lemma}[Pars refl]
+  \label{lemma:parsrefl}
+  For all terms $[[a]]$, $[[a =>+ a]]$.
+\end{lemma}
+
+A more conventional presentation would instead use full beta reduction
+instead of parallel reduction as the base for the definition of
+coherence. However, the diamond
 
@@
 \section{Consistency}
-\label{sec:logreldep}
+\label{sec:logrelproof}
 

### 2023-12-18 paper/paper.tex
Add proof about coherence

@@
 coherence suggests that it is an equivalence relation, and in fact, we
-will use coherence as the equational theory for our type conversion rule.
+will use coherence as the equational theory for our type conversion
+rule. Now, we prove that coherence is indeed an equivalence relation.
 
-The following lemma shows us that parallel reduction is
-reflexive. As a result, we only need to take its transitive closure
-for the resulting relation to also be reflexive.
+First, we show that coherence is reflexive through the following
+sequence of lemmas.
 \begin{lemma}[Par Refl]
@@
 
-A more conventional presentation would instead use full beta reduction
-instead of parallel reduction as the base for the definition of
-coherence. However, the diamond
+\begin{lemma}[Coherence refl]
+  \label{lemma:coherencerefl}
+  For all terms $[[a]]$, $[[a <=> a]]$.
+\end{lemma}
+
+Lemma~\ref{lemma:parrefl} can be proven by structural induction over
+the term $[[a]]$. Lemmas~\ref{lemma:parsrefl} and \ref{lemma:coherencerefl}
+immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.
+
+Symmetry of coherence immediately falls from its definition.
+\begin{lemma}[Coherence sym]
+  \label{lemma:coherencesym}
+  If $[[a <=> b]]$, then $[[b <=> a]]$.
+\end{lemma}
+
+Before we can prove transitivity, we need to show that parallel
+reduction satisfies the diamond property.
+\begin{lemma}[Par cong]
+  \label{lemma:parcong}
+  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
+\end{lemma}
+\begin{lemma}[Par diamond]
+  \label{lemma:pardiamond}
+  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
+  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
+\end{lemma}
+The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
+structural induction over the derivation of $[[a0 => a1]]$.
+Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
+over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
+the use of Lemma~\ref{lemma:parcong}.
+
+A relation that satisfies the
+diamond property must also be confluent, meaning that its transitive
+and reflexive closure is confluent.
+\begin{lemma}[Par confluent]
+  \label{lemma:parconfluent}
+  If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
+  $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
+\end{lemma}
+While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
+it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a
+=> b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).
+
+The transitivity of the coherence relation follows as a corollary of
+Lemma~\ref{lemma:parconfluent}.
+\begin{lemma}[Coherence trans]
+  \label{lemma:coherencetrans}
+  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
+\end{lemma}
+By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
+$[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
+$[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
+term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
+to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
+trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
+This concludes the proof that coherence is an equivalence relation.
+
+We note that a more conventional presentation of
+\rref{T-Conv} would instead use full beta reduction as the base for
+the definition of coherence. However, since full beta reduction
+doesn't satisfy the diamond property, one typically needs parallel
+reduction as an auxilliary definition to derive the confluence of full
+beta reduction. The formulation of \lang is slightly more economical.
 

### 2023-12-19 paper/paper.tex
Add figures for specification

@@
 \label{sec:spec}
+
+\begin{figure}[h]
+\[
+\begin{array}{lcll}
+\mathit{Natural\ numbers}\\
+[[i]],[[j]],[[n]] & ::= & [[0]]\ |\ [[S i]] & \mbox{zero, successor} \\ \\
+
+\mathit{Contexts}\\
+[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
+\mathit{Terms}\\
+[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
+                  & \mbox{universes, variables, empty type} \\
+            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
+                  & \mbox{function types, abstractions, applications} \\
+            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
+                  & \mbox{equality types, reflexivity proof, J eliminator} \\
+            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ |\  [[if a b0 b1]]
+                  & \mbox{boolean type, true, false, if}
+\end{array}
+\]
+  \caption{Syntax of \lang}
+  \label{fig:syntax}
+\end{figure}
+
+\begin{figure}[h]
+\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse, Bool, Eq, Refl, J, JRefl}
+\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{One, Step}
+\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
+\caption{Parallel reduction and coherence}
+\label{fig:par}
+\end{figure}
+
+\begin{figure}[h]
+\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
+\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Refl, J, Bool, True, False, If, Void}
+\caption{Syntactic typing for \lang}
+\label{fig:typing}
+\end{figure}
+
 In this section, we present the dynamics and statics of the

### 2023-12-19 paper/paper.tex
Add a paragraph to STLC LR proof

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{A Short and Mechanized Consistency Proof for a Dependent Type Theory}
+\title{A Short and Mechanized Consistency Proof for Dependent Type Theory}
 
@@
 \label{sec:logrelstlc}
+In this section, we give a quick review of the consistency proof of the
+simply typed lambda calculus augmented with the empty type through
+proof by logical relation.
+
+We give the full specification of the simply typed calculus in
+Figure~\ref{fig:stlcsyntax},\ref{fig:stlctyping},\ref{fig:stlcdynamics}
+without any further explanation since it is standard.
+
+The goal here is to show that the empty type cannot be
+inhabited. A naive attempt at proof by structural induction
 

### 2023-12-19 paper/paper.tex
Add figure for semantic typing (WIP)

@@
 \label{sec:logreldep}
+\begin{figure}[h]
+  \begin{equation*}
+    \begin{split}
+      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
+      [ [[Void]] ]_i &= \emptyset \\
+      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text { and } [[a<=>b]] \}\\
+    \end{split}
+  \end{equation*}
+  \caption{Logical relation for \lang}
+  \label{fig:logrel}
+\end{figure}
+
+In this section, we show the definition of the logical relation as a
+partial function in Figure~\ref{fig:logrel}. The logical relation can
+be viewed as a partial function that takes a closed type,
+a natural number $[[i]]$ representing the universe level, and
+returns a set of closed terms that \emph{semantically} inhabit the
+closed type, though we do not strictly enforce the closedness of
+either the type or the term.
 

### 2023-12-19 paper/paper.tex
Complete the logical relation

@@
     \begin{split}
+      [-]_= &:: \mathit{Term} \rightarrow \mathcal{P}(\mathit{Term})\\
+      [ [[Void]] ]_i &= \emptyset \\
       [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
-      [ [[Void]] ]_i &= \emptyset \\
-      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text { and } [[a<=>b]] \}\\
+      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
+      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
+      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
+      [ [[ A ]] ]_i &= \{ [[a]] \mid [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
     \end{split}

### 2023-12-20 paper/paper.tex
Minor

--- 2023-12-19 paper/paper.tex
@@
     \begin{split}
+      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
+      [ [[Void]] ]_i &= \emptyset \\
       [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
-      [ [[Void]] ]_i &= \emptyset \\
-      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text { and } [[a<=>b]] \}\\
+      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
+      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
+      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
+      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
     \end{split}
@@
 either the type or the term.
+
+The claim that $[ - ]_=$ is a partial function is not immediately
+obvious due to the last equation in its definition. The last equation
+ensures that the if some type $[[A]]$ is in the domain of the
+interpretation function, then any type $[[B]]$ that steps into $[[A]]$
+has the same interpretation as $[[A]]$. For example,
+since $[[if true Bool Void]]$ steps into $[[Bool]]$, we can say that
+$[[if true Bool Void]]$ is semantically inhabited by terms that
+reduces down to either $[[true]]$ or $[[false]]$ through the
+$[[Bool]]$ case of the interpretation function.
+
+However, due to the non-deterministic nature of parallel reduction,
+given some terms $[[A]]$, $[[B0]]$, $[[B1]]$ such that $[[A => B0]]$
+and $[[A => B1]]$, the meaning of $[[A]]$ could either be $[ [[B0]] ]_i$
+or $[ [[B1]] ]_i$. In fact, the definedness of $[ [[B0]] ]_i$ and $[
+[[B1]] ]_i$
+should not be taken as granted either. It is unclear for now whether
+the definedness of one implies the other.
+
+Fortunately,
 

### 2023-12-20 paper/paper.tex
Add definition for the recursor

--- 2023-12-19 paper/paper.tex
@@
     \begin{split}
+      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
+      [ [[Void]] ]_i &= \emptyset \\
       [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
-      [ [[Void]] ]_i &= \emptyset \\
-      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text { and } [[a<=>b]] \}\\
+      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
+      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
+      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
+      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
     \end{split}
   \end{equation*}
-  \caption{Logical relation for \lang}
+  \caption{Specification of the logical relation for \lang}
   \label{fig:logrel}
@@
 
-In this section, we show the definition of the logical relation as a
-partial function in Figure~\ref{fig:logrel}. The logical relation can
-be viewed as a partial function that takes a closed type,
-a natural number $[[i]]$ representing the universe level, and
-returns a set of closed terms that \emph{semantically} inhabit the
-closed type, though we do not strictly enforce the closedness of
-either the type or the term.
+In this section, we present the definition of the logical relation as a
+partial function.
+Figure~\ref{fig:logrel} shows the specification
+of the logical relation that we are about to define. The specification
+itself does not immediately induce a definition as it is unclear
+whether there exists a smallest partial function that satisfies the
+specification.
+Without such as proof, we would not be able to induct on the graph of
+the function or use any inversion principle.
+
+The construction of a function that meets the specification in
+Figure~\ref{fig:logrel} can be broken down into two steps. To build an
+infinite hierarchy of universes, we first define a recursor that constructs
+the interpretation of the universe level $[[n]]$ given a parameterized
+interpretation function for universe levels that are strictly less
+than $[[n]]$. The well-definedness of this inductive relation requires
+some extra justification as it is not strictly positive. Instead, we
+must manually prove the monotonicity of the underlying functor.
+Next, we use the well-foundedness of the less than relation to tie the
+knot and define the full interpretation function that meets the
+specification.
+
+For concision, we first fix the following parameters:
+\begin{equation*}
+  \begin{split}
+    i &:: \mathit{Nat} \\
+    I &:: \mathit{Nat} \rightarrow \mathit{Term} \times \mathcal{P}(\mathit{Term})
+  \end{split}
+\end{equation*}
+
+Now, consider the following function $F$:
+\begin{equation*}
+  \begin{split}
+    F &:: \mathit{Term} \times \mathcal{P}(\mathit{Term}) \rightarrow \mathit{Term} \times \mathcal{P}(\mathit{Term})\\
+    F(X) &= F_V  \cup F_B \cup F_I \cup F_\Pi(X) \cup F_U \cup F_S(X) \\
+    F_V &= \{ ([[Void]], \emptyset) \} \\
+    F_B &= \{ ([[Bool]], \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a
+          =>+ false]] \})  \} \\
+    F_I &= \{ ([[a ~ b : A]], \{ [[p]] \mid [[p =>+ refl]] \text{ and
+          } [[a<=>b]] \})  \} \\
+          % This definition is wrong because it'd include all pi types
+    F_{\Pi}(X) &= ([[Pi A B]], \{ [[b]] \mid \exists A\text{ and }S, (A,S) \in X,
+                   \forall [[a]], \text{ if } [[a]] \in S \text{, then
+                 } \\
+    F_U &= \{([[Set j]] , \{ [[A]] \mid [[A]] \in
+          dom(I_j) \}) \mid j < i \} \\
+    F_S(X) &= ([[A]], \{ [[a]] \mid \exists ([[B]], S) \in X, [[a]]
+          \in S \text{ and } [[A => B]] \})  \\
+  \end{split}
+\end{equation*}
+
+%  The logical relation can
+% be viewed as a partial function that takes a closed type,
+% a natural number $[[i]]$ representing the universe level, and
+% returns a set of closed terms that \emph{semantically} inhabit the
+% closed type, though we do not strictly enforce the closedness of
+% either the type or the term.
+
+% The claim that $[ - ]_=$ is a partial function is not immediately
+% obvious due to the last equation in its definition. The last equation
+% ensures that the if some type $[[A]]$ is in the domain of the
+% interpretation function, then any type $[[B]]$ that steps into $[[A]]$
+% has the same interpretation as $[[A]]$. For example,
+% since $[[if true Bool Void]]$ steps into $[[Bool]]$, we can say that
+% $[[if true Bool Void]]$ is semantically inhabited by terms that
+% reduces down to either $[[true]]$ or $[[false]]$ through the
+% $[[Bool]]$ case of the interpretation function.
+
+% However, due to the non-deterministic nature of parallel reduction,
+% given some terms $[[A]]$, $[[B0]]$, $[[B1]]$ such that $[[A => B0]]$
+% and $[[A => B1]]$, the meaning of $[[A]]$ could either be $[ [[B0]] ]_i$
+% or $[ [[B1]] ]_i$. In fact, the definedness of $[ [[B0]] ]_i$ and $[
+% [[B1]] ]_i$
+% should not be taken as granted either. It is unclear for now whether
+% the definedness of one implies the other.
+
+% Fortunately,
 

### 2023-12-22 paper/paper.tex
Set up ott to type set set theory

--- 2023-12-20 paper/paper.tex
@@
   \end{equation*}
-  \caption{Logical relation for \lang}
+  \caption{Specification of the logical relation for \lang}
   \label{fig:logrel}
@@
 
-In this section, we show the definition of the logical relation as a
-partial function in Figure~\ref{fig:logrel}. The logical relation can
-be viewed as a partial function that takes a closed type,
-a natural number $[[i]]$ representing the universe level, and
-returns a set of closed terms that \emph{semantically} inhabit the
-closed type, though we do not strictly enforce the closedness of
-either the type or the term.
+In this section, we present the definition of the logical relation as a
+partial function.
+Figure~\ref{fig:logrel} shows the specification
+of the logical relation that we are about to define. The specification
+itself does not immediately induce a definition as it is unclear
+whether there exists a smallest partial function that satisfies the
+specification.
 
-The claim that $[ - ]_=$ is a partial function is not immediately
-obvious due to the last equation in its definition. The last equation
-ensures that the if some type $[[A]]$ is in the domain of the
-interpretation function, then any type $[[B]]$ that steps into $[[A]]$
-has the same interpretation as $[[A]]$. For example,
-since $[[if true Bool Void]]$ steps into $[[Bool]]$, we can say that
-$[[if true Bool Void]]$ is semantically inhabited by terms that
-reduces down to either $[[true]]$ or $[[false]]$ through the
-$[[Bool]]$ case of the interpretation function.
-
-However, due to the non-deterministic nature of parallel reduction,
-given some terms $[[A]]$, $[[B0]]$, $[[B1]]$ such that $[[A => B0]]$
-and $[[A => B1]]$, the meaning of $[[A]]$ could either be $[ [[B0]] ]_i$
-or $[ [[B1]] ]_i$. In fact, the definedness of $[ [[B0]] ]_i$ and $[
-[[B1]] ]_i$
-should not be taken as granted either. It is unclear for now whether
-the definedness of one implies the other.
-
-Fortunately,
+The construction of a function that meets the specification in
+Figure~\ref{fig:logrel} can be broken down into two steps. To build an
+infinite hierarchy of universes, we first define a recursor that constructs
+the interpretation of the universe level $[[n]]$ given a parameterized
+interpretation function for universe levels that are strictly less
+than $[[n]]$.
+Next, we use the well-foundedness of the less than relation to tie the
+knot and define the full interpretation function that meets the
+specification.
+The resulting definition is then shown to be functional and thus
+justifies the notation $[ [[A]] ]_i$ since we know there would be an
+unambiguous output corresponding to the type $[[A]]$, assuming that
+$[[A]]$ is in the domain of the partial function.
 

### 2023-12-22 paper/paper.tex
Add logical relation (inductive ver) to the paper

@@
   \begin{split}
-      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
-      [ [[Void]] ]_i &= \emptyset \\
-      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
-      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
-      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
-      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
-      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
+    [[InterpR i A S]] &:= [[ Interp I i A S  ]] \\
+    [[I i]] &:= [[exists S , InterpR i A S]]
   \end{split}
 \end{equation*}
-  \caption{Specification of the logical relation for \lang}
+\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
+\caption{Logical relation for \lang}
 \label{fig:logrel}
 \end{figure}
+In this section, we define the logical relation for \lang{} in the
+form of an inductively defined relation (Figure~\ref{fig:logrel}). The
+logical relation takes the form $[[Interp I i A S]]$, where $[[I]]$ is a
+function that takes a natural number strictly less than $[[i]]$ to a set of
+terms, $[[i]]$ is a natural number representing the universe level,
+$[[A]]$ is a type, and $[[S]]$ a set of terms.
+% Intuitively, the relation $[[Interp I i A S]]$ takes $[[I]]$, an interpretation
+% for types from universes smaller than $[[Set i]]$ and builds the types
+% that semantically inhabit $[[Set i]]$.
+Ignoring $[[I]]$ for a moment, the notation $[[Interp I i A S]]$ reads
+that $[[A]]$ is a level-$[[i]]$ type
+semantically inhabited by terms from the set $[[S]]$. For example, the
+$[[Void]]$ type is not inhabited by any terms as it corresponds to the
+empty set, whereas the $[[Bool]]$ type is inhabted by the set of (not
+necessarily closed) terms that reduce to either $[[true]]$ or
+$[[false]]$. Not all terms have an interpretation since some terms do
+not correspond to types and therefore cannot be inhabited. For
+example, $[[true]]$ is a boolean value and therefore does not have a
+set corresponding to it.
 
-In this section, we present the definition of the logical relation as a
-partial function.
-Figure~\ref{fig:logrel} shows the specification
-of the logical relation that we are about to define. The specification
-itself does not immediately induce a definition as it is unclear
-whether there exists a smallest partial function that satisfies the
-specification.
+\Rref{T-Pi} requires some special attention. Let's focus first on the
+premises of \rref{T-Pi} and ignore the set that corresponds to $[[Pi A
+B]]$. For $[[Pi A B]]$ to have a meaning, we
+require that there exists some interpretation $[[S]]$ for the input
+type $[[A]]$. For each term $[[a]]$ in the set $[[S]]$, we also
+require that there exists some set that corresponds to the term $[[B
+{a}]]$. The function $[[F]]$ appearing in the premise is simply a
+result of skolemization. In fact, we can alternatively formulate the
+premise of \rref{T-Pi} as follows:
 
-The construction of a function that meets the specification in
-Figure~\ref{fig:logrel} can be broken down into two steps. To build an
-infinite hierarchy of universes, we first define a recursor that constructs
-the interpretation of the universe level $[[n]]$ given a parameterized
-interpretation function for universe levels that are strictly less
-than $[[n]]$.
-Next, we use the well-foundedness of the less than relation to tie the
-knot and define the full interpretation function that meets the
-specification.
-The resulting definition is then shown to be functional and thus
-justifies the notation $[ [[A]] ]_i$ since we know there would be an
-unambiguous output corresponding to the type $[[A]]$, assuming that
-$[[A]]$ is in the domain of the partial function.
+This alternative definition becomes problematic when we try to fill
+out the missing conclusion of the rule. With the existential variant,
+we cannot conveniently refer to the witness that corresponds
+to the interpretation of the type $[[B { a }]]$ from the premise. It
+is tempting to rewrite the conclusion as follows, though this
+alternative definition ...
 
+In a fully dependently typed system, we may perform computations at
+the type level. Given $[[empty |- a : A]]$, the type $[[A]]$ may not
+be in head form but can reduce to the $[[Bool]]$ type. \Rref{I-Red}
+allows us to assign a type-level computation a meaning based on what
+type value it reduces to eventually.
+
+The addition of \rref{I-Red} requires us to prove several inversion
+lemmas about the logical relation. For example, given a proof of
+
+% In the rules for $[[Interp I i A S]]$, we exclude cases where $[[A]]$
+% is neutral. Since we only care about consistency, it suffices to focus
+% only on terms that evaluate to closed terms.
 \section{Consistency}

### 2023-12-25 paper/paper.tex
Add the inversion lemmas

--- 2023-12-22 paper/paper.tex
@@
 \begin{figure}[h]
+\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
+\caption{Logical relation for \lang}
+\label{fig:logrel}
+\end{figure}
+In this section, we define the logical relation for \lang{} in the
+form of an inductively defined relation (Figure~\ref{fig:logrel}). The
+logical relation takes the form $[[Interp I i A S]]$. The
+metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
+numbers respectively, as introduced earlier in
+Figure~\ref{fig:syntax}.
+The metavariables $[[I]]$ and $[[S]]$ are
+sets with the following signatures:
 \begin{equation*}
   \begin{split}
-      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
-      [ [[Void]] ]_i &= \emptyset \\
-      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
-      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
-      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
-      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
-      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
+    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
+    [[S]] &\in [[PowerSet STm]]
   \end{split}
 \end{equation*}
-  \caption{Specification of the logical relation for \lang}
-  \label{fig:logrel}
+The function $[[I]]$ is a family of subsets of terms indexed by
+natural numbers strictly less than the parameter $[[i]]$, which
+represents the current universe level.  In \rref{I-Set}, function
+$[[I]]$ is used to define the meaning of
+universes that are strictly smaller than the current level $[[i]]$. The
+restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
+be predicative. Removing the ordering constraint would result in a
+system where one can encode Girard's paradox~\citep{girard-thesis}.
+
+\begin{figure}[h]
+\begin{equation*}
+  \begin{split}
+    [[InterpR i A S]] &:= [[ Interp F i A S  ]] \\
+    [[F i]] &:= [[exists S , InterpR i A S]]
+  \end{split}
+\end{equation*}
+\caption{Logical relation for all universe levels}
+\label{fig:logrelrec}
 \end{figure}
+To tie the knot and obtain an interpretation of all universe levels,
+we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
+using the well-foundedness of the strict less than relation on natural
+numbers.
+The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
+level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
+$[[S]]$. The way it's constructed enables us to reason about its
+property by nested induction over the less than relation and the
+structure of the relation $[[Interp I i A S]]$, though for the
+majority of the cases we do not need any information about the
+parameterized function $[[I]]$, where the property about $[[InterpR i
+A S]]$ follows as a corollary of a property about $[[Interp I i A S]]$
+with no assumptions imposed on $[[I]]$.
 
-In this section, we present the definition of the logical relation as a
-partial function.
-Figure~\ref{fig:logrel} shows the specification
-of the logical relation that we are about to define. The specification
-itself does not immediately induce a definition as it is unclear
-whether there exists a smallest partial function that satisfies the
-specification.
+For the rest of the section, we establish some important facts about
+the logical relation that will be useful for proving fundamental lemma
+in Section~\ref{sec:logrelproof}.
 
-The construction of a function that meets the specification in
-Figure~\ref{fig:logrel} can be broken down into two steps. To build an
-infinite hierarchy of universes, we first define a recursor that constructs
-the interpretation of the universe level $[[n]]$ given a parameterized
-interpretation function for universe levels that are strictly less
-than $[[n]]$.
-Next, we use the well-foundedness of the less than relation to tie the
-knot and define the full interpretation function that meets the
-specification.
-The resulting definition is then shown to be functional and thus
-justifies the notation $[ [[A]] ]_i$ since we know there would be an
-unambiguous output corresponding to the type $[[A]]$, assuming that
-$[[A]]$ is in the domain of the partial function.
+The relation $[[Interp I i A S]]$ satisfies the following inversion
+principles.
+\begin{lemma}[Void Inv]
+  \label{lemma:voidinv}
+  If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
+\end{lemma}
+\begin{lemma}[Bool Inv]
+  \label{lemma:boolinv}
+  If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
+\end{lemma}
+\begin{lemma}[Eq Inv]
+  \label{lemma:eqinv}
+  If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
+\end{lemma}
+\begin{lemma}[Pi Inv]
+  \label{lemma:piinv}
+  If $[[Interp I i Pi A B S]]$, then there exists $[[S]],[[F]]$ such
+  that the following conditions hold:
+  \begin{itemize}
+  \item $[[Interp I i A S ]]$
+  \item $[[F in S -> PowerSet STm]]$
+  \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+  \end{itemize}
+\end{lemma}
+\begin{lemma}
+  \label{lemma:setinv}
+  If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
+\end{lemma}
 

### 2023-12-27 paper/paper.tex
Add preservation (WIP)

@@
 
+From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
+single substitution property as a simple corollary.
+\begin{corollary}[Par subst]
+  \label{lemma:parsubst}
+  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
+\end{corollary}
+
 A relation that satisfies the
@@
 This concludes the proof that coherence is an equivalence relation.
+\begin{lemma}[Coherence Equivalence]
+  \label{lemma:coherenceequiv}
+  The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
+  transitivity and therefore is an equivalence relation.
+\end{lemma}
 
@@
 principles.
-\begin{lemma}[Void Inv]
-  \label{lemma:voidinv}
-  If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
-\end{lemma}
-\begin{lemma}[Bool Inv]
-  \label{lemma:boolinv}
-  If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
-\end{lemma}
-\begin{lemma}[Eq Inv]
-  \label{lemma:eqinv}
-  If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
-\end{lemma}
-\begin{lemma}[Pi Inv]
-  \label{lemma:piinv}
-  If $[[Interp I i Pi A B S]]$, then there exists $[[S]],[[F]]$ such
-  that the following conditions hold:
+\begin{lemma}[Inversion of the logical relation]
+  \label{lemma:interpinv}\leavevmode
+  \begin{enumerate}
+  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
+  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
+  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
+  \item If $[[Interp I i Pi A B S]]$, then there exists $[[S]],[[F]]$ such that:
     \begin{itemize}
@@
     \end{itemize}
+  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
+  \end{enumerate}
 \end{lemma}
-\begin{lemma}
-  \label{lemma:setinv}
-  If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
+\begin{proof}
+  We only show the most involved function case; the rest follows a
+  similar but simpler pattern. We start by
+  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
+  are only two possible cases we need to consider.
+  \begin{description}
+  \item[\Rref{I-Pi}:] Immediate.
+  \item[\Rref{I-Red}:] We know that there exists some $[[A0]]$ and
+    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$, and from the
+    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
+    \begin{itemize}
+    \item $[[Interp I i A0 S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
+    \end{itemize}
+    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
+    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
+    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
+    $[[F]]$ satisfies the desired properties:
+    \begin{itemize}
+    \item $[[Interp I i A S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+    \end{itemize}
+  \end{description}
+\end{proof}
+
+\Rref{I-Red} bakes into the logical relation the backward preservation
+property. The following property shows that preservation holds in the
+usual forward direction too.
+\begin{lemma}[Preservation of the logical relation]
+  \label{lemma:interppreservation}
+  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
 \end{lemma}
+\begin{proof}
+  We carry out the proof by induction over the derivation of $[[Interp
+  I i A S]]$.
+  \begin{description}
+  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
+    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
+    be $[[Void]]$ and the result trivially follows.
+  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
+  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
+      refl , a <=> b  }]]$ and, by inverting the derivation of
+    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
+    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
+    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
+    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
+      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
+    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
+      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
+    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
+    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
+    <=> b0]]$. The result then immediately follows from the fact that
+    coherence is an equivalence relation
+    (Lemma~\ref{lemma:coherenceequiv}).
+  \item[\Rref{I-Pi}:]
+  \end{description}
+\end{proof}
 

### 2023-12-27 paper/paper.tex
Finish typesetting the proof of determinism

@@
 dependent on the input variable. $[[Set i]]$ represents the universe
-type where $[[i]]$ ranges over the set of natural numbers. Finally,
+type where $[[i]]$ ranges over the set of natural numbers.
+ Finally,
 \lang also includes an intensional identity type $[[a ~ b : A]]$ whose
@@
 \begin{equation*}
-  \begin{split}
-    [[InterpR i A S]] &:= [[ Interp F i A S  ]] \\
-    [[F i]] &:= [[exists S , InterpR i A S]]
-  \end{split}
+    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
 \end{equation*}
@@
 using the well-foundedness of the strict less than relation on natural
-numbers.
+numbers (recall that
+the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
+numbers strictly less than $[[i]]$ as its input).
 The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
@@
   \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
-  \item If $[[Interp I i Pi A B S]]$, then there exists $[[S]],[[F]]$ such that:
+  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
     \begin{itemize}
@@
     \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
     \end{itemize}
@@
   \item[\Rref{I-Pi}:] Immediate.
-  \item[\Rref{I-Red}:] We know that there exists some $[[A0]]$ and
-    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$, and from the
+  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
+    We know that there exists some $[[A0]]$ and
+    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
+    A0 B0 S0]]$. From the
     induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
@@
     \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
+    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
     \end{itemize}
@@
     $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
-    $[[F]]$ satisfies the desired properties:
+    $[[F]]$ satisfies the following properties:
     \begin{itemize}
@@
     \end{itemize}
+    The properties above are exactly the preconditions needed to apply
+    \rref{I-Pi} for $[[Pi A B]]$ to finish off the proof.
   \end{description}
@@
     (Lemma~\ref{lemma:coherenceequiv}).
-  \item[\Rref{I-Pi}:]
+  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
+    such that:
+    \begin{itemize}
+    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
+        a #) }]]$
+    \item $[[Interp I i A S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+    \end{itemize}
+    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
+    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
+      a #) }]]$. By \rref{I-Pi}, it suffices to show:
+    \begin{itemize}
+    \item $[[Interp I i A0 S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
+    \end{itemize}
+    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
+    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
+    conditions follow immediately from the induction hypothesis.
+  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
+    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
+    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
+    property of parallel
+    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
+    $[[C]]$ such that $[[B0 => C]]$ and $[[B => C]]$. By the induction
+    hypothesis, we derive $[[Interp I i C S]]$ from $[[Interp I i
+    B0 S]]$. By \rref{I-Red} and
+    $[[B => C]]$, we conclude that $[[Interp I i B S]]$.
+  \end{description}
+\end{proof}
+From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
+derive the following corollary that two coherent types have the same
+interpretation.
+\begin{corollary}[Logical Relation for Coherent Types]
+  \label{lemma:logrelcoherence}
+  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
+\end{corollary}
+Next, we show that $[[Interp I i A S]]$ is in fact a partial function.
+\begin{lemma}[Logical relation is functional]
+  \label{lemma:logreldeter}
+  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
+\end{lemma}
+\begin{proof}
+  We start by inducting over the derivation of the first premise $[[Interp I i A
+  S0]]$.
+  \begin{description}
+  \item[\Rref{I-Void}:] We know that $[[Interp I i Void
+    emptyset]]$. Given $[[Interp I i Void S1]]$, our goal is to show that
+    $[[emptyset = S1]]$. This is immediate by applying the $[[Void]]$ case of
+    Lemma~\ref{lemma:interpinv} to $[[Interp I i Void S1]]$.
+  \item[\Rref{I-Bool, I-Eq, I-Set}:] Similar to the \rref{I-Void} case
+    by applying the matching case of Lemma~\ref{lemma:interpinv}
+    to $[[Interp I i A S1]]$.
+  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
+    such that:
+    \begin{itemize}
+    \item $[[Interp I i A S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+    \end{itemize}
+    Our goal is to show that given $[[Interp I i Pi A B S1]]$, we have
+    $[[S1 = { b | forall a, (# a in S implies b a in F
+      a #) }]]$.
+    By the function case of Lemma~\ref{lemma:interpinv}, there exists
+    some $[[S0]]$ and $[[F0]]$ such that:
+    \begin{itemize}
+    \item $[[Interp I i A S0 ]]$
+    \item $[[F0 in S0 -> PowerSet STm]]$
+    \item
+      $[[forall a, (# a in S0 implies Interp I i B { a } F0 a #)]]$
+    \item
+      $[[S1 = { b | forall a, (# a in S0 implies b a in F0 a #) }]]$
+    \end{itemize}
+    It suffices to show that $[[S = S0]]$ and $[[F = F0]]$. The
+    equality $[[S = S0]]$ immediately follows from the induction
+    hypothesis since $[[Interp I i A S]]$ and $[[Interp I i A
+    S0]]$. Therefore, functions $[[F]]$ and $[[F0]]$ have the same
+    domain and codomain and thus it suffices to show that forall $[[a in S]]$,
+    $[[F a = F0 a]]$. Suppose $[[a in S]]$, we must have $[[Interp I
+    i B {a} F a]]$ and $[[Interp I i B {a} F0 a]]$ from the two
+    $\forall$-quantified statements above. The equality $[[F a = F0 a]]$ then
+    immediately follows from the induction hypothesis.
+  \item[\Rref{I-Red}:] There exists some $[[B]]$ such that $[[A =>
+    B]]$ and $[[Interp I i B S0]]$. Our goal is to show that given
+    $[[Interp I i A S1]]$, we have $[[S0 = S1]]$. By
+    Lemma~\ref{lemma:interppreservation}, from $[[Interp I i A S1]]$
+    and $[[A => B]]$, we have $[[Interp I i B S1]]$. From the
+    induction hypothesis, we can conclude that $[[S0 = S1]]$ since
+    $[[Interp I i B S0]]$ and $[[Interp I i B S1]]$.
   \end{description}

### 2023-12-28 paper/paper.tex
Finish typesetting all proofs in semtyping.v

@@
 
-\section{Semantic Typing}
+\section{Logical Relation}
 \label{sec:logreldep}
@@
 level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
-$[[S]]$. The way it's constructed enables us to reason about its
-property by nested induction over the less than relation and the
-structure of the relation $[[Interp I i A S]]$, though for the
-majority of the cases we do not need any information about the
-parameterized function $[[I]]$, where the property about $[[InterpR i
-A S]]$ follows as a corollary of a property about $[[Interp I i A S]]$
-with no assumptions imposed on $[[I]]$.
+$[[S]]$. For the
+majority of the properties we are about to prove in this section, we
+do not need any information about the parameterized function $[[I]]$.
+Each property about $[[InterpR i A S]]$ follows as a corollary of
+a property about $[[Interp I i A S]]$ with no or few assumptions imposed on
+$[[I]]$. As a result, we usually state our lemmas in terms of
+$[[Interp I i A S]]$ without duplicating them in terms of $[[InterpR i
+A S]]$.
 
@@
 
-\section{Consistency}
+The next lemma shows that our logical relation satisfies
+cumulativity. That is, if a type has an interpretation at a lower
+universe level, then we can obtain the same interpretation at a higher
+universe level.
+\begin{lemma}[Logical relation cumulativity]
+  \label{lemma:logrelcumulativity}
+  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
+\end{lemma}
+\begin{proof}
+  Trivial by structural induction over the derivation of $[[Interp I
+  i0 A S]]$.
+\end{proof}
+Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
+implicitly assume that $[[I]]$ is defined on the set of natural
+numbers less than $[[i1]]$.
+
+\begin{corollary}[Logical relation is functional with different levels]
+  \label{lemma:logreldeterhet}
+  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
+\end{corollary}
+\begin{proof}
+  Immediate from Lemma~\ref{lemma:logreldeter} and
+  \ref{lemma:logrelcumulativity}.
+\end{proof}
+
+We say that a set of terms $[[S]]$ is closed under expansion if given
+$[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
+The final property we want to show is that the output set $[[S]]$ from
+the logical relation is closed under expansion. Unlike the previous
+lemmas, we need to constrain the function $[[I]]$ so its outputs are
+all closed under expansion.
+\begin{lemma}[Logical Relation Elements back preservation]
+  \label{lemma:logrelbackclos}
+  If $[[Interp I i A S]]$ and $[[I]]$ satisfies the property that
+  for all $[[i]]$, $[[I i]]$ is closed under expansion, then the set
+  $[[S]]$ is closed under expansion.
+\end{lemma}
+\begin{proof}
+  Trivial by structural induction over the derivation of $[[Interp I i
+  A S]]$. The function case requires the following simple fact about
+  parallel reduction:
+  If $[[b0 => b1]]$ then $[[b0 a => b1 a]]$ for all $[[a]]$. This fact
+  is not immediate from the definition of parallel reduction but
+  follows from Lemma~\ref{lemma:parrefl} and \rref{P-App}.
+\end{proof}
+
+\begin{corollary}
+  \label{lemma:logrelNbackclos}
+  If $[[InterpR i A S]]$, then $[[S]]$ is closed under expansion.
+\end{corollary}
+\begin{proof}
+  Immediate from Lemma~\ref{lemma:logrelbackclos}, the definition of
+  $[[InterpR i A S]]$, and \rref{I-Red}.
+\end{proof}
+
+\section{Semantic Typing and Consistency}
 \label{sec:logrelproof}

### 2023-12-28 paper/paper.tex
Add a paragraph to the semantic typing section

@@
 \label{sec:logrelproof}
+The logical relation we define in Figure~\ref{fig:logrel} does not
+include cases for variables. Likewise, for the base types such as
+boolean and equality, the output set $[[S]]$ contains only terms that
+evaluate to closed terms. To generalize our logical relation to open
+terms, we define the semantic typing judgment by closing the open
+terms with a substitution whose codomain consists of terms that
+respect the interpretation of the types from the context. The full
+definition of semantic typing ($[[ G |= a : A]]$) is presented in
+Figure~\ref{fig:semtyping}.
 

### 2023-12-30 paper/paper.tex
Add figure for semantic typing

@@
 \label{sec:logrelproof}
+\begin{figure}[h]
+  \begin{equation*}
+    \begin{split}
+      [[rho |= G]] &:= \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
+                     } [[InterpR j (up S i G i) { rho } S ]] \text{, then } [[rho i in S]] \\
+      [[G |= a : A]] &:= \forall [[rho]], \text{ if }[[rho |=
+                       G]]\text{ then there exists some } [[j]] \text{
+                       and } [[S]] \text{ such that } [[InterpR j A
+                       {rho} S]] \text{ and } [[a {rho} in S]] \\
+      [[|= G]] &:= \forall [[i < |G|]], \exists [[j]], [[drop S i G |= G i : Set j]]
+    \end{split}
+  \end{equation*}
+  \caption{Semantic Typing for \lang}
+  \label{fig:semtyping}
+\end{figure}
+
 The logical relation we define in Figure~\ref{fig:logrel} does not
@@
 respect the interpretation of the types from the context. The full
-definition of semantic typing ($[[ G |= a : A]]$) is presented in
-Figure~\ref{fig:semtyping}.
+definitions of well-formed substitution ($[[rho |= G]]$), semantic
+typing ($[[ G |= a : A]]$), and semantic context well-formedness
+($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
+
+Next, we show that semantic typing is implied by syntactic typing
+case-by-case through a series of lemmas.
+
+\begin{lemma}[ST-Var]
+  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= i : up S i G i]]$.
+\end{lemma}
+
+\begin{lemma}[ST-Set]
+  If $[[|= G]]$ and $[[i < j]]$, then $[[G |= Set i : Set j]]$.
+\end{lemma}
+
+\begin{lemma}[ST-Pi]
+  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
+  A B : Set i]]$.
+\end{lemma}
 

### 2023-12-30 paper/paper.tex
Type set the inversion lemma for set

@@
 
+\begin{lemma}[Set Inv]
+  \label{lemma:setinv}
+  The following two statements are equivalent:
+  \begin{itemize}
+  \item $[[G |= A : Set i]]$
+  \item For all $[[rho]]$, if $[[rho |= G]]$, then there exists
+    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  The forward direction is immediate by
+  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
+  and show that $[[G |= A : Set i]]$ given the second bullet.
+
+  Suppose $[[rho |= G]]$, then we know that there exists some $[[S]]$
+  such that $[[InterpR i (A {rho}) S]]$. By the definition of semantic
+  typing, it suffices to show that there exists some $[[j]]$ and
+  $[[S0]]$ such that  $[[InterpR j Set i S0]]$ and $[[A {rho} in
+  S0]]$.
+  Pick $[[S i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
+  ]]$ for $[[S0]]$ and it's trivial to verify the conditions hold.
+\end{proof}
+
 \begin{lemma}[ST-Var]
@@
 \begin{lemma}[ST-Set]
-  If $[[|= G]]$ and $[[i < j]]$, then $[[G |= Set i : Set j]]$.
+  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
 \end{lemma}
+\begin{proof}
+  Immedaite by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
+\end{proof}
 
@@
 \end{lemma}
+\begin{proof}
+  Suppose $[[rho |= G]]$. Applying
+% Our goal is to show that there exists some
+%   $[[S]]$ such that $[[InterpR i Pi A B S ]]$.
+%   From the premise $[[G |= A : Set i]]$, we know that there exists
+%   some $[[S0]]$ such that
+\end{proof}
 

### 2023-12-31 paper/paper.tex
Remove stlc review section

@@
 The key technique that underlies our consistency proof is proof by
-logical relation. In Section~\ref{sec:logrelstlc}, we review the
-technique in the context of the simply typed lambda calculus. In
+logical relation. In
 Section~\ref{sec:spec}, we present the dependent type theory of
@@
 type theory.
-
-\section{Review of Proof by Logical Relation}
-\label{sec:logrelstlc}
-In this section, we give a quick review of the consistency proof of the
-simply typed lambda calculus augmented with the empty type through
-proof by logical relation.
-
-We give the full specification of the simply typed calculus in
-Figure~\ref{fig:stlcsyntax},\ref{fig:stlctyping},\ref{fig:stlcdynamics}
-without any further explanation since it is standard.
-
-The goal here is to show that the empty type cannot be
-inhabited. A naive attempt at proof by structural induction
 
@@
 \begin{proof}
-  Suppose $[[rho |= G]]$. Applying
-% Our goal is to show that there exists some
-%   $[[S]]$ such that $[[InterpR i Pi A B S ]]$.
-%   From the premise $[[G |= A : Set i]]$, we know that there exists
-%   some $[[S0]]$ such that
+  Applying Lemma~\ref{lemma:setinv} to the
+  conclusion, it now suffices to show that given $[[rho |= G]]$, there
+  exists some $[[S]]$ such that $[[InterpR i Pi A B S]]$.
+  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
+  there exists some set $[[S0]]$ such that $[[InterpR i A S0]]$.
+  Let $[[F]]$ be the following relation:
+  \[ (a, S) \in [[F]] \iff [[InterpR i B {rho .: a} S]] \]
+  By Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a partial function with
+  the signature $[[STm -> PowerSet STm]]$.
+
+We
+  instantiate $[[S0]]$ with the set $\{ b \mid \forall [[ a in S0]],
+  \forall [[S1]], [[InterpR i B {a} S1]] \implies [[b a in S1]] \}$.
 \end{proof}

### 2023-12-31 paper/paper.tex
Finish rule ST-Pi

@@
   \[ (a, S) \in [[F]] \iff [[InterpR i B {rho .: a} S]] \]
-  By Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a partial function with
-  the signature $[[STm -> PowerSet STm]]$.
-
-We
-  instantiate $[[S0]]$ with the set $\{ b \mid \forall [[ a in S0]],
-  \forall [[S1]], [[InterpR i B {a} S1]] \implies [[b a in S1]] \}$.
+  By Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a partial
+function. From $[[G ++ A |= B : Set i]]$, we know that there must
+exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
+in S0]]$. As a result, $[[F]]$ is defined on every element of $[[S0]]$.
+We can now instantiate  $[[S]]$ with the set $[[ { b | forall a, (# a
+  in S0 implies b a in F a #) }]]$. The goal $[[InterpR i Pi A B S]]$
+ immediately follows from  \rref{I-Pi} and the definition of $[[F]]$.
 \end{proof}

### 2023-12-31 paper/paper.tex
Add the alt intro inv rules for pi

@@
 
+Lemma~\ref{lemma:logreldeter} enables us to derive the following
+alternative introduction and inversion principles for \rref{I-Pi}.
+\begin{lemma}[Pi Intro Alt]
+  \label{lemma:piintroalt}
+  Suppose the following two statements hold:
+  \begin{itemize}
+  \item $[[Interp I i A S]]$
+  \item $[[forall a, a in S implies (# exists S0 , Interp I i B {a} S0 #)]]$
+  \end{itemize}
+  Then we have $[[Interp I i Pi A B { b | forall a, (# a in S , forall
+    S0, (# Interp I i B {a} S0,  b a in F a #) #) }]]$
+\end{lemma}
+\begin{proof}
+  Let $[[F]]$ be the relation defined as follows:
+  \[ (a,S) \in [[F]] \iff [[Interp I i B {a} S]] \]
+  By the second bullet from the premise and
+  Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a function that is total
+  on the set $[[S]]$. The conclusion then trivially follows from
+  \rref{I-Pi}.
+\end{proof}
+
+\begin{lemma}[Pi Inv Alt]
+  \label{lemma:piinvalt}
+  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
+  such that the following constraints hold:
+  \begin{itemize}
+  \item $[[Interp I i A S0]]$
+  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
+    S1 #) #)]]$
+  \item $[[S = { b | forall a, (# a in S0 , forall
+    S1, (# Interp I i B {a} S1,  b a in F a #) #) }]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  Immedaite from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
+\end{proof}
+
+% The combination of Lemmas~\ref{lemma:piintroalt} and
+% \ref{lemma:piinvalt} reveals what we truly want to capture with
+% \rref{I-Pi}. A function type is semantically well-defined if its input
+% type is well-defined, and its output type is defined for each valid
+% inahbitant of its input type.
+
 The next lemma shows that our logical relation satisfies

### 2023-12-31 paper/paper.tex
Minor

@@
 
+The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
+counterpart $[[|- G]]$, is defined through a for all quantified statement rather
+than inductively over the context. It is easy to recover the same
+structural rules:
+\begin{lemma}[TODO]
+
+\end{lemma}
+
+Likewise, the semantic well-formendness judgment for substitutions
+satisfies similar structural rules.
+\begin{lemma}[Well-formed $[[rho]]$ cons]
+  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
+  $[[rho .: a |= G ++ A]]$.
+\end{lemma}
+
 \begin{lemma}[ST-Var]
@@
   there exists some set $[[S0]]$ such that $[[InterpR i A S0]]$.
-  Let $[[F]]$ be the following relation:
-  \[ (a, S) \in [[F]] \iff [[InterpR i B {rho .: a} S]] \]
-  By Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a partial
-function. From $[[G ++ A |= B : Set i]]$, we know that there must
+From $[[G ++ A |= B : Set i]]$, we know that there must
 exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
-in S0]]$. As a result, $[[F]]$ is defined on every element of $[[S0]]$.
-We can now instantiate  $[[S]]$ with the set $[[ { b | forall a, (# a
-  in S0 implies b a in F a #) }]]$. The goal $[[InterpR i Pi A B S]]$
- immediately follows from  \rref{I-Pi} and the definition of $[[F]]$.
+in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
 \end{proof}

### 2024-01-08 paper/paper.tex
Finish the fundamental theorem

@@
 fundamental lemma is established, logical consistency follows as a
-trivial corollary.
-
-The downside of a pen and paper proof is that the validity of the
-definitions themselves can be dubious at times. When working with a
-dependent type theory that has a universe hierarchy, an intuitive
-understanding of naive set theory no longer suffices as the size of
-sets matters. For example, the partial function representing the
-interpretation of types does not immediately correspond to a valid
-definition in a proof assistant. We cannot easily define it as a
-function as proof assistants such as Coq and Agda only accept total
-functions. We cannot naively encode the interpretation function as an
-inductive relation either since we violate the strict positivity
-constraint. A detailed discussion of how we work around such technical
-issues can be distracting. However, what can be equally distracting is
-following along a proof where even the validity of the statement is
-not trusted. As a result, we clearly mark the ``suspicious''
-definitions, and defer the discussion of these definitions
-until Section~\ref{sec:logrelmech}, where we give an overview of our
-short Coq mechanization proof and the various proof engineering
-techniques we adopt to make the mechanization for tractable.
-
+trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
+specifics related to the Coq mechanization of the proof described in earlier
+section.
 Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
@@
 \mathit{Natural\ numbers}\\
-[[i]],[[j]],[[n]] & ::= & [[0]]\ |\ [[S i]] & \mbox{zero, successor} \\ \\
+[[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
 
@@
             & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ |\  [[if a b0 b1]]
-                  & \mbox{boolean type, true, false, if}
+                  & \mbox{boolean type, true, false, if} \\ \\
+\mathit{Substitution}\\
+[[rho]] & \in & [[SNat -> STm]] &
 \end{array}
@@
 immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.
+
+The reflexivity of parallel reduction enables us to embed rules from
+call-by-name semantics into parallel reduction, as the following lemma
+shows.
+\begin{lemma}[Par AbsCbn] For all $[[A]], [[a]],$ and $[[b]]$,
+  \label{lemma:parabscbn}
+  $[[(\ A a) b => b {a}]]$
+\end{lemma}
+\begin{proof}
+  Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
+\end{proof}
 
@@
   \item $[[S = { b | forall a, (# a in S0 , forall
-    S1, (# Interp I i B {a} S1,  b a in F a #) #) }]]$
+    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
   \end{itemize}
@@
       [[rho |= G]] &:= \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
-                     } [[InterpR j (up S i G i) { rho } S ]] \text{, then } [[rho i in S]] \\
+                     } [[InterpR j (up Suc i G i) { rho } S ]] \text{, then } [[rho i in S]] \\
       [[G |= a : A]] &:= \forall [[rho]], \text{ if }[[rho |=
@@
                        {rho} S]] \text{ and } [[a {rho} in S]] \\
-      [[|= G]] &:= \forall [[i < |G|]], \exists [[j]], [[drop S i G |= G i : Set j]]
+      [[|= G]] &:= \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
     \end{split}
@@
 
-Next, we show that semantic typing is implied by syntactic typing
-case-by-case through a series of lemmas.
-
+The following lemma makes the statement $[[G |= A : Set i]]$ easier to
+work with.
 \begin{lemma}[Set Inv]
@@
 structural rules:
-\begin{lemma}[TODO]
-
+\begin{lemma}[Semantic context well-formedness cons]
+  \label{lemma:semwffcons}
+  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
 \end{lemma}
+\begin{proof}
+  By the definition of semantic context well-formedness, the goal is
+  to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i
+  (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
+  proven by case analysis on whether $[[i]]$ is zero.
+\end{proof}
 
@@
 \end{lemma}
+\begin{proof}
+  Start by unfolding the definition of $[[rho .: a |= G ++ A]]$ and
+  performing a case analysis similar to the proof of
+  Lemma~\ref{lemma:semwffcons}.  The case where the number is $[[0]]$
+  requires Lemma~\ref{lemma:logreldeterhet} to finish the proof.
+\end{proof}
 
+Next, we show some non-trivial cases of the fundamental theorem as
+top-level lemmas and leave the remaining cases as exercises for the reader.
 \begin{lemma}[ST-Var]
-  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= i : up S i G i]]$.
+  \label{lemma:stvar}
+  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= i : up Suc i G i]]$.
 \end{lemma}
+\begin{proof}
+  TODO: formulate renaming
+\end{proof}
 
 \begin{lemma}[ST-Set]
+  \label{lemma:stset}
   If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
@@
 \begin{lemma}[ST-Pi]
+  \label{lemma:stpi}
   If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
@@
   conclusion, it now suffices to show that given $[[rho |= G]]$, there
-  exists some $[[S]]$ such that $[[InterpR i Pi A B S]]$.
+  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
   From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
-  there exists some set $[[S0]]$ such that $[[InterpR i A S0]]$.
+  there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
 From $[[G ++ A |= B : Set i]]$, we know that there must
@@
 in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
+\end{proof}
+
+\begin{lemma}[ST-Abs]
+  \label{lemma:stabs}
+  If $[[G |= Pi A B : Set i]]$ and $[[G ++ A |= b : B]]$, then $[[G |=
+  \ A b : Pi A B]]$.
+\end{lemma}
+\begin{proof}
+  By unfolding the definition of $[[G |= \ A b : Pi A B]]$, we need to
+  show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
+  $[[S]]$ such that $[[InterpR i Pi A {rho} B {up rho} S]]$ and $[[\
+  A{rho} b{up rho} in S]]$.
+
+  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
+  i]]$, there exists some set $[[S]]$ such that
+  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
+  $[[\A{rho} b{up rho} in S
+  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
+  that all following conditions hold:
+  \begin{itemize}
+  \item $[[InterpR i A{rho} S0]]$
+  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
+    {rho .: a}
+    S1 #) #)]]$
+  \item $[[S = { b | forall a, (# a in S0 , forall
+      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
+  \end{itemize}
+  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
+  that given $[[a in S0]]$,
+  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
+  a in S1]]$.
+  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
+  under expansion. By Lemma~\ref{lemma:parabscbn}, since $[[( \A{rho} b{up rho} )
+  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
+  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
+  B]]$ and the fact that the logical relation is deterministic and
+  cumulative (Lemma~\ref{lemma:logreldeterhet}).
+\end{proof}
+
+\begin{lemma}[ST-App]
+  \label{lemma:stapp}
+  If $[[G |= b : Pi A B]]$ and $[[G |= a : A]]$, then $[[G |= b a : B {a}]]$.
+\end{lemma}
+\begin{proof}
+Suppose $[[rho |= G]]$. The goal is to show that there exists some
+$[[i]]$ and $[[S1]]$
+such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a} {rho}
+S1]]$, or equivalently, $[[InterpR i B {rho .: a {rho}} S1]]$ since
+$[[B {a}{rho}]] = [[B {rho .: a {rho}}]]$. By the premise $[[G |= b :
+Pi A B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
+there exists some $[[i]]$ and $[[S0]]$ such that:
+  \begin{itemize}
+  \item $[[InterpR i A{rho} S0]]$
+  \item $[[forall a0, (# a0 in S0 implies (# exists S1 , InterpR i B
+    {rho .: a0}
+    S1 #) #)]]$
+  \item $[[forall a0, (# a0 in S0 , forall
+      S1, (# InterpR i B {rho .: a0} S1,  b {rho} a0 in S1 #) #)]]$
+  \end{itemize}
+  Instantiating the variable $[[a0]]$ from the last two bullets with
+  the term $[[a {rho}]]$, the conclusion immediately follows.
+\end{proof}
+
+\begin{theorem}[The Fundamental Theorem]
+  \label{theorem:soundness}
+  If $[[G |- a : A]]$, then $[[G |= a : A]]$.
+\end{theorem}
+\begin{proof}
+  Proof by structural induction over the derivation of $[[G |- a :
+  A]]$. Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
+  can be used to discharge their syntactic counterpart
+  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
+  cases not covered by the lemmas are similar to the ones already
+shown or simpler and therefore omitted from the text.
 \end{proof}

### 2024-01-09 paper/paper.tex
Add auxiliary functions

@@
 \usepackage{xspace}
+\usepackage[para]{footmisc}
+\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
 \newcommand{\lang}{$\lambda^H$\xspace}
@@
 typing judgment for the set of lambda terms.
-In Section~\ref{sec:logrelproof}, we prove the fundamental lemma,
+In Section~\ref{sec:logrelproof}, we prove the fundamental theorem,
 which states that syntactic typing implies semantic typing. Once the
-fundamental lemma is established, logical consistency follows as a
+fundamental theorem is established, logical consistency follows as a
 trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
@@
                   & \mbox{boolean type, true, false, if} \\ \\
+\mathit{Renaming}\\
+[[xi]] & \in & [[SNat -> SNat]] & \\ \\
 \mathit{Substitution}\\
@@
   \label{fig:syntax}
+\end{figure}
+
+\begin{figure}[h]
+    \[
+      \begin{array}{lll}
+        \mathit{Curried\ Addition} \\
+        add(n) & := & m \mapsto n + m \\ \\
+        \mathit{Extension ([[xi]])} \\
+        [[(xi .: j) 0]]  & := & 0 \\
+        [[(xi .: j) Suc i]]  & := & [[xi j]] \\ \\
+
+        \mathit{Extension ([[rho]])} \\
+        [[(rho .: a) 0]]  & := & [[a]] \\
+        [[(rho .: a) Suc i]]  & := & [[rho i]] \\ \\
+
+        \mathit{Up ([[xi]])} \\
+        [[up xi]] & := & (add(1) \circ [[xi]]) , 0 \\ \\
+
+        \mathit{Renaming} \\
+        [[i {xi}]] & := & [[xi i]] \\
+        [[(Set i) {xi}]] & := & [[Set i]] \\
+        [[Void {xi}]] & := & [[Void]]\\
+        [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
+        [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
+        [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
+        [[Bool {xi}]] & := & [[Bool]] \\
+        [[true {xi}]] & := & [[true]] \\
+        [[false {xi}]] & := & [[false]] \\
+        [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
+        [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
+        [[refl {xi}]] & := & [[refl]] \\
+        [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\
+
+        \mathit{Shift} \\
+        [[up n A]] & := & A \langle add(n) \rangle \\ \\
+
+        \mathit{Shift\ one} \\
+        [[up1 A]] & := & \uparrow^1 [[A]] \\ \\
+
+        \mathit{Lookup} \\
+        [[(G ++ A) 0]] & := &  [[A]] \\
+        [[(G ++ A) Suc i]] & := & [[G i]] \\ \\
+
+        \mathit{Drop} \\
+        [[drop 0 G]] & := & [[G]] \\
+        [[drop Suc i (#G ++ A#)]] & := & [[drop i G]] \\ \\
+      \end{array}
+    \]
+  \caption{Auxiliary Functions over Syntax}
+  \label{fig:auxdef}
 \end{figure}
@@
 
-The syntax of \lang can be found in Figure~\ref{fig:syntax}. As a
+The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
+the unscoped DeBruijn representation to leave less ambiguity for the
+formulation of the system and simplify the reasoning about binders.
+Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
+syntax, including renaming, substitution, and operations over
+the typing context or substitution.
+As a
 dependent type theory, terms and types are collapsed into the same
@@
 represent its transitive closure, which in turn allows us to define
-the coherence relation $[[a <=> b]]$. % The $*$ symbol is more commonly
-% used for the transitive and reflexive closure of a relation. However,
-% since we can show that parallel reduction is reflexive, the transitive
-% closure of parallel reduction must also be reflexive.
+the coherence relation $[[a <=> b]]$.
 We say that two terms $[[a]]$
@@
 sequence of lemmas.
-\begin{lemma}[Par Refl]
+\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
   \label{lemma:parrefl}
@@
 
-\begin{lemma}[Coherence refl]
+\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
   \label{lemma:coherencerefl}
@@
 shows.
-\begin{lemma}[Par AbsCbn] For all $[[A]], [[a]],$ and $[[b]]$,
+\begin{lemma}[Par AbsCbn\footnote{\dotv{join.v}{P\_AppAbs\_cbn}}] For all $[[A]], [[a]],$ and $[[b]]$,
   \label{lemma:parabscbn}
@@
 Symmetry of coherence immediately falls from its definition.
-\begin{lemma}[Coherence sym]
+\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
   \label{lemma:coherencesym}
@@
 reduction satisfies the diamond property.
-\begin{lemma}[Par cong]
+\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
   \label{lemma:parcong}
@@
 \end{lemma}
-\begin{lemma}[Par diamond]
+\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
   \label{lemma:pardiamond}
@@
 single substitution property as a simple corollary.
-\begin{corollary}[Par subst]
+\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
   \label{lemma:parsubst}
@@
 and reflexive closure is confluent.
-\begin{lemma}[Par confluent]
+\begin{lemma}[Par confluent\footnote{\dotv{join.v}{pars\_confluent}}]
   \label{lemma:parconfluent}
@@
 Lemma~\ref{lemma:parconfluent}.
-\begin{lemma}[Coherence trans]
+\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
   \label{lemma:coherencetrans}
@@
 reduction as an auxilliary definition to derive the confluence of full
-beta reduction. The formulation of \lang is slightly more economical.
+beta reduction. Our formulation of \lang through parallel reduction
+is slightly more economical.
 
@@
 For the rest of the section, we establish some important facts about
-the logical relation that will be useful for proving fundamental lemma
+the logical relation that will be useful for proving fundamental theorem
 in Section~\ref{sec:logrelproof}.
@@
   \begin{enumerate}
-  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
-  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
-  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
-  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
+  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}}
+  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}}
+  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}}
+  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
     \begin{itemize}
@@
     \end{itemize}
-  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
+  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}}
   \end{enumerate}
@@
 usual forward direction too.
-\begin{lemma}[Preservation of the logical relation]
+\begin{lemma}[Preservation of the logical relation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
   \label{lemma:interppreservation}
@@
 interpretation.
-\begin{corollary}[Logical Relation for Coherent Types]
+\begin{corollary}[Logical Relation for Coherent Types\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
   \label{lemma:logrelcoherence}
@@
 Next, we show that $[[Interp I i A S]]$ is in fact a partial function.
-\begin{lemma}[Logical relation is functional]
+\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
   \label{lemma:logreldeter}
@@
 alternative introduction and inversion principles for \rref{I-Pi}.
-\begin{lemma}[Pi Intro Alt]
+\begin{lemma}[Pi Intro Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
   \label{lemma:piintroalt}
@@
 
-\begin{lemma}[Pi Inv Alt]
+\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
   \label{lemma:piinvalt}
@@
 universe level.
-\begin{lemma}[Logical relation cumulativity]
+\begin{lemma}[Logical relation cumulativity\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
   \label{lemma:logrelcumulativity}
@@
 
-\begin{corollary}[Logical relation is functional with different levels]
+\begin{corollary}[Logical relation is functional with different levels\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
   \label{lemma:logreldeterhet}
@@
 all closed under expansion.
-\begin{lemma}[Logical Relation Elements back preservation]
+\begin{lemma}[Logical Relation Elements back preservation\footnote{\dotv{semtyping.v}{InterpExt\_back\_clos}}]
   \label{lemma:logrelbackclos}
@@
 
-\begin{corollary}
+\begin{corollary}[Logical Relation Elements back preservation (rec)\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
   \label{lemma:logrelNbackclos}
@@
 work with.
-\begin{lemma}[Set Inv]
+\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}]
   \label{lemma:setinv}
@@
 satisfies similar structural rules.
-\begin{lemma}[Well-formed $[[rho]]$ cons]
+\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
   If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
@@
 
-\begin{theorem}[The Fundamental Theorem]
+\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
   \label{theorem:soundness}
-  If $[[G |- a : A]]$, then $[[G |= a : A]]$.
+  \begin{itemize}
+  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
+  \item If $[[|- G]]$, then $[[|= G]]$.
+  \end{itemize}
 \end{theorem}
 \begin{proof}
-  Proof by structural induction over the derivation of $[[G |- a :
-  A]]$. Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
+  Proof by mutual induction over the derivation of $[[G |- a :
+  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
+  from Lemma~\ref{lemma:semwffcons}.
+Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
   can be used to discharge their syntactic counterpart
@@
   shown or simpler and therefore omitted from the text.
+
+\end{proof}
+
+\begin{corollary}[Logical Consistency]
+  \label{corollary:consistency}
+  The judgment $[[empty |- a : Void ]]$ is not derivable.
+\end{corollary}
+\begin{proof}
+  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
 \end{proof}
@@
 \label{sec:logrelmech}
+The Coq mechanization follows an almost identical structure as the
+proofs presented in Section~\ref{sec:logrelproof}. The inductive
+definition of the logical relation in Figure~\ref{fig:logrel} requires
+the impredicativity of Coq's \texttt{Prop} sort since in \rref{I-Pi},
+the function $[[F]]$ can be later instantiated into the logical
+relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
 

### 2024-02-09 paper/paper.tex
Add more stuff to the intro

@@
 consistency proof for a dependently typed language that contains a
-rich set of features, including an infinite and cumulative universe
-hierarchy, natural numbers, and an intensional identity type. We have
+rich set of features, including an infinite cumulative universe
+hierarchy, booleans, and an intensional identity type. We have
 fully mechanized the consistency proof using the Coq proof assistant
@@
 % during evaluation.
-When a type system is used as a program logic where terms encode proofs, we
-want our type system to be logically consistent, meaning
-that the empty type is not inhabited.
+When a dependently type system is used as a program logic where terms
+encode proofs, we want our type system to be logically consistent,
+meaning that the empty type is not inhabited.
+
+The consistency proofs of various dependently typed systems, including Martin-Lof's
+type theory and the Calculus of Constructions, have long been
+available in the literature. In particular, recent works such as \citet{nbeincoq},
+\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
+results related to the decidability of type conversion or type
+checking of dependently typed systems, from which consistency can be
+derived as a corollary. The underlying technique of the forementioned
+works is proof by logical relation,
+which involves interpreting types as reducibility predicates,
+representing sets of terms satisfying certain properties with respect
+to the reduction relation. While the proof technique and the
+consistency result for dependent types are both well-established,
+there is a severe lack of rigorous and accessible material that shows
+how proof by logical relation can be applied to dependently typed
+systems.
+Introductory materials about logical relations such as
+\citet{skorstengaard2019introduction}, \citet{harper2016practical}
+start from logical relations for simply typed languages, and
+eventually extend the technique to build a relational model for System
+F in order to derive
+parametricity. \citet{harperkripke} gives a gentle introduction on how
+logical relations for closed terms can be extended to talk about open
+terms so we can derive properties such as normalization for well-typed
+open terms. In short, the introductory texts about logical relations
+cover systems and properties with varying degrees of complexity, from
+simply typed to polymorphicly typed, logical predicate to logical
+equivalence, closed terms to open terms. The glaring gap here is the
+lack of fully dependently typed systems where computations may appear
+at the type-level. In fact, it is far from obvious why proof by
+logical relation is even applicable to dependent types, since the type
+may very well-be a computation that is yet to be evaluated.
+While it is assuring that proof by logical relations for dependent
+types is available in mechanized forms, % the key to address the
+% complexities of dependent types is obscured in
+\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
+Kripke-style models that obscure the underlying technique of applying
+proof by logical relation to dependent types. In fact, the
+complexivity is self-evident from the size of their code base. For
+example, \citet{nbeincoq} involves 22 thousand lines of Coq code.
+
+The goal of this paper is to give a tutorial on proof by logical relation for
+dependent types in a simple and digestable format. % The language specification will
+% be given in Section~\ref{sec:spec}.
+The system we present is small but not minimal. We want to cover
+features that would require more interesting treatments under
+dependent types without introducing complexities that would have
+existed in its simply typed variant. We choose boolean
+types over natural numbers as our base type for simplicity, but include an
+intensional identity type in our type system as the interpretation
+of identity types requires extra side conditions that are not present
+in other types. Unlike \citet{nbeincoq,decagda,martin-lof-a-la-coq}
+but similar to \citet{anand2014towards}, we include an infinite
+hierarchy of universes since we want to not only have at least one
+predicative universe level to support type-level computations, but
+also avoid the unnecessary code duplication pointed out by
+\citet{nbeincoq}.
+We choose consistency as the property of interest since it is the one
+of the easiest metatheoretic property that requires proof by logical
+relation.
+
+Our consistency proof is short and fully mechanized. The proof scripts
+involve less than 1000 lines of Coq code. Furthermore, the mechanized
+proof closely corresponds to the informal
+
+% A naive attempt at proving consistency through induction over the
+% typing derivation would fail since the inductive hypothesis is not
+% strong enough to derive the consistency result. Instead, one typically
+% relies on the technique referred to as proof by logical relation to
+% interpret types as reducibility predicates to strengthen the inductive
+% hypothesis.
+
 This paper is specifically
@@
 system with an infinite universe hierarchy and support for large
-elimination. % The type system, presented in Section~\ref{sec:spec}, is
+elimination . % The type system, presented in Section~\ref{sec:spec}, is
 % most similar to Martin-Lof's predicate type theory with the minor
@@
 
-The proof of various dependently typed systems, including Martin-Lof's
-type theory, has long been
-established in the literature and therefore the consistency result of
-this paper is not novel. Rather,
+Rather,
 our goal is to present the proof in a form that is digestable by a

### 2024-02-10 paper/paper.tex
Minor

@@
 %%
-\documentclass[nonacm]{acmart}
+\documentclass[acmsmall]{acmart}
 \usepackage{ottalt}
@@
 checking of dependently typed systems, from which consistency can be
-derived as a corollary. The underlying technique of the forementioned
+derived as a corollary.
+
+The underlying technique of the forementioned
 works is proof by logical relation,
@@
 systems.
+
 Introductory materials about logical relations such as
@@
 simply typed to polymorphicly typed, logical predicate to logical
-equivalence, closed terms to open terms. The glaring gap here is the
+equivalence, closed terms to open terms.
+
+The glaring gap here is the
 lack of fully dependently typed systems where computations may appear
@@
 \citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
-Kripke-style models that obscure the underlying technique of applying
-proof by logical relation to dependent types. In fact, the
-complexivity is self-evident from the size of their code base. For
-example, \citet{nbeincoq} involves 22 thousand lines of Coq code.
+Kripke-style models that obscure the technique for addressing
+type-level computation. In fact, the added complexivity is evident
+from the size of their code base. For example, \citet{nbeincoq}
+involves 22 thousand lines of Coq code.
 
@@
 Our consistency proof is short and fully mechanized. The proof scripts
-involve less than 1000 lines of Coq code. Furthermore, the mechanized
-proof closely corresponds to the informal
+involve less than 1000 lines of Coq code. In fact, what is even more
+encouraging is that among the 1000
+lines of Coq code, 400 lines are related to the specification of the
+type system, semantics, and properties related to untyped lambda
+terms. The semantic type soundness proof through logical relation
+takes almost the same amount of code as our syntactic type soundness
+proof!
+Thanks to the conciseness of
+the proof, we are able to present most of its details in
+Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
+structure of our mechanization closely corresponds to the proof we
+present in the text, enabling us to label at the footnote each lemma
+directly to their counterpart in the proof script for the readers to
+reference and validate.
+
+The technique we use is most similar to the one from \citet{nbeincoq},
+which leverages impredicativity to define the logical
+relation. % In our presented proof, we use the informal
+% language of sets, where the use of impredicativity feels more natural
+% and less noticeable.
+Rather than framing impredicativity as a
+mechanism for encoding induction
+recursion~\citep{induction-recursion}, a scheme in which logical
+relation for dependent types is typically defined~\citep{todo}, we opt
+for a direct explanation in informal language of sets, where
+impredicativity manifests in the form of second-order logical
+formulas and thus more intuitive to understand for readers with some
+general mathematics background.
+
+In Section~\ref{sec:logrelmech}, we discuss the details about our
+mechanization, including our use of existing libraries such as
+Autosubst for handling bindings and CoqHammer for general-purpose
+automation. In Section~\ref{sec:discuss}, we show that extensions such
+as $\eta$ rules for functions and normalization for open terms are
+more or less orthogonal to the fact that the language is dependently
+typed. In our mechanized proof of said extensions, the overall
+structure of the logical relation is preserved, and the additional
+proof obligations are mostly related to properties about the untyped
+lambda calculus, all of which can be derived through syntactic
+means.
+In other words, once we have established the base technique for
+proof by logical relation for dependent types, we can factor out the
+complexities of an extension the are not specifically related to dependent
+types; such an extension, if desired, may be studied in the context of
+a simply typed language and later ported into a dependently typed setting.
+
+We hope our success at creating a short and mechanized proof for a
+relatively feature-complete language will encourage future researchers
+to leverage the tool of logical relation more often in mechanized
+proofs for dependent types.
 
@@
 
-This paper is specifically
-about establishing logical consistency for a fully dependently typed
-system with an infinite universe hierarchy and support for large
-elimination . % The type system, presented in Section~\ref{sec:spec}, is
+% This paper is specifically
+% about establishing logical consistency for a fully dependently typed
+% system with an infinite universe hierarchy and support for large
+% elimination .
+% The type system, presented in Section~\ref{sec:spec}, is
 % most similar to Martin-Lof's predicate type theory with the minor
@@
 
-Rather,
-our goal is to present the proof in a form that is digestable by a
-working type theorist and can be more readily mechanized in a proof
-assistant. Compared to existing efforts at mechanizing logical
-consistency or stronger properties such as existence of normal
-form~\citep{nbeincoq},
-decidable type checking~\citep{decagda}, our work is minimal since it requires very
-little scaffolding and therefore results in an extremely succinct
-proof of under 1000 lines of manually written Coq code for a dependent
-type theory that is reasonably complete in its features.
-
-The key technique that underlies our consistency proof is proof by
-logical relation. In
-Section~\ref{sec:spec}, we present the dependent type theory of
-interest. In Section~\ref{sec:logreldep}, we give the definition
-of the logical relation for the dependent type theory. Rather than
-presenting the logical relation as an inductive-recursive definition,
-we use the more elementary concept of a partial function to capture
-the interpretation of types. The alternative representation requires us
-to show that the set of equations indeed defines a partial function;
-that is, for each input, there should always be a unique
-output.
-From the interpretation function, we can define the semantic
-typing judgment for the set of lambda terms.
-In Section~\ref{sec:logrelproof}, we prove the fundamental theorem,
-which states that syntactic typing implies semantic typing. Once the
-fundamental theorem is established, logical consistency follows as a
-trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
-specifics related to the Coq mechanization of the proof described in earlier
-section.
-Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
-existing literature related to logical consistency about dependent
-type theory.
+% Rather,
+% our goal is to present the proof in a form that is digestable by a
+% working type theorist and can be more readily mechanized in a proof
+% assistant. Compared to existing efforts at mechanizing logical
+% consistency or stronger properties such as existence of normal
+% form~\citep{nbeincoq},
+% decidable type checking~\citep{decagda}, our work is minimal since it requires very
+% little scaffolding and therefore results in an extremely succinct
+% proof of under 1000 lines of manually written Coq code for a dependent
+% type theory that is reasonably complete in its features.
+
+% The key technique that underlies our consistency proof is proof by
+% logical relation. In
+% Section~\ref{sec:spec}, we present the dependent type theory of
+% interest. In Section~\ref{sec:logreldep}, we give the definition
+% of the logical relation for the dependent type theory. Rather than
+% presenting the logical relation as an inductive-recursive definition,
+% we use the more elementary concept of a partial function to capture
+% the interpretation of types. The alternative representation requires us
+% to show that the set of equations indeed defines a partial function;
+% that is, for each input, there should always be a unique
+% output.
+% From the interpretation function, we can define the semantic
+% typing judgment for the set of lambda terms.
+% In Section~\ref{sec:logrelproof}, we prove the fundamental theorem,
+% which states that syntactic typing implies semantic typing. Once the
+% fundamental theorem is established, logical consistency follows as a
+% trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
+% specifics related to the Coq mechanization of the proof described in earlier
+% section.
+% Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
+% existing literature related to logical consistency about dependent
+% type theory.
 
@@
 
-The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
+The syntax of \lang can be found in Figure~\ref{fig:syntax}. To help
+the reader understand the binding structure of the terms, we annotate
+the terms with their DeBruijn depth. For example, ....
+ We use
 the unscoped DeBruijn representation to leave less ambiguity for the

### 2024-02-10 paper/paper.tex
Add more to the intro

@@
 The system we present is small but not minimal. We want to cover
-features that would require more interesting treatments under
-dependent types without introducing complexities that would have
-existed in its simply typed variant. We choose boolean
+features that are only available or require special treatments under
+dependent types%  without introducing complexities that would have
+% existed in its simply typed variant
+. We choose boolean
 types over natural numbers as our base type for simplicity, but include an
 intensional identity type in our type system as the interpretation
-of identity types requires extra side conditions that are not present
-in other types. Unlike \citet{nbeincoq,decagda,martin-lof-a-la-coq}
+of identity types requires an extra side condition that is needed for
+indexed family of types. Unlike \citet{nbeincoq,decagda,martin-lof-a-la-coq}
 but similar to \citet{anand2014towards}, we include an infinite
@@
 \citet{nbeincoq}.
-We choose consistency as the property of interest since it is the one
-of the easiest metatheoretic property that requires proof by logical
-relation.
 
@@
 Autosubst for handling bindings and CoqHammer for general-purpose
-automation. In Section~\ref{sec:discuss}, we show that extensions such
+automation. In Section~\ref{sec:discuss}, we extend our logical
+relation to prove the stronger property that every well-typed term
+has a normal form. Most interestingly, we show that extensions such
 as $\eta$ rules for functions and normalization for open terms are
-more or less orthogonal to the fact that the language is dependently
-typed. In our mechanized proof of said extensions, the overall
-structure of the logical relation is preserved, and the additional
+mostly orthogonal to the fact that the language is dependently
+typed. In our mechanized proof, the overall
+structure of the logical relation remains unchanged, and the additional
 proof obligations are mostly related to properties about the untyped
@@
 
-The syntax of \lang can be found in Figure~\ref{fig:syntax}. To help
-the reader understand the binding structure of the terms, we annotate
-the terms with their DeBruijn depth. For example, ....
- We use
-the unscoped DeBruijn representation to leave less ambiguity for the
-formulation of the system and simplify the reasoning about binders.
-Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
-syntax, including renaming, substitution, and operations over
-the typing context or substitution.
+The syntax of \lang can be found in Figure~\ref{fig:syntax}. We will use
+the unscoped de Bruijn representation for the rest of the paper,
+adopting notations from \citet{autosubst2}, summarized in Figure~\ref{fig:auxdef}.
+Without providing
+the full definition of the renaming and substitution functions, it is
+impossible to tell the binding structure from the syntax
+alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
+with the de Bruijn depth of each term, though we note that the
+syntax we work with is unscoped and the choice does matter when
+we extend our logical relation to open terms in Section~\ref{sec:discuss}.
+
+% Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
+% syntax, including renaming, substitution, and operations over
+% the typing context or substitution.
 As a
@@
 will use coherence as the equational theory for our type conversion
-rule. Now, we prove that coherence is indeed an equivalence relation.
+rule.
+
+We sketch out some key properties about coherence without giving their
+proofs. Our technique for establishing those results is based on
+\citet{takahashi-parallel-reduction} and a modern exposition can be
+found in \citet{plfa22.08}.
+
+Now, we prove that coherence is indeed an equivalence relation.
 

### 2024-02-11 paper/paper.tex
Minor comments

@@
 relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
-represent its transitive closure, which in turn allows us to define
+represent its transitive and reflexive closure, which in turn allows us to define
 the coherence relation $[[a <=> b]]$.
@@
 proofs. Our technique for establishing those results is based on
-\citet{takahashi-parallel-reduction} and a modern exposition can be
-found in \citet{plfa22.08}.
+\citet{takahashi-parallel-reduction} and a modern exposition of the
+same technique can be found in \citet{plfa22.08}.
+
+TODO: remove the lengthy discussion below
 
@@
 
-We note that a more conventional presentation of
-\rref{T-Conv} would instead use full beta reduction as the base for
-the definition of coherence. However, since full beta reduction
-doesn't satisfy the diamond property, one typically needs parallel
-reduction as an auxilliary definition to derive the confluence of full
-beta reduction. Our formulation of \lang through parallel reduction
-is slightly more economical.
+Figure~\ref{fig:typing} gives the full typing rules for
+\lang{}. \Rref{T-Conv} uses the definition of coherence from earlier
+for type conversion. The use of an untyped relation for type
+conversion makes our formulation slightly different from languages
+such as MLTT, where the judgmental equality typically takes the form
+TODO, where terms ... and ... are known to be well-typed. The
+equivalence of such systems and a system that uses untyped equality
+are explored in detail in ...
+
+Without fancy eta laws, it is easy to embed a typed language into an
+untyped language.
+
+% We note that a more conventional presentation of
+% \rref{T-Conv} would instead use full beta reduction as the base for
+% the definition of coherence. However, since full beta reduction
+% doesn't satisfy the diamond property, one typically needs parallel
+% reduction as an auxilliary definition to derive the confluence of full
+% beta reduction. Our formulation of \lang through parallel reduction
+% is slightly more economical.
 

### 2024-02-12 paper/paper.tex
Add more text

@@
                   & \mbox{equality types, reflexivity proof, J eliminator} \\
-            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ |\  [[if a b0 b1]]
-                  & \mbox{boolean type, true, false, if} \\ \\
+            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
+                  & \mbox{boolean type, true, false} \\
+            & |   & [[if a b0 b1]]
+                  & \mbox{if} \\ \\
 \mathit{Renaming}\\
@@
 
-\begin{figure}[h]
-    \[
-      \begin{array}{lll}
-        \mathit{Curried\ Addition} \\
-        add(n) & := & m \mapsto n + m \\ \\
-        \mathit{Extension ([[xi]])} \\
-        [[(xi .: j) 0]]  & := & 0 \\
-        [[(xi .: j) Suc i]]  & := & [[xi j]] \\ \\
-
-        \mathit{Extension ([[rho]])} \\
-        [[(rho .: a) 0]]  & := & [[a]] \\
-        [[(rho .: a) Suc i]]  & := & [[rho i]] \\ \\
-
-        \mathit{Up ([[xi]])} \\
-        [[up xi]] & := & (add(1) \circ [[xi]]) , 0 \\ \\
-
-        \mathit{Renaming} \\
-        [[i {xi}]] & := & [[xi i]] \\
-        [[(Set i) {xi}]] & := & [[Set i]] \\
-        [[Void {xi}]] & := & [[Void]]\\
-        [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
-        [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
-        [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
-        [[Bool {xi}]] & := & [[Bool]] \\
-        [[true {xi}]] & := & [[true]] \\
-        [[false {xi}]] & := & [[false]] \\
-        [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
-        [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
-        [[refl {xi}]] & := & [[refl]] \\
-        [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\
-
-        \mathit{Shift} \\
-        [[up n A]] & := & A \langle add(n) \rangle \\ \\
-
-        \mathit{Shift\ one} \\
-        [[up1 A]] & := & \uparrow^1 [[A]] \\ \\
-
-        \mathit{Lookup} \\
-        [[(G ++ A) 0]] & := &  [[A]] \\
-        [[(G ++ A) Suc i]] & := & [[G i]] \\ \\
-
-        \mathit{Drop} \\
-        [[drop 0 G]] & := & [[G]] \\
-        [[drop Suc i (#G ++ A#)]] & := & [[drop i G]] \\ \\
-      \end{array}
-    \]
-  \caption{Auxiliary Functions over Syntax}
-  \label{fig:auxdef}
-\end{figure}
-
-\begin{figure}[h]
-\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse, Bool, Eq, Refl, J, JRefl}
-\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{One, Step}
-\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
-\caption{Parallel reduction and coherence}
-\label{fig:par}
-\end{figure}
-
-\begin{figure}[h]
-\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
-\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Refl, J, Bool, True, False, If, Void}
-\caption{Syntactic typing for \lang}
-\label{fig:typing}
-\end{figure}
+% \begin{figure}[h]
+%     \[
+%       \begin{array}{lll}
+%         \mathit{Curried\ Addition} \\
+%         add(n) & := & m \mapsto n + m \\ \\
+%         \mathit{Extension ([[xi]])} \\
+%         [[(xi .: j) 0]]  & := & 0 \\
+%         [[(xi .: j) Suc i]]  & := & [[xi j]] \\ \\
+
+%         \mathit{Extension ([[rho]])} \\
+%         [[(rho .: a) 0]]  & := & [[a]] \\
+%         [[(rho .: a) Suc i]]  & := & [[rho i]] \\ \\
+
+%         \mathit{Up ([[xi]])} \\
+%         [[up xi]] & := & (add(1) \circ [[xi]]) , 0 \\ \\
+
+%         \mathit{Renaming} \\
+%         [[i {xi}]] & := & [[xi i]] \\
+%         [[(Set i) {xi}]] & := & [[Set i]] \\
+%         [[Void {xi}]] & := & [[Void]]\\
+%         [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
+%         [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
+%         [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
+%         [[Bool {xi}]] & := & [[Bool]] \\
+%         [[true {xi}]] & := & [[true]] \\
+%         [[false {xi}]] & := & [[false]] \\
+%         [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
+%         [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
+%         [[refl {xi}]] & := & [[refl]] \\
+%         [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\
+
+%         \mathit{Shift} \\
+%         [[up n A]] & := & A \langle add(n) \rangle \\ \\
+
+%         \mathit{Shift\ one} \\
+%         [[up1 A]] & := & \uparrow^1 [[A]] \\ \\
+
+%         \mathit{Lookup} \\
+%         [[(G ++ A) 0]] & := &  [[A]] \\
+%         [[(G ++ A) Suc i]] & := & [[G i]] \\ \\
+
+%         \mathit{Drop} \\
+%         [[drop 0 G]] & := & [[G]] \\
+%         [[drop Suc i (#G ++ A#)]] & := & [[drop i G]] \\ \\
+%       \end{array}
+%     \]
+%   \caption{Auxiliary Functions over Syntax}
+%   \label{fig:auxdef}
+% \end{figure}
 
@@
 
+\begin{figure}[h]
+\begin{minipage}{0.9\textwidth}
+\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse,
+  Bool, Eq, Refl, J, JRefl}
+\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{One, Step}
+\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
+\end{minipage}
+\caption{Parallel reduction and coherence}
+\label{fig:par}
+\end{figure}
+
 Figure~\ref{fig:par} shows the definition of the parallel reduction
@@
 term $[[c]]$ through parallel reduction. The symmetric notation of
-coherence suggests that it is an equivalence relation, and in fact, we
-will use coherence as the equational theory for our type conversion
-rule.
-
-We sketch out some key properties about coherence without giving their
-proofs. Our technique for establishing those results is based on
-\citet{takahashi-parallel-reduction} and a modern exposition of the
-same technique can be found in \citet{plfa22.08}.
-
-TODO: remove the lengthy discussion below
-
-Now, we prove that coherence is indeed an equivalence relation.
-
-First, we show that coherence is reflexive through the following
-sequence of lemmas.
+coherence suggests that it is an equivalence relation.% , and in fact, we
+% will use coherence as the equational theory for our type conversion
+% rule.
+
+We sketch out some key properties about parallel reduction and
+coherence without giving their proofs. Our technique for establishing
+those results is based on \citet{takahashi-parallel-reduction} and a
+modern exposition of the same technique can be found in
+\citet{plfa22.08}.
 \begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
@@
 \end{lemma}
-
-\begin{lemma}[Pars refl]
-  \label{lemma:parsrefl}
-  For all terms $[[a]]$, $[[a =>+ a]]$.
-\end{lemma}
-
-\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
-  \label{lemma:coherencerefl}
-  For all terms $[[a]]$, $[[a <=> a]]$.
-\end{lemma}
-
-Lemma~\ref{lemma:parrefl} can be proven by structural induction over
-the term $[[a]]$. Lemmas~\ref{lemma:parsrefl} and \ref{lemma:coherencerefl}
-immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.
-
-The reflexivity of parallel reduction enables us to embed rules from
-call-by-name semantics into parallel reduction, as the following lemma
-shows.
-\begin{lemma}[Par AbsCbn\footnote{\dotv{join.v}{P\_AppAbs\_cbn}}] For all $[[A]], [[a]],$ and $[[b]]$,
-  \label{lemma:parabscbn}
-  $[[(\ A a) b => b {a}]]$
-\end{lemma}
-\begin{proof}
-  Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
-\end{proof}
-
-Symmetry of coherence immediately falls from its definition.
-\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
-  \label{lemma:coherencesym}
-  If $[[a <=> b]]$, then $[[b <=> a]]$.
-\end{lemma}
-
-Before we can prove transitivity, we need to show that parallel
-reduction satisfies the diamond property.
 \begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
@@
 \end{lemma}
-The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
-structural induction over the derivation of $[[a0 => a1]]$.
-Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
-over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
-the use of Lemma~\ref{lemma:parcong}.
-
-From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
-single substitution property as a simple corollary.
-\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
-  \label{lemma:parsubst}
-  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
-\end{corollary}
-
-A relation that satisfies the
-diamond property must also be confluent, meaning that its transitive
-and reflexive closure is confluent.
-\begin{lemma}[Par confluent\footnote{\dotv{join.v}{pars\_confluent}}]
-  \label{lemma:parconfluent}
-  If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
-  $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
-\end{lemma}
-While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
-it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a
-=> b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).
-
-The transitivity of the coherence relation follows as a corollary of
-Lemma~\ref{lemma:parconfluent}.
+\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
+  \label{lemma:coherencerefl}
+  For all terms $[[a]]$, $[[a <=> a]]$.
+\end{lemma}
+\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
+  \label{lemma:coherencesym}
+  If $[[a <=> b]]$, then $[[b <=> a]]$.
+\end{lemma}
 \begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
@@
 \end{lemma}
-By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
-$[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
-$[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
-term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
-to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
-trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
-This concludes the proof that coherence is an equivalence relation.
-\begin{lemma}[Coherence Equivalence]
-  \label{lemma:coherenceequiv}
-  The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
-  transitivity and therefore is an equivalence relation.
-\end{lemma}
+
+From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
+\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
+equivalence relation. It is then trivial to derive that coherence is
+exactly the same as the untyped $\beta$-equivalence.
+
+% TODO: remove the lengthy discussion below
+
+% Now, we prove that coherence is indeed an equivalence relation.
+
+% First, we show that coherence is reflexive through the following
+% sequence of lemmas.
+
+% Lemma~\ref{lemma:parrefl} can be proven by structural induction over
+% the term $[[a]]$. Lemmas~\ref{lemma:parsrefl} and \ref{lemma:coherencerefl}
+% immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.
+
+% The reflexivity of parallel reduction enables us to embed rules from
+% call-by-name semantics into parallel reduction, as the following lemma
+% shows.
+% \begin{lemma}[Par AbsCbn\footnote{\dotv{join.v}{P\_AppAbs\_cbn}}] For all $[[A]], [[a]],$ and $[[b]]$,
+%   \label{lemma:parabscbn}
+%   $[[(\ A a) b => b {a}]]$
+% \end{lemma}
+% \begin{proof}
+%   Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
+% \end{proof}
+
+% Symmetry of coherence immediately falls from its definition.
+
+% Before we can prove transitivity, we need to show that parallel
+% reduction satisfies the diamond property.
+
+% The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
+% structural induction over the derivation of $[[a0 => a1]]$.
+% Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
+% over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
+% the use of Lemma~\ref{lemma:parcong}.
+
+% From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
+% single substitution property as a simple corollary.
+% \begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
+%   \label{lemma:parsubst}
+%   If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
+% \end{corollary}
+
+% A relation that satisfies the
+% diamond property must also be confluent, meaning that its transitive
+% and reflexive closure is confluent.
+% \begin{lemma}[Par confluent\footnote{\dotv{join.v}{pars\_confluent}}]
+%   \label{lemma:parconfluent}
+%   If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
+%   $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
+% \end{lemma}
+% While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
+% it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a => b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).
+
+% The transitivity of the coherence relation follows as a corollary of
+% Lemma~\ref{lemma:parconfluent}.
+% By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
+% $[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
+% $[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
+% term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
+% to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
+% trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
+% This concludes the proof that coherence is an equivalence relation.
+% \begin{lemma}[Coherence Equivalence]
+%   \label{lemma:coherenceequiv}
+%   The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
+%   transitivity and therefore is an equivalence relation.
+% \end{lemma}
+
+\begin{figure}[ht]
+\begin{minipage}{0.9\textwidth}
+\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
+\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Refl, J, Bool, True, False, If, Void}
+\end{minipage}
+\caption{Syntactic typing for \lang}
+\label{fig:typing}
+\end{figure}
 
@@
 \lang{}. \Rref{T-Conv} uses the definition of coherence from earlier
-for type conversion. The use of an untyped relation for type
-conversion makes our formulation slightly different from languages
-such as MLTT, where the judgmental equality typically takes the form
-TODO, where terms ... and ... are known to be well-typed. The
-equivalence of such systems and a system that uses untyped equality
-are explored in detail in ...
-
-Without fancy eta laws, it is easy to embed a typed language into an
-untyped language.
+as our equality judgment for type conversion. The use of an untyped
+relation for type conversion makes our formulation slightly different
+from languages such as MLTT, where the judgmental equality typically
+takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
+A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
+\citet{siles2012pure} shows the equivalence of Barendregt's
+Pure Type System~\citep{barendregt1991introduction}, which employs
+untyped equality, and its variant that uses typed judgmental
+equality. Furthermore, it is trivial to embed a system with judgmental
+equality to a system with untyped equality by erasing typing
+information. As a result, it is easy to port our consistency result from our
+type system to a variant with typed judgmental equality
+as long as the typed system does not include $\eta$ laws
+that would require type annotations (e.g. the $\eta$-law for unit
+types) and we discuss in Section~\ref{sec:discuss} how we can extend
+our proof to handle those rules.
+
+Working with a system with untyped equality has the huge benefit that
+the confluence result for untyped parallel reduction
+(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
+resort to the complex syntatic (resp. semantic) technique from
+\citet{siles2012pure} (resp. \citet{decagda}) to resolve
+the circularity of subject reduction and $\Pi$-injectivity described
+in \citet{siles2012pure}. Section~\ref{sec:discuss} explains how we
+can generalize our technique to include $\eta$-law for functions and
+show the existence of normal form for well-typed (open and closed)
+terms, achieving a similar level of expressiveness of the type system
+and strength of metatheoretic property as \citet{decagda}.
+
+Finally, since our system has an infinite universe hierarchy, we can
+present the system à la Russell by using the same judgment form $[[G
+|- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
+is therefore no need to distinguish between big types and small types
+and duplicate our typing specification.
+
+% , the
+% statement that $\Gamma \vdash [[Pi A0 B0]] \equiv [[Pi A1 B1]]$
+% implies $\Gamma \vdash [[A0]] \equiv [[A1]]$ and
+% $\Gamma, [[A0]]\vdash [[B0]] \equiv [[B1]]$.
+
+% working with a system with untyped equality not only preserves the
+% same level of generality,
+% Of course,
+% but also enables us to derive confluence
+% (Lemma~\ref{lemma:parconfluent}) early on without having to use the
+% intricate techniques from \citet{lemma:}
+
+% TODO, where terms ... and ... are known to be
+% well-typed. The equivalence of such systems and a system that uses
+% untyped equality are explored in detail in ...
+
+% Without fancy eta laws, it is easy to embed a typed language into an
+% untyped language.
 
@@
 level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
-$[[S]]$. For the
-majority of the properties we are about to prove in this section, we
+$[[S]]$.
+
+By unfolding the definition from Figure~\ref{fig:logrelrec}, we can
+now show that the same introduction rules for $[[Interp I i A S]]$ is
+admissible for $[[InterpR i A S]]$. We give as example TODO:
+
+In most informal presentations, instead of defining the logical
+relation in two steps as we have shown above, the rules for $[[InterpR
+i A S]]$ is given directly, with the implicit understanding of
+inductive definitions nested inside a recursive function. We choose
+the more explicit definition not only because it is directly definable
+in existing proof assistants where inductive definitions must appear
+at the top level, but also because it makes clear the induction
+principle we are allowed to use when reasoning about $[[InterpR i A
+S]]$.
+
+The most general format of the induction principle over $[[InterpR i A
+S]]$ is first by strong induction over the universe level $[[i]]$
+followed by structural induction over $[[Interp I i A S]]$. As
+examples, ... (\rref{I-Set} and \rref{I-Red}).
+
+For the majority of the properties we are about to prove in this section, we
 do not need any information about the parameterized function $[[I]]$.
@@
 
-For the rest of the section, we establish some important facts about
+To derive consistency, it suffices to restrict $[[S]]$ to the set of
+terms to reduce to closed terms. \Rref{I-Void, I-Bool} are
+unsurprising; when considering only closed terms, the empty type
+should not be inhabited and therefore corresponds to the empty set,
+whereas the boolean type is semantically inhabited by terms that
+evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
+says that an equality type $[[a ~ b : A]]$ corresonds to the
+set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
+otherwise corresponds to the empty set. Side conditions like $[[a <=>
+b]]$ are typically required for indexed types, of which equality types
+are an instance.
+
+\Rref{I-Pi} is the most interesting. The precondition consists of a
+mysterious function $[[F]]$ that takes an element from the set $[[S]]$, the interpretation
+of the type $[[A]]$, and returns a subset of terms. $[[F]]$ is further
+subject to the constraint that $[[forall a, (# a in S implies Interp I
+i B { a } F a #)]]$. Ignoring the content below the horizontal bar of
+the derivation, the statements from the precondition are equivalent to
+the conjunction of the following statements:
+\begin{itemize}
+\item $[[Interp I i A S]]$
+\item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
+  [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
+\end{itemize}
+In general, let $R$ be a binary relation over the sets $[[A]]$ and
+$[[B]]$, it easy to verify that the following equivalence:
+\[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
+  \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
+  \in R\]
+
+The left-hand side is equivalent to the right-hand side, but has no
+mentioning of any functions. In the context of \rref{I-Pi}, through
+the equivalence, we can more easily tell what's happening: $[[Pi A
+B]]$ has a semantic interpretation of $[[A]]$ can be semantically
+interpreted as the set $[[S]]$, and for all terms $[[a in S]]$, the
+type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
+$[[B]]$, has a semantic interpretation.
+
+If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
+  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
+presentation with an explicit $[[F in S -> PowerSet STm]]$ is
+useful. We want to state the fact that $[[B {a}]]$ has a semantic
+interpretation of some set $[[S0]]$, but we also want to refer to the
+set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
+to. If we used the alternative representation of the preconditions, we
+would be forced to define our logical relation in the following way.
+
+TODO: put the wrong logical relation that violates ...
+
+Unfortunately, both versions not only violate the syntactic strict
+positivity constraint required in proof assistants, but
+are in fact genuinely not monotone over the domain of
+relations.
+Intuitively, the failure of monotonicity stems from the fact
+that the witness picked in the precondition is not necessarily the
+same witness being referred to in the post condition. While it might
+be possible to restrict the domain with additional constraints such as
+functionality and inversion properties, we opt for our current
+skolemized formulation of \rref{I-Pi} so we immediately obtain a
+well-defined inductive relation and a usable induction principle. The
+more intuitive introduction rules can be later shown to be admissible
+a posteriori.
+
+In the rest of the section, we establish some important facts about
 the logical relation that will be useful for proving fundamental theorem
@@
     \end{itemize}
-    The properties above are exactly the preconditions needed to apply
-    \rref{I-Pi} for $[[Pi A B]]$ to finish off the proof.
+    These properties are exactly what's required in the conclusion.
   \end{description}
@@
     <=> b0]]$. The result then immediately follows from the fact that
-    coherence is an equivalence relation
-    (Lemma~\ref{lemma:coherenceequiv}).
+    coherence is an equivalence relation.
   \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
@@
 interpretation.
-\begin{corollary}[Logical Relation for Coherent Types\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
+\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
   \label{lemma:logrelcoherence}
@@
   S0]]$.
-  Pick $[[S i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
+  Pick $[[Suc i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
   ]]$ for $[[S0]]$ and it's trivial to verify the conditions hold.
@@
 \label{sec:logrelmech}
-The Coq mechanization follows an almost identical structure as the
-proofs presented in Section~\ref{sec:logrelproof}. The inductive
+
+\subsection{Artifacts Specific to Coq}
+In this section, we discuss the Coq encoding of the definitions and proofs presented
+in Section~\ref{sec:logrelproof}  and the artifacts that are
+specific to Coq and may not appear in other proof assistants.
+
+The powerset $[[PowerSet STm]]$ is encoded as the type \texttt{tm ->
+Prop}, a predicate over \lang{} terms.
+The inductive
 definition of the logical relation in Figure~\ref{fig:logrel} requires
@@
 relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
+
+In Coq, there is a distinction between computable functions and
+relations that can later be proven to be functional. The former can be
+viewed as a strict subset of the latter in axiom-free Coq. To be more
+precise, given a relation \texttt{R : A -> B -> Prop} subject to the
+totality and functionality constraints ($\forall$ \texttt{a} $\in$
+\texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
+that \texttt{R a b} is inhabited), we do not immediately obtain a function
+\texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
+\texttt{R a (F a)} is inhabited. However, the functional side
+conditions of a relation is clunky to express and tend to block
+automation. A simple workaround is to assume the choice axiom, which
+is known to be consistent and allows us to induce a function \texttt{F
+: A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
+is functional. This approach would make our Coq development match
+the text version of our proof from Section~\ref{sec:logrelproof} more
+closely.
+
+However, we choose instead an axiom-free workaround and define
+\rref{T-Pi} as follows in our Coq mechanization.
+
+Instead of constraining \texttt{F} to be a function or a functional
+relation, we leave \texttt{F} as an unrestricted relation that is
+total on input from the set $[[S]]$ and ...TODO...
+
+This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
+before we even show that our logical relation is
+functional/deterministic, but does not affect the proof structure
+otherwise. We choose to keep the discrepancy since the skolemization
+process is typically expressed in terms of function symbols rather
+than relation symbols.
+
+\subsection{Automation}
+\label{sec:automation}
+Our Coq mechanization heavily uses automation, though instead of
+defining custom tactics, we rely mostly on off-the-shelf tools such as
+Autosubst 2 and CoqHammer.
+
+We use the Autosubst 2 framework to specify our syntax in HOAS and
+produce Coq syntax files in de Bruijn representation. Additionallly,
+Autosubst 2 provides a powerful tactic \texttt{asimpl} which
+can be used to prove the equivalence of two terms constructed using
+the primitive operators provided by the framework. Instead of
+examining carefully the freeness conditions of variables to decide
+which rewriting lemma to invoke, reasoning about substitution with
+Autosubst 2 boils down to reducing the problem to a goal of the
+
+\section{Discussion}
+\label{sec:discuss}
+
+\subsection{Use of Impredicativity}
+
+\subsection{Toward Function $\eta$-laws and Decidability of Type Conversion}
 

### 2024-02-13 paper/paper.tex
Add text about automation

@@
 \usepackage{ottalt}
+\usepackage{minted}
 \usepackage{xspace}
@@
 can be used to prove the equivalence of two terms constructed using
-the primitive operators provided by the framework. Instead of
-examining carefully the freeness conditions of variables to decide
-which rewriting lemma to invoke, reasoning about substitution with
-Autosubst 2 boils down to reducing the problem to a goal of the
-
+the primitive operators provided by the framework. This greatly
+simplifies the reasoning about substitution in our development as
+almost all substitution related properties are immediately discharged
+by \texttt{asimpl} without having to manually prove or even directly invoke any
+substitution related lemmas.
+
+For other automation tasks that are not specific to binding, we use
+the powerful \texttt{sauto} tactic provided by CoqHammer to write
+short and declarative proofs. For example, here is a one-line proof of
+the triangle property about parallel reduction, from which the diamond
+property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
+\begin{minted}{coq}
+Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
+Proof.
+  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
+Qed.
+\end{minted}
+In prose, the triangle property can be proven by induction over the
+graph of the \texttt{tstar} function, which stands for the Takahashi
+translation~\citep{takahashi-parallel-reduction}. \texttt{inv:Par} and
+\texttt{ctrs:Par} says that the proof involves inverting and
+constructing of the derivations of parallel
+reduction. The option \texttt{use:Par\_refl,par\_cong} says that the proof would
+require the reflexivity and congruence properties of parallel
+reduction that we have shown earlier.
+
+The automation here not only gives us a proof that is shorter and more
+resiliant to changes in definition, but also gives useful
+documentation for readers who wish to understand how the underlying
+proof works since the automation tactic is guaranteed to not use
+lemmas that are not specified in the \texttt{use} flag.
 \section{Discussion}
@@
 
-\subsection{Toward Function $\eta$-laws and Decidability of Type Conversion}
+\subsection{Toward Decidability of Type Conversion and $\eta$ laws}
+In this section, we sketch out how the logical relation can be extended
+show the existence of $\beta\eta$ normal forms for well-typed terms.
+In other words, our goal is to show that the following property holds
+in our system.
 

### 2024-02-13 paper/paper.tex
Add text about the normal form extension

@@
 
+\section{Toward Decidability of Type Conversion and $\eta$ laws}
+In this section, we sketch out how the logical relation from
+Section~\ref{sec:logreldep} can be extended show the existence of
+$\beta\eta$ normal forms for (open and closed) well-typed terms.
+
+The updated logical relation and its auxiliary definitions are shown
+in Figure~\ref{fig:logrelopen}. When extend parallel reduction to
+include $\eta$ reduction for functions. This effectively adds $\eta$
+laws for functions to our equational theory since coherence, the
+untyped relation used for type conversion, is built on top of parallel
+reduction.
+
+We can recover the same confluence result about parallel reduction
+using the standard techniques from \citet{barendregt, takahashi},
+though specifically for the de Bruijn representation, we need the
+following anti-renaming lemma about parallel reduction:
+\begin{lemma}[Par anti-renaming]
+  \label{lemma:parantirenaming}
+  If $[[a {xi} => b0]]$, then there exists some $[[b]]$ such that
+  $[[b {xi} = b0]]$ and $[[a => b]]$.
+\end{lemma}
+
+% The idea of a term blocked from $\beta$ reduction is
+% captured by $[[ne]]$, the set of neutral terms. We augment the
+% interpretation of booleans and the empty types with terms that
+% evaluate to neutral terms. Furthermore, types that are neutral terms
+% can also be assigned a meaning since they may be inhabited by other
+% neutral terms.
+The syntactic forms $[[ne]]$ and $[[nf]]$ capture the neutral terms
+and normal forms with respect to $\beta$-reduction, but can still take
+$\eta$ reduction steps. Furthermore, we can show that parallel
+preserves $[[ne]]$ and $[[nf]]$.
+
+The predicates $[[wne a]]$ and $[[wn
+a]]$ describe terms that can evaluate into $\beta$-neutral or
+$\beta$-normal form through parallel reduction.
+
+In the definition of the logical relation, we omit the rules for
+the function and universe cases since they remain identical to the
+original version in Figure~\ref{fig:logrel}.
+The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
+An open term of type $[[Bool]]$ does
+not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
+a variable, or more generally, a neutral term. Likewise, the
+$[[Void]]$ type, while remains uninhabited under an empty context, may
+be inherited by the set of neutral terms when there is a variable in
+the context that allows us to inhabit $[[Void]]$.
+
+The rule for equality type $[[a ~ b : A]]$ is augmented with the preconditions that
+$[[a]]$, $[[b]]$, and $[[A]]$ are all in their normal forms since
+otherwise our model would include equality types that are themselves
+not normalizing. Furthermore, the side condition $[[a <=> b]]$ is only
+available when the equality proof reduces to $[[refl]]$. If the proof
+term reduces to a neutral term, then there is nothing we can learn
+about the relationship between $[[a]]$ and $[[b]]$.
+
+Finally, in a
+non-empty context, a type itself may evaluate to a neutral term and in
+turn can only inhabited by neutral terms, thus the addition of \rref{I-Ne}.
+
+Interestingly, all the properties we have shown in
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} up to the
+fundamental lemma can be proven in the exact same order, where the new
+cases related due to \rref{I-Ne} and the augmentation of neutral terms
+to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
+Lemma~\ref{lemma:parnenf}.
+
+Before we can prove the fundamental theorem and derive the
+normalization property as its corollary, we need to additionally
+formulate and prove the adequacy property.
+
+Let us first start with some auxiliary lemmas and definitions.
+\begin{lemma}[Ext Wn]
+  \label{lemma:extwn}
+  If $[[wn a i]]$, then $[[wn a]]$.
+\end{lemma}
+\begin{proof}
+  By induction over the length of the reduction sequence in $[[wn a
+  i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
+  \ref{lemma:parnenf}.
+\end{proof}
+
+\begin{lemma}[wne wn]
+  \label{lemma:wnewn}
+  If $[[wne a]]$ and $[[wn b]]$, then $[[wne a b]]$.
+\end{lemma}
+\begin{proof}
+  Immediate by lexicographical induction over the length of the reduction sequences in
+  $[[wne a]]$ and $[[wn b]]$.
+\end{proof}
+
+\begin{definition}[CR]
+  Let $[[S]]$ be a set of lambda terms. We say that
+  \begin{itemize}
+  \item $[[S]] \in CR_1$ if $[[wne a implies a in S]]$
+  \item $[[S]] \in CR_2$ if $[[a in S implies wn a]]$
+  \item $[[S]] \in CR$ if $[[S]] \in CR_1$ and $[[S]] \in CR_2$
+  \end{itemize}
+\end{definition}
+
+\begin{lemma}[$CR_1$ for type]
+  \label{lemma:cr1ty}
+  If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
+\end{lemma}
+\begin{proof}
+  Immediately from \rref{I-Ne} and \rref{I-Par}.
+\end{proof}
+
+\begin{lemma}[$CR$ for interpreted sets]
+  \label{lemma:crel}
+  If $[[I]] \in CR$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $[[Interp I i A S]]$. The base
+  cases are all immediate.
+
+  In the function case ($[[Pi A B]]$), we use the induction
+  hypothesis to show that variables, which are special cases of
+  neutral terms, semantically inhabit the input type $[[A]]$. We apply the
+  function to an arbitrary variable and we can then use
+  Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.
+
+  To show
+  the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
+  such that $[[wne b]]$,
+  and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
+  us a term that semantically inhabits $[[B {a}]]$. From the induction
+  hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
+  holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
+  from the induction hypothesis, since every term that evaluates to
+  some neutral form semantically inhabits $[[B {a}]]$. Done.
+\end{proof}
+
+\begin{lemma}[$CR_2$ for type]
+  \label{lemma:cr2ty}
+  If $[[I]] \in CR$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $[[Interp I i A S]]$. The
+  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
+\end{proof}
+
+\begin{lemma}[$CR_2$ for types (Rec)]
+  \label{lemma:cr2tyrec}
+  If $[[InterpR i A S]]$, then $[[wn A]]$.
+\end{lemma}
+\begin{proof}
+  By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
+\end{proof}
+
+\begin{lemma}[$CR$ for terms (Rec)]
+  \label{lemma:crelrec}
+  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
+\end{lemma}
+\begin{proof}
+  After unfolding the definition of $[[InterpR i A S]]$, trivial by
+  Lemma~\ref{lemma:cr2ty}, \ref{lemma:crel} and \ref{lemma:cr1ty}.
+\end{proof}
+
+The formulation of the valuation and semantic well-typedness from
+Figure~\ref{fig:semtyping} and the fundamental lemma remains
+unchanged.
+
+The proof of the fundamental lemma is still carried out by induction
+over the typing derivation, where the additional neutral term related
+cases are handled by Lemmas\ref{lemma:cr2tyrec} and
+\ref{lemma:crelrec}.
+
+The normalization property then follows as a corollary of the
+fundamental theorem.
+\begin{theorem}[Existence of $\beta$-normal form]
+  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
+\end{theorem}
+\begin{proof}
+  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
+  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
+  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
+  We pick the valuation $[[rho]]$ as the term constructor that injects
+  natural numbers as variables. This is
+  allowed since Lemma~\ref{lemma:crelrec} says neutral terms
+  semantically inhabit any type.
+\end{proof}
+
 \section{Mechanization}
@@
 
-\subsection{Toward Decidability of Type Conversion and $\eta$ laws}
-In this section, we sketch out how the logical relation can be extended
-show the existence of $\beta\eta$ normal forms for well-typed terms.
-In other words, our goal is to show that the following property holds
-in our system.
-
 \section{Related Work}

### 2024-02-13 paper/paper.tex
Add more text to the extension section

@@
 impredicativity manifests in the form of second-order logical
-formulas and thus more intuitive to understand for readers with some
+formulas and thus more intuitive to understand for readers with
 general mathematics background.
 
-In Section~\ref{sec:logrelmech}, we discuss the details about our
-mechanization, including our use of existing libraries such as
-Autosubst for handling bindings and CoqHammer for general-purpose
-automation. In Section~\ref{sec:discuss}, we extend our logical
+In Section~\ref{sec:extension}, we extend our logical
 relation to prove the stronger property that every well-typed term
-has a normal form. Most interestingly, we show that extensions such
-as $\eta$ rules for functions and normalization for open terms are
+has a normal form. Most interestingly, we find that extending the
+language with
+$\eta$ rule for functions and the metatheoretic result to normalization for open terms are
 mostly orthogonal to the fact that the language is dependently
@@
 a simply typed language and later ported into a dependently typed setting.
+
+In Section~\ref{sec:logrelmech}, we discuss the details about our
+mechanization, including our use of existing libraries such as
+Autosubst for handling bindings and CoqHammer for general-purpose
+automation.
+
+Finally, in Section~\ref{sec:relatedwork}, we discuss how our proof
+relates to existing proofs by logical relations for dependent types in
+the literature.
 
@@
 \begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
-  \label{theorem:soundness}
+  \label{theorem:soundness}\leavevmode
   \begin{itemize}
@@
 \section{Toward Decidability of Type Conversion and $\eta$ laws}
+\label{sec:extension}
+\begin{figure}[h]
+  \[
+    \begin{array}{lcl}
+      % \beta\text{-}\mathit{neutral\ terms}\\
+      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
+                    f]] \\ \\
+      % \beta\text{-}\mathit{normal\ terms} \\
+      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
+                    |\ [[f ~ f : f]]\\
+            & |   & [[\\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
+    \end{array}
+  \]
+  \caption{$\beta$-neutral and normal forms}
+  \label{fig:nenf}
+\end{figure}
+\begin{figure}[h]
+  \caption{Extended logical relation}
+  \label{fig:logrelopen}
+\end{figure}
+
 In this section, we sketch out how the logical relation from
 Section~\ref{sec:logreldep} can be extended show the existence of
-$\beta\eta$ normal forms for (open and closed) well-typed terms.
+$\beta\eta$ normal forms for (open and closed) well-typed terms. We
+first extend parallel reduction to include $\eta$ reduction for
+functions.
+\begin{center}
+  \drule[width=2.5in]{P-AbsEta}
+\end{center}
+\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
+equational theory since coherence, the untyped relation used for type
+conversion, is built on top of parallel reduction. We can recover the
+same confluence result about parallel reduction using the standard
+techniques from \citet{barendregt:lambda-calculi-with-types,
+takahashi-parallel-reduction}, though specifically for the de Bruijn
+representation, we need the following anti-renaming lemma about
+parallel reduction.
+\begin{lemma}[Par anti-renaming]
+  \label{lemma:parantirenaming} If $[[a { xi } => b0]]$, then there
+exists some $[[b]]$ such that $[[b { xi } = b0]]$ and $[[a => b]]$.
+\end{lemma}
+
+The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the neutral terms
+and normal forms with respect to $\beta$-reduction, but can still take
+$\eta$ reduction steps. We sometimes use the judgment forms $[[ne a]]$
+and $[[nf a]]$ to indicate that there exists $[[e]]$ or $[[f]]$ such
+that $[[a = e]]$ or $[[a = f]]$.
+
+We can show that parallel reduction preserves $\beta$-normal and
+neutral forms.
+\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
+  \label{lemma:parnenf}
+  If $[[a => b]]$, then
+  \begin{itemize}
+  \item $[[ne a]]$ implies $[[ne b]]$
+  \item $[[nf a]]$ implies $[[nf b]]$
+  \end{itemize}
+\end{lemma}
+This lemma is mainly to show that \rref{P-AbsEta} does not introduce
+any new $\beta$ redex because $\eta$ reduction is the only real step
+we can take.
+
+The predicates $[[wne a]]$ and $[[wn
+a]]$ describe terms that can evaluate into $\beta$-neutral or
+$\beta$-normal form through parallel reduction and are defined as
+follows.
+\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
+\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
 
 The updated logical relation and its auxiliary definitions are shown
-in Figure~\ref{fig:logrelopen}. When extend parallel reduction to
-include $\eta$ reduction for functions. This effectively adds $\eta$
-laws for functions to our equational theory since coherence, the
-untyped relation used for type conversion, is built on top of parallel
-reduction.
-
-We can recover the same confluence result about parallel reduction
-using the standard techniques from \citet{barendregt, takahashi},
-though specifically for the de Bruijn representation, we need the
-following anti-renaming lemma about parallel reduction:
-\begin{lemma}[Par anti-renaming]
-  \label{lemma:parantirenaming}
-  If $[[a {xi} => b0]]$, then there exists some $[[b]]$ such that
-  $[[b {xi} = b0]]$ and $[[a => b]]$.
-\end{lemma}
+in Figure~\ref{fig:logrelopen}.
 
 % The idea of a term blocked from $\beta$ reduction is
-% captured by $[[ne]]$, the set of neutral terms. We augment the
+% captured by $[[e]]$, the set of neutral terms. We augment the
 % interpretation of booleans and the empty types with terms that
@@
 % neutral terms.
-The syntactic forms $[[ne]]$ and $[[nf]]$ capture the neutral terms
-and normal forms with respect to $\beta$-reduction, but can still take
-$\eta$ reduction steps. Furthermore, we can show that parallel
-preserves $[[ne]]$ and $[[nf]]$.
-
-The predicates $[[wne a]]$ and $[[wn
-a]]$ describe terms that can evaluate into $\beta$-neutral or
-$\beta$-normal form through parallel reduction.
 
@@
   \label{lemma:crel}
-  If $[[I]] \in CR$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
+  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
 \end{lemma}
@@
   \label{lemma:cr2ty}
-  If $[[I]] \in CR$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
+  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
 \end{lemma}
@@
 unchanged.
-
 The proof of the fundamental lemma is still carried out by induction
 over the typing derivation, where the additional neutral term related
-cases are handled by Lemmas\ref{lemma:cr2tyrec} and
+cases are handled by Lemmas~\ref{lemma:cr2tyrec} and
 \ref{lemma:crelrec}.
@@
 fundamental theorem.
-\begin{theorem}[Existence of $\beta$-normal form]
+\begin{corollary}[Existence of $\beta$-normal form]
+  \label{corollary:exbetanf}
   If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
-\end{theorem}
+\end{corollary}
 \begin{proof}
@@
   that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
-  We pick the valuation $[[rho]]$ as the term constructor that injects
-  natural numbers as variables. This is
-  allowed since Lemma~\ref{lemma:crelrec} says neutral terms
-  semantically inhabit any type.
-\end{proof}
+  We pick the valuation $[[rho]](i) = i$, which injects
+  natural numbers as term variables. This is
+  allowed since Lemma~\ref{lemma:crelrec} says neutral terms,
+  including variables, semantically inhabit any $[[S0]]$ where
+  $[[S0]]$ is the interpretation of some type. With our choice of
+  $[[rho]]$, we have $[[A {rho} = A]]$ and $[[a {rho} = a]]$. Then we
+  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
+  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
+  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
+\end{proof}
+From Corollary~\ref{corollary:exbetanf}, we can show that there exists
+$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
+preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
+and also strictly decreases the size of the term.
+
+Due to the non-deterministic nature of parallel reduction, we need to
+take a few more steps to convert the existence of $\beta\eta$-normal
+form into a decision procedure for type conversion. More specifically,
+we can show that a deterministic evaluation strategy such as
+leftmost-outermost reduction can always find the $\beta\eta$-normal
+form if there exists one. However, we omit such proofs since they can
+can be formulated on untyped lambda terms and thus orthogonal to the
+specifics of dependently typed systems. Instead, we redirect readers
+to \citet{factorization-essentially, takahashi-parallel-reduction} for
+the details.
+
+Finally, we want to point out that
+Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
+\ref{lemma:crelrec}, often referred to as adequacy of the logical
+relation, are in fact required in the normalization proof for simply
+typed languages~\citep{abel2019poplmark}. In the simply typed
+scenario, adequacy needs to be proven before the fundamental theorem
+so we can handle rules such as \rref{T-If} where the scrutinee is a
+neutral term. In a simply typed setting, Lemma~\ref{lemma:extwn} is
+used in the exact same way to show that lambda terms are themselves
+normalizing as we have done in Lemma~\ref{lemma:crel}.
+
+Dependent types make the proof slightly more complicated
+as we also need to know that every type has a normal form. Whereas the
+$CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
+derivable, the $CR_2$ property for types (Lemma~\ref{lemma:cr2ty})
+requires the $CR_1$ property about the interpreted sets
+(Lemma~\ref{lemma:crel}).
+
+Overall, despite the dependently typed setting,
+the extension of our logical relation to prove normalization of open
+\emph{and} closed terms closely mirrors the progression from
+normalization of closed terms~\citep{harpertait} to normalization of
+open terms~\citep{harperkripke} in the simply typed lambda calculus.
+It is in fact reassuring that once we have laid the foundational
+technique for handling dependent types in our logical relation, the
+further extensions more or less boil down to properties that can be
+derived through syntactic means, which tend to not get in the way of
+Gödel's second incompleteness theorem.
 
@@
 \label{sec:logrelmech}
+\begin{figure}[h]
+  TODO
+  \caption{Statistics of the Coq Development}
+  \label{fig:linecount}
+\end{figure}
+
+Figure~\ref{fig:linecount} shows the statics of our
+development, including the base consistency proof from
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
+extension from Section~\ref{sec:extension}.
 
@@
 graph of the \texttt{tstar} function, which stands for the Takahashi
-translation~\citep{takahashi-parallel-reduction}. \texttt{inv:Par} and
+translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
 \texttt{ctrs:Par} says that the proof involves inverting and
 constructing of the derivations of parallel
-reduction. The option \texttt{use:Par\_refl,par\_cong} says that the proof would
-require the reflexivity and congruence properties of parallel
-reduction that we have shown earlier.
+reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
+automation tactic to use
+the reflexivity and congruence properties of parallel
+reduction as lemmas.
 
@@
 proof works since the automation tactic is guaranteed to not use
-lemmas that are not specified in the \texttt{use} flag.
+lemmas or invert derivations that are not specified in the
+\texttt{use} or \texttt{inv} flag.
 \section{Discussion}

### 2024-02-14 paper/paper.tex
Add text to related work

@@
 syntax we work with is unscoped and the choice does matter when
-we extend our logical relation to open terms in Section~\ref{sec:discuss}.
+we extend our logical relation to open terms in Section~\ref{sec:extension}.
 
@@
 that would require type annotations (e.g. the $\eta$-law for unit
-types) and we discuss in Section~\ref{sec:discuss} how we can extend
-our proof to handle those rules.
+types). % and we discuss in Section~\ref{sec:extension} how we can extend
+% our proof to handle those rules.
 
@@
 the circularity of subject reduction and $\Pi$-injectivity described
-in \citet{siles2012pure}. Section~\ref{sec:discuss} explains how we
-can generalize our technique to include $\eta$-law for functions and
+in \citet{siles2012pure}. Section~\ref{sec:extension} explains how we
+generalize our technique to include $\eta$-law for functions and
 show the existence of normal form for well-typed (open and closed)
@@
 |- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
-is therefore no need to distinguish between big types and small types
+is no need to distinguish between big types and small types
 and duplicate our typing specification.
@@
 \texttt{use} or \texttt{inv} flag.
-\section{Discussion}
-\label{sec:discuss}
-
-\subsection{Use of Impredicativity}
 
@@
 \label{sec:relatedwork}
+\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
+\citet{barendregt:lambda-calculi-with-types} are some of the earlier
+works that establish metatheoretic results strong enough to derive
+consistency for dependently typed systems. Some of these techniques
+are still in use in more recent works. For example, the technique for
+proving strong normalization by \citet{geuvers1994short} is adopted by
+\citet{moon2021graded} to show the same property to a dependently
+typed system extended with modalities.
+
+More recently, there have been efforts at mechanizing consistency
+or decidability of type conversion proofs using proof assistants such
+as Coq and Agda.
 

### 2024-02-14 paper/paper.tex
Add more text to related work

@@
 lack of fully dependently typed systems where computations may appear
-at the type-level. In fact, it is far from obvious why proof by
+at the type-level. It is far from obvious why proof by
 logical relation is even applicable to dependent types, since the type
@@
 Kripke-style models that obscure the technique for addressing
-type-level computation. In fact, the added complexivity is evident
+type-level computation. The added complexivity is evident
 from the size of their code base. For example, \citet{nbeincoq}
@@
   Bool, Eq, Refl, J, JRefl}
-\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{One, Step}
+\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
 \drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
@@
 term $[[c]]$ through parallel reduction. The symmetric notation of
-coherence suggests that it is an equivalence relation.% , and in fact, we
-% will use coherence as the equational theory for our type conversion
-% rule.
+coherence suggests that it is an equivalence relation.
 
@@
 positivity constraint required in proof assistants, but
-are in fact genuinely not monotone over the domain of
+are genuinely non-monotone over the domain of
 relations.
@@
 are still in use in more recent works. For example, the technique for
-proving strong normalization by \citet{geuvers1994short} is adopted by
+proving strong normalization by \citet{geuvers1994short} is adapted by
 \citet{moon2021graded} to show the same property to a dependently
@@
 
-More recently, there have been efforts at mechanizing consistency
-or decidability of type conversion proofs using proof assistants such
-as Coq and Agda.
+\citet{decagda} mechanizes in Agda the decidability of type
+conversion rule for a dependently typed language with one predicative
+universe level and typed judgmental equality with function
+$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
+uses a Kripke-style logical relation parameterized over an
+type-directed equivalence relation satisfying certain properties. We
+are able to avoid the complexivity of a Kripke-style logical relation
+by defining our logical relation in terms of untyped parallel
+reduction, which does not impose any scoping constraint on the
+terms. \citet{decagda} defines their logical relation using the
+induction-recursion scheme, which is available in Agda but not in Coq.
+\citet{martin-lof-a-la-coq} manages to encode the logical relation
+from \citet{decagda} in the predicative fragment of Coq using a
+special way of encoding induction recursion. Their work further
+extends the decidability of type conversion result from~\citet{decagda} to the decidability
+type checking of a bidrectional type system.
+
+\citet{anand2014towards} mechanizes the metatheory of
+Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
+extensional type theory with features such as dependent functions,
+inductive types, and a full universe hierarchy. \citet{nbeincoq}
+mechanizes the normalization-by-evaluation algorithm in Coq for a
+dependently typed language with one predicative universe, similar to
+\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
+\citet{anand2014towards} and \citet{nbeincoq} leverage the
+impredicative \texttt{Prop} sort of Coq to define the interpretation
+of dependent function types and thus are closely related to our
+mechanization. However, instead of explaining the purely inductive
+logical relation as a convoluted workaround to the lack of
+induction-recursion in Coq, we give a self-contained explanation of
+our logical relation.
+
+Finally, it is worth pointing out that all systems we have discussed
+so far builds a relational model for their logical relation. Recall
+that our logical relation takes the form $[[InterpR i A S]]$ where
+$[[S]]$ is a set of terms. In a relational model, each type is
+interpreted as a partial equivalence relation over terms rather than a
+set. A relational model is necessary for an extensional type theory
+such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
+use a relational model either to justify the function $\eta$-law or to
+derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
+law for function is baked into the untyped reduction
+relation and therefore a relational model is not needed.
+
+The dependently typed systems we have discussed so far also vary
+greatly in expressiveness. For example, the Calculus of
+Constructions~\citep{CoC} and MLTT without universe levels are unable
+to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
+identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
+the mechanized systems we have discussed so far, only
+\citet{anand2014towards} has a full universe hierarchy.
+Our object language \lang{}, while small, has a decent coverage of
+features commonly seen in dependently typed languages, including
+dependent pattern matching, dependent function types, a full universe
+hierarchy, an intensional identity type. We leave features such as
+type-directed reduction (necessary for unit $\eta$-law) and
+impredicative sorts as part of our future work.
 

### 2024-02-14 paper/paper.tex
Use array for semtyping definitions

@@
 \begin{figure}[h]
-  \begin{equation*}
-    \begin{split}
-      [[rho |= G]] &:= \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
+\[
+\begin{array}{lcl}
+
+      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
                      } [[InterpR j (up Suc i G i) { rho } S ]] \text{, then } [[rho i in S]] \\
-      [[G |= a : A]] &:= \forall [[rho]], \text{ if }[[rho |=
+      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
                        G]]\text{ then there exists some } [[j]] \text{
@@
                        {rho} S]] \text{ and } [[a {rho} in S]] \\
-      [[|= G]] &:= \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
-    \end{split}
-  \end{equation*}
+      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
+\end{array}
+\]
   \caption{Semantic Typing for \lang}
@@
 
+\section{Conclusion}
+\label{sec:conclusion}
+
 % Type soundness can be proven through a syntactic

### 2024-02-14 paper/paper.tex
Update the if rule rendering
Add the admissible pi alt rule in natural deduction form

@@
 mechanism for encoding induction
-recursion~\citep{induction-recursion}, a scheme in which logical
-relation for dependent types is typically defined~\citep{todo}, we opt
-for a direct explanation in informal language of sets, where
-impredicativity manifests in the form of second-order logical
-formulas and thus more intuitive to understand for readers with
-general mathematics background.
+recursion~\citep{induction-recursion-dybjer}, a scheme in which
+semantic models (including logical relations) for dependent types can
+be defined, we opt for a direct explanation in informal language of
+sets, where impredicativity manifests in the form of second-order
+logical formulas and thus more intuitive to understand for readers
+with general mathematics background.
 
@@
 \drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
-\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Refl, J, Bool, True, False, If, Void}
+\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
+  Refl, J, If, Bool, True, False, Void}
 \end{minipage}
@@
 By unfolding the definition from Figure~\ref{fig:logrelrec}, we can
-now show that the same introduction rules for $[[Interp I i A S]]$ is
-admissible for $[[InterpR i A S]]$. We give as example TODO:
+now show that the same introduction rules for $[[Interp I i A S]]$ are
+admissible for $[[InterpR i A S]]$, by
+instantiating $[[I]]$ with $[[I i]] := [[{A | exists S , InterpR i A
+  S}]]$, the same function $[[I]]$ used in the definition of $[[InterpR i A
+S]]$. We give as example the following rules.
+\begin{center}
+\drule[]{IR-Void} \qquad \drule[]{IR-Set}
+\end{center}
 
@@
 relation in two steps as we have shown above, the rules for $[[InterpR
-i A S]]$ is given directly, with the implicit understanding of
-inductive definitions nested inside a recursive function. We choose
+i A S]]$ are given directly, with the implicit understanding that the
+relation is an inductive definition nested inside a recursive
+function over the universe level $[[i]]$. We choose
 the more explicit definition not only because it is directly definable
@@
 
-The most general format of the induction principle over $[[InterpR i A
-S]]$ is first by strong induction over the universe level $[[i]]$
-followed by structural induction over $[[Interp I i A S]]$. As
-examples, ... (\rref{I-Set} and \rref{I-Red}).
+% The paragraph below might be useful but I don't know how to phrase
+% it well
+
+% The most general format of the induction principle over
+% $[[InterpR i A S]]$ is first by strong induction over the universe level $[[i]]$
+% followed by structural induction over $[[Interp I i A S]]$. As
+% examples, ... (\rref{I-Set} and \rref{I-Red}).
 
@@
 In general, let $R$ be a binary relation over the sets $[[A]]$ and
-$[[B]]$, it easy to verify that the following equivalence:
+$[[B]]$, it easy to verify the following equivalence, which is used to
+justify the skolemization process~\citep{skolemization}.
 \[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
@@
   \in R\]
-
 The left-hand side is equivalent to the right-hand side, but has no
-mentioning of any functions. In the context of \rref{I-Pi}, through
-the equivalence, we can more easily tell what's happening: $[[Pi A
-B]]$ has a semantic interpretation of $[[A]]$ can be semantically
-interpreted as the set $[[S]]$, and for all terms $[[a in S]]$, the
+mentioning of any functions.
+In the context of \rref{I-Pi}, through
+the equivalence, we can more easily tell what's happening: the
+function type $[[Pi A B]]$ has a semantic interpretation if its input
+type $[[A]]$ can be semantically
+interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
 type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
@@
 set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
-to. If we used the alternative representation of the preconditions, we
-would be forced to define our logical relation in the following way.
-
-TODO: put the wrong logical relation that violates ...
-
-Unfortunately, both versions not only violate the syntactic strict
+to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
+witness in the precondition for $[[a in S]]$.
+With the alternative representation that is free of the function $[[F]]$, we
+might want to reformulate our logical relation in the following way.
+\begin{center}
+  \drule[]{I-PiAlt}
+\end{center}
+Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
 positivity constraint required in proof assistants, but
-are genuinely non-monotone over the domain of
-relations.
+is genuinely non-monotone when written as an endofunction over the
+domain of relations.
 Intuitively, the failure of monotonicity stems from the fact
@@
 skolemized formulation of \rref{I-Pi} so we immediately obtain a
-well-defined inductive relation and a usable induction principle. The
-more intuitive introduction rules can be later shown to be admissible
-a posteriori.
+well-defined inductive relation and a usable induction
+principle. However, as we will see in Lemma~\ref{lemma:piintroalt} to
+show the admissibility of \rref{I-Pi} a posteriori.
 
@@
 
-Lemma~\ref{lemma:logreldeter} enables us to derive the following
-alternative introduction and inversion principles for \rref{I-Pi}.
+Lemma~\ref{lemma:logreldeter} enables us to show the admissibility of
+\rref{I-PiAlt} and its related inversion principle.
 \begin{lemma}[Pi Intro Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
@@
 \begin{figure}[h]
+  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
   \caption{Extended logical relation}

### 2024-02-15 paper/paper.tex
Minor tweaks

@@
 plethora of introductory materials about logical relation in the
-context of simply typed lambda calculus and System F, a streamlined
+context of simply typed or polymorphic lambda calculus, a streamlined
 presentation of proof by logical relation for a dependently language
-is lacking in comparison. In this paper, I present a short
+is lacking. In this paper, I present a short
 consistency proof for a dependently typed language that contains a
-rich set of features, including an infinite cumulative universe
+rich set of features, including a full cumulative universe
 hierarchy, booleans, and an intensional identity type. We have
@@
 terms so we can derive properties such as normalization for well-typed
-open terms. In short, the introductory texts about logical relations
+open terms. Overall, the introductory texts about logical relations
 cover systems and properties with varying degrees of complexity, from
@@
 type-level computation. The added complexivity is evident
-from the size of their code base. For example, \citet{nbeincoq}
-involves 22 thousand lines of Coq code.
+from the size of their code base. All three developments involve 20
+thousand to 30 thousand lines of Coq or Agda code.
 
 The goal of this paper is to give a tutorial on proof by logical relation for
-dependent types in a simple and digestable format. % The language specification will
-% be given in Section~\ref{sec:spec}.
-The system we present is small but not minimal. We want to cover
-features that are only available or require special treatments under
-dependent types%  without introducing complexities that would have
-% existed in its simply typed variant
-. We choose boolean
+dependent types in a simple and digestable format.
+The system we present is small but covers
+important features that are only available or require special treatments in
+dependently typed systems. We choose boolean
 types over natural numbers as our base type for simplicity, but include an
-intensional identity type in our type system as the interpretation
-of identity types requires an extra side condition that is needed for
-indexed family of types. Unlike \citet{nbeincoq,decagda,martin-lof-a-la-coq}
-but similar to \citet{anand2014towards}, we include an infinite
-hierarchy of universes since we want to not only have at least one
-predicative universe level to support type-level computations, but
-also avoid the unnecessary code duplication pointed out by
-\citet{nbeincoq}.
+intensional identity type in our type system to show how indexed
+types are treated in a logical relational proof. Unlike
+\citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
+\citet{anand2014towards}, we include an infinite hierarchy of
+universes to not only support type-level computations, but also avoid
+the unnecessary code duplication pointed out by \citet{nbeincoq} when
+big and small types are treated non-uniformly.
 
 Our consistency proof is short and fully mechanized. The proof scripts
-involve less than 1000 lines of Coq code. In fact, what is even more
+involve less than 1000 lines of manually written Coq code. In fact, what we find
 encouraging is that among the 1000
@@
 Thanks to the conciseness of
-the proof, we are able to present most of its details in
+the proof, we are able to present it in detail in
 Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
@@
 which leverages impredicativity to define the logical
-relation. % In our presented proof, we use the informal
-% language of sets, where the use of impredicativity feels more natural
-% and less noticeable.
+relation as a partial function.
 Rather than framing impredicativity as a
@@
 semantic models (including logical relations) for dependent types can
-be defined, we opt for a direct explanation in informal language of
-sets, where impredicativity manifests in the form of second-order
-logical formulas and thus more intuitive to understand for readers
-with general mathematics background.
+be defined, we opt for a direct explanation through the informal
+language of sets, where impredicativity manifests in the form of
+second-order logical formulas and thus more intuitive to grasp
+for readers with a general mathematical background.
 
@@
 automation.
-
-Finally, in Section~\ref{sec:relatedwork}, we discuss how our proof
+In Section~\ref{sec:relatedwork}, we discuss how our proof
 relates to existing proofs by logical relations for dependent types in
@@
 \end{lemma}
+\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
+  \label{lemma:parsubst}
+  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
+\end{corollary}
 \begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
@@
 % single substitution property as a simple corollary.
-% \begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
-%   \label{lemma:parsubst}
-%   If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
-% \end{corollary}
 
@@
 \citet{siles2012pure} (resp. \citet{decagda}) to resolve
-the circularity of subject reduction and $\Pi$-injectivity described
-in \citet{siles2012pure}. Section~\ref{sec:extension} explains how we
+the circularity of subject reduction and $\Pi$-injectivity.
+% described
+% in \citet{siles2012pure}
+Section~\ref{sec:extension} explains how we
 generalize our technique to include $\eta$-law for functions and
@@
 \end{equation*}
-The function $[[I]]$ is a family of subsets of terms indexed by
+The function $[[I]]$ is a family of sets of terms indexed by
 natural numbers strictly less than the parameter $[[i]]$, which
@@
 
-By unfolding the definition from Figure~\ref{fig:logrelrec}, we can
-now show that the same introduction rules for $[[Interp I i A S]]$ are
+With the definition from Figure~\ref{fig:logrelrec}, we can
+show that the same introduction rules for $[[Interp I i A S]]$ are
 admissible for $[[InterpR i A S]]$, by
@@
 well-defined inductive relation and a usable induction
-principle. However, as we will see in Lemma~\ref{lemma:piintroalt} to
-show the admissibility of \rref{I-Pi} a posteriori.
+principle. However, as we will see in Lemma~\ref{lemma:piintroalt}, it
+is possible to show the admissibility of \rref{I-PiAlt} a posteriori.
 
@@
 dependent pattern matching, dependent function types, a full universe
-hierarchy, an intensional identity type. We leave features such as
+hierarchy, an intensional identity type. We hope features such as
 type-directed reduction (necessary for unit $\eta$-law) and
-impredicative sorts as part of our future work.
+impredicative sorts can be implemented in a similarly streamlined
+fashion and we will leave those as part of our future work.
 

### 2024-02-15 paper/paper.tex
Add the coq pi rule

@@
 mechanization, including our use of existing libraries such as
-Autosubst for handling bindings and CoqHammer for general-purpose
+Autosubst 2~\citep{autosubst2} for handling bindings and CoqHammer~\citep{czajka2018hammer} for general-purpose
 automation.
@@
 conditions of a relation is clunky to express and tend to block
-automation. A simple workaround is to assume the choice axiom, which
-is known to be consistent and allows us to induce a function \texttt{F
+automation. A simple workaround is to assume the axiom of unique choice, which
+is known to be consistent with Coq and allows us to induce a function \texttt{F
 : A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
@@
 However, we choose instead an axiom-free workaround and define
-\rref{T-Pi} as follows in our Coq mechanization.
-
-Instead of constraining \texttt{F} to be a function or a functional
-relation, we leave \texttt{F} as an unrestricted relation that is
-total on input from the set $[[S]]$ and ...TODO...
-
-This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
-before we even show that our logical relation is
-functional/deterministic, but does not affect the proof structure
-otherwise. We choose to keep the discrepancy since the skolemization
-process is typically expressed in terms of function symbols rather
-than relation symbols.
+\rref{I-Pi} as follows in our Coq mechanization.
+\begin{center}
+  \drule[]{I-PiCoq}
+\end{center}
+It should be easy to verify that the preconditions of
+\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
+equivalent. After establishing Lemma~{lemma:logreldeter}, it is
+possible to further show that the conclusions of the rules are
+equivalent, too.
+
+% This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
+% before we even show that our logical relation is
+% functional/deterministic, but does not affect the proof structure
+% otherwise.
+We choose to keep this discrepancy between the Coq development and the
+logical relation presented in Section~\ref{sec:logreldep} since the
+skolemization process is more intuitively expressed in terms of
+function symbols rather than relation symbols. Otherwise, we do not
+see a clear advantage of \rref{I-PiCoq} over \rref{I-Pi} in set
+theory, where there is no distinction between computable functions and
+functions in general.
 
@@
 defining custom tactics, we rely mostly on off-the-shelf tools such as
-Autosubst 2 and CoqHammer.
+Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.
 
@@
 translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
-\texttt{ctrs:Par} says that the proof involves inverting and
+\texttt{ctrs:Par} say that the proof involves inverting and
 constructing of the derivations of parallel

### 2024-02-15 paper/paper.tex
Add auxiliary functions for substitution

@@
 \orcid{0009-0006-8717-2498}
-\affiliation{%
+\affiliation{
   \institution{University of Pennsylvania}
@@
 \email{liuyiyun@seas.upenn.edu}
+
+\author{Stephanie Weirich}
+\orcid{0000-0002-6756-9168}
+\affiliation{
+  \institution{University of Pennsylvania}
+  \city{Philadelphia}
+  \country{USA}
+}
+\email{sweirich@seas.upenn.edu}
 
@@
 \mathit{Terms}\\
-[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
+[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var i]]\  |\ [[Void]]
                   & \mbox{universes, variables, empty type} \\
@@
 %         \mathit{Renaming} \\
-%         [[i {xi}]] & := & [[xi i]] \\
+%         [[var i {xi}]] & := & [[xi i]] \\
 %         [[(Set i) {xi}]] & := & [[Set i]] \\
@@
 
-%         \mathit{Shift} \\
-%         [[up n A]] & := & A \langle add(n) \rangle \\ \\
-
-%         \mathit{Shift\ one} \\
-%         [[up1 A]] & := & \uparrow^1 [[A]] \\ \\
-
 %         \mathit{Lookup} \\
@@
 
-The syntax of \lang can be found in Figure~\ref{fig:syntax}. We will use
-the unscoped de Bruijn representation for the rest of the paper,
-adopting notations from \citet{autosubst2}, summarized in Figure~\ref{fig:auxdef}.
-Without providing
-the full definition of the renaming and substitution functions, it is
-impossible to tell the binding structure from the syntax
-alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
-with the de Bruijn depth of each term, though we note that the
-syntax we work with is unscoped and the choice does matter when
-we extend our logical relation to open terms in Section~\ref{sec:extension}.
+The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
+the unscoped de Bruijn representation for both our Coq development and
+the informal presentation in our paper. We find de Bruijn
+representation advantageous for specification since it leaves very
+little ambiguity about the variable freeness side
+conditions, making our proof more easily reproducible. Furthermore, as we discuss in
+Section~\ref{sec:automation}, de Bruijn representation is much more
+amenable to automated reasoning in proof assistants.
+We adopt from \citet{autosubst} the notations for simultaneous
+renaming, substitution, and other auxiliary definitions related to our
+syntax, summarized in Figure~\ref{fig:auxdef}.
+
+\begin{figure}[ht]
+  \begin{equation*}
+    \begin{split}
+      \begin{array}{lll}
+        \mathit{Identity (Ren)} \\
+        [[id i]] & := & [[i]] \\ \\
+        \mathit{Identity (Subst)} \\
+        [[idtm i]] & := & [[var i]] \\ \\
+        \mathit{Cons (Ren)} \\
+        ([[xi .: n]])([[0]]) & := & [[n]] \\
+        ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
+        \mathit{Cons (Subst)} \\
+        ([[rho .: a]])([[0]]) & := & [[a]] \\
+        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
+      \end{array}
+    \end{split}
+    \qquad \qquad
+    \begin{split}
+      \begin{array}{lll}
+        \mathit{Curried\ Add} \\
+        [[up n i]] & := & [[n]] + [[i]] \\ \\
+        \mathit{Renaming} \\
+        [[a < xi >]] & := & \ldots \\ \\
+        \mathit{Up} \\
+        [[( up rho ) 0]] & := & [[var 0]] \\
+        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
+        \mathit{Substitution} \\
+        [[var i { rho }  ]] & := &  [[rho i]] \\
+        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
+        [[a { rho }  ]] & := & \ldots \\
+      \end{array}
+    \end{split}
+  \end{equation*}
+  \caption{Auxiliary Functions over Syntax}
+  \label{fig:auxdef}
+\end{figure}
+
+% Without providing
+% the full definition of the renaming and substitution functions, it is
+% impossible to tell the binding structure from the syntax
+% alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
+% with the de Bruijn depth of each term, though we note that the
+% syntax we work with is unscoped and the choice does matter when
+% we extend our logical relation to open terms in Section~\ref{sec:extension}.
 
@@
 elimination, the ability to compute a type using a term as input. For
-example, the function $[[\ Bool if 0 Bool Bool -> Bool]]$ returns
+example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
 either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
@@
 skolemized formulation of \rref{I-Pi} so we immediately obtain a
-well-defined inductive relation and a usable induction
-principle. However, as we will see in Lemma~\ref{lemma:piintroalt}, it
-is possible to show the admissibility of \rref{I-PiAlt} a posteriori.
+well-defined inductive relation and a usable induction principle. The
+slight disadvantage of \rref{I-Pi} is that we need to construct the
+function $[[F]]$ each time we apply it. Therefore, in
+Lemma~\ref{lemma:piintroalt}, we show the admissibility of
+\rref{I-PiAlt} a posteriori and use that as a more convenient
+introduction rule for the rest of our proofs.
 
@@
       [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
-                     } [[InterpR j (up Suc i G i) { rho } S ]] \text{, then } [[rho i in S]] \\
+                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
       [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
@@
   \label{lemma:stvar}
-  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= i : up Suc i G i]]$.
+  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
 \end{lemma}
@@
   \label{lemma:extwn}
-  If $[[wn a i]]$, then $[[wn a]]$.
+  If $[[wn a var i]]$, then $[[wn a]]$.
 \end{lemma}
@@
   By induction over the length of the reduction sequence in $[[wn a
-  i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
+  var i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
   \ref{lemma:parnenf}.

### 2024-02-15 paper/paper.tex
Tweak the notation for renamign and substitution

@@
 renaming, substitution, and other auxiliary definitions related to our
-syntax, summarized in Figure~\ref{fig:auxdef}.
-
+syntax, summarized in Figure~\ref{fig:auxdef}. We omit most of the
+definitions of renaming and substitution and only show the definition
+of a few representative cases of substitution.
 \begin{figure}[ht]
@@
       \begin{array}{lll}
-        \mathit{Identity (Ren)} \\
+        \mathit{IdentityRen} \\
         [[id i]] & := & [[i]] \\ \\
-        \mathit{Identity (Subst)} \\
+        \mathit{IdentityTm} \\
         [[idtm i]] & := & [[var i]] \\ \\
-        \mathit{Cons (Ren)} \\
+        \mathit{ConsRen ([[xi .: n]])} \\
         ([[xi .: n]])([[0]]) & := & [[n]] \\
         ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
-        \mathit{Cons (Subst)} \\
+        \mathit{ConsSubst ([[rho .: a]])} \\
         ([[rho .: a]])([[0]]) & := & [[a]] \\
         ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
+        \mathit{Curried\ Add} \\
+        [[up n i]] & := & [[n]] + [[i]]
       \end{array}
@@
       \begin{array}{lll}
-        \mathit{Curried\ Add} \\
-        [[up n i]] & := & [[n]] + [[i]] \\ \\
-        \mathit{Renaming} \\
-        [[a < xi >]] & := & \ldots \\ \\
-        \mathit{Up} \\
+        \mathit{UpRen} \\
+        [[( up xi ) 0]] & := & [[0]] \\
+        [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
+        \mathit{UpTm} \\
         [[( up rho ) 0]] & := & [[var 0]] \\
         [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
-        \mathit{Substitution} \\
+        \mathit{Renaming ([[a < xi >]])} \\
+        [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
+        [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
+        \ldots \\ \\
+        \mathit{Substitution ([[ a { rho }  ]])} \\
         [[var i { rho }  ]] & := &  [[rho i]] \\
         [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
-        [[a { rho }  ]] & := & \ldots \\
+        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
+        \ldots
       \end{array}

### 2024-02-15 paper/paper.tex
Minor
Add admissible gray boxes

@@
 \usepackage{xspace}
+\usepackage{tcolorbox}
 \usepackage[para]{footmisc}
+\definecolor{lightgray}{gray}{0.75}
 \newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
@@
         [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
-        \mathit{UpTm} \\
-        [[( up rho ) 0]] & := & [[var 0]] \\
-        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
         \mathit{Renaming ([[a < xi >]])} \\
@@
         \ldots \\ \\
+        \mathit{UpTm} \\
+        [[( up rho ) 0]] & := & [[var 0]] \\
+        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
         \mathit{Substitution ([[ a { rho }  ]])} \\
@@
 Figure~\ref{fig:typing} gives the full typing rules for
-\lang{}. \Rref{T-Conv} uses the definition of coherence from earlier
+\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
+admissible syntactically, though some of them are required to
+strengthen the inductive hypothesis of the fundamental theorem.
+In \rref{T-Var}, $[[G i]]$ is the partial function defined
+with the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
+i ]] := [[G i]] $. The precondition $[[i < | G | ]]$, where $[[ | G |
+]]$ represents the length of the context $[[G]]$, ensures that $[[G
+i]]$ has a defined output.
+\Rref{T-Conv} uses the definition of coherence from earlier
 as our equality judgment for type conversion. The use of an untyped
 relation for type conversion makes our formulation slightly different
-from languages such as MLTT, where the judgmental equality typically
+from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
 takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
@@
 untyped equality, and its variant that uses typed judgmental
-equality. Furthermore, it is trivial to embed a system with judgmental
+equality. It is trivial to embed a system with judgmental
 equality to a system with untyped equality by erasing typing
@@
 In this section, we define the logical relation for \lang{} in the
-form of an inductively defined relation (Figure~\ref{fig:logrel}). The
-logical relation takes the form $[[Interp I i A S]]$. The
+form of an inductively defined relation (Figure~\ref{fig:logrel}).
+Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
+The logical relation takes the form $[[Interp I i A S]]$. The
 metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
@@
 The metavariables $[[I]]$ and $[[S]]$ are
-sets with the following signatures:
+sets with the following signatures.
 \begin{equation*}
@@
   \item $[[G |= A : Set i]]$
-  \item For all $[[rho]]$, if $[[rho |= G]]$, then there exists
+  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
     $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
@@
 \begin{lemma}[Par anti-renaming]
-  \label{lemma:parantirenaming} If $[[a { xi } => b0]]$, then there
-exists some $[[b]]$ such that $[[b { xi } = b0]]$ and $[[a => b]]$.
+  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
+exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
 \end{lemma}
@@
 Interestingly, all the properties we have shown in
-Section~\ref{sec:logreldep} and \ref{sec:logrelproof} up to the
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the
 fundamental lemma can be proven in the exact same order, where the new
-cases related due to \rref{I-Ne} and the augmentation of neutral terms
+cases due to \rref{I-Ne} and the augmentation of neutral terms
 to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
@@
   \begin{itemize}
-  \item $[[S]] \in CR_1$ if $[[wne a implies a in S]]$
-  \item $[[S]] \in CR_2$ if $[[a in S implies wn a]]$
-  \item $[[S]] \in CR$ if $[[S]] \in CR_1$ and $[[S]] \in CR_2$
+  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
+  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
+  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
   \end{itemize}
@@
   After unfolding the definition of $[[InterpR i A S]]$, trivial by
-  Lemma~\ref{lemma:cr2ty}, \ref{lemma:crel} and \ref{lemma:cr1ty}.
+  Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
 \end{proof}
@@
   that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
-  We pick the valuation $[[rho]](i) = i$, which injects
-  natural numbers as term variables. This is
-  allowed since Lemma~\ref{lemma:crelrec} says neutral terms,
+  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
+  natural numbers as term variables. The side condition $[[idtm |=
+  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
   including variables, semantically inhabit any $[[S0]]$ where
   $[[S0]]$ is the interpretation of some type. With our choice of
-  $[[rho]]$, we have $[[A {rho} = A]]$ and $[[a {rho} = a]]$. Then we
+  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
   know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
@@
 relation, are in fact required in the normalization proof for simply
-typed languages~\citep{abel2019poplmark}. In the simply typed
+typed languages~\citep{abel2019poplmark}. Similar to the simply typed
 scenario, adequacy needs to be proven before the fundamental theorem
 so we can handle rules such as \rref{T-If} where the scrutinee is a
-neutral term. In a simply typed setting, Lemma~\ref{lemma:extwn} is
-used in the exact same way to show that lambda terms are themselves
-normalizing as we have done in Lemma~\ref{lemma:crel}.
+neutral term. In \citet{abel2019poplmark}, a variant of
+Lemma~\ref{lemma:extwn} is used in the exact same way to show that
+lambda terms are themselves normalizing as we have done in
+Lemma~\ref{lemma:crel}.
 
@@
 \rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
-equivalent. After establishing Lemma~{lemma:logreldeter}, it is
+equivalent. After establishing Lemma~\ref{lemma:logreldeter}, it is
 possible to further show that the conclusions of the rules are
@@
 impredicative sorts can be implemented in a similarly streamlined
-fashion and we will leave those as part of our future work.
+fashion and we will leave those extensions as part of our future work.
 

### 2024-02-15 paper/paper.tex
Add more discussion

@@
 the circularity of subject reduction and $\Pi$-injectivity.
-% described
-% in \citet{siles2012pure}
 Section~\ref{sec:extension} explains how we
@@
 
-\section{Related Work}
+\section{Related Work and Discussion}
 \label{sec:relatedwork}
@@
 uses a Kripke-style logical relation parameterized over an
-type-directed equivalence relation satisfying certain properties. We
-are able to avoid the complexivity of a Kripke-style logical relation
-by defining our logical relation in terms of untyped parallel
-reduction, which does not impose any scoping constraint on the
-terms. \citet{decagda} defines their logical relation using the
+type-directed equivalence relation satisfying certain
+properties. Their logical relation is defined using the
 induction-recursion scheme, which is available in Agda but not in Coq.
@@
 fashion and we will leave those extensions as part of our future work.
+
+% To end this section, we
+
+The final question we want to address is: why is our proof,
+even with the extension from Section~\ref{sec:extension}, so much
+shorter than the proofs from \citet{decagda, nbeincoq,
+  martin-lof-a-la-coq}?
+First, unlike developments that mechanize the correctness of a clever
+type conversion algorithm, we only show the existence of
+normal form for open and closed terms and state our properties in
+terms of the untyped small-step reduction relation. This removes a lot
+of scaffolding in the specification of our type system and our logical
+relation.
+
+This contrasts our development with
+\citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
+difficult property to prove due to the complexity of the equality
+judgment used to define the logical relation. In our system, the
+irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
+direct consequence of Lemma~\ref{lemma:interppreservation}, which is
+straightforward after we prove the classical results about parallel
+reduction in Section~\ref{sec:spec}. In other words, we keep the
+the logical relation itself minimal at the cost of extra lemmas that can be proven
+through syntactic means independently from the logical relation.
+For example, as commented near the end of Section~\ref{sec:extension},
+our normalization property (Corollary~\ref{corollary:exbetanf})
+% can induce a decision procedure for type conversion by reducing two
+% terms to $\beta\eta$-normal form and check for their syntactic
+% equivalence. This decision procedure would be inefficient compared to
+% NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
+% to weak head normal and checks for the equivalence of subterms
+% recursively
+does not immediately induce an efficient decision procedure, but it is
+possible to recover an efficient algorithm by reasoning about untyped
+lambda terms.
+
+% the concision of
+% our proof can be largely attributed to the fact that we bake into the
+% logical relation only the minimum amount of information that is the
+% most conducive to the proof of the metatheoretic result.
+
+Another key to our proof is the very early establishment of the confluence
+property of the paralell reduction relation
+(Lemma~\ref{lemma:pardiamond}). This property is used in
+Lemma~\ref{lemma:interppreservation} to show that the interpretation
+of a type is preserved under evaluation. That is, if $[[InterpR i A
+S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
+In a system with typed directed reduction, the confluence result is
+harder to establish since confluence requires subject reduction, which
+circularly depends on $\Pi$-injectivity, a consequence of
+confluence~\citep{siles2012pure}. There are different workarounds to
+this problem. \citet{siles2012pure} proposes a syntactic approach by
+defining a type system where the confluence result is
+directly provable and later show that the system of interest is
+equivalent to the system with the confluence property.
+\citet{decagda} uses a relational model and the relational counterpart
+of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
+model is parameterized by a typed indexed equivalence relation and
+is reused to prove derive different properties in their
+development. Another perhaps simpler approach is to simply
+extend the judgmental equality with rules about
+$\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
+reduction to be proven independently from confluence and
+$\Pi$-injectivity can later be shown to be admissible. We find the
+approach from \citep{weirich:systemd} the most lightweight as it
+allows us to derive confluence early on even in systems with typed
+conversion.
+
+In cases where the judgmental equality does not leverage type information heavily
+(e.g. $\eta$-law for the unit type), it might be possible to show that
+$\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
+$[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
+property holds (we believe this is likely true for both
+\citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
+terms of untyped parallel reduction to show the decidability of type
+conversion. Otherwise, to fully model the system, we need to bake
+typing information into our logical relation and we need to redefine
+our logical relation in Kripke-style so we can relax the scoping
+constraint in order to prove adequacy, though a relational model would
+still be unnecessary as long as the $\eta$ laws are baked into the
+typed directed relation.
+
+Since confluence plays such as key role in our proof, we are uncertain
+how to generalize our technique to systems where confluence does not
+hold, and we will investigate those scenarios in our future work.
+
+% We are able to avoid the complexivity of a Kripke-style logical relation
+% by defining our logical relation in terms of untyped parallel
+% reduction, which does not impose any scoping constraint on the
+% terms.
+
+% Why is it that the justification of $\eta$ law
+% requires a Kripke-style relational model in all three developments
+% but not ours? Are we ``cheating'' by defining \lang{}'s equational
+% theory using an untyped conversion rule? Instead of answering the
+% questions one by one, we clarify the design of our type system and our
+% proof technique so the answers to those questions are self-evident.
+
+% The key to the shortness of
 

### 2024-02-16 paper/paper.tex
Minor
Add renaming properties of the logical relation

@@
 typing ($[[ G |= a : A]]$), and semantic context well-formedness
-($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
+($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}. In the
+definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
+the typing context obtained by dropping the last $[[i]]$ elements of
+$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
+returns the empty list.
 
@@
 Next, we show some non-trivial cases of the fundamental theorem as
-top-level lemmas and leave the remaining cases as exercises for the reader.
+top-level lemmas and leave the remaining cases as exercises for the
+reader.
+
+First, we formulate the definition of valid renamings and prove that
+semantic typing satisfies renaming. We say that $[[xi]]$ is
+valid from the context $[[G]]$ to the context $[[D]]$ if the
+following condition holds.
+\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
+  [[D ren xi
+  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]
+
+\begin{lemma}[Truncate is valid]
+  \label{lemma:validtruncate}
+  If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i
+  G]]$ to $[[G]]$.
+\end{lemma}
+\begin{proof}
+  By the definition of a valid renaming, we must show that given $[[j
+  < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
+  \begin{itemize}
+  \item $[[ren up i j]] = i + j  < [[| G | ]]$
+  \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
+  \end{itemize}
+  The first bullet point is immediate from the fact that $[[j]] <
+  [[|G|]] - [[i]]$. The second bullet follows from unfolding the
+  definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
+  $[[i]] + [[j]] < [[|G|]]$.
+\end{proof}
+
+\begin{lemma}[Renaming for $[[rho |= G]]$]
+  \label{lemma:renamingval}
+  If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
+  $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
+\end{lemma}
+\begin{proof}
+  Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
+  that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i
+  < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.
+
+  By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
+have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho
+} ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |=
+D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
+S]]$ and $[[rho ren xi i in S]]$. Done.
+\end{proof}
+
+\begin{lemma}[Renaming for $[[G |= a : A]]$]
+  \label{lemma:semrenaming}
+  If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
+  $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
+\end{lemma}
+\begin{proof}
+  Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
+\end{proof}
+
 \begin{lemma}[ST-Var]
@@
 \begin{proof}
-  TODO: formulate renaming
+  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
+need to show that there exists some $[[j]]$ and $[[S]]$ such that
+  \begin{itemize}
+  \item $[[InterpR j G i < up Suc i > { rho } S]]$
+  \item $[[rho i in S]]$
+  \end{itemize}
+  From the definition of $[[|= G]]$, we know that there exists some
+  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. By
+  Lemma~\ref{lemma:validtruncate}, we know that $[[up Suc i]]$ is a
+  valid renaming from $[[drop Suc i G]]$ to $[[G]]$. By
+  Lemma~\ref{lemma:semrenaming}, we deduce $[[G |= G i < up Suc
+  i > : Set j]]$. By the identity from Lemma~\ref{lemma:setinv}, we
+  know that $[[InterpR j G i < up Suc i > { rho } S]]$. It now
+  suffices to show $[[rho i in S]]$, but that is immediate from the
+  definition of $[[rho |= G]]$.
 \end{proof}

### 2024-02-18 paper/paper.tex

@@
 First, we formulate the definition of valid renamings and prove that
-semantic typing satisfies renaming. We say that $[[xi]]$ is
+semantic typing satisfies renaming so we can weaken the context when
+reasoning about the variable case of the fundamental lemma
+(Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G
+++ D]]$, it is easy to show that we can extract some valuation
+$[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
+``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :
+A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
+$[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
+after weakening the context; this implication holds because $[[rho |=
+G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
+can make use of the premise $[[G |= a : A]]$ to derive what we need
+for the conclusion. We recommend the readers to skip the proofs of
+Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
+during the first read as long as they have an intuitive understanding
+of what the renaming property is meant to capture.
+
+We say that $[[xi]]$ is
 valid from the context $[[G]]$ to the context $[[D]]$ if the
@@
 \begin{figure}[h]
-  TODO
+  \begin{tabular}{ c |  c  | c  }
+    & Consistency & Normalization \\
+    \hline
+    Library (Autosubst 2)  & 491 & - \\
+    Syntactic typing (specification) &  69 & - \\
+    Renaming (common)  & 46 & -  \\
+    Syntactic soundness & 660 &  - \\
+    Untyped reduction & 350 & 834 \\
+    Logical relation & 342 & 515 \\
+    Semantic typing and soundness & 169 & 197 \\
+  \end{tabular}
   \caption{Statistics of the Coq Development}
@@
 Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
-extension from Section~\ref{sec:extension}.
+extension to normalization for open and closed terms from
+Section~\ref{sec:extension}.
+
+Since the normalization and the consistency development share the same
+syntactic and typing specification, they differ in only the categories
+that are related to the logical relation. We use - as a marker that
+the line count is the same.
+
+Autosubst 2 takes our syntax specification in higher-order abstract
+syntax and generates Coq syntax specification, renaming and
+substitution functions, and lemmas and tactics that allow reasoning
+about those functions. The auto-generated files and the library header
+files are counted toward the \emph{Library (Autosubst 2)} category.
+
+Orthogonal to the development of our consistency and normalization
+proof, we prove the syntactic soundness of our system through subject
+reduction and progress. The syntactic soundness proof shares the
+definition of renaming with our semantic soundness proof, which is
+factored out as a separate file under the Renaming (common) category.
 

### 2024-02-18 paper/paper.tex
comments throughout the paper

@@
+\newif\ifcomments     %% include author discussion
+\newif\ifanonymous    %% include author identities
+\newif\ifextended     %% include appendix
+\newif\ifsubmission   %% prepare the submitted version
+\newif\ifpublic       %% version available for posting / final version
+
+\commentstrue         %% toggle comments here
+\extendedfalse
+\anonymoustrue
+
+\submissionfalse     %% at most one of these must be true (neither for draft version)
+\publicfalse          %% but if you want to see comments, these should both be off
+
+% If we are going to make a version public, i.e. on arXiv, we should 
+% make sure that there are no comments and our names are on it.
+\ifpublic
+\submissionfalse
+\commentsfalse
+\anonymousfalse
+\fi
+
+%% If we are submission, make sure there are no comments and our names 
+%% are NOT on it.
+\ifsubmission
+\publicfalse
+\commentsfalse
+\anonymoustrue
+\fi
+
 %%
@@
 %%
-\documentclass[acmsmall]{acmart}
+\documentclass[acmsmall,screen=true,
+\ifpublic review=false\else,review=true\fi
+  ,anonymous=\ifanonymous true\else false\fi]{acmart}
 \usepackage{ottalt}
@@
 \usepackage[para]{footmisc}
-\definecolor{lightgray}{gray}{0.75}
+\definecolor{lightgray}{gray}{0.85}
 \newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
@@
     Bib\TeX}}}
+
+\usepackage{draft}
+
+\ifcomments
+\newnote{scw}{blue} % Stephanie Weirich
+\newnote{yl}{purple} % Yiyun Liu
+\else
+\newcommand{\scw}[1]{}
+\newcommand{\yl}[1]{}
+\fi
 
@@
 \citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
-results related to the decidability of type conversion or type
+results\scw{stronger than what?} related to the decidability of type conversion or type
 checking of dependently typed systems, from which consistency can be
@@
 consistency result for dependent types are both well-established,
-there is a severe lack of rigorous and accessible material that shows
+there is a lack of rigorous and accessible material that shows
 how proof by logical relation can be applied to dependently typed
 systems.
+\scw{You are choosing to focus on mechanized logical relations proofs for 
+dependent type theories, not consistency proofs for dependent type theories. 
+Why? (And your title shuld match}
 
@@
 \citet{skorstengaard2019introduction}, \citet{harper2016practical}
-start from logical relations for simply typed languages, and
+start from logical relations for simply typed languages,\scw{proving 
+results such as \ldots} and
 eventually extend the technique to build a relational model for System
 F in order to derive
-parametricity. \citet{harperkripke} gives a gentle introduction on how
+parametricity.
+\scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
+Does Aspinall and Hofmann's chapter of ATTAPL include a logical relation?}
+\scw{Seems like you should also mention POPLmark reloaded somehere too?}
+ \citet{harperkripke} gives a gentle introduction on how
 logical relations for closed terms can be extended to talk about open
@@
 The goal of this paper is to give a tutorial on proof by logical relation for
-dependent types in a simple and digestable format.
+dependent types in a simple and digestable format. \scw{This sentence should come as 
+early as possible.}
 The system we present is small but covers
-important features that are only available or require special treatments in
+important features\scw{list these features explicitly: 
+large eliminations, indexed types, others? \ldots}  that are only available or require special treatments in
 dependently typed systems. We choose boolean
@@
 intensional identity type in our type system to show how indexed
-types are treated in a logical relational proof. Unlike
+types are treated in a logical relational proof. \scw{Talk about 
+identity type first.} Unlike
 \citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
@@
 
-In Section~\ref{sec:extension}, we extend our logical
+In Section~\ref{sec:extension}, \scw{what about sections 2-4?}
+
+ we extend our logical
 relation to prove the stronger property that every well-typed term
@@
 
+\scw{The end of this section needs to include an explicit summar of the 
+contributions of the paper. }
+
 \section{Specification of a Dependent Type Theory}
@@
 \]
-  \caption{Syntax of \lang}
+  \caption{Syntax of \lang \scw{Where does the name come from?}}
   \label{fig:syntax}
@@
 
-The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
+The syntax of \lang can be found in Figure~\ref{fig:syntax}. 
+\scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
+We use
 the unscoped de Bruijn representation for both our Coq development and
-the informal presentation in our paper. We find de Bruijn
+the informal presentation in our paper.  \scw{Give an overview of the figure}
+\scw{Define a convention that separates your use of natural numbers. Sometimes they 
+are universe levels and sometimes they are de Bruijn indices. Better to have separate
+metavariables for each. (And it is possible to generalize universe levels to structures other than natural numbers. Ask Jonathan.}
+We find de Bruijn
 representation advantageous for specification since it leaves very
@@
 renaming, substitution, and other auxiliary definitions related to our
-syntax, summarized in Figure~\ref{fig:auxdef}. We omit most of the
+syntax, summarized in Figure~\ref{fig:auxdef}. 
+\scw{Need to explain the notations in the text. What are these operations? 
+What do readers need to understand about them? Help me understand the figure}
+\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
+indices work.}
+We omit most of the
 definitions of renaming and substitution and only show the definition
@@
         \ldots \\ \\
-        \mathit{UpTm} \\
+        \mathit{UpTm \scw{shifting? lifting?}} \\
         [[( up rho ) 0]] & := & [[var 0]] \\
@@
         [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
+        && \fbox{ \scw{case for $\lambda$ } } \\
         \ldots
@@
  Finally,
-\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
+\lang\scw{Don't begin a sentence with a symbol} also includes an intensional identity type $[[a ~ b : A]]$ whose
 proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
@@
 is the term whose type is to be casted.
+\scw{Example of what you can use identity types for? Or an example of 
+a program that uses J?}
 
 \lang is expressive enough to support large
-elimination, the ability to compute a type using a term as input. For
+eliminations, the ability to compute a type using a term as input. For
 example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
@@
 
+\scw{motivate parallel reduction. Why are you telling us this now. Is this 
+an explanation of definitional equality and its properties?}
 Figure~\ref{fig:par} shows the definition of the parallel reduction
@@
 represent its transitive and reflexive closure, which in turn allows us to define
-the coherence relation $[[a <=> b]]$.
+the coherence relation $[[a <=> b]]$.\scw{where is the terminology ``coherence''
+from? I'm not familiar with it. Maybe you can replace it with ``definitional 
+equivalence'' as that is how it is used in \lang?}
 We say that two terms $[[a]]$
@@
 term $[[c]]$ through parallel reduction. The symmetric notation of
-coherence suggests that it is an equivalence relation.
+coherence suggests\scw{but you prove it below?} it is an equivalence relation.
 
@@
 equivalence relation. It is then trivial to derive that coherence is
-exactly the same as the untyped $\beta$-equivalence.
+exactly the same as untyped $\beta$-equivalence.\scw{cite barendregt?}
 
@@
 strengthen the inductive hypothesis of the fundamental theorem.
+\scw{Is the precondition for T-ABS really admissible? Isn't it required 
+to ensure predicativity? Have you proved it? Have you proved regularity?}
+\scw{Do you have a formation rule for identity types?}
+
 In \rref{T-Var}, $[[G i]]$ is the partial function defined
 with the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
-i ]] := [[G i]] $. The precondition $[[i < | G | ]]$, where $[[ | G |
+i ]] := [[G i]] $.  \scw{explain the renaming in the conclusion of 
+this lemma}
+The precondition $[[i < | G | ]]$, where $[[ | G |
 ]]$ represents the length of the context $[[G]]$, ensures that $[[G
 i]]$ has a defined output.
+
 \Rref{T-Conv} uses the definition of coherence from earlier
@@
 % our proof to handle those rules.
+\scw{Maybe move more of this paragraph and the next to the discussion/related
+  work section, and add a forward pointer here.}
+\scw{Just say that like Coq, the definition of the system uses untyped 
+equality}
 
@@
 and duplicate our typing specification.
+\scw{Is there an advantage for Tarski universes even with infinite 
+hierarchy? or no?}
 
@@
 \section{Logical Relation}
+\scw{This section pre-supposes that we want to define a logical relation, 
+but doesn't precisely state what we want to use it to prove, and why a logical
+relation is suitable. (And why the property you want to prove is difficult to 
+show!) Should add more motivation here.}
+
 \label{sec:logreldep}
@@
 Figure~\ref{fig:syntax}.
+\scw{Because the type of $I$ depends on $i$, I'd prefer a syntax that 
+keeps them in order. }
+\scw{Many introductory texts define the relation as a recursive function 
+over type structure, or step-indices. You use an inductive relation instead, why?}
+\scw{Can we view this inductive relation as the graph of the partial function
+that is defined recursively over types?}
+\scw{What is this form extensible too? impredicative quantification? recursive 
+types? }
+\scw{Is it worth observing here that this definition is not over sets of typed
+terms. That it characterizes all terms that look like booleans (i.e. evaluate to 
+true or false) or all terms that look like proofs (i.e. evaluate to refl). The 
+fact that there is no connection between p and a and b in the I-Eq case is strange
+looking. Need to explain.  }
+
 The metavariables $[[I]]$ and $[[S]]$ are
@@
 system where one can encode Girard's paradox~\citep{girard-thesis}.
+\scw{Spell this out a bit more about what goes wrong. Where does the 
+proof break? Also spell out how this enforces predicativity.}
 
@@
 \label{fig:logrelrec}
+\scw{I didn't understand this figure without reading the text below.}
 \end{figure}
@@
 To derive consistency, it suffices to restrict $[[S]]$ to the set of
-terms to reduce to closed terms. \Rref{I-Void, I-Bool} are
+terms that reduce to closed terms. \Rref{I-Void, I-Bool} are
 unsurprising; when considering only closed terms, the empty type
@@
 justify the skolemization process~\citep{skolemization}.
+\scw{Is this true constructively? If not, is that ok?}
 \[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
@@
 The relation $[[Interp I i A S]]$ satisfies the following inversion
-principles.
+principles.\scw{Put footnotes on lemma name, or on (1), (2), etc}
 \begin{lemma}[Inversion of the logical relation]
@@
 \begin{proof}
+  \scw{Why is this proof interesting and worth typesetting? Should
+    it be in an appendix so that it doesn't interrupt the flow of 
+    reading?}
   We only show the most involved function case; the rest follows a
@@
     \end{itemize}
-    These properties are exactly what's required in the conclusion.
+    These properties are exactly what's\scw{avoid contractions} required in the conclusion.
   \end{description}
@@
 interpretation.
+\scw{I'm not used to this lemma being called irrelevance. Intuitively, 
+it says that the definition of the logical relation respects
+definitional equality. }
 \begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
@@
 \end{corollary}
+\scw{Need a transition here}
 Next, we show that $[[Interp I i A S]]$ is in fact a partial function.
@@
 \begin{proof}
-  Immedaite from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
+  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
 \end{proof}
@@
 all closed under expansion.
+\scw{How is this related to the comment before 3.2 (i.e. the property
+derived from I-Red?}
 \begin{lemma}[Logical Relation Elements back preservation\footnote{\dotv{semtyping.v}{InterpExt\_back\_clos}}]
@@
 boolean and equality, the output set $[[S]]$ contains only terms that
-evaluate to closed terms. To generalize our logical relation to open
+evaluate to closed terms.
+
+\scw{What is the motivation for this section? Why do we want to generalize
+the logical relation?}
+ To generalize our logical relation to open
 terms, we define the semantic typing judgment by closing the open
@@
 typing ($[[ G |= a : A]]$), and semantic context well-formedness
-($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}. In the
+($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}. 
+\scw{explain all parts of this figure. it's pretty intimidating.}
+In the
 definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
@@
 work with.
-\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}]
+\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}\scw{spell out the whole word}]
   \label{lemma:setinv}
@@
   Pick $[[Suc i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
-  ]]$ for $[[S0]]$ and it's trivial to verify the conditions hold.
+  ]]$ for $[[S0]]$ and it is trivial to verify the conditions hold.
 \end{proof}
@@
 The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
-counterpart $[[|- G]]$, is defined through a for all quantified statement rather
-than inductively over the context. It is easy to recover the same
+counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
+than inductively over the context. \scw{why is that the case?} It is easy to recover the same
 structural rules:
@@
 \end{lemma}
+\scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about 
+Ctx-empty?}
 \begin{proof}
@@
 top-level lemmas and leave the remaining cases as exercises for the
-reader.
+reader.\scw{I dislike leaving ``exercises to the reader''}
 
@@
 
+\scw{Make this a definition? Would it be useful to create notation 
+  for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
 We say that $[[xi]]$ is
@@
   $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
+\scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
 \end{lemma}
@@
 D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
-S]]$ and $[[rho ren xi i in S]]$. Done.
+S]]$ and $[[rho ren xi i in S]]$. 
 \end{proof}
@@
 \end{proof}
+
+\scw{Need a transition here that you are starting to explain the semantic typing 
+rules.}
 
@@
 \begin{proof}
-  Immedaite by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
+  Immediate by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
 \end{proof}
@@
 
+\scw{Spell this out in a little more detail.}
+
 \begin{corollary}[Logical Consistency]
@@
 
+\scw{Put this introduction before the figure in the section.}
 In this section, we sketch out how the logical relation from
@@
 \end{center}
+\scw{What happened to the domain type in the $\lambda$ expression? This is 
+not your syntax...}
 \Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
@@
 $\beta$-normal form through parallel reduction and are defined as
-follows.
+follows.\scw{What does $[[wne]]$ stand for?}
 \[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
@@
 
+\scw{What does CR stand for?}
 \begin{definition}[CR]
@@
 
+\scw{Name and foreshadow these lemmas when they are introduced so that 
+it is easier to refer to them in the discussion.}
 Finally, we want to point out that
@@
 definition of the logical relation in Figure~\ref{fig:logrel} requires
-the impredicativity of Coq's \texttt{Prop} sort since in \rref{I-Pi},
+the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
 the function $[[F]]$ can be later instantiated into the logical
@@
 possible to further show that the conclusions of the rules are
-equivalent, too.
+equivalent, too.\scw{If it is easy, you should have already done it. 
+Unless it is really long and boring.}
 
@@
 We choose to keep this discrepancy between the Coq development and the
-logical relation presented in Section~\ref{sec:logreldep} since the
+description of the logical relation presented in Section~\ref{sec:logreldep} since the
 skolemization process is more intuitively expressed in terms of
@@
 theory, where there is no distinction between computable functions and
-functions in general.
+functions in general. \scw{not sure I follow this last bit}
 
@@
 produce Coq syntax files in de Bruijn representation. Additionallly,
-Autosubst 2 provides a powerful tactic \texttt{asimpl} which
+Autosubst 2 provides a powerful tactic \texttt{asimpl} that
 can be used to prove the equivalence of two terms constructed using
@@
 by \texttt{asimpl} without having to manually prove or even directly invoke any
-substitution related lemmas.
+substitution related lemmas. \scw{You are only referring to lemmas about 
+how substitution relates to syntax, right? You still need to manually 
+invoke lemmas about substitution and judgements.}
 
@@
 property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
+\scw{you have space, so repeat the mathematical statement of the lemma}
 \begin{minted}{coq}
@@
 reduction as lemmas.
+\scw{What does \texttt{lq:on} and \texttt{hauto} mean?}
 
@@
 \label{sec:relatedwork}
+
+\scw{Maybe it would be useful to include a chart here, so that readers can
+  easily keep track of the features of the various languages.  i.e. which ones
+  include large eliminations? type-directed equivalence? impredicative prop?
+  inductive datatypes? what are their line counts?}
+
 \citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
@@
 works that establish metatheoretic results strong enough to derive
-consistency for dependently typed systems. Some of these techniques
+consistency for dependently typed systems. 
+\scw{Should give Martin L\"of credit for first consistency proof.}
+\scw{What techniques do these three proofs use? How is it different from 
+your proof? What about Luo, which 
+introduced a universe hierarchy? What about categories with families, or 
+other ways of showing consistency?}
+Some of these techniques
 are still in use in more recent works. For example, the technique for
@@
 relation and therefore a relational model is not needed.
+\scw{would it be difficult to extend your proof to be a PER?}
 
@@
 % To end this section, we
-
+\scw{should this be a new subsection?}
 The final question we want to address is: why is our proof,
@@
 irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
-direct consequence of Lemma~\ref{lemma:interppreservation}, which is
-straightforward after we prove the classical results about parallel
+direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to 
+lemma name} which is
+straightforward after we prove the results about parallel
 reduction in Section~\ref{sec:spec}. In other words, we keep the
@@
 allows us to derive confluence early on even in systems with typed
-conversion.
+conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}
 
@@
 still be unnecessary as long as the $\eta$ laws are baked into the
-typed directed relation.
+typed directed relation. \scw{not sure I understand this last bit}
 
@@
 hold, and we will investigate those scenarios in our future work.
+\scw{What systems are you thinking of? Or do you mean more generally 
+that this technique only works for languages where the definition of 
+equality is based on joinability of parallel reduction?}
 

### 2024-02-18 paper/paper.tex
Fix citation for takahashi

@@
 $\beta$-normal form through parallel reduction and are defined as
-follows.\scw{What does $[[wne]]$ stand for?}
+follows.\scw{What does $wne$ stand for?}
 \[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]

### 2024-02-18 paper/paper.tex
Minor

@@
   \begin{enumerate}
-  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}}
-  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}}
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
   \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}}

### 2024-02-19 paper/paper.tex
Cite tapl; fix universe hierarchy

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{A Short and Mechanized Consistency Proof for Dependent Type Theory}
+\title{Short and Mechanized Logical Relation for Dependent Type Theory}
 
@@
 available in the literature. In particular, recent works such as \citet{nbeincoq},
-\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
-results\scw{stronger than what?} related to the decidability of type conversion or type
-checking of dependently typed systems, from which consistency can be
-derived as a corollary.
+\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize the
+correctness of the NbE algorithm, the decidability of type
+conversion, decidability of type
+checking respectively for dependently typed systems. From these
+properties, consistency can be derived as a corollary.
 
@@
 
-Introductory materials about logical relations such as
-\citet{skorstengaard2019introduction}, \citet{harper2016practical}
-start from logical relations for simply typed languages,\scw{proving 
-results such as \ldots} and
-eventually extend the technique to build a relational model for System
-F in order to derive
-parametricity.
+Introductory materials about logical relations or standard textbooks such as
+\citet{skorstengaard2019introduction}, \citet{harper2016practical},
+and \citet{pierce2002types}
+talk about logical relations for simply or polymorphically typed
+languages, proving results such as termination for closed terms and parametricity.
 \scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
-Does Aspinall and Hofmann's chapter of ATTAPL include a logical relation?}
+Does Aspinall and Hofmann's chapter of ATTAPL include a logical
+relation?}
+\yl{skorstengaard is from the OPLSS notes. Cited ATTAPL}
+\yl{Don't have a copy of }
 \scw{Seems like you should also mention POPLmark reloaded somehere too?}
- \citet{harperkripke} gives a gentle introduction on how
-logical relations for closed terms can be extended to talk about open
-terms so we can derive properties such as normalization for well-typed
-open terms. Overall, the introductory texts about logical relations
+\citet{pierce2004advanced}, \citet{harperkripke}, and \citet{abel2019poplmark} show how
+a Kripke-style logical relation can be used to include scoping
+information in the logical relation and
+derive properties such as the existence of normal form or strong
+normalization for open and closed terms in simply typed languages.
+Overall, the introductory texts about logical relations
 cover systems and properties with varying degrees of complexity, from
@@
 \drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
-  Refl, J, If, Bool, True, False, Void}
+  Refl, Eq, J, If, Bool, True, False, Void}
 \end{minipage}
@@
 to ensure predicativity? Have you proved it? Have you proved regularity?}
+\yl{I proved regularity. \Rref{T-Abs} is admissible if cumulativity
+  was formulated correctly (which wasn't the case in the version
+  of the paper you read). I revised \rref{T-Univ} so the premise is
+  not admissible }
 \scw{Do you have a formation rule for identity types?}
+\yl{Added}
 

### 2024-02-19 paper/paper.tex
Fix duplicate lambda
Add text in intro about section 3

@@
 dependent types in a simple and digestable format. \scw{This sentence should come as
-early as possible.}
-The system we present is small but covers
-important features\scw{list these features explicitly: 
-large eliminations, indexed types, others? \ldots}  that are only available or require special treatments in
-dependently typed systems. We choose boolean
-types over natural numbers as our base type for simplicity, but include an
-intensional identity type in our type system to show how indexed
-types are treated in a logical relational proof. \scw{Talk about 
-identity type first.} Unlike
-\citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
-\citet{anand2014towards}, we include an infinite hierarchy of
-universes to not only support type-level computations, but also avoid
-the unnecessary code duplication pointed out by \citet{nbeincoq} when
-big and small types are treated non-uniformly.
+early as possible.} Dependently typed systems comes in many flavors and
+vary greatly in their expressiveness.
+In Section~\ref{sec:spec}, we introduce \lang{}, a dependently typed
+language that is small but relatively complete in features \scw{list these features explicitly:
+large eliminations, indexed types, others? \ldots} related to
+dependently types, including large elimination, a full universe
+hierarchy, an intensional identity type and a boolean base type, both
+of which are equipped with dependent elimination forms (e.g. J eliminator
+for the identity type). In particular, the inclusion of identity
+types, which is absent from \citet{decagda}, \citet{nbeincoq},
+helps demonstrate how indexed types are handled when defining the
+logical relation.
+
+% We choose boolean
+% types over natural numbers as our base type for simplicity, but include an
+% intensional identity type in our type system to show how indexed
+% types are treated in a logical relational proof. \scw{Talk about
+% identity type first.} Unlike
+% \citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
+% \citet{anand2014towards}, we include an infinite hierarchy of
+% universes to not only support type-level computations, but also avoid
+% the unnecessary code duplication pointed out by \citet{nbeincoq} when
+% big and small types are treated non-uniformly.
+In Section~\ref{sec:logreldep}, we give the definition of our logical
+relation as an inductively defined relation. We first show that the
+logical relation respects evaluation (referred to as the irrelevance
+property in \citet{martin-lof-a-la-coq}), then proceed to
+prove that the logical relation corresponds to a partial function in
+the sense that the each type has a unique interpretation if there
+exists one.
+
+In Section~\ref{sec:logrelproof}, we use the standard technique that
+is commonly seen in ...
 
@@
 In Section~\ref{sec:extension}, \scw{what about sections 2-4?}
-
  we extend our logical
@@
                     |\ [[f ~ f : f]]\\
-            & |   & [[\\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
+            & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
     \end{array}

### 2024-02-20 paper/paper.tex
Rewrite the first part of the intro

@@
 \section{Introduction}
+In this paper, we present a short and mechanized consistency proof by
+logical relation for \lang{}, a Martin-Löf style dependent type theory
+with large elimination, a full predicative universe hierarchy, an
+intensional identity type and a boolean base type, both of which
+equipped with dependent elimination forms. % The mechanization is
+% extremely short and involves less than 1000 lines of manually written
+% Coq code.
+
+Our goal is to show the fundamental technique required to carry out proof by logical
+relation in a dependently typed setting where logical relation can no
+longer be defined as a simple recursive function over the structure of
+types. Compared to the many mechanized proofs by logical relation for
+predicative dependent type
+theory~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards},
+whose proof developments range from 20,000 to 400,000 lines of code,
+our Coq development is the most lightweight at less than 1,000 lines.
+The big difference in the size of the developments does not
+necessarily imply that our proof technique leads to a more concise
+proof due to the differing expressiveness of the languages and the
+different metatheoretic results being established. For
+example, \citet{anand2014towards} mechanizes the metatheory of
+Nuprl~\citep{constable1986implementing}, which, on top of all the
+features that \lang{} supports, includes W-types and partial types and
+requires a PER semantics to model its extensional typed
+equality. The object language from \citet{decagda} does not support
+identity types and only has one predicative universe, but additionally
+supports $\Sigma$ types. However, for a researcher who wishes to learn
+the underlying techniques for mechanized logical relation for
+dependent type theory, a small development like ours is much easier to
+navigate and understand. Furthermore, we show how our simple
+consistency proof can be easily extended to show the existence of
+$\beta\eta$-normal form for well-typed open \emph{and} closed terms,
+giving us a metatheoretic result almost as strong as the one from
+\citet{decagda} at the moderate cost of around 500 lines of extra code.
+
+We present our definitions and proofs in set theory notation and
+terminology rather than in Coq syntax to make the material accessible
+to readers with general mathematical background. However, we make the
+typeset definitions as close as possible to our mechanization so the
+definitions and the proofs can be easily encoded in a proof
+assistant.
+We find the close correspondence
+between our typeset presentation and the Coq development advantageous
+for a couple of reasons.
+
 % Depending on its application, we care about certain metatheoretic

### 2024-02-20 paper/paper.tex
Minor

@@
 terminology rather than in Coq syntax to make the material accessible
-to readers with general mathematical background. However, we make the
+to readers with general mathematical background. However, notation
+aside, the typeset representation is purposefully made to be closer to
+our mechanization so less massaging of the definitions is required to
+encode our proofs in a proof assistant.  The close correspondence
+between a typeset proof and a mechanized proof comes at the potential
+risk that the artifacts specific to theorem provers are reflected upon
+the presented proof. However, for proofs by logical relation for
+dependent types, we find our approach advantageous for a few
+reasons.
+First, unlike properties that are derivable through
+syntactic means, proof by logical relation is demanding on the
+strength of the metatheory. An informal proof that attempts to be
+agnostic or merely ambiguous about the underlying metatheory
+
+However, we make the
 typeset definitions as close as possible to our mechanization so the
@@
 between our typeset presentation and the Coq development advantageous
-for a couple of reasons.
+for a couple of reasons. First, compared to other proof techniques,
+such as the use of preservation and progress to derive syntactic
+soundness, defining a logical relation for a dependently typed
+language is demanding on the strength of the metatheory.
 

### 2024-02-20 paper/paper.tex
Rewrite the intro

--- 2024-02-20 paper/paper.tex
@@
 our Coq development is the most lightweight at less than 1,000 lines.
+
 The big difference in the size of the developments does not
@@
 terminology rather than in Coq syntax to make the material accessible
-to readers with general mathematical background. However, we make the
-typeset definitions as close as possible to our mechanization so the
-definitions and the proofs can be easily encoded in a proof
+to readers with general mathematical background.
+However, notation aside, the typeset representation is
+purposefully made to be closer to our mechanization so less massaging
+of the definitions is required to encode our proofs in a proof
 assistant.
-We find the close correspondence
-between our typeset presentation and the Coq development advantageous
-for a couple of reasons.
-
-% Depending on its application, we care about certain metatheoretic
-% properties about a type system. As a programming language, we may care
-% about type soundness, which states that a well-typed never gets stuck
-% during evaluation.
-When a dependently type system is used as a program logic where terms
-encode proofs, we want our type system to be logically consistent,
-meaning that the empty type is not inhabited.
-
-The consistency proofs of various dependently typed systems, including Martin-Lof's
-type theory and the Calculus of Constructions, have long been
-available in the literature. In particular, recent works such as \citet{nbeincoq},
-\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize the
-correctness of the NbE algorithm, the decidability of type
-conversion, decidability of type
-checking respectively for dependently typed systems. From these
-properties, consistency can be derived as a corollary.
-
-The underlying technique of the forementioned
-works is proof by logical relation,
-which involves interpreting types as reducibility predicates,
-representing sets of terms satisfying certain properties with respect
-to the reduction relation. While the proof technique and the
-consistency result for dependent types are both well-established,
-there is a lack of rigorous and accessible material that shows
-how proof by logical relation can be applied to dependently typed
-systems.
-\scw{You are choosing to focus on mechanized logical relations proofs for
-dependent type theories, not consistency proofs for dependent type theories.
-Why? (And your title shuld match}
-
-Introductory materials about logical relations or standard textbooks such as
-\citet{skorstengaard2019introduction}, \citet{harper2016practical},
-and \citet{pierce2002types}
-talk about logical relations for simply or polymorphically typed
-languages, proving results such as termination for closed terms and parametricity.
-\scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
-Does Aspinall and Hofmann's chapter of ATTAPL include a logical
-relation?}
-\yl{skorstengaard is from the OPLSS notes. Cited ATTAPL}
-\yl{Don't have a copy of }
-\scw{Seems like you should also mention POPLmark reloaded somehere too?}
-\citet{pierce2004advanced}, \citet{harperkripke}, and \citet{abel2019poplmark} show how
-a Kripke-style logical relation can be used to include scoping
-information in the logical relation and
-derive properties such as the existence of normal form or strong
-normalization for open and closed terms in simply typed languages.
-Overall, the introductory texts about logical relations
-cover systems and properties with varying degrees of complexity, from
-simply typed to polymorphicly typed, logical predicate to logical
-equivalence, closed terms to open terms.
-
-The glaring gap here is the
-lack of fully dependently typed systems where computations may appear
-at the type-level. It is far from obvious why proof by
-logical relation is even applicable to dependent types, since the type
-may very well-be a computation that is yet to be evaluated.
-While it is assuring that proof by logical relations for dependent
-types is available in mechanized forms, % the key to address the
-% complexities of dependent types is obscured in
-\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
-Kripke-style models that obscure the technique for addressing
-type-level computation. The added complexivity is evident
-from the size of their code base. All three developments involve 20
-thousand to 30 thousand lines of Coq or Agda code.
-
-The goal of this paper is to give a tutorial on proof by logical relation for
-dependent types in a simple and digestable format. \scw{This sentence should come as
-early as possible.} Dependently typed systems comes in many flavors and
-vary greatly in their expressiveness.
-In Section~\ref{sec:spec}, we introduce \lang{}, a dependently typed
-language that is small but relatively complete in features \scw{list these features explicitly:
-large eliminations, indexed types, others? \ldots} related to
-dependently types, including large elimination, a full universe
-hierarchy, an intensional identity type and a boolean base type, both
-of which are equipped with dependent elimination forms (e.g. J eliminator
-for the identity type). In particular, the inclusion of identity
-types, which is absent from \citet{decagda}, \citet{nbeincoq},
-helps demonstrate how indexed types are handled when defining the
-logical relation.
-
-% We choose boolean
-% types over natural numbers as our base type for simplicity, but include an
-% intensional identity type in our type system to show how indexed
-% types are treated in a logical relational proof. \scw{Talk about
-% identity type first.} Unlike
-% \citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
-% \citet{anand2014towards}, we include an infinite hierarchy of
-% universes to not only support type-level computations, but also avoid
-% the unnecessary code duplication pointed out by \citet{nbeincoq} when
-% big and small types are treated non-uniformly.
-In Section~\ref{sec:logreldep}, we give the definition of our logical
-relation as an inductively defined relation. We first show that the
-logical relation respects evaluation (referred to as the irrelevance
-property in \citet{martin-lof-a-la-coq}), then proceed to
-prove that the logical relation corresponds to a partial function in
-the sense that the each type has a unique interpretation if there
-exists one.
-
-In Section~\ref{sec:logrelproof}, we use the standard technique that
-is commonly seen in ...
-
-Our consistency proof is short and fully mechanized. The proof scripts
-involve less than 1000 lines of manually written Coq code. In fact, what we find
-encouraging is that among the 1000
-lines of Coq code, 400 lines are related to the specification of the
-type system, semantics, and properties related to untyped lambda
-terms. The semantic type soundness proof through logical relation
-takes almost the same amount of code as our syntactic type soundness
-proof!
-Thanks to the conciseness of
-the proof, we are able to present it in detail in
-Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
-structure of our mechanization closely corresponds to the proof we
-present in the text, enabling us to label at the footnote each lemma
+The close correspondence between a typeset proof and a mechanized
+proof comes at the potential risk that the artifacts specific to
+theorem provers may become visible in the presented proof.
+However, for proofs by logical relation for
+dependent types, we find our approach advantageous for a few reasons.
+First, unlike properties that are derivable through
+syntactic means, proof by logical relation is demanding on the
+strength of the metatheory. An informal proof that attempts to be
+agnostic or ambiguous about the underlying metatheory % may struggle to even
+% convey the validity of their definition and
+requires substantial
+effort from the reader to even understand whether it is theoretically
+possible to encode such proofs in a proof assistant.
+Second, while a proof assistant may reject perfectly valid
+definitions, it helps us keep our definition precise and
+unambiguous. A precise definition is crucial as the reasoning
+principles we are allowed to use is directly tied to the definition,
+which is sometimes obscured by concise notations that diverge too far
+from a formal proof.
+Finally, our approach does not result in a verbose proof on paper, thanks to
+the already short mechanization. On top of that, We make conservative
+simplifications to our typeset proof such as avoiding the clunky
+syntax for well-founded recursion in Coq, but we keep the overall
+structure of our typeset proofs consistent with our mechanization.
+The consistency enables us to label at the footnote each lemma
 directly to their counterpart in the proof script for the readers to
@@
 
-The technique we use is most similar to the one from \citet{nbeincoq},
-which leverages impredicativity to define the logical
-relation as a partial function.
-Rather than framing impredicativity as a
-mechanism for encoding induction
-recursion~\citep{induction-recursion-dybjer}, a scheme in which
-semantic models (including logical relations) for dependent types can
-be defined, we opt for a direct explanation through the informal
-language of sets, where impredicativity manifests in the form of
-second-order logical formulas and thus more intuitive to grasp
-for readers with a general mathematical background.
-
-In Section~\ref{sec:extension}, \scw{what about sections 2-4?}
- we extend our logical
-relation to prove the stronger property that every well-typed term
-has a normal form. Most interestingly, we find that extending the
-language with
-$\eta$ rule for functions and the metatheoretic result to normalization for open terms are
-mostly orthogonal to the fact that the language is dependently
-typed. In our mechanized proof, the overall
-structure of the logical relation remains unchanged, and the additional
-proof obligations are mostly related to properties about the untyped
-lambda calculus, all of which can be derived through syntactic
-means.
-In other words, once we have established the base technique for
-proof by logical relation for dependent types, we can factor out the
-complexities of an extension the are not specifically related to dependent
-types; such an extension, if desired, may be studied in the context of
-a simply typed language and later ported into a dependently typed setting.
+In Section~\ref{sec:spec}, we introduce \lang{}, the object language
+of interest, and establish some elementary properties about parallel
+reduction, which will be used as the equational theory of our language
+and play a central role in defining our logical relation. The design
+of \lang{} is reminiscent of Martin-Löf style predicatvie type
+systems. The untyped conversion rule of \lang{} is inspired by Pure
+Type Systems~\citet{barendregt1991introduction}, though we discuss how
+our proof can be easily adjusted for systems with typed judgmental
+equality.  We design \lang{} to incorporate some of the most common
+features of dependent type theory, including an intensional identity
+type, a boolean type, large elimination, dependent elimination, and a
+full universe hierarchy. We believe a good coverage of features will
+make our proof applicable to a broad range of type systems.
+
+In Section~\ref{sec:logreldep}, we formulate the consistency property
+and motivate why proof by logical relation is needed to derive it.
+We define our logical relation as an inductively defined relation. We first
+show that the logical relation respects evaluation (referred to as the
+irrelevance property in \citet{martin-lof-a-la-coq}), then proceed to
+prove that the logical relation is a partial function; that is, each
+type can have at most one unique interpretation.
+
+In Section~\ref{sec:logrelproof}, we give the definition of semantic
+typing in terms of our logical relation so we can state and prove the
+fundamental theorem, which states that syntactic typing implies
+semantic typing, from which consistency follows as a corollary.
+
+In Section~\ref{sec:extension}, we extend our logical relation to
+prove the stronger property that every well-typed term has a
+$\beta\eta$-normal form. After laying the foundation for handling
+type-level computation in Section~\ref{sec:logreldep}, the extension
+to include open terms in the logical relation requries mechanical
+changes that directly mirrors the same extension for a simply typed
+language. We use this extension to demonstrate the idea that once we
+have established the base technique for proof by logical relation for
+dependent types, we can factor out the complexities of an extension
+the are not specifically related to dependent types; such an
+extension, if desired, may be studied in the context of a simply typed
+language and later ported into a dependently typed setting. We believe
+it is possible to extend our logical relation to a Kripke-style and
+relational model in a similar fashion.
 
@@
 mechanization, including our use of existing libraries such as
-Autosubst 2~\citep{autosubst2} for handling bindings and CoqHammer~\citep{czajka2018hammer} for general-purpose
-automation.
-In Section~\ref{sec:relatedwork}, we discuss how our proof
-relates to existing proofs by logical relations for dependent types in
-the literature.
+Autosubst 2~\citep{autosubst2} for handling bindings and
+CoqHammer~\citep{czajka2018hammer} for general-purpose automation.
+
+In Section~\ref{sec:relatedwork}, we discuss how our proof relates to
+existing proofs by logical relations and other proof techniques for
+proving consistency and normalization.
 
@@
 proofs for dependent types.
+% % For
+% % example, \citet{geuvers1994short} requires impredicativity to encode
+% % Calculus of Constructions, an impredicative dependent type theory, but
+% % also relies on the classification property to define the logical
+% % relation over well-formed types and kinds. Their simple
+% % presentation of the logical relation as an inductively defined
+% % function over types, reminiscent of logical relations for simply typed
+% % languages,
+
+% However, we make the
+% typeset definitions as close as possible to our mechanization so the
+% definitions and the proofs can be easily encoded in a proof
+% assistant.
+% We find the close correspondence
+% between our typeset presentation and the Coq development advantageous
+% for a couple of reasons. First, compared to other proof techniques,
+% such as the use of preservation and progress to derive syntactic
+% soundness, defining a logical relation for a dependently typed
+% language is demanding on the strength of the metatheory.
+
+% % Depending on its application, we care about certain metatheoretic
+% % properties about a type system. As a programming language, we may care
+% % about type soundness, which states that a well-typed never gets stuck
+% % during evaluation.
+% When a dependently type system is used as a program logic where terms
+% encode proofs, we want our type system to be logically consistent,
+% meaning that the empty type is not inhabited.
+
+% The consistency proofs of various dependently typed systems, including Martin-Lof's
+% type theory and the Calculus of Constructions, have long been
+% available in the literature. In particular, recent works such as \citet{nbeincoq},
+% \citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize the
+% correctness of the NbE algorithm, the decidability of type
+% conversion, decidability of type
+% checking respectively for dependently typed systems. From these
+% properties, consistency can be derived as a corollary.
+
+% The underlying technique of the forementioned
+% works is proof by logical relation,
+% which involves interpreting types as reducibility predicates,
+% representing sets of terms satisfying certain properties with respect
+% to the reduction relation. While the proof technique and the
+% consistency result for dependent types are both well-established,
+% there is a lack of rigorous and accessible material that shows
+% how proof by logical relation can be applied to dependently typed
+% systems.
+% \scw{You are choosing to focus on mechanized logical relations proofs for
+% dependent type theories, not consistency proofs for dependent type theories.
+% Why? (And your title shuld match}
+
+% Introductory materials about logical relations or standard textbooks such as
+% \citet{skorstengaard2019introduction}, \citet{harper2016practical},
+% and \citet{pierce2002types}
+% talk about logical relations for simply or polymorphically typed
+% languages, proving results such as termination for closed terms and parametricity.
+% \scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
+% Does Aspinall and Hofmann's chapter of ATTAPL include a logical
+% relation?}
+% \yl{skorstengaard is from the OPLSS notes. Cited ATTAPL}
+% \yl{Don't have a copy of }
+% \scw{Seems like you should also mention POPLmark reloaded somehere too?}
+% \citet{pierce2004advanced}, \citet{harperkripke}, and \citet{abel2019poplmark} show how
+% a Kripke-style logical relation can be used to include scoping
+% information in the logical relation and
+% derive properties such as the existence of normal form or strong
+% normalization for open and closed terms in simply typed languages.
+% Overall, the introductory texts about logical relations
+% cover systems and properties with varying degrees of complexity, from
+% simply typed to polymorphicly typed, logical predicate to logical
+% equivalence, closed terms to open terms.
+
+% The glaring gap here is the
+% lack of fully dependently typed systems where computations may appear
+% at the type-level. It is far from obvious why proof by
+% logical relation is even applicable to dependent types, since the type
+% may very well-be a computation that is yet to be evaluated.
+% While it is assuring that proof by logical relations for dependent
+% types is available in mechanized forms, % the key to address the
+% % complexities of dependent types is obscured in
+% \citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
+% Kripke-style models that obscure the technique for addressing
+% type-level computation. The added complexivity is evident
+% from the size of their code base. All three developments involve 20
+% thousand to 30 thousand lines of Coq or Agda code.
+
+% The goal of this paper is to give a tutorial on proof by logical relation for
+% dependent types in a simple and digestable format. \scw{This sentence should come as
+% early as possible.} Dependently typed systems comes in many flavors and
+% vary greatly in their expressiveness.
+% In Section~\ref{sec:spec}, we introduce \lang{}, a dependently typed
+% language that is small but relatively complete in features \scw{list these features explicitly:
+% large eliminations, indexed types, others? \ldots} related to
+% dependently types, including large elimination, a full universe
+% hierarchy, an intensional identity type and a boolean base type, both
+% of which are equipped with dependent elimination forms (e.g. J eliminator
+% for the identity type). In particular, the inclusion of identity
+% types, which is absent from \citet{decagda}, \citet{nbeincoq},
+% helps demonstrate how indexed types are handled when defining the
+% logical relation.
+
+% % We choose boolean
+% % types over natural numbers as our base type for simplicity, but include an
+% % intensional identity type in our type system to show how indexed
+% % types are treated in a logical relational proof. \scw{Talk about
+% % identity type first.} Unlike
+% % \citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
+% % \citet{anand2014towards}, we include an infinite hierarchy of
+% % universes to not only support type-level computations, but also avoid
+% % the unnecessary code duplication pointed out by \citet{nbeincoq} when
+% % big and small types are treated non-uniformly.
+
+% In Section~\ref{sec:logrelproof}, we use the standard technique that
+% is commonly seen in ...
+
+% Our consistency proof is short and fully mechanized. The proof scripts
+% involve less than 1000 lines of manually written Coq code. In fact, what we find
+% encouraging is that among the 1000
+% lines of Coq code, 400 lines are related to the specification of the
+% type system, semantics, and properties related to untyped lambda
+% terms. The semantic type soundness proof through logical relation
+% takes almost the same amount of code as our syntactic type soundness
+% proof!
+% Thanks to the conciseness of
+% the proof, we are able to present it in detail in
+% Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
+% structure of our mechanization closely corresponds to the proof we
+% present in the text, enabling us to label at the footnote each lemma
+% directly to their counterpart in the proof script for the readers to
+% reference and validate.
+
+% The technique we use is most similar to the one from \citet{nbeincoq},
+% which leverages impredicativity to define the logical
+% relation as a partial function.
+% Rather than framing impredicativity as a
+% mechanism for encoding induction
+% recursion~\citep{induction-recursion-dybjer}, a scheme in which
+% semantic models (including logical relations) for dependent types can
+% be defined, we opt for a direct explanation through the informal
+% language of sets, where impredicativity manifests in the form of
+% second-order logical formulas and thus more intuitive to grasp
+% for readers with a general mathematical background.
 
@@
 \mathit{Terms}\\
-[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var i]]\  |\ [[Void]]
+[[a]],[[b]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                   & \mbox{universes, variables, empty type} \\

### 2024-02-20 paper/paper.tex
Revise the text about de bruijn related functions (WIP)

@@
                   & \mbox{if} \\ \\
-\mathit{Renaming}\\
-[[xi]] & \in & [[SNat -> SNat]] & \\ \\
+% \mathit{Renaming}\\
+% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
 \mathit{Substitution}\\
@@
 the unscoped de Bruijn representation for both our Coq development and
-the informal presentation in our paper.  \scw{Give an overview of the figure}
+the informal presentation in our paper since it is more amenable to
+automation (Section~\ref{sec:automation}) and leaves no ambiguity
+about variable freeness conditions.
+
+ \scw{Give an overview of the figure}
 \scw{Define a convention that separates your use of natural numbers. Sometimes they
@@
 metavariables for each. (And it is possible to generalize universe levels to structures other than natural numbers. Ask Jonathan.}
-We find de Bruijn
-representation advantageous for specification since it leaves very
-little ambiguity about the variable freeness side
-conditions, making our proof more easily reproducible. Furthermore, as we discuss in
-Section~\ref{sec:automation}, de Bruijn representation is much more
-amenable to automated reasoning in proof assistants.
-We adopt from \citet{autosubst} the notations for simultaneous
-renaming, substitution, and other auxiliary definitions related to our
-syntax, summarized in Figure~\ref{fig:auxdef}.
+As a dependent type theory, terms and types are collapsed into the same
+syntactic category. The type $[[Set i]]$ represent universe
+types and $[[var n]]$ represent de Bruijn variables. While
+$[[i]],[[j]],$ and $[[n]]$ are all metavariables representing natural
+numbers, we always use $[[n]]$ to represent term variables that
+participate in substitution and $[[i]],[[j]]$ to represent universe levels.
+Dependent functions take the form $[[Pi A B]]$ and
+we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
+dependent on the input variable.
+Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
+proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
+$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
+is the term whose type is to be casted.
+\scw{Example of what you can use identity types for? Or an example of
+a program that uses J?}
+
+% We find de Bruijn
+% representation advantageous for specification since it leaves very
+% little ambiguity about the variable freeness side
+% conditions, making our proof more easily reproducible. Furthermore, as we discuss in
+% Section~\ref{sec:automation}, de Bruijn representation is much more
+% amenable to automated reasoning in proof assistants.
+
 \scw{Need to explain the notations in the text. What are these operations?
@@
 indices work.}
-We omit most of the
-definitions of renaming and substitution and only show the definition
-of a few representative cases of substitution.
+% We omit most of the
+% definitions of renaming and substitution and only show the definition
+% of a few representative cases of substitution.
 \begin{figure}[ht]
@@
       \begin{array}{lll}
-        \mathit{IdentityRen} \\
-        [[id i]] & := & [[i]] \\ \\
+        % \mathit{IdentityRen} \\
+        % [[id i]] & := & [[i]] \\ \\
         \mathit{IdentityTm} \\
-        [[idtm i]] & := & [[var i]] \\ \\
-        \mathit{ConsRen ([[xi .: n]])} \\
-        ([[xi .: n]])([[0]]) & := & [[n]] \\
-        ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
+        [[idtm n]] & := & [[var n]] \\ \\
+        % \mathit{ConsRen ([[xi .: n]])} \\
+        % ([[xi .: n]])([[0]]) & := & [[n]] \\
+        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
         \mathit{ConsSubst ([[rho .: a]])} \\
@@
         \mathit{Curried\ Add} \\
-        [[up n i]] & := & [[n]] + [[i]]
+        [[up n i]] & := & \ottkw{v}_{[[n]] + [[i]]}
       \end{array}
@@
       \begin{array}{lll}
-        \mathit{UpRen} \\
-        [[( up xi ) 0]] & := & [[0]] \\
-        [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
-        \mathit{Renaming ([[a < xi >]])} \\
-        [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
-        [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
-        \ldots \\ \\
-        \mathit{UpTm \scw{shifting? lifting?}} \\
+        % \mathit{UpRen} \\
+        % [[( up xi ) 0]] & := & [[0]] \\
+        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
+        % \mathit{Renaming ([[a < xi >]])} \\
+        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
+        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
+        % \ldots \\ \\
+        \mathit{Lifting} \\
         [[( up rho ) 0]] & := & [[var 0]] \\
@@
         [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
-        && \fbox{ \scw{case for $\lambda$ } } \\
+        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
         \ldots
@@
 % the typing context or substitution.
-As a
-dependent type theory, terms and types are collapsed into the same
-syntactic category. Dependent functions take the form $[[Pi A B]]$ and
-we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
-dependent on the input variable. $[[Set i]]$ represents the universe
-type where $[[i]]$ ranges over the set of natural numbers.
- Finally,
-\lang\scw{Don't begin a sentence with a symbol} also includes an intensional identity type $[[a ~ b : A]]$ whose
-proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
-$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
-is the term whose type is to be casted.
-\scw{Example of what you can use identity types for? Or an example of
-a program that uses J?}
-
-\lang is expressive enough to support large
+
+The \lang language is expressive enough to support large
 eliminations, the ability to compute a type using a term as input. For
@@
 is $[[true]]$ or $[[false]]$.
+
+We adapt from \citet{autosubst2} the notations for simultaneous
+renaming, substitution, and other auxiliary definitions used in
+our paper, summarized in Figure~\ref{fig:auxdef}. We need those
+definitions to specify the reduction relation and the typing relation
+for \lang{}. The metavariable $[[rho]]$ represents substitutions,
+mappings from de Bruijn indices to terms.
+One example of a substitution is
+the function $\ottkw{id}_{tm}$, which takes a de Bruijn index directly
+injects it as a term variable. The $\uparrow^{[[n]]}$ operator is
+more general and adds $[[n]]$ to the index before injecting it as a
+term. The $[[(rho .: a)]]$ operation allows us to extend a substitution
+$[[rho]]$ with an extra element $[[a]]$; the extended substitution maps the variable
+$[[var 0]]$ to $[[a]]$ and variables $[[var Suc i]]$ to $[[rho i]]$.
+
+The substitution operator, which takes the form $[[a {rho}]]$,
+traverses the syntax of $[[a]]$ and replaces each variable $[[var i]]$
+with the term $[[rho i]]$. When traversing under binders (e.g. in the
+$[[(\ a) { rho }]]$ case), the $\Uparrow$ operator prevents $[[var
+0]]$, the bound variable, from being replaced by
+$[[rho]]$. From the definition of $[[up rho]]$, we see that it keeps
+$[[var 0]]$ unchanged during substitution and replaces each $[[var
+Suc i]]$ with $[[rho i {up 1}]]$. The extra shifting with $[[up 1]]$
+is required so the variables in terms from the codomain of $[[rho]]$ skip over
+the newly introduced binder to refer to the correct binding
+location. The substitution operator is referred to as simultaneous
+substitution as it substitutes all variables at once. It is possible
+to recover single substitution by composing the extension operator and
+the identity substitution: $[[a { b }]] := [[a { idtm .: b }]]$.
 

### 2024-02-21 paper/paper.tex
Revise section 2

@@
         ([[rho .: a]])([[0]]) & := & [[a]] \\
-        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
+        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
         \mathit{Curried\ Add} \\
-        [[up n i]] & := & \ottkw{v}_{[[n]] + [[i]]}
+        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
       \end{array}
@@
         [[( up rho ) 0]] & := & [[var 0]] \\
-        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
+        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
         \mathit{Substitution ([[ a { rho }  ]])} \\
-        [[var i { rho }  ]] & := &  [[rho i]] \\
+        [[var n { rho }  ]] & := &  [[rho n]] \\
         [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
@@
   an explanation of definitional equality and its properties?}
-Figure~\ref{fig:par} shows the definition of the parallel reduction
-relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
-represent its transitive and reflexive closure, which in turn allows us to define
-the coherence relation $[[a <=> b]]$.\scw{where is the terminology ``coherence''
-from? I'm not familiar with it. Maybe you can replace it with ``definitional
-equivalence'' as that is how it is used in \lang?}
-We say that two terms $[[a]]$
-and $[[b]]$ are coherent if they can eventually reduce to some common
-term $[[c]]$ through parallel reduction. The symmetric notation of
-coherence suggests\scw{but you prove it below?} it is an equivalence relation.
-
-We sketch out some key properties about parallel reduction and
-coherence without giving their proofs. Our technique for establishing
-those results is based on \citet{takahashi-parallel-reduction} and a
-modern exposition of the same technique can be found in
-\citet{plfa22.08}.
+Before we specify the typing rules, we need to first specify its
+untyped equational theory. We choose to represent the untyped
+equational theory in an algorithmic style using the notion of parallel
+reduction, defined in Figure~\ref{fig:par}. Parallel reduction takes
+the form $[[a => b]]$ and we denote its transitive as $[[a =>+
+b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
+denoted by the notation $[[a0 <=> a1]]$, if there exists some term
+$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
+notation of convertibility suggests that it is an equivalence
+relation. We sketch out the sequence of lemmas required to derive this
+property below and omit the details. Our technique for proving the
+properties about parallel reduction is based on
+\citet{takahashi-parallel-reduction}. A modern exposition of the
+same technique can be found in \citet{plfa22.08}.
 \begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
@@
 \end{lemma}
-
 From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
 \ref{lemma:coherencetrans}, we conclude that coherence is indeed an
-equivalence relation. It is then trivial to derive that coherence is
-exactly the same as untyped $\beta$-equivalence.\scw{cite barendregt?}
+equivalence relation. It is then trivial to show that all untyped
+$\beta$-equivalence rules are admissible, though we omit this proof
+from our mechanization and will from now on focus on parallel
+reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
+makes it easier to work with. A detailed argument of the equivalence
+between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
+\citet{barendregt:lambda-calculi-with-types} and
+\citet{takahashi-parallel-reduction}.
+
+\begin{figure}[h]
+\begin{minipage}{0.9\textwidth}
+\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
+\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
+  Refl, Eq, J, If, Bool, True, False, Void}
+\end{minipage}
+\caption{Syntactic typing for \lang}
+\label{fig:typing}
+\end{figure}
+% Figure~\ref{fig:par} shows the definition of the parallel reduction
+% relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
+% represent its transitive and reflexive closure, which in turn allows us to define
+% the coherence relation $[[a <=> b]]$.\scw{where is the terminology ``coherence''
+% from? I'm not familiar with it. Maybe you can replace it with ``definitional
+% equivalence'' as that is how it is used in \lang?}
+% We say that two terms $[[a]]$
+% and $[[b]]$ are coherent if they can eventually reduce to some common
+% term $[[c]]$ through parallel reduction. The symmetric notation of
+% coherence suggests\scw{but you prove it below?} it is an equivalence relation.
+
+% We sketch out some key properties about parallel reduction and
+% coherence without giving their proofs. Our technique for establishing
+% those results is based on
+
+% \scw{cite barendregt?}
+% \yl{resolved}
 
@@
 
-\begin{figure}[ht]
-\begin{minipage}{0.9\textwidth}
-\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
-\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
-  Refl, Eq, J, If, Bool, True, False, Void}
-\end{minipage}
-\caption{Syntactic typing for \lang}
-\label{fig:typing}
-\end{figure}
-
 Figure~\ref{fig:typing} gives the full typing rules for
@@
 strengthen the inductive hypothesis of the fundamental theorem.
-\scw{Is the precondition for T-ABS really admissible? Isn't it required
-to ensure predicativity? Have you proved it? Have you proved regularity?}
-\yl{I proved regularity. \Rref{T-Abs} is admissible if cumulativity
-  was formulated correctly (which wasn't the case in the version
-  of the paper you read). I revised \rref{T-Univ} so the premise is
-  not admissible }
-\scw{Do you have a formation rule for identity types?}
-\yl{Added}
-
-In \rref{T-Var}, $[[G i]]$ is the partial function defined
-with the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
-i ]] := [[G i]] $.  \scw{explain the renaming in the conclusion of
-this lemma}
-The precondition $[[i < | G | ]]$, where $[[ | G |
+% \scw{Is the precondition for T-ABS really admissible? Isn't it required
+% to ensure predicativity? Have you proved it? Have you proved regularity?}
+% \yl{I proved regularity. \Rref{T-Abs} is admissible if cumulativity
+%   was formulated correctly (which wasn't the case in the version
+%   of the paper you read). I revised \rref{T-Univ} so the premise is
+%   not admissible }
+% \scw{Do you have a formation rule for identity types?}
+% \yl{Added}
+
+In \rref{T-Var}, $[[G n]]$ is the partial function defined
+by the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
+n ]] := [[G n]]$.
+The precondition $[[n < | G | ]]$, where $[[ | G |
 ]]$ represents the length of the context $[[G]]$, ensures that $[[G
-i]]$ has a defined output.
-
-\Rref{T-Conv} uses the definition of coherence from earlier
+n]]$ has a defined output. The type $[[G n]]$ needs to be shifted
+by $[[Suc n]]$ since the types stored in the context $[[G]]$ are
+scoped differently. In the typing judgment $[[G |- a : A]]$, we want
+both $[[a]]$ and $[[A]]$ to have the same scope. However, the closer a
+type is to the head of $[[G]]$, the smaller de Bruijn it has since
+there is less variables in the context it can refer to (see \rref{Ctx-Cons}). Shifting $[[G
+n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
+the full context $[[G]]$ rather than a truncated version of $[[G]]$.
+
+\Rref{T-Conv} uses the convertibility relation from earlier
 as our equality judgment for type conversion. The use of an untyped
-relation for type conversion makes our formulation slightly different
+relation for type conversion is reminiscent of Barendregt's Pure Type
+Systems~\citet{barendregt1991introduction} and makes our formulation different
 from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
-takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
-A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
+takes the form $\Gamma \vdash a \equiv b : A$, from which one can
+usually derive  $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving
+subject reduction.
+
 \citet{siles2012pure} shows the equivalence of Barendregt's
@@
 untyped equality, and its variant that uses typed judgmental
-equality. It is trivial to embed a system with judgmental
-equality to a system with untyped equality by erasing typing
-information. As a result, it is easy to port our consistency result from our
-type system to a variant with typed judgmental equality
-as long as the typed system does not include $\eta$ laws
-that would require type annotations (e.g. the $\eta$-law for unit
-types). % and we discuss in Section~\ref{sec:extension} how we can extend
+equality. This assures us that we do not lose much generality working
+with a system with untyped conversion. We include a detailed
+discussion in Section~\ref{sec:relatedwork} on how type-directed and
+untyped equality affect the applicability of our technique.
+
+% As a result, there is no loss of generality from
+
+% It is trivial to embed a system with judgmental
+% equality to a system with untyped equality by erasing typing
+% information. As a result, it is easy to port our consistency result from our
+% type system to a variant with typed judgmental equality
+% as long as the typed system does not include $\eta$ laws
+% that would require type annotations (e.g. the $\eta$-law for unit
+% types).
+% and we discuss in Section~\ref{sec:extension} how we can extend
 % our proof to handle those rules.
-\scw{Maybe move more of this paragraph and the next to the discussion/related
-  work section, and add a forward pointer here.}
-\scw{Just say that like Coq, the definition of the system uses untyped
-equality}
-
-Working with a system with untyped equality has the huge benefit that
-the confluence result for untyped parallel reduction
-(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
-resort to the complex syntatic (resp. semantic) technique from
-\citet{siles2012pure} (resp. \citet{decagda}) to resolve
-the circularity of subject reduction and $\Pi$-injectivity.
-Section~\ref{sec:extension} explains how we
-generalize our technique to include $\eta$-law for functions and
-show the existence of normal form for well-typed (open and closed)
-terms, achieving a similar level of expressiveness of the type system
-and strength of metatheoretic property as \citet{decagda}.
-
-Finally, since our system has an infinite universe hierarchy, we can
-present the system à la Russell by using the same judgment form $[[G
-|- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
-is no need to distinguish between big types and small types
-and duplicate our typing specification.
-\scw{Is there an advantage for Tarski universes even with infinite
-hierarchy? or no?}
+\yl{Unresolved: is it ok to say our system's conversion rule is untyped like
+Coq?}
+% \scw{Maybe move more of this paragraph and the next to the discussion/related
+%   work section, and add a forward pointer here.}
+% \scw{Just say that like Coq, the definition of the system uses untyped
+% equality}
+
+% Working with a system with untyped equality has the huge benefit that
+% the confluence result for untyped parallel reduction
+% (Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
+% resort to the complex syntatic (resp. semantic) technique from
+% \citet{siles2012pure} (resp. \citet{decagda}) to resolve
+% the circularity of subject reduction and $\Pi$-injectivity.
+% Section~\ref{sec:extension} explains how we
+% generalize our technique to include $\eta$-law for functions and
+% show the existence of normal form for well-typed (open and closed)
+% terms, achieving a similar level of expressiveness of the type system
+% and strength of metatheoretic property as \citet{decagda}.
+
+% Finally, since our system has an infinite universe hierarchy, we can
+% present the system à la Russell by using the same judgment form
+%$[[G |- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
+% is no need to distinguish between big types and small types
+% and duplicate our typing specification.
+% \scw{Is there an advantage for Tarski universes even with infinite
+% hierarchy? or no?}
 
@@
 \section{Logical Relation}
-\scw{This section pre-supposes that we want to define a logical relation,
-but doesn't precisely state what we want to use it to prove, and why a logical
-relation is suitable. (And why the property you want to prove is difficult to
-show!) Should add more motivation here.}
+% \scw{This section pre-supposes that we want to define a logical relation,
+% but doesn't precisely state what we want to use it to prove, and why a logical
+% relation is suitable. (And why the property you want to prove is difficult to
+% show!) Should add more motivation here.}
+Before we define our logical relation, we first formally specify the
+consistency property that we want to prove.
+\begin{theorem}[Logical Consistency]
+  \label{theorem:consistency}
+  The judgment $[[empty |- a : Void ]]$ is not derivable.
+\end{theorem}
+The property can be formulated in a simply typed language, where
+$[[Void]]$ is similarly defined as a type that has no term. A related
+property, referred to as the termination property (for closed terms),
+is commonly used in introductory materials such as
+\citet{skorstengaard2019introduction}, \citet{pierce2002types}, and
+\citet{harpertait} to motivate the need for a logical relation.
+
+A naive attempt to proving Theorem~\ref{theorem:consistency} by
+induction on the derivation $[[empty |- a : Void]]$ would succeed at
+almost all cases except for \rref{T-App}. In the application
+case, we are given $[[empty |- b : Pi A B]]$ and $[[empty |- a : A]]$, and
+the equality that $[[B {a} = Bool]]$. Our goal is to show that
+$[[empty |- b a : Void]]$ is not possible. However, note that there is
+nothing we know of $[[b]]$ or $[[a]]$ from the induction hypothesis
+because neither $[[Pi A B]]$ nor $[[A]]$ is equal to $[[Bool]]$.
+We have no way of deriving a contradiction from $[[empty |- b a :
+Void]]$. The takeaway from this failed attempt is that, in order to
+derive the consistency, we need to know something about types other
+than $[[Void]]$. From a pragmatic point of view, proof by logical
+relation can be seen as a sophisticated way of strengthening the
+induction hypothesis. From the strengthened property, the fundamental
+theorem, we will be able to derive consistency as a corollary.
 
@@
 \end{figure}
-In this section, we define the logical relation for \lang{} in the
-form of an inductively defined relation (Figure~\ref{fig:logrel}).
-Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
-The logical relation takes the form $[[Interp I i A S]]$. The
-metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
+The logical relation for \lang{}, which takes the form $[[Interp I i A
+S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
+% The logical relation takes the form $[[Interp I i A S]]$.
+Themetavariables $[[A]]$ and $[[i]]$ stand for terms and natural
 numbers respectively, as introduced earlier in
 Figure~\ref{fig:syntax}.
-\scw{Because the type of $I$ depends on $i$, I'd prefer a syntax that
-keeps them in order. }
+Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
 \scw{Many introductory texts define the relation as a recursive function
@@
 
-\begin{figure}[h]
-\begin{equation*}
-    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
-\end{equation*}
-\caption{Logical relation for all universe levels}
-\label{fig:logrelrec}
-\scw{I didn't understand this figure without reading the text below.}
-\end{figure}
 To tie the knot and obtain an interpretation of all universe levels,
@@
 $[[S]]$.
+
+\begin{figure}[h]
+\begin{equation*}
+    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
+\end{equation*}
+\caption{Logical relation for all universe levels}
+\label{fig:logrelrec}
+\scw{I didn't understand this figure without reading the text below.}
+\end{figure}
 
@@
 
-\begin{corollary}[Logical Consistency]
-  \label{corollary:consistency}
-  The judgment $[[empty |- a : Void ]]$ is not derivable.
-\end{corollary}
 \begin{proof}

### 2024-02-21 paper/paper.tex
Revise section 3 (WIP)

@@
 % The logical relation takes the form $[[Interp I i A S]]$.
-Themetavariables $[[A]]$ and $[[i]]$ stand for terms and natural
+Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
 numbers respectively, as introduced earlier in
 Figure~\ref{fig:syntax}.
-Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
 \scw{Many introductory texts define the relation as a recursive function
@@
 \end{equation*}
+The notation $[[PowerSet STm]]$ denotes the powerset of the set of
+\lang{} terms.
 The function $[[I]]$ is a family of sets of terms indexed by
@@
 restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
-be predicative. Removing the ordering constraint would result in a
-system where one can encode Girard's paradox~\citep{girard-thesis}.
-\scw{Spell this out a bit more about what goes wrong. Where does the
-proof break? Also spell out how this enforces predicativity.}
+be predicative and we will explain the reason shortly.
 
@@
 we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
-using the well-foundedness of the strict less than relation on natural
+using the well-foundedness of $<$ relation on natural
 numbers (recall that
@@
 \label{fig:logrelrec}
-\scw{I didn't understand this figure without reading the text below.}
 \end{figure}
-
-With the definition from Figure~\ref{fig:logrelrec}, we can
+The definition of in Figure~\ref{fig:logrelrec} explains how the $[[j
+< i]]$ constraint in \rref{I-Set} makes our system predicative; the
+interpretation of the $[[i]]_{th}$ universe is only dependent on
+universes at strictly lower than $[[i]]$, which have been defined earlier.
+Removing the ordering constraint would result in a
+system where one can encode Girard's
+paradox~\citep{girard-thesis}. Our proof would no longer work with the
+restriction removed since the definition of $[[InterpR i A S]]$ would
+not be well-founded when $[[Interp I i A S]]$ calls $[[I]]$ on
+universe levels greater than or equal to $[[i]]$, which are yet to be defined.
+
+By unfolding the definition in Figure~\ref{fig:logrelrec}, we
 show that the same introduction rules for $[[Interp I i A S]]$ are
@@
 
-In the rest of the section, we establish some important facts about
-the logical relation that will be useful for proving fundamental theorem
-in Section~\ref{sec:logrelproof}.
-
-The relation $[[Interp I i A S]]$ satisfies the following inversion
-principles.\scw{Put footnotes on lemma name, or on (1), (2), etc}
+In the rest of the section, we prove four important facts about
+our logical relation: irrelevance (Lemma~\ref{lemma:logrelcoherence}),
+functionality (Lemma~\ref{lemma:logreldeter}), cumulativity
+(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure
+property (Lemma~\ref{lemma:logrelNbackclos}).
+
+First, we prove a family of simple properties, which we refer to as
+inversion principles for our logical relation. Given $[[Interp I i A
+S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi A0
+B0]]$, the inversion lemma allows us to say something about the set
+$[[S]]$. Its proof is simple, but we sketch out the case for
+functions to help readers confirm their understanding of \rref{I-Pi}.
 \begin{lemma}[Inversion of the logical relation]
@@
   \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
-  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}}
-  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
     \begin{itemize}
@@
     \end{itemize}
-  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}}
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
   \end{enumerate}
@@
 \begin{proof}
-  \scw{Why is this proof interesting and worth typesetting? Should
-    it be in an appendix so that it doesn't interrupt the flow of
-    reading?}
-  We only show the most involved function case; the rest follows a
-  similar but simpler pattern. We start by
-  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
+  As mentioned earlier, we only show the inversion property for the
+  function type.
+  We start by inducting over the derivation of $[[Interp I i Pi A B S]]$. There
   are only two possible cases we need to consider.
@@
     \end{itemize}
-    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
+    By inverting the derivation of $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
     $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
@@
     \end{itemize}
-    These properties are exactly what's\scw{avoid contractions} required in the conclusion.
+    These properties are exactly the ones from the conclusion.
   \end{description}
@@
 \Rref{I-Red} bakes into the logical relation the backward preservation
-property. The following property shows that preservation holds in the
-usual forward direction too.
+property. That is, given $[[Interp I i A S]]$, if $[[B =>+ A]]$, then
+$[[Interp I i B S]]$ also holds. The following property shows that
+preservation holds in the usual forward direction too.
 \begin{lemma}[Preservation of the logical relation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
@@
 
+\appendix
+
+\subsection{More detailed proofs}
+\begin{lemma}[Preservation of the logical relation]
+  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
+\end{lemma}
+\begin{proof}
+  We carry out the proof by induction over the derivation of $[[Interp
+  I i A S]]$.
+  \begin{description}
+  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
+    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
+    be $[[Void]]$ and the result trivially follows.
+  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
+  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
+      refl , a <=> b  }]]$ and, by inverting the derivation of
+    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
+    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
+    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
+    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
+      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
+    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
+      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
+    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
+    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
+    <=> b0]]$. The result then immediately follows from the fact that
+    coherence is an equivalence relation.
+  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
+    such that:
+    \begin{itemize}
+    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
+        a #) }]]$
+    \item $[[Interp I i A S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+    \end{itemize}
+    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
+    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
+      a #) }]]$. By \rref{I-Pi}, it suffices to show:
+    \begin{itemize}
+    \item $[[Interp I i A0 S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
+    \end{itemize}
+    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
+    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
+    conditions follow immediately from the induction hypothesis.
+  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
+    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
+    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
+    property of parallel
+    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
+    $[[C]]$ such that $[[B0 => C]]$ and $[[B => C]]$. By the induction
+    hypothesis, we derive $[[Interp I i C S]]$ from $[[Interp I i
+    B0 S]]$. By \rref{I-Red} and
+    $[[B => C]]$, we conclude that $[[Interp I i B S]]$.
+  \end{description}
+\end{proof}
+
 %%

### 2024-02-21 paper/paper.tex
Revise section 3

@@
 
+The complexity of applying proof by logical relation to dependent types stem
+from the fact that the logical relation is much harder to define. In
+simply typed languages, the logical relation is defined as a recursive
+function over the type $[[A]]$. In dependent types, the type
+$[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
+this type, we need to first reduce it to $[[Bool]]$. However, we
+cannot write a function that performs the reduction because we do not
+know the termination of well-typed terms a priori. As a result, we
+define the logical relation as an inductively defined relation,
+reminiscent of how we specify the reduction graph of a partial
+function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.
+
 \label{sec:logreldep}
@@
 looking. Need to explain.  }
-
 The metavariables $[[I]]$ and $[[S]]$ are
@@
 b]]$ are typically required for indexed types, of which equality types
-are an instance.
-
-\Rref{I-Pi} is the most interesting. The precondition consists of a
-mysterious function $[[F]]$ that takes an element from the set $[[S]]$, the interpretation
-of the type $[[A]]$, and returns a subset of terms. $[[F]]$ is further
-subject to the constraint that $[[forall a, (# a in S implies Interp I
-i B { a } F a #)]]$. Ignoring the content below the horizontal bar of
-the derivation, the statements from the precondition are equivalent to
-the conjunction of the following statements:
+are an instance. \Rref{I-Par} enables us to reduce our types in order
+to assign meanings. Recall the type expression $[[(\ var 0)
+Bool]]$. \Rref{I-Par} says to know that
+$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
+show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
+Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
+  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Par}
+and \rref{I-Bool}.
+
+\Rref{I-Pi} is the most interesting. To explain what it means, we
+need to perform a few steps of transformations to its precondition.
+The precondition consists of a
+mysterious relation $[[R]]$ over the set $[[S]]$, an interpretation
+of the type $[[A]]$, and returns a subset of terms. $[[R]]$ is further
+subject to the following two constraints:
+\begin{itemize}
+\item $[[forall a, (# exists S0  , ( a , S0 ) in R #)]]$
+\item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
+\end{itemize}
+The first condition allows us to induce a function $[[F in S ->
+PowerSet STm]]$ such that $[[(a , F a) in R]]$. Then we can easily
+show that $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$.
+
+For the other direction, given a function $[[F]]$ satisfying the
+property $[[forall a, (# a in S implies Interp I i B { a } F a #) ]]$, it
+is possible to go in the other direction and define a relation $[[R]]=[[
+{ ( a , F a )  | a in S }]]$. It is easy to verify that $[[R]]$ satisfy
+the the constraint required in the precondition of \rref{I-Pi}.
+
+Furthermore, the skolemization process allows us to conclude the
+following equivalence:
+\[ \exists F, [[forall a, (# a in S implies Interp I i B { a } F a #)]] \iff [[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]  \]
+% Ignoring the content below the horizontal bar of
+% the derivation, we claim the statements from the precondition are equivalent to
+% the conjunction of the following statements:
+% \begin{itemize}
+% \item $[[Interp I i A S]]$
+% \item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
+%   [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
+% \end{itemize}
+% In general, let $R$ be a binary relation over the sets $[[A]]$ and
+% $[[B]]$, it easy to verify the following equivalence, which is used to
+% justify the skolemization process~\citep{skolemization}.
+% \scw{Is this true constructively? If not, is that ok?}
+% \[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
+%   \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
+%   \in R\]
+% The left-hand side is equivalent to the right-hand side, but has no
+% mentioning of any functions.
+The precondition of \rref{I-Pi} therefore can be formulated as the
+following statements:
 \begin{itemize}
 \item $[[Interp I i A S]]$
-\item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
-  [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
+\item $[[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]$
 \end{itemize}
-In general, let $R$ be a binary relation over the sets $[[A]]$ and
-$[[B]]$, it easy to verify the following equivalence, which is used to
-justify the skolemization process~\citep{skolemization}.
-\scw{Is this true constructively? If not, is that ok?}
-\[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
-  \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
-  \in R\]
-The left-hand side is equivalent to the right-hand side, but has no
-mentioning of any functions.
-In the context of \rref{I-Pi}, through
-the equivalence, we can more easily tell what's happening: the
-function type $[[Pi A B]]$ has a semantic interpretation if its input
-type $[[A]]$ can be semantically
+From this alternative formulation, we can more easily tell what is
+happening in \rref{I-Pi}: the
+function type $[[Pi A B]]$ has an interpretation if its input
+type $[[A]]$ can be
 interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
 type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
-$[[B]]$, has a semantic interpretation.
-
-If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
-  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
-presentation with an explicit $[[F in S -> PowerSet STm]]$ is
+$[[B]]$, has some semantic interpretation.
+
+If we look at $[[Interp I i Pi A B { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$, the conclusion of \rref{I-Pi}, we see why the
+presentation with an explicit $[[R in S * PowerSet STm]]$ is
 useful. We want to state the fact that $[[B {a}]]$ has a semantic
@@
 set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
-to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
-witness in the precondition for $[[a in S]]$.
-With the alternative representation that is free of the function $[[F]]$, we
-might want to reformulate our logical relation in the following way.
+to. With the relation $[[R]]$ in scope, we can refer to the
+witness sets $[[S0]]$ such that $[[Interp I i B {a} S0]]$ for $[[a in S]]$.
+With the alternative representation that is free of the relation $[[R]]$, we
+might want to reformulate our logical relation in the following more
+concise but equivalent form, whose admissibility follows directly from
+\rref{I-Pi}.
 \begin{center}
@@
 \end{center}
+\begin{lemma}
+  \label{lemma:piintroalt}
+  \Rref{I-PiAlt} is admissible.
+\end{lemma}
+\begin{proof}
+Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
+, S0 ) | a in S , Interp I i B { a } S0 }]]$.
+\end{proof}
+
 Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
@@
 functionality and inversion properties, we opt for our current
-skolemized formulation of \rref{I-Pi} so we immediately obtain a
+formulation of \rref{I-Pi} so we immediately obtain a
 well-defined inductive relation and a usable induction principle. The
 slight disadvantage of \rref{I-Pi} is that we need to construct the
-function $[[F]]$ each time we apply it. Therefore, in
-Lemma~\ref{lemma:piintroalt}, we show the admissibility of
-\rref{I-PiAlt} a posteriori and use that as a more convenient
-introduction rule for the rest of our proofs.
+function $[[R]]$ each time we apply it, though this is mitigated by
+the immediate admissibility of \rref{I-PiAlt}.
 
@@
   \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
-  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S1]]$, then there exists $[[S]],[[R]]$ such that:
     \begin{itemize}
     \item $[[Interp I i A S ]]$
-    \item $[[F in S -> PowerSet STm]]$
-    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
-    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
+    \item $[[R in S * PowerSet STm]]$
+    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
+    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
+    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
     \end{itemize}
@@
   \item[\Rref{I-Pi}:] Immediate.
-  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
+  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S1]]$.
     We know that there exists some $[[A0]]$ and
     $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
-    A0 B0 S0]]$. From the
-    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
+    A0 B0 S1]]$. From the
+    induction hypothesis, there exists $[[S]]$ and $[[R]]$ such that :
     \begin{itemize}
     \item $[[Interp I i A0 S ]]$
-    \item $[[F in S -> PowerSet STm]]$
-    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
-    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
+    \item $[[R in S * PowerSet STm]]$
+    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
+    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B0 { a } S0 #)]]$
+    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
     \end{itemize}
@@
     $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
-    $[[F]]$ satisfies the following properties:
+    $[[R]]$ additionally satisfies the following properties:
     \begin{itemize}
     \item $[[Interp I i A S ]]$
-    \item $[[F in S -> PowerSet STm]]$
-    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
     \end{itemize}
-    These properties are exactly the ones from the conclusion.
+    These properties are exactly what we need to finish the proof.
   \end{description}
@@
   \label{lemma:interppreservation}
+  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
+\end{lemma}
+\begin{proof}
+  We carry out the proof by induction over the derivation of $[[Interp
+  I i A S]]$.
+
+  The only interesting case is \Rref{I-Red}. Given that
+  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
+  for all  $[[B]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
+  S]]$. By the diamond property of parallel reduction
+  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
+  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
+  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
+  B0 S]]$. By \rref{I-Par} and $[[B1 => B]]$, we conclude that
+  $[[Interp I i B S]]$.
+
+  The remaining cases all fall from induction hypotheses and basic
+  properties about convertibility and parallel reduction we have
+  established in Section~\ref{sec:spec}.
+\end{proof}
+From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
+derive the following corollary that two convertible types can always interpret
+into the same set. We adopt the terminology from \citet{martin-lof-a-la-coq}
+and refer to this property as irrelevance.
+\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
+  \label{lemma:logrelcoherence}
+  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
+\end{corollary}
+% \scw{Need a transition here}
+
+Since the definition of our logical relation is an inductive relation,
+it is not immediately obvious why each type $[[A]]$ can only uniquely
+corresponds to one set $[[S]]$. The following lemma shows that our
+logical relation is indeed functional.
+\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
+  \label{lemma:logreldeter}
+  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
+\end{lemma}
+\begin{proof}
+  The proof proceeds by induction over the derivation of the first
+  premise $[[Interp I i A S0]]$.
+  All cases that are not \rref{I-Red} follow immediately from the
+  Lemma~\ref{lemma:interpinv}.
+
+  For \rref{I-Red}, we are given that there exists some $[[B]]$ such
+  that $[[A => B]]$ and $[[Interp I i B S0]]$. Our goal is to show
+  that given $[[Interp I i A S1]]$ for some $[[S1]]$, we have $[[S0 =
+  S1]]$. By the preservation property
+  (Lemma~\ref{lemma:interppreservation}),
+  we know that $[[Interp I i B S1]]$ since $[[A => B]]$. The statement
+  $[[S0 = S1]]$ then immediately follows from the induction hypothesis.
+\end{proof}
+
+Lemma~\ref{lemma:logreldeter} enables us to show the following
+improved inversion lemma for function types whose statement is free of
+the relation $[[R]]$, analogous to the admissible rule \rref{I-PiAlt}.
+\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
+  \label{lemma:piinvalt}
+  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
+  such that the following constraints hold:
+  \begin{itemize}
+  \item $[[Interp I i A S0]]$
+  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
+    S1 #) #)]]$
+  \item $[[S = { b | forall a, (# a in S0 , forall
+    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
+\end{proof}
+
+The next lemma shows that our logical relation satisfies
+cumulativity. That is, if a type has an interpretation at a lower
+universe level, then we can obtain the same interpretation at a higher
+universe level.
+\begin{lemma}[Logical relation cumulativity\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
+  \label{lemma:logrelcumulativity}
+  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
+\end{lemma}
+\begin{proof}
+  Trivial by structural induction over the derivation of $[[Interp I
+  i0 A S]]$.
+\end{proof}
+Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
+implicitly assume that $[[I]]$ is defined on the set of natural
+numbers less than $[[i1]]$.
+
+\begin{corollary}[Logical relation is functional with different levels\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
+  \label{lemma:logreldeterhet}
+  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
+\end{corollary}
+\begin{proof}
+  Immediate from Lemma~\ref{lemma:logreldeter} and
+  \ref{lemma:logrelcumulativity}.
+\end{proof}
+
+We say that a set of terms $[[S]]$ is closed under expansion if given
+$[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
+The final property we want to show is that the output set $[[S]]$ from
+the logical relation is closed under expansion. Unlike the previous
+lemmas, we directly state the lemma
+in terms of $[[InterpR i A S]]$ rather than $[[Interp I i A S]]$
+because we need to know something about $[[I]]$ for this property to
+hold in the \rref{I-Set} case.
+
+\yl{Can lift this lemma somewhere earlier in the section since it doesn't really
+  depend on anything}
+\begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpExt\_back\_clos}}]
+  \label{lemma:logrelbackclos}
+  If $[[InterpR i A S]]$, then the set $[[S]]$ is closed under expansion.
+\end{lemma}
+\begin{proof}
+  By the definition of $[[InterpR i A S]]$ from
+  Figure~\ref{fig:logrelrec}, we unfold $[[InterpR i A S]]$ by one
+  step into $[[ Interp I i A S  ]]$ where $[[I i]] := [[{A | exists S
+    , InterpR i A S}]]$.
+  We then proceed by induction over the derivation of $[[Interp I i A S]]$.
+
+  All cases are trivial except for the \rref{I-Set} case, where we
+  want to show that the set $[[I j]]$ is closed under expansion for
+  all $[[j < i]]$. However, by the definition of $[[I]]$, we know that
+  $[[A in I j]]$ if only if there exists some $[[S]]$ such that
+  $[[Interp I i A S]]$. By \rref{I-Par}, we must also have $[[B in I
+  j]]$ for all $[[B => A]]$.
+\end{proof}
+
+\section{Semantic Typing and Consistency}
+\label{sec:logrelproof}
+\begin{figure}[h]
+\[
+\begin{array}{lcl}
+
+      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
+                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
+      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
+                       G]]\text{ then there exists some } [[j]] \text{
+                       and } [[S]] \text{ such that } [[InterpR j A
+                       {rho} S]] \text{ and } [[a {rho} in S]] \\
+      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
+\end{array}
+\]
+  \caption{Semantic Typing for \lang}
+  \label{fig:semtyping}
+\end{figure}
+
+The logical relation we define in Figure~\ref{fig:logrel} does not
+include cases for variables. Likewise, for the base types such as
+boolean and equality, the output set $[[S]]$ contains only terms that
+evaluate to closed terms.
+
+\scw{What is the motivation for this section? Why do we want to generalize
+the logical relation?}
+To generalize our logical relation to open
+terms, we define the semantic typing judgment by closing the open
+terms with a substitution whose codomain consists of terms that
+respect the interpretation of the types from the context. The full
+definitions of well-formed substitution ($[[rho |= G]]$), semantic
+typing ($[[ G |= a : A]]$), and semantic context well-formedness
+($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
+\scw{explain all parts of this figure. it's pretty intimidating.}
+In the
+definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
+the typing context obtained by dropping the last $[[i]]$ elements of
+$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
+returns the empty list.
+
+The following lemma makes the statement $[[G |= A : Set i]]$ easier to
+work with.
+\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}\scw{spell out the whole word}]
+  \label{lemma:setinv}
+  The following two statements are equivalent:
+  \begin{itemize}
+  \item $[[G |= A : Set i]]$
+  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
+    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  The forward direction is immediate by
+  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
+  and show that $[[G |= A : Set i]]$ given the second bullet.
+
+  Suppose $[[rho |= G]]$, then we know that there exists some $[[S]]$
+  such that $[[InterpR i (A {rho}) S]]$. By the definition of semantic
+  typing, it suffices to show that there exists some $[[j]]$ and
+  $[[S0]]$ such that  $[[InterpR j Set i S0]]$ and $[[A {rho} in
+  S0]]$.
+  Pick $[[Suc i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
+  ]]$ for $[[S0]]$ and it is trivial to verify the conditions hold.
+\end{proof}
+
+The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
+counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
+than inductively over the context. \scw{why is that the case?} It is easy to recover the same
+structural rules:
+\begin{lemma}[Semantic context well-formedness cons]
+  \label{lemma:semwffcons}
+  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
+\end{lemma}
+\scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
+Ctx-empty?}
+\begin{proof}
+  By the definition of semantic context well-formedness, the goal is
+  to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i
+  (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
+  proven by case analysis on whether $[[i]]$ is zero.
+\end{proof}
+
+Likewise, the semantic well-formendness judgment for substitutions
+satisfies similar structural rules.
+\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
+  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
+  $[[rho .: a |= G ++ A]]$.
+\end{lemma}
+\begin{proof}
+  Start by unfolding the definition of $[[rho .: a |= G ++ A]]$ and
+  performing a case analysis similar to the proof of
+  Lemma~\ref{lemma:semwffcons}.  The case where the number is $[[0]]$
+  requires Lemma~\ref{lemma:logreldeterhet} to finish the proof.
+\end{proof}
+
+Next, we show some non-trivial cases of the fundamental theorem as
+top-level lemmas and leave the remaining cases as exercises for the
+reader.\scw{I dislike leaving ``exercises to the reader''}
+
+First, we formulate the definition of valid renamings and prove that
+semantic typing satisfies renaming so we can weaken the context when
+reasoning about the variable case of the fundamental lemma
+(Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G
+++ D]]$, it is easy to show that we can extract some valuation
+$[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
+``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :
+A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
+$[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
+after weakening the context; this implication holds because $[[rho |=
+G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
+can make use of the premise $[[G |= a : A]]$ to derive what we need
+for the conclusion. We recommend the readers to skip the proofs of
+Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
+during the first read as long as they have an intuitive understanding
+of what the renaming property is meant to capture.
+
+\scw{Make this a definition? Would it be useful to create notation
+  for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
+We say that $[[xi]]$ is
+valid from the context $[[G]]$ to the context $[[D]]$ if the
+following condition holds.
+\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
+  [[D ren xi
+  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]
+
+\begin{lemma}[Truncate is valid]
+  \label{lemma:validtruncate}
+  If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i
+  G]]$ to $[[G]]$.
+\end{lemma}
+\begin{proof}
+  By the definition of a valid renaming, we must show that given $[[j
+  < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
+  \begin{itemize}
+  \item $[[ren up i j]] = i + j  < [[| G | ]]$
+  \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
+  \end{itemize}
+  The first bullet point is immediate from the fact that $[[j]] <
+  [[|G|]] - [[i]]$. The second bullet follows from unfolding the
+  definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
+  $[[i]] + [[j]] < [[|G|]]$.
+\end{proof}
+
+\begin{lemma}[Renaming for $[[rho |= G]]$]
+  \label{lemma:renamingval}
+  If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
+  $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
+\scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
+\end{lemma}
+\begin{proof}
+  Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
+  that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i
+  < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.
+
+  By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
+have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho
+} ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |=
+D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
+S]]$ and $[[rho ren xi i in S]]$.
+\end{proof}
+
+\begin{lemma}[Renaming for $[[G |= a : A]]$]
+  \label{lemma:semrenaming}
+  If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
+  $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
+\end{lemma}
+\begin{proof}
+  Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
+\end{proof}
+
+\scw{Need a transition here that you are starting to explain the semantic typing
+rules.}
+
+\begin{lemma}[ST-Var]
+  \label{lemma:stvar}
+  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
+\end{lemma}
+\begin{proof}
+  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
+need to show that there exists some $[[j]]$ and $[[S]]$ such that
+  \begin{itemize}
+  \item $[[InterpR j G i < up Suc i > { rho } S]]$
+  \item $[[rho i in S]]$
+  \end{itemize}
+  From the definition of $[[|= G]]$, we know that there exists some
+  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. By
+  Lemma~\ref{lemma:validtruncate}, we know that $[[up Suc i]]$ is a
+  valid renaming from $[[drop Suc i G]]$ to $[[G]]$. By
+  Lemma~\ref{lemma:semrenaming}, we deduce $[[G |= G i < up Suc
+  i > : Set j]]$. By the identity from Lemma~\ref{lemma:setinv}, we
+  know that $[[InterpR j G i < up Suc i > { rho } S]]$. It now
+  suffices to show $[[rho i in S]]$, but that is immediate from the
+  definition of $[[rho |= G]]$.
+\end{proof}
+
+\begin{lemma}[ST-Set]
+  \label{lemma:stset}
+  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
+\end{lemma}
+\begin{proof}
+  Immediate by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
+\end{proof}
+
+\begin{lemma}[ST-Pi]
+  \label{lemma:stpi}
+  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
+  A B : Set i]]$.
+\end{lemma}
+\begin{proof}
+  Applying Lemma~\ref{lemma:setinv} to the
+  conclusion, it now suffices to show that given $[[rho |= G]]$, there
+  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
+  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
+  there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
+From $[[G ++ A |= B : Set i]]$, we know that there must
+exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
+in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
+\end{proof}
+
+\begin{lemma}[ST-Abs]
+  \label{lemma:stabs}
+  If $[[G |= Pi A B : Set i]]$ and $[[G ++ A |= b : B]]$, then $[[G |=
+  \ A b : Pi A B]]$.
+\end{lemma}
+\begin{proof}
+  By unfolding the definition of $[[G |= \ A b : Pi A B]]$, we need to
+  show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
+  $[[S]]$ such that $[[InterpR i Pi A {rho} B {up rho} S]]$ and $[[\
+  A{rho} b{up rho} in S]]$.
+
+  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
+  i]]$, there exists some set $[[S]]$ such that
+  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
+  $[[\A{rho} b{up rho} in S
+  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
+  that all following conditions hold:
+  \begin{itemize}
+  \item $[[InterpR i A{rho} S0]]$
+  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
+    {rho .: a}
+    S1 #) #)]]$
+  \item $[[S = { b | forall a, (# a in S0 , forall
+      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
+  \end{itemize}
+  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
+  that given $[[a in S0]]$,
+  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
+  a in S1]]$.
+  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
+  under expansion. By Lemma~\ref{lemma:parabscbn}, since $[[( \A{rho} b{up rho} )
+  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
+  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
+  B]]$ and the fact that the logical relation is deterministic and
+  cumulative (Lemma~\ref{lemma:logreldeterhet}).
+\end{proof}
+
+\begin{lemma}[ST-App]
+  \label{lemma:stapp}
+  If $[[G |= b : Pi A B]]$ and $[[G |= a : A]]$, then $[[G |= b a : B {a}]]$.
+\end{lemma}
+\begin{proof}
+Suppose $[[rho |= G]]$. The goal is to show that there exists some
+$[[i]]$ and $[[S1]]$
+such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a} {rho}
+S1]]$, or equivalently, $[[InterpR i B {rho .: a {rho}} S1]]$ since
+$[[B {a}{rho}]] = [[B {rho .: a {rho}}]]$. By the premise $[[G |= b :
+Pi A B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
+there exists some $[[i]]$ and $[[S0]]$ such that:
+  \begin{itemize}
+  \item $[[InterpR i A{rho} S0]]$
+  \item $[[forall a0, (# a0 in S0 implies (# exists S1 , InterpR i B
+    {rho .: a0}
+    S1 #) #)]]$
+  \item $[[forall a0, (# a0 in S0 , forall
+      S1, (# InterpR i B {rho .: a0} S1,  b {rho} a0 in S1 #) #)]]$
+  \end{itemize}
+  Instantiating the variable $[[a0]]$ from the last two bullets with
+  the term $[[a {rho}]]$, the conclusion immediately follows.
+\end{proof}
+
+\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
+  \label{theorem:soundness}\leavevmode
+  \begin{itemize}
+  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
+  \item If $[[|- G]]$, then $[[|= G]]$.
+  \end{itemize}
+\end{theorem}
+\begin{proof}
+  Proof by mutual induction over the derivation of $[[G |- a :
+  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
+  from Lemma~\ref{lemma:semwffcons}.
+Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
+  can be used to discharge their syntactic counterpart
+  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
+  cases not covered by the lemmas are similar to the ones already
+  shown or simpler and therefore omitted from the text.
+
+\end{proof}
+
+\scw{Spell this out in a little more detail.}
+
+\begin{proof}
+  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
+\end{proof}
+
+\section{Toward Decidability of Type Conversion and $\eta$ laws}
+\label{sec:extension}
+\begin{figure}[h]
+  \[
+    \begin{array}{lcl}
+      % \beta\text{-}\mathit{neutral\ terms}\\
+      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
+                    f]] \\ \\
+      % \beta\text{-}\mathit{normal\ terms} \\
+      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
+                    |\ [[f ~ f : f]]\\
+            & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
+    \end{array}
+  \]
+  \caption{$\beta$-neutral and normal forms}
+  \label{fig:nenf}
+\end{figure}
+\begin{figure}[h]
+  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
+  \caption{Extended logical relation}
+  \label{fig:logrelopen}
+\end{figure}
+
+\scw{Put this introduction before the figure in the section.}
+In this section, we sketch out how the logical relation from
+Section~\ref{sec:logreldep} can be extended show the existence of
+$\beta\eta$ normal forms for (open and closed) well-typed terms. We
+first extend parallel reduction to include $\eta$ reduction for
+functions.
+\begin{center}
+  \drule[width=2.5in]{P-AbsEta}
+\end{center}
+\scw{What happened to the domain type in the $\lambda$ expression? This is
+not your syntax...}
+\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
+equational theory since coherence, the untyped relation used for type
+conversion, is built on top of parallel reduction. We can recover the
+same confluence result about parallel reduction using the standard
+techniques from \citet{barendregt:lambda-calculi-with-types,
+takahashi-parallel-reduction}, though specifically for the de Bruijn
+representation, we need the following anti-renaming lemma about
+parallel reduction.
+\begin{lemma}[Par anti-renaming]
+  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
+exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
+\end{lemma}
+
+The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the neutral terms
+and normal forms with respect to $\beta$-reduction, but can still take
+$\eta$ reduction steps. We sometimes use the judgment forms $[[ne a]]$
+and $[[nf a]]$ to indicate that there exists $[[e]]$ or $[[f]]$ such
+that $[[a = e]]$ or $[[a = f]]$.
+
+We can show that parallel reduction preserves $\beta$-normal and
+neutral forms.
+\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
+  \label{lemma:parnenf}
+  If $[[a => b]]$, then
+  \begin{itemize}
+  \item $[[ne a]]$ implies $[[ne b]]$
+  \item $[[nf a]]$ implies $[[nf b]]$
+  \end{itemize}
+\end{lemma}
+This lemma is mainly to show that \rref{P-AbsEta} does not introduce
+any new $\beta$ redex because $\eta$ reduction is the only real step
+we can take.
+
+The predicates $[[wne a]]$ and $[[wn
+a]]$ describe terms that can evaluate into $\beta$-neutral or
+$\beta$-normal form through parallel reduction and are defined as
+follows.\scw{What does $wne$ stand for?}
+\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
+\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
+
+The updated logical relation and its auxiliary definitions are shown
+in Figure~\ref{fig:logrelopen}.
+
+% The idea of a term blocked from $\beta$ reduction is
+% captured by $[[e]]$, the set of neutral terms. We augment the
+% interpretation of booleans and the empty types with terms that
+% evaluate to neutral terms. Furthermore, types that are neutral terms
+% can also be assigned a meaning since they may be inhabited by other
+% neutral terms.
+
+In the definition of the logical relation, we omit the rules for
+the function and universe cases since they remain identical to the
+original version in Figure~\ref{fig:logrel}.
+The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
+An open term of type $[[Bool]]$ does
+not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
+a variable, or more generally, a neutral term. Likewise, the
+$[[Void]]$ type, while remains uninhabited under an empty context, may
+be inherited by the set of neutral terms when there is a variable in
+the context that allows us to inhabit $[[Void]]$.
+
+The rule for equality type $[[a ~ b : A]]$ is augmented with the preconditions that
+$[[a]]$, $[[b]]$, and $[[A]]$ are all in their normal forms since
+otherwise our model would include equality types that are themselves
+not normalizing. Furthermore, the side condition $[[a <=> b]]$ is only
+available when the equality proof reduces to $[[refl]]$. If the proof
+term reduces to a neutral term, then there is nothing we can learn
+about the relationship between $[[a]]$ and $[[b]]$.
+
+Finally, in a
+non-empty context, a type itself may evaluate to a neutral term and in
+turn can only inhabited by neutral terms, thus the addition of \rref{I-Ne}.
+
+Interestingly, all the properties we have shown in
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the
+fundamental lemma can be proven in the exact same order, where the new
+cases due to \rref{I-Ne} and the augmentation of neutral terms
+to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
+Lemma~\ref{lemma:parnenf}.
+
+Before we can prove the fundamental theorem and derive the
+normalization property as its corollary, we need to additionally
+formulate and prove the adequacy property.
+
+Let us first start with some auxiliary lemmas and definitions.
+\begin{lemma}[Ext Wn]
+  \label{lemma:extwn}
+  If $[[wn a var i]]$, then $[[wn a]]$.
+\end{lemma}
+\begin{proof}
+  By induction over the length of the reduction sequence in $[[wn a
+  var i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
+  \ref{lemma:parnenf}.
+\end{proof}
+
+\begin{lemma}[wne wn]
+  \label{lemma:wnewn}
+  If $[[wne a]]$ and $[[wn b]]$, then $[[wne a b]]$.
+\end{lemma}
+\begin{proof}
+  Immediate by lexicographical induction over the length of the reduction sequences in
+  $[[wne a]]$ and $[[wn b]]$.
+\end{proof}
+
+\scw{What does CR stand for?}
+\begin{definition}[CR]
+  Let $[[S]]$ be a set of lambda terms. We say that
+  \begin{itemize}
+  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
+  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
+  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
+  \end{itemize}
+\end{definition}
+
+\begin{lemma}[$CR_1$ for type]
+  \label{lemma:cr1ty}
+  If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
+\end{lemma}
+\begin{proof}
+  Immediately from \rref{I-Ne} and \rref{I-Par}.
+\end{proof}
+
+\begin{lemma}[$CR$ for interpreted sets]
+  \label{lemma:crel}
+  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $[[Interp I i A S]]$. The base
+  cases are all immediate.
+
+  In the function case ($[[Pi A B]]$), we use the induction
+  hypothesis to show that variables, which are special cases of
+  neutral terms, semantically inhabit the input type $[[A]]$. We apply the
+  function to an arbitrary variable and we can then use
+  Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.
+
+  To show
+  the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
+  such that $[[wne b]]$,
+  and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
+  us a term that semantically inhabits $[[B {a}]]$. From the induction
+  hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
+  holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
+  from the induction hypothesis, since every term that evaluates to
+  some neutral form semantically inhabits $[[B {a}]]$. Done.
+\end{proof}
+
+\begin{lemma}[$CR_2$ for type]
+  \label{lemma:cr2ty}
+  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $[[Interp I i A S]]$. The
+  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
+\end{proof}
+
+\begin{lemma}[$CR_2$ for types (Rec)]
+  \label{lemma:cr2tyrec}
+  If $[[InterpR i A S]]$, then $[[wn A]]$.
+\end{lemma}
+\begin{proof}
+  By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
+\end{proof}
+
+\begin{lemma}[$CR$ for terms (Rec)]
+  \label{lemma:crelrec}
+  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
+\end{lemma}
+\begin{proof}
+  After unfolding the definition of $[[InterpR i A S]]$, trivial by
+  Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
+\end{proof}
+
+The formulation of the valuation and semantic well-typedness from
+Figure~\ref{fig:semtyping} and the fundamental lemma remains
+unchanged.
+The proof of the fundamental lemma is still carried out by induction
+over the typing derivation, where the additional neutral term related
+cases are handled by Lemmas~\ref{lemma:cr2tyrec} and
+\ref{lemma:crelrec}.
+
+The normalization property then follows as a corollary of the
+fundamental theorem.
+\begin{corollary}[Existence of $\beta$-normal form]
+  \label{corollary:exbetanf}
+  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
+\end{corollary}
+\begin{proof}
+  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
+  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
+  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
+  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
+  natural numbers as term variables. The side condition $[[idtm |=
+  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
+  including variables, semantically inhabit any $[[S0]]$ where
+  $[[S0]]$ is the interpretation of some type. With our choice of
+  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
+  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
+  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
+  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
+\end{proof}
+From Corollary~\ref{corollary:exbetanf}, we can show that there exists
+$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
+preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
+and also strictly decreases the size of the term.
+
+Due to the non-deterministic nature of parallel reduction, we need to
+take a few more steps to convert the existence of $\beta\eta$-normal
+form into a decision procedure for type conversion. More specifically,
+we can show that a deterministic evaluation strategy such as
+leftmost-outermost reduction can always find the $\beta\eta$-normal
+form if there exists one. However, we omit such proofs since they can
+can be formulated on untyped lambda terms and thus orthogonal to the
+specifics of dependently typed systems. Instead, we redirect readers
+to \citet{factorization-essentially, takahashi-parallel-reduction} for
+the details.
+
+\scw{Name and foreshadow these lemmas when they are introduced so that
+it is easier to refer to them in the discussion.}
+Finally, we want to point out that
+Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
+\ref{lemma:crelrec}, often referred to as adequacy of the logical
+relation, are in fact required in the normalization proof for simply
+typed languages~\citep{abel2019poplmark}. Similar to the simply typed
+scenario, adequacy needs to be proven before the fundamental theorem
+so we can handle rules such as \rref{T-If} where the scrutinee is a
+neutral term. In \citet{abel2019poplmark}, a variant of
+Lemma~\ref{lemma:extwn} is used in the exact same way to show that
+lambda terms are themselves normalizing as we have done in
+Lemma~\ref{lemma:crel}.
+
+Dependent types make the proof slightly more complicated
+as we also need to know that every type has a normal form. Whereas the
+$CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
+derivable, the $CR_2$ property for types (Lemma~\ref{lemma:cr2ty})
+requires the $CR_1$ property about the interpreted sets
+(Lemma~\ref{lemma:crel}).
+
+Overall, despite the dependently typed setting,
+the extension of our logical relation to prove normalization of open
+\emph{and} closed terms closely mirrors the progression from
+normalization of closed terms~\citep{harpertait} to normalization of
+open terms~\citep{harperkripke} in the simply typed lambda calculus.
+It is in fact reassuring that once we have laid the foundational
+technique for handling dependent types in our logical relation, the
+further extensions more or less boil down to properties that can be
+derived through syntactic means, which tend to not get in the way of
+Gödel's second incompleteness theorem.
+
+\section{Mechanization}
+\label{sec:logrelmech}
+\begin{figure}[h]
+  \begin{tabular}{ c |  c  | c  }
+    & Consistency & Normalization \\
+    \hline
+    Library (Autosubst 2)  & 491 & - \\
+    Syntactic typing (specification) &  69 & - \\
+    Renaming (common)  & 46 & -  \\
+    Syntactic soundness & 660 &  - \\
+    Untyped reduction & 350 & 834 \\
+    Logical relation & 342 & 515 \\
+    Semantic typing and soundness & 169 & 197 \\
+  \end{tabular}
+  \caption{Statistics of the Coq Development}
+  \label{fig:linecount}
+\end{figure}
+
+Figure~\ref{fig:linecount} shows the statics of our
+development, including the base consistency proof from
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
+extension to normalization for open and closed terms from
+Section~\ref{sec:extension}.
+
+Since the normalization and the consistency development share the same
+syntactic and typing specification, they differ in only the categories
+that are related to the logical relation. We use - as a marker that
+the line count is the same.
+
+Autosubst 2 takes our syntax specification in higher-order abstract
+syntax and generates Coq syntax specification, renaming and
+substitution functions, and lemmas and tactics that allow reasoning
+about those functions. The auto-generated files and the library header
+files are counted toward the \emph{Library (Autosubst 2)} category.
+
+Orthogonal to the development of our consistency and normalization
+proof, we prove the syntactic soundness of our system through subject
+reduction and progress. The syntactic soundness proof shares the
+definition of renaming with our semantic soundness proof, which is
+factored out as a separate file under the Renaming (common) category.
+
+\subsection{Artifacts Specific to Coq}
+In this section, we discuss the Coq encoding of the definitions and proofs presented
+in Section~\ref{sec:logrelproof}  and the artifacts that are
+specific to Coq and may not appear in other proof assistants.
+
+The powerset $[[PowerSet STm]]$ is encoded as the type \texttt{tm ->
+Prop}, a predicate over \lang{} terms.
+The inductive
+definition of the logical relation in Figure~\ref{fig:logrel} requires
+the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
+the function $[[F]]$ can be later instantiated into the logical
+relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
+
+In Coq, there is a distinction between computable functions and
+relations that can later be proven to be functional. The former can be
+viewed as a strict subset of the latter in axiom-free Coq. To be more
+precise, given a relation \texttt{R : A -> B -> Prop} subject to the
+totality and functionality constraints ($\forall$ \texttt{a} $\in$
+\texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
+that \texttt{R a b} is inhabited), we do not immediately obtain a function
+\texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
+\texttt{R a (F a)} is inhabited. However, the functional side
+conditions of a relation is clunky to express and tend to block
+automation. A simple workaround is to assume the axiom of unique choice, which
+is known to be consistent with Coq and allows us to induce a function \texttt{F
+: A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
+is functional. This approach would make our Coq development match
+the text version of our proof from Section~\ref{sec:logrelproof} more
+closely.
+
+However, we choose instead an axiom-free workaround and define
+\rref{I-Pi} as follows in our Coq mechanization.
+\begin{center}
+  \drule[]{I-PiCoq}
+\end{center}
+It should be easy to verify that the preconditions of
+\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
+equivalent. After establishing Lemma~\ref{lemma:logreldeter}, it is
+possible to further show that the conclusions of the rules are
+equivalent, too.\scw{If it is easy, you should have already done it.
+Unless it is really long and boring.}
+
+% This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
+% before we even show that our logical relation is
+% functional/deterministic, but does not affect the proof structure
+% otherwise.
+We choose to keep this discrepancy between the Coq development and the
+description of the logical relation presented in Section~\ref{sec:logreldep} since the
+skolemization process is more intuitively expressed in terms of
+function symbols rather than relation symbols. Otherwise, we do not
+see a clear advantage of \rref{I-PiCoq} over \rref{I-Pi} in set
+theory, where there is no distinction between computable functions and
+functions in general. \scw{not sure I follow this last bit}
+
+\subsection{Automation}
+\label{sec:automation}
+Our Coq mechanization heavily uses automation, though instead of
+defining custom tactics, we rely mostly on off-the-shelf tools such as
+Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.
+
+We use the Autosubst 2 framework to specify our syntax in HOAS and
+produce Coq syntax files in de Bruijn representation. Additionallly,
+Autosubst 2 provides a powerful tactic \texttt{asimpl} that
+can be used to prove the equivalence of two terms constructed using
+the primitive operators provided by the framework. This greatly
+simplifies the reasoning about substitution in our development as
+almost all substitution related properties are immediately discharged
+by \texttt{asimpl} without having to manually prove or even directly invoke any
+substitution related lemmas. \scw{You are only referring to lemmas about
+how substitution relates to syntax, right? You still need to manually
+invoke lemmas about substitution and judgements.}
+
+For other automation tasks that are not specific to binding, we use
+the powerful \texttt{sauto} tactic provided by CoqHammer to write
+short and declarative proofs. For example, here is a one-line proof of
+the triangle property about parallel reduction, from which the diamond
+property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
+\scw{you have space, so repeat the mathematical statement of the lemma}
+\begin{minted}{coq}
+Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
+Proof.
+  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
+Qed.
+\end{minted}
+In prose, the triangle property can be proven by induction over the
+graph of the \texttt{tstar} function, which stands for the Takahashi
+translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
+\texttt{ctrs:Par} say that the proof involves inverting and
+constructing of the derivations of parallel
+reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
+automation tactic to use
+the reflexivity and congruence properties of parallel
+reduction as lemmas.
+\scw{What does \texttt{lq:on} and \texttt{hauto} mean?}
+
+The automation here not only gives us a proof that is shorter and more
+resiliant to changes in definition, but also gives useful
+documentation for readers who wish to understand how the underlying
+proof works since the automation tactic is guaranteed to not use
+lemmas or invert derivations that are not specified in the
+\texttt{use} or \texttt{inv} flag.
+
+\section{Related Work and Discussion}
+\label{sec:relatedwork}
+
+\scw{Maybe it would be useful to include a chart here, so that readers can
+  easily keep track of the features of the various languages.  i.e. which ones
+  include large eliminations? type-directed equivalence? impredicative prop?
+  inductive datatypes? what are their line counts?}
+
+\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
+\citet{barendregt:lambda-calculi-with-types} are some of the earlier
+works that establish metatheoretic results strong enough to derive
+consistency for dependently typed systems.
+\scw{Should give Martin L\"of credit for first consistency proof.}
+\scw{What techniques do these three proofs use? How is it different from
+your proof? What about Luo, which
+introduced a universe hierarchy? What about categories with families, or
+other ways of showing consistency?}
+Some of these techniques
+are still in use in more recent works. For example, the technique for
+proving strong normalization by \citet{geuvers1994short} is adapted by
+\citet{moon2021graded} to show the same property to a dependently
+typed system extended with modalities.
+
+\citet{decagda} mechanizes in Agda the decidability of type
+conversion rule for a dependently typed language with one predicative
+universe level and typed judgmental equality with function
+$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
+uses a Kripke-style logical relation parameterized over an
+type-directed equivalence relation satisfying certain
+properties. Their logical relation is defined using the
+induction-recursion scheme, which is available in Agda but not in Coq.
+\citet{martin-lof-a-la-coq} manages to encode the logical relation
+from \citet{decagda} in the predicative fragment of Coq using a
+special way of encoding induction recursion. Their work further
+extends the decidability of type conversion result from~\citet{decagda} to the decidability
+type checking of a bidrectional type system.
+
+\citet{anand2014towards} mechanizes the metatheory of
+Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
+extensional type theory with features such as dependent functions,
+inductive types, and a full universe hierarchy. \citet{nbeincoq}
+mechanizes the normalization-by-evaluation algorithm in Coq for a
+dependently typed language with one predicative universe, similar to
+\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
+\citet{anand2014towards} and \citet{nbeincoq} leverage the
+impredicative \texttt{Prop} sort of Coq to define the interpretation
+of dependent function types and thus are closely related to our
+mechanization. However, instead of explaining the purely inductive
+logical relation as a convoluted workaround to the lack of
+induction-recursion in Coq, we give a self-contained explanation of
+our logical relation.
+
+Finally, it is worth pointing out that all systems we have discussed
+so far builds a relational model for their logical relation. Recall
+that our logical relation takes the form $[[InterpR i A S]]$ where
+$[[S]]$ is a set of terms. In a relational model, each type is
+interpreted as a partial equivalence relation over terms rather than a
+set. A relational model is necessary for an extensional type theory
+such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
+use a relational model either to justify the function $\eta$-law or to
+derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
+law for function is baked into the untyped reduction
+relation and therefore a relational model is not needed.
+\scw{would it be difficult to extend your proof to be a PER?}
+
+The dependently typed systems we have discussed so far also vary
+greatly in expressiveness. For example, the Calculus of
+Constructions~\citep{CoC} and MLTT without universe levels are unable
+to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
+identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
+the mechanized systems we have discussed so far, only
+\citet{anand2014towards} has a full universe hierarchy.
+Our object language \lang{}, while small, has a decent coverage of
+features commonly seen in dependently typed languages, including
+dependent pattern matching, dependent function types, a full universe
+hierarchy, an intensional identity type. We hope features such as
+type-directed reduction (necessary for unit $\eta$-law) and
+impredicative sorts can be implemented in a similarly streamlined
+fashion and we will leave those extensions as part of our future work.
+
+% To end this section, we
+\scw{should this be a new subsection?}
+The final question we want to address is: why is our proof,
+even with the extension from Section~\ref{sec:extension}, so much
+shorter than the proofs from \citet{decagda, nbeincoq,
+  martin-lof-a-la-coq}?
+First, unlike developments that mechanize the correctness of a clever
+type conversion algorithm, we only show the existence of
+normal form for open and closed terms and state our properties in
+terms of the untyped small-step reduction relation. This removes a lot
+of scaffolding in the specification of our type system and our logical
+relation.
+
+This contrasts our development with
+\citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
+difficult property to prove due to the complexity of the equality
+judgment used to define the logical relation. In our system, the
+irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
+direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
+lemma name} which is
+straightforward after we prove the results about parallel
+reduction in Section~\ref{sec:spec}. In other words, we keep the
+the logical relation itself minimal at the cost of extra lemmas that can be proven
+through syntactic means independently from the logical relation.
+For example, as commented near the end of Section~\ref{sec:extension},
+our normalization property (Corollary~\ref{corollary:exbetanf})
+% can induce a decision procedure for type conversion by reducing two
+% terms to $\beta\eta$-normal form and check for their syntactic
+% equivalence. This decision procedure would be inefficient compared to
+% NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
+% to weak head normal and checks for the equivalence of subterms
+% recursively
+does not immediately induce an efficient decision procedure, but it is
+possible to recover an efficient algorithm by reasoning about untyped
+lambda terms.
+
+% the concision of
+% our proof can be largely attributed to the fact that we bake into the
+% logical relation only the minimum amount of information that is the
+% most conducive to the proof of the metatheoretic result.
+
+Another key to our proof is the very early establishment of the confluence
+property of the paralell reduction relation
+(Lemma~\ref{lemma:pardiamond}). This property is used in
+Lemma~\ref{lemma:interppreservation} to show that the interpretation
+of a type is preserved under evaluation. That is, if $[[InterpR i A
+S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
+In a system with typed directed reduction, the confluence result is
+harder to establish since confluence requires subject reduction, which
+circularly depends on $\Pi$-injectivity, a consequence of
+confluence~\citep{siles2012pure}. There are different workarounds to
+this problem. \citet{siles2012pure} proposes a syntactic approach by
+defining a type system where the confluence result is
+directly provable and later show that the system of interest is
+equivalent to the system with the confluence property.
+\citet{decagda} uses a relational model and the relational counterpart
+of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
+model is parameterized by a typed indexed equivalence relation and
+is reused to prove derive different properties in their
+development. Another perhaps simpler approach is to simply
+extend the judgmental equality with rules about
+$\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
+reduction to be proven independently from confluence and
+$\Pi$-injectivity can later be shown to be admissible. We find the
+approach from \citep{weirich:systemd} the most lightweight as it
+allows us to derive confluence early on even in systems with typed
+conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}
+
+In cases where the judgmental equality does not leverage type information heavily
+(e.g. $\eta$-law for the unit type), it might be possible to show that
+$\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
+$[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
+property holds (we believe this is likely true for both
+\citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
+terms of untyped parallel reduction to show the decidability of type
+conversion. Otherwise, to fully model the system, we need to bake
+typing information into our logical relation and we need to redefine
+our logical relation in Kripke-style so we can relax the scoping
+constraint in order to prove adequacy, though a relational model would
+still be unnecessary as long as the $\eta$ laws are baked into the
+typed directed relation. \scw{not sure I understand this last bit}
+
+Since confluence plays such as key role in our proof, we are uncertain
+how to generalize our technique to systems where confluence does not
+hold, and we will investigate those scenarios in our future work.
+\scw{What systems are you thinking of? Or do you mean more generally
+that this technique only works for languages where the definition of
+equality is based on joinability of parallel reduction?}
+
+% We are able to avoid the complexivity of a Kripke-style logical relation
+% by defining our logical relation in terms of untyped parallel
+% reduction, which does not impose any scoping constraint on the
+% terms.
+
+% Why is it that the justification of $\eta$ law
+% requires a Kripke-style relational model in all three developments
+% but not ours? Are we ``cheating'' by defining \lang{}'s equational
+% theory using an untyped conversion rule? Instead of answering the
+% questions one by one, we clarify the design of our type system and our
+% proof technique so the answers to those questions are self-evident.
+
+% The key to the shortness of
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+% Type soundness can be proven through a syntactic
+% approach~\citep{syntacticsoundness} as a corollary of two properties:
+% progress and preservation. % The syntactic type soundness proof
+% % varies in complexity depending on the underlying type
+% % system. For example, a type system that tracks information flow would
+% % require additional structural rules related to security levels. In
+% % this paper, we focus on one specific type of complexity: the
+% In Figure~\ref{fig:stlcsoundness}, we summarize the structure of the
+% syntactic type soundness proof for the simply typed lambda
+% calculus. Each lemma can be proven by structural induction over the
+% typing derivation, while using the previous established results as
+% lemmas for specific cases that do not immediately follow from the
+% induction hypothesis. If we make our language more complex by adding
+% full dependent type support, the overall structure remains almost
+% identical.
+
+% NbE in Coq
+
+%%
+%% The next two lines define the bibliography style to be used, and
+%% the bibliography file.
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{refs}
+
+\appendix
+
+\subsection{More detailed proofs}
+\begin{lemma}[Preservation of the logical relation]
   If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
@@
 \end{proof}
-From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
-derive the following corollary that two coherent types have the same
-interpretation.
-\scw{I'm not used to this lemma being called irrelevance. Intuitively,
-it says that the definition of the logical relation respects
-definitional equality. }
-\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
-  \label{lemma:logrelcoherence}
-  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
-\end{corollary}
-\scw{Need a transition here}
-Next, we show that $[[Interp I i A S]]$ is in fact a partial function.
+
 \begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
-  \label{lemma:logreldeter}
   If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
@@
 
-Lemma~\ref{lemma:logreldeter} enables us to show the admissibility of
-\rref{I-PiAlt} and its related inversion principle.
-\begin{lemma}[Pi Intro Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
-  \label{lemma:piintroalt}
-  Suppose the following two statements hold:
-  \begin{itemize}
-  \item $[[Interp I i A S]]$
-  \item $[[forall a, a in S implies (# exists S0 , Interp I i B {a} S0 #)]]$
-  \end{itemize}
-  Then we have $[[Interp I i Pi A B { b | forall a, (# a in S , forall
-    S0, (# Interp I i B {a} S0,  b a in F a #) #) }]]$
-\end{lemma}
-\begin{proof}
-  Let $[[F]]$ be the relation defined as follows:
-  \[ (a,S) \in [[F]] \iff [[Interp I i B {a} S]] \]
-  By the second bullet from the premise and
-  Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a function that is total
-  on the set $[[S]]$. The conclusion then trivially follows from
-  \rref{I-Pi}.
-\end{proof}
-
-\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
-  \label{lemma:piinvalt}
-  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
-  such that the following constraints hold:
-  \begin{itemize}
-  \item $[[Interp I i A S0]]$
-  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
-    S1 #) #)]]$
-  \item $[[S = { b | forall a, (# a in S0 , forall
-    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
-\end{proof}
-
-% The combination of Lemmas~\ref{lemma:piintroalt} and
-% \ref{lemma:piinvalt} reveals what we truly want to capture with
-% \rref{I-Pi}. A function type is semantically well-defined if its input
-% type is well-defined, and its output type is defined for each valid
-% inahbitant of its input type.
-
-The next lemma shows that our logical relation satisfies
-cumulativity. That is, if a type has an interpretation at a lower
-universe level, then we can obtain the same interpretation at a higher
-universe level.
-\begin{lemma}[Logical relation cumulativity\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
-  \label{lemma:logrelcumulativity}
-  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
-\end{lemma}
-\begin{proof}
-  Trivial by structural induction over the derivation of $[[Interp I
-  i0 A S]]$.
-\end{proof}
-Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
-implicitly assume that $[[I]]$ is defined on the set of natural
-numbers less than $[[i1]]$.
-
-\begin{corollary}[Logical relation is functional with different levels\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
-  \label{lemma:logreldeterhet}
-  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
-\end{corollary}
-\begin{proof}
-  Immediate from Lemma~\ref{lemma:logreldeter} and
-  \ref{lemma:logrelcumulativity}.
-\end{proof}
-
-We say that a set of terms $[[S]]$ is closed under expansion if given
-$[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
-The final property we want to show is that the output set $[[S]]$ from
-the logical relation is closed under expansion. Unlike the previous
-lemmas, we need to constrain the function $[[I]]$ so its outputs are
-all closed under expansion.
-\scw{How is this related to the comment before 3.2 (i.e. the property
-derived from I-Red?}
-\begin{lemma}[Logical Relation Elements back preservation\footnote{\dotv{semtyping.v}{InterpExt\_back\_clos}}]
-  \label{lemma:logrelbackclos}
-  If $[[Interp I i A S]]$ and $[[I]]$ satisfies the property that
-  for all $[[i]]$, $[[I i]]$ is closed under expansion, then the set
-  $[[S]]$ is closed under expansion.
-\end{lemma}
-\begin{proof}
-  Trivial by structural induction over the derivation of $[[Interp I i
-  A S]]$. The function case requires the following simple fact about
-  parallel reduction:
-  If $[[b0 => b1]]$ then $[[b0 a => b1 a]]$ for all $[[a]]$. This fact
-  is not immediate from the definition of parallel reduction but
-  follows from Lemma~\ref{lemma:parrefl} and \rref{P-App}.
-\end{proof}
-
-\begin{corollary}[Logical Relation Elements back preservation (rec)\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
-  \label{lemma:logrelNbackclos}
-  If $[[InterpR i A S]]$, then $[[S]]$ is closed under expansion.
-\end{corollary}
-\begin{proof}
-  Immediate from Lemma~\ref{lemma:logrelbackclos}, the definition of
-  $[[InterpR i A S]]$, and \rref{I-Red}.
-\end{proof}
-
-\section{Semantic Typing and Consistency}
-\label{sec:logrelproof}
-\begin{figure}[h]
-\[
-\begin{array}{lcl}
-
-      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
-                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
-      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
-                       G]]\text{ then there exists some } [[j]] \text{
-                       and } [[S]] \text{ such that } [[InterpR j A
-                       {rho} S]] \text{ and } [[a {rho} in S]] \\
-      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
-\end{array}
-\]
-  \caption{Semantic Typing for \lang}
-  \label{fig:semtyping}
-\end{figure}
-
-The logical relation we define in Figure~\ref{fig:logrel} does not
-include cases for variables. Likewise, for the base types such as
-boolean and equality, the output set $[[S]]$ contains only terms that
-evaluate to closed terms.
-
-\scw{What is the motivation for this section? Why do we want to generalize
-the logical relation?}
-To generalize our logical relation to open
-terms, we define the semantic typing judgment by closing the open
-terms with a substitution whose codomain consists of terms that
-respect the interpretation of the types from the context. The full
-definitions of well-formed substitution ($[[rho |= G]]$), semantic
-typing ($[[ G |= a : A]]$), and semantic context well-formedness
-($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
-\scw{explain all parts of this figure. it's pretty intimidating.}
-In the
-definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
-the typing context obtained by dropping the last $[[i]]$ elements of
-$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
-returns the empty list.
-
-The following lemma makes the statement $[[G |= A : Set i]]$ easier to
-work with.
-\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}\scw{spell out the whole word}]
-  \label{lemma:setinv}
-  The following two statements are equivalent:
-  \begin{itemize}
-  \item $[[G |= A : Set i]]$
-  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
-    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  The forward direction is immediate by
-  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
-  and show that $[[G |= A : Set i]]$ given the second bullet.
-
-  Suppose $[[rho |= G]]$, then we know that there exists some $[[S]]$
-  such that $[[InterpR i (A {rho}) S]]$. By the definition of semantic
-  typing, it suffices to show that there exists some $[[j]]$ and
-  $[[S0]]$ such that  $[[InterpR j Set i S0]]$ and $[[A {rho} in
-  S0]]$.
-  Pick $[[Suc i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
-  ]]$ for $[[S0]]$ and it is trivial to verify the conditions hold.
-\end{proof}
-
-The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
-counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
-than inductively over the context. \scw{why is that the case?} It is easy to recover the same
-structural rules:
-\begin{lemma}[Semantic context well-formedness cons]
-  \label{lemma:semwffcons}
-  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
-\end{lemma}
-\scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
-Ctx-empty?}
-\begin{proof}
-  By the definition of semantic context well-formedness, the goal is
-  to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i
-  (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
-  proven by case analysis on whether $[[i]]$ is zero.
-\end{proof}
-
-Likewise, the semantic well-formendness judgment for substitutions
-satisfies similar structural rules.
-\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
-  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
-  $[[rho .: a |= G ++ A]]$.
-\end{lemma}
-\begin{proof}
-  Start by unfolding the definition of $[[rho .: a |= G ++ A]]$ and
-  performing a case analysis similar to the proof of
-  Lemma~\ref{lemma:semwffcons}.  The case where the number is $[[0]]$
-  requires Lemma~\ref{lemma:logreldeterhet} to finish the proof.
-\end{proof}
-
-Next, we show some non-trivial cases of the fundamental theorem as
-top-level lemmas and leave the remaining cases as exercises for the
-reader.\scw{I dislike leaving ``exercises to the reader''}
-
-First, we formulate the definition of valid renamings and prove that
-semantic typing satisfies renaming so we can weaken the context when
-reasoning about the variable case of the fundamental lemma
-(Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G
-++ D]]$, it is easy to show that we can extract some valuation
-$[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
-``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :
-A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
-$[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
-after weakening the context; this implication holds because $[[rho |=
-G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
-can make use of the premise $[[G |= a : A]]$ to derive what we need
-for the conclusion. We recommend the readers to skip the proofs of
-Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
-during the first read as long as they have an intuitive understanding
-of what the renaming property is meant to capture.
-
-\scw{Make this a definition? Would it be useful to create notation
-  for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
-We say that $[[xi]]$ is
-valid from the context $[[G]]$ to the context $[[D]]$ if the
-following condition holds.
-\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
-  [[D ren xi
-  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]
-
-\begin{lemma}[Truncate is valid]
-  \label{lemma:validtruncate}
-  If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i
-  G]]$ to $[[G]]$.
-\end{lemma}
-\begin{proof}
-  By the definition of a valid renaming, we must show that given $[[j
-  < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
-  \begin{itemize}
-  \item $[[ren up i j]] = i + j  < [[| G | ]]$
-  \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
-  \end{itemize}
-  The first bullet point is immediate from the fact that $[[j]] <
-  [[|G|]] - [[i]]$. The second bullet follows from unfolding the
-  definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
-  $[[i]] + [[j]] < [[|G|]]$.
-\end{proof}
-
-\begin{lemma}[Renaming for $[[rho |= G]]$]
-  \label{lemma:renamingval}
-  If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
-  $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
-\scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
-\end{lemma}
-\begin{proof}
-  Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
-  that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i
-  < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.
-
-  By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
-have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho
-} ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |=
-D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
-S]]$ and $[[rho ren xi i in S]]$.
-\end{proof}
-
-\begin{lemma}[Renaming for $[[G |= a : A]]$]
-  \label{lemma:semrenaming}
-  If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
-  $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
-\end{lemma}
-\begin{proof}
-  Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
-\end{proof}
-
-\scw{Need a transition here that you are starting to explain the semantic typing
-rules.}
-
-\begin{lemma}[ST-Var]
-  \label{lemma:stvar}
-  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
-\end{lemma}
-\begin{proof}
-  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
-need to show that there exists some $[[j]]$ and $[[S]]$ such that
-  \begin{itemize}
-  \item $[[InterpR j G i < up Suc i > { rho } S]]$
-  \item $[[rho i in S]]$
-  \end{itemize}
-  From the definition of $[[|= G]]$, we know that there exists some
-  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. By
-  Lemma~\ref{lemma:validtruncate}, we know that $[[up Suc i]]$ is a
-  valid renaming from $[[drop Suc i G]]$ to $[[G]]$. By
-  Lemma~\ref{lemma:semrenaming}, we deduce $[[G |= G i < up Suc
-  i > : Set j]]$. By the identity from Lemma~\ref{lemma:setinv}, we
-  know that $[[InterpR j G i < up Suc i > { rho } S]]$. It now
-  suffices to show $[[rho i in S]]$, but that is immediate from the
-  definition of $[[rho |= G]]$.
-\end{proof}
-
-\begin{lemma}[ST-Set]
-  \label{lemma:stset}
-  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
-\end{lemma}
-\begin{proof}
-  Immediate by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
-\end{proof}
-
-\begin{lemma}[ST-Pi]
-  \label{lemma:stpi}
-  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
-  A B : Set i]]$.
-\end{lemma}
-\begin{proof}
-  Applying Lemma~\ref{lemma:setinv} to the
-  conclusion, it now suffices to show that given $[[rho |= G]]$, there
-  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
-  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
-  there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
-From $[[G ++ A |= B : Set i]]$, we know that there must
-exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
-in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
-\end{proof}
-
-\begin{lemma}[ST-Abs]
-  \label{lemma:stabs}
-  If $[[G |= Pi A B : Set i]]$ and $[[G ++ A |= b : B]]$, then $[[G |=
-  \ A b : Pi A B]]$.
-\end{lemma}
-\begin{proof}
-  By unfolding the definition of $[[G |= \ A b : Pi A B]]$, we need to
-  show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
-  $[[S]]$ such that $[[InterpR i Pi A {rho} B {up rho} S]]$ and $[[\
-  A{rho} b{up rho} in S]]$.
-
-  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
-  i]]$, there exists some set $[[S]]$ such that
-  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
-  $[[\A{rho} b{up rho} in S
-  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
-  that all following conditions hold:
-  \begin{itemize}
-  \item $[[InterpR i A{rho} S0]]$
-  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
-    {rho .: a}
-    S1 #) #)]]$
-  \item $[[S = { b | forall a, (# a in S0 , forall
-      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
-  \end{itemize}
-  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
-  that given $[[a in S0]]$,
-  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
-  a in S1]]$.
-  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
-  under expansion. By Lemma~\ref{lemma:parabscbn}, since $[[( \A{rho} b{up rho} )
-  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
-  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
-  B]]$ and the fact that the logical relation is deterministic and
-  cumulative (Lemma~\ref{lemma:logreldeterhet}).
-\end{proof}
-
-\begin{lemma}[ST-App]
-  \label{lemma:stapp}
-  If $[[G |= b : Pi A B]]$ and $[[G |= a : A]]$, then $[[G |= b a : B {a}]]$.
-\end{lemma}
-\begin{proof}
-Suppose $[[rho |= G]]$. The goal is to show that there exists some
-$[[i]]$ and $[[S1]]$
-such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a} {rho}
-S1]]$, or equivalently, $[[InterpR i B {rho .: a {rho}} S1]]$ since
-$[[B {a}{rho}]] = [[B {rho .: a {rho}}]]$. By the premise $[[G |= b :
-Pi A B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
-there exists some $[[i]]$ and $[[S0]]$ such that:
-  \begin{itemize}
-  \item $[[InterpR i A{rho} S0]]$
-  \item $[[forall a0, (# a0 in S0 implies (# exists S1 , InterpR i B
-    {rho .: a0}
-    S1 #) #)]]$
-  \item $[[forall a0, (# a0 in S0 , forall
-      S1, (# InterpR i B {rho .: a0} S1,  b {rho} a0 in S1 #) #)]]$
-  \end{itemize}
-  Instantiating the variable $[[a0]]$ from the last two bullets with
-  the term $[[a {rho}]]$, the conclusion immediately follows.
-\end{proof}
-
-\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
-  \label{theorem:soundness}\leavevmode
-  \begin{itemize}
-  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
-  \item If $[[|- G]]$, then $[[|= G]]$.
-  \end{itemize}
-\end{theorem}
-\begin{proof}
-  Proof by mutual induction over the derivation of $[[G |- a :
-  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
-  from Lemma~\ref{lemma:semwffcons}.
-Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
-  can be used to discharge their syntactic counterpart
-  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
-  cases not covered by the lemmas are similar to the ones already
-  shown or simpler and therefore omitted from the text.
-
-\end{proof}
-
-\scw{Spell this out in a little more detail.}
-
-\begin{proof}
-  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
-\end{proof}
-
-\section{Toward Decidability of Type Conversion and $\eta$ laws}
-\label{sec:extension}
-\begin{figure}[h]
-  \[
-    \begin{array}{lcl}
-      % \beta\text{-}\mathit{neutral\ terms}\\
-      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
-                    f]] \\ \\
-      % \beta\text{-}\mathit{normal\ terms} \\
-      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
-                    |\ [[f ~ f : f]]\\
-            & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
-    \end{array}
-  \]
-  \caption{$\beta$-neutral and normal forms}
-  \label{fig:nenf}
-\end{figure}
-\begin{figure}[h]
-  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
-  \caption{Extended logical relation}
-  \label{fig:logrelopen}
-\end{figure}
-
-\scw{Put this introduction before the figure in the section.}
-In this section, we sketch out how the logical relation from
-Section~\ref{sec:logreldep} can be extended show the existence of
-$\beta\eta$ normal forms for (open and closed) well-typed terms. We
-first extend parallel reduction to include $\eta$ reduction for
-functions.
-\begin{center}
-  \drule[width=2.5in]{P-AbsEta}
-\end{center}
-\scw{What happened to the domain type in the $\lambda$ expression? This is
-not your syntax...}
-\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
-equational theory since coherence, the untyped relation used for type
-conversion, is built on top of parallel reduction. We can recover the
-same confluence result about parallel reduction using the standard
-techniques from \citet{barendregt:lambda-calculi-with-types,
-takahashi-parallel-reduction}, though specifically for the de Bruijn
-representation, we need the following anti-renaming lemma about
-parallel reduction.
-\begin{lemma}[Par anti-renaming]
-  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
-exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
-\end{lemma}
-
-The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the neutral terms
-and normal forms with respect to $\beta$-reduction, but can still take
-$\eta$ reduction steps. We sometimes use the judgment forms $[[ne a]]$
-and $[[nf a]]$ to indicate that there exists $[[e]]$ or $[[f]]$ such
-that $[[a = e]]$ or $[[a = f]]$.
-
-We can show that parallel reduction preserves $\beta$-normal and
-neutral forms.
-\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
-  \label{lemma:parnenf}
-  If $[[a => b]]$, then
-  \begin{itemize}
-  \item $[[ne a]]$ implies $[[ne b]]$
-  \item $[[nf a]]$ implies $[[nf b]]$
-  \end{itemize}
-\end{lemma}
-This lemma is mainly to show that \rref{P-AbsEta} does not introduce
-any new $\beta$ redex because $\eta$ reduction is the only real step
-we can take.
-
-The predicates $[[wne a]]$ and $[[wn
-a]]$ describe terms that can evaluate into $\beta$-neutral or
-$\beta$-normal form through parallel reduction and are defined as
-follows.\scw{What does $wne$ stand for?}
-\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
-\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
-
-The updated logical relation and its auxiliary definitions are shown
-in Figure~\ref{fig:logrelopen}.
-
-% The idea of a term blocked from $\beta$ reduction is
-% captured by $[[e]]$, the set of neutral terms. We augment the
-% interpretation of booleans and the empty types with terms that
-% evaluate to neutral terms. Furthermore, types that are neutral terms
-% can also be assigned a meaning since they may be inhabited by other
-% neutral terms.
-
-In the definition of the logical relation, we omit the rules for
-the function and universe cases since they remain identical to the
-original version in Figure~\ref{fig:logrel}.
-The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
-An open term of type $[[Bool]]$ does
-not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
-a variable, or more generally, a neutral term. Likewise, the
-$[[Void]]$ type, while remains uninhabited under an empty context, may
-be inherited by the set of neutral terms when there is a variable in
-the context that allows us to inhabit $[[Void]]$.
-
-The rule for equality type $[[a ~ b : A]]$ is augmented with the preconditions that
-$[[a]]$, $[[b]]$, and $[[A]]$ are all in their normal forms since
-otherwise our model would include equality types that are themselves
-not normalizing. Furthermore, the side condition $[[a <=> b]]$ is only
-available when the equality proof reduces to $[[refl]]$. If the proof
-term reduces to a neutral term, then there is nothing we can learn
-about the relationship between $[[a]]$ and $[[b]]$.
-
-Finally, in a
-non-empty context, a type itself may evaluate to a neutral term and in
-turn can only inhabited by neutral terms, thus the addition of \rref{I-Ne}.
-
-Interestingly, all the properties we have shown in
-Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the
-fundamental lemma can be proven in the exact same order, where the new
-cases due to \rref{I-Ne} and the augmentation of neutral terms
-to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
-Lemma~\ref{lemma:parnenf}.
-
-Before we can prove the fundamental theorem and derive the
-normalization property as its corollary, we need to additionally
-formulate and prove the adequacy property.
-
-Let us first start with some auxiliary lemmas and definitions.
-\begin{lemma}[Ext Wn]
-  \label{lemma:extwn}
-  If $[[wn a var i]]$, then $[[wn a]]$.
-\end{lemma}
-\begin{proof}
-  By induction over the length of the reduction sequence in $[[wn a
-  var i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
-  \ref{lemma:parnenf}.
-\end{proof}
-
-\begin{lemma}[wne wn]
-  \label{lemma:wnewn}
-  If $[[wne a]]$ and $[[wn b]]$, then $[[wne a b]]$.
-\end{lemma}
-\begin{proof}
-  Immediate by lexicographical induction over the length of the reduction sequences in
-  $[[wne a]]$ and $[[wn b]]$.
-\end{proof}
-
-\scw{What does CR stand for?}
-\begin{definition}[CR]
-  Let $[[S]]$ be a set of lambda terms. We say that
-  \begin{itemize}
-  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
-  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
-  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
-  \end{itemize}
-\end{definition}
-
-\begin{lemma}[$CR_1$ for type]
-  \label{lemma:cr1ty}
-  If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
-\end{lemma}
-\begin{proof}
-  Immediately from \rref{I-Ne} and \rref{I-Par}.
-\end{proof}
-
-\begin{lemma}[$CR$ for interpreted sets]
-  \label{lemma:crel}
-  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
-\end{lemma}
-\begin{proof}
-  By induction over the derivation of $[[Interp I i A S]]$. The base
-  cases are all immediate.
-
-  In the function case ($[[Pi A B]]$), we use the induction
-  hypothesis to show that variables, which are special cases of
-  neutral terms, semantically inhabit the input type $[[A]]$. We apply the
-  function to an arbitrary variable and we can then use
-  Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.
-
-  To show
-  the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
-  such that $[[wne b]]$,
-  and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
-  us a term that semantically inhabits $[[B {a}]]$. From the induction
-  hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
-  holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
-  from the induction hypothesis, since every term that evaluates to
-  some neutral form semantically inhabits $[[B {a}]]$. Done.
-\end{proof}
-
-\begin{lemma}[$CR_2$ for type]
-  \label{lemma:cr2ty}
-  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
-\end{lemma}
-\begin{proof}
-  By induction over the derivation of $[[Interp I i A S]]$. The
-  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
-\end{proof}
-
-\begin{lemma}[$CR_2$ for types (Rec)]
-  \label{lemma:cr2tyrec}
-  If $[[InterpR i A S]]$, then $[[wn A]]$.
-\end{lemma}
-\begin{proof}
-  By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
-\end{proof}
-
-\begin{lemma}[$CR$ for terms (Rec)]
-  \label{lemma:crelrec}
-  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
-\end{lemma}
-\begin{proof}
-  After unfolding the definition of $[[InterpR i A S]]$, trivial by
-  Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
-\end{proof}
-
-The formulation of the valuation and semantic well-typedness from
-Figure~\ref{fig:semtyping} and the fundamental lemma remains
-unchanged.
-The proof of the fundamental lemma is still carried out by induction
-over the typing derivation, where the additional neutral term related
-cases are handled by Lemmas~\ref{lemma:cr2tyrec} and
-\ref{lemma:crelrec}.
-
-The normalization property then follows as a corollary of the
-fundamental theorem.
-\begin{corollary}[Existence of $\beta$-normal form]
-  \label{corollary:exbetanf}
-  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
-\end{corollary}
-\begin{proof}
-  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
-  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
-  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
-  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
-  natural numbers as term variables. The side condition $[[idtm |=
-  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
-  including variables, semantically inhabit any $[[S0]]$ where
-  $[[S0]]$ is the interpretation of some type. With our choice of
-  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
-  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
-  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
-  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
-\end{proof}
-From Corollary~\ref{corollary:exbetanf}, we can show that there exists
-$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
-preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
-and also strictly decreases the size of the term.
-
-Due to the non-deterministic nature of parallel reduction, we need to
-take a few more steps to convert the existence of $\beta\eta$-normal
-form into a decision procedure for type conversion. More specifically,
-we can show that a deterministic evaluation strategy such as
-leftmost-outermost reduction can always find the $\beta\eta$-normal
-form if there exists one. However, we omit such proofs since they can
-can be formulated on untyped lambda terms and thus orthogonal to the
-specifics of dependently typed systems. Instead, we redirect readers
-to \citet{factorization-essentially, takahashi-parallel-reduction} for
-the details.
-
-\scw{Name and foreshadow these lemmas when they are introduced so that
-it is easier to refer to them in the discussion.}
-Finally, we want to point out that
-Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
-\ref{lemma:crelrec}, often referred to as adequacy of the logical
-relation, are in fact required in the normalization proof for simply
-typed languages~\citep{abel2019poplmark}. Similar to the simply typed
-scenario, adequacy needs to be proven before the fundamental theorem
-so we can handle rules such as \rref{T-If} where the scrutinee is a
-neutral term. In \citet{abel2019poplmark}, a variant of
-Lemma~\ref{lemma:extwn} is used in the exact same way to show that
-lambda terms are themselves normalizing as we have done in
-Lemma~\ref{lemma:crel}.
-
-Dependent types make the proof slightly more complicated
-as we also need to know that every type has a normal form. Whereas the
-$CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
-derivable, the $CR_2$ property for types (Lemma~\ref{lemma:cr2ty})
-requires the $CR_1$ property about the interpreted sets
-(Lemma~\ref{lemma:crel}).
-
-Overall, despite the dependently typed setting,
-the extension of our logical relation to prove normalization of open
-\emph{and} closed terms closely mirrors the progression from
-normalization of closed terms~\citep{harpertait} to normalization of
-open terms~\citep{harperkripke} in the simply typed lambda calculus.
-It is in fact reassuring that once we have laid the foundational
-technique for handling dependent types in our logical relation, the
-further extensions more or less boil down to properties that can be
-derived through syntactic means, which tend to not get in the way of
-Gödel's second incompleteness theorem.
-
-\section{Mechanization}
-\label{sec:logrelmech}
-\begin{figure}[h]
-  \begin{tabular}{ c |  c  | c  }
-    & Consistency & Normalization \\
-    \hline
-    Library (Autosubst 2)  & 491 & - \\
-    Syntactic typing (specification) &  69 & - \\
-    Renaming (common)  & 46 & -  \\
-    Syntactic soundness & 660 &  - \\
-    Untyped reduction & 350 & 834 \\
-    Logical relation & 342 & 515 \\
-    Semantic typing and soundness & 169 & 197 \\
-  \end{tabular}
-  \caption{Statistics of the Coq Development}
-  \label{fig:linecount}
-\end{figure}
-
-Figure~\ref{fig:linecount} shows the statics of our
-development, including the base consistency proof from
-Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
-extension to normalization for open and closed terms from
-Section~\ref{sec:extension}.
-
-Since the normalization and the consistency development share the same
-syntactic and typing specification, they differ in only the categories
-that are related to the logical relation. We use - as a marker that
-the line count is the same.
-
-Autosubst 2 takes our syntax specification in higher-order abstract
-syntax and generates Coq syntax specification, renaming and
-substitution functions, and lemmas and tactics that allow reasoning
-about those functions. The auto-generated files and the library header
-files are counted toward the \emph{Library (Autosubst 2)} category.
-
-Orthogonal to the development of our consistency and normalization
-proof, we prove the syntactic soundness of our system through subject
-reduction and progress. The syntactic soundness proof shares the
-definition of renaming with our semantic soundness proof, which is
-factored out as a separate file under the Renaming (common) category.
-
-\subsection{Artifacts Specific to Coq}
-In this section, we discuss the Coq encoding of the definitions and proofs presented
-in Section~\ref{sec:logrelproof}  and the artifacts that are
-specific to Coq and may not appear in other proof assistants.
-
-The powerset $[[PowerSet STm]]$ is encoded as the type \texttt{tm ->
-Prop}, a predicate over \lang{} terms.
-The inductive
-definition of the logical relation in Figure~\ref{fig:logrel} requires
-the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
-the function $[[F]]$ can be later instantiated into the logical
-relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
-
-In Coq, there is a distinction between computable functions and
-relations that can later be proven to be functional. The former can be
-viewed as a strict subset of the latter in axiom-free Coq. To be more
-precise, given a relation \texttt{R : A -> B -> Prop} subject to the
-totality and functionality constraints ($\forall$ \texttt{a} $\in$
-\texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
-that \texttt{R a b} is inhabited), we do not immediately obtain a function
-\texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
-\texttt{R a (F a)} is inhabited. However, the functional side
-conditions of a relation is clunky to express and tend to block
-automation. A simple workaround is to assume the axiom of unique choice, which
-is known to be consistent with Coq and allows us to induce a function \texttt{F
-: A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
-is functional. This approach would make our Coq development match
-the text version of our proof from Section~\ref{sec:logrelproof} more
-closely.
-
-However, we choose instead an axiom-free workaround and define
-\rref{I-Pi} as follows in our Coq mechanization.
-\begin{center}
-  \drule[]{I-PiCoq}
-\end{center}
-It should be easy to verify that the preconditions of
-\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
-equivalent. After establishing Lemma~\ref{lemma:logreldeter}, it is
-possible to further show that the conclusions of the rules are
-equivalent, too.\scw{If it is easy, you should have already done it.
-Unless it is really long and boring.}
-
-% This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
-% before we even show that our logical relation is
-% functional/deterministic, but does not affect the proof structure
-% otherwise.
-We choose to keep this discrepancy between the Coq development and the
-description of the logical relation presented in Section~\ref{sec:logreldep} since the
-skolemization process is more intuitively expressed in terms of
-function symbols rather than relation symbols. Otherwise, we do not
-see a clear advantage of \rref{I-PiCoq} over \rref{I-Pi} in set
-theory, where there is no distinction between computable functions and
-functions in general. \scw{not sure I follow this last bit}
-
-\subsection{Automation}
-\label{sec:automation}
-Our Coq mechanization heavily uses automation, though instead of
-defining custom tactics, we rely mostly on off-the-shelf tools such as
-Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.
-
-We use the Autosubst 2 framework to specify our syntax in HOAS and
-produce Coq syntax files in de Bruijn representation. Additionallly,
-Autosubst 2 provides a powerful tactic \texttt{asimpl} that
-can be used to prove the equivalence of two terms constructed using
-the primitive operators provided by the framework. This greatly
-simplifies the reasoning about substitution in our development as
-almost all substitution related properties are immediately discharged
-by \texttt{asimpl} without having to manually prove or even directly invoke any
-substitution related lemmas. \scw{You are only referring to lemmas about
-how substitution relates to syntax, right? You still need to manually
-invoke lemmas about substitution and judgements.}
-
-For other automation tasks that are not specific to binding, we use
-the powerful \texttt{sauto} tactic provided by CoqHammer to write
-short and declarative proofs. For example, here is a one-line proof of
-the triangle property about parallel reduction, from which the diamond
-property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
-\scw{you have space, so repeat the mathematical statement of the lemma}
-\begin{minted}{coq}
-Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
-Proof.
-  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
-Qed.
-\end{minted}
-In prose, the triangle property can be proven by induction over the
-graph of the \texttt{tstar} function, which stands for the Takahashi
-translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
-\texttt{ctrs:Par} say that the proof involves inverting and
-constructing of the derivations of parallel
-reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
-automation tactic to use
-the reflexivity and congruence properties of parallel
-reduction as lemmas.
-\scw{What does \texttt{lq:on} and \texttt{hauto} mean?}
-
-The automation here not only gives us a proof that is shorter and more
-resiliant to changes in definition, but also gives useful
-documentation for readers who wish to understand how the underlying
-proof works since the automation tactic is guaranteed to not use
-lemmas or invert derivations that are not specified in the
-\texttt{use} or \texttt{inv} flag.
-
-\section{Related Work and Discussion}
-\label{sec:relatedwork}
-
-\scw{Maybe it would be useful to include a chart here, so that readers can
-  easily keep track of the features of the various languages.  i.e. which ones
-  include large eliminations? type-directed equivalence? impredicative prop?
-  inductive datatypes? what are their line counts?}
-
-\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
-\citet{barendregt:lambda-calculi-with-types} are some of the earlier
-works that establish metatheoretic results strong enough to derive
-consistency for dependently typed systems.
-\scw{Should give Martin L\"of credit for first consistency proof.}
-\scw{What techniques do these three proofs use? How is it different from
-your proof? What about Luo, which
-introduced a universe hierarchy? What about categories with families, or
-other ways of showing consistency?}
-Some of these techniques
-are still in use in more recent works. For example, the technique for
-proving strong normalization by \citet{geuvers1994short} is adapted by
-\citet{moon2021graded} to show the same property to a dependently
-typed system extended with modalities.
-
-\citet{decagda} mechanizes in Agda the decidability of type
-conversion rule for a dependently typed language with one predicative
-universe level and typed judgmental equality with function
-$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
-uses a Kripke-style logical relation parameterized over an
-type-directed equivalence relation satisfying certain
-properties. Their logical relation is defined using the
-induction-recursion scheme, which is available in Agda but not in Coq.
-\citet{martin-lof-a-la-coq} manages to encode the logical relation
-from \citet{decagda} in the predicative fragment of Coq using a
-special way of encoding induction recursion. Their work further
-extends the decidability of type conversion result from~\citet{decagda} to the decidability
-type checking of a bidrectional type system.
-
-\citet{anand2014towards} mechanizes the metatheory of
-Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
-extensional type theory with features such as dependent functions,
-inductive types, and a full universe hierarchy. \citet{nbeincoq}
-mechanizes the normalization-by-evaluation algorithm in Coq for a
-dependently typed language with one predicative universe, similar to
-\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
-\citet{anand2014towards} and \citet{nbeincoq} leverage the
-impredicative \texttt{Prop} sort of Coq to define the interpretation
-of dependent function types and thus are closely related to our
-mechanization. However, instead of explaining the purely inductive
-logical relation as a convoluted workaround to the lack of
-induction-recursion in Coq, we give a self-contained explanation of
-our logical relation.
-
-Finally, it is worth pointing out that all systems we have discussed
-so far builds a relational model for their logical relation. Recall
-that our logical relation takes the form $[[InterpR i A S]]$ where
-$[[S]]$ is a set of terms. In a relational model, each type is
-interpreted as a partial equivalence relation over terms rather than a
-set. A relational model is necessary for an extensional type theory
-such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
-use a relational model either to justify the function $\eta$-law or to
-derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
-law for function is baked into the untyped reduction
-relation and therefore a relational model is not needed.
-\scw{would it be difficult to extend your proof to be a PER?}
-
-The dependently typed systems we have discussed so far also vary
-greatly in expressiveness. For example, the Calculus of
-Constructions~\citep{CoC} and MLTT without universe levels are unable
-to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
-identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
-the mechanized systems we have discussed so far, only
-\citet{anand2014towards} has a full universe hierarchy.
-Our object language \lang{}, while small, has a decent coverage of
-features commonly seen in dependently typed languages, including
-dependent pattern matching, dependent function types, a full universe
-hierarchy, an intensional identity type. We hope features such as
-type-directed reduction (necessary for unit $\eta$-law) and
-impredicative sorts can be implemented in a similarly streamlined
-fashion and we will leave those extensions as part of our future work.
-
-% To end this section, we
-\scw{should this be a new subsection?}
-The final question we want to address is: why is our proof,
-even with the extension from Section~\ref{sec:extension}, so much
-shorter than the proofs from \citet{decagda, nbeincoq,
-  martin-lof-a-la-coq}?
-First, unlike developments that mechanize the correctness of a clever
-type conversion algorithm, we only show the existence of
-normal form for open and closed terms and state our properties in
-terms of the untyped small-step reduction relation. This removes a lot
-of scaffolding in the specification of our type system and our logical
-relation.
-
-This contrasts our development with
-\citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
-difficult property to prove due to the complexity of the equality
-judgment used to define the logical relation. In our system, the
-irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
-direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
-lemma name} which is
-straightforward after we prove the results about parallel
-reduction in Section~\ref{sec:spec}. In other words, we keep the
-the logical relation itself minimal at the cost of extra lemmas that can be proven
-through syntactic means independently from the logical relation.
-For example, as commented near the end of Section~\ref{sec:extension},
-our normalization property (Corollary~\ref{corollary:exbetanf})
-% can induce a decision procedure for type conversion by reducing two
-% terms to $\beta\eta$-normal form and check for their syntactic
-% equivalence. This decision procedure would be inefficient compared to
-% NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
-% to weak head normal and checks for the equivalence of subterms
-% recursively
-does not immediately induce an efficient decision procedure, but it is
-possible to recover an efficient algorithm by reasoning about untyped
-lambda terms.
-
-% the concision of
-% our proof can be largely attributed to the fact that we bake into the
-% logical relation only the minimum amount of information that is the
-% most conducive to the proof of the metatheoretic result.
-
-Another key to our proof is the very early establishment of the confluence
-property of the paralell reduction relation
-(Lemma~\ref{lemma:pardiamond}). This property is used in
-Lemma~\ref{lemma:interppreservation} to show that the interpretation
-of a type is preserved under evaluation. That is, if $[[InterpR i A
-S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
-In a system with typed directed reduction, the confluence result is
-harder to establish since confluence requires subject reduction, which
-circularly depends on $\Pi$-injectivity, a consequence of
-confluence~\citep{siles2012pure}. There are different workarounds to
-this problem. \citet{siles2012pure} proposes a syntactic approach by
-defining a type system where the confluence result is
-directly provable and later show that the system of interest is
-equivalent to the system with the confluence property.
-\citet{decagda} uses a relational model and the relational counterpart
-of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
-model is parameterized by a typed indexed equivalence relation and
-is reused to prove derive different properties in their
-development. Another perhaps simpler approach is to simply
-extend the judgmental equality with rules about
-$\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
-reduction to be proven independently from confluence and
-$\Pi$-injectivity can later be shown to be admissible. We find the
-approach from \citep{weirich:systemd} the most lightweight as it
-allows us to derive confluence early on even in systems with typed
-conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}
-
-In cases where the judgmental equality does not leverage type information heavily
-(e.g. $\eta$-law for the unit type), it might be possible to show that
-$\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
-$[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
-property holds (we believe this is likely true for both
-\citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
-terms of untyped parallel reduction to show the decidability of type
-conversion. Otherwise, to fully model the system, we need to bake
-typing information into our logical relation and we need to redefine
-our logical relation in Kripke-style so we can relax the scoping
-constraint in order to prove adequacy, though a relational model would
-still be unnecessary as long as the $\eta$ laws are baked into the
-typed directed relation. \scw{not sure I understand this last bit}
-
-Since confluence plays such as key role in our proof, we are uncertain
-how to generalize our technique to systems where confluence does not
-hold, and we will investigate those scenarios in our future work.
-\scw{What systems are you thinking of? Or do you mean more generally
-that this technique only works for languages where the definition of
-equality is based on joinability of parallel reduction?}
-
-% We are able to avoid the complexivity of a Kripke-style logical relation
-% by defining our logical relation in terms of untyped parallel
-% reduction, which does not impose any scoping constraint on the
-% terms.
-
-% Why is it that the justification of $\eta$ law
-% requires a Kripke-style relational model in all three developments
-% but not ours? Are we ``cheating'' by defining \lang{}'s equational
-% theory using an untyped conversion rule? Instead of answering the
-% questions one by one, we clarify the design of our type system and our
-% proof technique so the answers to those questions are self-evident.
-
-% The key to the shortness of
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-% Type soundness can be proven through a syntactic
-% approach~\citep{syntacticsoundness} as a corollary of two properties:
-% progress and preservation. % The syntactic type soundness proof
-% % varies in complexity depending on the underlying type
-% % system. For example, a type system that tracks information flow would
-% % require additional structural rules related to security levels. In
-% % this paper, we focus on one specific type of complexity: the
-% In Figure~\ref{fig:stlcsoundness}, we summarize the structure of the
-% syntactic type soundness proof for the simply typed lambda
-% calculus. Each lemma can be proven by structural induction over the
-% typing derivation, while using the previous established results as
-% lemmas for specific cases that do not immediately follow from the
-% induction hypothesis. If we make our language more complex by adding
-% full dependent type support, the overall structure remains almost
-% identical.
-
-% NbE in Coq
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{refs}
-
-\appendix
-
-\subsection{More detailed proofs}
-\begin{lemma}[Preservation of the logical relation]
-  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
-\end{lemma}
-\begin{proof}
-  We carry out the proof by induction over the derivation of $[[Interp
-  I i A S]]$.
-  \begin{description}
-  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
-    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
-    be $[[Void]]$ and the result trivially follows.
-  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
-  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
-      refl , a <=> b  }]]$ and, by inverting the derivation of
-    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
-    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
-    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
-    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
-      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
-    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
-      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
-    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
-    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
-    <=> b0]]$. The result then immediately follows from the fact that
-    coherence is an equivalence relation.
-  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
-    such that:
-    \begin{itemize}
-    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
-        a #) }]]$
-    \item $[[Interp I i A S ]]$
-    \item $[[F in S -> PowerSet STm]]$
-    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
-    \end{itemize}
-    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
-    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
-      a #) }]]$. By \rref{I-Pi}, it suffices to show:
-    \begin{itemize}
-    \item $[[Interp I i A0 S ]]$
-    \item $[[F in S -> PowerSet STm]]$
-    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
-    \end{itemize}
-    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
-    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
-    conditions follow immediately from the induction hypothesis.
-  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
-    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
-    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
-    property of parallel
-    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
-    $[[C]]$ such that $[[B0 => C]]$ and $[[B => C]]$. By the induction
-    hypothesis, we derive $[[Interp I i C S]]$ from $[[Interp I i
-    B0 S]]$. By \rref{I-Red} and
-    $[[B => C]]$, we conclude that $[[Interp I i B S]]$.
-  \end{description}
-\end{proof}
-
 %%

### 2024-02-21 paper/paper.tex
Delete the appendix (for now)

@@
 
-\appendix
-
-\subsection{More detailed proofs}
-\begin{lemma}[Preservation of the logical relation]
-  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
-\end{lemma}
-\begin{proof}
-  We carry out the proof by induction over the derivation of $[[Interp
-  I i A S]]$.
-  \begin{description}
-  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
-    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
-    be $[[Void]]$ and the result trivially follows.
-  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
-  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
-      refl , a <=> b  }]]$ and, by inverting the derivation of
-    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
-    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
-    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
-    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
-      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
-    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
-      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
-    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
-    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
-    <=> b0]]$. The result then immediately follows from the fact that
-    coherence is an equivalence relation.
-  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
-    such that:
-    \begin{itemize}
-    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
-        a #) }]]$
-    \item $[[Interp I i A S ]]$
-    \item $[[F in S -> PowerSet STm]]$
-    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
-    \end{itemize}
-    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
-    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
-      a #) }]]$. By \rref{I-Pi}, it suffices to show:
-    \begin{itemize}
-    \item $[[Interp I i A0 S ]]$
-    \item $[[F in S -> PowerSet STm]]$
-    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
-    \end{itemize}
-    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
-    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
-    conditions follow immediately from the induction hypothesis.
-  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
-    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
-    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
-    property of parallel
-    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
-    $[[C]]$ such that $[[B0 => C]]$ and $[[B => C]]$. By the induction
-    hypothesis, we derive $[[Interp I i C S]]$ from $[[Interp I i
-    B0 S]]$. By \rref{I-Red} and
-    $[[B => C]]$, we conclude that $[[Interp I i B S]]$.
-  \end{description}
-\end{proof}
-
-\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
-  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
-\end{lemma}
-\begin{proof}
-  We start by inducting over the derivation of the first premise $[[Interp I i A
-  S0]]$.
-  \begin{description}
-  \item[\Rref{I-Void}:] We know that $[[Interp I i Void
-    emptyset]]$. Given $[[Interp I i Void S1]]$, our goal is to show that
-    $[[emptyset = S1]]$. This is immediate by applying the $[[Void]]$ case of
-    Lemma~\ref{lemma:interpinv} to $[[Interp I i Void S1]]$.
-  \item[\Rref{I-Bool, I-Eq, I-Set}:] Similar to the \rref{I-Void} case
-    by applying the matching case of Lemma~\ref{lemma:interpinv}
-    to $[[Interp I i A S1]]$.
-  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
-    such that:
-    \begin{itemize}
-    \item $[[Interp I i A S ]]$
-    \item $[[F in S -> PowerSet STm]]$
-    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
-    \end{itemize}
-    Our goal is to show that given $[[Interp I i Pi A B S1]]$, we have
-    $[[S1 = { b | forall a, (# a in S implies b a in F
-      a #) }]]$.
-    By the function case of Lemma~\ref{lemma:interpinv}, there exists
-    some $[[S0]]$ and $[[F0]]$ such that:
-    \begin{itemize}
-    \item $[[Interp I i A S0 ]]$
-    \item $[[F0 in S0 -> PowerSet STm]]$
-    \item
-      $[[forall a, (# a in S0 implies Interp I i B { a } F0 a #)]]$
-    \item
-      $[[S1 = { b | forall a, (# a in S0 implies b a in F0 a #) }]]$
-    \end{itemize}
-    It suffices to show that $[[S = S0]]$ and $[[F = F0]]$. The
-    equality $[[S = S0]]$ immediately follows from the induction
-    hypothesis since $[[Interp I i A S]]$ and $[[Interp I i A
-    S0]]$. Therefore, functions $[[F]]$ and $[[F0]]$ have the same
-    domain and codomain and thus it suffices to show that forall $[[a in S]]$,
-    $[[F a = F0 a]]$. Suppose $[[a in S]]$, we must have $[[Interp I
-    i B {a} F a]]$ and $[[Interp I i B {a} F0 a]]$ from the two
-    $\forall$-quantified statements above. The equality $[[F a = F0 a]]$ then
-    immediately follows from the induction hypothesis.
-  \item[\Rref{I-Red}:] There exists some $[[B]]$ such that $[[A =>
-    B]]$ and $[[Interp I i B S0]]$. Our goal is to show that given
-    $[[Interp I i A S1]]$, we have $[[S0 = S1]]$. By
-    Lemma~\ref{lemma:interppreservation}, from $[[Interp I i A S1]]$
-    and $[[A => B]]$, we have $[[Interp I i B S1]]$. From the
-    induction hypothesis, we can conclude that $[[S0 = S1]]$ since
-    $[[Interp I i B S0]]$ and $[[Interp I i B S1]]$.
-  \end{description}
-\end{proof}
-
 %%

### 2024-02-21 paper/paper.tex
Reorder the figures in Section 5

@@
 \label{sec:extension}
+In this section, we sketch out how the logical relation from
+Section~\ref{sec:logreldep} can be extended show the existence of
+$\beta\eta$ normal forms for (open and closed) well-typed terms. We
+first extend parallel reduction to include $\eta$ reduction for
+functions.
+\begin{center}
+  \drule[width=2.5in]{P-AbsEta}
+\end{center}
+
+\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
+equational theory since coherence, the untyped relation used for type
+conversion, is built on top of parallel reduction. We can recover the
+same confluence result about parallel reduction using the standard
+techniques from \citet{barendregt:lambda-calculi-with-types,
+takahashi-parallel-reduction}, though specifically for the de Bruijn
+representation, we need the following anti-renaming lemma about
+parallel reduction.
+\begin{lemma}[Par anti-renaming]
+  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
+exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
+\end{lemma}
+
 \begin{figure}[h]
@@
 \end{figure}
-\begin{figure}[h]
-  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
-  \caption{Extended logical relation}
-  \label{fig:logrelopen}
-\end{figure}
-
-\scw{Put this introduction before the figure in the section.}
-In this section, we sketch out how the logical relation from
-Section~\ref{sec:logreldep} can be extended show the existence of
-$\beta\eta$ normal forms for (open and closed) well-typed terms. We
-first extend parallel reduction to include $\eta$ reduction for
-functions.
-\begin{center}
-  \drule[width=2.5in]{P-AbsEta}
-\end{center}
-\scw{What happened to the domain type in the $\lambda$ expression? This is
-not your syntax...}
-\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
-equational theory since coherence, the untyped relation used for type
-conversion, is built on top of parallel reduction. We can recover the
-same confluence result about parallel reduction using the standard
-techniques from \citet{barendregt:lambda-calculi-with-types,
-takahashi-parallel-reduction}, though specifically for the de Bruijn
-representation, we need the following anti-renaming lemma about
-parallel reduction.
-\begin{lemma}[Par anti-renaming]
-  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
-exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
-\end{lemma}
 
@@
 \[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
+
+\begin{figure}[h]
+  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
+  \caption{Extended logical relation}
+  \label{fig:logrelopen}
+\end{figure}
 

### 2024-02-21 paper/paper.tex
some work on the introduction. not done yet

--- 2024-02-21 paper/paper.tex
@@
 \section{Introduction}
-In this paper, we present a short and mechanized consistency proof by
-logical relation for \lang{}, a Martin-Löf style dependent type theory
-with large elimination, a full predicative universe hierarchy, an
-intensional identity type and a boolean base type, both of which
-equipped with dependent elimination forms. % The mechanization is
-% extremely short and involves less than 1000 lines of manually written
-% Coq code.
-
-Our goal is to show the fundamental technique required to carry out proof by logical
-relation in a dependently typed setting where logical relation can no
-longer be defined as a simple recursive function over the structure of
-types. Compared to the many mechanized proofs by logical relation for
-predicative dependent type
-theory~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards},
-whose proof developments range from 20,000 to 400,000 lines of code,
-our Coq development is the most lightweight at less than 1,000 lines.
-
-The big difference in the size of the developments does not
-necessarily imply that our proof technique leads to a more concise
-proof due to the differing expressiveness of the languages and the
-different metatheoretic results being established. For
-example, \citet{anand2014towards} mechanizes the metatheory of
-Nuprl~\citep{constable1986implementing}, which, on top of all the
-features that \lang{} supports, includes W-types and partial types and
-requires a PER semantics to model its extensional typed
-equality. The object language from \citet{decagda} does not support
-identity types and only has one predicative universe, but additionally
-supports $\Sigma$ types. However, for a researcher who wishes to learn
-the underlying techniques for mechanized logical relation for
-dependent type theory, a small development like ours is much easier to
-navigate and understand. Furthermore, we show how our simple
-consistency proof can be easily extended to show the existence of
-$\beta\eta$-normal form for well-typed open \emph{and} closed terms,
-giving us a metatheoretic result almost as strong as the one from
-\citet{decagda} at the moderate cost of around 500 lines of extra code.
-
-We present our definitions and proofs in set theory notation and
-terminology rather than in Coq syntax to make the material accessible
-to readers with general mathematical background.
-However, notation aside, the typeset representation is
-purposefully made to be closer to our mechanization so less massaging
-of the definitions is required to encode our proofs in a proof
-assistant.
-The close correspondence between a typeset proof and a mechanized
-proof comes at the potential risk that the artifacts specific to
-theorem provers may become visible in the presented proof.
-However, for proofs by logical relation for
-dependent types, we find our approach advantageous for a few reasons.
-First, unlike properties that are derivable through
-syntactic means, proof by logical relation is demanding on the
-strength of the metatheory. An informal proof that attempts to be
-agnostic or ambiguous about the underlying metatheory % may struggle to even
+In this paper, we present a \emph{short}, \emph{mechanized} proof of logical
+consistency for \lang{}, a Martin-Löf style dependent type theory with large
+eliminations, a full predicative universe hierarchy, an intensional identity
+type and a boolean base type, both of which are equipped with dependent
+elimination forms.
+
+Our goal with this work is to demonstrate how the proof technique of
+\emph{logical relations} can be applied in dependently-typed setting.
+Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
+languages with simple or polymorphic types. In that context, logical relations
+can be define as simple recursive functions over the structure of types, or
+(in the case of recursive types) defined over the evaluation steps of the
+computation.
+
+Neither of these techniques make sense for predicative dependent type theory,
+so a young researcher might be excused for thinking that logical
+relations-based arguments are not available.  \scw{Well, actually, we did use
+  a step-indexed relation for Trellys\ldots} But this is not the case. Recent
+authors have developed tour-de-force mechanizations for the metatheory of
+modern proof
+assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
+have used logical-relations-based reasoning as part of their
+arguments. However, at the same time, these definitions have been overshadowed
+by their magnitude of the surrounding developments. Because these proofs show
+diverse results about real systems and algorithms, these developments range in
+size from 20,000 to 400,000 lines of code, and the techniques are not
+accessible to a casual reader.
+
+Our proof, designed to capture the essence of this proof technique in this
+setting, requires less than 1000 lines of code. We have achieved this
+significant reduction through a number of means: the careful selection of the
+features that we include in the object type type system and the results that
+we prove about it, in addition to the judicious use of automation.  Our
+language is small, but includes enough to be illustrative. For example, we
+eschew inductive datatypes or W-types, the but we do include propositionality
+equality and booleans in order to capture the challenges presented by indexed
+types and dependent pattern matching. We do not show the decidability of type
+checking, nor do we develop a PER semantics, but we do demonstrate how our
+consistency proof can be extended (at a moderate cost of 500 lines of code) to
+show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
+closed terms. We include a full predicative universe hiercharchy and
+type-level computation, to demonstrate the logical strength of the
+approach. \scw{Say that we don't have impredicativity, but neither does anyone
+  else?}
+%%
+%% SCW: save specific comparisons with other systems for later
+%%
+% The big difference in the size of the developments does not
+% necessarily imply that our proof technique leads to a more concise
+% proof due to the differing expressiveness of the languages and the
+% different metatheoretic results being established. For
+% example, \citet{anand2014towards} mechanizes the metatheory of
+% Nuprl~\citep{constable1986implementing}, which, on top of all the
+% features that \lang{} supports, includes W-types and partial types and
+% requires a PER semantics to model its extensional typed
+% equality. The object language from \citet{decagda} does not support
+% identity types and only has one predicative universe, but additionally
+% supports $\Sigma$ types. However, for a researcher who wishes to learn
+% the underlying techniques for mechanized logical relation for
+% dependent type theory, a small development like ours is much easier to
+% navigate and understand. Furthermore, we show how our simple
+% consistency proof can be easily extended to show the existence of
+% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
+% giving us a metatheoretic result almost as strong as the one from
+% \citet{decagda} at the moderate cost of around 500 lines of extra code.
+
+The result of our work is a development that an interested researcher can
+navigate and understand. We accompany this mechanized proof with an informal
+description, presented here using set theory notation and terminology so that
+the material is accessible to readers with general mathematical background,
+That said, our explanations do not stray too far away from our proof scripts.
+We label each lemma directly to their counterpart in the proof script,
+anticipating that readers may wish to reference and validate our formalism.
+Our typeset representation are purposefully made to directly follow our
+mechanical proofs, while avoiding, as much as possible, the introduction 
+of artifacts that are specific to certain theorem prover.
+
+Not only does this connection aid readers that wish to, like us, adopt proof
+assistants for their day-to-day use. We also find that this precision is
+important for conveying the proof techique itself. Unlike properties that are
+derivable through syntactic means, proofs by logical relation make demands on
+the strength of the logic in which they are expressed. An informal proof that
+attempts to be agnostic or ambiguous about the underlying
+metatheory % may struggle to even
 % convey the validity of their definition and
@@
 effort from the reader to even understand whether it is theoretically
-possible to encode such proofs in a proof assistant.
-Second, while a proof assistant may reject perfectly valid
-definitions, it helps us keep our definition precise and
-unambiguous. A precise definition is crucial as the reasoning
-principles we are allowed to use is directly tied to the definition,
-which is sometimes obscured by concise notations that diverge too far
-from a formal proof.
+possible to encode such proofs in a given ambient logic, such as those found 
+in modern proof assistants.
+
+% The close correspondence between a typeset proof and a mechanized
+% proof comes at the potential risk that the artifacts specific to
+% theorem provers may become visible in the presented proof.
+%However, for proofs by logical relation for
+%dependent types, we find our approach advantageous for a few reasons.
+Furthermore, while a proof assistant may reject perfectly valid definitions,
+it helps us keep our definition precise and unambiguous. A precise definition
+is crucial as the reasoning principles we are allowed to use is directly tied
+to the definition, which is sometimes obscured by concise notations that
+diverge too far from a formal proof.
+
 Finally, our approach does not result in a verbose proof on paper, thanks to
-the already short mechanization. On top of that, We make conservative
-simplifications to our typeset proof such as avoiding the clunky
-syntax for well-founded recursion in Coq, but we keep the overall
-structure of our typeset proofs consistent with our mechanization.
-The consistency enables us to label at the footnote each lemma
-directly to their counterpart in the proof script for the readers to
-reference and validate.
+the already short mechanization. While we do make conservative simplifications
+to our typeset proof, such as avoiding the clunky syntax for well-founded
+recursion in Coq, we keep the overall structure of our typeset proofs
+consistent with our mechanization.
 
@@
 
-The complexity of applying proof by logical relation to dependent types stem
-from the fact that the logical relation is much harder to define. In
-simply typed languages, the logical relation is defined as a recursive
-function over the type $[[A]]$. In dependent types, the type
-$[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
-this type, we need to first reduce it to $[[Bool]]$. However, we
-cannot write a function that performs the reduction because we do not
-know the termination of well-typed terms a priori. As a result, we
-define the logical relation as an inductively defined relation,
-reminiscent of how we specify the reduction graph of a partial
-function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.
-
 \label{sec:logreldep}
@@
 looking. Need to explain.  }
+
 The metavariables $[[I]]$ and $[[S]]$ are
@@
 b]]$ are typically required for indexed types, of which equality types
-are an instance. \Rref{I-Par} enables us to reduce our types in order
-to assign meanings. Recall the type expression $[[(\ var 0)
-Bool]]$. \Rref{I-Par} says to know that
-$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
-show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
-Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
-  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Par}
-and \rref{I-Bool}.
-
-\Rref{I-Pi} is the most interesting. To explain what it means, we
-need to perform a few steps of transformations to its precondition.
-The precondition consists of a
-mysterious relation $[[R]]$ over the set $[[S]]$, an interpretation
-of the type $[[A]]$, and returns a subset of terms. $[[R]]$ is further
-subject to the following two constraints:
-\begin{itemize}
-\item $[[forall a, (# exists S0  , ( a , S0 ) in R #)]]$
-\item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
-\end{itemize}
-The first condition allows us to induce a function $[[F in S ->
-PowerSet STm]]$ such that $[[(a , F a) in R]]$. Then we can easily
-show that $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$.
-
-For the other direction, given a function $[[F]]$ satisfying the
-property $[[forall a, (# a in S implies Interp I i B { a } F a #) ]]$, it
-is possible to go in the other direction and define a relation $[[R]]=[[
-{ ( a , F a )  | a in S }]]$. It is easy to verify that $[[R]]$ satisfy
-the the constraint required in the precondition of \rref{I-Pi}.
-
-Furthermore, the skolemization process allows us to conclude the
-following equivalence:
-\[ \exists F, [[forall a, (# a in S implies Interp I i B { a } F a #)]] \iff [[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]  \]
-% Ignoring the content below the horizontal bar of
-% the derivation, we claim the statements from the precondition are equivalent to
-% the conjunction of the following statements:
-% \begin{itemize}
-% \item $[[Interp I i A S]]$
-% \item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
-%   [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
-% \end{itemize}
-% In general, let $R$ be a binary relation over the sets $[[A]]$ and
-% $[[B]]$, it easy to verify the following equivalence, which is used to
-% justify the skolemization process~\citep{skolemization}.
-% \scw{Is this true constructively? If not, is that ok?}
-% \[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
-%   \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
-%   \in R\]
-% The left-hand side is equivalent to the right-hand side, but has no
-% mentioning of any functions.
-The precondition of \rref{I-Pi} therefore can be formulated as the
-following statements:
+are an instance.
+
+\Rref{I-Pi} is the most interesting. The precondition consists of a
+mysterious function $[[F]]$ that takes an element from the set $[[S]]$, the interpretation
+of the type $[[A]]$, and returns a subset of terms. $[[F]]$ is further
+subject to the constraint that $[[forall a, (# a in S implies Interp I
+i B { a } F a #)]]$. Ignoring the content below the horizontal bar of
+the derivation, the statements from the precondition are equivalent to
+the conjunction of the following statements:
 \begin{itemize}
 \item $[[Interp I i A S]]$
-\item $[[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]$
+\item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
+  [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
 \end{itemize}
-From this alternative formulation, we can more easily tell what is
-happening in \rref{I-Pi}: the
-function type $[[Pi A B]]$ has an interpretation if its input
-type $[[A]]$ can be
+In general, let $R$ be a binary relation over the sets $[[A]]$ and
+$[[B]]$, it easy to verify the following equivalence, which is used to
+justify the skolemization process~\citep{skolemization}.
+\scw{Is this true constructively? If not, is that ok?}
+\[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
+  \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
+  \in R\]
+The left-hand side is equivalent to the right-hand side, but has no
+mentioning of any functions.
+In the context of \rref{I-Pi}, through
+the equivalence, we can more easily tell what's happening: the
+function type $[[Pi A B]]$ has a semantic interpretation if its input
+type $[[A]]$ can be semantically
 interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
 type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
-$[[B]]$, has some semantic interpretation.
-
-If we look at $[[Interp I i Pi A B { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$, the conclusion of \rref{I-Pi}, we see why the
-presentation with an explicit $[[R in S * PowerSet STm]]$ is
+$[[B]]$, has a semantic interpretation.
+
+If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
+  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
+presentation with an explicit $[[F in S -> PowerSet STm]]$ is
 useful. We want to state the fact that $[[B {a}]]$ has a semantic
@@
 set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
-to. With the relation $[[R]]$ in scope, we can refer to the
-witness sets $[[S0]]$ such that $[[Interp I i B {a} S0]]$ for $[[a in S]]$.
-With the alternative representation that is free of the relation $[[R]]$, we
-might want to reformulate our logical relation in the following more
-concise but equivalent form, whose admissibility follows directly from
-\rref{I-Pi}.
+to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
+witness in the precondition for $[[a in S]]$.
+With the alternative representation that is free of the function $[[F]]$, we
+might want to reformulate our logical relation in the following way.
 \begin{center}
@@
 \end{center}
-\begin{lemma}
-  \label{lemma:piintroalt}
-  \Rref{I-PiAlt} is admissible.
-\end{lemma}
-\begin{proof}
-Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
-, S0 ) | a in S , Interp I i B { a } S0 }]]$.
-\end{proof}
-
 Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
@@
 functionality and inversion properties, we opt for our current
-formulation of \rref{I-Pi} so we immediately obtain a
+skolemized formulation of \rref{I-Pi} so we immediately obtain a
 well-defined inductive relation and a usable induction principle. The
 slight disadvantage of \rref{I-Pi} is that we need to construct the
-function $[[R]]$ each time we apply it, though this is mitigated by
-the immediate admissibility of \rref{I-PiAlt}.
+function $[[F]]$ each time we apply it. Therefore, in
+Lemma~\ref{lemma:piintroalt}, we show the admissibility of
+\rref{I-PiAlt} a posteriori and use that as a more convenient
+introduction rule for the rest of our proofs.
 
@@
   \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
-  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S1]]$, then there exists $[[S]],[[R]]$ such that:
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
     \begin{itemize}
     \item $[[Interp I i A S ]]$
-    \item $[[R in S * PowerSet STm]]$
-    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
-    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
     \end{itemize}
@@
   \item[\Rref{I-Pi}:] Immediate.
-  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S1]]$.
+  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
     We know that there exists some $[[A0]]$ and
     $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
-    A0 B0 S1]]$. From the
-    induction hypothesis, there exists $[[S]]$ and $[[R]]$ such that :
+    A0 B0 S0]]$. From the
+    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
     \begin{itemize}
     \item $[[Interp I i A0 S ]]$
-    \item $[[R in S * PowerSet STm]]$
-    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B0 { a } S0 #)]]$
-    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
+    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
     \end{itemize}
@@
     $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
-    $[[R]]$ additionally satisfies the following properties:
+    $[[F]]$ satisfies the following properties:
     \begin{itemize}
     \item $[[Interp I i A S ]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
     \end{itemize}
-    These properties are exactly what we need to finish the proof.
+    These properties are exactly the ones from the conclusion.
   \end{description}
@@
   \label{lemma:interppreservation}
-  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
-\end{lemma}
-\begin{proof}
-  We carry out the proof by induction over the derivation of $[[Interp
-  I i A S]]$.
-
-  The only interesting case is \Rref{I-Red}. Given that
-  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
-  for all  $[[B]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
-  S]]$. By the diamond property of parallel reduction
-  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
-  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
-  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
-  B0 S]]$. By \rref{I-Par} and $[[B1 => B]]$, we conclude that
-  $[[Interp I i B S]]$.
-
-  The remaining cases all fall from induction hypotheses and basic
-  properties about convertibility and parallel reduction we have
-  established in Section~\ref{sec:spec}.
-\end{proof}
-From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
-derive the following corollary that two convertible types can always interpret
-into the same set. We adopt the terminology from \citet{martin-lof-a-la-coq}
-and refer to this property as irrelevance.
-\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
-  \label{lemma:logrelcoherence}
-  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
-\end{corollary}
-% \scw{Need a transition here}
-
-Since the definition of our logical relation is an inductive relation,
-it is not immediately obvious why each type $[[A]]$ can only uniquely
-corresponds to one set $[[S]]$. The following lemma shows that our
-logical relation is indeed functional.
-\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
-  \label{lemma:logreldeter}
-  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
-\end{lemma}
-\begin{proof}
-  The proof proceeds by induction over the derivation of the first
-  premise $[[Interp I i A S0]]$.
-  All cases that are not \rref{I-Red} follow immediately from the
-  Lemma~\ref{lemma:interpinv}.
-
-  For \rref{I-Red}, we are given that there exists some $[[B]]$ such
-  that $[[A => B]]$ and $[[Interp I i B S0]]$. Our goal is to show
-  that given $[[Interp I i A S1]]$ for some $[[S1]]$, we have $[[S0 =
-  S1]]$. By the preservation property
-  (Lemma~\ref{lemma:interppreservation}),
-  we know that $[[Interp I i B S1]]$ since $[[A => B]]$. The statement
-  $[[S0 = S1]]$ then immediately follows from the induction hypothesis.
-\end{proof}
-
-Lemma~\ref{lemma:logreldeter} enables us to show the following
-improved inversion lemma for function types whose statement is free of
-the relation $[[R]]$, analogous to the admissible rule \rref{I-PiAlt}.
-\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
-  \label{lemma:piinvalt}
-  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
-  such that the following constraints hold:
-  \begin{itemize}
-  \item $[[Interp I i A S0]]$
-  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
-    S1 #) #)]]$
-  \item $[[S = { b | forall a, (# a in S0 , forall
-    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
-\end{proof}
-
-The next lemma shows that our logical relation satisfies
-cumulativity. That is, if a type has an interpretation at a lower
-universe level, then we can obtain the same interpretation at a higher
-universe level.
-\begin{lemma}[Logical relation cumulativity\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
-  \label{lemma:logrelcumulativity}
-  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
-\end{lemma}
-\begin{proof}
-  Trivial by structural induction over the derivation of $[[Interp I
-  i0 A S]]$.
-\end{proof}
-Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
-implicitly assume that $[[I]]$ is defined on the set of natural
-numbers less than $[[i1]]$.
-
-\begin{corollary}[Logical relation is functional with different levels\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
-  \label{lemma:logreldeterhet}
-  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
-\end{corollary}
-\begin{proof}
-  Immediate from Lemma~\ref{lemma:logreldeter} and
-  \ref{lemma:logrelcumulativity}.
-\end{proof}
-
-We say that a set of terms $[[S]]$ is closed under expansion if given
-$[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
-The final property we want to show is that the output set $[[S]]$ from
-the logical relation is closed under expansion. Unlike the previous
-lemmas, we directly state the lemma
-in terms of $[[InterpR i A S]]$ rather than $[[Interp I i A S]]$
-because we need to know something about $[[I]]$ for this property to
-hold in the \rref{I-Set} case.
-
-\yl{Can lift this lemma somewhere earlier in the section since it doesn't really
-  depend on anything}
-\begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpExt\_back\_clos}}]
-  \label{lemma:logrelbackclos}
-  If $[[InterpR i A S]]$, then the set $[[S]]$ is closed under expansion.
-\end{lemma}
-\begin{proof}
-  By the definition of $[[InterpR i A S]]$ from
-  Figure~\ref{fig:logrelrec}, we unfold $[[InterpR i A S]]$ by one
-  step into $[[ Interp I i A S  ]]$ where $[[I i]] := [[{A | exists S
-    , InterpR i A S}]]$.
-  We then proceed by induction over the derivation of $[[Interp I i A S]]$.
-
-  All cases are trivial except for the \rref{I-Set} case, where we
-  want to show that the set $[[I j]]$ is closed under expansion for
-  all $[[j < i]]$. However, by the definition of $[[I]]$, we know that
-  $[[A in I j]]$ if only if there exists some $[[S]]$ such that
-  $[[Interp I i A S]]$. By \rref{I-Par}, we must also have $[[B in I
-  j]]$ for all $[[B => A]]$.
-\end{proof}
-
-\section{Semantic Typing and Consistency}
-\label{sec:logrelproof}
-\begin{figure}[h]
-\[
-\begin{array}{lcl}
-
-      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
-                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
-      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
-                       G]]\text{ then there exists some } [[j]] \text{
-                       and } [[S]] \text{ such that } [[InterpR j A
-                       {rho} S]] \text{ and } [[a {rho} in S]] \\
-      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
-\end{array}
-\]
-  \caption{Semantic Typing for \lang}
-  \label{fig:semtyping}
-\end{figure}
-
-The logical relation we define in Figure~\ref{fig:logrel} does not
-include cases for variables. Likewise, for the base types such as
-boolean and equality, the output set $[[S]]$ contains only terms that
-evaluate to closed terms.
-
-\scw{What is the motivation for this section? Why do we want to generalize
-the logical relation?}
-To generalize our logical relation to open
-terms, we define the semantic typing judgment by closing the open
-terms with a substitution whose codomain consists of terms that
-respect the interpretation of the types from the context. The full
-definitions of well-formed substitution ($[[rho |= G]]$), semantic
-typing ($[[ G |= a : A]]$), and semantic context well-formedness
-($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
-\scw{explain all parts of this figure. it's pretty intimidating.}
-In the
-definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
-the typing context obtained by dropping the last $[[i]]$ elements of
-$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
-returns the empty list.
-
-The following lemma makes the statement $[[G |= A : Set i]]$ easier to
-work with.
-\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}\scw{spell out the whole word}]
-  \label{lemma:setinv}
-  The following two statements are equivalent:
-  \begin{itemize}
-  \item $[[G |= A : Set i]]$
-  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
-    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  The forward direction is immediate by
-  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
-  and show that $[[G |= A : Set i]]$ given the second bullet.
-
-  Suppose $[[rho |= G]]$, then we know that there exists some $[[S]]$
-  such that $[[InterpR i (A {rho}) S]]$. By the definition of semantic
-  typing, it suffices to show that there exists some $[[j]]$ and
-  $[[S0]]$ such that  $[[InterpR j Set i S0]]$ and $[[A {rho} in
-  S0]]$.
-  Pick $[[Suc i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
-  ]]$ for $[[S0]]$ and it is trivial to verify the conditions hold.
-\end{proof}
-
-The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
-counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
-than inductively over the context. \scw{why is that the case?} It is easy to recover the same
-structural rules:
-\begin{lemma}[Semantic context well-formedness cons]
-  \label{lemma:semwffcons}
-  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
-\end{lemma}
-\scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
-Ctx-empty?}
-\begin{proof}
-  By the definition of semantic context well-formedness, the goal is
-  to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i
-  (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
-  proven by case analysis on whether $[[i]]$ is zero.
-\end{proof}
-
-Likewise, the semantic well-formendness judgment for substitutions
-satisfies similar structural rules.
-\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
-  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
-  $[[rho .: a |= G ++ A]]$.
-\end{lemma}
-\begin{proof}
-  Start by unfolding the definition of $[[rho .: a |= G ++ A]]$ and
-  performing a case analysis similar to the proof of
-  Lemma~\ref{lemma:semwffcons}.  The case where the number is $[[0]]$
-  requires Lemma~\ref{lemma:logreldeterhet} to finish the proof.
-\end{proof}
-
-Next, we show some non-trivial cases of the fundamental theorem as
-top-level lemmas and leave the remaining cases as exercises for the
-reader.\scw{I dislike leaving ``exercises to the reader''}
-
-First, we formulate the definition of valid renamings and prove that
-semantic typing satisfies renaming so we can weaken the context when
-reasoning about the variable case of the fundamental lemma
-(Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G
-++ D]]$, it is easy to show that we can extract some valuation
-$[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
-``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :
-A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
-$[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
-after weakening the context; this implication holds because $[[rho |=
-G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
-can make use of the premise $[[G |= a : A]]$ to derive what we need
-for the conclusion. We recommend the readers to skip the proofs of
-Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
-during the first read as long as they have an intuitive understanding
-of what the renaming property is meant to capture.
-
-\scw{Make this a definition? Would it be useful to create notation
-  for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
-We say that $[[xi]]$ is
-valid from the context $[[G]]$ to the context $[[D]]$ if the
-following condition holds.
-\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
-  [[D ren xi
-  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]
-
-\begin{lemma}[Truncate is valid]
-  \label{lemma:validtruncate}
-  If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i
-  G]]$ to $[[G]]$.
-\end{lemma}
-\begin{proof}
-  By the definition of a valid renaming, we must show that given $[[j
-  < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
-  \begin{itemize}
-  \item $[[ren up i j]] = i + j  < [[| G | ]]$
-  \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
-  \end{itemize}
-  The first bullet point is immediate from the fact that $[[j]] <
-  [[|G|]] - [[i]]$. The second bullet follows from unfolding the
-  definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
-  $[[i]] + [[j]] < [[|G|]]$.
-\end{proof}
-
-\begin{lemma}[Renaming for $[[rho |= G]]$]
-  \label{lemma:renamingval}
-  If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
-  $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
-\scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
-\end{lemma}
-\begin{proof}
-  Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
-  that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i
-  < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.
-
-  By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
-have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho
-} ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |=
-D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
-S]]$ and $[[rho ren xi i in S]]$.
-\end{proof}
-
-\begin{lemma}[Renaming for $[[G |= a : A]]$]
-  \label{lemma:semrenaming}
-  If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
-  $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
-\end{lemma}
-\begin{proof}
-  Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
-\end{proof}
-
-\scw{Need a transition here that you are starting to explain the semantic typing
-rules.}
-
-\begin{lemma}[ST-Var]
-  \label{lemma:stvar}
-  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
-\end{lemma}
-\begin{proof}
-  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
-need to show that there exists some $[[j]]$ and $[[S]]$ such that
-  \begin{itemize}
-  \item $[[InterpR j G i < up Suc i > { rho } S]]$
-  \item $[[rho i in S]]$
-  \end{itemize}
-  From the definition of $[[|= G]]$, we know that there exists some
-  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. By
-  Lemma~\ref{lemma:validtruncate}, we know that $[[up Suc i]]$ is a
-  valid renaming from $[[drop Suc i G]]$ to $[[G]]$. By
-  Lemma~\ref{lemma:semrenaming}, we deduce $[[G |= G i < up Suc
-  i > : Set j]]$. By the identity from Lemma~\ref{lemma:setinv}, we
-  know that $[[InterpR j G i < up Suc i > { rho } S]]$. It now
-  suffices to show $[[rho i in S]]$, but that is immediate from the
-  definition of $[[rho |= G]]$.
-\end{proof}
-
-\begin{lemma}[ST-Set]
-  \label{lemma:stset}
-  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
-\end{lemma}
-\begin{proof}
-  Immediate by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
-\end{proof}
-
-\begin{lemma}[ST-Pi]
-  \label{lemma:stpi}
-  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
-  A B : Set i]]$.
-\end{lemma}
-\begin{proof}
-  Applying Lemma~\ref{lemma:setinv} to the
-  conclusion, it now suffices to show that given $[[rho |= G]]$, there
-  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
-  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
-  there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
-From $[[G ++ A |= B : Set i]]$, we know that there must
-exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
-in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
-\end{proof}
-
-\begin{lemma}[ST-Abs]
-  \label{lemma:stabs}
-  If $[[G |= Pi A B : Set i]]$ and $[[G ++ A |= b : B]]$, then $[[G |=
-  \ A b : Pi A B]]$.
-\end{lemma}
-\begin{proof}
-  By unfolding the definition of $[[G |= \ A b : Pi A B]]$, we need to
-  show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
-  $[[S]]$ such that $[[InterpR i Pi A {rho} B {up rho} S]]$ and $[[\
-  A{rho} b{up rho} in S]]$.
-
-  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
-  i]]$, there exists some set $[[S]]$ such that
-  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
-  $[[\A{rho} b{up rho} in S
-  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
-  that all following conditions hold:
-  \begin{itemize}
-  \item $[[InterpR i A{rho} S0]]$
-  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
-    {rho .: a}
-    S1 #) #)]]$
-  \item $[[S = { b | forall a, (# a in S0 , forall
-      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
-  \end{itemize}
-  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
-  that given $[[a in S0]]$,
-  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
-  a in S1]]$.
-  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
-  under expansion. By Lemma~\ref{lemma:parabscbn}, since $[[( \A{rho} b{up rho} )
-  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
-  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
-  B]]$ and the fact that the logical relation is deterministic and
-  cumulative (Lemma~\ref{lemma:logreldeterhet}).
-\end{proof}
-
-\begin{lemma}[ST-App]
-  \label{lemma:stapp}
-  If $[[G |= b : Pi A B]]$ and $[[G |= a : A]]$, then $[[G |= b a : B {a}]]$.
-\end{lemma}
-\begin{proof}
-Suppose $[[rho |= G]]$. The goal is to show that there exists some
-$[[i]]$ and $[[S1]]$
-such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a} {rho}
-S1]]$, or equivalently, $[[InterpR i B {rho .: a {rho}} S1]]$ since
-$[[B {a}{rho}]] = [[B {rho .: a {rho}}]]$. By the premise $[[G |= b :
-Pi A B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
-there exists some $[[i]]$ and $[[S0]]$ such that:
-  \begin{itemize}
-  \item $[[InterpR i A{rho} S0]]$
-  \item $[[forall a0, (# a0 in S0 implies (# exists S1 , InterpR i B
-    {rho .: a0}
-    S1 #) #)]]$
-  \item $[[forall a0, (# a0 in S0 , forall
-      S1, (# InterpR i B {rho .: a0} S1,  b {rho} a0 in S1 #) #)]]$
-  \end{itemize}
-  Instantiating the variable $[[a0]]$ from the last two bullets with
-  the term $[[a {rho}]]$, the conclusion immediately follows.
-\end{proof}
-
-\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
-  \label{theorem:soundness}\leavevmode
-  \begin{itemize}
-  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
-  \item If $[[|- G]]$, then $[[|= G]]$.
-  \end{itemize}
-\end{theorem}
-\begin{proof}
-  Proof by mutual induction over the derivation of $[[G |- a :
-  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
-  from Lemma~\ref{lemma:semwffcons}.
-Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
-  can be used to discharge their syntactic counterpart
-  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
-  cases not covered by the lemmas are similar to the ones already
-  shown or simpler and therefore omitted from the text.
-
-\end{proof}
-
-\scw{Spell this out in a little more detail.}
-
-\begin{proof}
-  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
-\end{proof}
-
-\section{Toward Decidability of Type Conversion and $\eta$ laws}
-\label{sec:extension}
-\begin{figure}[h]
-  \[
-    \begin{array}{lcl}
-      % \beta\text{-}\mathit{neutral\ terms}\\
-      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
-                    f]] \\ \\
-      % \beta\text{-}\mathit{normal\ terms} \\
-      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
-                    |\ [[f ~ f : f]]\\
-            & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
-    \end{array}
-  \]
-  \caption{$\beta$-neutral and normal forms}
-  \label{fig:nenf}
-\end{figure}
-\begin{figure}[h]
-  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
-  \caption{Extended logical relation}
-  \label{fig:logrelopen}
-\end{figure}
-
-\scw{Put this introduction before the figure in the section.}
-In this section, we sketch out how the logical relation from
-Section~\ref{sec:logreldep} can be extended show the existence of
-$\beta\eta$ normal forms for (open and closed) well-typed terms. We
-first extend parallel reduction to include $\eta$ reduction for
-functions.
-\begin{center}
-  \drule[width=2.5in]{P-AbsEta}
-\end{center}
-\scw{What happened to the domain type in the $\lambda$ expression? This is
-not your syntax...}
-\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
-equational theory since coherence, the untyped relation used for type
-conversion, is built on top of parallel reduction. We can recover the
-same confluence result about parallel reduction using the standard
-techniques from \citet{barendregt:lambda-calculi-with-types,
-takahashi-parallel-reduction}, though specifically for the de Bruijn
-representation, we need the following anti-renaming lemma about
-parallel reduction.
-\begin{lemma}[Par anti-renaming]
-  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
-exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
-\end{lemma}
-
-The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the neutral terms
-and normal forms with respect to $\beta$-reduction, but can still take
-$\eta$ reduction steps. We sometimes use the judgment forms $[[ne a]]$
-and $[[nf a]]$ to indicate that there exists $[[e]]$ or $[[f]]$ such
-that $[[a = e]]$ or $[[a = f]]$.
-
-We can show that parallel reduction preserves $\beta$-normal and
-neutral forms.
-\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
-  \label{lemma:parnenf}
-  If $[[a => b]]$, then
-  \begin{itemize}
-  \item $[[ne a]]$ implies $[[ne b]]$
-  \item $[[nf a]]$ implies $[[nf b]]$
-  \end{itemize}
-\end{lemma}
-This lemma is mainly to show that \rref{P-AbsEta} does not introduce
-any new $\beta$ redex because $\eta$ reduction is the only real step
-we can take.
-
-The predicates $[[wne a]]$ and $[[wn
-a]]$ describe terms that can evaluate into $\beta$-neutral or
-$\beta$-normal form through parallel reduction and are defined as
-follows.\scw{What does $wne$ stand for?}
-\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
-\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
-
-The updated logical relation and its auxiliary definitions are shown
-in Figure~\ref{fig:logrelopen}.
-
-% The idea of a term blocked from $\beta$ reduction is
-% captured by $[[e]]$, the set of neutral terms. We augment the
-% interpretation of booleans and the empty types with terms that
-% evaluate to neutral terms. Furthermore, types that are neutral terms
-% can also be assigned a meaning since they may be inhabited by other
-% neutral terms.
-
-In the definition of the logical relation, we omit the rules for
-the function and universe cases since they remain identical to the
-original version in Figure~\ref{fig:logrel}.
-The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
-An open term of type $[[Bool]]$ does
-not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
-a variable, or more generally, a neutral term. Likewise, the
-$[[Void]]$ type, while remains uninhabited under an empty context, may
-be inherited by the set of neutral terms when there is a variable in
-the context that allows us to inhabit $[[Void]]$.
-
-The rule for equality type $[[a ~ b : A]]$ is augmented with the preconditions that
-$[[a]]$, $[[b]]$, and $[[A]]$ are all in their normal forms since
-otherwise our model would include equality types that are themselves
-not normalizing. Furthermore, the side condition $[[a <=> b]]$ is only
-available when the equality proof reduces to $[[refl]]$. If the proof
-term reduces to a neutral term, then there is nothing we can learn
-about the relationship between $[[a]]$ and $[[b]]$.
-
-Finally, in a
-non-empty context, a type itself may evaluate to a neutral term and in
-turn can only inhabited by neutral terms, thus the addition of \rref{I-Ne}.
-
-Interestingly, all the properties we have shown in
-Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the
-fundamental lemma can be proven in the exact same order, where the new
-cases due to \rref{I-Ne} and the augmentation of neutral terms
-to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
-Lemma~\ref{lemma:parnenf}.
-
-Before we can prove the fundamental theorem and derive the
-normalization property as its corollary, we need to additionally
-formulate and prove the adequacy property.
-
-Let us first start with some auxiliary lemmas and definitions.
-\begin{lemma}[Ext Wn]
-  \label{lemma:extwn}
-  If $[[wn a var i]]$, then $[[wn a]]$.
-\end{lemma}
-\begin{proof}
-  By induction over the length of the reduction sequence in $[[wn a
-  var i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
-  \ref{lemma:parnenf}.
-\end{proof}
-
-\begin{lemma}[wne wn]
-  \label{lemma:wnewn}
-  If $[[wne a]]$ and $[[wn b]]$, then $[[wne a b]]$.
-\end{lemma}
-\begin{proof}
-  Immediate by lexicographical induction over the length of the reduction sequences in
-  $[[wne a]]$ and $[[wn b]]$.
-\end{proof}
-
-\scw{What does CR stand for?}
-\begin{definition}[CR]
-  Let $[[S]]$ be a set of lambda terms. We say that
-  \begin{itemize}
-  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
-  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
-  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
-  \end{itemize}
-\end{definition}
-
-\begin{lemma}[$CR_1$ for type]
-  \label{lemma:cr1ty}
-  If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
-\end{lemma}
-\begin{proof}
-  Immediately from \rref{I-Ne} and \rref{I-Par}.
-\end{proof}
-
-\begin{lemma}[$CR$ for interpreted sets]
-  \label{lemma:crel}
-  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
-\end{lemma}
-\begin{proof}
-  By induction over the derivation of $[[Interp I i A S]]$. The base
-  cases are all immediate.
-
-  In the function case ($[[Pi A B]]$), we use the induction
-  hypothesis to show that variables, which are special cases of
-  neutral terms, semantically inhabit the input type $[[A]]$. We apply the
-  function to an arbitrary variable and we can then use
-  Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.
-
-  To show
-  the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
-  such that $[[wne b]]$,
-  and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
-  us a term that semantically inhabits $[[B {a}]]$. From the induction
-  hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
-  holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
-  from the induction hypothesis, since every term that evaluates to
-  some neutral form semantically inhabits $[[B {a}]]$. Done.
-\end{proof}
-
-\begin{lemma}[$CR_2$ for type]
-  \label{lemma:cr2ty}
-  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
-\end{lemma}
-\begin{proof}
-  By induction over the derivation of $[[Interp I i A S]]$. The
-  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
-\end{proof}
-
-\begin{lemma}[$CR_2$ for types (Rec)]
-  \label{lemma:cr2tyrec}
-  If $[[InterpR i A S]]$, then $[[wn A]]$.
-\end{lemma}
-\begin{proof}
-  By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
-\end{proof}
-
-\begin{lemma}[$CR$ for terms (Rec)]
-  \label{lemma:crelrec}
-  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
-\end{lemma}
-\begin{proof}
-  After unfolding the definition of $[[InterpR i A S]]$, trivial by
-  Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
-\end{proof}
-
-The formulation of the valuation and semantic well-typedness from
-Figure~\ref{fig:semtyping} and the fundamental lemma remains
-unchanged.
-The proof of the fundamental lemma is still carried out by induction
-over the typing derivation, where the additional neutral term related
-cases are handled by Lemmas~\ref{lemma:cr2tyrec} and
-\ref{lemma:crelrec}.
-
-The normalization property then follows as a corollary of the
-fundamental theorem.
-\begin{corollary}[Existence of $\beta$-normal form]
-  \label{corollary:exbetanf}
-  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
-\end{corollary}
-\begin{proof}
-  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
-  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
-  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
-  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
-  natural numbers as term variables. The side condition $[[idtm |=
-  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
-  including variables, semantically inhabit any $[[S0]]$ where
-  $[[S0]]$ is the interpretation of some type. With our choice of
-  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
-  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
-  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
-  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
-\end{proof}
-From Corollary~\ref{corollary:exbetanf}, we can show that there exists
-$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
-preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
-and also strictly decreases the size of the term.
-
-Due to the non-deterministic nature of parallel reduction, we need to
-take a few more steps to convert the existence of $\beta\eta$-normal
-form into a decision procedure for type conversion. More specifically,
-we can show that a deterministic evaluation strategy such as
-leftmost-outermost reduction can always find the $\beta\eta$-normal
-form if there exists one. However, we omit such proofs since they can
-can be formulated on untyped lambda terms and thus orthogonal to the
-specifics of dependently typed systems. Instead, we redirect readers
-to \citet{factorization-essentially, takahashi-parallel-reduction} for
-the details.
-
-\scw{Name and foreshadow these lemmas when they are introduced so that
-it is easier to refer to them in the discussion.}
-Finally, we want to point out that
-Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
-\ref{lemma:crelrec}, often referred to as adequacy of the logical
-relation, are in fact required in the normalization proof for simply
-typed languages~\citep{abel2019poplmark}. Similar to the simply typed
-scenario, adequacy needs to be proven before the fundamental theorem
-so we can handle rules such as \rref{T-If} where the scrutinee is a
-neutral term. In \citet{abel2019poplmark}, a variant of
-Lemma~\ref{lemma:extwn} is used in the exact same way to show that
-lambda terms are themselves normalizing as we have done in
-Lemma~\ref{lemma:crel}.
-
-Dependent types make the proof slightly more complicated
-as we also need to know that every type has a normal form. Whereas the
-$CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
-derivable, the $CR_2$ property for types (Lemma~\ref{lemma:cr2ty})
-requires the $CR_1$ property about the interpreted sets
-(Lemma~\ref{lemma:crel}).
-
-Overall, despite the dependently typed setting,
-the extension of our logical relation to prove normalization of open
-\emph{and} closed terms closely mirrors the progression from
-normalization of closed terms~\citep{harpertait} to normalization of
-open terms~\citep{harperkripke} in the simply typed lambda calculus.
-It is in fact reassuring that once we have laid the foundational
-technique for handling dependent types in our logical relation, the
-further extensions more or less boil down to properties that can be
-derived through syntactic means, which tend to not get in the way of
-Gödel's second incompleteness theorem.
-
-\section{Mechanization}
-\label{sec:logrelmech}
-\begin{figure}[h]
-  \begin{tabular}{ c |  c  | c  }
-    & Consistency & Normalization \\
-    \hline
-    Library (Autosubst 2)  & 491 & - \\
-    Syntactic typing (specification) &  69 & - \\
-    Renaming (common)  & 46 & -  \\
-    Syntactic soundness & 660 &  - \\
-    Untyped reduction & 350 & 834 \\
-    Logical relation & 342 & 515 \\
-    Semantic typing and soundness & 169 & 197 \\
-  \end{tabular}
-  \caption{Statistics of the Coq Development}
-  \label{fig:linecount}
-\end{figure}
-
-Figure~\ref{fig:linecount} shows the statics of our
-development, including the base consistency proof from
-Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
-extension to normalization for open and closed terms from
-Section~\ref{sec:extension}.
-
-Since the normalization and the consistency development share the same
-syntactic and typing specification, they differ in only the categories
-that are related to the logical relation. We use - as a marker that
-the line count is the same.
-
-Autosubst 2 takes our syntax specification in higher-order abstract
-syntax and generates Coq syntax specification, renaming and
-substitution functions, and lemmas and tactics that allow reasoning
-about those functions. The auto-generated files and the library header
-files are counted toward the \emph{Library (Autosubst 2)} category.
-
-Orthogonal to the development of our consistency and normalization
-proof, we prove the syntactic soundness of our system through subject
-reduction and progress. The syntactic soundness proof shares the
-definition of renaming with our semantic soundness proof, which is
-factored out as a separate file under the Renaming (common) category.
-
-\subsection{Artifacts Specific to Coq}
-In this section, we discuss the Coq encoding of the definitions and proofs presented
-in Section~\ref{sec:logrelproof}  and the artifacts that are
-specific to Coq and may not appear in other proof assistants.
-
-The powerset $[[PowerSet STm]]$ is encoded as the type \texttt{tm ->
-Prop}, a predicate over \lang{} terms.
-The inductive
-definition of the logical relation in Figure~\ref{fig:logrel} requires
-the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
-the function $[[F]]$ can be later instantiated into the logical
-relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
-
-In Coq, there is a distinction between computable functions and
-relations that can later be proven to be functional. The former can be
-viewed as a strict subset of the latter in axiom-free Coq. To be more
-precise, given a relation \texttt{R : A -> B -> Prop} subject to the
-totality and functionality constraints ($\forall$ \texttt{a} $\in$
-\texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
-that \texttt{R a b} is inhabited), we do not immediately obtain a function
-\texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
-\texttt{R a (F a)} is inhabited. However, the functional side
-conditions of a relation is clunky to express and tend to block
-automation. A simple workaround is to assume the axiom of unique choice, which
-is known to be consistent with Coq and allows us to induce a function \texttt{F
-: A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
-is functional. This approach would make our Coq development match
-the text version of our proof from Section~\ref{sec:logrelproof} more
-closely.
-
-However, we choose instead an axiom-free workaround and define
-\rref{I-Pi} as follows in our Coq mechanization.
-\begin{center}
-  \drule[]{I-PiCoq}
-\end{center}
-It should be easy to verify that the preconditions of
-\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
-equivalent. After establishing Lemma~\ref{lemma:logreldeter}, it is
-possible to further show that the conclusions of the rules are
-equivalent, too.\scw{If it is easy, you should have already done it.
-Unless it is really long and boring.}
-
-% This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
-% before we even show that our logical relation is
-% functional/deterministic, but does not affect the proof structure
-% otherwise.
-We choose to keep this discrepancy between the Coq development and the
-description of the logical relation presented in Section~\ref{sec:logreldep} since the
-skolemization process is more intuitively expressed in terms of
-function symbols rather than relation symbols. Otherwise, we do not
-see a clear advantage of \rref{I-PiCoq} over \rref{I-Pi} in set
-theory, where there is no distinction between computable functions and
-functions in general. \scw{not sure I follow this last bit}
-
-\subsection{Automation}
-\label{sec:automation}
-Our Coq mechanization heavily uses automation, though instead of
-defining custom tactics, we rely mostly on off-the-shelf tools such as
-Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.
-
-We use the Autosubst 2 framework to specify our syntax in HOAS and
-produce Coq syntax files in de Bruijn representation. Additionallly,
-Autosubst 2 provides a powerful tactic \texttt{asimpl} that
-can be used to prove the equivalence of two terms constructed using
-the primitive operators provided by the framework. This greatly
-simplifies the reasoning about substitution in our development as
-almost all substitution related properties are immediately discharged
-by \texttt{asimpl} without having to manually prove or even directly invoke any
-substitution related lemmas. \scw{You are only referring to lemmas about
-how substitution relates to syntax, right? You still need to manually
-invoke lemmas about substitution and judgements.}
-
-For other automation tasks that are not specific to binding, we use
-the powerful \texttt{sauto} tactic provided by CoqHammer to write
-short and declarative proofs. For example, here is a one-line proof of
-the triangle property about parallel reduction, from which the diamond
-property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
-\scw{you have space, so repeat the mathematical statement of the lemma}
-\begin{minted}{coq}
-Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
-Proof.
-  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
-Qed.
-\end{minted}
-In prose, the triangle property can be proven by induction over the
-graph of the \texttt{tstar} function, which stands for the Takahashi
-translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
-\texttt{ctrs:Par} say that the proof involves inverting and
-constructing of the derivations of parallel
-reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
-automation tactic to use
-the reflexivity and congruence properties of parallel
-reduction as lemmas.
-\scw{What does \texttt{lq:on} and \texttt{hauto} mean?}
-
-The automation here not only gives us a proof that is shorter and more
-resiliant to changes in definition, but also gives useful
-documentation for readers who wish to understand how the underlying
-proof works since the automation tactic is guaranteed to not use
-lemmas or invert derivations that are not specified in the
-\texttt{use} or \texttt{inv} flag.
-
-\section{Related Work and Discussion}
-\label{sec:relatedwork}
-
-\scw{Maybe it would be useful to include a chart here, so that readers can
-  easily keep track of the features of the various languages.  i.e. which ones
-  include large eliminations? type-directed equivalence? impredicative prop?
-  inductive datatypes? what are their line counts?}
-
-\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
-\citet{barendregt:lambda-calculi-with-types} are some of the earlier
-works that establish metatheoretic results strong enough to derive
-consistency for dependently typed systems.
-\scw{Should give Martin L\"of credit for first consistency proof.}
-\scw{What techniques do these three proofs use? How is it different from
-your proof? What about Luo, which
-introduced a universe hierarchy? What about categories with families, or
-other ways of showing consistency?}
-Some of these techniques
-are still in use in more recent works. For example, the technique for
-proving strong normalization by \citet{geuvers1994short} is adapted by
-\citet{moon2021graded} to show the same property to a dependently
-typed system extended with modalities.
-
-\citet{decagda} mechanizes in Agda the decidability of type
-conversion rule for a dependently typed language with one predicative
-universe level and typed judgmental equality with function
-$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
-uses a Kripke-style logical relation parameterized over an
-type-directed equivalence relation satisfying certain
-properties. Their logical relation is defined using the
-induction-recursion scheme, which is available in Agda but not in Coq.
-\citet{martin-lof-a-la-coq} manages to encode the logical relation
-from \citet{decagda} in the predicative fragment of Coq using a
-special way of encoding induction recursion. Their work further
-extends the decidability of type conversion result from~\citet{decagda} to the decidability
-type checking of a bidrectional type system.
-
-\citet{anand2014towards} mechanizes the metatheory of
-Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
-extensional type theory with features such as dependent functions,
-inductive types, and a full universe hierarchy. \citet{nbeincoq}
-mechanizes the normalization-by-evaluation algorithm in Coq for a
-dependently typed language with one predicative universe, similar to
-\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
-\citet{anand2014towards} and \citet{nbeincoq} leverage the
-impredicative \texttt{Prop} sort of Coq to define the interpretation
-of dependent function types and thus are closely related to our
-mechanization. However, instead of explaining the purely inductive
-logical relation as a convoluted workaround to the lack of
-induction-recursion in Coq, we give a self-contained explanation of
-our logical relation.
-
-Finally, it is worth pointing out that all systems we have discussed
-so far builds a relational model for their logical relation. Recall
-that our logical relation takes the form $[[InterpR i A S]]$ where
-$[[S]]$ is a set of terms. In a relational model, each type is
-interpreted as a partial equivalence relation over terms rather than a
-set. A relational model is necessary for an extensional type theory
-such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
-use a relational model either to justify the function $\eta$-law or to
-derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
-law for function is baked into the untyped reduction
-relation and therefore a relational model is not needed.
-\scw{would it be difficult to extend your proof to be a PER?}
-
-The dependently typed systems we have discussed so far also vary
-greatly in expressiveness. For example, the Calculus of
-Constructions~\citep{CoC} and MLTT without universe levels are unable
-to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
-identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
-the mechanized systems we have discussed so far, only
-\citet{anand2014towards} has a full universe hierarchy.
-Our object language \lang{}, while small, has a decent coverage of
-features commonly seen in dependently typed languages, including
-dependent pattern matching, dependent function types, a full universe
-hierarchy, an intensional identity type. We hope features such as
-type-directed reduction (necessary for unit $\eta$-law) and
-impredicative sorts can be implemented in a similarly streamlined
-fashion and we will leave those extensions as part of our future work.
-
-% To end this section, we
-\scw{should this be a new subsection?}
-The final question we want to address is: why is our proof,
-even with the extension from Section~\ref{sec:extension}, so much
-shorter than the proofs from \citet{decagda, nbeincoq,
-  martin-lof-a-la-coq}?
-First, unlike developments that mechanize the correctness of a clever
-type conversion algorithm, we only show the existence of
-normal form for open and closed terms and state our properties in
-terms of the untyped small-step reduction relation. This removes a lot
-of scaffolding in the specification of our type system and our logical
-relation.
-
-This contrasts our development with
-\citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
-difficult property to prove due to the complexity of the equality
-judgment used to define the logical relation. In our system, the
-irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
-direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
-lemma name} which is
-straightforward after we prove the results about parallel
-reduction in Section~\ref{sec:spec}. In other words, we keep the
-the logical relation itself minimal at the cost of extra lemmas that can be proven
-through syntactic means independently from the logical relation.
-For example, as commented near the end of Section~\ref{sec:extension},
-our normalization property (Corollary~\ref{corollary:exbetanf})
-% can induce a decision procedure for type conversion by reducing two
-% terms to $\beta\eta$-normal form and check for their syntactic
-% equivalence. This decision procedure would be inefficient compared to
-% NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
-% to weak head normal and checks for the equivalence of subterms
-% recursively
-does not immediately induce an efficient decision procedure, but it is
-possible to recover an efficient algorithm by reasoning about untyped
-lambda terms.
-
-% the concision of
-% our proof can be largely attributed to the fact that we bake into the
-% logical relation only the minimum amount of information that is the
-% most conducive to the proof of the metatheoretic result.
-
-Another key to our proof is the very early establishment of the confluence
-property of the paralell reduction relation
-(Lemma~\ref{lemma:pardiamond}). This property is used in
-Lemma~\ref{lemma:interppreservation} to show that the interpretation
-of a type is preserved under evaluation. That is, if $[[InterpR i A
-S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
-In a system with typed directed reduction, the confluence result is
-harder to establish since confluence requires subject reduction, which
-circularly depends on $\Pi$-injectivity, a consequence of
-confluence~\citep{siles2012pure}. There are different workarounds to
-this problem. \citet{siles2012pure} proposes a syntactic approach by
-defining a type system where the confluence result is
-directly provable and later show that the system of interest is
-equivalent to the system with the confluence property.
-\citet{decagda} uses a relational model and the relational counterpart
-of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
-model is parameterized by a typed indexed equivalence relation and
-is reused to prove derive different properties in their
-development. Another perhaps simpler approach is to simply
-extend the judgmental equality with rules about
-$\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
-reduction to be proven independently from confluence and
-$\Pi$-injectivity can later be shown to be admissible. We find the
-approach from \citep{weirich:systemd} the most lightweight as it
-allows us to derive confluence early on even in systems with typed
-conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}
-
-In cases where the judgmental equality does not leverage type information heavily
-(e.g. $\eta$-law for the unit type), it might be possible to show that
-$\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
-$[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
-property holds (we believe this is likely true for both
-\citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
-terms of untyped parallel reduction to show the decidability of type
-conversion. Otherwise, to fully model the system, we need to bake
-typing information into our logical relation and we need to redefine
-our logical relation in Kripke-style so we can relax the scoping
-constraint in order to prove adequacy, though a relational model would
-still be unnecessary as long as the $\eta$ laws are baked into the
-typed directed relation. \scw{not sure I understand this last bit}
-
-Since confluence plays such as key role in our proof, we are uncertain
-how to generalize our technique to systems where confluence does not
-hold, and we will investigate those scenarios in our future work.
-\scw{What systems are you thinking of? Or do you mean more generally
-that this technique only works for languages where the definition of
-equality is based on joinability of parallel reduction?}
-
-% We are able to avoid the complexivity of a Kripke-style logical relation
-% by defining our logical relation in terms of untyped parallel
-% reduction, which does not impose any scoping constraint on the
-% terms.
-
-% Why is it that the justification of $\eta$ law
-% requires a Kripke-style relational model in all three developments
-% but not ours? Are we ``cheating'' by defining \lang{}'s equational
-% theory using an untyped conversion rule? Instead of answering the
-% questions one by one, we clarify the design of our type system and our
-% proof technique so the answers to those questions are self-evident.
-
-% The key to the shortness of
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-% Type soundness can be proven through a syntactic
-% approach~\citep{syntacticsoundness} as a corollary of two properties:
-% progress and preservation. % The syntactic type soundness proof
-% % varies in complexity depending on the underlying type
-% % system. For example, a type system that tracks information flow would
-% % require additional structural rules related to security levels. In
-% % this paper, we focus on one specific type of complexity: the
-% In Figure~\ref{fig:stlcsoundness}, we summarize the structure of the
-% syntactic type soundness proof for the simply typed lambda
-% calculus. Each lemma can be proven by structural induction over the
-% typing derivation, while using the previous established results as
-% lemmas for specific cases that do not immediately follow from the
-% induction hypothesis. If we make our language more complex by adding
-% full dependent type support, the overall structure remains almost
-% identical.
-
-% NbE in Coq
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{refs}
-
-\appendix
-
-\subsection{More detailed proofs}
-\begin{lemma}[Preservation of the logical relation]
   If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
@@
 \end{proof}
-
+From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
+derive the following corollary that two coherent types have the same
+interpretation.
+\scw{I'm not used to this lemma being called irrelevance. Intuitively,
+it says that the definition of the logical relation respects
+definitional equality. }
+\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
+  \label{lemma:logrelcoherence}
+  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
+\end{corollary}
+\scw{Need a transition here}
+Next, we show that $[[Interp I i A S]]$ is in fact a partial function.
 \begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
+  \label{lemma:logreldeter}
   If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
@@
 
+Lemma~\ref{lemma:logreldeter} enables us to show the admissibility of
+\rref{I-PiAlt} and its related inversion principle.
+\begin{lemma}[Pi Intro Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
+  \label{lemma:piintroalt}
+  Suppose the following two statements hold:
+  \begin{itemize}
+  \item $[[Interp I i A S]]$
+  \item $[[forall a, a in S implies (# exists S0 , Interp I i B {a} S0 #)]]$
+  \end{itemize}
+  Then we have $[[Interp I i Pi A B { b | forall a, (# a in S , forall
+    S0, (# Interp I i B {a} S0,  b a in F a #) #) }]]$
+\end{lemma}
+\begin{proof}
+  Let $[[F]]$ be the relation defined as follows:
+  \[ (a,S) \in [[F]] \iff [[Interp I i B {a} S]] \]
+  By the second bullet from the premise and
+  Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a function that is total
+  on the set $[[S]]$. The conclusion then trivially follows from
+  \rref{I-Pi}.
+\end{proof}
+
+\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
+  \label{lemma:piinvalt}
+  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
+  such that the following constraints hold:
+  \begin{itemize}
+  \item $[[Interp I i A S0]]$
+  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
+    S1 #) #)]]$
+  \item $[[S = { b | forall a, (# a in S0 , forall
+    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
+\end{proof}
+
+% The combination of Lemmas~\ref{lemma:piintroalt} and
+% \ref{lemma:piinvalt} reveals what we truly want to capture with
+% \rref{I-Pi}. A function type is semantically well-defined if its input
+% type is well-defined, and its output type is defined for each valid
+% inahbitant of its input type.
+
+The next lemma shows that our logical relation satisfies
+cumulativity. That is, if a type has an interpretation at a lower
+universe level, then we can obtain the same interpretation at a higher
+universe level.
+\begin{lemma}[Logical relation cumulativity\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
+  \label{lemma:logrelcumulativity}
+  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
+\end{lemma}
+\begin{proof}
+  Trivial by structural induction over the derivation of $[[Interp I
+  i0 A S]]$.
+\end{proof}
+Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
+implicitly assume that $[[I]]$ is defined on the set of natural
+numbers less than $[[i1]]$.
+
+\begin{corollary}[Logical relation is functional with different levels\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
+  \label{lemma:logreldeterhet}
+  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
+\end{corollary}
+\begin{proof}
+  Immediate from Lemma~\ref{lemma:logreldeter} and
+  \ref{lemma:logrelcumulativity}.
+\end{proof}
+
+We say that a set of terms $[[S]]$ is closed under expansion if given
+$[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
+The final property we want to show is that the output set $[[S]]$ from
+the logical relation is closed under expansion. Unlike the previous
+lemmas, we need to constrain the function $[[I]]$ so its outputs are
+all closed under expansion.
+\scw{How is this related to the comment before 3.2 (i.e. the property
+derived from I-Red?}
+\begin{lemma}[Logical Relation Elements back preservation\footnote{\dotv{semtyping.v}{InterpExt\_back\_clos}}]
+  \label{lemma:logrelbackclos}
+  If $[[Interp I i A S]]$ and $[[I]]$ satisfies the property that
+  for all $[[i]]$, $[[I i]]$ is closed under expansion, then the set
+  $[[S]]$ is closed under expansion.
+\end{lemma}
+\begin{proof}
+  Trivial by structural induction over the derivation of $[[Interp I i
+  A S]]$. The function case requires the following simple fact about
+  parallel reduction:
+  If $[[b0 => b1]]$ then $[[b0 a => b1 a]]$ for all $[[a]]$. This fact
+  is not immediate from the definition of parallel reduction but
+  follows from Lemma~\ref{lemma:parrefl} and \rref{P-App}.
+\end{proof}
+
+\begin{corollary}[Logical Relation Elements back preservation (rec)\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
+  \label{lemma:logrelNbackclos}
+  If $[[InterpR i A S]]$, then $[[S]]$ is closed under expansion.
+\end{corollary}
+\begin{proof}
+  Immediate from Lemma~\ref{lemma:logrelbackclos}, the definition of
+  $[[InterpR i A S]]$, and \rref{I-Red}.
+\end{proof}
+
+\section{Semantic Typing and Consistency}
+\label{sec:logrelproof}
+\begin{figure}[h]
+\[
+\begin{array}{lcl}
+
+      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
+                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
+      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
+                       G]]\text{ then there exists some } [[j]] \text{
+                       and } [[S]] \text{ such that } [[InterpR j A
+                       {rho} S]] \text{ and } [[a {rho} in S]] \\
+      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
+\end{array}
+\]
+  \caption{Semantic Typing for \lang}
+  \label{fig:semtyping}
+\end{figure}
+
+The logical relation we define in Figure~\ref{fig:logrel} does not
+include cases for variables. Likewise, for the base types such as
+boolean and equality, the output set $[[S]]$ contains only terms that
+evaluate to closed terms.
+
+\scw{What is the motivation for this section? Why do we want to generalize
+the logical relation?}
+To generalize our logical relation to open
+terms, we define the semantic typing judgment by closing the open
+terms with a substitution whose codomain consists of terms that
+respect the interpretation of the types from the context. The full
+definitions of well-formed substitution ($[[rho |= G]]$), semantic
+typing ($[[ G |= a : A]]$), and semantic context well-formedness
+($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
+\scw{explain all parts of this figure. it's pretty intimidating.}
+In the
+definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
+the typing context obtained by dropping the last $[[i]]$ elements of
+$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
+returns the empty list.
+
+The following lemma makes the statement $[[G |= A : Set i]]$ easier to
+work with.
+\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}\scw{spell out the whole word}]
+  \label{lemma:setinv}
+  The following two statements are equivalent:
+  \begin{itemize}
+  \item $[[G |= A : Set i]]$
+  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
+    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  The forward direction is immediate by
+  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
+  and show that $[[G |= A : Set i]]$ given the second bullet.
+
+  Suppose $[[rho |= G]]$, then we know that there exists some $[[S]]$
+  such that $[[InterpR i (A {rho}) S]]$. By the definition of semantic
+  typing, it suffices to show that there exists some $[[j]]$ and
+  $[[S0]]$ such that  $[[InterpR j Set i S0]]$ and $[[A {rho} in
+  S0]]$.
+  Pick $[[Suc i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
+  ]]$ for $[[S0]]$ and it is trivial to verify the conditions hold.
+\end{proof}
+
+The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
+counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
+than inductively over the context. \scw{why is that the case?} It is easy to recover the same
+structural rules:
+\begin{lemma}[Semantic context well-formedness cons]
+  \label{lemma:semwffcons}
+  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
+\end{lemma}
+\scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
+Ctx-empty?}
+\begin{proof}
+  By the definition of semantic context well-formedness, the goal is
+  to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i
+  (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
+  proven by case analysis on whether $[[i]]$ is zero.
+\end{proof}
+
+Likewise, the semantic well-formendness judgment for substitutions
+satisfies similar structural rules.
+\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
+  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
+  $[[rho .: a |= G ++ A]]$.
+\end{lemma}
+\begin{proof}
+  Start by unfolding the definition of $[[rho .: a |= G ++ A]]$ and
+  performing a case analysis similar to the proof of
+  Lemma~\ref{lemma:semwffcons}.  The case where the number is $[[0]]$
+  requires Lemma~\ref{lemma:logreldeterhet} to finish the proof.
+\end{proof}
+
+Next, we show some non-trivial cases of the fundamental theorem as
+top-level lemmas and leave the remaining cases as exercises for the
+reader.\scw{I dislike leaving ``exercises to the reader''}
+
+First, we formulate the definition of valid renamings and prove that
+semantic typing satisfies renaming so we can weaken the context when
+reasoning about the variable case of the fundamental lemma
+(Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G
+++ D]]$, it is easy to show that we can extract some valuation
+$[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
+``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :
+A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
+$[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
+after weakening the context; this implication holds because $[[rho |=
+G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
+can make use of the premise $[[G |= a : A]]$ to derive what we need
+for the conclusion. We recommend the readers to skip the proofs of
+Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
+during the first read as long as they have an intuitive understanding
+of what the renaming property is meant to capture.
+
+\scw{Make this a definition? Would it be useful to create notation
+  for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
+We say that $[[xi]]$ is
+valid from the context $[[G]]$ to the context $[[D]]$ if the
+following condition holds.
+\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
+  [[D ren xi
+  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]
+
+\begin{lemma}[Truncate is valid]
+  \label{lemma:validtruncate}
+  If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i
+  G]]$ to $[[G]]$.
+\end{lemma}
+\begin{proof}
+  By the definition of a valid renaming, we must show that given $[[j
+  < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
+  \begin{itemize}
+  \item $[[ren up i j]] = i + j  < [[| G | ]]$
+  \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
+  \end{itemize}
+  The first bullet point is immediate from the fact that $[[j]] <
+  [[|G|]] - [[i]]$. The second bullet follows from unfolding the
+  definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
+  $[[i]] + [[j]] < [[|G|]]$.
+\end{proof}
+
+\begin{lemma}[Renaming for $[[rho |= G]]$]
+  \label{lemma:renamingval}
+  If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
+  $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
+\scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
+\end{lemma}
+\begin{proof}
+  Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
+  that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i
+  < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.
+
+  By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
+have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho
+} ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |=
+D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
+S]]$ and $[[rho ren xi i in S]]$.
+\end{proof}
+
+\begin{lemma}[Renaming for $[[G |= a : A]]$]
+  \label{lemma:semrenaming}
+  If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
+  $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
+\end{lemma}
+\begin{proof}
+  Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
+\end{proof}
+
+\scw{Need a transition here that you are starting to explain the semantic typing
+rules.}
+
+\begin{lemma}[ST-Var]
+  \label{lemma:stvar}
+  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
+\end{lemma}
+\begin{proof}
+  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
+need to show that there exists some $[[j]]$ and $[[S]]$ such that
+  \begin{itemize}
+  \item $[[InterpR j G i < up Suc i > { rho } S]]$
+  \item $[[rho i in S]]$
+  \end{itemize}
+  From the definition of $[[|= G]]$, we know that there exists some
+  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. By
+  Lemma~\ref{lemma:validtruncate}, we know that $[[up Suc i]]$ is a
+  valid renaming from $[[drop Suc i G]]$ to $[[G]]$. By
+  Lemma~\ref{lemma:semrenaming}, we deduce $[[G |= G i < up Suc
+  i > : Set j]]$. By the identity from Lemma~\ref{lemma:setinv}, we
+  know that $[[InterpR j G i < up Suc i > { rho } S]]$. It now
+  suffices to show $[[rho i in S]]$, but that is immediate from the
+  definition of $[[rho |= G]]$.
+\end{proof}
+
+\begin{lemma}[ST-Set]
+  \label{lemma:stset}
+  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
+\end{lemma}
+\begin{proof}
+  Immediate by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
+\end{proof}
+
+\begin{lemma}[ST-Pi]
+  \label{lemma:stpi}
+  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
+  A B : Set i]]$.
+\end{lemma}
+\begin{proof}
+  Applying Lemma~\ref{lemma:setinv} to the
+  conclusion, it now suffices to show that given $[[rho |= G]]$, there
+  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
+  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
+  there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
+From $[[G ++ A |= B : Set i]]$, we know that there must
+exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
+in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
+\end{proof}
+
+\begin{lemma}[ST-Abs]
+  \label{lemma:stabs}
+  If $[[G |= Pi A B : Set i]]$ and $[[G ++ A |= b : B]]$, then $[[G |=
+  \ A b : Pi A B]]$.
+\end{lemma}
+\begin{proof}
+  By unfolding the definition of $[[G |= \ A b : Pi A B]]$, we need to
+  show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
+  $[[S]]$ such that $[[InterpR i Pi A {rho} B {up rho} S]]$ and $[[\
+  A{rho} b{up rho} in S]]$.
+
+  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
+  i]]$, there exists some set $[[S]]$ such that
+  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
+  $[[\A{rho} b{up rho} in S
+  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
+  that all following conditions hold:
+  \begin{itemize}
+  \item $[[InterpR i A{rho} S0]]$
+  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
+    {rho .: a}
+    S1 #) #)]]$
+  \item $[[S = { b | forall a, (# a in S0 , forall
+      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
+  \end{itemize}
+  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
+  that given $[[a in S0]]$,
+  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
+  a in S1]]$.
+  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
+  under expansion. By Lemma~\ref{lemma:parabscbn}, since $[[( \A{rho} b{up rho} )
+  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
+  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
+  B]]$ and the fact that the logical relation is deterministic and
+  cumulative (Lemma~\ref{lemma:logreldeterhet}).
+\end{proof}
+
+\begin{lemma}[ST-App]
+  \label{lemma:stapp}
+  If $[[G |= b : Pi A B]]$ and $[[G |= a : A]]$, then $[[G |= b a : B {a}]]$.
+\end{lemma}
+\begin{proof}
+Suppose $[[rho |= G]]$. The goal is to show that there exists some
+$[[i]]$ and $[[S1]]$
+such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a} {rho}
+S1]]$, or equivalently, $[[InterpR i B {rho .: a {rho}} S1]]$ since
+$[[B {a}{rho}]] = [[B {rho .: a {rho}}]]$. By the premise $[[G |= b :
+Pi A B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
+there exists some $[[i]]$ and $[[S0]]$ such that:
+  \begin{itemize}
+  \item $[[InterpR i A{rho} S0]]$
+  \item $[[forall a0, (# a0 in S0 implies (# exists S1 , InterpR i B
+    {rho .: a0}
+    S1 #) #)]]$
+  \item $[[forall a0, (# a0 in S0 , forall
+      S1, (# InterpR i B {rho .: a0} S1,  b {rho} a0 in S1 #) #)]]$
+  \end{itemize}
+  Instantiating the variable $[[a0]]$ from the last two bullets with
+  the term $[[a {rho}]]$, the conclusion immediately follows.
+\end{proof}
+
+\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
+  \label{theorem:soundness}\leavevmode
+  \begin{itemize}
+  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
+  \item If $[[|- G]]$, then $[[|= G]]$.
+  \end{itemize}
+\end{theorem}
+\begin{proof}
+  Proof by mutual induction over the derivation of $[[G |- a :
+  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
+  from Lemma~\ref{lemma:semwffcons}.
+Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
+  can be used to discharge their syntactic counterpart
+  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
+  cases not covered by the lemmas are similar to the ones already
+  shown or simpler and therefore omitted from the text.
+
+\end{proof}
+
+\scw{Spell this out in a little more detail.}
+
+\begin{proof}
+  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
+\end{proof}
+
+\section{Toward Decidability of Type Conversion and $\eta$ laws}
+\label{sec:extension}
+\begin{figure}[h]
+  \[
+    \begin{array}{lcl}
+      % \beta\text{-}\mathit{neutral\ terms}\\
+      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
+                    f]] \\ \\
+      % \beta\text{-}\mathit{normal\ terms} \\
+      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
+                    |\ [[f ~ f : f]]\\
+            & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
+    \end{array}
+  \]
+  \caption{$\beta$-neutral and normal forms}
+  \label{fig:nenf}
+\end{figure}
+\begin{figure}[h]
+  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
+  \caption{Extended logical relation}
+  \label{fig:logrelopen}
+\end{figure}
+
+\scw{Put this introduction before the figure in the section.}
+In this section, we sketch out how the logical relation from
+Section~\ref{sec:logreldep} can be extended show the existence of
+$\beta\eta$ normal forms for (open and closed) well-typed terms. We
+first extend parallel reduction to include $\eta$ reduction for
+functions.
+\begin{center}
+  \drule[width=2.5in]{P-AbsEta}
+\end{center}
+\scw{What happened to the domain type in the $\lambda$ expression? This is
+not your syntax...}
+\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
+equational theory since coherence, the untyped relation used for type
+conversion, is built on top of parallel reduction. We can recover the
+same confluence result about parallel reduction using the standard
+techniques from \citet{barendregt:lambda-calculi-with-types,
+takahashi-parallel-reduction}, though specifically for the de Bruijn
+representation, we need the following anti-renaming lemma about
+parallel reduction.
+\begin{lemma}[Par anti-renaming]
+  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
+exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
+\end{lemma}
+
+The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the neutral terms
+and normal forms with respect to $\beta$-reduction, but can still take
+$\eta$ reduction steps. We sometimes use the judgment forms $[[ne a]]$
+and $[[nf a]]$ to indicate that there exists $[[e]]$ or $[[f]]$ such
+that $[[a = e]]$ or $[[a = f]]$.
+
+We can show that parallel reduction preserves $\beta$-normal and
+neutral forms.
+\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
+  \label{lemma:parnenf}
+  If $[[a => b]]$, then
+  \begin{itemize}
+  \item $[[ne a]]$ implies $[[ne b]]$
+  \item $[[nf a]]$ implies $[[nf b]]$
+  \end{itemize}
+\end{lemma}
+This lemma is mainly to show that \rref{P-AbsEta} does not introduce
+any new $\beta$ redex because $\eta$ reduction is the only real step
+we can take.
+
+The predicates $[[wne a]]$ and $[[wn
+a]]$ describe terms that can evaluate into $\beta$-neutral or
+$\beta$-normal form through parallel reduction and are defined as
+follows.\scw{What does $wne$ stand for?}
+\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
+\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
+
+The updated logical relation and its auxiliary definitions are shown
+in Figure~\ref{fig:logrelopen}.
+
+% The idea of a term blocked from $\beta$ reduction is
+% captured by $[[e]]$, the set of neutral terms. We augment the
+% interpretation of booleans and the empty types with terms that
+% evaluate to neutral terms. Furthermore, types that are neutral terms
+% can also be assigned a meaning since they may be inhabited by other
+% neutral terms.
+
+In the definition of the logical relation, we omit the rules for
+the function and universe cases since they remain identical to the
+original version in Figure~\ref{fig:logrel}.
+The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
+An open term of type $[[Bool]]$ does
+not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
+a variable, or more generally, a neutral term. Likewise, the
+$[[Void]]$ type, while remains uninhabited under an empty context, may
+be inherited by the set of neutral terms when there is a variable in
+the context that allows us to inhabit $[[Void]]$.
+
+The rule for equality type $[[a ~ b : A]]$ is augmented with the preconditions that
+$[[a]]$, $[[b]]$, and $[[A]]$ are all in their normal forms since
+otherwise our model would include equality types that are themselves
+not normalizing. Furthermore, the side condition $[[a <=> b]]$ is only
+available when the equality proof reduces to $[[refl]]$. If the proof
+term reduces to a neutral term, then there is nothing we can learn
+about the relationship between $[[a]]$ and $[[b]]$.
+
+Finally, in a
+non-empty context, a type itself may evaluate to a neutral term and in
+turn can only inhabited by neutral terms, thus the addition of \rref{I-Ne}.
+
+Interestingly, all the properties we have shown in
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the
+fundamental lemma can be proven in the exact same order, where the new
+cases due to \rref{I-Ne} and the augmentation of neutral terms
+to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
+Lemma~\ref{lemma:parnenf}.
+
+Before we can prove the fundamental theorem and derive the
+normalization property as its corollary, we need to additionally
+formulate and prove the adequacy property.
+
+Let us first start with some auxiliary lemmas and definitions.
+\begin{lemma}[Ext Wn]
+  \label{lemma:extwn}
+  If $[[wn a var i]]$, then $[[wn a]]$.
+\end{lemma}
+\begin{proof}
+  By induction over the length of the reduction sequence in $[[wn a
+  var i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
+  \ref{lemma:parnenf}.
+\end{proof}
+
+\begin{lemma}[wne wn]
+  \label{lemma:wnewn}
+  If $[[wne a]]$ and $[[wn b]]$, then $[[wne a b]]$.
+\end{lemma}
+\begin{proof}
+  Immediate by lexicographical induction over the length of the reduction sequences in
+  $[[wne a]]$ and $[[wn b]]$.
+\end{proof}
+
+\scw{What does CR stand for?}
+\begin{definition}[CR]
+  Let $[[S]]$ be a set of lambda terms. We say that
+  \begin{itemize}
+  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
+  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
+  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
+  \end{itemize}
+\end{definition}
+
+\begin{lemma}[$CR_1$ for type]
+  \label{lemma:cr1ty}
+  If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
+\end{lemma}
+\begin{proof}
+  Immediately from \rref{I-Ne} and \rref{I-Par}.
+\end{proof}
+
+\begin{lemma}[$CR$ for interpreted sets]
+  \label{lemma:crel}
+  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $[[Interp I i A S]]$. The base
+  cases are all immediate.
+
+  In the function case ($[[Pi A B]]$), we use the induction
+  hypothesis to show that variables, which are special cases of
+  neutral terms, semantically inhabit the input type $[[A]]$. We apply the
+  function to an arbitrary variable and we can then use
+  Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.
+
+  To show
+  the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
+  such that $[[wne b]]$,
+  and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
+  us a term that semantically inhabits $[[B {a}]]$. From the induction
+  hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
+  holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
+  from the induction hypothesis, since every term that evaluates to
+  some neutral form semantically inhabits $[[B {a}]]$. Done.
+\end{proof}
+
+\begin{lemma}[$CR_2$ for type]
+  \label{lemma:cr2ty}
+  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $[[Interp I i A S]]$. The
+  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
+\end{proof}
+
+\begin{lemma}[$CR_2$ for types (Rec)]
+  \label{lemma:cr2tyrec}
+  If $[[InterpR i A S]]$, then $[[wn A]]$.
+\end{lemma}
+\begin{proof}
+  By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
+\end{proof}
+
+\begin{lemma}[$CR$ for terms (Rec)]
+  \label{lemma:crelrec}
+  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
+\end{lemma}
+\begin{proof}
+  After unfolding the definition of $[[InterpR i A S]]$, trivial by
+  Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
+\end{proof}
+
+The formulation of the valuation and semantic well-typedness from
+Figure~\ref{fig:semtyping} and the fundamental lemma remains
+unchanged.
+The proof of the fundamental lemma is still carried out by induction
+over the typing derivation, where the additional neutral term related
+cases are handled by Lemmas~\ref{lemma:cr2tyrec} and
+\ref{lemma:crelrec}.
+
+The normalization property then follows as a corollary of the
+fundamental theorem.
+\begin{corollary}[Existence of $\beta$-normal form]
+  \label{corollary:exbetanf}
+  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
+\end{corollary}
+\begin{proof}
+  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
+  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
+  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
+  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
+  natural numbers as term variables. The side condition $[[idtm |=
+  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
+  including variables, semantically inhabit any $[[S0]]$ where
+  $[[S0]]$ is the interpretation of some type. With our choice of
+  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
+  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
+  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
+  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
+\end{proof}
+From Corollary~\ref{corollary:exbetanf}, we can show that there exists
+$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
+preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
+and also strictly decreases the size of the term.
+
+Due to the non-deterministic nature of parallel reduction, we need to
+take a few more steps to convert the existence of $\beta\eta$-normal
+form into a decision procedure for type conversion. More specifically,
+we can show that a deterministic evaluation strategy such as
+leftmost-outermost reduction can always find the $\beta\eta$-normal
+form if there exists one. However, we omit such proofs since they can
+can be formulated on untyped lambda terms and thus orthogonal to the
+specifics of dependently typed systems. Instead, we redirect readers
+to \citet{factorization-essentially, takahashi-parallel-reduction} for
+the details.
+
+\scw{Name and foreshadow these lemmas when they are introduced so that
+it is easier to refer to them in the discussion.}
+Finally, we want to point out that
+Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
+\ref{lemma:crelrec}, often referred to as adequacy of the logical
+relation, are in fact required in the normalization proof for simply
+typed languages~\citep{abel2019poplmark}. Similar to the simply typed
+scenario, adequacy needs to be proven before the fundamental theorem
+so we can handle rules such as \rref{T-If} where the scrutinee is a
+neutral term. In \citet{abel2019poplmark}, a variant of
+Lemma~\ref{lemma:extwn} is used in the exact same way to show that
+lambda terms are themselves normalizing as we have done in
+Lemma~\ref{lemma:crel}.
+
+Dependent types make the proof slightly more complicated
+as we also need to know that every type has a normal form. Whereas the
+$CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
+derivable, the $CR_2$ property for types (Lemma~\ref{lemma:cr2ty})
+requires the $CR_1$ property about the interpreted sets
+(Lemma~\ref{lemma:crel}).
+
+Overall, despite the dependently typed setting,
+the extension of our logical relation to prove normalization of open
+\emph{and} closed terms closely mirrors the progression from
+normalization of closed terms~\citep{harpertait} to normalization of
+open terms~\citep{harperkripke} in the simply typed lambda calculus.
+It is in fact reassuring that once we have laid the foundational
+technique for handling dependent types in our logical relation, the
+further extensions more or less boil down to properties that can be
+derived through syntactic means, which tend to not get in the way of
+Gödel's second incompleteness theorem.
+
+\section{Mechanization}
+\label{sec:logrelmech}
+\begin{figure}[h]
+  \begin{tabular}{ c |  c  | c  }
+    & Consistency & Normalization \\
+    \hline
+    Library (Autosubst 2)  & 491 & - \\
+    Syntactic typing (specification) &  69 & - \\
+    Renaming (common)  & 46 & -  \\
+    Syntactic soundness & 660 &  - \\
+    Untyped reduction & 350 & 834 \\
+    Logical relation & 342 & 515 \\
+    Semantic typing and soundness & 169 & 197 \\
+  \end{tabular}
+  \caption{Statistics of the Coq Development}
+  \label{fig:linecount}
+\end{figure}
+
+Figure~\ref{fig:linecount} shows the statics of our
+development, including the base consistency proof from
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
+extension to normalization for open and closed terms from
+Section~\ref{sec:extension}.
+
+Since the normalization and the consistency development share the same
+syntactic and typing specification, they differ in only the categories
+that are related to the logical relation. We use - as a marker that
+the line count is the same.
+
+Autosubst 2 takes our syntax specification in higher-order abstract
+syntax and generates Coq syntax specification, renaming and
+substitution functions, and lemmas and tactics that allow reasoning
+about those functions. The auto-generated files and the library header
+files are counted toward the \emph{Library (Autosubst 2)} category.
+
+Orthogonal to the development of our consistency and normalization
+proof, we prove the syntactic soundness of our system through subject
+reduction and progress. The syntactic soundness proof shares the
+definition of renaming with our semantic soundness proof, which is
+factored out as a separate file under the Renaming (common) category.
+
+\subsection{Artifacts Specific to Coq}
+In this section, we discuss the Coq encoding of the definitions and proofs presented
+in Section~\ref{sec:logrelproof}  and the artifacts that are
+specific to Coq and may not appear in other proof assistants.
+
+The powerset $[[PowerSet STm]]$ is encoded as the type \texttt{tm ->
+Prop}, a predicate over \lang{} terms.
+The inductive
+definition of the logical relation in Figure~\ref{fig:logrel} requires
+the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
+the function $[[F]]$ can be later instantiated into the logical
+relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
+
+In Coq, there is a distinction between computable functions and
+relations that can later be proven to be functional. The former can be
+viewed as a strict subset of the latter in axiom-free Coq. To be more
+precise, given a relation \texttt{R : A -> B -> Prop} subject to the
+totality and functionality constraints ($\forall$ \texttt{a} $\in$
+\texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
+that \texttt{R a b} is inhabited), we do not immediately obtain a function
+\texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
+\texttt{R a (F a)} is inhabited. However, the functional side
+conditions of a relation is clunky to express and tend to block
+automation. A simple workaround is to assume the axiom of unique choice, which
+is known to be consistent with Coq and allows us to induce a function \texttt{F
+: A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
+is functional. This approach would make our Coq development match
+the text version of our proof from Section~\ref{sec:logrelproof} more
+closely.
+
+However, we choose instead an axiom-free workaround and define
+\rref{I-Pi} as follows in our Coq mechanization.
+\begin{center}
+  \drule[]{I-PiCoq}
+\end{center}
+It should be easy to verify that the preconditions of
+\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
+equivalent. After establishing Lemma~\ref{lemma:logreldeter}, it is
+possible to further show that the conclusions of the rules are
+equivalent, too.\scw{If it is easy, you should have already done it.
+Unless it is really long and boring.}
+
+% This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
+% before we even show that our logical relation is
+% functional/deterministic, but does not affect the proof structure
+% otherwise.
+We choose to keep this discrepancy between the Coq development and the
+description of the logical relation presented in Section~\ref{sec:logreldep} since the
+skolemization process is more intuitively expressed in terms of
+function symbols rather than relation symbols. Otherwise, we do not
+see a clear advantage of \rref{I-PiCoq} over \rref{I-Pi} in set
+theory, where there is no distinction between computable functions and
+functions in general. \scw{not sure I follow this last bit}
+
+\subsection{Automation}
+\label{sec:automation}
+Our Coq mechanization heavily uses automation, though instead of
+defining custom tactics, we rely mostly on off-the-shelf tools such as
+Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.
+
+We use the Autosubst 2 framework to specify our syntax in HOAS and
+produce Coq syntax files in de Bruijn representation. Additionallly,
+Autosubst 2 provides a powerful tactic \texttt{asimpl} that
+can be used to prove the equivalence of two terms constructed using
+the primitive operators provided by the framework. This greatly
+simplifies the reasoning about substitution in our development as
+almost all substitution related properties are immediately discharged
+by \texttt{asimpl} without having to manually prove or even directly invoke any
+substitution related lemmas. \scw{You are only referring to lemmas about
+how substitution relates to syntax, right? You still need to manually
+invoke lemmas about substitution and judgements.}
+
+For other automation tasks that are not specific to binding, we use
+the powerful \texttt{sauto} tactic provided by CoqHammer to write
+short and declarative proofs. For example, here is a one-line proof of
+the triangle property about parallel reduction, from which the diamond
+property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
+\scw{you have space, so repeat the mathematical statement of the lemma}
+\begin{minted}{coq}
+Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
+Proof.
+  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
+Qed.
+\end{minted}
+In prose, the triangle property can be proven by induction over the
+graph of the \texttt{tstar} function, which stands for the Takahashi
+translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
+\texttt{ctrs:Par} say that the proof involves inverting and
+constructing of the derivations of parallel
+reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
+automation tactic to use
+the reflexivity and congruence properties of parallel
+reduction as lemmas.
+\scw{What does \texttt{lq:on} and \texttt{hauto} mean?}
+
+The automation here not only gives us a proof that is shorter and more
+resiliant to changes in definition, but also gives useful
+documentation for readers who wish to understand how the underlying
+proof works since the automation tactic is guaranteed to not use
+lemmas or invert derivations that are not specified in the
+\texttt{use} or \texttt{inv} flag.
+
+\section{Related Work and Discussion}
+\label{sec:relatedwork}
+
+\scw{Maybe it would be useful to include a chart here, so that readers can
+  easily keep track of the features of the various languages.  i.e. which ones
+  include large eliminations? type-directed equivalence? impredicative prop?
+  inductive datatypes? what are their line counts?}
+
+\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
+\citet{barendregt:lambda-calculi-with-types} are some of the earlier
+works that establish metatheoretic results strong enough to derive
+consistency for dependently typed systems.
+\scw{Should give Martin L\"of credit for first consistency proof.}
+\scw{What techniques do these three proofs use? How is it different from
+your proof? What about Luo, which
+introduced a universe hierarchy? What about categories with families, or
+other ways of showing consistency?}
+Some of these techniques
+are still in use in more recent works. For example, the technique for
+proving strong normalization by \citet{geuvers1994short} is adapted by
+\citet{moon2021graded} to show the same property to a dependently
+typed system extended with modalities.
+
+\citet{decagda} mechanizes in Agda the decidability of type
+conversion rule for a dependently typed language with one predicative
+universe level and typed judgmental equality with function
+$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
+uses a Kripke-style logical relation parameterized over an
+type-directed equivalence relation satisfying certain
+properties. Their logical relation is defined using the
+induction-recursion scheme, which is available in Agda but not in Coq.
+\citet{martin-lof-a-la-coq} manages to encode the logical relation
+from \citet{decagda} in the predicative fragment of Coq using a
+special way of encoding induction recursion. Their work further
+extends the decidability of type conversion result from~\citet{decagda} to the decidability
+type checking of a bidrectional type system.
+
+\citet{anand2014towards} mechanizes the metatheory of
+Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
+extensional type theory with features such as dependent functions,
+inductive types, and a full universe hierarchy. \citet{nbeincoq}
+mechanizes the normalization-by-evaluation algorithm in Coq for a
+dependently typed language with one predicative universe, similar to
+\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
+\citet{anand2014towards} and \citet{nbeincoq} leverage the
+impredicative \texttt{Prop} sort of Coq to define the interpretation
+of dependent function types and thus are closely related to our
+mechanization. However, instead of explaining the purely inductive
+logical relation as a convoluted workaround to the lack of
+induction-recursion in Coq, we give a self-contained explanation of
+our logical relation.
+
+Finally, it is worth pointing out that all systems we have discussed
+so far builds a relational model for their logical relation. Recall
+that our logical relation takes the form $[[InterpR i A S]]$ where
+$[[S]]$ is a set of terms. In a relational model, each type is
+interpreted as a partial equivalence relation over terms rather than a
+set. A relational model is necessary for an extensional type theory
+such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
+use a relational model either to justify the function $\eta$-law or to
+derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
+law for function is baked into the untyped reduction
+relation and therefore a relational model is not needed.
+\scw{would it be difficult to extend your proof to be a PER?}
+
+The dependently typed systems we have discussed so far also vary
+greatly in expressiveness. For example, the Calculus of
+Constructions~\citep{CoC} and MLTT without universe levels are unable
+to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
+identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
+the mechanized systems we have discussed so far, only
+\citet{anand2014towards} has a full universe hierarchy.
+Our object language \lang{}, while small, has a decent coverage of
+features commonly seen in dependently typed languages, including
+dependent pattern matching, dependent function types, a full universe
+hierarchy, an intensional identity type. We hope features such as
+type-directed reduction (necessary for unit $\eta$-law) and
+impredicative sorts can be implemented in a similarly streamlined
+fashion and we will leave those extensions as part of our future work.
+
+% To end this section, we
+\scw{should this be a new subsection?}
+The final question we want to address is: why is our proof,
+even with the extension from Section~\ref{sec:extension}, so much
+shorter than the proofs from \citet{decagda, nbeincoq,
+  martin-lof-a-la-coq}?
+First, unlike developments that mechanize the correctness of a clever
+type conversion algorithm, we only show the existence of
+normal form for open and closed terms and state our properties in
+terms of the untyped small-step reduction relation. This removes a lot
+of scaffolding in the specification of our type system and our logical
+relation.
+
+This contrasts our development with
+\citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
+difficult property to prove due to the complexity of the equality
+judgment used to define the logical relation. In our system, the
+irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
+direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
+lemma name} which is
+straightforward after we prove the results about parallel
+reduction in Section~\ref{sec:spec}. In other words, we keep the
+the logical relation itself minimal at the cost of extra lemmas that can be proven
+through syntactic means independently from the logical relation.
+For example, as commented near the end of Section~\ref{sec:extension},
+our normalization property (Corollary~\ref{corollary:exbetanf})
+% can induce a decision procedure for type conversion by reducing two
+% terms to $\beta\eta$-normal form and check for their syntactic
+% equivalence. This decision procedure would be inefficient compared to
+% NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
+% to weak head normal and checks for the equivalence of subterms
+% recursively
+does not immediately induce an efficient decision procedure, but it is
+possible to recover an efficient algorithm by reasoning about untyped
+lambda terms.
+
+% the concision of
+% our proof can be largely attributed to the fact that we bake into the
+% logical relation only the minimum amount of information that is the
+% most conducive to the proof of the metatheoretic result.
+
+Another key to our proof is the very early establishment of the confluence
+property of the paralell reduction relation
+(Lemma~\ref{lemma:pardiamond}). This property is used in
+Lemma~\ref{lemma:interppreservation} to show that the interpretation
+of a type is preserved under evaluation. That is, if $[[InterpR i A
+S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
+In a system with typed directed reduction, the confluence result is
+harder to establish since confluence requires subject reduction, which
+circularly depends on $\Pi$-injectivity, a consequence of
+confluence~\citep{siles2012pure}. There are different workarounds to
+this problem. \citet{siles2012pure} proposes a syntactic approach by
+defining a type system where the confluence result is
+directly provable and later show that the system of interest is
+equivalent to the system with the confluence property.
+\citet{decagda} uses a relational model and the relational counterpart
+of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
+model is parameterized by a typed indexed equivalence relation and
+is reused to prove derive different properties in their
+development. Another perhaps simpler approach is to simply
+extend the judgmental equality with rules about
+$\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
+reduction to be proven independently from confluence and
+$\Pi$-injectivity can later be shown to be admissible. We find the
+approach from \citep{weirich:systemd} the most lightweight as it
+allows us to derive confluence early on even in systems with typed
+conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}
+
+In cases where the judgmental equality does not leverage type information heavily
+(e.g. $\eta$-law for the unit type), it might be possible to show that
+$\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
+$[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
+property holds (we believe this is likely true for both
+\citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
+terms of untyped parallel reduction to show the decidability of type
+conversion. Otherwise, to fully model the system, we need to bake
+typing information into our logical relation and we need to redefine
+our logical relation in Kripke-style so we can relax the scoping
+constraint in order to prove adequacy, though a relational model would
+still be unnecessary as long as the $\eta$ laws are baked into the
+typed directed relation. \scw{not sure I understand this last bit}
+
+Since confluence plays such as key role in our proof, we are uncertain
+how to generalize our technique to systems where confluence does not
+hold, and we will investigate those scenarios in our future work.
+\scw{What systems are you thinking of? Or do you mean more generally
+that this technique only works for languages where the definition of
+equality is based on joinability of parallel reduction?}
+
+% We are able to avoid the complexivity of a Kripke-style logical relation
+% by defining our logical relation in terms of untyped parallel
+% reduction, which does not impose any scoping constraint on the
+% terms.
+
+% Why is it that the justification of $\eta$ law
+% requires a Kripke-style relational model in all three developments
+% but not ours? Are we ``cheating'' by defining \lang{}'s equational
+% theory using an untyped conversion rule? Instead of answering the
+% questions one by one, we clarify the design of our type system and our
+% proof technique so the answers to those questions are self-evident.
+
+% The key to the shortness of
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+% Type soundness can be proven through a syntactic
+% approach~\citep{syntacticsoundness} as a corollary of two properties:
+% progress and preservation. % The syntactic type soundness proof
+% % varies in complexity depending on the underlying type
+% % system. For example, a type system that tracks information flow would
+% % require additional structural rules related to security levels. In
+% % this paper, we focus on one specific type of complexity: the
+% In Figure~\ref{fig:stlcsoundness}, we summarize the structure of the
+% syntactic type soundness proof for the simply typed lambda
+% calculus. Each lemma can be proven by structural induction over the
+% typing derivation, while using the previous established results as
+% lemmas for specific cases that do not immediately follow from the
+% induction hypothesis. If we make our language more complex by adding
+% full dependent type support, the overall structure remains almost
+% identical.
+
+% NbE in Coq
+
+%%
+%% The next two lines define the bibliography style to be used, and
+%% the bibliography file.
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{refs}
+
+\appendix
+
+\subsection{More detailed proofs}
+\begin{lemma}[Preservation of the logical relation]
+  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
+\end{lemma}
+\begin{proof}
+  We carry out the proof by induction over the derivation of $[[Interp
+  I i A S]]$.
+  \begin{description}
+  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
+    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
+    be $[[Void]]$ and the result trivially follows.
+  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
+  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
+      refl , a <=> b  }]]$ and, by inverting the derivation of
+    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
+    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
+    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
+    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
+      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
+    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
+      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
+    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
+    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
+    <=> b0]]$. The result then immediately follows from the fact that
+    coherence is an equivalence relation.
+  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
+    such that:
+    \begin{itemize}
+    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
+        a #) }]]$
+    \item $[[Interp I i A S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
+    \end{itemize}
+    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
+    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
+      a #) }]]$. By \rref{I-Pi}, it suffices to show:
+    \begin{itemize}
+    \item $[[Interp I i A0 S ]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
+    \end{itemize}
+    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
+    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
+    conditions follow immediately from the induction hypothesis.
+  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
+    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
+    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
+    property of parallel
+    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
+    $[[C]]$ such that $[[B0 => C]]$ and $[[B => C]]$. By the induction
+    hypothesis, we derive $[[Interp I i C S]]$ from $[[Interp I i
+    B0 S]]$. By \rref{I-Red} and
+    $[[B => C]]$, we conclude that $[[Interp I i B S]]$.
+  \end{description}
+\end{proof}
+
 %%

### 2024-02-21 paper/paper.tex
Add explanation for valuation

@@
 
-The logical relation we define in Figure~\ref{fig:logrel} does not
-include cases for variables. Likewise, for the base types such as
-boolean and equality, the output set $[[S]]$ contains only terms that
-evaluate to closed terms.
-
-\scw{What is the motivation for this section? Why do we want to generalize
-the logical relation?}
-To generalize our logical relation to open
-terms, we define the semantic typing judgment by closing the open
-terms with a substitution whose codomain consists of terms that
-respect the interpretation of the types from the context. The full
+% The logical relation we define in Figure~\ref{fig:logrel} does not
+% include cases for variables. Likewise, for the base types such as
+% boolean and equality, the output set $[[S]]$ contains only terms that
+% evaluate to closed terms.
+
+% \scw{What is the motivation for this section? Why do we want to generalize
+% the logical relation?}
+In this section, we define the notion of semantic typing based on the
+logical relation we have defined in Section~\ref{sec:logreldep} and
+prove the fundamental lemma, which states that syntactic typing
+implies semantic typing.
+
+The necessity of semantic typing as an extra layer of definition on
+top of the logic relation can be understood in simply typed
+languages~\citep{skorstengaard2019introduction, harpertait,
+  pierce2002types}. In our setting, attempting to show that $[[empty
+|- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ will fail in
+\rref{T-Abs}, where induction hypothesis is not helpful since the body
+of the lambda term is typed under a non-empty context. Through
+the definition of semantic typing, we can state a strengthened
+property that is actually provable.
+
+% To generalize our logical relation to open
+% terms, we define the semantic typing judgment by closing the open
+% terms with a substitution whose codomain consists of terms that
+% respect the interpretation of the types from the context.
+The full
 definitions of well-formed substitution ($[[rho |= G]]$), semantic
@@
 ($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
-\scw{explain all parts of this figure. it's pretty intimidating.}
+
+The $[[rho |= G]]$ judgment defines the semantic well-formedness of
+a substitution $[[rho]]$ with respect to a context
+$[[G]]$. Intuitively, for variable $[[i]]$ in the context (thus the
+restriction $[[i < | G |]]$), $[[rho i]]$ is a term that inhabits all
+possible interpretations of the type $[[G i]]$ from the
+context. However, since we are in a dependently typed system, there
+may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
+  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
+operation $[[up Suc i]]$ can be viewed as an artifact of the de Bruijn
+representation to lift the type $[[G i]]$ to the right scope before
+substitution can be carried out.
+
 In the
@@
 $\beta$-normal form through parallel reduction and are defined as
-follows.\scw{What does $wne$ stand for?}
+follows.
 \[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
 \[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
+We can read $[[wne a]]$ and $[[wn a]]$ as weakly normalizing to a neutral or a normal form respectively.
 
@@
 in Figure~\ref{fig:logrelopen}.
-
 % The idea of a term blocked from $\beta$ reduction is
@@
 % neutral terms.
-
 In the definition of the logical relation, we omit the rules for

### 2024-02-22 paper/paper.tex
Revisions to Sec 4-6

@@
 
-\scw{Need to explain the notations in the text. What are these operations?
-What do readers need to understand about them? Help me understand the figure}
+% \scw{Need to explain the notations in the text. What are these operations?
+% What do readers need to understand about them? Help me understand
+% the figure} \yl{resolved}
 \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
@@
 \label{sec:logrelproof}
+\yl{todo: rename all variables from i,j to n}
 \begin{figure}[h]
@@
 |- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ will fail in
-\rref{T-Abs}, where induction hypothesis is not helpful since the body
+\rref{T-Abs}, where the induction hypothesis is not helpful since the body
 of the lambda term is typed under a non-empty context. Through
@@
 
-The $[[rho |= G]]$ judgment defines the semantic well-formedness of
+The $[[rho |= G]]$ notation denotes the semantic well-formedness of
 a substitution $[[rho]]$ with respect to a context
@@
 possible interpretations of the type $[[G i]]$ from the
-context. However, since we are in a dependently typed system, there
+context.
+Since we are in a dependently typed system, there
 may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
@@
 substitution can be carried out.
-
-In the
+Finally, despite the for all quantification over the interpreted set
+$[[S]]$ and the universe level $[[j]]$ in the definition of $[[rho |=
+G]]$, we only need show that there exists some $[[j]]$ and $[[S]]$
+such that $[[InterpR j (G i < up Suc i > ) { rho } S]]$ and $[[rho i
+in S]]$. This idea is encapsulated in the second of the two structural
+law about $[[rho |= G]]$.
+\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\gamma$\_ok\_nil}}]
+  \label{lemma:rhowfempty}
+  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
+\end{lemma}
+\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
+  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
+  $[[rho .: a |= G ++ A]]$.
+\end{lemma}
+
+We refer to $[[G |= a : A]]$ as semantic well-typedness. Its definition says the term $[[a]]$ can be
+semantically typed $[[A]]$ under the context $[[G]]$ if for all
+substitutions $[[rho]]$ such that $[[rho |= G]]$, the type $[[A { rho }]]$ can be interpreted
+as the set $[[S]]$, and $[[a { rho } in S]]$. Our definition of
+semantic well-typedness is standard, though dependent types add a
+small twist that we apply the $[[rho]]$ to $[[A]]$ and require that
+$[[A { rho }]]$ has some interpretation, which is not always the case
+for arbitrary types.
+
+Analogous to $[[|- G]]$, the syntactic well-formedness of the typing
+context, $[[|= G]]$ defines when a context is semantically
+well-formed. In the
 definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
@@
 $[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
-returns the empty list.
+returns the empty list. Intuitively, the definition of $[[|= G]]$
+simply says all types in the context must correspond to some set
+according to the logical relation. The truncation to the context
+$[[G]]$ is due to the different scoping of the types in the
+context. An alternative way of defining $[[|= G]]$ would look more
+similar to $[[|- G]]$, though we find it easier to define $[[|= G]]$
+through for all quantification of $[[i]]$ and recover the structural
+rules as lemmas.
+\begin{lemma}[Semantic context well-formedness cons]
+  \label{lemma:semwffcons}
+  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
+\end{lemma}
+\begin{lemma}[Semantic context well-formedness empty]
+  \label{lemma:semwffempty}
+  $[[|= G]]$ holds when $[[G]]$ is empty.
+\end{lemma}
 
@@
 work with.
-\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}\scw{spell out the whole word}]
+\begin{lemma}[Set Inversion\footnote{\dotv{soundness.v}{SemWt\_Univ}}]
   \label{lemma:setinv}
@@
 
-The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
-counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
-than inductively over the context. \scw{why is that the case?} It is easy to recover the same
-structural rules:
-\begin{lemma}[Semantic context well-formedness cons]
-  \label{lemma:semwffcons}
-  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
-\end{lemma}
-\scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
-Ctx-empty?}
-\begin{proof}
-  By the definition of semantic context well-formedness, the goal is
-  to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i
-  (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
-  proven by case analysis on whether $[[i]]$ is zero.
-\end{proof}
-
-Likewise, the semantic well-formendness judgment for substitutions
-satisfies similar structural rules.
-\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
-  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
-  $[[rho .: a |= G ++ A]]$.
-\end{lemma}
-\begin{proof}
-  Start by unfolding the definition of $[[rho .: a |= G ++ A]]$ and
-  performing a case analysis similar to the proof of
-  Lemma~\ref{lemma:semwffcons}.  The case where the number is $[[0]]$
-  requires Lemma~\ref{lemma:logreldeterhet} to finish the proof.
-\end{proof}
+% The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
+% counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
+% than inductively over the context. \scw{why is that the case?} It is easy to recover the same
+% structural rules:
+% \scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
+% Ctx-empty?}
+% \begin{proof}
+%   By the definition of semantic context well-formedness, the goal is
+%   to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i   (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
+%   proven by case analysis on whether $[[i]]$ is zero.
+% \end{proof}
 
 Next, we show some non-trivial cases of the fundamental theorem as
-top-level lemmas and leave the remaining cases as exercises for the
-reader.\scw{I dislike leaving ``exercises to the reader''}
-
-First, we formulate the definition of valid renamings and prove that
-semantic typing satisfies renaming so we can weaken the context when
-reasoning about the variable case of the fundamental lemma
-(Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G
-++ D]]$, it is easy to show that we can extract some valuation
-$[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
-``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :
-A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
-$[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
-after weakening the context; this implication holds because $[[rho |=
-G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
-can make use of the premise $[[G |= a : A]]$ to derive what we need
-for the conclusion. We recommend the readers to skip the proofs of
-Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
-during the first read as long as they have an intuitive understanding
-of what the renaming property is meant to capture.
-
-\scw{Make this a definition? Would it be useful to create notation
-  for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
-We say that $[[xi]]$ is
-valid from the context $[[G]]$ to the context $[[D]]$ if the
-following condition holds.
-\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
-  [[D ren xi
-  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]
-
-\begin{lemma}[Truncate is valid]
-  \label{lemma:validtruncate}
-  If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i
-  G]]$ to $[[G]]$.
-\end{lemma}
-\begin{proof}
-  By the definition of a valid renaming, we must show that given $[[j
-  < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
-  \begin{itemize}
-  \item $[[ren up i j]] = i + j  < [[| G | ]]$
-  \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
-  \end{itemize}
-  The first bullet point is immediate from the fact that $[[j]] <
-  [[|G|]] - [[i]]$. The second bullet follows from unfolding the
-  definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
-  $[[i]] + [[j]] < [[|G|]]$.
-\end{proof}
-
-\begin{lemma}[Renaming for $[[rho |= G]]$]
-  \label{lemma:renamingval}
-  If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
-  $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
-\scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
-\end{lemma}
-\begin{proof}
-  Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
-  that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i
-  < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.
-
-  By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
-have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho
-} ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |=
-D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
-S]]$ and $[[rho ren xi i in S]]$.
-\end{proof}
-
-\begin{lemma}[Renaming for $[[G |= a : A]]$]
-  \label{lemma:semrenaming}
-  If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
-  $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
-\end{lemma}
-\begin{proof}
-  Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
-\end{proof}
-
-\scw{Need a transition here that you are starting to explain the semantic typing
-rules.}
-
+top-level lemmas.
 \begin{lemma}[ST-Var]
@@
   From the definition of $[[|= G]]$, we know that there exists some
-  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. By
-  Lemma~\ref{lemma:validtruncate}, we know that $[[up Suc i]]$ is a
-  valid renaming from $[[drop Suc i G]]$ to $[[G]]$. By
-  Lemma~\ref{lemma:semrenaming}, we deduce $[[G |= G i < up Suc
-  i > : Set j]]$. By the identity from Lemma~\ref{lemma:setinv}, we
-  know that $[[InterpR j G i < up Suc i > { rho } S]]$. It now
+  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. This is not
+  immediately useful since $[[rho |= G]]$, not the truncated
+  context. We claim, without giving the proof, that from $[[drop Suc i
+  G |= G i  : Set j]]$, we can derive $[[G |= G i {up Suc i} : Set
+  j]]$. A rigorous justification would require us to formally state
+  and show renaming/weakening lemma for semantic typing. We instead
+  provide an informal justification since the proof of renaming
+  is uninteresting and only requires unfolding definitions, but its
+  statement introduces unnecessary line noise related to shifting.
+
+  Intuitively, a substitution $[[rho]]$
+  that closes over a context should always induce a truncated
+  substituion for each truncated context. As a result, if
+  some term $[[b]]$ is semantically well-typed under some truncated
+  context $[[G]]$, then it must also be semantically well-typed under
+  the context $[[G]]$ up to shifting.
+
+  Knowing that $[[G |= G i {up Suc i} : Set j]]$ holds, we can then
+  apply Lemma~\ref{lemma:setinv} to show that
+  $[[InterpR j G i < up Suc i > { rho } S]]$. It now
   suffices to show $[[rho i in S]]$, but that is immediate from the
@@
 \end{proof}
+
+% Next, we show some non-trivial cases of the fundamental theorem as
+% top-level lemmas.
+% We first formulate the definition of valid renamings and prove that
+% semantic typing satisfies renaming so we can weaken the context when
+% reasoning about the variable case of the fundamental lemma
+% (Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G ++ D]]$, it is easy to show that we can extract some valuation
+% $[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
+% ``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
+% $[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
+% after weakening the context; this implication holds because $[[rho |=G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
+% can make use of the premise $[[G |= a : A]]$ to derive what we need
+% for the conclusion. We recommend the readers to skip the proofs of
+% Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
+% during the first read as long as they have an intuitive understanding
+% of what the renaming property is meant to capture.
+
+% \scw{Make this a definition? Would it be useful to create notation
+%   for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
+% We say that $[[xi]]$ is
+% valid from the context $[[G]]$ to the context $[[D]]$ if the
+% following condition holds.
+% \[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
+%   [[D ren xi  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]
+
+% \begin{lemma}[Truncate is valid]
+%   \label{lemma:validtruncate}
+%   If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i   G]]$ to $[[G]]$.
+% \end{lemma}
+% \begin{proof}
+%   By the definition of a valid renaming, we must show that given $[[j   < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
+%   \begin{itemize}
+%   \item $[[ren up i j]] = i + j  < [[| G | ]]$
+%   \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
+%   \end{itemize}
+%   The first bullet point is immediate from the fact that $[[j]] <
+%   [[|G|]] - [[i]]$. The second bullet follows from unfolding the
+%   definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
+%   $[[i]] + [[j]] < [[|G|]]$.
+% \end{proof}
+
+% \begin{lemma}[Renaming for $[[rho |= G]]$]
+%   \label{lemma:renamingval}
+%   If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
+%   $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
+% \scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
+% \end{lemma}
+% \begin{proof}
+%   Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
+%   that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i   < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.
+
+%   By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
+% have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho } ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$.
+From $[[rho |= D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho } S]]$ and $[[rho ren xi i in S]]$.
+% \end{proof}
+
+% \begin{lemma}[Renaming for $[[G |= a : A]]$]
+%   \label{lemma:semrenaming}
+%   If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
+%   $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
+% \end{lemma}
+% \begin{proof}
+%   Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
+% \end{proof}
+
+% \scw{Need a transition here that you are starting to explain the semantic typing
+% rules.}
 
@@
   By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
-  under expansion. By Lemma~\ref{lemma:parabscbn}, since $[[( \A{rho} b{up rho} )
+  under expansion. Since $[[( \A{rho} b{up rho} )
   a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
@@
   shown or simpler and therefore omitted from the text.
-
 \end{proof}
 
-\scw{Spell this out in a little more detail.}
+Recall the logical consistency property
+(Theorem~\ref{theorem:consistency}), which states that the judgment
+$[[empty |- a : Void ]]$ is not derivable. We now give a proof of the
+property using the fundamental lemma.
 
 \begin{proof}
-  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
+  Suppose $[[empty |- a : Void]]$ is derivable, then by the
+  fundamental lemma, we have $[[empty |= a : Void]]$, which states
+  that for all $[[rho |= empty]]$, and for all $[[j]]$, $[[S]]$ such
+  that $[[InterpR j Void S]]$, we have $[[a {rho} in S]]$. By
+  Lemma~\ref{lemma:rhowfempty}, any $[[rho]]$ we pick trivially
+  satisfies $[[rho |= G]]$. For convenience, we pick $[[rho]]$ as
+  $[[idtm]]$ (though any $[[rho]]$ would work since
+  $[[empty |- a : Void]]$ ensures there is no free variable in
+  $[[a]]$), we have $[[a {idtm}]] = [[a in S]]$. By the $[[Void]]$
+  case of the inversion property (Lemma~\ref{lemma:interpinv}), we
+  know that $[[S]]$ must be the empty set and therefore contradicts
+  the assupmtion that $[[a in S]]$. Done.
 \end{proof}
@@
 
-\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
-equational theory since coherence, the untyped relation used for type
+\Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
+equational theory since convertibility ($[[a <=> b]]$), the untyped relation used for type
 conversion, is built on top of parallel reduction. We can recover the
@@
 \end{lemma}
+Both the confluence and anti-renaming properties require structural
+induction over the size of terms to obtain a useful induction
+hypotehsis in the $\eta$ case.
 
@@
 \[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
-We can read $[[wne a]]$ and $[[wn a]]$ as weakly normalizing to a neutral or a normal form respectively.
+We can read $[[wne a]]$ and $[[wn a]]$ as the term $[[a]]$ weakly normalizes to a neutral or a normal form respectively.
 
@@
 
-The updated logical relation and its auxiliary definitions are shown
+The updated logical relation is shown
 in Figure~\ref{fig:logrelopen}.
@@
 normalization property as its corollary, we need to additionally
-formulate and prove the adequacy property.
+formulate and prove the adequacy properties.
 
@@
   \label{lemma:extwn}
-  If $[[wn a var i]]$, then $[[wn a]]$.
+  If $[[wn (a var i)]]$, then $[[wn a]]$.
 \end{lemma}
 \begin{proof}
-  By induction over the length of the reduction sequence in $[[wn a
-  var i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
+  By induction over the length of the reduction sequence in $[[wn (a
+  var i)]]$. The conclusion follows from Lemma~\ref{lemma:parantirenaming} and
   \ref{lemma:parnenf}.
@@
 
-\scw{What does CR stand for?}
+To state the adequacy lemmas, it is useful to first define the notion
+of reducibility candidates (shortened as $CR$), predicates over sets
+of lambda terms that capture the properties we need from our logical
+relation.
 \begin{definition}[CR]
@@
 
-\begin{lemma}[$CR_1$ for type]
+\begin{lemma}[Adequacy: $CR_1$ for type]
   \label{lemma:cr1ty}
@@
 
-\begin{lemma}[$CR$ for interpreted sets]
+\begin{lemma}[Adequacy: $CR$ for interpreted sets]
   \label{lemma:crel}
@@
 
-\begin{lemma}[$CR_2$ for type]
+\begin{lemma}[Adequacy: $CR_2$ for type]
   \label{lemma:cr2ty}
@@
 
-\begin{lemma}[$CR_2$ for types (Rec)]
+\begin{lemma}[Adequacy: $CR_2$ for types (Rec)]
   \label{lemma:cr2tyrec}
@@
 
-\begin{lemma}[$CR$ for terms (Rec)]
+\begin{lemma}[Adequacy: $CR$ for terms (Rec)]
   \label{lemma:crelrec}
@@
 \end{proof}
-From Corollary~\ref{corollary:exbetanf}, we can show that there exists
-$\beta\eta$-normal forms for well-typed terms, since $\eta$ reduction
-preserves $\beta$-normal form (implied by Lemma~\ref{lemma:parnenf})
-and also strictly decreases the size of the term.
+\begin{corollary}[Existence of $\beta\eta$-normal form]
+\label{corollary:exbetaeta}
+If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
+\end{corollary}
+\begin{proof}
+  From Corollary~\ref{corollary:exbetanf}, we can easily conclude the
+  existence of $\beta\eta$-normal forms for well-typed terms since
+  $\eta$-reduction preserves $\beta$-normal form (implied by
+  Lemma~\ref{lemma:parnenf}) and also strictly decreases the size of
+  the term (evident from \rref{P-AbsEta}).
+\end{proof}
 
@@
 leftmost-outermost reduction can always find the $\beta\eta$-normal
-form if there exists one. However, we omit such proofs since they can
-can be formulated on untyped lambda terms and thus orthogonal to the
-specifics of dependently typed systems. Instead, we redirect readers
-to \citet{factorization-essentially, takahashi-parallel-reduction} for
-the details.
+form if there exists one. Then the termination of that strategy
+immediately follows from Corollary~\ref{corollary:exbetaeta} However, we omit such proofs since they can
+can be formulated on untyped lambda terms and thus are orthogonal to
+the specifics of dependently typed systems. Instead, we redirect
+readers to \citet{factorization-essentially,
+takahashi-parallel-reduction} for the details.
 
@@
 Finally, we want to point out that
-Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
-\ref{lemma:crelrec}, often referred to as adequacy of the logical
-relation, are in fact required in the normalization proof for simply
-typed languages~\citep{abel2019poplmark}. Similar to the simply typed
-scenario, adequacy needs to be proven before the fundamental theorem
-so we can handle rules such as \rref{T-If} where the scrutinee is a
-neutral term. In \citet{abel2019poplmark}, a variant of
-Lemma~\ref{lemma:extwn} is used in the exact same way to show that
-lambda terms are themselves normalizing as we have done in
-Lemma~\ref{lemma:crel}.
-
-Dependent types make the proof slightly more complicated
-as we also need to know that every type has a normal form. Whereas the
+the adequacy properties (Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
+\ref{lemma:crelrec}) are in fact required in the normalization proof
+for simply typed languages~\citep{abel2019poplmark}. Similar to the
+simply typed scenario, adequacy needs to be proven before the
+fundamental theorem so we can handle elimination rules such as
+\rref{T-If} and \rref{T-App} where the scrutinee is a neutral term. In
+\citet{abel2019poplmark}, a variant of Lemma~\ref{lemma:extwn} is used
+in the exact same way to show the normalization of lambda forms
+as we have done in Lemma~\ref{lemma:crel}.
+
+Dependent types make the normalization proof slightly more complicated
+as we also need to know that every type has a normal form.
+Whereas the
 $CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
@@
 requires the $CR_1$ property about the interpreted sets
-(Lemma~\ref{lemma:crel}).
+(Lemma~\ref{lemma:crel}) to be derived first.
 
@@
 technique for handling dependent types in our logical relation, the
-further extensions more or less boil down to properties that can be
-derived through syntactic means, which tend to not get in the way of
-Gödel's second incompleteness theorem.
+extension to open terms mostly boil down to properties that can be
+independently derived from the logical relation through syntactic
+means.
 
@@
     \hline
-    Library (Autosubst 2)  & 491 & - \\
+    % Library (Autosubst 2)  & 491 & - \\
     Syntactic typing (specification) &  69 & - \\
-    Renaming (common)  & 46 & -  \\
-    Syntactic soundness & 660 &  - \\
+    Renaming & 46 & -  \\
+    % Syntactic soundness & 660 &  - \\
     Untyped reduction & 350 & 834 \\
@@
 
-The powerset $[[PowerSet STm]]$ is encoded as the type \texttt{tm ->
+% The inductive
+% definition of the logical relation in Figure~\ref{fig:logrel} requires
+% the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
+% the function $[[F]]$ can be later instantiated into the logical
+% relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
+
+In our proof presented in Section~\ref{sec:logreldep}, the relation
+$[[R]]$ in \rref{I-Pi} requires Coq's impredicative \texttt{Prop} sort
+since $[[R]]$ can later be instantiated into something defined in
+terms of the logical relation itself (e.g. in
+Lemma~\ref{lemma:piintroalt}). However, in the mechanized proof, the
+definition of $[[Interp I i A S]]$ can in fact be encoded without the
+use of impredicativity by consistently replacing the use of
+\texttt{Prop} with Coq's predicative sort \texttt{Type}. One can even
+use this \texttt{Type} variant of the logical relation to define
+interpretation for any finite number of universes. The lack of
+impredicativity becomes troublesome only when we attempt to define
+$[[InterpR i A S]]$, the top-level logical relation defined in
+Figure~\ref{fig:logrelrec} that recursively calls itself at smaller
+universe levels. As a result, the one feature that truly requires
+impredicativity from \lang{} is its countable universe
+hierarchy, though lemmas such as the admissibility of \rref{I-PiAlt}
+(Lemma~\ref{lemma:piintroalt}) not only simply our mechanization, but
+also give better intuition on what our logical relation means.
+
+The definition of $[[Interp I i A S]]$ has an almost one-to-one
+correspondence to the Coq definition. The main difference is the
+specification of $[[I]]$. In the presentation in
+Section~\ref{sec:logreldep}, we define $[[I]]$ as a function over
+numbers less than $[[i]]$, the universe level. In Coq, we only require
+$[[I]]$ to be function with the set of natural numbers as its domain.
+In the Coq encoding of $[[InterpR i A S]]$ (Figure~\ref{fig:logrelrec}), we define $[[I]]$ as follows.
+\begin{equation*}
+  \begin{split}
+    [[I]]   &\in [[SNat -> PowerSet STm]] \\
+    [[I j]] &= [[{A | exists S , InterpR j A S}]] \text{, if } j < i \\
+    [[I j]] &= [[emptyset]] \text{, } j \geq i
+  \end{split}
+\end{equation*}
+Since $[[I]]$ is only applied to numbers strictly less than $[[i]]$ in
+\rref{I-Set}, we can retroactively show that the set we return in the $j
+\geq i$ case is junk data that does not affect the result of the logical
+relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Figure~\ref{fig:logrelrec}.
+
+Finally, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$
+are equal, it suffices to show the extensional property that $\forall
+x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
+occassionally in our presented proofs.
+In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm ->
 Prop}, a predicate over \lang{} terms.
-The inductive
-definition of the logical relation in Figure~\ref{fig:logrel} requires
-the impredicativity of Coq's \texttt{Prop} sort since\scw{or ``so that''?} in \rref{I-Pi},
-the function $[[F]]$ can be later instantiated into the logical
-relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
-
-In Coq, there is a distinction between computable functions and
-relations that can later be proven to be functional. The former can be
-viewed as a strict subset of the latter in axiom-free Coq. To be more
-precise, given a relation \texttt{R : A -> B -> Prop} subject to the
-totality and functionality constraints ($\forall$ \texttt{a} $\in$
-\texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
-that \texttt{R a b} is inhabited), we do not immediately obtain a function
-\texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
-\texttt{R a (F a)} is inhabited. However, the functional side
-conditions of a relation is clunky to express and tend to block
-automation. A simple workaround is to assume the axiom of unique choice, which
-is known to be consistent with Coq and allows us to induce a function \texttt{F
-: A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
-is functional. This approach would make our Coq development match
-the text version of our proof from Section~\ref{sec:logrelproof} more
-closely.
-
-However, we choose instead an axiom-free workaround and define
-\rref{I-Pi} as follows in our Coq mechanization.
-\begin{center}
-  \drule[]{I-PiCoq}
-\end{center}
-It should be easy to verify that the preconditions of
-\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
-equivalent. After establishing Lemma~\ref{lemma:logreldeter}, it is
-possible to further show that the conclusions of the rules are
-equivalent, too.\scw{If it is easy, you should have already done it.
-Unless it is really long and boring.}
+In axiom-free Coq, predicates do not come with the extensionality
+property. Given two predicates $P$ and $Q$, we can prove that $\forall
+x, P(x) \iff Q(x)$, but we can not concldue that $P = Q$. While
+predicate extensionality is not necessary in our development, we
+assume it as an axiom since it is known to be consistent with Coq's
+metatheory and helps bridge the gap between our mechanization and the
+proofs we present.
+
+% In Coq, there is a distinction between computable functions and
+% relations that can later be proven to be functional. The former can be
+% viewed as a strict subset of the latter in axiom-free Coq. To be more
+% precise, given a relation \texttt{R : A -> B -> Prop} subject to the
+% totality and functionality constraints ($\forall$ \texttt{a} $\in$
+% \texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
+% that \texttt{R a b} is inhabited), we do not immediately obtain a function
+% \texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
+% \texttt{R a (F a)} is inhabited. However, the functional side
+% conditions of a relation is clunky to express and tend to block
+% automation. A simple workaround is to assume the axiom of unique choice, which
+% is known to be consistent with Coq and allows us to induce a function \texttt{F
+% : A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
+% is functional. This approach would make our Coq development match
+% the text version of our proof from Section~\ref{sec:logrelproof} more
+% closely.
+
+% However, we choose instead an axiom-free workaround and define
+% \rref{I-Pi} as follows in our Coq mechanization.
+% \begin{center}
+%   \drule[]{I-PiCoq}
+% \end{center}
+% It should be easy to verify that the preconditions of
+% \rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
+% equivalent. After establishing Lemma~\ref{lemma:logreldeter}, it is
+% possible to further show that the conclusions of the rules are
+% equivalent, too.\scw{If it is easy, you should have already done it.
+% Unless it is really long and boring.}
 
@@
 % otherwise.
-We choose to keep this discrepancy between the Coq development and the
-description of the logical relation presented in Section~\ref{sec:logreldep} since the
-skolemization process is more intuitively expressed in terms of
-function symbols rather than relation symbols. Otherwise, we do not
-see a clear advantage of \rref{I-PiCoq} over \rref{I-Pi} in set
-theory, where there is no distinction between computable functions and
-functions in general. \scw{not sure I follow this last bit}
+% We choose to keep this discrepancy between the Coq development and the
+% description of the logical relation presented in Section~\ref{sec:logreldep} since the
+% skolemization process is more intuitively expressed in terms of
+% function symbols rather than relation symbols. Otherwise, we do not
+% see a clear advantage of \rref{I-PiCoq} over \rref{I-Pi} in set
+% theory, where there is no distinction between computable functions and
+% functions in general. \scw{not sure I follow this last bit}
 
@@
 simplifies the reasoning about substitution in our development as
-almost all substitution related properties are immediately discharged
-by \texttt{asimpl} without having to manually prove or even directly invoke any
-substitution related lemmas. \scw{You are only referring to lemmas about
-how substitution relates to syntax, right? You still need to manually
-invoke lemmas about substitution and judgements.}
+almost all substitution related properties about the syntax are
+immediately discharged by \texttt{asimpl}.
 
@@
 property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
-\scw{you have space, so repeat the mathematical statement of the lemma}
+The triangle property states
+that if $[[a => b]]$, then $[[b]]\Rightarrow [[a]]^*$, where $[[a]]^*$
+is the Takahashi translation~\citep{takahashi-parallel-reduction}
+which roughly corresponds to simultaneous reduction of the redexes in
+$[[a]]$, excluding the new redexes that appear as a result of
+reduction.
 \begin{minted}{coq}
@@
 In prose, the triangle property can be proven by induction over the
-graph of the \texttt{tstar} function, which stands for the Takahashi
-translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
+graph of \mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par} and
 \texttt{ctrs:Par} say that the proof involves inverting and
@@
 reduction as lemmas.
-\scw{What does \texttt{lq:on} and \texttt{hauto} mean?}
-
-The automation here not only gives us a proof that is shorter and more
-resiliant to changes in definition, but also gives useful
+The flag \texttt{lq:on} toggles tunes CoqHammer's search
+algorithm, though we never need to specify those options
+manually. Instead, we first invoke the \texttt{sauto} tactic indirectly through the
+\texttt{best} tactic provided by CoqHammer where we specify only the
+\texttt{inv}, \texttt{ctrs}, and lemmas we want to use. The
+\texttt{best} tactic iterates through possible configurations and
+provides us a replacement to the \texttt{best} tactic with the
+tuned performance flags to save time for future re-execution of the
+tactic.
+
+The automation provided by CoqHammer not only gives us a proof that is
+shorter and more resiliant to changes, but also gives useful
 documentation for readers who wish to understand how the underlying

### 2024-02-22 paper/paper.tex
Minor comments

@@
 relations-based arguments are not available.  \scw{Well, actually, we did use
-  a step-indexed relation for Trellys\ldots} But this is not the case. Recent
+  a step-indexed relation for Trellys\ldots}\yl{The core calculus of Trellys' logical fragment
+  is simple type + identity so I don't think that counts..} But this is not the case. Recent
 authors have developed tour-de-force mechanizations for the metatheory of
@@
 approach. \scw{Say that we don't have impredicativity, but neither does anyone
-  else?}
+  else?}\yl{Nobody else but Barras. I'll find the citation}
 %%

### 2024-02-22 paper/paper.tex
Add a conclusion

@@
 \label{sec:conclusion}
+In this work, we show a short and mechanized proof of consistency
+through proof by logical relation for a fully dependently typed
+language with a full universe hierarchy, an intensional identity type,
+and large elimination. We show the extensibility of our approach by
+proving the existence of $\beta\eta$-normal forms with only small
+and mechanical changes to our proof development. Our Coq mechanization
+leverages existing Coq libraries for reasoning about metatheory and
+general-purpose automation, allowing us to significantly reduce the
+verbosity typically associated with mechanized proofs and recover a
+declarative proof style that is usually only available in pen and
+paper. Our development shows that proof by logical relation for
+dependent types should be more accessible since it does not always
+require months of effort to implement. We hope our proof can inspire
+researchers to more actively mechanize results such as consistency,
+normalization for dependent type theory.
 

### 2024-02-22 paper/paper.tex
Address some comments

@@
 
- \scw{Give an overview of the figure}
-\scw{Define a convention that separates your use of natural numbers. Sometimes they
-are universe levels and sometimes they are de Bruijn indices. Better to have separate
-metavariables for each. (And it is possible to generalize universe levels to structures other than natural numbers. Ask Jonathan.}
+%  \scw{Give an overview of the figure}
+% \scw{Define a convention that separates your use of natural numbers. Sometimes they
+% are universe levels and sometimes they are de Bruijn indices. Better to have separate
+% metavariables for each. (And it is possible to generalize universe
+% levels to structures other than natural numbers. Ask Jonathan.} \yl{resolved?}
 As a dependent type theory, terms and types are collapsed into the same
@@
 % the figure} \yl{resolved}
-\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
-indices work.}
+% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
+% indices work.} \yl{resolved?}
 % We omit most of the
@@
 
-\scw{motivate parallel reduction. Why are you telling us this now. Is this
-  an explanation of definitional equality and its properties?}
+% \scw{motivate parallel reduction. Why are you telling us this now. Is this
+%   an explanation of definitional equality and its properties?}
 Before we specify the typing rules, we need to first specify its
@@
 over type structure, or step-indices. You use an inductive relation instead, why?}
-\scw{Can we view this inductive relation as the graph of the partial function
-that is defined recursively over types?}
-\scw{What is this form extensible too? impredicative quantification? recursive
-types? }
+% \scw{Can we view this inductive relation as the graph of the partial function
+% that is defined recursively over types?} \yl{reolsved}
+% \scw{What is this form extensible too? impredicative quantification? recursive
+% types? } \yl{resolved? impredicativity is hard}
 \scw{Is it worth observing here that this definition is not over sets of typed
@@
 (Lemma~\ref{lemma:logrelcumulativity}), and the backward closure
-property (Lemma~\ref{lemma:logrelNbackclos}).
+property (Lemma~\ref{lemma:logrelbackclos}).
 
@@
 to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
-identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
+identity types. % \citet{nbeincoq} lacks $\eta$-law for functions.
+Among
 the mechanized systems we have discussed so far, only

### 2024-02-22 paper/paper.tex
Minor wording

@@
 property $[[forall a, (# a in S implies Interp I i B { a } F a #) ]]$, it
-is possible to go in the other direction and define a relation $[[R]]=[[
+is possible to  define a relation $[[R]]=[[
 { ( a , F a )  | a in S }]]$. It is easy to verify that $[[R]]$ satisfy
@@
 
-Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
-positivity constraint required in proof assistants, but
-is genuinely non-monotone when written as an endofunction over the
+Unfortunately, we cannot define the function case of our logical
+relation directly using \rref{I-PiAlt} since it not only violates the
+syntactic strict positivity constraint required in proof assistants,
+but is genuinely non-monotone when written as an endofunction over the
 domain of relations.

### 2024-02-22 paper/paper.tex
minor wording in intro

--- 2024-02-22 paper/paper.tex
@@
 \section{Introduction}
-In this paper, we present a \emph{short}, \emph{mechanized} proof of logical
-consistency for \lang{}, a Martin-Löf style dependent type theory with large
-eliminations, a full predicative universe hierarchy, an intensional identity
-type and a boolean base type, both of which are equipped with dependent
-elimination forms.
+In this paper, we present a \emph{short} and \emph{mechanized} proof of
+logical consistency for \lang{}, a dependent type theory with a full
+predicative universe hierarchy, large eliminations, an intensional identity
+type, a boolean base type, and dependent elimination forms.
 
 Our goal with this work is to demonstrate how the proof technique of
-\emph{logical relations} can be applied in dependently-typed setting.
+\emph{logical relations} can be applied to reason about dependent type theories.
 Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
-languages with simple or polymorphic types. In that context, logical relations
+systems with simple or polymorphic types. In that context, logical relations
 can be define as simple recursive functions over the structure of types, or
@@
 Neither of these techniques make sense for predicative dependent type theory,
-so a young researcher might be excused for thinking that logical
-relations-based arguments are not available.  \scw{Well, actually, we did use
-  a step-indexed relation for Trellys\ldots}\yl{The core calculus of Trellys' logical fragment
-  is simple type + identity so I don't think that counts..} But this is not the case. Recent
-authors have developed tour-de-force mechanizations for the metatheory of
-modern proof
+so a young researcher might be excused for thinking that proofs that use
+logical relations are not applicable.  \scw{Well, actually, we did use a
+  step-indexed relation for Trellys\ldots}\yl{The core calculus of Trellys'
+  logical fragment is simple type + identity so I don't think that counts..}
+But this is not the case. Recent authors have developed tour-de-force
+mechanizations for the metatheory of modern proof
 assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
-have used logical-relations-based reasoning as part of their
-arguments. However, at the same time, these definitions have been overshadowed
-by their magnitude of the surrounding developments. Because these proofs show
-diverse results about real systems and algorithms, these developments range in
-size from 20,000 to 400,000 lines of code, and the techniques are not
-accessible to a casual reader.
+have relied on logical relations defined as part of their
+developments. However, because these proofs show diverse results about real
+systems and algorithms, these developments range in size from 20,000 to
+400,000 lines of code, these definitions are easy to miss and difficult to
+isolate from their surrounding contexts. As a result, they are not accessible
+to casual readers.
 

### 2024-02-23 paper/paper.tex
Typo
remove eta from section 5

@@
   $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
-  for all  $[[B]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
+  for all  $[[B1]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
   S]]$. By the diamond property of parallel reduction
@@
 
-\section{Toward Decidability of Type Conversion and $\eta$ laws}
+\section{Existence of $\beta$-normal forms}
 \label{sec:extension}
 In this section, we sketch out how the logical relation from
-Section~\ref{sec:logreldep} can be extended show the existence of
-$\beta\eta$ normal forms for (open and closed) well-typed terms. We
-first extend parallel reduction to include $\eta$ reduction for
-functions.
-\begin{center}
-  \drule[width=2.5in]{P-AbsEta}
-\end{center}
-
-\Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
-equational theory since convertibility ($[[a <=> b]]$), the untyped relation used for type
-conversion, is built on top of parallel reduction. We can recover the
-same confluence result about parallel reduction using the standard
-techniques from \citet{barendregt:lambda-calculi-with-types,
-takahashi-parallel-reduction}, though specifically for the de Bruijn
-representation, we need the following anti-renaming lemma about
-parallel reduction.
-\begin{lemma}[Par anti-renaming]
-  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
-exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
-\end{lemma}
-Both the confluence and anti-renaming properties require structural
-induction over the size of terms to obtain a useful induction
-hypotehsis in the $\eta$ case.
+Section~\ref{sec:logreldep} can be extended to show the existence of
+$\beta$ normal forms for (open and closed) well-typed terms.
+\scw{Why is this a sketch? We've done the whole proof and we will
+explain it all, right?}
+
+The goal of this section is to prove that all well-typed terms have
+$\beta$-normal forms. In other words, we can repeatedly use our parallel
+reduction relation to reduce any term to a unique normal form, where no
+further reductions can be applied. This result can be used to show that our
+type conversion relation is decidable, as computation of the normal form will
+always terminate.
 
@@
   \[
-    \begin{array}{lcl}
-      % \beta\text{-}\mathit{neutral\ terms}\\
+    \begin{array}{llcl}
+       \beta\text{-}\mathit{neutral\ terms} &
       [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                     f]] \\ \\
-      % \beta\text{-}\mathit{normal\ terms} \\
+      \beta\text{-}\mathit{normal\ terms} &
       [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
                     |\ [[f ~ f : f]]\\
-            & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
+            & & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
     \end{array}
@@
 
-The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the neutral terms
-and normal forms with respect to $\beta$-reduction, but can still take
-$\eta$ reduction steps. We sometimes use the judgment forms $[[ne a]]$
-and $[[nf a]]$ to indicate that there exists $[[e]]$ or $[[f]]$ such
-that $[[a = e]]$ or $[[a = f]]$.
-
-We can show that parallel reduction preserves $\beta$-normal and
-neutral forms.
-\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
-  \label{lemma:parnenf}
-  If $[[a => b]]$, then
-  \begin{itemize}
-  \item $[[ne a]]$ implies $[[ne b]]$
-  \item $[[nf a]]$ implies $[[nf b]]$
-  \end{itemize}
-\end{lemma}
-This lemma is mainly to show that \rref{P-AbsEta} does not introduce
-any new $\beta$ redex because $\eta$ reduction is the only real step
-we can take.
+The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
+neutral terms and normal forms with respect to $\beta$-reduction. We sometimes
+use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
+$[[e]]$ or $[[f]]$ such that $[[a = e]]$ or $[[a = f]]$.
 
@@
 \end{proof}
-\begin{corollary}[Existence of $\beta\eta$-normal form]
-\label{corollary:exbetaeta}
-If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
-\end{corollary}
-\begin{proof}
-  From Corollary~\ref{corollary:exbetanf}, we can easily conclude the
-  existence of $\beta\eta$-normal forms for well-typed terms since
-  $\eta$-reduction preserves $\beta$-normal form (implied by
-  Lemma~\ref{lemma:parnenf}) and also strictly decreases the size of
-  the term (evident from \rref{P-AbsEta}).
-\end{proof}
-
-Due to the non-deterministic nature of parallel reduction, we need to
-take a few more steps to convert the existence of $\beta\eta$-normal
-form into a decision procedure for type conversion. More specifically,
-we can show that a deterministic evaluation strategy such as
-leftmost-outermost reduction can always find the $\beta\eta$-normal
-form if there exists one. Then the termination of that strategy
-immediately follows from Corollary~\ref{corollary:exbetaeta} However, we omit such proofs since they can
-can be formulated on untyped lambda terms and thus are orthogonal to
-the specifics of dependently typed systems. Instead, we redirect
-readers to \citet{factorization-essentially,
+
+Due to the non-deterministic nature of parallel reduction, we need to take a
+few more steps to convert the existence of $\beta$-normal form into a decision
+procedure for type conversion. More specifically, we can show that a
+deterministic evaluation strategy such as leftmost-outermost reduction can
+always find the $\beta$-normal form if there exists one. Then the termination
+of that strategy immediately follows from Corollary~\ref{corollary:exbetanf}
+However, we omit such proofs since they can can be formulated on untyped
+lambda terms and thus are orthogonal to the specifics of dependently typed
+systems. Instead, we redirect readers to \citet{factorization-essentially,
   takahashi-parallel-reduction} for the details.
@@
 means.
+
+\section{Existence of $\beta\eta$-normal forms}
+\scw{relocated text from previous section. Not a complete description yet.}
+
+We first extend parallel reduction to include $\eta$ reduction for functions.
+\begin{center}
+  \drule[width=2.5in]{P-AbsEta}
+\end{center}
+
+\Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
+equational theory since convertibility ($[[a <=> b]]$), the untyped relation
+used for type conversion, is built on top of parallel reduction. We can
+recover the same confluence result about parallel reduction using the standard
+techniques from \citet{barendregt:lambda-calculi-with-types,
+  takahashi-parallel-reduction}, though specifically for the de Bruijn
+representation, we need the following anti-renaming lemma about parallel
+reduction.
+\begin{lemma}[Par anti-renaming]
+  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
+exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
+\end{lemma}
+Both the confluence and anti-renaming properties require structural
+induction over the size of terms to obtain a useful induction
+hypotehsis in the $\eta$ case.
+
+We can show that parallel reduction preserves $\beta$-normal and
+neutral forms.
+\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
+  \label{lemma:parnenf}
+  If $[[a => b]]$, then
+  \begin{itemize}
+  \item $[[ne a]]$ implies $[[ne b]]$
+  \item $[[nf a]]$ implies $[[nf b]]$
+  \end{itemize}
+\end{lemma}
+This lemma is mainly to show that \rref{P-AbsEta} does not introduce
+any new $\beta$ redex because $\eta$ reduction is the only real step
+we can take.
+
+\begin{corollary}[Existence of $\beta\eta$-normal form]
+\label{corollary:exbetaeta}
+If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
+\end{corollary}
+\begin{proof}
+  From Corollary~\ref{corollary:exbetanf}, we can easily conclude the
+  existence of $\beta\eta$-normal forms for well-typed terms since
+  $\eta$-reduction preserves $\beta$-normal form (implied by
+  Lemma~\ref{lemma:parnenf}) and also strictly decreases the size of
+  the term (evident from \rref{P-AbsEta}).
+\end{proof}
 

### 2024-02-23 paper/paper.tex
Merge adequacy lemmas into a single one

@@
 normalization property as its corollary, we need to additionally
-formulate and prove the adequacy properties.
+formulate and prove the adequacy property.
 
 Let us first start with some auxiliary lemmas and definitions.
-\begin{lemma}[Ext Wn]
+\begin{lemma}[Wn extensionality]
   \label{lemma:extwn}
@@
 
-To state the adequacy lemmas, it is useful to first define the notion
+To state the adequacy lemma, we first define the notion
 of reducibility candidates (shortened as $CR$), predicates over sets
@@
 
-\begin{lemma}[Adequacy: $CR_1$ for type]
-  \label{lemma:cr1ty}
-  If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
+\yl{Merged all the separate adequacy related properties into a single
+  property proved through nested induction. (The mechanization has
+  been adjusted accordingly)}
+We now state and prove the adequacy lemma. We give some more details
+since we prove it through nested induction.
+\begin{lemma}[Adequacy\footnote{\dotv{soundnessopen.v}{soundness}}]
+  \label{lemma:adequacy}
+  If $[[InterpR i A S]]$, then we have $[[S]] \in CR$.
 \end{lemma}
 \begin{proof}
-  Immediately from \rref{I-Ne} and \rref{I-Par}.
+  We start by strong induction over $[[i]]$. We are given the
+  induction hypothesis that for all $[[j < i]]$, $[[InterpR j A S]]$
+  implies $[[S]] \in CR$. Our goal is to show $[[InterpR i A S]]$
+  implies $[[S]] \in CR$.
+
+  By its definition in Figure~\ref{fig:logrelrec}, we have the
+  equality $[[InterpR i A S]] = [[Interp I i A S]]$  where $[[I i]] :=
+  [[{A | exists S , InterpR i A S}]]$.
+  We then proceed by structural over the derivation of $[[Interp I i A
+  S]]$. The only interesting cases are \rref{I-Pi} and \rref{I-Set}.
+  The function case requires Lemmas~\ref{lemma:extwn} and
+  \ref{lemma:wnewn} we have shown earlier.
+
+  The \rref{I-Set} case is the most involved. We need to show that
+  for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
+  \in CR$. We immediately know that $[[{A | exists S, InterpR j A S
+  }]] \in CR_1$ by \rref{I-Ne}. It remains to show that $[[{A | exists S, InterpR j A S
+  }]] \in CR_2$, or equivalently, for all $[[A]]$, $[[InterpR j A S]]$
+  implies $[[wn A]]$. Suppose $[[InterpR j A S]]$ for an arbitrary
+  $[[A]]$. We have $[[InterpR j A S]]=[[Interp I j A S]]$ where $[[I]]$
+  has the same definition from earlier but its domain restricted to
+  numbers less than $[[j]]$. We perform another induction on
+  $[[j]]$. All cases are trivial except for the case for
+  \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
+  A]]$. To derive $[[wn Pi A B]]$, it remains to show $[[wn B]]$. We
+  use the outermost induction hypothesis to show that $[[var 0]]$
+  semantically inhabits $[[A]]$, from which we derive $[[wn B {var 0}]]$
+  and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
 \end{proof}
 
-\begin{lemma}[Adequacy: $CR$ for interpreted sets]
-  \label{lemma:crel}
-  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
-\end{lemma}
-\begin{proof}
-  By induction over the derivation of $[[Interp I i A S]]$. The base
-  cases are all immediate.
-
-  In the function case ($[[Pi A B]]$), we use the induction
-  hypothesis to show that variables, which are special cases of
-  neutral terms, semantically inhabit the input type $[[A]]$. We apply the
-  function to an arbitrary variable and we can then use
-  Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.
-
-  To show
-  the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
-  such that $[[wne b]]$,
-  and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
-  us a term that semantically inhabits $[[B {a}]]$. From the induction
-  hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
-  holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
-  from the induction hypothesis, since every term that evaluates to
-  some neutral form semantically inhabits $[[B {a}]]$. Done.
-\end{proof}
-
-\begin{lemma}[Adequacy: $CR_2$ for type]
-  \label{lemma:cr2ty}
-  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
-\end{lemma}
-\begin{proof}
-  By induction over the derivation of $[[Interp I i A S]]$. The
-  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
-\end{proof}
-
-\begin{lemma}[Adequacy: $CR_2$ for types (Rec)]
-  \label{lemma:cr2tyrec}
-  If $[[InterpR i A S]]$, then $[[wn A]]$.
-\end{lemma}
-\begin{proof}
-  By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
-\end{proof}
-
-\begin{lemma}[Adequacy: $CR$ for terms (Rec)]
-  \label{lemma:crelrec}
-  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
-\end{lemma}
-\begin{proof}
-  After unfolding the definition of $[[InterpR i A S]]$, trivial by
-  Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
-\end{proof}
+% \begin{lemma}[Adequacy: $CR_1$ for type]
+%   \label{lemma:cr1ty}
+%   If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
+% \end{lemma}
+% \begin{proof}
+%   Immediately from \rref{I-Ne} and \rref{I-Par}.
+% \end{proof}
+
+% \begin{lemma}[Adequacy: $CR$ for interpreted sets]
+%   \label{lemma:crel}
+%   If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
+% \end{lemma}
+% \begin{proof}
+%   By induction over the derivation of $[[Interp I i A S]]$. The base
+%   cases are all immediate.
+
+%   In the function case ($[[Pi A B]]$), we use the induction
+%   hypothesis to show that variables, which are special cases of
+%   neutral terms, semantically inhabit the input type $[[A]]$. We apply the
+%   function to an arbitrary variable and we can then use
+%   Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.
+
+%   To show
+%   the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
+%   such that $[[wne b]]$,
+%   and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
+%   us a term that semantically inhabits $[[B {a}]]$. From the induction
+%   hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
+%   holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
+%   from the induction hypothesis, since every term that evaluates to
+%   some neutral form semantically inhabits $[[B {a}]]$. Done.
+% \end{proof}
+
+% \begin{lemma}[Adequacy: $CR_2$ for type]
+%   \label{lemma:cr2ty}
+%   If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
+% \end{lemma}
+% \begin{proof}
+%   By induction over the derivation of $[[Interp I i A S]]$. The
+%   function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
+% \end{proof}
+
+% \begin{lemma}[Adequacy: $CR_2$ for types (Rec)]
+%   \label{lemma:cr2tyrec}
+%   If $[[InterpR i A S]]$, then $[[wn A]]$.
+% \end{lemma}
+% \begin{proof}
+%   By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
+% \end{proof}
+
+% \begin{lemma}[Adequacy: $CR$ for terms (Rec)]
+%   \label{lemma:crelrec}
+%   If $[[InterpR i A S]]$, then $[[S]] \in CR$.
+% \end{lemma}
+% \begin{proof}
+%   After unfolding the definition of $[[InterpR i A S]]$, trivial by
+%   Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
+% \end{proof}
 
@@
 over the typing derivation, where the additional neutral term related
-cases are handled by Lemmas~\ref{lemma:cr2tyrec} and
-\ref{lemma:crelrec}.
+cases are handled by Lemma~\ref{lemma:adequacy}, the adequacy property.
 
@@
 
-Dependent types make the normalization proof slightly more complicated
-as we also need to know that every type has a normal form.
-Whereas the
-$CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
-derivable, the $CR_2$ property for types (Lemma~\ref{lemma:cr2ty})
-requires the $CR_1$ property about the interpreted sets
-(Lemma~\ref{lemma:crel}) to be derived first.
+Dependent types make the adequacy proof slightly more complicated
+as we also need to know that every type has a normal form, not just
+the terms.
+This complicates our proof specifically in the \rref{I-Set} case for
+Lemma~\ref{lemma:adequacy}.
 

### 2024-02-23 paper/paper.tex
Minor

@@
 \label{sec:extension}
-In this section, we sketch out how the logical relation from
+In this section, we show how the logical relation from
 Section~\ref{sec:logreldep} can be extended to show the existence of
@@
 \scw{Why is this a sketch? We've done the whole proof and we will
-explain it all, right?}
+explain it all, right?}\yl{s/sketch out/show. I used sketch because we
+didn't spell out the full proofs}
 

### 2024-02-23 paper/paper.tex
Move  anti renaming and par preservation back to section 5

@@
 \begin{minipage}{0.9\textwidth}
-\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse,
-  Bool, Eq, Refl, J, JRefl}
+\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
 \drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
@@
 \end{minipage}
-\caption{Parallel reduction and coherence}
+\caption{Parallel reduction ($\beta$-rules only) and convertibility }
 \label{fig:par}
@@
 equational theory in an algorithmic style using the notion of parallel
-reduction, defined in Figure~\ref{fig:par}. Parallel reduction takes
+reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted). Parallel reduction takes
 the form $[[a => b]]$ and we denote its transitive as $[[a =>+
@@
 
+\yl{Moved renaming back to Sec 5 because they are needed for the
+  non-eta development too}
 Let us first start with some auxiliary lemmas and definitions.
+\begin{lemma}[Par anti-renaming]
+  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
+exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
+\end{lemma}
+
+We can show that parallel reduction preserves $\beta$-normal and
+neutral forms.
+\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
+  \label{lemma:parnenf}
+  If $[[a => b]]$, then
+  \begin{itemize}
+  \item $[[ne a]]$ implies $[[ne b]]$
+  \item $[[nf a]]$ implies $[[nf b]]$
+  \end{itemize}
+\end{lemma}
+
 \begin{lemma}[Wn extensionality]
@@
   natural numbers as term variables. The side condition $[[idtm |=
-  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
+  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
   including variables, semantically inhabit any $[[S0]]$ where
@@
   know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
-  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
+  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
   conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
@@
 Finally, we want to point out that
-the adequacy properties (Lemmas~\ref{lemma:cr1ty},~\ref{lemma:cr2tyrec}, and
-\ref{lemma:crelrec}) are in fact required in the normalization proof
+the adequacy property (Lemma~\ref{lemma:adequacy}) are in fact required in the normalization proof
 for simply typed languages~\citep{abel2019poplmark}. Similar to the
@@
 reduction.
-\begin{lemma}[Par anti-renaming]
-  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
-exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
-\end{lemma}
+
 Both the confluence and anti-renaming properties require structural
@@
 
-We can show that parallel reduction preserves $\beta$-normal and
-neutral forms.
-\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
-  \label{lemma:parnenf}
-  If $[[a => b]]$, then
-  \begin{itemize}
-  \item $[[ne a]]$ implies $[[ne b]]$
-  \item $[[nf a]]$ implies $[[nf b]]$
-  \end{itemize}
-\end{lemma}
 This lemma is mainly to show that \rref{P-AbsEta} does not introduce

### 2024-02-23 paper/paper.tex
Update line count and minor rewording

@@
 fundamental theorem so we can handle elimination rules such as
-\rref{T-If} and \rref{T-App} where the scrutinee is a neutral term. In
-\citet{abel2019poplmark}, a variant of Lemma~\ref{lemma:extwn} is used
-in the exact same way to show the normalization of lambda forms
-as we have done in Lemma~\ref{lemma:crel}.
-
+\rref{T-If} and \rref{T-App} where the scrutinee is a neutral term. % In
+% \citet{abel2019poplmark}, a variant of Lemma~\ref{lemma:extwn} is used
+% in the exact same way to show the normalization of lambda forms
 Dependent types make the adequacy proof slightly more complicated
@@
 This complicates our proof specifically in the \rref{I-Set} case for
-Lemma~\ref{lemma:adequacy}.
+our adequacy property (Lemma~\ref{lemma:adequacy}).
 
@@
     % Library (Autosubst 2)  & 491 & - \\
-    Syntactic typing (specification) &  69 & - \\
+    Syntactic typing (specification) &  68 & - \\
     Renaming & 46 & -  \\
     % Syntactic soundness & 660 &  - \\
-    Untyped reduction & 350 & 834 \\
-    Logical relation & 342 & 515 \\
-    Semantic typing and soundness & 169 & 197 \\
+    Untyped reduction & 349 & - \\
+    Neutral and normal forms & N/A & 269 \\
+    Logical relation & 331 & 426 \\
+    Semantic typing and soundness & 169 & 199 \\
+    \hline
+    Total & 963 & 1357 \\
   \end{tabular}

### 2024-02-23 paper/paper.tex
comments on what we still need to say in the first three sections

@@
 computation.
+\scw{need one sentence to motivate why logical relations are important}
 
@@
 \label{sec:spec}
+
+\scw{This section needs to say:
+  \begin{itemize}
+\item definitional equality is untyped so that we can prove 
+  properties about it independent of the type system
+\item definitional equality is based on parallel reduction so that we can take
+  advantage of the algorithmic structure later. This definition is easier to
+  invert because there are fewer derivation. That is important when reasoning
+  about our logical relation.
+\item can prove that parallel reduction is equivalent to other definitions of
+  equality using standard means (cite Barendregt's book).  (we go halfway
+  there by showing it is an equivalence relation). Can also prove equivalent
+  to typed relation. (cite Siles, also CoreSpec?)
+\item All of the proofs in this section are standard, well known, and use
+  techniques that are well-suited for proof assistants. In fact, de Bruijn's
+  paper that introduces de Bruijn indices was part of a mechanized confluence
+  proof for the untyped lambda calculus.
+\item extensions: forward reference to eta equivalence for functions in later
+  section. type-directed equivalence is future work
+  \end{itemize}
+}
 
@@
 % show!) Should add more motivation here.}
+\scw{We need to explicitly point out that the key ideas of this paper are 
+  discussed, here, in this section. 
+  We need to explicitly remark on why logical relations are difficult to 
+  define for dependent type theory and explain why this setting is more 
+  difficult than with simple types (STLC) or with polymorphic types (System F).
+  \begin{itemize}
+  \item Dependent function types (result type mentions argument value, so can't define 
+    relation by recursion over type structure)
+  \item Definitional equality (not all types look like types)
+  \item Identity types(?)
+  \end{itemize}
+  Should we be more explicit in our comparison with Girard's trick for polymorphic type?
+  There, the definition stays recursive because it doesn't substitute for the variables 
+  in the function types. But that approach is not available in this setting, because not 
+  all quantified things are types. And we might need that information to interpret, say, 
+  identity types in the right way.
+}
+\scw{ We also need to explicitly point out that our logical relation is untyped. 
+  This has two benefits: it allows semantic typing to be meaningful independent from 
+  syntactic typing (cite Derek, forward reference to next section) and it avoids 
+  significant bookkeeping, especially in the case of Kripke logical relations (we need to define 
+  what these are).
+  Is there a cost to an untyped relation? 
+}
 Before we define our logical relation, we first formally specify the

### 2024-02-23 paper/paper.tex
another comment

@@
 
+\scw{The goal of this section is also to demonstrate that our logical relations 
+  proof technique can be extended to reason about the (parallel) reduction of 
+  open terms, not just the reduction of terms after closing 
+  substitutions. This is important, in particular, for dependently-typed languages 
+  because type checking involves reasoning about open terms. But even simple type 
+  systems may wish to employ this sort of reasoning, in the case of, say, reasoning 
+  about compiler correctness in open environments. (Does Amal do anthing like this?)
+}
+
 The goal of this section is to prove that all well-typed terms have
@@
 of lambda terms that capture the properties we need from our logical
-relation.
+relation. \scw{Need to provide intuition for these two properties that need 
+to be proved similtaneously. Also, cite where this form of proof originally 
+appeared? This part is standard.}
 \begin{definition}[CR]
@@
 files are counted toward the \emph{Library (Autosubst 2)} category.
+\scw{this line is no longer in the figure}
 

### 2024-02-23 paper/paper.tex
Revise related work

@@
 substitution functions, and lemmas and tactics that allow reasoning
-about those functions. The auto-generated files and the library header
-files are counted toward the \emph{Library (Autosubst 2)} category.
-\scw{this line is no longer in the figure}
+about those functions. The auto-generated syntax file by Autosubst 2
+and library header files are omitted from the figure.
 
@@
 
-\section{Related Work and Discussion}
+\section{Related Work}
 \label{sec:relatedwork}
@@
 
-\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
+Proof by logical relations does not come with a precise definition. In
+the most general sense, a logical relation can be viewed as a
+practical technique that uses a type-indexed relation to assign
+meanings to types and strengthen the induction hypothesis for the
+property of interest. The original idea of this technique can be
+traced back to
+\citet{tait1967:reducibility}. \citet{tait1967:reducibility} maps
+types to sets of terms satisfying certain properties related to reduction.
+The same idea is explained in \citet{girard1989proofs} and extended to
+prove the strong normalization property of System F.
+
+Tait's method has been successfully applied to dependently typed
+languages.  \citet{Martin-Lof-1973}, \citet{luo1990extended},
+\citet{geuvers1994short}, and
 \citet{barendregt:lambda-calculi-with-types} are some of the earlier
-works that establish metatheoretic results strong enough to derive
-consistency for dependently typed systems.
+works that prove strong
+normalization for different dependently type systems using Tait's notion of
+reducibility.
+
 \scw{Should give Martin L\"of credit for first consistency proof.}
@@
 other ways of showing consistency?}
-Some of these techniques
-are still in use in more recent works. For example, the technique for
-proving strong normalization by \citet{geuvers1994short} is adapted by
-\citet{moon2021graded} to show the same property to a dependently
-typed system extended with modalities.
+
+% Some of these proofs are still reapplied in more recent works. For
+% example, the technique for proving strong normalization by
+% \citet{geuvers1994short} is adapted by \citet{moon2021graded} to show
+% the normalization property for a dependently typed system extended with
+% modalities.
+
+The pen and paper representation of the logical relation proofs
+can be challenging to adapt to a theorem prover since a lot of details
+are hidden behind the concise notations.
+As an example, \citet{geuvers1994short} presents the interpretation for types as
+an inductively defined total function over the set of syntactically
+well-formed types. Despite the notation for the logical relation as a
+simply typed function that takes a type and returns some set, the
+interpretation function in reality is a dependent function whose
+return type depends on the derivation of the well-typedness of its
+input. The well-typedness derivation and the proof of the
+classification theorem are needed relevantly in the body
+of the interpretation function to decide whether an argument of an
+application should be erased during interpretation. Due to the
+impredicativity of the object language, \citet{geuvers1994short}'s
+proof cannot be encoded in Agda, which has a predicative
+metatheory. Due to the use of proof-relevant derivations, even in
+Coq, a proof assistant that supports impredicativity, one would need
+to constantly juggle between the impredicative but irrelevant sort
+\texttt{Prop} sort and the predicative but relevant sort
+\texttt{Type}.
+
+More recent works such as \citet{Abel12}
+and\citet{abel2008betaeta} make their definitions more explicit and
+precise and thus easier to encode in proof assistants. Our definition
+of logical relation is reminiscent to their definition of a
+semantic universe hierarchy, though our logical relation is defined to
+be closed under expansion with respect to parallel reduction rather
+than weak head reduction. Furthermore, \citet{Abel12} and
+\citet{abel2008betaeta} use the semantic universe hierachy as a
+measure to define Kripke-style logical relations, from which they
+derive the correctness of their conversion algorithms. In our work, we
+use the semantic universe hierarchy directly because it is sufficient
+for deriving consistency and later normalization.
 
@@
 universe level and typed judgmental equality with function
-$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
-uses a Kripke-style logical relation parameterized over an
+$\eta$-law. They
+use a Kripke-style logical relation parameterized over a
 type-directed equivalence relation satisfying certain
-properties. Their logical relation is defined using the
-induction-recursion scheme, which is available in Agda but not in Coq.
+properties to facillitate the reuse of their definition. The
+logical relation is defined using the induction-recursion scheme,
+which is available in Agda but not in Coq.
 \citet{martin-lof-a-la-coq} manages to encode the logical relation
-from \citet{decagda} in the predicative fragment of Coq using a
-special way of encoding induction recursion. Their work further
-extends the decidability of type conversion result from~\citet{decagda} to the decidability
-type checking of a bidrectional type system.
+from \citet{decagda} in the predicative fragment of Coq and further
+extends the decidability of type conversion result
+from~\citet{decagda} to the decidability type checking of a
+bidrectional type system.
 
@@
 of dependent function types and thus are closely related to our
-mechanization. However, instead of explaining the purely inductive
+mechanization. However, instead of explaining the inductive
 logical relation as a convoluted workaround to the lack of
@@
 
-Finally, it is worth pointing out that all systems we have discussed
-so far builds a relational model for their logical relation. Recall
-that our logical relation takes the form $[[InterpR i A S]]$ where
-$[[S]]$ is a set of terms. In a relational model, each type is
-interpreted as a partial equivalence relation over terms rather than a
-set. A relational model is necessary for an extensional type theory
-such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
-use a relational model either to justify the function $\eta$-law or to
-derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
-law for function is baked into the untyped reduction
-relation and therefore a relational model is not needed.
-\scw{would it be difficult to extend your proof to be a PER?}
-
-The dependently typed systems we have discussed so far also vary
-greatly in expressiveness. For example, the Calculus of
-Constructions~\citep{CoC} and MLTT without universe levels are unable
-to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
-identity types. % \citet{nbeincoq} lacks $\eta$-law for functions.
-Among
-the mechanized systems we have discussed so far, only
-\citet{anand2014towards} has a full universe hierarchy.
-Our object language \lang{}, while small, has a decent coverage of
-features commonly seen in dependently typed languages, including
-dependent pattern matching, dependent function types, a full universe
-hierarchy, an intensional identity type. We hope features such as
-type-directed reduction (necessary for unit $\eta$-law) and
-impredicative sorts can be implemented in a similarly streamlined
-fashion and we will leave those extensions as part of our future work.
-
-% To end this section, we
-\scw{should this be a new subsection?}
-The final question we want to address is: why is our proof,
-even with the extension from Section~\ref{sec:extension}, so much
-shorter than the proofs from \citet{decagda, nbeincoq,
-  martin-lof-a-la-coq}?
-First, unlike developments that mechanize the correctness of a clever
-type conversion algorithm, we only show the existence of
+Finally, \citet{barras2010sets, Wang2013SemanticsOI} assign
+set-theoretic semantics to dependent type theory in Coq. Unlike the
+previous mechanization efforts, which primarily focus on predicative
+type theory and more direct reducibility models,
+\citet{barras2010sets, Wang2013SemanticsOI} tackle extensions of
+$CC^\omega$, a system that incorporates a predicative universe on top
+of the impredicative sort in Calculus of Constructions. We choose to
+focus on a simple term model so we do not have to take the extra step
+of mechanizing mathematical objects such as sets and domains.
+
+\section{Discussion}
+\label{sec:discuss}
+\yl{I'm not sure how strong of a statement we can make about our proof
+technique. We've already demonstrated how to address type-level
+dependency when defining a logical relation through a simple
+example. Claiming that our proof structure is better seems quite
+ambitious, but I think there's a middle point where we claim that
+adding moderate features like typed reduction doesn't instantly make
+our code size expand all the way from 1000 to 20,000 without saying
+the other developments are just verbose for no good reason}
+The short consistency proof we
+present achieves the goal of ilustrating the technique of proof by
+logical relation for dependently typed languages.
+In this section, we discuss what attributes to the shortness of our
+proof.
+% focus on the following question: why is our proof,
+% even with its extension to the existence $\beta\eta$-normal form, so much
+% shorter than the other mechanized results from \citet{decagda,
+% nbeincoq, martin-lof-a-la-coq}?
+
+First, unlike developments that mechanize the correctness of a
+type-directed conversion algorithm, we only show the existence of
 normal form for open and closed terms and state our properties in
 terms of the untyped small-step reduction relation. This removes a lot
-of scaffolding in the specification of our type system and our logical
-relation.
-
-This contrasts our development with
-\citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
-difficult property to prove due to the complexity of the equality
-judgment used to define the logical relation. In our system, the
-irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
-direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
-lemma name} which is
-straightforward after we prove the results about parallel
-reduction in Section~\ref{sec:spec}. In other words, we keep the
-the logical relation itself minimal at the cost of extra lemmas that can be proven
-through syntactic means independently from the logical relation.
-For example, as commented near the end of Section~\ref{sec:extension},
-our normalization property (Corollary~\ref{corollary:exbetanf})
-% can induce a decision procedure for type conversion by reducing two
-% terms to $\beta\eta$-normal form and check for their syntactic
-% equivalence. This decision procedure would be inefficient compared to
-% NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
-% to weak head normal and checks for the equivalence of subterms
-% recursively
-does not immediately induce an efficient decision procedure, but it is
-possible to recover an efficient algorithm by reasoning about untyped
-lambda terms.
+of scaffolding related to the specification of the algorithm and the
+extra proof obligation that the algorithm is sound and complete with
+respect to the declarative specification of the type system.
+
+Second, our logical relation, unlike the ones from \citet{decagda,
+  nbeincoq, martin-lof-a-la-coq, anand2014towards}, maps from types to
+predicates rather than relations. The need for a relational model is
+directly related to the metatheoretic property one wants to
+prove. \citet{anand2014towards} requires a relational model to capture
+the extensionality of their type system, wheras \citet{decagda,
+  martin-lof-a-la-coq, nbeincoq} uses a relational model to derive the
+injectivity of $\Pi$ types and justify the validity of
+$\eta$-conversion among other properties.
+Since \lang{} uses an untyped conversion rule, type conversion can be
+done through the \emph{normalize-and-compare}
+strategy described in \citet{pierce2004advanced}. The decidability of
+normalize-and-compare is implied by the existence of
+$\beta\eta$-normal form, which follows from our logical
+predicate.
+
+In \rref{I-Par, I-Bool}, we close over our semantically valid types
+and terms using the non-deterministic parallel reduction relation,
+which relates to full reduction, whereas \citet{decagda,nbeincoq} all
+refer to the deterministic weak head reduction. The use of a
+non-deterministic reduction strategy comes at the cost that we need
+confluence to prove the functionality of our logical relation, but
+immediately gives us a semantic justification of the conversion rule
+(Lemma~\ref{lemma:logrelcoherence}).
+% In particular, we obtain the confluence
+% result of reduction at a very early stage before we even define our
+% logical relation.
+
+With untyped conversion,
+we sidestep the relational, Kripke-style logical relation found in
+other mechanized proofs. However, our early dependence on confluence
+before the fundamental theorem is established can be alarming.
+In a system with type-directed reduction,
+confluence is not immediately available since it depends on
+depends on $\Pi$-injectivity, which is usually only proven after the
+fundamental theorem.
+Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
+problem that allow us to recover confluence property independently
+from the logical relation. \citet{siles2012pure} generalizes the
+notion of Type Parallel One Step Reduction from \citet{adams2006pure}
+to syntactically prove $\Pi$-injectivity for arbitrary Pure Tyep
+Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
+equational theory, and thus allowing subject reduction to be proven
+independently from confluence. Therefore, we believe that even in a
+system where type-directed reduction is required (e.g. a system with
+the unit $\eta$-law) in the logical relation, the proof can still be carried
+out in a structure similar to the one we have presented.
+
+% Finally, it is worth pointing out that all systems we have discussed
+% so far builds a relational model for their logical relation.
+
+% Recall
+% that our logical relation takes the form $[[InterpR i A S]]$ where
+% $[[S]]$ is a set of terms. In a relational model, each type is
+% interpreted as a partial equivalence relation over terms rather than a
+% set. A relational model is necessary for an extensional type theory
+% such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
+% use a relational model either to justify the function $\eta$-law or to
+% derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
+% law for function is baked into the untyped reduction
+% relation and therefore a relational model is not needed.
+% \scw{would it be difficult to extend your proof to be a PER?}
+
+% The dependently typed systems we have discussed so far also vary
+% greatly in expressiveness. For example, the Calculus of
+% Constructions~\citep{CoC} and MLTT without universe levels are unable
+% to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
+% identity types. % \citet{nbeincoq} lacks $\eta$-law for functions.
+% Among
+% the mechanized systems we have discussed so far, only
+% \citet{anand2014towards} has a full universe hierarchy.
+% Our object language \lang{}, while small, has a decent coverage of
+% features commonly seen in dependently typed languages, including
+% dependent pattern matching, dependent function types, a full universe
+% hierarchy, an intensional identity type. We hope features such as
+% type-directed reduction (necessary for unit $\eta$-law) and
+% impredicative sorts can be implemented in a similarly streamlined
+% fashion and we will leave those extensions as part of our future work.
+
+% To end this section, we
+% \scw{should this be a new subsection?}
+% The final question we want to address is: why is our proof,
+% even with the extension from Section~\ref{sec:extension}, so much
+% shorter than the proofs from \citet{decagda, nbeincoq,
+%   martin-lof-a-la-coq}?
+
+% This contrasts our development with
+% \citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
+% difficult property to prove due to the complexity of the equality
+% judgment used to define the logical relation. In our system, the
+% irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
+% direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
+% lemma name} which is
+% straightforward after we prove the results about parallel
+% reduction in Section~\ref{sec:spec}. In other words, we keep the
+% the logical relation itself minimal at the cost of extra lemmas that can be proven
+% through syntactic means independently from the logical relation.
+% For example, as commented near the end of Section~\ref{sec:extension},
+% our normalization property (Corollary~\ref{corollary:exbetanf})
+% % can induce a decision procedure for type conversion by reducing two
+% % terms to $\beta\eta$-normal form and check for their syntactic
+% % equivalence. This decision procedure would be inefficient compared to
+% % NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
+% % to weak head normal and checks for the equivalence of subterms
+% % recursively
+% does not immediately induce an efficient decision procedure, but it is
+% possible to recover an efficient algorithm by reasoning about untyped
+% lambda terms.
 
@@
 
-Another key to our proof is the very early establishment of the confluence
-property of the paralell reduction relation
-(Lemma~\ref{lemma:pardiamond}). This property is used in
-Lemma~\ref{lemma:interppreservation} to show that the interpretation
-of a type is preserved under evaluation. That is, if $[[InterpR i A
-S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
-In a system with typed directed reduction, the confluence result is
-harder to establish since confluence requires subject reduction, which
-circularly depends on $\Pi$-injectivity, a consequence of
-confluence~\citep{siles2012pure}. There are different workarounds to
-this problem. \citet{siles2012pure} proposes a syntactic approach by
-defining a type system where the confluence result is
-directly provable and later show that the system of interest is
-equivalent to the system with the confluence property.
-\citet{decagda} uses a relational model and the relational counterpart
-of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
-model is parameterized by a typed indexed equivalence relation and
-is reused to prove derive different properties in their
-development. Another perhaps simpler approach is to simply
-extend the judgmental equality with rules about
-$\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
-reduction to be proven independently from confluence and
-$\Pi$-injectivity can later be shown to be admissible. We find the
-approach from \citep{weirich:systemd} the most lightweight as it
-allows us to derive confluence early on even in systems with typed
-conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}
-
-In cases where the judgmental equality does not leverage type information heavily
-(e.g. $\eta$-law for the unit type), it might be possible to show that
-$\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
-$[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
-property holds (we believe this is likely true for both
-\citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
-terms of untyped parallel reduction to show the decidability of type
-conversion. Otherwise, to fully model the system, we need to bake
-typing information into our logical relation and we need to redefine
-our logical relation in Kripke-style so we can relax the scoping
-constraint in order to prove adequacy, though a relational model would
-still be unnecessary as long as the $\eta$ laws are baked into the
-typed directed relation. \scw{not sure I understand this last bit}
-
-Since confluence plays such as key role in our proof, we are uncertain
-how to generalize our technique to systems where confluence does not
-hold, and we will investigate those scenarios in our future work.
-\scw{What systems are you thinking of? Or do you mean more generally
-that this technique only works for languages where the definition of
-equality is based on joinability of parallel reduction?}
+% Another key to our proof is the very early establishment of the confluence
+% property of the paralell reduction relation
+% (Lemma~\ref{lemma:pardiamond}). This property is used in
+% Lemma~\ref{lemma:interppreservation} to show that the interpretation
+% of a type is preserved under evaluation. That is, if $[[InterpR i A S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
+% In a system with typed directed reduction, the confluence result is
+% harder to establish since confluence requires subject reduction, which
+% circularly depends on $\Pi$-injectivity, a consequence of
+% confluence~\citep{siles2012pure}. There are different workarounds to
+% this problem. \citet{siles2012pure} proposes a syntactic approach by
+% defining a type system where the confluence result is
+% directly provable and later show that the system of interest is
+% equivalent to the system with the confluence property.
+% \citet{decagda} uses a relational model and the relational counterpart
+% of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
+% model is parameterized by a typed indexed equivalence relation and
+% is reused to prove derive different properties in their
+% development. Another perhaps simpler approach is to simply
+% extend the judgmental equality with rules about
+% $\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
+% reduction to be proven independently from confluence and
+% $\Pi$-injectivity can later be shown to be admissible. We find the
+% approach from \citep{weirich:systemd} the most lightweight as it
+% allows us to derive confluence early on even in systems with typed
+% conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}
+
+% In cases where the judgmental equality does not leverage type information heavily
+% (e.g. $\eta$-law for the unit type), it might be possible to show that
+% $\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
+% $[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
+% property holds (we believe this is likely true for both
+% \citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
+% terms of untyped parallel reduction to show the decidability of type
+% conversion. Otherwise, to fully model the system, we need to bake
+% typing information into our logical relation and we need to redefine
+% our logical relation in Kripke-style so we can relax the scoping
+% constraint in order to prove adequacy, though a relational model would
+% still be unnecessary as long as the $\eta$ laws are baked into the
+% typed directed relation. \scw{not sure I understand this last bit}
+
+% Since confluence plays such as key role in our proof, we are uncertain
+% how to generalize our technique to systems where confluence does not
+% hold, and we will investigate those scenarios in our future work.
+% \scw{What systems are you thinking of? Or do you mean more generally
+% that this technique only works for languages where the definition of
+% equality is based on joinability of parallel reduction?}
 

### 2024-02-23 paper/paper.tex
Minor

@@
 The short consistency proof we
-present achieves the goal of ilustrating the technique of proof by
-logical relation for dependently typed languages.
-In this section, we discuss what attributes to the shortness of our
-proof.
+present achieves the goal of demonstrating the technique of proof by
+logical relation for dependently typed languages. However, what
+remains unanswered is what makes our development significantly shorter
+compared to others. Are we proving simpler results, or is our proof
+technique genuinely more efficient?
 % focus on the following question: why is our proof,
@@
 
-First, unlike developments that mechanize the correctness of a
+First, the metatheoretic property we prove is indeed simpler.
+Unlike developments that mechanize the correctness of a
 type-directed conversion algorithm, we only show the existence of
@@
 respect to the declarative specification of the type system.
-
-Second, our logical relation, unlike the ones from \citet{decagda,
+As a result, our logical relation, unlike the ones from \citet{decagda,
   nbeincoq, martin-lof-a-la-coq, anand2014towards}, maps from types to
@@
 prove. \citet{anand2014towards} requires a relational model to capture
-the extensionality of their type system, wheras \citet{decagda,
+the extensionality of their type system, whereas \citet{decagda,
   martin-lof-a-la-coq, nbeincoq} uses a relational model to derive the
@@
 
-In \rref{I-Par, I-Bool}, we close over our semantically valid types
-and terms using the non-deterministic parallel reduction relation,
+In terms of our proof technique, we close over our semantically valid types
+and terms in \rref{I-Par, I-Bool} using the non-deterministic parallel reduction relation,
 which relates to full reduction, whereas \citet{decagda,nbeincoq} all
@@
 problem that allow us to recover confluence property independently
-from the logical relation. \citet{siles2012pure} generalizes the
+from the logical relation. \citet{siles2012pure} generalize the
 notion of Type Parallel One Step Reduction from \citet{adams2006pure}
@@
 Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
-equational theory, and thus allowing subject reduction to be proven
+equational theory, thus allowing subject reduction to be proven
 independently from confluence. Therefore, we believe that even in a
@@
 out in a structure similar to the one we have presented.
+
+Finally, despite our earlier claim that our metatheoretic properties
+are not as strong as some of the related work, we can strengthen our
+results through syntactic means. For example, the
+normalize-and-compare strategy does not induce an efficient algorithm
+for type conversion, like the ones from \citet{decagda} and
+\citet{martin-lof-a-la-coq}. However, with the standard syntactic
+techniques from \citet{takahashi-parallel-reduction,
+factorization-essentially}, we can prove that leftmost-outermost
+reduction is a normalizing reduction strategy, from which we can
+separately show the correctness a more efficient algorithm that
+reduces the two terms to weak head normal form before recursively
+comparing their subcomponents. We believe factoring such properties
+out of a logical relation is valuable, as it helps us identify the
+part of our proof that requires extra strength from the metatheory.
 

### 2024-02-24 paper/paper.tex
Add todos
pass through mechanization section

@@
     Bib\TeX}}}
+
+%% to allow unicode characters in minted code
+\usepackage{newunicodechar}
+\newunicodechar{⇒}{$\Rightarrow$}
 
@@
 \begin{figure}[h]
-  \begin{tabular}{ c |  c  | c  }
-    & Consistency & Normalization \\
+  \begin{tabular}{ l |  c  | c | c }
+    & Consistency & Normalization & Syntactic metatheory \\
     \hline
     % Library (Autosubst 2)  & 491 & - \\
-    Syntactic typing (specification) &  68 & - \\
-    Renaming & 46 & -  \\
-    % Syntactic soundness & 660 &  - \\
-    Untyped reduction & 349 & - \\
-    Neutral and normal forms & N/A & 269 \\
-    Logical relation & 331 & 426 \\
-    Semantic typing and soundness & 169 & 199 \\
+    Syntactic typing (specification) &  68 & - & - \\
+    Renaming & 46 & -  & - \\
+    Untyped reduction & 349 & - & - \\
+    Neutral and normal forms & N/A & 269 & N/A \\
+    Logical relation & 331 & 426 & N/A \\
+    Semantic typing and soundness & 169 & 199 & N/A \\
+    Syntactic soundness & N/A & N/A  & 660 \\
     \hline
-    Total & 963 & 1357 \\
+    Total & 963 & 1357 & 1123 \\
   \end{tabular}
@@
 
-Figure~\ref{fig:linecount} shows the statics of our
+Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
+code\footnote{calculated by the \texttt{cloc} tool} for each file of our
 development, including the base consistency proof from
-Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
-extension to normalization for open and closed terms from
-Section~\ref{sec:extension}.
-
-Since the normalization and the consistency development share the same
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the extension to
+normalization for open and closed terms from Section~\ref{sec:extension}.
+
+Because the normalization and the consistency development share the same
 syntactic and typing specification, they differ in only the categories
-that are related to the logical relation. We use - as a marker that
+that are relevant to the logical relation. We use - as a marker that
 the line count is the same.
@@
 Orthogonal to the development of our consistency and normalization
-proof, we prove the syntactic soundness of our system through subject
+proof, we have also proven the syntactic soundness of our system through subject
 reduction and progress. The syntactic soundness proof shares the
@@
 factored out as a separate file under the Renaming (common) category.
+\scw{Should add this line count as another column, for comparison. I've done so, 
+but may need to update the numbers. }
 
@@
 In this section, we discuss the Coq encoding of the definitions and proofs presented
-in Section~\ref{sec:logrelproof}  and the artifacts that are
-specific to Coq and may not appear in other proof assistants.
+in Section~\ref{sec:logrelproof} and the artifacts of this encoding that are
+specific to Coq.
 
@@
 terms of the logical relation itself (e.g. in
-Lemma~\ref{lemma:piintroalt}). However, in the mechanized proof, the
-definition of $[[Interp I i A S]]$ can in fact be encoded without the
-use of impredicativity by consistently replacing the use of
-\texttt{Prop} with Coq's predicative sort \texttt{Type}. One can even
-use this \texttt{Type} variant of the logical relation to define
-interpretation for any finite number of universes. The lack of
-impredicativity becomes troublesome only when we attempt to define
+Lemma~\ref{lemma:piintroalt}). \scw{I'm having trouble connecting the rule in 
+the figure with rule \texttt{InterpExtFun} in the coq development. Would it make 
+sense to include the definition of ProdSpace? What is the Coq type of R?}
+
+In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
+\texttt{Prop}, when \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S}
+has type \texttt {tm -> Prop}.  However, if desired, we could consistently
+replace the use of \texttt{Prop} with Coq's predicative sort
+\texttt{Type}. This variation of the definition could be used to define the
+interpretation for any \emph{finite} number of universes. The use of
+\texttt{Type} becomes troublesome only when we attempt to define
 $[[InterpR i A S]]$, the top-level logical relation defined in
-Figure~\ref{fig:logrelrec} that recursively calls itself at smaller
-universe levels. As a result, the one feature that truly requires
-impredicativity from \lang{} is its countable universe
-hierarchy, though lemmas such as the admissibility of \rref{I-PiAlt}
-(Lemma~\ref{lemma:piintroalt}) not only simply our mechanization, but
-also give better intuition on what our logical relation means.
+Figure~\ref{fig:logrelrec} that recursively calls itself at smaller universe
+levels. Therefore, the one feature of \lang{} that truly requires impredicativity 
+is its countable universe hierarchy. \scw{I'm not sure the last part of this sentence
+adds much: ``Although, lemmas such as the
+admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only simply
+our mechanization, but also give better intuition on what our logical relation
+means.'' Yes, we need impredicativity to define these lemmas, but without impredicativity
+our logical relation means something else.}
 
@@
 correspondence to the Coq definition. The main difference is the
-specification of $[[I]]$. In the presentation in
+specification of $[[I]]$. In 
 Section~\ref{sec:logreldep}, we define $[[I]]$ as a function over
@@
     [[I]]   &\in [[SNat -> PowerSet STm]] \\
-    [[I j]] &= [[{A | exists S , InterpR j A S}]] \text{, if } j < i \\
-    [[I j]] &= [[emptyset]] \text{, } j \geq i
+    [[I j]] &= 
+     \begin{cases}
+      \ [[{A | exists S , InterpR j A S}]] & \text{when } j < i \\
+      \ [[emptyset]] & \text{otherwise} 
+    \end{cases}
   \end{split}
@@
 \begin{minted}{coq}
-Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
+Lemma par_triangle a : forall b, (a ⇒ b) -> (b ⇒ tstar a).
 Proof.
@@
 \end{minted}
-In prose, the triangle property can be proven by induction over the
-graph of \mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par} and
-\texttt{ctrs:Par} say that the proof involves inverting and
-constructing of the derivations of parallel
-reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
-automation tactic to use
-the reflexivity and congruence properties of parallel
-reduction as lemmas.
-The flag \texttt{lq:on} toggles tunes CoqHammer's search
-algorithm, though we never need to specify those options
-manually. Instead, we first invoke the \texttt{sauto} tactic indirectly through the
-\texttt{best} tactic provided by CoqHammer where we specify only the
-\texttt{inv}, \texttt{ctrs}, and lemmas we want to use. The
-\texttt{best} tactic iterates through possible configurations and
-provides us a replacement to the \texttt{best} tactic with the
-tuned performance flags to save time for future re-execution of the
-tactic.
-
-The automation provided by CoqHammer not only gives us a proof that is
-shorter and more resiliant to changes, but also gives useful
-documentation for readers who wish to understand how the underlying
-proof works since the automation tactic is guaranteed to not use
-lemmas or invert derivations that are not specified in the
-\texttt{use} or \texttt{inv} flag.
+In prose, the triangle property can be proven by induction over the graph of
+\mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par}
+and \texttt{ctrs:Par} say that the proof involves inverting and constructing
+of the derivations of parallel reduction. The option
+\texttt{use:Par\_refl,par\_cong} allows the automation tactic to use the
+reflexivity and congruence properties of parallel reduction as lemmas.
+
+The flag \texttt{lq:on} tunes CoqHammer's search algorithm.  While this flag
+appears arcane, when developing our proof scripts we never specify this option
+manually. Instead, we first invoke the \texttt{best} tactic provided by
+CoqHammer, specifying only the \texttt{inv}, \texttt{ctrs}, and lemmas that we
+want to use. The \texttt{best} tactic then iterates through possible
+configurations and provides us with a replacement with the tuned performance
+flags that save time for future re-execution of the proof script.
+
+The automation provided by CoqHammer not only gives us a proof that is shorter
+and more resiliant to changes, but also gives useful documentation for readers
+who wish to understand how the underlying proof. Although automation performs
+extensive search, it does not use lemmas or invert derivations that are not
+specified in the \texttt{use} or \texttt{inv} flag.
+
+\subsection{Use of impredicativity}
+\yl{Mention that logrel itself doesn't require impredicativty}
+\scw{Didn't we say this above?}
+
+\subsection{Extraction of conversion algorithm}
+\yl{technically future work. Not sure whereelse to put it}
 

### 2024-02-24 paper/paper.tex
Add more details to Section 6 (beta eta)

@@
 \section{Existence of $\beta\eta$-normal forms}
-\scw{relocated text from previous section. Not a complete description yet.}
-
-We first extend parallel reduction to include $\eta$ reduction for functions.
+We can easily incorporate the function $\eta$ law to the equational
+theory of \lang{} by adding the following parallel reduction rule.
 \begin{center}
@@
 \end{center}
-
-\Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
-equational theory since convertibility ($[[a <=> b]]$), the untyped relation
-used for type conversion, is built on top of parallel reduction. We can
-recover the same confluence result about parallel reduction using the standard
-techniques from \citet{barendregt:lambda-calculi-with-types,
-  takahashi-parallel-reduction}, though specifically for the de Bruijn
-representation, we need the following anti-renaming lemma about parallel
-reduction.
-
-Both the confluence and anti-renaming properties require structural
-induction over the size of terms to obtain a useful induction
-hypotehsis in the $\eta$ case.
-
-This lemma is mainly to show that \rref{P-AbsEta} does not introduce
-any new $\beta$ redex because $\eta$ reduction is the only real step
-we can take.
-
+% \Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
+% equational theory since convertibility ($[[a <=> b]]$), the untyped relation
+% used for type conversion, is built on top of parallel reduction.
+We can recover the same confluence result about parallel reduction
+using the standard techniques from
+\citet{barendregt:lambda-calculi-with-types,
+takahashi-parallel-reduction}, though anti-renaming
+(Lemma~\ref{lemma:parantirenaming}) needs to be proven before the
+diamond property (Lemma~\ref{lemma:pardiamond}). Another complication
+is that confluence and anti-renaming now requires induction on a size
+metric of lambda terms so the $\eta$ case has a useful induction hypothesis.
+
+Note that the specification of our logical relation does not require
+any changes. We can continue using $\ottkw{ne}$ and $\ottkw{nf}$ to
+represent $\beta$-neutral and normal forms. By reproving the
+fundamental lemma, we learn that every well-typed term as a
+$\beta$-normal form. Lemma~\ref{lemma:parnenf} still holds with our
+extended parallel reduction. In the presence of the $\eta$ reduction
+rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
+preserves $\beta$-normal forms (i.e. does not produce new
+$\beta$-redexes). Furthermore, since the $\eta$ reduction rule for
+functions strictly decreases the size of the term, the existence of
+$\beta\eta$ normal form trivially follows.
 \begin{corollary}[Existence of $\beta\eta$-normal form]
@@
 \end{corollary}
-\begin{proof}
-  From Corollary~\ref{corollary:exbetanf}, we can easily conclude the
-  existence of $\beta\eta$-normal forms for well-typed terms since
-  $\eta$-reduction preserves $\beta$-normal form (implied by
-  Lemma~\ref{lemma:parnenf}) and also strictly decreases the size of
-  the term (evident from \rref{P-AbsEta}).
-\end{proof}
+Due to some technical
+details, our proof about the existence of $\beta\eta$ form does not
+immediately induce a decision procedure for type conversion in
+\lang{}. We discuss this issue and its workaround in
+Section~\ref{sec:conversionalgo}.
 
@@
 \subsection{Extraction of conversion algorithm}
+\label{sec:conversionalgo}
 \yl{technically future work. Not sure whereelse to put it}

### 2024-02-24 paper/paper.tex
pass through discussion section. questions.

--- 2024-02-24 paper/paper.tex
@@
 \yl{I'm not sure how strong of a statement we can make about our proof
-technique. We've already demonstrated how to address type-level
-dependency when defining a logical relation through a simple
-example. Claiming that our proof structure is better seems quite
-ambitious, but I think there's a middle point where we claim that
-adding moderate features like typed reduction doesn't instantly make
-our code size expand all the way from 1000 to 20,000 without saying
-the other developments are just verbose for no good reason}
-The short consistency proof we
-present achieves the goal of demonstrating the technique of proof by
-logical relation for dependently typed languages. However, what
-remains unanswered is what makes our development significantly shorter
-compared to others. Are we proving simpler results, or is our proof
+  technique. We've already demonstrated how to address type-level dependency
+  when defining a logical relation through a simple example. Claiming that our
+  proof structure is better seems quite ambitious, but I think there's a
+  middle point where we claim that adding moderate features like typed
+  reduction doesn't instantly make our code size expand all the way from 1000
+  to 20,000 without saying the other developments are just verbose for no good
+  reason} The short consistency proof we present achieves the goal of
+demonstrating the technique of proof by logical relation for dependently typed
+languages. However, what remains unanswered is what makes our development
+significantly shorter compared to others. Are we proving simpler results, for
+smaller languages, are we making more use of automation, or is our proof
 technique genuinely more efficient?
@@
 
-First, the metatheoretic property we prove is indeed simpler.
-Unlike developments that mechanize the correctness of a
-type-directed conversion algorithm, we only show the existence of
-normal form for open and closed terms and state our properties in
-terms of the untyped small-step reduction relation. This removes a lot
-of scaffolding related to the specification of the algorithm and the
-extra proof obligation that the algorithm is sound and complete with
+First, the metatheoretic properties that we prove are indeed simpler.  Unlike
+developments that mechanize the correctness of a type-directed conversion
+algorithm, we only show the existence of normal forms for open and closed
+terms and state our properties in terms of an untyped reduction relation. This
+avoids a lot of the scaffolding related to the specification of the algorithm
+and the proof obligations that the algorithm is sound and complete with
 respect to the declarative specification of the type system.
@@
   nbeincoq, martin-lof-a-la-coq, anand2014towards}, maps from types to
-predicates rather than relations. The need for a relational model is
+predicates rather than relations. \scw{How does this follow? Why do we need unary relations where they need binary relations?} The need for a relational model is
 directly related to the metatheoretic property one wants to
@@
 
+\scw{I'm getting confused by this paragraph. Does this reorganization sense:
+  Our language is simpler than Nuprl, because it doesn't have extensional
+  equality. It is simpler than Agda, because it doesn't have type-directed
+  equality. Both of these cases require the definition of a binary logical
+  relation, that defines a notion of semantic equality between terms. This
+  relation justifies the injectivity of $\Pi$ types and justify the validity
+  of $\eta$-conversion among other properties.}  \scw{Furthermore our proof is
+  also simpler because we don't need prove the correctness of the NBE
+  algorithm, which is used to show the decidability of Agda's type-directed
+  equivalence. Therefore, we don't need to define this algorithm and show that
+  it is sound and complete with respect to the type-directed
+  equality. Instead, to show the decidability of our untyped equivalence, we
+  need only show that terms have $\beta\eta$ normal forms. }
+
 In terms of our proof technique, we close over our semantically valid types
-and terms in \rref{I-Par, I-Bool} using the non-deterministic parallel reduction relation,
-which relates to full reduction, whereas \citet{decagda,nbeincoq} all
-refer to the deterministic weak head reduction. The use of a
-non-deterministic reduction strategy comes at the cost that we need
-confluence to prove the functionality of our logical relation, but
-immediately gives us a semantic justification of the conversion rule
-(Lemma~\ref{lemma:logrelcoherence}).
+and terms in \rref{I-Par, I-Bool} using the non-deterministic parallel
+reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak
+head reduction relation. Our use of a non-deterministic reduction strategy
+means that we need confluence to prove the functionality of our logical
+relation. However, the benefit is that it immediately gives us a semantic
+justification of the conversion rule (Lemma~\ref{lemma:logrelcoherence}).
 % In particular, we obtain the confluence
@@
 we sidestep the relational, Kripke-style logical relation found in
-other mechanized proofs. However, our early dependence on confluence
+other mechanized proofs. \scw{Need to define Kripke-style. Also the other 
+proofs need Kripke style because they are defining typed relations, not untyped 
+relations. }
+However, our early dependence on confluence
 before the fundamental theorem is established can be alarming.
 In a system with type-directed reduction,
-confluence is not immediately available since it depends on
+confluence is not immediately available because it 
 depends on $\Pi$-injectivity, which is usually only proven after the
-fundamental theorem.
+fundamental theorem.\scw{confluence depends on Pi injectivity? I thought 
+it was only needed for subject reduction}
 Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
@@
 notion of Type Parallel One Step Reduction from \citet{adams2006pure}
-to syntactically prove $\Pi$-injectivity for arbitrary Pure Tyep
+to syntactically prove $\Pi$-injectivity for arbitrary Pure Type
 Systems. \citet{weirich:systemd} add $\Pi$-injectivity to their
@@
 the unit $\eta$-law) in the logical relation, the proof can still be carried
-out in a structure similar to the one we have presented.
+out in a structure similar to the one we have presented. \scw{Not sure that 
+I understand this paragraph}
 
@@
 part of our proof that requires extra strength from the metatheory.
+\scw{Why don't we just use leftmost-outermost reduction in the first place? 
+Do we even need nondeterministic parallel reduction?}
 

### 2024-02-24 paper/paper.tex
Fill in the code extraction section

@@
 
-\subsection{Use of impredicativity}
-\yl{Mention that logrel itself doesn't require impredicativty}
-\scw{Didn't we say this above?}
-
-\subsection{Extraction of conversion algorithm}
+\subsection{Extraction of a conversion algorithm}
 \label{sec:conversionalgo}
-\yl{technically future work. Not sure whereelse to put it}
+A constructive proof of the existence of $\beta\eta$-normal form
+for well-typed terms (Corollary~\ref{corollary:exbetaeta}) induces a
+normalization algorithm. From this normalization procedure, we can
+derive a normalize-and-compare algorithm. Given two well-typed terms
+$[[a]]$ and $[[b]]$, to know whether $[[a <=> b]]$, we apply the
+normalization algorithm on $[[a]]$ and $[[b]]$ to obtain $\beta\eta$
+normal forms $[[f0]]$ and $[[f1]]$. The algorithm then returns true
+exactly when $[[f0]]$ and $[[f1]]$ are syntactically equal. This
+algorithm is referred to as normalize-and-compare by
+\citet{pierce2004advanced}.
+
+The soundness of the algorithm is immediate. The completeness of the
+algorithm is justified the confluence property of the untyped
+reduction relation. Suppose $[[a <=> b]]$, $[[a =>+ f0]]$, $[[b =>+
+f1]]$ but $[[f0]]$ is syntactically distinct from each
+other. By the transitivity of convertibility, we have $[[f0 <=> f1]]$.
+Because $[[f0]]$ and $[[f1]]$ are both in
+$\beta\eta$-normal forms and can only reduce to themselves, we must
+have $[[f0 = f1]]$. Contradiction.
+
+In our development, since our countable universe hierarchy
+relies on impredicativity from the metatheory, we encode our
+properties in Coq's \texttt{Prop} sort, which is not very suitable for
+code execution. On the other hand, due to our heavy reliance on
+automated proof search, even if we were able to extract an algorithm from \texttt{Prop},
+the algorithm would be unpredictable because its definition depends on
+the specific choices made by the proof search algorithm.
+
+However, with a little more effort, it is possible to recover a
+precise algorithm. The first step is to define a deterministic
+small-step reduction relation that is normalizing; that is, the
+relation can always find a normal form if there exists one. A good
+candidate is the leftmost-outermost reduction strategy. Its
+normalizing property is standard and can be proven using the
+factorization technique discussed in
+\citet{takahashi-parallel-reduction, factorization-essentially}. By
+composing the existence of normal form and the fact that the
+deterministic relation is normalizing, we can derive show the
+accessibility of the deterministic reduction relation
+(\mintedinline{coq}{Acc} in Coq), and use the accessibility proof as
+an induction metric to define an executable algorithm.
 

### 2024-02-24 paper/paper.tex
Typo

@@
 accessibility of the deterministic reduction relation
-(\mintedinline{coq}{Acc} in Coq), and use the accessibility proof as
+(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
 an induction metric to define an executable algorithm.

### 2024-02-24 paper/paper.tex
another comment in the discussion section

--- 2024-02-24 paper/paper.tex
@@
   to 20,000 without saying the other developments are just verbose for no good
-  reason} The short consistency proof we present achieves the goal of
+  reason} 
+\scw{I think we can find reasons for much of the differences. Am I missing any?
+While it would be difficult to assign numbers to each of the deltas, I think it 
+is believable that when put together they add up to a lot.
+\begin{itemize}
+\item We don't include inductive or coinductive datatypes. We don't include 
+cumulativity. We don't include Prop. We don't include universe polymorphism.
+\item We state our equality algorithmically instead of declaratively. On one
+  hand, this gives us automatic inversion principles when working with
+  definition. Furthermore, we don't need to prove the equivalence between an
+  algorithmic version and a declarative specification.
+\item Our equality requires a simple decision algorithm and isn't type directed.
+\item We don't prove decidability of type checking. (And, it is not provable
+  for our system, because we lack type annotations on functions. We should
+  point this out.)
+\item Our logical relation is unary and untyped. The latter means that we don't 
+  require the bookkeeping of a Kripke logical relation when reasoning about 
+  open terms. I don't know why unary relations are shorter.
+\item CoqHammer leads to short proofs.
+\end{itemize}
+}
+
+The short consistency proof we present achieves the goal of
 demonstrating the technique of proof by logical relation for dependently typed

### 2024-02-24 paper/paper.tex
Answer some questions

@@
 fundamental theorem.\scw{confluence depends on Pi injectivity? I thought
-it was only needed for subject reduction}
+it was only needed for subject reduction}\yl{it's
+transitive. Confluence depends on subject reduction, which in turn
+depends on pi injectivity. Maybe it's worth spelling out the details}
 Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
@@
 equational theory, thus allowing subject reduction to be proven
-independently from confluence. Therefore, we believe that even in a
-system where type-directed reduction is required (e.g. a system with
-the unit $\eta$-law) in the logical relation, the proof can still be carried
-out in a structure similar to the one we have presented. \scw{Not sure that
-I understand this paragraph}
+independently from confluence. By adopting these techniques that allow
+us to derive confluence early even for systems with type-directed
+reduction, we believe our proof technique can significantly shorten
+the existing logical relation proofs for systems with typed judgmental
+equality and more complex $\eta$ laws. We leave that as part of our
+future work.
+% Therefore, we believe that even in a
+% system where type-directed reduction is required (e.g. a system with
+% the unit $\eta$-law) in the logical relation, the proof can still be carried
+% out in a structure similar to the one we have presented.
+\scw{Not sure that I understand this paragraph}
+\yl{Reworded slightly to emphasize it's future work that we haven't
+  done and it is speculative}
 

### 2024-02-24 paper/paper.tex
Another comment
Answer more questions

@@
 \end{itemize}
+\yl{
+  Just one more technical point to add, though it's in the text already:
+  the logical relation is closed backward by full reduction rather
+  than weak-head/deterministic reduction. This requires an early
+  confluence result to show that the logrel is
+  deterministic/functional but simplifies everything else
+  (e.g. conversion is justified immediately by our preservation
+  theorem, but that is not the case if you use weak head reduction).
+
+  Also, regarding the first point, cumulativity only exists in
+  Barras's work. Inductive, (maybe coinductive?), can be found in
+  nuprl, metacoq, and maybe Barras's work.
+
+  The 20,000 - 30,000 LoC mechanization are all about small languages
+  with pretty much the same features as our language except for your
+  second and third bullet point. martin-lof a la coq, Abel's work, and
+  nbe in coq aren't that richer in feature otherwise. None includes
+  cumulativity (they only have one predicative universe)
+
+  The 400,000 NuPRL in Coq probably falls into a different cateogry
+  because they are trying to mechanize a full practical language
+}
+
 }
@@
 Do we even need nondeterministic parallel reduction?}
+\yl{The conversion uses full reduction. Nondeterministic reduction
+  makes it harder to show that convertible types have the same
+  meaning. Maybe it would require us to prove factorization in our
+  development but it definitely simplifies the determinism proof
+  (confluence is no longer required before the fundamental
+  lemma). }
 

### 2024-02-24 paper/paper.tex
Fix compile error

@@
 \end{itemize}
-\yl{
-  Just one more technical point to add, though it's in the text already:
+}
+\yl{ Just one more technical point to add, though it's in the text already:
   the logical relation is closed backward by full reduction rather
@@
   (e.g. conversion is justified immediately by our preservation
-  theorem, but that is not the case if you use weak head reduction).
-
+  theorem, but that is not the case if you use weak head reduction).\\ \\
   Also, regarding the first point, cumulativity only exists in
   Barras's work. Inductive, (maybe coinductive?), can be found in
-  nuprl, metacoq, and maybe Barras's work.
-
+  nuprl, metacoq, and maybe Barras's work.\\ \\
   The 20,000 - 30,000 LoC mechanization are all about small languages
@@
   nbe in coq aren't that richer in feature otherwise. None includes
-  cumulativity (they only have one predicative universe)
-
+  cumulativity (they only have one predicative universe)\\ \\
   The 400,000 NuPRL in Coq probably falls into a different cateogry
-  because they are trying to mechanize a full practical language
-}
-
-}
+  because they are trying to mechanize a full practical language}
 

### 2024-02-24 paper/paper.tex
Remove some metavariables

@@
 \mathit{Terms}\\
-[[a]],[[b]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
+[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                   & \mbox{universes, variables, empty type} \\
@@
                   & \mbox{function types, abstractions, applications} \\
-            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
+            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                   & \mbox{equality types, reflexivity proof, J eliminator} \\
@@
 \]
-  \caption{Syntax of \lang \scw{Where does the name come from?}}
+  \caption{Syntax of \lang \scw{Where does the name come from?}\yl{H =
+    hierarchy though it needs a better name}}
   \label{fig:syntax}
 \end{figure}
-
-% \begin{figure}[h]
-%     \[
-%       \begin{array}{lll}
-%         \mathit{Curried\ Addition} \\
-%         add(n) & := & m \mapsto n + m \\ \\
-%         \mathit{Extension ([[xi]])} \\
-%         [[(xi .: j) 0]]  & := & 0 \\
-%         [[(xi .: j) Suc i]]  & := & [[xi j]] \\ \\
-
-%         \mathit{Extension ([[rho]])} \\
-%         [[(rho .: a) 0]]  & := & [[a]] \\
-%         [[(rho .: a) Suc i]]  & := & [[rho i]] \\ \\
-
-%         \mathit{Up ([[xi]])} \\
-%         [[up xi]] & := & (add(1) \circ [[xi]]) , 0 \\ \\
-
-%         \mathit{Renaming} \\
-%         [[var i {xi}]] & := & [[xi i]] \\
-%         [[(Set i) {xi}]] & := & [[Set i]] \\
-%         [[Void {xi}]] & := & [[Void]]\\
-%         [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
-%         [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
-%         [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
-%         [[Bool {xi}]] & := & [[Bool]] \\
-%         [[true {xi}]] & := & [[true]] \\
-%         [[false {xi}]] & := & [[false]] \\
-%         [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
-%         [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
-%         [[refl {xi}]] & := & [[refl]] \\
-%         [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\
-
-%         \mathit{Lookup} \\
-%         [[(G ++ A) 0]] & := &  [[A]] \\
-%         [[(G ++ A) Suc i]] & := & [[G i]] \\ \\
-
-%         \mathit{Drop} \\
-%         [[drop 0 G]] & := & [[G]] \\
-%         [[drop Suc i (#G ++ A#)]] & := & [[drop i G]] \\ \\
-%       \end{array}
-%     \]
-%   \caption{Auxiliary Functions over Syntax}
-%   \label{fig:auxdef}
-% \end{figure}
 
@@
 The syntax of \lang can be found in Figure~\ref{fig:syntax}.
-\scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
+% \scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
+% \yl{need at least 4 vars for J. removed the unnecessary t and C}
 We use
@@
 Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
-proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
-$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
+proofs can be eliminated by the J-eliminator $[[J c a b p]]$, where
+$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[c]]$
 is the term whose type is to be casted.
@@
       \begin{array}{lll}
-        % \mathit{UpRen} \\
-        % [[( up xi ) 0]] & := & [[0]] \\
-        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
-        % \mathit{Renaming ([[a < xi >]])} \\
-        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
-        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
-        % \ldots \\ \\
         \mathit{Lifting} \\

### 2024-02-24 paper/paper.tex
pass through section 4

--- 2024-02-24 paper/paper.tex
@@
 \yl{todo: rename all variables from i,j to n}
-\begin{figure}[h]
-\[
-\begin{array}{lcl}
-
-      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
-                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
-      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
-                       G]]\text{ then there exists some } [[j]] \text{
-                       and } [[S]] \text{ such that } [[InterpR j A
-                       {rho} S]] \text{ and } [[a {rho} in S]] \\
-      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
-\end{array}
-\]
-  \caption{Semantic Typing for \lang}
-  \label{fig:semtyping}
-\end{figure}
+\scw{Would it make sense to define the notation $a \in [\![A]\!]^i$ when
+there exists some $S$ such that $[[InterpR i A S]]$ and $[[a in S]]$ ?}
+% \begin{figure}[h]
+% \[
+% \begin{array}{lcl}
+
+%       [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
+%                      } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
+%       [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |= G]]
+%    \text{ then there exists some } [[j]] \text{
+%                        and } [[S]] \text{ such that } [[InterpR j A {rho} S]]
+%        \text{ and } [[a {rho} in S]] \\
+%       [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
+% \end{array}
+% \]
+%   \caption{Semantic Typing for \lang}
+%   \label{fig:semtyping}
+% \end{figure}x
 
@@
 
-% \scw{What is the motivation for this section? Why do we want to generalize
-% the logical relation?}
-In this section, we define the notion of semantic typing based on the
-logical relation we have defined in Section~\ref{sec:logreldep} and
-prove the fundamental lemma, which states that syntactic typing
-implies semantic typing.
-
-The necessity of semantic typing as an extra layer of definition on
-top of the logic relation can be understood in simply typed
+In this section, we show that all closed, well-typed terms are contained
+within their type-indexed sets. In other words, $[[empty |- a : A]]$ implies
+$[[InterpR i A S]]$ and $[[a in S]]$.  This result gives us consistency
+because we know that $[[InterpR i Void S]]$ is defined, and that $[[S]]$ must
+be the empty set. Therefore, if there were some closed, well-typed term of type
+$[[Void]]$, it would need to be a member of the empty set, a contradiction.
+
+To prove this result, we define a notion of semantic typing based on the
+logical relation we have defined in Section~\ref{sec:logreldep} and prove the
+fundamental lemma, which states that syntactic typing implies semantic typing.
+Semantic typing extends our logical relation from being a (type-indexed)
+family of predicates on closed terms, to a type-indexed family of predicates
+on open terms.
+
+The necessity of semantic typing as an extra layer of
+definition on top of the logic relation can be understood in simply typed
 languages~\citep{skorstengaard2019introduction, harpertait,
-  pierce2002types}. In our setting, attempting to show that $[[empty
-|- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ will fail in
-\rref{T-Abs}, where the induction hypothesis is not helpful since the body
-of the lambda term is typed under a non-empty context. Through
-the definition of semantic typing, we can state a strengthened
-property that is actually provable.
+  pierce2002types}. In our setting, attempting to show that
+$[[empty |- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ will fail in
+\rref{T-Abs}, where the induction hypothesis is not helpful since the body of
+the lambda term is typed under a non-empty context. Through the definition of
+semantic typing, we can state a strengthened property that is actually
+provable.
+
+\begin{definition}[Semantic well-formed substitution\footnote{\dotv{soundness.v}{$\rho$\_ok}}]
+Define $[[rho |= G]]$ when 
+\[ \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
+                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \]
+\end{definition}
 
@@
 % respect the interpretation of the types from the context.
-The full
-definitions of well-formed substitution ($[[rho |= G]]$), semantic
-typing ($[[ G |= a : A]]$), and semantic context well-formedness
-($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
-
-The $[[rho |= G]]$ notation denotes the semantic well-formedness of
-a substitution $[[rho]]$ with respect to a context
-$[[G]]$. Intuitively, for variable $[[i]]$ in the context (thus the
-restriction $[[i < | G |]]$), $[[rho i]]$ is a term that inhabits all
-possible interpretations of the type $[[G i]]$ from the
-context.
+% The full
+% definitions of well-formed substitution ($[[rho |= G]]$), semantic
+% typing ($[[ G |= a : A]]$), and semantic context well-formedness
+% ($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
+
+The $[[rho |= G]]$ notation denotes the semantic well-formedness of a
+substitution $[[rho]]$ with respect to a context $[[G]]$. Intuitively, for any
+variable $[[i]]$ in the context (thus the restriction $[[i < | G |]]$),
+$[[rho i]]$ is a term that inhabits all possible interpretations of the type
+$[[G i]]$ from the context.
+
 Since we are in a dependently typed system, there
@@
 substitution can be carried out.
+
 Finally, despite the for all quantification over the interpreted set
@@
 law about $[[rho |= G]]$.
-\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\gamma$\_ok\_nil}}]
+\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
   \label{lemma:rhowfempty}
@@
 \end{lemma}
-\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
+\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\rho$\_ok\_cons}}]
   If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
@@
 
-We refer to $[[G |= a : A]]$ as semantic well-typedness. Its definition says the term $[[a]]$ can be
-semantically typed $[[A]]$ under the context $[[G]]$ if for all
-substitutions $[[rho]]$ such that $[[rho |= G]]$, the type $[[A { rho }]]$ can be interpreted
-as the set $[[S]]$, and $[[a { rho } in S]]$. Our definition of
-semantic well-typedness is standard, though dependent types add a
-small twist that we apply the $[[rho]]$ to $[[A]]$ and require that
-$[[A { rho }]]$ has some interpretation, which is not always the case
-for arbitrary types.
-
-Analogous to $[[|- G]]$, the syntactic well-formedness of the typing
-context, $[[|= G]]$ defines when a context is semantically
-well-formed. In the
-definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
+We next define semantic well-typedness.
+
+\begin{definition}[Semantic typing\footnote{\dotv{soundness.v}{SemWt}}]
+Define $[[G |= a : A]]$  when
+\[ \forall [[rho]], \text{ if }[[rho |=
+                       G]]\text{ then there exists some } [[j]] \text{
+                       and } [[S]] \text{ such that } [[InterpR j A
+                       {rho} S]] \text{ and } [[a {rho} in S]]  \]
+\end{definition}
+
+This definition says the term $[[a]]$ can be semantically typed $[[A]]$ under
+the context $[[G]]$ if for all substitutions $[[rho]]$ such that
+$[[rho |= G]]$, the type $[[A { rho }]]$ can be interpreted as the set
+$[[S]]$, and $[[a { rho } in S]]$. Our definition of semantic well-typedness
+is standard, though dependent types add a small twist that we apply the
+$[[rho]]$ to $[[A]]$ and require that $[[A { rho }]]$ has some interpretation.
+
+Finally, we define semantic well-formedness for contexts, analogous to the 
+relation $[[|-G]]$.
+
+\begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
+Define $[[|= G]]$ when
+\[
+   \forall [[i < |G|]],\text{ there exists some } [[j]] \text{ such that } [[drop Suc i G |= G i : Set j]] \]
+\end{definition}
+
+In this definition, we use the notation $[[drop i G]]$ to denote
 the typing context obtained by dropping the last $[[i]]$ elements of
-$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
-returns the empty list. Intuitively, the definition of $[[|= G]]$
-simply says all types in the context must correspond to some set
-according to the logical relation. The truncation to the context
-$[[G]]$ is due to the different scoping of the types in the
-context. An alternative way of defining $[[|= G]]$ would look more
-similar to $[[|- G]]$, though we find it easier to define $[[|= G]]$
-through for all quantification of $[[i]]$ and recover the structural
-rules as lemmas.
+$[[G]]$. When $[[G]]$ has fewer than $[[i]]$ elements, $[[drop i G]]$ returns
+the empty list. Intuitively, $[[|= G]]$ says that all
+types in the context must correspond to some set according to the logical
+relation. The truncation to the context $[[G]]$ is due to the telescopic
+scoping of the types in the context. An alternative approach would be to define $[[|= G]]$
+inductively, similar to $[[|- G]]$, but we find it easier to use this definition and
+recover the structural rules as lemmas.
 \begin{lemma}[Semantic context well-formedness cons]
@@
 Next, we show some non-trivial cases of the fundamental theorem as
-top-level lemmas.
+top-level lemmas. For example, we can define the semantic analogue to the 
+syntactic typing rule for variables (\rref{T-Var}).
 \begin{lemma}[ST-Var]
@@
   the context $[[G]]$ up to shifting.
-
+%
   Knowing that $[[G |= G i {up Suc i} : Set j]]$ holds, we can then
@@
   A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
-  from Lemma~\ref{lemma:semwffcons}.
-Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
-  can be used to discharge their syntactic counterpart
+  from Lemma~\ref{lemma:semwffcons}. The semantic typing rules
+(Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp})
+  can be used to discharge their syntactic counterparts
   (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
   cases not covered by the lemmas are similar to the ones already
-  shown or simpler and therefore omitted from the text.
+  shown.
 \end{proof}
@@
   satisfies $[[rho |= G]]$. For convenience, we pick $[[rho]]$ as
-  $[[idtm]]$ (though any $[[rho]]$ would work since
+  $[[idtm]]$, though any $[[rho]]$ would work since
   $[[empty |- a : Void]]$ ensures there is no free variable in
-  $[[a]]$), we have $[[a {idtm}]] = [[a in S]]$. By the $[[Void]]$
+  $[[a]]$. We have $[[a {idtm}]] = [[a in S]]$. By the $[[Void]]$
   case of the inversion property (Lemma~\ref{lemma:interpinv}), we
   know that $[[S]]$ must be the empty set and therefore contradicts
-  the assupmtion that $[[a in S]]$. Done.
+  the assumption that $[[a in S]]$.
 \end{proof}

### 2024-02-24 paper/paper.tex
Rename Coherence to convertibility

@@
 to recover single substitution by composing the extension operator and
-the identity substitution: $[[a { b }]] := [[a { idtm .: b }]]$.
+the identity substitution: $[[a { b }]] := [[a { idtm .: b }]]$. Since
+the definition of semantic typing in Section~\ref{sec:logrelproof}
+relies on simultaneous substitution, we find it more convenient to
+formulate simultaneous substitution directly rather than recovering it
+from single substitution.
 
@@
 \drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
-\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
+\drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
 \end{minipage}
@@
 equational theory in an algorithmic style using the notion of parallel
-reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted). Parallel reduction takes
+reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted).
+ Parallel reduction takes
 the form $[[a => b]]$ and we denote its transitive as $[[a =>+
@@
 \end{lemma}
-\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
+\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
   \label{lemma:coherencerefl}
@@
 \end{lemma}
-\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
+\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
   \label{lemma:coherencesym}
@@
 \end{lemma}
-\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
+\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
   \label{lemma:coherencetrans}
@@
 From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
-\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
+\ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
 equivalence relation. It is then trivial to show that all untyped
@@
 \citet{takahashi-parallel-reduction}.
+
+Untyped equality allows us to show the
+equivalence between our more structured notion of convertibility
+ ($[[a <=> b]]$) and the declarative, untyped $\beta$-equivalence
+independently from the type system and the logical relation through
+well-established syntactic approaches. The directedness of the
+convertibility relation will be leveraged in
+Section~\ref{sec:logreldep} to justify the conversion rule of our
+typing judgment, which we will introduce shortly.
 
@@
 \end{figure}
-% Figure~\ref{fig:par} shows the definition of the parallel reduction
-% relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
-% represent its transitive and reflexive closure, which in turn allows us to define
-% the coherence relation $[[a <=> b]]$.\scw{where is the terminology ``coherence''
-% from? I'm not familiar with it. Maybe you can replace it with ``definitional
-% equivalence'' as that is how it is used in \lang?}
-% We say that two terms $[[a]]$
-% and $[[b]]$ are coherent if they can eventually reduce to some common
-% term $[[c]]$ through parallel reduction. The symmetric notation of
-% coherence suggests\scw{but you prove it below?} it is an equivalence relation.
-
-% We sketch out some key properties about parallel reduction and
-% coherence without giving their proofs. Our technique for establishing
-% those results is based on
-
-% \scw{cite barendregt?}
-% \yl{resolved}
-
-% TODO: remove the lengthy discussion below
-
-% Now, we prove that coherence is indeed an equivalence relation.
-
-% First, we show that coherence is reflexive through the following
-% sequence of lemmas.
-
-% Lemma~\ref{lemma:parrefl} can be proven by structural induction over
-% the term $[[a]]$. Lemmas~\ref{lemma:parsrefl} and \ref{lemma:coherencerefl}
-% immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.
-
-% The reflexivity of parallel reduction enables us to embed rules from
-% call-by-name semantics into parallel reduction, as the following lemma
-% shows.
-% \begin{lemma}[Par AbsCbn\footnote{\dotv{join.v}{P\_AppAbs\_cbn}}] For all $[[A]], [[a]],$ and $[[b]]$,
-%   \label{lemma:parabscbn}
-%   $[[(\ A a) b => b {a}]]$
-% \end{lemma}
-% \begin{proof}
-%   Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
-% \end{proof}
-
-% Symmetry of coherence immediately falls from its definition.
-
-% Before we can prove transitivity, we need to show that parallel
-% reduction satisfies the diamond property.
-
-% The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
-% structural induction over the derivation of $[[a0 => a1]]$.
-% Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
-% over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
-% the use of Lemma~\ref{lemma:parcong}.
-
-% From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
-% single substitution property as a simple corollary.
-
-% A relation that satisfies the
-% diamond property must also be confluent, meaning that its transitive
-% and reflexive closure is confluent.
-% \begin{lemma}[Par confluent\footnote{\dotv{join.v}{pars\_confluent}}]
-%   \label{lemma:parconfluent}
-%   If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
-%   $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
-% \end{lemma}
-% While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
-% it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a => b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).
-
-% The transitivity of the coherence relation follows as a corollary of
-% Lemma~\ref{lemma:parconfluent}.
-% By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
-% $[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
-% $[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
-% term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
-% to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
-% trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
-% This concludes the proof that coherence is an equivalence relation.
-% \begin{lemma}[Coherence Equivalence]
-%   \label{lemma:coherenceequiv}
-%   The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
-%   transitivity and therefore is an equivalence relation.
-% \end{lemma}
 
@@
 strengthen the inductive hypothesis of the fundamental theorem.
-% \scw{Is the precondition for T-ABS really admissible? Isn't it required
-% to ensure predicativity? Have you proved it? Have you proved regularity?}
-% \yl{I proved regularity. \Rref{T-Abs} is admissible if cumulativity
-%   was formulated correctly (which wasn't the case in the version
-%   of the paper you read). I revised \rref{T-Univ} so the premise is
-%   not admissible }
-% \scw{Do you have a formation rule for identity types?}
-% \yl{Added}
 
@@
 equality. This assures us that we do not lose much generality working
-with a system with untyped conversion. We include a detailed
-discussion in Section~\ref{sec:relatedwork} on how type-directed and
-untyped equality affect the applicability of our technique.
+with a system with untyped conversion.
+We include a detailed discussion in Section~\ref{sec:discuss} on how
+type-directed and untyped equality affect the applicability of our
+technique.
 
@@
 % our proof to handle those rules.
-\yl{Unresolved: is it ok to say our system's conversion rule is untyped like
-Coq?}
 % \scw{Maybe move more of this paragraph and the next to the discussion/related
@@
 }
+\yl{I don't think identity types and dependent function types should
+  be listed as reasons why the logrel can't be a recursive
+  function. System DE has both dependent function types and identity
+  types but the logical fragment doesn't have type-level computation
+  so it can be defined structurally by specifying some kind of size metric
+}
 \scw{ We also need to explicitly point out that our logical relation is untyped.
@@
 }
+
 Before we define our logical relation, we first formally specify the
@@
 simply typed languages, the logical relation is defined as a recursive
-function over the type $[[A]]$. In dependent types, the type
+function over the type $[[A]]$.
+In dependent types, the type
 $[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
@@
 \section{Existence of $\beta\eta$-normal forms}
+\label{sec:betaeta}
 We can easily incorporate the function $\eta$ law to the equational

### 2024-02-24 paper/paper.tex
Give a full explanation of the admissibility of i-pialt early on

@@
 Figure~\ref{fig:syntax}.
-\scw{Many introductory texts define the relation as a recursive function
-over type structure, or step-indices. You use an inductive relation instead, why?}
+% \scw{Many introductory texts define the relation as a recursive function
+% over type structure, or step-indices. You use an inductive relation instead, why?}
+%\yl{resolved}
 % \scw{Can we view this inductive relation as the graph of the partial function
@@
 
-\Rref{I-Pi} is the most interesting. To explain what it means, we
-need to perform a few steps of transformations to its precondition.
-The precondition consists of a
-mysterious relation $[[R]]$ over the set $[[S]]$, an interpretation
-of the type $[[A]]$, and returns a subset of terms. $[[R]]$ is further
-subject to the following two constraints:
-\begin{itemize}
-\item $[[forall a, (# exists S0  , ( a , S0 ) in R #)]]$
-\item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
-\end{itemize}
-The first condition allows us to induce a function $[[F in S ->
-PowerSet STm]]$ such that $[[(a , F a) in R]]$. Then we can easily
-show that $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$.
-
-For the other direction, given a function $[[F]]$ satisfying the
-property $[[forall a, (# a in S implies Interp I i B { a } F a #) ]]$, it
-is possible to  define a relation $[[R]]=[[
-{ ( a , F a )  | a in S }]]$. It is easy to verify that $[[R]]$ satisfy
-the the constraint required in the precondition of \rref{I-Pi}.
-
-Furthermore, the skolemization process allows us to conclude the
-following equivalence:
-\[ \exists F, [[forall a, (# a in S implies Interp I i B { a } F a #)]] \iff [[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]  \]
-% Ignoring the content below the horizontal bar of
-% the derivation, we claim the statements from the precondition are equivalent to
-% the conjunction of the following statements:
-% \begin{itemize}
-% \item $[[Interp I i A S]]$
-% \item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
-%   [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
-% \end{itemize}
-% In general, let $R$ be a binary relation over the sets $[[A]]$ and
-% $[[B]]$, it easy to verify the following equivalence, which is used to
-% justify the skolemization process~\citep{skolemization}.
-% \scw{Is this true constructively? If not, is that ok?}
-% \[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
-%   \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
-%   \in R\]
-% The left-hand side is equivalent to the right-hand side, but has no
-% mentioning of any functions.
-The precondition of \rref{I-Pi} therefore can be formulated as the
-following statements:
-\begin{itemize}
-\item $[[Interp I i A S]]$
-\item $[[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]$
-\end{itemize}
-From this alternative formulation, we can more easily tell what is
-happening in \rref{I-Pi}: the
-function type $[[Pi A B]]$ has an interpretation if its input
-type $[[A]]$ can be
-interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
-type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
-$[[B]]$, has some semantic interpretation.
-
-If we look at $[[Interp I i Pi A B { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$, the conclusion of \rref{I-Pi}, we see why the
-presentation with an explicit $[[R in S * PowerSet STm]]$ is
-useful. We want to state the fact that $[[B {a}]]$ has a semantic
-interpretation of some set $[[S0]]$, but we also want to refer to the
-set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
-to. With the relation $[[R]]$ in scope, we can refer to the
-witness sets $[[S0]]$ such that $[[Interp I i B {a} S0]]$ for $[[a in S]]$.
-With the alternative representation that is free of the relation $[[R]]$, we
-might want to reformulate our logical relation in the following more
-concise but equivalent form, whose admissibility follows directly from
-\rref{I-Pi}.
+\Rref{I-Pi} is the most complicated rule in our logical
+relation. Instead of explaining it directly, we consider the following
+simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
 \begin{center}
@@
 \end{proof}
+Compared to the \rref{I-Pi}, which contains the mysterious relation
+$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
+a well-behaved dependent function type. The precondition of
+\rref{I-PiAlt} says that
+the function type $[[Pi A B]]$ has an interpretation if its input
+type $[[A]]$ can be
+interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
+type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
+$[[B]]$, has some semantic interpretation. In its conclusion, we say
+that a term $[[b]]$ belongs to an interpretation of $[[Pi A B]]$ if
+for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
+application form $[[b a]]$ belongs to all possible interpretations of
+$[[B {a}]]$ (the pre-condition ensures at least one interpretation
+exists for each $[[B {a}]]$ where $[[a in S]]$).
+
+In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
+these two rules are equivalent in the sense that every derivation of
+involving \rref{I-Pi} can be systematically replaced by
+\rref{I-PiAlt}. This equivalence follows directly from the fact
+that the logical relation is a partial function, a result we will show
+in Lemma~\ref{lemma:logreldeter}. The preconditions of \rref{I-Pi},
+when combined with the functionality of the logical relation, uniquely
+determine the relation $[[R in S * PowerSet STm]]$ to be the relation
+$[[{ (a , S0 ) | a in S , Interp I  i B { a } S0 }]]$. This result is
+formally shown through the improved inversion lemma for function types
+(Lemma~\ref{lemma:piinvalt}).
 
 Unfortunately, we cannot define the function case of our logical
-relation directly using \rref{I-PiAlt} since it not only violates the
+relation directly using \rref{I-PiAlt} since the occurrence of
+$[[Interp I i B {a} S0]]$ in its conclusion not only violates the
 syntactic strict positivity constraint required in proof assistants,
-but is genuinely non-monotone when written as an endofunction over the
-domain of relations.
+but is genuinely non-monotone when we rewrite the inductive definition
+as an endofunction over the domain of relations.
 Intuitively, the failure of monotonicity stems from the fact
 that the witness picked in the precondition is not necessarily the
-same witness being referred to in the post condition. While it might
+same witness being referred to in the post condition as the relation
+grows, whereas the
+relation $[[R]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ and
+prevents the set of witnesses from growing. While it might
 be possible to restrict the domain with additional constraints such as
-functionality and inversion properties, we opt for our current
+functionality and inversion properties to justify the existence of a
+fixed point with \rref{I-PiAlt}, we opt for our current
 formulation of \rref{I-Pi} so we immediately obtain a
@@
 function $[[R]]$ each time we apply it, though this is mitigated by
-the immediate admissibility of \rref{I-PiAlt}.
+the immediate admissibility of \rref{I-PiAlt} and the inversion
+principle (Lemma~\ref{lemma:piinvalt}) we derive after proving functionality.
 
@@
   we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
-  B0 S]]$. By \rref{I-Par} and $[[B1 => B]]$, we conclude that
+  B0 S]]$. By \rref{I-Red} and $[[B1 => B]]$, we conclude that
   $[[Interp I i B S]]$.
@@
   $[[A in I j]]$ if only if there exists some $[[S]]$ such that
-  $[[Interp I i A S]]$. By \rref{I-Par}, we must also have $[[B in I
+  $[[Interp I i A S]]$. By \rref{I-Red}, we must also have $[[B in I
   j]]$ for all $[[B => A]]$.
@@
 % \begin{proof}
-%   Immediately from \rref{I-Ne} and \rref{I-Par}.
+%   Immediately from \rref{I-Ne} and \rref{I-Red}.
 % \end{proof}
@@
 In terms of our proof technique, we close over our semantically valid types
-and terms in \rref{I-Par, I-Bool} using the non-deterministic parallel
+and terms in \rref{I-Red, I-Bool} using the non-deterministic parallel
 reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak

### 2024-02-25 paper/paper.tex
Take a pass and fix minor things

@@
 Our goal with this work is to demonstrate how the proof technique of
-\emph{logical relations} can be applied to reason about dependent type theories.
-Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
+\emph{logical relations} can be applied to reason about dependent type
+theories.
+Tutorial material on this topic~\citep{skorstengaard2019introduction,
+  harpertait, harperkripke, pierce2002types, pierce2004advanced,harper2016practical} is primarily focussed on
 systems with simple or polymorphic types. In that context, logical relations
@@
 so a young researcher might be excused for thinking that proofs that use
-logical relations are not applicable.  \scw{Well, actually, we did use a
-  step-indexed relation for Trellys\ldots}\yl{The core calculus of Trellys'
-  logical fragment is simple type + identity so I don't think that counts..}
+logical relations are not applicable.
 But this is not the case. Recent authors have developed tour-de-force
@@
 type-level computation, to demonstrate the logical strength of the
-approach. \scw{Say that we don't have impredicativity, but neither does anyone
-  else?}\yl{Nobody else but Barras. I'll find the citation}
+approach. % \scw{Say that we don't have impredicativity, but neither does anyone
+  % else?}\yl{Nobody else but Barras. I'll find the citation}
 %%
@@
 \end{lemma}
-From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
+From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
 \ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
@@
 n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
-the full context $[[G]]$ rather than a truncated version of $[[G]]$.
+the full context $[[G]]$ rather than a truncated version of
+$[[G]]$. In \rref{T-J}, the elimination form for the idnetity type,
+$[[a]]$ and $[[A]]$ need to be shifted in the context for a similar
+reason; an extra type $[[A]]$ appears after the $[[G]]$ and therefore
+the terms $[[a]]$ and $[[A]]$ in the identity type both need to be shifted
+to refer to the same binders.
 
@@
 
-To derive consistency, it suffices to restrict $[[S]]$ to the set of
-terms that reduce to closed terms. \Rref{I-Void, I-Bool} are
-unsurprising; when considering only closed terms, the empty type
-should not be inhabited and therefore corresponds to the empty set,
-whereas the boolean type is semantically inhabited by terms that
-evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
+% To derive consistency, it suffices to restrict $[[S]]$ to the set of
+% terms that reduce to closed terms.
+\Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
+of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
+example, the $[[Void]]$ type should not have any inhabitants under the
+empty context, where as the $[[Bool]]$ type corresponds to terms that
+evaluate to $[[true]]$ or $[[false]]$. However, the characterization
+of $[[Bool]]$ in our logical relation does not always correspond to
+well-typed or even closed terms. For example, we have $[[if false var 0
+true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
+  true \/ a =>+ false }]]$ since it evalutes to $[[true]]$.
+The independence of syntactic typing in our logical relation allows
+our semantic typing definition in Section~\ref{sec:logrelproof} to be
+meaningful on its own. Furthermore, not having to embed scoping
+information into the logical relation avoids extra bookkeeping and the
+need for a Kripke-style logical relation.
+\yl{Not sure what to cite from Derek Dreyer. I know his blog post
+  about semantic type soundness but is there a good paper to cite? one
+of the rust papers?}
+
+% unsurprising; when considering only closed terms, the empty type
+% should not be inhabited and therefore corresponds to the empty set,
+% whereas the boolean type is semantically inhabited by terms that
+% evaluate to the boolean values $[[true]]$ or $[[false]]$.
+\Rref{I-Eq}
 says that an equality type $[[a ~ b : A]]$ corresonds to the
@@
 \end{center}
-\begin{lemma}
+\begin{lemma}[I-PiAlt admissibility]
   \label{lemma:piintroalt}
@@
 be possible to restrict the domain with additional constraints such as
-functionality and inversion properties to justify the existence of a
-fixed point with \rref{I-PiAlt}, we opt for our current
+functionality and inversion properties to justify the well-definedness of our
+inductive relation with \rref{I-PiAlt}, we opt for our current
 formulation of \rref{I-Pi} so we immediately obtain a
@@
 function $[[R]]$ each time we apply it, though this is mitigated by
-the immediate admissibility of \rref{I-PiAlt} and the inversion
-principle (Lemma~\ref{lemma:piinvalt}) we derive after proving functionality.
+the immediate admissibility of \rref{I-PiAlt} and the alternative $\Pi$ inversion
+principle (Lemma~\ref{lemma:piinvalt}), which we derive after proving functionality.
 
@@
 the relation $[[R]]$, analogous to the admissible rule \rref{I-PiAlt}.
-\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
+\begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
   \label{lemma:piinvalt}
@@
 $\beta$ normal forms for (open and closed) well-typed terms.
-\scw{Why is this a sketch? We've done the whole proof and we will
-explain it all, right?}\yl{s/sketch out/show. I used sketch because we
-didn't spell out the full proofs}
 
@@
   \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
-  \caption{Extended logical relation}
+  \caption{Extended logical relation (new and changed rules)}
   \label{fig:logrelopen}
@@
 \end{lemma}
+Lemma~\ref{lemma:parnenf} could have been strengthened to say that if
+$[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
+$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes,
+parallel reduction can not take any real reduction steps and can only
+step into a term itself. However, for the purpose of our proof,
+Lemma~\ref{lemma:parnenf} is sufficient. Furthermore,
+Lemma~\ref{lemma:parnenf}, in its current weaker form, would still hold after we extend our vequational
+theory with the function $\eta$ rule, where parallel reduction can
+take $\eta$ steps but still preserves $\beta\eta$-normal form.
 
@@
   By induction over the length of the reduction sequence in $[[wn (a
-  var i)]]$. The conclusion follows from Lemma~\ref{lemma:parantirenaming} and
+  var i)]]$. The conclusion follows from Lemmas~\ref{lemma:parantirenaming} and
   \ref{lemma:parnenf}.
@@
   numbers less than $[[j]]$. We perform another induction on
-  $[[j]]$. All cases are trivial except for the case for
+  the derivation of $[[Interp I j A S]]$. All cases are trivial except for the case for
   \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
@@
 The formulation of the valuation and semantic well-typedness from
-Figure~\ref{fig:semtyping} and the fundamental lemma remains
+Section~\ref{sec:logrelproof} and the fundamental lemma remains
 unchanged.
@@
 
-Due to the non-deterministic nature of parallel reduction, we need to take a
-few more steps to convert the existence of $\beta$-normal form into a decision
-procedure for type conversion. More specifically, we can show that a
-deterministic evaluation strategy such as leftmost-outermost reduction can
-always find the $\beta$-normal form if there exists one. Then the termination
-of that strategy immediately follows from Corollary~\ref{corollary:exbetanf}
-However, we omit such proofs since they can can be formulated on untyped
-lambda terms and thus are orthogonal to the specifics of dependently typed
-systems. Instead, we redirect readers to \citet{factorization-essentially,
-  takahashi-parallel-reduction} for the details.
-
-\scw{Name and foreshadow these lemmas when they are introduced so that
-it is easier to refer to them in the discussion.}
+\yl{commented out the following paragraph because I accidentally wrote
+the same thing twice. Will do a comparison and if this is better
+written I'll replace the part from the $\beta\eta$ section with it}
+% Due to the non-deterministic nature of parallel reduction, we need to take a
+% few more steps to convert the existence of $\beta$-normal form into a decision
+% procedure for type conversion. More specifically, we can show that a
+% deterministic evaluation strategy such as leftmost-outermost reduction can
+% always find the $\beta$-normal form if there exists one. Then the termination
+% of that strategy immediately follows from Corollary~\ref{corollary:exbetanf}
+% However, we omit such proofs since they can can be formulated on untyped
+% lambda terms and thus are orthogonal to the specifics of dependently typed
+% systems. Instead, we redirect readers to \citet{factorization-essentially,
+%   takahashi-parallel-reduction} for the details.
+
 Finally, we want to point out that
@@
 sense to include the definition of ProdSpace? What is the Coq type of R?}
+\yl{I think I might just inline the definition of ProdSpace and then
+  rename the variables from the Coq code to match the ones from the paper}
 
@@
 property. Given two predicates $P$ and $Q$, we can prove that $\forall
-x, P(x) \iff Q(x)$, but we can not concldue that $P = Q$. While
+x, P(x) \iff Q(x)$, but we can not conclude that $P = Q$. While
 predicate extensionality is not necessary in our development, we
@@
 reduction relation. Suppose $[[a <=> b]]$, $[[a =>+ f0]]$, $[[b =>+
-f1]]$ but $[[f0]]$ is syntactically distinct from each
-other. By the transitivity of convertibility, we have $[[f0 <=> f1]]$.
+f1]]$ but $[[f0]]$ is syntactically distinct from $[[f1]]$. By the transitivity of convertibility, we have $[[f0 <=> f1]]$.
 Because $[[f0]]$ and $[[f1]]$ are both in
@@
 composing the existence of normal form and the fact that the
-deterministic relation is normalizing, we can derive show the
+deterministic relation is normalizing, we can derive the
 accessibility of the deterministic reduction relation
 (\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
-an induction metric to define an executable algorithm.
+an induction metric to define an executable algorithm\footnote{Coq's
+  singleton elimination principle allows \mintinline{coq}{Acc}, a
+  \texttt{Prop} data type with a single constructor, to be eliminated
+  to construct runtime relevant data}.
 
@@
 precise and thus easier to encode in proof assistants. Our definition
-of logical relation is reminiscent to their definition of a
+of logical relation is reminiscent of their definition of a
 semantic universe hierarchy, though our logical relation is defined to
@@
   need only show that terms have $\beta\eta$ normal forms. }
+\yl{Makes perfect sense!}
 

### 2024-02-25 paper/paper.tex
Add the discussion about parallel reduction vs lo/wh reduction

@@
 
-In terms of our proof technique, we close over our semantically valid types
-and terms in \rref{I-Red, I-Bool} using the non-deterministic parallel
-reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak
-head reduction relation. Our use of a non-deterministic reduction strategy
-means that we need confluence to prove the functionality of our logical
-relation. However, the benefit is that it immediately gives us a semantic
-justification of the conversion rule (Lemma~\ref{lemma:logrelcoherence}).
+On the other hand, our definition of our logical relation does
+contribute to a more concise proof.
+In \rref{I-Red, I-Bool}, we choose parallel reduction, a full
+reduction relation, to close over our semantic interpretation of types
+and terms. Parallel reduction is non-deterministic, but it satisfies
+useful structural properties such as congruence
+(Lemma~\ref{lemma:parcong}) and the diamond property
+(Lemma~\ref{lemma:pardiamond}). We pay the price of using a
+non-deterministic reduction relation when we want to prove that our
+logical relation is a partial function; because of \rref{I-Red}, we
+can have $[[A => B0]]$ and $[[A => B1]]$, where $[[B0]]$ and $[[B1]]$
+each have their separate interpretations that we have to prove to be
+equal. Fortunately, this complexity is easily reconciled by the
+diamond property, which is easy to derive syntactically.
+\citet{decagda,nbeincoq} employ a deterministic weak
+head reduction relation. The use of a deterministic reduction relation
+makes the functionality of a logical relation trivial to prove, but
+the downside is that weak head reduction (denoted by $\leadsto^{wh}$)
+or leftmost-outermost reduction (denoted by $\leadsto^{lo}$) does
+not satisfy even the weaker substitution property
+(Lemma~\ref{lemma:parsubst}). That is, given a term $[[A]]$
+and $a \leadsto^{lo} b$, we do not necessarily have $[[A {a}]]
+\leadsto^{lo} [[A {b}]]$ since the variable might appear to the
+right of some redex. Eventually, one has to prove how the
+deterministic reduction relates to a full non-deterministic reduction
+relation to prove the fundamental theorem. In our setup, this would amount
+to proving the factorization theorem of the deterministic reduction
+relation with respect to full non-deterministic reduction. Instead,
+using parallel reduction in \rref{I-Red} allows us to delay the
+factorization property until we need to extract a normalization
+algorithm from our fundamental theorem (Section~\ref{sec:conversionalgo}).
+
+% In terms of our proof technique, the choice of parallel reduction, a full reduction
+% relation, to close over our semantic interpretation in \rref{I-Red,I-Bool} has a non-negligible effect on the
+% size of our development.
+
+% we close over our semantically valid types
+% and terms in \rref{I-Red, I-Bool} using the non-deterministic parallel
+% reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak
+% head reduction relation. Our use of a non-deterministic reduction strategy
+% means that we need confluence to prove the functionality of our logical
+% relation. However, the benefit is that it immediately gives us a semantic
+% justification of the conversion rule (Lemma~\ref{lemma:logrelcoherence}).
 % In particular, we obtain the confluence
@@
 
-% Finally, it is worth pointing out that all systems we have discussed
-% so far builds a relational model for their logical relation.
-
-% Recall
-% that our logical relation takes the form $[[InterpR i A S]]$ where
-% $[[S]]$ is a set of terms. In a relational model, each type is
-% interpreted as a partial equivalence relation over terms rather than a
-% set. A relational model is necessary for an extensional type theory
-% such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
-% use a relational model either to justify the function $\eta$-law or to
-% derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
-% law for function is baked into the untyped reduction
-% relation and therefore a relational model is not needed.
-% \scw{would it be difficult to extend your proof to be a PER?}
-
-% The dependently typed systems we have discussed so far also vary
-% greatly in expressiveness. For example, the Calculus of
-% Constructions~\citep{CoC} and MLTT without universe levels are unable
-% to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
-% identity types. % \citet{nbeincoq} lacks $\eta$-law for functions.
-% Among
-% the mechanized systems we have discussed so far, only
-% \citet{anand2014towards} has a full universe hierarchy.
-% Our object language \lang{}, while small, has a decent coverage of
-% features commonly seen in dependently typed languages, including
-% dependent pattern matching, dependent function types, a full universe
-% hierarchy, an intensional identity type. We hope features such as
-% type-directed reduction (necessary for unit $\eta$-law) and
-% impredicative sorts can be implemented in a similarly streamlined
-% fashion and we will leave those extensions as part of our future work.
-
-% To end this section, we
-% \scw{should this be a new subsection?}
-% The final question we want to address is: why is our proof,
-% even with the extension from Section~\ref{sec:extension}, so much
-% shorter than the proofs from \citet{decagda, nbeincoq,
-%   martin-lof-a-la-coq}?
-
-% This contrasts our development with
-% \citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
-% difficult property to prove due to the complexity of the equality
-% judgment used to define the logical relation. In our system, the
-% irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
-% direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
-% lemma name} which is
-% straightforward after we prove the results about parallel
-% reduction in Section~\ref{sec:spec}. In other words, we keep the
-% the logical relation itself minimal at the cost of extra lemmas that can be proven
-% through syntactic means independently from the logical relation.
-% For example, as commented near the end of Section~\ref{sec:extension},
-% our normalization property (Corollary~\ref{corollary:exbetanf})
-% % can induce a decision procedure for type conversion by reducing two
-% % terms to $\beta\eta$-normal form and check for their syntactic
-% % equivalence. This decision procedure would be inefficient compared to
-% % NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
-% % to weak head normal and checks for the equivalence of subterms
-% % recursively
-% does not immediately induce an efficient decision procedure, but it is
-% possible to recover an efficient algorithm by reasoning about untyped
-% lambda terms.
-
-% the concision of
-% our proof can be largely attributed to the fact that we bake into the
-% logical relation only the minimum amount of information that is the
-% most conducive to the proof of the metatheoretic result.
-
-% Another key to our proof is the very early establishment of the confluence
-% property of the paralell reduction relation
-% (Lemma~\ref{lemma:pardiamond}). This property is used in
-% Lemma~\ref{lemma:interppreservation} to show that the interpretation
-% of a type is preserved under evaluation. That is, if $[[InterpR i A S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
-% In a system with typed directed reduction, the confluence result is
-% harder to establish since confluence requires subject reduction, which
-% circularly depends on $\Pi$-injectivity, a consequence of
-% confluence~\citep{siles2012pure}. There are different workarounds to
-% this problem. \citet{siles2012pure} proposes a syntactic approach by
-% defining a type system where the confluence result is
-% directly provable and later show that the system of interest is
-% equivalent to the system with the confluence property.
-% \citet{decagda} uses a relational model and the relational counterpart
-% of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
-% model is parameterized by a typed indexed equivalence relation and
-% is reused to prove derive different properties in their
-% development. Another perhaps simpler approach is to simply
-% extend the judgmental equality with rules about
-% $\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
-% reduction to be proven independently from confluence and
-% $\Pi$-injectivity can later be shown to be admissible. We find the
-% approach from \citep{weirich:systemd} the most lightweight as it
-% allows us to derive confluence early on even in systems with typed
-% conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}
-
-% In cases where the judgmental equality does not leverage type information heavily
-% (e.g. $\eta$-law for the unit type), it might be possible to show that
-% $\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
-% $[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
-% property holds (we believe this is likely true for both
-% \citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
-% terms of untyped parallel reduction to show the decidability of type
-% conversion. Otherwise, to fully model the system, we need to bake
-% typing information into our logical relation and we need to redefine
-% our logical relation in Kripke-style so we can relax the scoping
-% constraint in order to prove adequacy, though a relational model would
-% still be unnecessary as long as the $\eta$ laws are baked into the
-% typed directed relation. \scw{not sure I understand this last bit}
-
-% Since confluence plays such as key role in our proof, we are uncertain
-% how to generalize our technique to systems where confluence does not
-% hold, and we will investigate those scenarios in our future work.
-% \scw{What systems are you thinking of? Or do you mean more generally
-% that this technique only works for languages where the definition of
-% equality is based on joinability of parallel reduction?}
-
-% We are able to avoid the complexivity of a Kripke-style logical relation
-% by defining our logical relation in terms of untyped parallel
-% reduction, which does not impose any scoping constraint on the
-% terms.
-
-% Why is it that the justification of $\eta$ law
-% requires a Kripke-style relational model in all three developments
-% but not ours? Are we ``cheating'' by defining \lang{}'s equational
-% theory using an untyped conversion rule? Instead of answering the
-% questions one by one, we clarify the design of our type system and our
-% proof technique so the answers to those questions are self-evident.
-
-% The key to the shortness of
-
 \section{Conclusion}

### 2024-02-25 paper/paper.tex
add comment about verification of Coq's type checker
pass through section 5

@@
 In this section, we show how the logical relation from
-Section~\ref{sec:logreldep} can be extended to show the existence of
-$\beta$ normal forms for (open and closed) well-typed terms.
-
-\scw{The goal of this section is also to demonstrate that our logical relations
-  proof technique can be extended to reason about the (parallel) reduction of
-  open terms, not just the reduction of terms after closing
-  substitutions. This is important, in particular, for dependently-typed languages
-  because type checking involves reasoning about open terms. But even simple type
-  systems may wish to employ this sort of reasoning, in the case of, say, reasoning
-  about compiler correctness in open environments. (Does Amal do anthing like this?)
+Section~\ref{sec:logreldep} can be extended to show the existence of $\beta$
+normal forms for (open and closed) well-typed terms.  In other words, we prove
+that it is possible to repeatedly use the parallel reduction relation to
+reduce any term to its unique normal form, where no further (non-identity)
+reductions can be applied. This result can be used to show that our type
+conversion relation is decidable, as computation of the normal form will
+always terminate.
+
+The goal of this section is also to demonstrate that our logical relations
+proof technique can be extended to reason about the reduction properties of
+open terms, not just the reduction of terms after closing substitutions.
+Reasoning about open terms is particularly important for dependently-typed
+languages because type checking involves working with open terms.
+\scw{
+  Add when we can find a reference:
+  However, even non dependently-typed languages employ such techniques, 
+  especially in the case of relational semantics. 
 }
-
-The goal of this section is to prove that all well-typed terms have
-$\beta$-normal forms. In other words, we can repeatedly use our parallel
-reduction relation to reduce any term to a unique normal form, where no
-further reductions can be applied. This result can be used to show that our
-type conversion relation is decidable, as computation of the normal form will
-always terminate.
+While this extension employs well-known techniques, it continues to be short and 
+demonstrates the robustness of our initial framework.
+
+We begin this part with a description of the $\beta$-normal forms 
+of \lang{}.
 
@@
 The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
-neutral terms and normal forms with respect to $\beta$-reduction. We sometimes
+neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here?
+We can be specific and say that that the only reductions available for these terms are identity reductions and cite
+\footnote{\dotv{normalform.v}{nf\_refl}} }. Instead of the 
+metavariables $e$ and $f$, we also
 use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
@@
 follows.
-\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
-\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
-We can read $[[wne a]]$ and $[[wn a]]$ as the term $[[a]]$ weakly normalizes to a neutral or a normal form respectively.
+\[
+\begin{array}{llcl}
+\mbox{\emph{weakly normalizes to a neutral form}} & [[wne a]] &\iff& \exists [[e]], [[a =>+ e]] \\
+\mbox{\emph{weakly normalizes to a normal form}}  & [[wn a]]  &\iff& \exists [[f]], [[a =>+ f]] \\
+\end{array}
+\]
 
@@
   \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
-  \caption{Extended logical relation (new and changed rules)}
+  \caption{Extended logical relation (new and changed rules) }
   \label{fig:logrelopen}
@@
 
-The updated logical relation is shown
-in Figure~\ref{fig:logrelopen}.
+The updated logical relation is shown in
+Figure~\ref{fig:logrelopen}.~\footnote{\dotv{semtypingopen.v}{InterpExt}}
+There is one new rule in this figure, \rref{I-Ne}. In a non-empty context, a
+type itself may evaluate to a neutral term and in turn can only be inhabited
+by neutral terms.  Otherwise, the rest of the rules in this figure are updates
+to the analogous rules in Figure~\ref{fig:logrel}. Note that, we omit the
+rules for the function and universe cases because they are identical to the
+original version.
+
 % The idea of a term blocked from $\beta$ reduction is
@@
 % neutral terms.
-In the definition of the logical relation, we omit the rules for
-the function and universe cases since they remain identical to the
-original version in Figure~\ref{fig:logrel}.
-The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
-An open term of type $[[Bool]]$ does
-not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
-a variable, or more generally, a neutral term. Likewise, the
-$[[Void]]$ type, while remains uninhabited under an empty context, may
-be inherited by the set of neutral terms when there is a variable in
-the context that allows us to inhabit $[[Void]]$.
-
-The rule for equality type $[[a ~ b : A]]$ is augmented with the preconditions that
-$[[a]]$, $[[b]]$, and $[[A]]$ are all in their normal forms since
+The changes to \rref{I-Bool} and \rref{I-Void}
+follow the same pattern: an open term of type $[[Bool]]$ does not necessarily
+reduce to $[[true]]$ or $[[false]]$, but may reduce to a variable, or more
+generally, a neutral term. Likewise, while the $[[Void]]$ type remains
+uninhabited under an empty context, it may be inhabited when there is a
+variable in the context that has type $[[Void]]$ or that can be eliminated to
+type $[[Void]]$.
+
+The rule for equality type $[[a ~ b : A]]$ is augmented with the precondition that
+$[[a]]$, $[[b]]$, and $[[A]]$ are all normal forms because
 otherwise our model would include equality types that are themselves
-not normalizing. Furthermore, the side condition $[[a <=> b]]$ is only
-available when the equality proof reduces to $[[refl]]$. If the proof
-term reduces to a neutral term, then there is nothing we can learn
+not normalizing.
+\scw{I don't see this precondition in the figure!}
+Furthermore, the side condition $[[a <=> b]]$ is only
+required/available when the equality proof reduces to $[[refl]]$. If the proof
+term reduces to a neutral term, then there is nothing we need to show/can learn
 about the relationship between $[[a]]$ and $[[b]]$.
 
-Finally, in a
-non-empty context, a type itself may evaluate to a neutral term and in
-turn can only inhabited by neutral terms, thus the addition of \rref{I-Ne}.
-
-Interestingly, all the properties we have shown in
-Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the
-fundamental lemma can be proven in the exact same order, where the new
-cases due to \rref{I-Ne} and the augmentation of neutral terms
-to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
-Lemma~\ref{lemma:parnenf}.
-
-Before we can prove the fundamental theorem and derive the
-normalization property as its corollary, we need to additionally
-formulate and prove the adequacy property.
-
-\yl{Moved renaming back to Sec 5 because they are needed for the
-  non-eta development too}
-Let us first start with some auxiliary lemmas and definitions.
-\begin{lemma}[Par anti-renaming]
+Because we are working with open terms, we need a few additional syntactic
+lemmas about reduction. First, a renaming $\xi$ is a generalization of
+weakening when working with de Bruijn indices. It consistently renumbers the
+variables that appears in terms. If a renamed term has been reduced, we can
+always recover the result of the reduction without the renaming.
+%
+\scw{This lemma is only used to prove \texttt{wn\_antirenaming}, which is then
+used more generally. Maybe we should replace it with that?}
+\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
   \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
@@
 neutral forms.
-\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
+\begin{lemma}[Par preserves $\beta$-neutral and normal forms\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
   \label{lemma:parnenf}
@@
 $[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
-$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes,
-parallel reduction can not take any real reduction steps and can only
-step into a term itself. However, for the purpose of our proof,
-Lemma~\ref{lemma:parnenf} is sufficient. Furthermore,
-Lemma~\ref{lemma:parnenf}, in its current weaker form, would still hold after we extend our vequational
-theory with the function $\eta$ rule, where parallel reduction can
-take $\eta$ steps but still preserves $\beta\eta$-normal form.
-
-\begin{lemma}[Wn extensionality]
+$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes, parallel
+reduction can not take any real reduction steps and can only step into a term
+itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
+sufficient.  In fact, all the properties we have shown in
+Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the fundamental
+lemma can be proven in the same order, where the new cases due to \rref{I-Ne}
+and the augmentation of neutral terms to \rref{I-Void, I-Eq, I-Bool} can be
+immediately discharged by Lemma~\ref{lemma:parnenf}.
+
+Furthermore, Lemma~\ref{lemma:parnenf}, in its current weaker form, would
+still hold after we extend our equational theory with the function $\eta$
+rule, where parallel reduction can take $\eta$ steps but still preserves
+$\beta\eta$-normal form.
+
+We also need to know that the $[[wne a]]$ and $[[wn a]]$ relations can be
+justified compositionally. For example, an application has a neutral form when
+the function has a neutral form and the argument has a normal form.
+
+\begin{lemma}[Wne application\footnote{\dotv{normalform.v}{wne\_app}}]
+  \label{lemma:wnewn}
+  If $[[wne a]]$ and $[[wn b]]$, then $[[wne (a b)]]$.
+\end{lemma}
+\begin{proof}
+  Immediate by lexicographical induction over the length of the reduction sequences in
+  $[[wne a]]$ and $[[wn b]]$.
+\end{proof}
+
+Furthermore, if we know that an application of a term to a variable 
+has a normal form, then we know tha the term must have a normal form.
+\begin{lemma}[Wn extensionality\footnote{\dotv{normalform.v}{ext\_wn}}]
   \label{lemma:extwn}
@@
 
-\begin{lemma}[wne wn]
-  \label{lemma:wnewn}
-  If $[[wne a]]$ and $[[wn b]]$, then $[[wne a b]]$.
-\end{lemma}
-\begin{proof}
-  Immediate by lexicographical induction over the length of the reduction sequences in
-  $[[wne a]]$ and $[[wn b]]$.
-\end{proof}
-
-To state the adequacy lemma, we first define the notion
-of reducibility candidates (shortened as $CR$), predicates over sets
-of lambda terms that capture the properties we need from our logical
-relation. \scw{Need to provide intuition for these two properties that need
-to be proved similtaneously. Also, cite where this form of proof originally
+Before we can prove the fundamental theorem and derive the normalization
+property as its corollary, we need to additionally formulate and prove an
+\emph{adequacy property} about the logical relation.  This property, that the
+interpretation of each type is a \emph{reducibility candidate}, allows us to
+conclude that every term in each interpretation has a normal form. In the
+previous section, we only needed a property of the interpretation of the
+$[[Void]]$ type. However, for this section, we need to know something about
+the interpretation of every type.
+
+Furthermore, to prove this adequacy property, we need to strengthen it to also
+give us more information about neutral terms as we proceed by induction. In
+particular, we need to know that all terms that reduce to neutral forms are
+contained within the interpretation.
+\scw{Also, cite where this form of proof originally
 appeared? This part is standard.}
-\begin{definition}[CR]
-  Let $[[S]]$ be a set of lambda terms. We say that
+Therefore, we formally define when a set is a \emph{reducibility candidate}
+(shortened as $CR$) as follows.
+\begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
+  Let $[[S]]$ be a set of lambda terms. We say that $[[S]] \in CR$ 
+  if and only if conditions $CR_1$ and $CR_2$ hold.
   \begin{itemize}
-  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
-  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
-  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
+  \item $[[S]] \in CR_1 \iff\ [[forall a, (#  wne a implies a in S #)]]$ 
+  \item $[[S]] \in CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$   
   \end{itemize}
@@
 
-\yl{Merged all the separate adequacy related properties into a single
-  property proved through nested induction. (The mechanization has
-  been adjusted accordingly)}
-We now state and prove the adequacy lemma. We give some more details
-since we prove it through nested induction.
-\begin{lemma}[Adequacy\footnote{\dotv{soundnessopen.v}{soundness}}]
+We now state and prove the adequacy lemma.
+\begin{lemma}[Adequacy\footnote{\dotv{semtypingopen.v}{adequacy}}]
   \label{lemma:adequacy}
@@
   \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
-  A]]$. To derive $[[wn Pi A B]]$, it remains to show $[[wn B]]$. We
+  A]]$. To derive $[[wn (Pi A B)]]$, it remains to show $[[wn B]]$. We
   use the outermost induction hypothesis to show that $[[var 0]]$
-  semantically inhabits $[[A]]$, from which we derive $[[wn B {var 0}]]$
+  semantically inhabits $[[A]]$, from which we derive $[[wn (B {var 0})]]$
   and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
@@
 unchanged.
-The proof of the fundamental lemma is still carried out by induction
+The proof of the fundamental lemma\footnote{\dotv{soundnessopen.v}{soundness}} is still carried out by induction
 over the typing derivation, where the additional neutral term related
@@
 fundamental theorem.
-\begin{corollary}[Existence of $\beta$-normal form]
+\begin{corollary}[Existence of $\beta$-normal form\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
   \label{corollary:exbetanf}
@@
 
-Finally, we want to point out that
-the adequacy property (Lemma~\ref{lemma:adequacy}) are in fact required in the normalization proof
-for simply typed languages~\citep{abel2019poplmark}. Similar to the
-simply typed scenario, adequacy needs to be proven before the
+The extension of our logical relation to prove normalization of open
+\emph{and} closed terms closely mirrors the progression from
+normalization of closed terms~\citep{harpertait} to normalization of
+open terms~\citep{harperkripke} in the simply typed lambda calculus.
+Indeed, a mechanization of the latter proof appears in ~\citet{abel2019poplmark}.
+In this setting, as above, adequacy must be proven before the
 fundamental theorem so we can handle elimination rules such as
@@
 % in the exact same way to show the normalization of lambda forms
-Dependent types make the adequacy proof slightly more complicated
-as we also need to know that every type has a normal form, not just
-the terms.
-This complicates our proof specifically in the \rref{I-Set} case for
+%
+Dependent types make the adequacy proof slightly more complicated because we
+also need to know that every \emph{type} has a normal form, not just
+terms. This complicates our proof specifically in the \rref{I-Set} case for
 our adequacy property (Lemma~\ref{lemma:adequacy}).
 
-Overall, despite the dependently typed setting,
-the extension of our logical relation to prove normalization of open
-\emph{and} closed terms closely mirrors the progression from
-normalization of closed terms~\citep{harpertait} to normalization of
-open terms~\citep{harperkripke} in the simply typed lambda calculus.
-It is in fact reassuring that once we have laid the foundational
-technique for handling dependent types in our logical relation, the
-extension to open terms mostly boil down to properties that can be
-independently derived from the logical relation through syntactic
-means.
+Overall, despite the dependently typed setting, it is in fact reassuring that
+once we have laid the foundational technique for handling dependent types in
+our logical relation, the extension to open terms mostly boil down to
+properties that can be independently derived from the logical relation through
+syntactic means.
 
@@
   because they are trying to mechanize a full practical language}
+\scw{The Coq-Coq-Correct paper (extended version) includes a (predicative) universe hierarchy, universe polymorphism, inductive/coinductive types.
+But they don't show consistency.
+Their development is 300k LOC.}
 

### 2024-02-25 paper/paper.tex
Add feature matrix
more room for columns

@@
   ,anonymous=\ifanonymous true\else false\fi]{acmart}
+\usepackage{fontspec}
+\newfontfamily\boxedsymbols{DejaVu Sans}
 \usepackage{ottalt}
@@
 
+\def\rot{\rotatebox}
+\newcommand\header[1]{\rotatebox{90}{{#1}}}
+
+\begin{figure}[h]
+
+% \begin{minipage}{0.8\textwidth}
+  \begin{tabular}{ l |  l  | l | l | l }
+      & \header{Universes} & \header{Inductives} & \header{Convertibility}
+      & \header{Large Elim} \\
+    \hline
+    \lang{} & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
+    \hline
+    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} \\
+    \hline
+    POPLmark2 & 1 & Nat & T & {\boxedsymbols ✓} \\
+    \hline
+    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} \\
+    \hline
+    Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
+    \hline
+    Coq-in-Coq & $\mathbb{P}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}
+    %   & \lang{} & \citet{anand2014towards} &
+    %                                          \citet{barras2012semantical}
+    % & \citet{martin-lof-a-la-coq}  \\
+    % \hline
+    % Universe & Countable & Countable & Impredicative + Countable & One\\
+    % % \hline
+    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
+    % \hline
+    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
+    % \hline
+    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
+  \end{tabular}
+% \end{minipage}
+
+  \begin{tabular}{l}
+  \\
+    NBE-in-Coq \citet{nbeincoq} \\
+    POPLmark reloaded \citet{decagda} \\
+    MLTT-a-la-Coq \citet{martin-lof-a-la-coq} \\
+    Nuprl \citet{anand2014towards} \\
+  \\ 
+  Universes: Countable ($\mathbb{N}$), One (1), Impredicative ($\mathbb{P}$) \\
+  Convertibility: Typed (T), Untyped (U), Extensional (E) \\
+  \end{tabular}
+
+  \caption{Overview of Language Features}
+  \label{fig:linecount}
+\end{figure}
+
 \scw{Maybe it would be useful to include a chart here, so that readers can

### 2024-02-25 paper/paper.tex
some subsections to organize related work
Extend the feature matrix with Zombie

@@
     \lang{} & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
-    \hline
+    % \hline
+    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗}\\
     NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} \\
-    \hline
-    POPLmark2 & 1 & Nat & T & {\boxedsymbols ✓} \\
-    \hline
+    % \hline
+    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} \\
+    % \hline
     MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} \\
-    \hline
-    Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
-    \hline
-    Coq-in-Coq & $\mathbb{P}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}
+    % \hline
+    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
+    % \hline
+    CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
     %   & \lang{} & \citet{anand2014towards} &
@@
   \\
+    $\lambda^\theta$ (logical fragment only) \citet{casinghino:combining-proofs-programs} \\
     NBE-in-Coq \citet{nbeincoq} \\
-    POPLmark reloaded \citet{decagda} \\
+    $\lambda^{\Pi U\mathbb{N}}$ \citet{decagda} \\
     MLTT-a-la-Coq \citet{martin-lof-a-la-coq} \\
-    Nuprl \citet{anand2014towards} \\
+    Core Nuprl \citet{anand2014towards} \\
+    CIC \citet{barras2012semantical} \\
   \\
-  Universes: Countable ($\mathbb{N}$), One (1), Impredicative ($\mathbb{P}$) \\
+  Universes: Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
   Convertibility: Typed (T), Untyped (U), Extensional (E) \\
@@
 
-  \caption{Overview of Language Features}
+  \caption{Feature matrix for dependently typed languages with
+    mechanized proofs by logical relation}
   \label{fig:linecount}
@@
 
+\subsection{Logical relations}
 Proof by logical relations does not come with a precise definition. In
@@
 
+\subsection{Mechanized logical relations about dependent types}
+
 \citet{decagda} mechanizes in Agda the decidability of type
@@
 our logical relation.
+
+\subsection{Other mechanized metatheory of dependent types}
 

### 2024-02-26 paper/paper.tex
Add more related work

@@
 \begin{equation*}
-    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
+    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I j]]
+    := [[{A | exists S , InterpR j A S}]] \text{ for } [[j < i]]
 \end{equation*}
@@
 interpretation of the $[[i]]_{th}$ universe is only dependent on
-universes at strictly lower than $[[i]]$, which have been defined earlier.
+universes strictly lower than $[[i]]$, which have been defined earlier.
 Removing the ordering constraint would result in a
@@
 b]]$ are typically required for indexed types, of which equality types
-are an instance. \Rref{I-Par} enables us to reduce our types in order
+are an instance. \Rref{I-Red} enables us to reduce our types in order
 to assign meanings. Recall the type expression $[[(\ var 0)
-Bool]]$. \Rref{I-Par} says to know that
+Bool]]$. \Rref{I-Red} says to know that
 $[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
@@
 Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
-  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Par}
+  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
 and \rref{I-Bool}.
@@
     mechanized proofs by logical relation}
-  \label{fig:linecount}
+  \label{fig:featurematrix}
 \end{figure}
 
-\scw{Maybe it would be useful to include a chart here, so that readers can
-  easily keep track of the features of the various languages.  i.e. which ones
-  include large eliminations? type-directed equivalence? impredicative prop?
-  inductive datatypes? what are their line counts?}
+% \scw{Maybe it would be useful to include a chart here, so that readers can
+%   easily keep track of the features of the various languages.  i.e. which ones
+%   include large eliminations? type-directed equivalence? impredicative prop?
+%   inductive datatypes? what are their line counts?}
+% \yl{resolved}
 
@@
 \scw{Should give Martin L\"of credit for first consistency proof.}
-\scw{What techniques do these three proofs use? How is it different from
-your proof? What about Luo, which
-introduced a universe hierarchy? What about categories with families, or
-other ways of showing consistency?}
+% \scw{What techniques do these three proofs use? How is it different from
+% your proof? What about Luo, which
+% introduced a universe hierarchy? What about categories with families, or
+% other ways of showing consistency?}
+% \yl{resolved}
 
@@
 derive the correctness of their conversion algorithms. In our work, we
-use the semantic universe hierarchy directly because it is sufficient
+use the semantic universe hierarchy directly in our definition of
+semantic typing because it is sufficient
 for deriving consistency and later normalization.
@@
 \subsection{Mechanized logical relations about dependent types}
+In Figure~\ref{fig:featurematrix}, we give an overview of features of
+the object languages from mechanized logical relations about dependent
+types. The table is far from exhaustive. For example,
+\citet{casinghino:combining-proofs-programs} and
+\citet{anand2014towards} both have support for partial
+programs. However, we include features that we believe to be most
+impactful to the definition of the logical relation.
+
+\citet{casinghino:combining-proofs-programs} introduce $\lambda^\theta$, a dependently typed
+programming language that uses modality to distinguish between logical
+proofs and programs. The programmatic fragment
+includes recursive data types and supports general recursion at the
+cost of introducing divergence. Its logical fragment supports a
+proof-irrelevant identity type, which can be used to reason about the
+behavior of its programmatic fragment.
+However, the $\lambda^\theta$
+language is limited in its expressiveness since it does not support
+type-level computation or polymorphism and therefore is not a fully dependently
+typed language according to \citet{abel2013normalization}.
+The consistency proof of $\lambda^\theta$'s logical fragment has been
+mechanized in Coq through a step-indexed logical relation;
+step-indexing is required to model the programmatic fragment, which
+interacts with the logical fragment.
+The lack of polymorphism and type-level computation means their
+logical relation can be defined recursively for well-formed types using
+a size metric, which has been explored in~\citet{liu2023dependently}.
 
@@
 of mechanizing mathematical objects such as sets and domains.
+
+There are other mechanized developments for dependently typed systems
+that only involve properties that are derivable through syntactic
+means. For example, \citet{coqcoqcorrect2019} proves the correctness
+of a type checker for the Polymorphic, Cumulative Calculus of
+Inductive Constructions (PCUIC), Coq's core calculus, assuming the
+strong normalization property of the object
+language. \citet{weirich:systemd} define System D, a core calculus of
+dependent Haskell, and prove the syntactic type soundness of the type
+system. Compared to properties such as consistency, normalization, and
+parametricity, syntactic type soundness is less intimidating for
+dependent types since one is most unlikely to run into issues related
+to the strength of the metatheory.
 

### 2024-02-26 paper/paper.tex
Shrink to 25 pages

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Short and Mechanized Logical Relation for Dependent Type Theory}
+\title{Short and Mechanized Logical Relation for Dependent Types}
 
@@
 presentation of proof by logical relation for a dependently language
-is lacking. In this paper, I present a short
+is lacking. In this paper, we present a short
 consistency proof for a dependently typed language that contains a
-rich set of features, including a full cumulative universe
-hierarchy, booleans, and an intensional identity type. We have
+rich set of features, including a full universe
+hierarchy, booleans, and an intensional identity type. We show that
+the logical relation can be easily extended to prove the existence of
+$\beta\eta$-normal forms.
+We have
 fully mechanized the consistency proof using the Coq proof assistant
-in under 1000 lines of code.
+in under 1000 lines of code, with 500 lines of additional code for the
+$\beta\eta$-normal form extension.
 \end{abstract}
@@
 language is small, but includes enough to be illustrative. For example, we
-eschew inductive datatypes or W-types, the but we do include propositionality
+eschew inductive datatypes or W-types, but we do include propositionality
 equality and booleans in order to capture the challenges presented by indexed
@@
 and play a central role in defining our logical relation. The design
-of \lang{} is reminiscent of Martin-Löf style predicatvie type
+of \lang{} is reminiscent of Martin-Löf style predicative type
 systems. The untyped conversion rule of \lang{} is inspired by Pure
@@
 \drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
-\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
-\drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
+% \drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
+% \drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
 \end{minipage}
-\caption{Parallel reduction ($\beta$-rules only) and convertibility }
+\caption{Parallel reduction ($\beta$-rules only)  }
 \label{fig:par}
@@
 the full context $[[G]]$ rather than a truncated version of
-$[[G]]$. In \rref{T-J}, the elimination form for the idnetity type,
-$[[a]]$ and $[[A]]$ need to be shifted in the context for a similar
+$[[G]]$. In \rref{T-J}, the elimination form for the identity type,
+the terms $[[a]]$ and $[[A]]$ need to be shifted in the context for a similar
 reason; an extra type $[[A]]$ appears after the $[[G]]$ and therefore
@@
 
-The complexity of applying proof by logical relation to dependent types stem
+The complexity of applying proof by logical relation to dependent types stems
 from the fact that the logical relation is much harder to define. In
@@
 
-We say that a set of terms $[[S]]$ is closed under expansion if given
-$[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
+\begin{definition}[Sets closed under expansion]
+  We say that a set of terms $[[S]]$ is closed under expansion if
+  given $[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
+\end{definition}
 The final property we want to show is that the output set $[[S]]$ from
@@
   depend on anything}
-\begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpExt\_back\_clos}}]
+\begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
   \label{lemma:logrelbackclos}
@@
 Define $[[rho |= G]]$ when
-\[ \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
-                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \]
+\[ \forall n\ i\ S, \text{ if }[[n < |G|]]\text{ and
+                     } [[InterpR i (G n < up Suc n > ) { rho } S ]] \text{, then } [[rho n in S]] \]
 \end{definition}
@@
 substitution $[[rho]]$ with respect to a context $[[G]]$. Intuitively, for any
-variable $[[i]]$ in the context (thus the restriction $[[i < | G |]]$),
-$[[rho i]]$ is a term that inhabits all possible interpretations of the type
-$[[G i]]$ from the context.
+variable $[[n]]$ in the context (thus the restriction $[[n < | G |]]$),
+$[[rho n]]$ is a term that inhabits all possible interpretations of the type
+$[[G n]]$ from the context.
 
 Since we are in a dependently typed system, there
-may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
+may be variables in $[[G n]]$, too. The type $[[(G n < up Suc n > ) {
   rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
-operation $[[up Suc i]]$ can be viewed as an artifact of the de Bruijn
-representation to lift the type $[[G i]]$ to the right scope before
+operation $[[up Suc n]]$ can be viewed as an artifact of the de Bruijn
+representation to lift the type $[[G n]]$ to the right scope before
 substitution can be carried out.
@@
 Finally, despite the for all quantification over the interpreted set
-$[[S]]$ and the universe level $[[j]]$ in the definition of $[[rho |=
-G]]$, we only need show that there exists some $[[j]]$ and $[[S]]$
-such that $[[InterpR j (G i < up Suc i > ) { rho } S]]$ and $[[rho i
-in S]]$. This idea is encapsulated in the second of the two structural
+$[[S]]$ and the universe level $[[i]]$ in the definition of $[[rho |=
+G]]$, we only need show that there exists some $[[i]]$ and $[[S]]$
+such that $[[InterpR i (G n < up Suc n > ) { rho } S]]$ and $[[rho n
+in S]]$ since the logical relation is a partial function (Lemma~\ref{lemma:logreldeter}). This idea is encapsulated in the second of the two structural
 law about $[[rho |= G]]$.
@@
 \[
-   \forall [[i < |G|]],\text{ there exists some } [[j]] \text{ such that } [[drop Suc i G |= G i : Set j]] \]
+   \forall [[n < |G|]],\text{ there exists some } [[i]] \text{ such that } [[drop Suc n G |= G n : Set i]] \]
 \end{definition}
 
-In this definition, we use the notation $[[drop i G]]$ to denote
-the typing context obtained by dropping the last $[[i]]$ elements of
-$[[G]]$. When $[[G]]$ has fewer than $[[i]]$ elements, $[[drop i G]]$ returns
+In this definition, we use the notation $[[drop n G]]$ to denote
+the typing context obtained by dropping the last $[[n]]$ elements of
+$[[G]]$. When $[[G]]$ has fewer than $[[i]]$ elements, $[[drop n G]]$ returns
 the empty list. Intuitively, $[[|= G]]$ says that all
@@
 recover the structural rules as lemmas.
+\begin{lemma}[Semantic context well-formedness empty]
+  \label{lemma:semwffempty}
+  $[[|= G]]$ holds when $[[G]]$ is empty.
+\end{lemma}
 \begin{lemma}[Semantic context well-formedness cons]
@@
   If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
-\end{lemma}
-\begin{lemma}[Semantic context well-formedness empty]
-  \label{lemma:semwffempty}
-  $[[|= G]]$ holds when $[[G]]$ is empty.
 \end{lemma}
@@
   \label{lemma:stvar}
-  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
+  If $[[|= G]]$ and $[[n < |G|]]$, then $[[G |= var n : G n < up Suc n  >]]$.
 \end{lemma}
@@
   Suppose $[[rho |= G]]$. By the definition of semantic typing, we
-need to show that there exists some $[[j]]$ and $[[S]]$ such that
+need to show that there exists some $[[i]]$ and $[[S]]$ such that
   \begin{itemize}
-  \item $[[InterpR j G i < up Suc i > { rho } S]]$
-  \item $[[rho i in S]]$
+  \item $[[InterpR i G n < up Suc n > { rho } S]]$
+  \item $[[rho n in S]]$
   \end{itemize}
   From the definition of $[[|= G]]$, we know that there exists some
-  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. This is not
+  $[[i]]$ such that $[[drop Suc n G |= G n : Set i]]$. This is not
   immediately useful since $[[rho |= G]]$, not the truncated
-  context. We claim, without giving the proof, that from $[[drop Suc i
-  G |= G i  : Set j]]$, we can derive $[[G |= G i {up Suc i} : Set
-  j]]$. A rigorous justification would require us to formally state
+  context. We claim, without giving the proof, that from $[[drop Suc n
+  G |= G n  : Set i]]$, we can derive $[[G |= G n {up Suc n} : Set
+  i]]$. A rigorous justification would require us to formally state
   and show renaming/weakening lemma for semantic typing. We instead
@@
 %
-  Knowing that $[[G |= G i {up Suc i} : Set j]]$ holds, we can then
+  Knowing that $[[G |= G n {up Suc n} : Set i]]$ holds, we can then
   apply Lemma~\ref{lemma:setinv} to show that
-  $[[InterpR j G i < up Suc i > { rho } S]]$. It now
-  suffices to show $[[rho i in S]]$, but that is immediate from the
+  $[[InterpR i G n < up Suc n > { rho } S]]$. It now
+  suffices to show $[[rho n in S]]$, but that is immediate from the
   definition of $[[rho |= G]]$.
@@
 \begin{proof}
-  Immediate by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
+  Immediate by Lemma~\ref{lemma:setinv} and \rref{IR-Set}.
 \end{proof}
@@
 exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
-in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
+in S0]]$. The conclusion immediately follows from the admissible \rref{I-PiAlt}.
 \end{proof}
@@
   $[[\A{rho} b{up rho} in S
-  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
+  ]]$. By Lemma~\ref{lemma:piinvalt}, the alternative inversion
+  principle for \rref{I-Pi}, there exists some $[[S0]]$ such
   that all following conditions hold:
@@
   A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
-  from Lemma~\ref{lemma:semwffcons}. The semantic typing rules
+  from Lemmas~\ref{lemma:semwffempty} and \ref{lemma:semwffcons}. The semantic typing rules
 (Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp})
@@
 rule, where parallel reduction can take $\eta$ steps but still preserves
-$\beta\eta$-normal form.
+$\beta$-normal form.
 
@@
 Therefore, we formally define when a set is a \emph{reducibility candidate}
-(shortened as $CR$) as follows.
+(shortened as $CR$) as follows. Our definition of $CR$ is inspired by
+\citet{girard1989proofs}, but not identical since we only care about
+weak normalization.
 \begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
@@
 \begin{figure}[h]
+  \begin{minipage}{0.9\textwidth}
   \begin{tabular}{ l |  c  | c | c }
@@
   \end{tabular}
+  \end{minipage}
   \caption{Statistics of the Coq Development}
@@
 In axiom-free Coq, predicates do not come with the extensionality
-property. Given two predicates $P$ and $Q$, we can prove that $\forall
-x, P(x) \iff Q(x)$, but we can not conclude that $P = Q$. While
+property. Given two predicates $P$ and $Q$, we can not conclude that
+$P = Q$ when given a proof of $\forall
+x, P(x) \iff Q(x)$. While
 predicate extensionality is not necessary in our development, we
@@
 and more resiliant to changes, but also gives useful documentation for readers
-who wish to understand how the underlying proof. Although automation performs
+who wish to understand the mechanized proof. Although automation performs
 extensive search, it does not use lemmas or invert derivations that are not
@@
   \begin{tabular}{ l |  l  | l | l | l }
-      & \header{Universes} & \header{Inductives} & \header{Convertibility}
+      & \header{Universes} & \header{Inductives} & \header{Conversion}
       & \header{Large Elim} \\
@@
   Universes: Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
-  Convertibility: Typed (T), Untyped (U), Extensional (E) \\
+  Conversion: Typed (T), Untyped (U), Extensional (E) \\
   \end{tabular}
@@
 programming language that uses modality to distinguish between logical
-proofs and programs. The programmatic fragment
-includes recursive data types and supports general recursion at the
-cost of introducing divergence. Its logical fragment supports a
-proof-irrelevant identity type, which can be used to reason about the
-behavior of its programmatic fragment.
-However, the $\lambda^\theta$
-language is limited in its expressiveness since it does not support
-type-level computation or polymorphism and therefore is not a fully dependently
-typed language according to \citet{abel2013normalization}.
+proofs and programs. % The programmatic fragment
+% includes recursive data types and supports general recursion at the
+% cost of introducing divergence.
+% However, the $\lambda^\theta$
+% language is limited in its expressiveness since it does not support
+% type-level computation or polymorphism and therefore is not a fully dependently
+% typed language according to \citet{abel2013normalization}.
 The consistency proof of $\lambda^\theta$'s logical fragment has been
@@
 
+\ifextended
 Finally, despite our earlier claim that our metatheoretic properties
@@
   lemma). }
+\fi
 

### 2024-02-26 paper/paper.tex
pass through introduction. We still need to refine our contributions summary

@@
 Our goal with this work is to demonstrate how the proof technique of
-\emph{logical relations} can be applied to reason about dependent type
-theories.
-Tutorial material on this topic~\citep{skorstengaard2019introduction,
-  harpertait, harperkripke, pierce2002types, pierce2004advanced,harper2016practical} is primarily focussed on
-systems with simple or polymorphic types. In that context, logical relations
+\emph{syntatic logical relations} can be applied to reason about dependent
+type theories.  Logical relations are a powerful proof technique for reasoning
+about lambda calculi and their operational semantics. They have been used to
+show diverse properties such as strong normalization, contextual equivalence,
+representation independence, noninterference, compiler correctness, and the
+decidability of conversion algorithms. \scw{add citations? emphasize that we are working 
+with syntactic logical relations.}
+
+Tutorial material on logical relations~\citep{skorstengaard2019introduction,
+  harpertait, harperkripke, pierce2002types,
+  pierce2004advanced,harper2016practical} is primarily focussed on systems
+with simple or polymorphic types. In that context, syntactic logical relations
 can be define as simple recursive functions over the structure of types, or
 (in the case of recursive types) defined over the evaluation steps of the
-computation.
-\scw{need one sentence to motivate why logical relations are important}
-
-Neither of these techniques make sense for predicative dependent type theory,
-so a young researcher might be excused for thinking that proofs that use
-logical relations are not applicable.
-But this is not the case. Recent authors have developed tour-de-force
+computation. However, neither of these techniques can be used to define a
+logical relation in the context of a predicative dependent type theory, so a
+young researcher might be excused for thinking that proofs that use logical
+relations are not applicable for such languages.
+
+But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs for dependent type theories}
+Recent authors have developed tour-de-force
 mechanizations for the metatheory of modern proof
@@
 systems and algorithms, these developments range in size from 20,000 to
-400,000 lines of code, these definitions are easy to miss and difficult to
-isolate from their surrounding contexts. As a result, they are not accessible
+400,000 lines of code. As a result, their use of logical relations arguments
+difficult to isolate from their surrounding contexts, and thus, not accessible
 to casual readers.
 
-Our proof, designed to capture the essence of this proof technique in this
-setting, requires less than 1000 lines of code. We have achieved this
-significant reduction through a number of means: the careful selection of the
-features that we include in the object type type system and the results that
-we prove about it, in addition to the judicious use of automation.  Our
+Thus, our paper provides a gentle and accessible introduction to a powerful
+technique for dependent type theories. To promote the use of machine-assisted
+reasoning, our development is accompanied by a short mechanized proof script,
+of less than 1000 lines of code, developed using the Coq/Rocq proof
+assistant~\cite{coq}.
+
+We have streamlined our proof through a number of means: the careful selection
+of the features that we include in the object type type system and the results
+that we prove about it, in addition to the judicious use of automation.  Our
 language is small, but includes enough to be illustrative. For example, we
@@
 mechanical proofs, while avoiding, as much as possible, the introduction
-of artifacts that are specific to certain theorem prover.
+of artifacts that are specific to theorem provers.
 
@@
 
-In Section~\ref{sec:spec}, we introduce \lang{}, the object language
-of interest, and establish some elementary properties about parallel
-reduction, which will be used as the equational theory of our language
-and play a central role in defining our logical relation. The design
-of \lang{} is reminiscent of Martin-Löf style predicative type
-systems. The untyped conversion rule of \lang{} is inspired by Pure
-Type Systems~\citet{barendregt1991introduction}, though we discuss how
-our proof can be easily adjusted for systems with typed judgmental
-equality.  We design \lang{} to incorporate some of the most common
-features of dependent type theory, including an intensional identity
-type, a boolean type, large elimination, dependent elimination, and a
-full universe hierarchy. We believe a good coverage of features will
-make our proof applicable to a broad range of type systems.
-
-In Section~\ref{sec:logreldep}, we formulate the consistency property
-and motivate why proof by logical relation is needed to derive it.
-We define our logical relation as an inductively defined relation. We first
-show that the logical relation respects evaluation (referred to as the
-irrelevance property in \citet{martin-lof-a-la-coq}), then proceed to
-prove that the logical relation is a partial function; that is, each
-type can have at most one unique interpretation.
-
-In Section~\ref{sec:logrelproof}, we give the definition of semantic
-typing in terms of our logical relation so we can state and prove the
-fundamental theorem, which states that syntactic typing implies
-semantic typing, from which consistency follows as a corollary.
-
-In Section~\ref{sec:extension}, we extend our logical relation to
-prove the stronger property that every well-typed term has a
-$\beta\eta$-normal form. After laying the foundation for handling
-type-level computation in Section~\ref{sec:logreldep}, the extension
-to include open terms in the logical relation requries mechanical
-changes that directly mirrors the same extension for a simply typed
-language. We use this extension to demonstrate the idea that once we
-have established the base technique for proof by logical relation for
-dependent types, we can factor out the complexities of an extension
-the are not specifically related to dependent types; such an
-extension, if desired, may be studied in the context of a simply typed
-language and later ported into a dependently typed setting. We believe
-it is possible to extend our logical relation to a Kripke-style and
-relational model in a similar fashion.
+The key contributions of each section of our paper are as follows.
+\scw{I'm still not happy with this overview / contributions list. }
+
+In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type theory of
+interest, reminiscent of Martin-Löf style predicative type systems. A key
+design choice that impacts our proofs is the use of an untyped conversion
+rule, inspired by Pure Type Systems~\citet{barendregt1991introduction}, and
+specified through parallel reduction to normal form~\citet{barendregt}.
+Because we believe that a good coverage of features makes our proof applicable
+to a broad range of type systems, \lang{} incorporates some of the most common
+features of dependent type theory, including an intensional identity type, a
+boolean type, large elimination, dependent elimination, and a full universe
+hierarchy. While our use of untyped conversion is nonstandard for MLTT, we
+discuss how our proof can be adjusted for systems with a typed, judgmental
+equality.
+
+In Section~\ref{sec:logreldep}, we formulate the consistency property and
+motivate where a proof by logical relations is required. The most important
+contribution of this section is the definition of the logical relation as an
+inductively defined relation and demonstrate how to reason about this
+definition. We also show that the logical relation respects evaluation
+(referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
+prove that the logical relation is a partial function; that is, each type can
+have at most one unique interpretation.
+
+In Section~\ref{sec:logrelproof}, shows that our definitions can be the
+subject of well-established reasoning techniques for logical relations. In
+particular, we define a notion of semantic typing in terms of our logical
+relation and prove the fundamental theorem, which states that syntactic typing
+implies semantic typing. Consistency follows from this result as a corollary.
+
+In Section~\ref{sec:extension}, extends the definition of the logical relation
+to include terms with free variables, and proves that every well-typed term
+has a $\beta$-normal form. After laying the foundation for handling type-level
+computation in Section~\ref{sec:logreldep}, this extension requries mechanical
+changes that directly mirror an analogous extension for a simply typed
+language. We use this section to demonstrate the idea that once we have
+established the base techniques, we can port ideas that have been studied in
+the context of a simpler system to the dependently-typed setting.  We believe
+it is possible to extend our logical relation to a Kripke-style and relational
+model in a similar fashion.
+
+In Section~\ref{sec:eta} we show how to incorporate the function $\eta$-law
+into the conversion rule for \lang{}.
 
@@
 relatively feature-complete language will encourage future researchers
-to leverage the tool of logical relation more often in mechanized
+to leverage the tool of logical relations in mechanized
 proofs for dependent types.
@@
 % type theory.
-
-\scw{The end of this section needs to include an explicit summar of the
-contributions of the paper. }
 
@@
 dependent type theory whose logical consistency will be proven in
-Section~\ref{sec:logrelproof}. For concision, we refer to this system
+Section~\ref{sec:logrelproof}. For conciseness, we refer to this system
 as \lang.
@@
 The syntax of \lang can be found in Figure~\ref{fig:syntax}.
-% \scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
-% \yl{need at least 4 vars for J. removed the unnecessary t and C}
-We use
-the unscoped de Bruijn representation for both our Coq development and
+We use the unscoped de Bruijn representation for both our Coq development and
 the informal presentation in our paper since it is more amenable to

### 2024-02-26 paper/paper.tex
pass through section 2. added more explanation to the type system.

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Short and Mechanized Logical Relation for Dependent Types}
+\title{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}
 
@@
 \begin{array}{lcll}
-\mathit{Natural\ numbers}\\
-[[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
-
-\mathit{Contexts}\\
-[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
+% \mathit{Natural\ numbers}\\
+% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
 \mathit{Terms}\\
@@
                   & \mbox{universes, variables, empty type} \\
-            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
+            & |   & [[Pi A B]]\ |\ [[\ a]]\ |\ [[a b]]
                   & \mbox{function types, abstractions, applications} \\
@@
             & |   & [[if a b0 b1]]
-                  & \mbox{if} \\ \\
+                  & \mbox{conditional expression} \\ \\
 % \mathit{Renaming}\\
 % [[xi]] & \in & [[SNat -> SNat]] & \\ \\
-\mathit{Substitution}\\
-[[rho]] & \in & [[SNat -> STm]] &
+\mathit{Substitutions}\\
+[[rho]] & \in & [[SNat -> STm]] & \\ \\ 
+\mathit{Typing Contexts}\\
+ [[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
 \end{array}
@@
 
-In this section, we present the dynamics and statics of the
-dependent type theory whose logical consistency will be proven in
-Section~\ref{sec:logrelproof}. For conciseness, we refer to this system
-as \lang.
-
-The syntax of \lang can be found in Figure~\ref{fig:syntax}.
-We use the unscoped de Bruijn representation for both our Coq development and
-the informal presentation in our paper since it is more amenable to
-automation (Section~\ref{sec:automation}) and leaves no ambiguity
-about variable freeness conditions.
-
-%  \scw{Give an overview of the figure}
-% \scw{Define a convention that separates your use of natural numbers. Sometimes they
-% are universe levels and sometimes they are de Bruijn indices. Better to have separate
-% metavariables for each. (And it is possible to generalize universe
-% levels to structures other than natural numbers. Ask Jonathan.} \yl{resolved?}
-As a dependent type theory, terms and types are collapsed into the same
-syntactic category. The type $[[Set i]]$ represent universe
-types and $[[var n]]$ represent de Bruijn variables. While
-$[[i]],[[j]],$ and $[[n]]$ are all metavariables representing natural
-numbers, we always use $[[n]]$ to represent term variables that
-participate in substitution and $[[i]],[[j]]$ to represent universe levels.
-Dependent functions take the form $[[Pi A B]]$ and
-we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
-dependent on the input variable.
-Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
-proofs can be eliminated by the J-eliminator $[[J c a b p]]$, where
-$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[c]]$
-is the term whose type is to be casted.
-\scw{Example of what you can use identity types for? Or an example of
-a program that uses J?}
+In this section, we present the dependent type theory whose logical
+consistency will be proven in Section~\ref{sec:logrelproof}. For conciseness,
+we refer to this system as \lang.
+
+The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
+type theory, terms and types are collapsed into the same syntactic
+category. The type $[[Set i]]$ represent universe types and $[[var n]]$
+represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
+$[[n]]$ are all metavariables representing natural numbers, we always use
+$[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
+variables.  We use de Bruijn indices for both our Coq development and the
+informal presentation here because it is more amenable to automation
+(Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
+conditions. As a result, variables do not appear in binding forms, such as in
+abstractions $[[ \ a ]]$ and dependent function types $[[Pi A B]]$.  This latter
+form binds a new variable of type $[[A]]$ that can appear in the codomain of
+the type $[[B]]$.  We use the notation $[[A -> B]]$ when the output type
+$[[B]]$ is not dependent on the input variable.
+We include in \lang{} the intensional identity type
+$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
+$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
+$[[b]]$, and $[[c]]$ is the term whose type is to be casted.
+Finally, \lang{} includes booleans, with standard syntax.
+\scw{Example of
+ what you can use identity types for? Or an example of a program that uses
+J?}
 
@@
         % [[id i]] & := & [[i]] \\ \\
-        \mathit{IdentityTm} \\
+        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} \mathsf{id}_{\mathit{tm}}} \\
         [[idtm n]] & := & [[var n]] \\ \\
+        \multicolumn{3}{l}{\mbox{\emph{Weakening substitution: }} \uparrow^{n}} \\
+        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]} \\ \\
+        %
         % \mathit{ConsRen ([[xi .: n]])} \\
@@
         % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
-        \mathit{ConsSubst ([[rho .: a]])} \\
+        %
+        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: a]]) } \\
         ([[rho .: a]])([[0]]) & := & [[a]] \\
         ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
-        \mathit{Curried\ Add} \\
-        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
       \end{array}
@@
       \begin{array}{lll}
-        \mathit{Lifting} \\
+        \multicolumn{3}{l}{\mbox{\emph{Lifting operation: }} \Uparrow\rho} \\
         [[( up rho ) 0]] & := & [[var 0]] \\
         [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
-        \mathit{Substitution ([[ a { rho }  ]])} \\
+        \multicolumn{3}{l}{\mbox{\emph{Substitution operation:}} ([[ a { rho } ]])} \\
         [[var n { rho }  ]] & := &  [[rho n]] \\
@@
 eliminations, the ability to compute a type using a term as input. For
-example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
+example, the function $[[\ if var 0 Bool Bool -> Bool]]$ returns
 either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
@@
 
-We adapt from \citet{autosubst2} the notations for simultaneous
-renaming, substitution, and other auxiliary definitions used in
-our paper, summarized in Figure~\ref{fig:auxdef}. We need those
-definitions to specify the reduction relation and the typing relation
-for \lang{}. The metavariable $[[rho]]$ represents substitutions,
-mappings from de Bruijn indices to terms.
-One example of a substitution is
-the function $\ottkw{id}_{tm}$, which takes a de Bruijn index directly
-injects it as a term variable. The $\uparrow^{[[n]]}$ operator is
-more general and adds $[[n]]$ to the index before injecting it as a
-term. The $[[(rho .: a)]]$ operation allows us to extend a substitution
-$[[rho]]$ with an extra element $[[a]]$; the extended substitution maps the variable
-$[[var 0]]$ to $[[a]]$ and variables $[[var Suc i]]$ to $[[rho i]]$.
+We adapt from \citet{autosubst2} the notations for simultaneous renaming,
+substitution, and other auxiliary definitions used in our paper, summarized in
+Figure~\ref{fig:auxdef}. We need these definitions to specify the reduction
+and typing relations for \lang{}. The metavariable $[[rho]]$ represents
+substitutions, which are mappings from de Bruijn indices to terms.  One
+example of a substitution is the function $\ottkw{id}_{tm}$, the identity
+substitution that maps each index to the corresponding term variable.  The
+$\uparrow^{[[n]]}$ substitution is more general and adds $[[n]]$ to the index
+before injecting it as a term. The extension operation, $[[(rho .: a)]]$,
+allows us to extend a substitution $[[rho]]$ with an extra element $[[a]]$;
+the result maps the variable $[[var 0]]$ to $[[a]]$ and variables
+$[[var Suc i]]$ to $[[rho i]]$.
 
@@
 $[[var 0]]$ unchanged during substitution and replaces each $[[var
-Suc i]]$ with $[[rho i {up 1}]]$. The extra shifting with $[[up 1]]$
+Suc i]]$ with $[[rho i {up 1}]]$. The shifting with $[[up 1]]$
 is required so the variables in terms from the codomain of $[[rho]]$ skip over
 the newly introduced binder to refer to the correct binding
-location. The substitution operator is referred to as simultaneous
-substitution as it substitutes all variables at once. It is possible
-to recover single substitution by composing the extension operator and
-the identity substitution: $[[a { b }]] := [[a { idtm .: b }]]$. Since
-the definition of semantic typing in Section~\ref{sec:logrelproof}
-relies on simultaneous substitution, we find it more convenient to
-formulate simultaneous substitution directly rather than recovering it
-from single substitution.
+location.
+
+The substitution operator is referred to as simultaneous substitution as it
+substitutes for all variables at once. It is possible to recover single
+substitution by composing the extension operator and the identity
+substitution: $[[a { b }]] := [[a { idtm .: b }]]$. In our proofs, we find it
+more convenient to formulate simultaneous substitution directly rather than
+recovering it from single substitution.  In particular, this shows up in the
+definition of semantic typing in Section~\ref{sec:logrelproof}, which relies
+on simultaneous substitution.
+
+\subsection{Definitional equality via parallel reduction}
 
@@
 %   an explanation of definitional equality and its properties?}
-Before we specify the typing rules, we need to first specify its
-untyped equational theory. We choose to represent the untyped
-equational theory in an algorithmic style using the notion of parallel
-reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted).
- Parallel reduction takes
-the form $[[a => b]]$ and we denote its transitive as $[[a =>+
-b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
-denoted by the notation $[[a0 <=> a1]]$, if there exists some term
-$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
-notation of convertibility suggests that it is an equivalence
-relation. We sketch out the sequence of lemmas required to derive this
-property below and omit the details. Our technique for proving the
-properties about parallel reduction is based on
-\citet{takahashi-parallel-reduction}. A modern exposition of the
+Before we specify the typing rules, we first specify the equational theory
+used in its conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). This
+relation is often called ``definitional equality'' in dependent type theories
+because it defines the equivalence that the syntactic type system works
+``up-to''.
+
+For \lang{}, we use an untyped equational theory, defined in an algorithmic
+style using the notion of parallel reduction, written $[[a => b]]$, and shown
+in Figure~\ref{fig:par}. (For brevity, reflexivity and congruence rules are
+omitted from this figure). The notation $[[a =>+ b]]$ indicates the transitive
+closure of parallel reduction.
+
+\begin{definition}[Convertibility]
+  Two terms $[[a0]]$ and $[[a1]]$ are \emph{convertible}, written
+  $[[a0 <=> a1]]$, if there exists some term $[[b]]$ such that $[[a0 =>+ b]]$
+  and $[[a1 =>+ b]]$.
+\end{definition}
+
+Convertibility is an equivalence relation. We sketch the sequence of lemmas
+required to derive this property below and omit the details. The key step in
+proving transitivity, is showing that parallel reduction satisfies the diamond
+property.  Our technique for proving the properties about parallel reduction
+is based on \citet{takahashi-parallel-reduction}. A modern exposition of the
 same technique can be found in \citet{plfa22.08}.
+
 \begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
@@
 \end{lemma}
-From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
-\ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
-equivalence relation. It is then trivial to show that all untyped
-$\beta$-equivalence rules are admissible, though we omit this proof
-from our mechanization and will from now on focus on parallel
-reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
-makes it easier to work with. A detailed argument of the equivalence
-between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
+% From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
+% \ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
+% equivalence relation.
+
+Our definition of convertibility for \lang{} is unusual in that it is directly
+defined via parallel reduction, instead of using the related notion of
+$\beta$-equivalence~\cite{barendregt:pts,cic}. This choice does not change the
+language definition; a detailed argument of the equivalence between
+$[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
 \citet{barendregt:lambda-calculi-with-types} and
-\citet{takahashi-parallel-reduction}.
-
-Untyped equality allows us to show the
-equivalence between our more structured notion of convertibility
- ($[[a <=> b]]$) and the declarative, untyped $\beta$-equivalence
-independently from the type system and the logical relation through
-well-established syntactic approaches. The directedness of the
-convertibility relation will be leveraged in
-Section~\ref{sec:logreldep} to justify the conversion rule of our
-typing judgment, which we will introduce shortly.
+\citet{takahashi-parallel-reduction}. However, this choice simplifies later
+proofs, as we discuss in Section~\ref{XXX}. \scw{Forward reference to further
+  discussion, comparing with whnf reduction?}
+
+Our definition of equality is untyped: the judgement does not require the two
+terms to type check and have the same type. The use of an untyped relation for
+type conversion is reminiscent of Barendregt's Pure Type
+Systems~\citet{barendregt1991introduction} and makes our formulation different
+from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental
+equality takes the form $\Gamma \vdash a \equiv b : A$, from which one can
+usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
+reduction. By working with an untyped judgement, we can establish its
+properties independently from the type system and the logical relation, using
+well-established syntactic approaches.
+
+\citet{siles2012pure} shows the equivalence of Barendregt's Pure Type
+System~\citep{barendregt1991introduction}, which employs untyped equality, and
+its variant that uses typed judgmental equality. This assures us that we do
+not lose generality working with a system with untyped
+conversion. Furthermore, discuss how this definition can be extended with
+$\eta$-equivalence of functions in Section~\ref{XXX}. Finally, we compare this
+definition with type-directed approaches to equality in
+Section~\ref{sec:discuss}.
+
+\subsection{Syntactic Typing}
 
@@
 \drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
-  Refl, Eq, J, If, Bool, True, False, Void}
+  Eq, Refl, J, If, Bool, True, False, Void}
 \end{minipage}
@@
 
-Figure~\ref{fig:typing} gives the full typing rules for
-\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
-admissible syntactically, though some of them are required to
-strengthen the inductive hypothesis of the fundamental theorem.
-
-In \rref{T-Var}, $[[G n]]$ is the partial function defined
-by the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
-n ]] := [[G n]]$.
-The precondition $[[n < | G | ]]$, where $[[ | G |
-]]$ represents the length of the context $[[G]]$, ensures that $[[G
-n]]$ has a defined output. The type $[[G n]]$ needs to be shifted
-by $[[Suc n]]$ since the types stored in the context $[[G]]$ are
-scoped differently. In the typing judgment $[[G |- a : A]]$, we want
-both $[[a]]$ and $[[A]]$ to have the same scope. However, the closer a
-type is to the head of $[[G]]$, the smaller de Bruijn it has since
-there is less variables in the context it can refer to (see \rref{Ctx-Cons}). Shifting $[[G
-n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
-the full context $[[G]]$ rather than a truncated version of
-$[[G]]$. In \rref{T-J}, the elimination form for the identity type,
-the terms $[[a]]$ and $[[A]]$ need to be shifted in the context for a similar
-reason; an extra type $[[A]]$ appears after the $[[G]]$ and therefore
-the terms $[[a]]$ and $[[A]]$ in the identity type both need to be shifted
-to refer to the same binders.
+Figure~\ref{fig:typing} gives the full typing rules for \lang{}. The premises
+wrapped in \colorbox{lightgray}{gray} boxes can be shown to be admissible
+syntactically, though some of them are required to strengthen the inductive
+hypothesis of the fundamental theorem.
+
+In \rref{T-Var}, $[[G n]]$ is the partial function defined by the equations
+$[[ (G ++ A) 0 ]] := [[A]]$ and $[[ (G ++ A) Suc n ]] := [[G n]]$.  The
+precondition $[[n < | G | ]]$, where $[[ | G | ]]$ represents the length of
+the context $[[G]]$, ensures that $[[G n]]$ has a defined output. The type
+$[[G n]]$ needs to be shifted by $[[Suc n]]$ since the types stored in the
+context $[[G]]$ are scoped differently. In the typing judgment
+$[[G |- a : A]]$, we want both $[[a]]$ and $[[A]]$ to have the same
+scope. However, the closer a type is to the head of $[[G]]$, the smaller de
+Bruijn it has since there is less variables in the context it can refer to
+(see \rref{Ctx-Cons}). Shifting $[[G n]]$ by $[[Suc n]]$ weakens the type so
+it is scoped consistently with the full context $[[G]]$ rather than a
+truncated version of $[[G]]$.
+
+Although the use of de Bruijn indices complicates the variable rule, the rules
+that involve binding, such as \rref{T-Pi,T-Abs} are straightforward.
 
 \Rref{T-Conv} uses the convertibility relation from earlier
-as our equality judgment for type conversion. The use of an untyped
-relation for type conversion is reminiscent of Barendregt's Pure Type
-Systems~\citet{barendregt1991introduction} and makes our formulation different
-from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
-takes the form $\Gamma \vdash a \equiv b : A$, from which one can
-usually derive  $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving
-subject reduction.
-
-\citet{siles2012pure} shows the equivalence of Barendregt's
-Pure Type System~\citep{barendregt1991introduction}, which employs
-untyped equality, and its variant that uses typed judgmental
-equality. This assures us that we do not lose much generality working
-with a system with untyped conversion.
-We include a detailed discussion in Section~\ref{sec:discuss} on how
-type-directed and untyped equality affect the applicability of our
-technique.
+as our equality judgment for type conversion.
+
+The elimination form for booleans, \rref{T-If} demonstrates dependent pattern
+matching.  The result type of this expression $[[A {a}]]$, is composed of some
+\emph{motive} $[[A]]$, a type where its single free variable has been replaced
+with the condition of the if expression. When typing the true branch, this
+substitution replaces the variable by $[[true]]$, and similar for the false
+branch. As a result, the type system communicates the information gained from
+the run-time test to each of the branches of the expression.
+
+A similar sort of dependent pattern matching occurs when eliminating identity
+types. Such types are checked for well-formedness with \rref{T-Eq} and
+introduced by \rref{T-Refl}. In \rref{T-J}, the elimination form, the subterm
+$p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
+is the body of the elimination form. In this rule, $[[B]]$ is called a
+\emph{motive} and has two free variables.  The substitution for these
+variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
+witnessing the information gained through dependent pattern matching.  In the
+typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
+for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
+therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
+shifted to refer to the same binders.
 
@@
     Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
-    % \hline
+    \hline
     CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
+    CoreSpec & \\
+    PCUIC & \\
     %   & \lang{} & \citet{anand2014towards} &
@@
 
-  \begin{tabular}{l}
+  \begin{tabular}{ll}
   \\
-    $\lambda^\theta$ (logical fragment only) \citet{casinghino:combining-proofs-programs} \\
-    NBE-in-Coq \citet{nbeincoq} \\
-    $\lambda^{\Pi U\mathbb{N}}$ \citet{decagda} \\
-    MLTT-a-la-Coq \citet{martin-lof-a-la-coq} \\
-    Core Nuprl \citet{anand2014towards} \\
-    CIC \citet{barras2012semantical} \\
+    $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
+    NBE-in-Coq& \citet{nbeincoq} \\
+    $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
+    MLTT-a-la-Coq &\citet{martin-lof-a-la-coq} \\
+    Core Nuprl &\citet{anand2014towards} \\
+    CIC &\citet{barras2012semantical} \\
+    CoreSpec & \\
+    PCUIC & \\
+
   \\
-  Universes: Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
-  Conversion: Typed (T), Untyped (U), Extensional (E) \\
+  Universes: &Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
+  Conversion:& Typed (T), Untyped (U), Extensional (E) \\
   \end{tabular}

### 2024-02-26 paper/paper.tex
turn off comments. fix proof.
Extend the eta section

@@
 
-\commentstrue         %% toggle comments here
+\commentsfalse         %% toggle comments here
 \extendedfalse
@@
 \end{corollary}
-Due to some technical
-details, our proof about the existence of $\beta\eta$ form does not
+
+In \citet{nbeincoq,decagda,martin-lof-a-la-coq}, a relational model is
+needed to justify the $\eta$-law for functions. Instead, we prove
+confluence for our parallel reduction extended with $\eta$-law for functions
+and continue using a logical predicate (i.e. a 1-ary logical relation) to justify
+$\eta$-law as part of our equational theory.
+A well-known issue with our approach is the failure of syntactic
+confluence when the lambda term contains type annotations. A simple
+counterexample is $[[\- B ((\- A a) {up 1} var 0)]]$; depending on
+whether \rref{P-AbsEta} is performed on the whole term or
+\rref{P-AppAbs} is used on the inner $\beta$ redex, we end up with the
+terms $[[\- B a]]$ or $[[\- A a]]$, where $[[A]]$ and $[[B]]$ are not
+necessarily syntactically equal terms. \citet{choudhury:ddc} resolve
+this problem by stating their confluence result in terms of an
+equivalence relation that quotients out parts of the terms that are
+computationally irrelevant; the annotations of lambda terms are
+ignored since the behavior of a lambda term is not affected by its
+type annotation. We believe the same approach is applicable to our
+proof if we add type annotations to our lambda terms.
+
+The bigger issue is extensions such as $\eta$-laws for unit and
+products. Surjective pairing, for example, is an equational law that
+is known to cause trouble since it is not confluent for untyped lambda
+terms~\citet{KLOP198997}. The relational, type-annotated, and Kripke-style models from
+\citet{nbeincoq,decagda,martin-lof-a-la-coq} can be more easily
+extended to support these rules.
+We note, however, that the issue with $\eta$ rules is not exclusive to dependently
+typed languages and has been studied in more limited languages that
+are either simply
+typed~\citep{pierce2004advanced,pfenning1997computation} or
+dependently typed but without large
+elimination~\citep{harper2005equivalence,
+abel2005untypedconvsurjective}. Common workarounds include
+type-directed conversion and shifting the focus to obtaining
+$\eta$-long forms~\citet{Abel12}.
+We believe our simple proof will help readers
+see how techniques from these simpler settings generalize to fully
+dependently typed languages and better understand the more complex
+mechanized developments.
+
+Finally, due to some technical
+details, our mechanized proof about the existence of $\beta\eta$ form does not
 immediately induce a decision procedure for type conversion in
@@
 Section~\ref{sec:conversionalgo}.
-
 \section{Mechanization}
@@
 dependently typed language with one predicative universe, similar to
-\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
-\citet{anand2014towards} and \citet{nbeincoq} leverage the
+\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
+type system has no elimination form for natural numbers, the
+only base type from the object language, large elimination is not
+supported despite the one predicative universe.
+Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
 impredicative \texttt{Prop} sort of Coq to define the interpretation
 of dependent function types and thus are closely related to our
-mechanization. However, instead of explaining the inductive
-logical relation as a convoluted workaround to the lack of
-induction-recursion in Coq, we give a self-contained explanation of
-our logical relation.
+mechanization. \citet{anand2014towards} further shows it is possible
+to encode a finite universe hierarchy without the use of
+either impredicativity or induction-recursion.
+% \yl{commented out because I think this paragraph is too subjective}
+% However, instead of explaining the inductive
+% logical relation as a convoluted workaround to the lack of
+% induction-recursion in Coq, we give a self-contained explanation of
+% our logical relation.
 
@@
 reduction, we believe our proof technique can significantly shorten
-the existing logical relation proofs for systems with typed judgmental
-equality and more complex $\eta$ laws. We leave that as part of our
-future work.
+the existing logical relation proofs for systems with typed
+judgmental. We leave that as part of our future work.
 % Therefore, we believe that even in a

### 2024-02-26 paper/paper.tex
Fix typos and change language name to be more descriptive

@@
 \newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
-\newcommand{\lang}{$\lambda^H$\xspace}
+\newcommand{\lang}{$\lambda^{U\mathbb{B}\ottkw{Id}}$\xspace}
 \inputott{rules}
@@
 context of simply typed or polymorphic lambda calculus, a streamlined
-presentation of proof by logical relation for a dependently language
+presentation of proof by logical relation for a dependently typed language
 is lacking. In this paper, we present a short
@@
 Our goal with this work is to demonstrate how the proof technique of
-\emph{syntatic logical relations} can be applied to reason about dependent
+\emph{syntactic logical relations} can be applied to reason about dependent
 type theories.  Logical relations are a powerful proof technique for reasoning
@@
   harpertait, harperkripke, pierce2002types,
-  pierce2004advanced,harper2016practical} is primarily focussed on systems
+  pierce2004advanced,harper2016practical} is primarily focused on systems
 with simple or polymorphic types. In that context, syntactic logical relations
@@
 We have streamlined our proof through a number of means: the careful selection
-of the features that we include in the object type type system and the results
+of the features that we include in the object type system and the results
 that we prove about it, in addition to the judicious use of automation.  Our
 language is small, but includes enough to be illustrative. For example, we
-eschew inductive datatypes or W-types, but we do include propositionality
+eschew inductive datatypes or W-types, but we do include propositional
 equality and booleans in order to capture the challenges presented by indexed
@@
 show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
-closed terms. We include a full predicative universe hiercharchy and
+closed terms. We include a full predicative universe hierarchy and
 type-level computation, to demonstrate the logical strength of the
@@
 That said, our explanations do not stray too far away from our proof scripts.
-We label each lemma directly to their counterpart in the proof script,
+We label each lemma directly to its counterpart in the proof script,
 anticipating that readers may wish to reference and validate our formalism.
-Our typeset representation are purposefully made to directly follow our
+Our typeset representations are purposefully made to directly follow our
 mechanical proofs, while avoiding, as much as possible, the introduction
@@
 assistants for their day-to-day use. We also find that this precision is
-important for conveying the proof techique itself. Unlike properties that are
+important for conveying the proof technique itself. Unlike properties that are
 derivable through syntactic means, proofs by logical relation make demands on
@@
 has a $\beta$-normal form. After laying the foundation for handling type-level
-computation in Section~\ref{sec:logreldep}, this extension requries mechanical
+computation in Section~\ref{sec:logreldep}, this extension requires mechanical
 changes that directly mirror an analogous extension for a simply typed
@@
 
-In Section~\ref{sec:eta} we show how to incorporate the function $\eta$-law
+In Section~\ref{sec:betaeta} we show how to incorporate the function $\eta$-law
 into the conversion rule for \lang{}.
@@
 proving consistency and normalization.
+
+In Section~\ref{sec:discuss}, we explain the various factors that contribute
+to our short proof.
 
@@
 true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
-  true \/ a =>+ false }]]$ since it evalutes to $[[true]]$.
+  true \/ a =>+ false }]]$ since it evaluates to $[[true]]$.
 The independence of syntactic typing in our logical relation allows
@@
 \Rref{I-Eq}
-says that an equality type $[[a ~ b : A]]$ corresonds to the
+says that an equality type $[[a ~ b : A]]$ corresponds to the
 set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
@@
 interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
-type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
+type $[[B {a}]]$, obtained by substituting $[[a]]$ into the output type
 $[[B]]$, has some semantic interpretation. In its conclusion, we say
@@
   that closes over a context should always induce a truncated
-  substituion for each truncated context. As a result, if
+  substitution for each truncated context. As a result, if
   some term $[[b]]$ is semantically well-typed under some truncated
@@
 Furthermore, if we know that an application of a term to a variable
-has a normal form, then we know tha the term must have a normal form.
+has a normal form, then we know the term must have a normal form.
 \begin{lemma}[Wn extensionality\footnote{\dotv{normalform.v}{ext\_wn}}]
@@
 confluence for our parallel reduction extended with $\eta$-law for functions
-and continue using a logical predicate (i.e. a 1-ary logical relation) to justify
+and continue using a logical predicate (i.e. a unary logical relation) to justify
 $\eta$-law as part of our equational theory.
@@
 x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
-occassionally in our presented proofs.
+occasionally in our presented proofs.
 In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm ->
@@
 We use the Autosubst 2 framework to specify our syntax in HOAS and
-produce Coq syntax files in de Bruijn representation. Additionallly,
+produce Coq syntax files in de Bruijn representation. Additionally,
 Autosubst 2 provides a powerful tactic \texttt{asimpl} that
@@
 The automation provided by CoqHammer not only gives us a proof that is shorter
-and more resiliant to changes, but also gives useful documentation for readers
+and more resilient to changes, but also gives useful documentation for readers
 who wish to understand the mechanized proof. Although automation performs
@@
 than weak head reduction. Furthermore, \citet{Abel12} and
-\citet{abel2008betaeta} use the semantic universe hierachy as a
+\citet{abel2008betaeta} use the semantic universe hierarchy as a
 measure to define Kripke-style logical relations, from which they
@@
 type-directed equivalence relation satisfying certain
-properties to facillitate the reuse of their definition. The
+properties to facilitate the reuse of their definition. The
 logical relation is defined using the induction-recursion scheme,
@@
 from~\citet{decagda} to the decidability type checking of a
-bidrectional type system.
+bidirectional type system.
 
@@
   cumulativity (they only have one predicative universe)\\ \\
-  The 400,000 NuPRL in Coq probably falls into a different cateogry
+  The 400,000 NuPRL in Coq probably falls into a different category
   because they are trying to mechanize a full practical language}

### 2024-02-26 paper/paper.tex
Rewrite the "First,..." from Sec.9 discussion

@@
 
-\commentsfalse         %% toggle comments here
+\commentstrue         %% toggle comments here
 \extendedfalse
@@
 
+\subsection{Definition of the Logical Relation}
 \label{sec:logreldep}
@@
 
-In the rest of the section, we prove four important facts about
+\subsection{Properties about the Logical Relation}
+In this section, we prove four important facts about
 our logical relation: irrelevance (Lemma~\ref{lemma:logrelcoherence}),
@@
 
-First, the metatheoretic properties that we prove are indeed simpler.  Unlike
-developments that mechanize the correctness of a type-directed conversion
-algorithm, we only show the existence of normal forms for open and closed
-terms and state our properties in terms of an untyped reduction relation. This
-avoids a lot of the scaffolding related to the specification of the algorithm
-and the proof obligations that the algorithm is sound and complete with
-respect to the declarative specification of the type system.
-As a result, our logical relation, unlike the ones from \citet{decagda,
-  nbeincoq, martin-lof-a-la-coq, anand2014towards}, maps from types to
-predicates rather than relations. \scw{How does this follow? Why do we need unary relations where they need binary relations?} The need for a relational model is
-directly related to the metatheoretic property one wants to
-prove. \citet{anand2014towards} requires a relational model to capture
-the extensionality of their type system, whereas \citet{decagda,
-  martin-lof-a-la-coq, nbeincoq} uses a relational model to derive the
-injectivity of $\Pi$ types and justify the validity of
-$\eta$-conversion among other properties.
-Since \lang{} uses an untyped conversion rule, type conversion can be
-done through the \emph{normalize-and-compare}
-strategy described in \citet{pierce2004advanced}. The decidability of
-normalize-and-compare is implied by the existence of
-$\beta\eta$-normal form, which follows from our logical
-predicate.
-
+% First, the metatheoretic properties that we prove are indeed simpler.  Unlike
+% developments that mechanize the correctness of a type-directed conversion
+% algorithm, we only show the existence of normal forms for open and closed
+% terms and state our properties in terms of an untyped reduction relation. This
+% avoids a lot of the scaffolding related to the specification of the algorithm
+% and the proof obligations that the algorithm is sound and complete with
+% respect to the declarative specification of the type system.
+% As a result, our logical relation, unlike the ones from \citet{decagda,
+%   nbeincoq, martin-lof-a-la-coq, anand2014towards}, maps from types to
+% predicates rather than relations. \scw{How does this follow? Why do we need unary relations where they need binary relations?} The need for a relational model is
+% directly related to the metatheoretic property one wants to
+% prove. \citet{anand2014towards} requires a relational model to capture
+% the extensionality of their type system, whereas \citet{decagda,
+%   martin-lof-a-la-coq, nbeincoq} uses a relational model to derive the
+% injectivity of $\Pi$ types and justify the validity of
+% $\eta$-conversion among other properties.
+% Since \lang{} uses an untyped conversion rule, type conversion can be
+% done through the \emph{normalize-and-compare}
+% strategy described in \citet{pierce2004advanced}. The decidability of
+% normalize-and-compare is implied by the existence of
+% $\beta\eta$-normal form, which follows from our logical
+% predicate.
+
+First, the metatheoretic properties that we prove are indeed
+simpler. Compared to Core Nuprl\citet{anand2014towards}, our system
+lacks extensionality, which would require a relational model to
+justify consistency. Compared to the systems from
+\citet{decagda,nbeincoq, martin-lof-a-la-coq}, the conversion rule for
+\lang{} is untyped and therefore we do not
+need a Kripke-style relational model to prove $\Pi$-injectivity
+among other properties. Furthermore, we prove the existence of normal
+forms, which induces a simple \emph{normalize-and-compare}
+procedure for type
+conversion~\citet{pierce2004advanced}. \citet{nbeincoq, decagda}, on
+the other hand, need
+to show how their algorithmic conversion procedure is sound and complete
+with respect to their respective declarative equational theory.
 \scw{I'm getting confused by this paragraph. Does this reorganization sense:
@@
   need only show that terms have $\beta\eta$ normal forms. }
-\yl{Makes perfect sense!}
-
-On the other hand, our definition of our logical relation does
+\yl{Makes sense. Though Abel's work doesn't use nbe but a recursive
+  binary algorithm. Rewrote the paragraph above and commented out the original}
+
+Second, the definition of our logical relation does
 contribute to a more concise proof.
@@
 each have their separate interpretations that we have to prove to be
-equal. Fortunately, this complexity is easily reconciled by the
+equal. Fortunately, this complexity is reconciled by the
 diamond property, which is easy to derive syntactically.
@@
 not satisfy even the weaker substitution property
-(Lemma~\ref{lemma:parsubst}). That is, given a term $[[A]]$
-and $a \leadsto^{lo} b$, we do not necessarily have $[[A {a}]]
-\leadsto^{lo} [[A {b}]]$ since the variable might appear to the
-right of some redex. Eventually, one has to prove how the
+(Lemma~\ref{lemma:parsubst}). % That is, given a term $[[A]]$
+% and $a \leadsto^{lo} b$, we do not necessarily have $[[A {a}]]
+% \leadsto^{lo} [[A {b}]]$ since the variable might appear to the
+% right of some redex.
+With this alternative formulation, we would have to prove how the
 deterministic reduction relates to a full non-deterministic reduction
-relation to prove the fundamental theorem. In our setup, this would amount
+relation to prove the fundamental theorem. This would amount
 to proving the factorization theorem of the deterministic reduction
@@
 % out in a structure similar to the one we have presented.
-\scw{Not sure that I understand this paragraph}
-\yl{Reworded slightly to emphasize it's future work that we haven't
-  done and it is speculative}
+% \scw{Not sure that I understand this paragraph}
+% \yl{Reworded slightly to emphasize it's future work that we haven't
+%   done and it is speculative}
 

### 2024-02-26 paper/paper.tex
Fix the typos pointed out by Cassia

@@
 with simple or polymorphic types. In that context, syntactic logical relations
-can be define as simple recursive functions over the structure of types, or
+can be defined as simple recursive functions over the structure of types, or
 (in the case of recursive types) defined over the evaluation steps of the
@@
 systems and algorithms, these developments range in size from 20,000 to
-400,000 lines of code. As a result, their use of logical relations arguments
+400,000 lines of code. As a result, their use of logical relations are
 difficult to isolate from their surrounding contexts, and thus, not accessible
@@
 Not only does this connection aid readers that wish to, like us, adopt proof
-assistants for their day-to-day use. We also find that this precision is
+assistants for their day-to-day use, we also find that this precision is
 important for conveying the proof technique itself. Unlike properties that are
@@
 it helps us keep our definition precise and unambiguous. A precise definition
-is crucial as the reasoning principles we are allowed to use is directly tied
+is crucial as the reasoning principles we are allowed to use are directly tied
 to the definition, which is sometimes obscured by concise notations that
@@
 design choice that impacts our proofs is the use of an untyped conversion
-rule, inspired by Pure Type Systems~\citet{barendregt1991introduction}, and
-specified through parallel reduction to normal form~\citet{barendregt}.
+rule, inspired by Pure Type Systems~\citep{barendregt1991introduction}, and
+specified through parallel
+reduction~\citep{takahashi-parallel-reduction, barendregt:lambda-calculi-with-types} % to normal form~\citet{barendregt}
+.
 Because we believe that a good coverage of features makes our proof applicable
@@
 contribution of this section is the definition of the logical relation as an
-inductively defined relation and demonstrate how to reason about this
+inductively defined relation and demonstration of how to reason about this
 definition. We also show that the logical relation respects evaluation
@@
 
-In Section~\ref{sec:logrelproof}, shows that our definitions can be the
+In Section~\ref{sec:logrelproof}, we show that our definitions can be the
 subject of well-established reasoning techniques for logical relations. In
@@
 
-In Section~\ref{sec:extension}, extends the definition of the logical relation
-to include terms with free variables, and proves that every well-typed term
+In Section~\ref{sec:extension}, we extend the definition of the logical relation
+to include terms with free variables, and prove that every well-typed term
 has a $\beta$-normal form. After laying the foundation for handling type-level
-computation in Section~\ref{sec:logreldep}, this extension requires mechanical
+computation in Section~\ref{sec:logreldep}, we show how this extension requires mechanical
 changes that directly mirror an analogous extension for a simply typed
@@
 
-In Section~\ref{sec:betaeta} we show how to incorporate the function $\eta$-law
+In Section~\ref{sec:betaeta}, we show how to incorporate the function $\eta$-law
 into the conversion rule for \lang{}.
@@
 First, the metatheoretic properties that we prove are indeed
-simpler. Compared to Core Nuprl\citet{anand2014towards}, our system
+simpler. Compared to Core Nuprl~\citep{anand2014towards}, our system
 lacks extensionality, which would require a relational model to

### 2024-02-26 paper/paper.tex
comment about intro. its not done yet

@@
 proofs for dependent types.
+
+\scw{TODO: this introduction is still not done. At this point, I think the
+  first part has gotten a little long. We could say what we mean more
+  concisely. We also need to work on the overview at the end of this section
+  --- it need to be more descriptive about contributions and less of a table
+  of contents.}
+
 % % For

### 2024-02-27 paper/paper.tex
pass through eta section

@@
 \label{sec:betaeta}
+\scw{Need an introduction to this section. Why do we care about this new 
+rule? What does this section demonstrate?}
+\scw{The way we describe this
+section should more clearly indicate that we have a mechanized version of this 
+proof. And include pointers on where readers might refer to it.}
 We can easily incorporate the function $\eta$ law to the equational
@@
 % used for type conversion, is built on top of parallel reduction.
-We can recover the same confluence result about parallel reduction
-using the standard techniques from
-\citet{barendregt:lambda-calculi-with-types,
+We can recover the same confluence result about parallel reduction using the
+standard techniques from \citet{barendregt:lambda-calculi-with-types,
   takahashi-parallel-reduction}, though anti-renaming
-(Lemma~\ref{lemma:parantirenaming}) needs to be proven before the
-diamond property (Lemma~\ref{lemma:pardiamond}). Another complication
-is that confluence and anti-renaming now requires induction on a size
-metric of lambda terms so the $\eta$ case has a useful induction hypothesis.
-
-Note that the specification of our logical relation does not require
-any changes. We can continue using $\ottkw{ne}$ and $\ottkw{nf}$ to
-represent $\beta$-neutral and normal forms. By reproving the
-fundamental lemma, we learn that every well-typed term as a
-$\beta$-normal form. Lemma~\ref{lemma:parnenf} still holds with our
-extended parallel reduction. In the presence of the $\eta$ reduction
+(Lemma~\ref{lemma:parantirenaming}) must be proven before the diamond property
+(Lemma~\ref{lemma:pardiamond}). Another complication is that
+confluence\scw{diamond property?} and anti-renaming are now proven through
+induction on a size metric of lambda terms; \rref{P-AbsEta} reduces a term
+that is not a strict subterm.
+
+Note that, after this extension, the specification of our logical relation does not require
+any updates. As before, $\ottkw{ne}$ and $\ottkw{nf}$ 
+represent $\beta$-neutral and $\beta-$normal forms, and the 
+fundamental lemma shows us that every well-typed term has a
+$\beta$-normal form. However, in the presence of the $\eta$ reduction
 rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
@@
 
-In \citet{nbeincoq,decagda,martin-lof-a-la-coq}, a relational model is
-needed to justify the $\eta$-law for functions. Instead, we prove
-confluence for our parallel reduction extended with $\eta$-law for functions
-and continue using a logical predicate (i.e. a unary logical relation) to justify
-$\eta$-law as part of our equational theory.
+%% SCW: This is out of place, and redundant with the next paragraph.
+% In contrast to our approach, related work \cite{nbeincoq,decagda,martin-lof-a-la-coq}
+% employs a relational model to
+% justify the $\eta$-law for functions. 
+% Instead, we prove
+% confluence for our parallel reduction extended with $\eta$-law for functions
+% and continue using a logical predicate (i.e. a unary logical relation) to justify
+% $\eta$-law as part of our equational theory.
 A well-known issue with our approach is the failure of syntactic
@@
 type annotation. We believe the same approach is applicable to our
-proof if we add type annotations to our lambda terms.
+proof.
 
 The bigger issue is extensions such as $\eta$-laws for unit and
-products. Surjective pairing, for example, is an equational law that
-is known to cause trouble since it is not confluent for untyped lambda
-terms~\citet{KLOP198997}. The relational, type-annotated, and Kripke-style models from
+products. Surjective pairing, for example, is not confluent for untyped lambda
+terms~\cite{KLOP198997}. The relational, type-annotated, and Kripke-style models from
 \citet{nbeincoq,decagda,martin-lof-a-la-coq} can be more easily
@@
 dependently typed but without large
-elimination~\citep{harper2005equivalence,
+eliminations~\citep{harper2005equivalence,
 abel2005untypedconvsurjective}. Common workarounds include
 type-directed conversion and shifting the focus to obtaining
-$\eta$-long forms~\citet{Abel12}.
-We believe our simple proof will help readers
-see how techniques from these simpler settings generalize to fully
-dependently typed languages and better understand the more complex
-mechanized developments.
-
-Finally, due to some technical
+$\eta$-long forms~\cite{Abel12}.
+
+While not without limitations, our simple proof demonstrates the core
+building blocks of more complex arguments, thus paving the way for experimentation 
+and eventual extension to more expressive systems.\scw{I tried to reword your 
+sentence, but I am still not happy with it.}
+
+\scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this 
+sentence elsewhere? I'd like to use the previous sentence as a conclusion for 
+this section.}
+Finally, due to technical
 details, our mechanized proof about the existence of $\beta\eta$ form does not

### 2024-02-27 paper/paper.tex
pass through 8.1

@@
 Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
-code\footnote{calculated by the \texttt{cloc} tool} for each file of our
+code\footnote{calculated by the \texttt{cloc} tool, available from \url{https://github.com/AlDanial/cloc}.} for each file of our
 development, including the base consistency proof from
@@
 
-\def\rot{\rotatebox}
+% \scw{Maybe it would be useful to include a chart here, so that readers can
+%   easily keep track of the features of the various languages.  i.e. which ones
+%   include large eliminations? type-directed equivalence? impredicative prop?
+%   inductive datatypes? what are their line counts?}
+% \yl{resolved}
+
+\subsection{Logical relations for dependent types}
+In the most general sense, a logical relation can be viewed as a
+practical technique that uses a type-indexed relation to assign
+meanings to types and strengthen the induction hypothesis for the
+property of interest. The original idea of this technique can be
+traced back to
+\citet{tait1967:reducibility}. This proof maps
+types to sets of terms satisfying certain properties related to reduction.
+The same idea is explained in \citet{girard1989proofs} and extended to
+prove the strong normalization property of System F.
+
+Tait's method has been successfully applied to dependently typed
+languages to prove strong normalization \cite{Martin-Lof-1973,luo1990extended,geuvers1994short, barendregt:lambda-calculi-with-types}.
+
+However, the pen-and-paper representation of logical relations proofs
+can be challenging to adapt to a theorem prover since many details
+are hidden behind concise notations.
+For example, \citet{geuvers1994short} presents the interpretation for types as
+an inductively defined total function over the set of syntactically
+well-formed types. \scw{The issue is that Geuver's metalogic is (untyped) set 
+theory, but Coq and Agda use type theory.}
+In untyped set theory, it makes sense to define the logical relation as a
+simply-typed function that takes a type and returns some set; however
+in constructive type theory, the metalogic of Coq and Agda, the
+interpretation function must be a dependently-typed function whose
+return type depends on the derivation of the well-typedness of its
+input. The well-typedness derivation and the proof of the
+classification theorem\scw{what is the classification theorem}
+are examined in the body
+of the interpretation function to decide whether an argument of an
+application should be erased during interpretation. As a result, this 
+definition causes difficulties for modern proof assistants.
+Due to the
+impredicativity of the object language, \citet{geuvers1994short}'s
+proof cannot be encoded in Agda, which has a predicative
+metatheory. Due to the use of proof-relevant derivations, even in
+Coq, a proof assistant that supports impredicativity, one would need
+to constantly juggle between the impredicative but irrelevant sort
+\texttt{Prop} sort and the predicative but relevant sort
+\texttt{Type}.
+
+More recent work such as \citet{Abel12}
+and \citet{abel2008betaeta} make their definitions more explicit and
+precise and thus more directly encodable in proof assistants. 
+Our logical relation resembles their definition of a
+semantic universe hierarchy, although we close our relation 
+under expansion with respect to parallel reduction rather
+than weak-head reduction. \scw{Why is this important?}
+Furthermore, \citet{Abel12} and
+\citet{abel2008betaeta} use their semantic universe hierarchy as a
+measure to define Kripke-style logical relations, from which they
+derive the correctness of their conversion algorithms. In our work, we
+use the semantic universe hierarchy directly in our definition of
+semantic typing because it is sufficient for our purposes
+(consistency and normalization).
+
+\subsection{Mechanized logical relations for dependent types}
+
+\newcommand\dheader[1]{\rotatebox{60}{{#1}}}
 \newcommand\header[1]{\rotatebox{90}{{#1}}}
@@
     Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
-    \hline
-    CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
-    CoreSpec & \\
-    PCUIC & \\
+    % \hline
+    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
+    % CoreSpec & \\
+    % PCUIC & \\
     %   & \lang{} & \citet{anand2014towards} &
@@
 
-% \scw{Maybe it would be useful to include a chart here, so that readers can
-%   easily keep track of the features of the various languages.  i.e. which ones
-%   include large eliminations? type-directed equivalence? impredicative prop?
-%   inductive datatypes? what are their line counts?}
-% \yl{resolved}
-
-\subsection{Logical relations}
-Proof by logical relations does not come with a precise definition. In
-the most general sense, a logical relation can be viewed as a
-practical technique that uses a type-indexed relation to assign
-meanings to types and strengthen the induction hypothesis for the
-property of interest. The original idea of this technique can be
-traced back to
-\citet{tait1967:reducibility}. \citet{tait1967:reducibility} maps
-types to sets of terms satisfying certain properties related to reduction.
-The same idea is explained in \citet{girard1989proofs} and extended to
-prove the strong normalization property of System F.
-
-Tait's method has been successfully applied to dependently typed
-languages.  \citet{Martin-Lof-1973}, \citet{luo1990extended},
-\citet{geuvers1994short}, and
-\citet{barendregt:lambda-calculi-with-types} are some of the earlier
-works that prove strong
-normalization for different dependently type systems using Tait's notion of
-reducibility.
-
-\scw{Should give Martin L\"of credit for first consistency proof.}
-% \scw{What techniques do these three proofs use? How is it different from
-% your proof? What about Luo, which
-% introduced a universe hierarchy? What about categories with families, or
-% other ways of showing consistency?}
-% \yl{resolved}
-
-% Some of these proofs are still reapplied in more recent works. For
-% example, the technique for proving strong normalization by
-% \citet{geuvers1994short} is adapted by \citet{moon2021graded} to show
-% the normalization property for a dependently typed system extended with
-% modalities.
-
-The pen and paper representation of the logical relation proofs
-can be challenging to adapt to a theorem prover since a lot of details
-are hidden behind the concise notations.
-As an example, \citet{geuvers1994short} presents the interpretation for types as
-an inductively defined total function over the set of syntactically
-well-formed types. Despite the notation for the logical relation as a
-simply typed function that takes a type and returns some set, the
-interpretation function in reality is a dependent function whose
-return type depends on the derivation of the well-typedness of its
-input. The well-typedness derivation and the proof of the
-classification theorem are needed relevantly in the body
-of the interpretation function to decide whether an argument of an
-application should be erased during interpretation. Due to the
-impredicativity of the object language, \citet{geuvers1994short}'s
-proof cannot be encoded in Agda, which has a predicative
-metatheory. Due to the use of proof-relevant derivations, even in
-Coq, a proof assistant that supports impredicativity, one would need
-to constantly juggle between the impredicative but irrelevant sort
-\texttt{Prop} sort and the predicative but relevant sort
-\texttt{Type}.
-
-More recent works such as \citet{Abel12}
-and\citet{abel2008betaeta} make their definitions more explicit and
-precise and thus easier to encode in proof assistants. Our definition
-of logical relation is reminiscent of their definition of a
-semantic universe hierarchy, though our logical relation is defined to
-be closed under expansion with respect to parallel reduction rather
-than weak head reduction. Furthermore, \citet{Abel12} and
-\citet{abel2008betaeta} use the semantic universe hierarchy as a
-measure to define Kripke-style logical relations, from which they
-derive the correctness of their conversion algorithms. In our work, we
-use the semantic universe hierarchy directly in our definition of
-semantic typing because it is sufficient
-for deriving consistency and later normalization.
-
-\subsection{Mechanized logical relations about dependent types}
 In Figure~\ref{fig:featurematrix}, we give an overview of features of

### 2024-02-27 paper/paper.tex
midstream check in in 8.2

@@
 
-\newcommand\dheader[1]{\rotatebox{60}{{#1}}}
-\newcommand\header[1]{\rotatebox{90}{{#1}}}
+%\newcommand\dheader[1]{\rotatebox{60}{{#1}}}
+\newcommand\header[1]{\rotatebox{60}{{#1}}}
 
@@
 % \begin{minipage}{0.8\textwidth}
-  \begin{tabular}{ l |  l  | l | l | l }
+  \begin{tabular}{| l |  l  | l | l | l | }
+    \hline
       & \header{Universes} & \header{Inductives} & \header{Conversion}
@@
     \hline
-    \lang{} & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
+    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
     % \hline
@@
     Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
-    % \hline
+    \hline 
     % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
@@
   \end{tabular}
-% \end{minipage}
 
   \begin{tabular}{ll}
+  \textbf{Features:} \\
+  Universes: & Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
+  Inductives: & Identity types (Id), Natural numbers (Nat), W-types~\cite{w-types} \\
+  Conversion:& Typed (T), Untyped (U), Extensional (E) \\
     \\
+  \textbf{Citations:}\\
   $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
@@
   Core Nuprl &\citet{anand2014towards} \\
-    CIC &\citet{barras2012semantical} \\
-    CoreSpec & \\
-    PCUIC & \\
-
-  \\
-  Universes: &Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
-  Conversion:& Typed (T), Untyped (U), Extensional (E) \\
   \end{tabular}

### 2024-02-27 paper/paper.tex
Apply some of Jonathan's feedback

@@
 eschew inductive datatypes or W-types, but we do include propositional
-equality and booleans in order to capture the challenges presented by indexed
+equality and booleans to capture the challenges presented by indexed
 types and dependent pattern matching. We do not show the decidability of type
-checking, nor do we develop a PER semantics, but we do demonstrate how our
+checking, nor do we develop a PER semantics, but we prove logical consistency, which states that empty types are not inhabited in an empty context, and demonstrate how our
 consistency proof can be extended (at a moderate cost of 500 lines of code) to
-show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
+show the existence of $\beta\eta$-normal forms for well-typed open \emph{and}
 closed terms. We include a full predicative universe hierarchy and
-type-level computation, to demonstrate the logical strength of the
+type-level computation to demonstrate the logical strength of the
 approach. % \scw{Say that we don't have impredicativity, but neither does anyone
@@
 navigate and understand. We accompany this mechanized proof with an informal
-description, presented here using set theory notation and terminology so that
-the material is accessible to readers with general mathematical background,
+description, presented here using set-theoretic notation and terminology so that
+the material is accessible to readers with a general mathematical background.
 That said, our explanations do not stray too far away from our proof scripts.
-We label each lemma directly to its counterpart in the proof script,
+We link each lemma directly to its counterpart in the proof script,
 anticipating that readers may wish to reference and validate our formalism.
-Our typeset representations are purposefully made to directly follow our
-mechanical proofs, while avoiding, as much as possible, the introduction
-of artifacts that are specific to theorem provers.
+The typeset proofs purposefully follow the
+mechanized proofs closely while avoiding, as much as possible, artifacts specific to Coq.
 
@@
 it helps us keep our definition precise and unambiguous. A precise definition
-is crucial as the reasoning principles we are allowed to use are directly tied
+is crucial, as the reasoning principles we are allowed to use are directly tied
 to the definition, which is sometimes obscured by concise notations that
@@
 rule, inspired by Pure Type Systems~\citep{barendregt1991introduction}, and
-specified through parallel
+specified by parallel
 reduction~\citep{takahashi-parallel-reduction, barendregt:lambda-calculi-with-types} % to normal form~\citet{barendregt}
@@
 motivate where a proof by logical relations is required. The most important
-contribution of this section is the definition of the logical relation as an
-inductively defined relation and demonstration of how to reason about this
+contributions of this section are the definition of the logical relation as an
+inductively defined relation and demonstrating how to reason about this
 definition. We also show that the logical relation respects evaluation
 (referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
-prove that the logical relation is a partial function; that is, each type can
+prove that the logical relation is a partial function: that is, each type can
 have at most one unique interpretation.
@@
 established the base techniques, we can port ideas that have been studied in
-the context of a simpler system to the dependently-typed setting.  We believe
+the context of a simpler system to the dependently typed setting.  We believe
 it is possible to extend our logical relation to a Kripke-style and relational
@@
 
-In Section~\ref{sec:betaeta}, we show how to incorporate the function $\eta$-law
+In Section~\ref{sec:betaeta}, we show how to incorporate the function $\eta$ law
 into the conversion rule for \lang{}.
 
-In Section~\ref{sec:logrelmech}, we discuss the details about our
+In Section~\ref{sec:logrelmech}, we discuss the details of our
 mechanization, including our use of existing libraries such as
@@
         % [[id i]] & := & [[i]] \\ \\
-        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} \mathsf{id}_{\mathit{tm}}} \\
+        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
         [[idtm n]] & := & [[var n]] \\ \\
@@
         [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
-        \multicolumn{3}{l}{\mbox{\emph{Substitution operation:}} ([[ a { rho } ]])} \\
+        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
         [[var n { rho }  ]] & := &  [[rho n]] \\
@@
 
-We adapt from \citet{autosubst2} the notations for simultaneous renaming,
+We adapt from \citet{autosubst2} notation for simultaneous renaming,
 substitution, and other auxiliary definitions used in our paper, summarized in
@@
 used in its conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). This
-relation is often called ``definitional equality'' in dependent type theories
+relation is often called \textit{definitional equality} in dependent type theories
 because it defines the equivalence that the syntactic type system works
-``up-to''.
+up to.
 
@@
 \citet{takahashi-parallel-reduction}. However, this choice simplifies later
-proofs, as we discuss in Section~\ref{XXX}. \scw{Forward reference to further
+proofs, as we discuss in Section~\ref{sec:discuss}. \scw{Forward reference to further
   discussion, comparing with whnf reduction?}
@@
 not lose generality working with a system with untyped
-conversion. Furthermore, discuss how this definition can be extended with
-$\eta$-equivalence of functions in Section~\ref{XXX}. Finally, we compare this
+conversion.
+\yl{removed the eta equivalence sentence because it might give the
+  wrong message that we extended Siles' work with eta}
+% Furthermore, discuss how this definition can be extended with
+% $\eta$-equivalence of functions in Section~\ref{XXX}
+Finally, we compare this
 definition with type-directed approaches to equality in
@@
 Although the use of de Bruijn indices complicates the variable rule, the rules
-that involve binding, such as \rref{T-Pi,T-Abs} are straightforward.
+that involve binding, such as \rref{T-Pi,T-Abs}, are straightforward.
 
@@
 
-The elimination form for booleans, \rref{T-If} demonstrates dependent pattern
-matching.  The result type of this expression $[[A {a}]]$, is composed of some
+The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
+matching.  The result type of this expression $[[A {a}]]$ is composed of some
 \emph{motive} $[[A]]$, a type where its single free variable has been replaced
 with the condition of the if expression. When typing the true branch, this
-substitution replaces the variable by $[[true]]$, and similar for the false
+substitution replaces the variable by $[[true]]$, and similarly for the false
 branch. As a result, the type system communicates the information gained from
-the run-time test to each of the branches of the expression.
+the % run-time \yl{commented out run-time per Jonathan's comment}
+test to each of the branches of the expression.
 
@@
 $p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
-is the body of the elimination form. In this rule, $[[B]]$ is called a
-\emph{motive} and has two free variables.  The substitution for these
+is the body of the elimination form. In this rule, $[[B]]$ is the motive and has two free variables.  The substitution for these
 variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,

### 2024-02-27 paper/paper.tex
Minor

@@
 In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
-\texttt{Prop}, when \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S}
+\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S}
 has type \texttt {tm -> Prop}.  However, if desired, we could consistently
@@
 In axiom-free Coq, predicates do not come with the extensionality
-property. Given two predicates $P$ and $Q$, we can not conclude that
+property. Given two predicates $P$ and $Q$, we cannot conclude that
 $P = Q$ when given a proof of $\forall
@@
 $\beta\eta$-normal forms and can only reduce to themselves, we must
-have $[[f0 = f1]]$. Contradiction.
+have $[[f0 = f1]]$, which is a contradiction.
 
@@
 
-However, with a little more effort, it is possible to recover a
+However, % with a little more effort,
+it is possible to recover a
 precise algorithm. The first step is to define a deterministic
@@
 \citet{takahashi-parallel-reduction, factorization-essentially}. By
-composing the existence of normal form and the fact that the
+composing the existence of normal forms and the fact that the
 deterministic relation is normalizing, we can derive the
@@
   \texttt{Prop} data type with a single constructor, to be eliminated
-  to construct runtime relevant data}.
+  to construct runtime relevant data.}.
 
@@
 
-\citet{decagda} mechanizes in Agda the decidability of type
+\citet{decagda} mechanize in Agda the decidability of type
 conversion rule for a dependently typed language with one predicative
 universe level and typed judgmental equality with function
-$\eta$-law. They
+$\eta$ law. They
 use a Kripke-style logical relation parameterized over a
@@
 
-\citet{anand2014towards} mechanizes the metatheory of
+\citet{anand2014towards} mechanize the metatheory of
 Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
@@
 inductive types, and a full universe hierarchy. \citet{nbeincoq}
-mechanizes the normalization-by-evaluation algorithm in Coq for a
+mechanize the normalization-by-evaluation algorithm in Coq for a
 dependently typed language with one predicative universe, similar to

### 2024-02-27 paper/paper.tex
Add citations

@@
 about lambda calculi and their operational semantics. They have been used to
-show diverse properties such as strong normalization, contextual equivalence,
-representation independence, noninterference, compiler correctness, and the
-decidability of conversion algorithms. \scw{add citations? emphasize that we are working
+show diverse properties such as strong
+normalization~\citep{girard1989proofs,
+  geuvers1994short}, contextual equivalence~\citep{constable1986implementing},
+representation independence~\citep{pitts1998existential}, noninterference~\citep{bowman2015noninterference}, compiler correctness~\citep{benton2009biorthogonality,perconti2014compiler}, and the
+decidability of conversion algorithms~\citep{harper2005equivalence,Abel12,abel2013normalization}. \scw{% add citations?
+  emphasize that we are working
 with syntactic logical relations.}
@@
 input. The well-typedness derivation and the proof of the
-classification theorem\scw{what is the classification theorem}
+classification theorem\scw{what is the classification theorem}\yl{It
+  says all well-formed terms A (either G |- A : .. or G |- .. : A)
+then A is either an object, a type, or a kind. The classification is
+used to guide erasure}
 are examined in the body
@@
 under expansion with respect to parallel reduction rather
-than weak-head reduction. \scw{Why is this important?}
+than weak-head reduction. \scw{Why is this important?} \yl{if you are
+  referring to the sentence about weak-head reduction, it can
+  be deleted because it's discussed in Section 9 already}
 Furthermore, \citet{Abel12} and

### 2024-02-27 paper/paper.tex
more related work

--- 2024-02-27 paper/paper.tex
@@
 % \begin{minipage}{0.8\textwidth}
-  \begin{tabular}{| l |  l  | l | l | l | }
+  \begin{tabular}{| l |  l  | l | l | l | l| }
     \hline
       & \header{Universes} & \header{Inductives} & \header{Conversion}
-      & \header{Large Elim} \\
+      & \header{Large Elim} & Main results \\
     \hline
-    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
+    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} 
+    & Consistency and normalization \\
     % \hline
-    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗}\\
-    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} \\
+    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & Consistency \\
+    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & Correctness of NBE \\
+
     % \hline
-    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} \\
+    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} & Decidability of conversion \\
     % \hline
-    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} \\
+    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
     % \hline
-    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
+    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & TODO \\
     \hline
@@
   \begin{tabular}{ll}
-  \textbf{Features:} \\
+    \\
   Universes: & Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
-  Inductives: & Identity types (Id), Natural numbers (Nat), W-types~\cite{w-types} \\
-  Conversion:& Typed (T), Untyped (U), Extensional (E) \\
+  Inductives: & Identity types (Id), Natural numbers (Nat), W-types \\
+  Conversion:& Untyped (U), Typed (T),  Extensional (E) \\
+  Large Eliminations:& included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
   \\
-  \textbf{Citations:}\\
   $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
@@
 
-In Figure~\ref{fig:featurematrix}, we give an overview of features of
-the object languages from mechanized logical relations about dependent
-types. The table is far from exhaustive. For example,
+Figure~\ref{fig:featurematrix} presents several mechanized proofs 
+that feature logical-relations arguments for dependently-typed languages.
+Each of these proofs is significantly larger than than our development; but they 
+also prove more results about different object languages.
+The table provides a comparison between the various features of their 
+object languages, but is not exhaustive. For example,
 \citet{casinghino:combining-proofs-programs} and
@@
 which is available in Agda but not in Coq.
-\citet{martin-lof-a-la-coq} manages to encode the logical relation
+\citet{martin-lof-a-la-coq} transports the logical relation
 from \citet{decagda} in the predicative fragment of Coq and further
@@
 extensional type theory with features such as dependent functions,
-inductive types, and a full universe hierarchy. \citet{nbeincoq}
+inductive types, and a full universe hierarchy. \scw{What do they prove?}
+\citet{nbeincoq}
 mechanize the normalization-by-evaluation algorithm in Coq for a
@@
 to encode a finite universe hierarchy without the use of
-either impredicativity or induction-recursion.
+either impredicativity or induction-recursion.\scw{How did the Nuprl 
+proof include a full hierarchy? Did they use impredicativity?}
 % \yl{commented out because I think this paragraph is too subjective}
@@
 
-Finally, \citet{barras2010sets, Wang2013SemanticsOI} assign
+\citet{barras2010sets, Wang2013SemanticsOI} assign
 set-theoretic semantics to dependent type theory in Coq. Unlike the
-previous mechanization efforts, which primarily focus on predicative
+previous efforts, which primarily focus on predicative
 type theory and more direct reducibility models,
@@
 of the impredicative sort in Calculus of Constructions. We choose to
-focus on a simple term model so we do not have to take the extra step
+focus on a syntactic term model so we do not have to take the extra step
 of mechanizing mathematical objects such as sets and domains.
 
-There are other mechanized developments for dependently typed systems
-that only involve properties that are derivable through syntactic
-means. For example, \citet{coqcoqcorrect2019} proves the correctness
-of a type checker for the Polymorphic, Cumulative Calculus of
-Inductive Constructions (PCUIC), Coq's core calculus, assuming the
-strong normalization property of the object
+There are other mechanized developments for dependently typed systems that
+only involve properties that are derivable through syntactic means. For
+example, \citet{coqcoqcorrect2019} proves the correctness of a type checker
+for the Polymorphic, Cumulative Calculus of Inductive Constructions (PCUIC),
+Coq's core calculus, assuming the strong normalization property of the object
 language. \citet{weirich:systemd} define System D, a core calculus of
@@
 system. Compared to properties such as consistency, normalization, and
-parametricity, syntactic type soundness is less intimidating for
-dependent types since one is most unlikely to run into issues related
-to the strength of the metatheory.
+parametricity, the most difficult part for syntactic type soundness is the
+consistency of definitional (not propositional) equality. In their setting,
+the result follows from the confluence of parallel reduction.
 

### 2024-02-27 paper/paper.tex
Minor

@@
 \]
-  \caption{Syntax of \lang \scw{Where does the name come from?}\yl{H =
-    hierarchy though it needs a better name}}
+  \caption{Syntax of \lang}
   \label{fig:syntax}
@@
 \citet{takahashi-parallel-reduction}. However, this choice simplifies later
-proofs, as we discuss in Section~\ref{sec:discuss}. \scw{Forward reference to further
-  discussion, comparing with whnf reduction?}
+proofs, as we discuss in Section~\ref{sec:discuss}. % \scw{Forward reference to further
+  % discussion, comparing with whnf reduction?} \yl{resolved?}
 
@@
 \yl{removed the eta equivalence sentence because it might give the
-  wrong message that we extended Siles' work with eta}
-% Furthermore, discuss how this definition can be extended with
-% $\eta$-equivalence of functions in Section~\ref{XXX}
+  wrong message that we extended Siles' work with eta:
+ Furthermore, discuss how this definition can be extended with
+ $\eta$-equivalence of functions in Section~\ref{XXX} }
 Finally, we compare this
@@
 % types? } \yl{resolved? impredicativity is hard}
-\scw{Is it worth observing here that this definition is not over sets of typed
-terms. That it characterizes all terms that look like booleans (i.e. evaluate to
-true or false) or all terms that look like proofs (i.e. evaluate to refl). The
-fact that there is no connection between p and a and b in the I-Eq case is strange
-looking. Need to explain.  }
+% \scw{Is it worth observing here that this definition is not over sets of typed
+% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
+% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
+% fact that there is no connection between p and a and b in the I-Eq case is strange
+% looking. Need to explain.  }
+% \yl{resolved}
 The metavariables $[[I]]$ and $[[S]]$ are
@@
 \label{sec:logrelproof}
-\yl{todo: rename all variables from i,j to n}
 \scw{Would it make sense to define the notation $a \in [\![A]\!]^i$ when

### 2024-02-27 paper/paper.tex
Add the missing entry for nuprl
Add new overview

@@
 consistent with our mechanization.
+
+\yl{new overview starts here}
+Here is an overview of our contribution.
+\begin{itemize}
+\item In Section~\ref{sec:spec},
+we present the type system of \lang{},
+a dependently typed system with an untyped conversion rule in the
+style of Pure Type Systems~\citep{barendregt1991introduction}. The
+\lang{} language supports many common features of a dependent
+type theory: a countable universe hierarchy, a boolean base
+type, an intensional identity type, and support for large elimination.
+\item In Section~\ref{sec:logreldep}, we formulate logical consistency for \lang{}, the property of
+interest, to motivate a logical relation, and proceed to define it as
+an inductive relation which we later prove to be a partial
+function. Based on the definition of the logical relation, we define
+a notion of semantic typing and prove the fundamental theorem, from
+which logical consistency follows as a corollary
+(Section~\ref{sec:logrelproof}). Thanks to the good coverage of
+features in our object language, our proof showcases the
+special treatment required to model those features, thus making
+our proof applicable to a broad range of type systems.
+\item We strengthen our logical relation to prove the existence of
+  $\beta$-normal forms (Section~\ref{sec:extension}) and
+  $\beta\eta$-normal forms (Section~\ref{sec:betaeta}) for well-typed
+  open \emph{and} closed terms. The modifications made to the logical
+  relation are small and closely mirror an analogous extension for a
+  simply typed language. We use these extensions to demonstrate the idea
+  that once we have established the base techniques, we can port ideas
+  that have been studied in the context of a simpler system to the
+  dependently typed setting.
+\item We mechanize all our proofs using the Coq proof assistant, with
+  963 lines of code for the consistency proof and a moderate increase
+  to 1357 lines of code for the normalization proof.
+  We discuss the specifics
+  related to our choice of Coq as our metatheory, including our use of
+  off-the-shelf semantic engineering infrastructure and automation
+  tools, in Section~\ref{sec:logrelmech}.
+  Our proof scripts, with comments, are available to reviewers as
+  supplementary materials. We use Coq's custom notations to ease the
+  effort of matching the pen and paper representation to our formalism.
+ \yl{I want to say something
+  about the benefits of using existing libraries such as Autosubst 2
+  and CoqHammer rather than rolling your own undocumented libraries that nobody
+  want to (or can) reuse.}
+Our development relies on off-the-shelf tools...
+\item We discuss in Section~\ref{sec:relatedwork} how our proof
+relates to existing proofs by logical relations and other proof
+techniques for proving consistency and normalization. In
+Section~\ref{sec:discuss}, we give an in-depth explanation on how
+various design decisions (e.g. the choice of untyped equality) affect
+the size of our proof and its extensibility to features such as
+type-directed conversion.
+\end{itemize}
+We hope our success at creating a short and mechanized proof for a
+relatively feature-complete language will encourage future researchers
+to leverage the tool of logical relations in mechanized
+proofs for dependent types.
+
+\yl{new overview ends here}
 
@@
     % \hline
-    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & TODO \\
+    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & Consistency \\
     \hline

### 2024-02-27 paper/paper.tex

@@
 \mathit{Terms}\\
-[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
+[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[x]]\  |\ [[Void]]
                   & \mbox{universes, variables, empty type} \\
-            & |   & [[Pi A B]]\ |\ [[\ a]]\ |\ [[a b]]
+            & |   & [[Pi x : A . B]]\ |\ [[\ x . a]]\ |\ [[a b]]
                   & \mbox{function types, abstractions, applications} \\
@@
 \mathit{Substitutions}\\
-[[rho]] & \in & [[SNat -> STm]] & \\ \\
+[[rho]] & \in & [[SVar -> STm]] & \\ \\
 \mathit{Typing Contexts}\\
- [[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
+ [[G]]       & ::= & [[empty]]\ |\ [[G ++ x : A]] &  \\ \\
 \end{array}
@@
 type theory, terms and types are collapsed into the same syntactic
-category. The type $[[Set i]]$ represent universe types and $[[var n]]$
-represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
-$[[n]]$ are all metavariables representing natural numbers, we always use
-$[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
-variables.  We use de Bruijn indices for both our Coq development and the
-informal presentation here because it is more amenable to automation
-(Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
-conditions. As a result, variables do not appear in binding forms, such as in
-abstractions $[[ \ a ]]$ and dependent function types $[[Pi A B]]$.  This latter
+category. The type $[[Set i]]$ represent universe types.%  and $[[x]]$
+% represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
+% $[[n]]$ are all metavariables representing natural numbers, we always use
+% $[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
+% variables.
+% We use de Bruijn indices for our Coq development, but use nominal
+% representation in the paper because it is more amenable to automation
+% (Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
+% conditions.
+\yl{need to fix the text}
+As a result, variables do not appear in binding forms, such as in
+abstractions $[[ \ x . a ]]$ and dependent function types $[[Pi x : A . B]]$.  This latter
 form binds a new variable of type $[[A]]$ that can appear in the codomain of
@@
       \begin{array}{lll}
-        % \mathit{IdentityRen} \\
-        % [[id i]] & := & [[i]] \\ \\
         \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
-        [[idtm n]] & := & [[var n]] \\ \\
-        \multicolumn{3}{l}{\mbox{\emph{Weakening substitution: }} \uparrow^{n}} \\
-        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]} \\ \\
-        %
-        % \mathit{ConsRen ([[xi .: n]])} \\
-        % ([[xi .: n]])([[0]]) & := & [[n]] \\
-        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
-        %
-        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: a]]) } \\
-        ([[rho .: a]])([[0]]) & := & [[a]] \\
-        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
+        [[idtm x]] & := & [[x]] \\ \\
+        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho
+        .: x -> a]]) } \\
+        ([[rho .: x -> a]])([[x]]) & := & [[a]] \\
+        ([[rho .: x -> a]])([[y]]) & := & [[rho y]] \text { if  } y
+                                          \neq x\\ \\
       \end{array}
@@
       \begin{array}{lll}
-        \multicolumn{3}{l}{\mbox{\emph{Lifting operation: }} \Uparrow\rho} \\
-        [[( up rho ) 0]] & := & [[var 0]] \\
-        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
         \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
-        [[var n { rho }  ]] & := &  [[rho n]] \\
-        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
+        [[x { rho }  ]] & := &  [[rho x]] \\
+        [[(Pi x : A . B) { rho }]] & := & [[Pi x : A { rho } . B { rho .: x -> x }]] \\
         [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
-        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
+        [[(\ x . a) { rho }]] & := & [[\ x . (a { rho .: x -> x })]] \\
         \ldots
@@
 eliminations, the ability to compute a type using a term as input. For
-example, the function $[[\ if var 0 Bool Bool -> Bool]]$ returns
-either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
-is $[[true]]$ or $[[false]]$.
-
+example, the function $[[\ x . if x Bool Void]]$ returns
+either $[[Bool]]$ or $[[Void]]$ depending on whether its
+input is $[[true]]$ or $[[false]]$.
+
+\yl{Need to fix the de bruijn text}
 We adapt from \citet{autosubst2} notation for simultaneous renaming,
@@
 example of a substitution is the function $\ottkw{id}_{tm}$, the identity
-substitution that maps each index to the corresponding term variable.  The
-$\uparrow^{[[n]]}$ substitution is more general and adds $[[n]]$ to the index
-before injecting it as a term. The extension operation, $[[(rho .: a)]]$,
-allows us to extend a substitution $[[rho]]$ with an extra element $[[a]]$;
-the result maps the variable $[[var 0]]$ to $[[a]]$ and variables
-$[[var Suc i]]$ to $[[rho i]]$.
+substitution that maps each index to the corresponding term variable.  % The
+% $...$ substitution is more general and adds $[[n]]$ to the index
+% before injecting it as a term.
+The extension operation, $[[(rho .: x -> a)]]$,
+allows us to update the substitution $[[rho]]$ to map the variable
+$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.
 
 The substitution operator, which takes the form $[[a {rho}]]$,
-traverses the syntax of $[[a]]$ and replaces each variable $[[var i]]$
-with the term $[[rho i]]$. When traversing under binders (e.g. in the
-$[[(\ a) { rho }]]$ case), the $\Uparrow$ operator prevents $[[var
-0]]$, the bound variable, from being replaced by
-$[[rho]]$. From the definition of $[[up rho]]$, we see that it keeps
-$[[var 0]]$ unchanged during substitution and replaces each $[[var
-Suc i]]$ with $[[rho i {up 1}]]$. The shifting with $[[up 1]]$
-is required so the variables in terms from the codomain of $[[rho]]$ skip over
-the newly introduced binder to refer to the correct binding
-location.
+traverses the syntax of $[[a]]$ and replaces each variable $[[x]]$
+with the term $[[rho x]]$. When traversing under binders (e.g. in the
+$[[(\ x . a) { rho }]]$ case), we extend $[[rho]]$ to map the bound
+variable $[[x]]$ to itself, protecting it from being replaced.
 
@@
 substitution by composing the extension operator and the identity
-substitution: $[[a { b }]] := [[a { idtm .: b }]]$. In our proofs, we find it
+substitution: $[[a { b / x }]] := [[a { idtm .: x -> b }]]$.
+
+\yl{Explain why iterated substitution is not good enough}
+
+In our proofs, we find it
 more convenient to formulate simultaneous substitution directly rather than
@@
   \label{lemma:parcong}
-  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
+  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 / x } => a1 {
+    b1 / x }]]$.
 \end{lemma}
@@
   \label{lemma:parsubst}
-  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
+  If $[[a0 => a1]]$, then $[[a0 {b / x } => a1 {b / x}]]$ for arbitrary $[[b]]$.
 \end{corollary}
@@
 
-In \rref{T-Var}, $[[G n]]$ is the partial function defined by the equations
-$[[ (G ++ A) 0 ]] := [[A]]$ and $[[ (G ++ A) Suc n ]] := [[G n]]$.  The
-precondition $[[n < | G | ]]$, where $[[ | G | ]]$ represents the length of
-the context $[[G]]$, ensures that $[[G n]]$ has a defined output. The type
-$[[G n]]$ needs to be shifted by $[[Suc n]]$ since the types stored in the
-context $[[G]]$ are scoped differently. In the typing judgment
-$[[G |- a : A]]$, we want both $[[a]]$ and $[[A]]$ to have the same
-scope. However, the closer a type is to the head of $[[G]]$, the smaller de
-Bruijn it has since there is less variables in the context it can refer to
-(see \rref{Ctx-Cons}). Shifting $[[G n]]$ by $[[Suc n]]$ weakens the type so
-it is scoped consistently with the full context $[[G]]$ rather than a
-truncated version of $[[G]]$.
-
-Although the use of de Bruijn indices complicates the variable rule, the rules
-that involve binding, such as \rref{T-Pi,T-Abs}, are straightforward.
+\yl{Var, T-Pi, T-Abs are self-explanatory now?}
+% In \rref{T-Var}, $G n$ is the partial function defined by the equations
+% $ (G ++ A) 0 := A$ and $ (G ++ A) Suc n  := G n$.  The
+% precondition $n < | G | $, where $ | G | $ represents the length of
+% the context $G$, ensures that $G n$ has a defined output. The type
+% $G n$ needs to be shifted by $Suc n$ since the types stored in the
+% context $G$ are scoped differently. In the typing judgment
+% $G |- a : A$, we want both $a$ and $A$ to have the same
+% scope. However, the closer a type is to the head of $G$, the smaller de
+% Bruijn it has since there is less variables in the context it can refer to
+% (see \rref{Ctx-Cons}). Shifting $G n$ by $Suc n$ weakens the type so
+% it is scoped consistently with the full context $G$ rather than a
+% truncated version of $G$.
+
+% Although the use of de Bruijn indices complicates the variable rule, the rules
+% that involve binding, such as \rref{T-Pi,T-Abs}, are straightforward.
 
@@
 The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
-matching.  The result type of this expression $[[A {a}]]$ is composed of some
+matching.  The result type of this expression $[[A {a / x}]]$ is composed of some
 \emph{motive} $[[A]]$, a type where its single free variable has been replaced
@@
 % hierarchy? or no?}
-
-% , the
-% statement that $\Gamma \vdash [[Pi A0 B0]] \equiv [[Pi A1 B1]]$
-% implies $\Gamma \vdash [[A0]] \equiv [[A1]]$ and
-% $\Gamma, [[A0]]\vdash [[B0]] \equiv [[B1]]$.
 
@@
 almost all cases except for \rref{T-App}. In the application
-case, we are given $[[empty |- b : Pi A B]]$ and $[[empty |- a : A]]$, and
-the equality that $[[B {a} = Bool]]$. Our goal is to show that
+case, we are given $[[empty |- b : Pi x : A . B]]$ and $[[empty |- a : A]]$, and
+the equality that $[[B {a / x} = Bool]]$. Our goal is to show that
 $[[empty |- b a : Void]]$ is not possible. However, note that there is
 nothing we know of $[[b]]$ or $[[a]]$ from the induction hypothesis
-because neither $[[Pi A B]]$ nor $[[A]]$ is equal to $[[Bool]]$.
+because neither $[[Pi x : A . B]]$ nor $[[A]]$ is equal to $[[Bool]]$.
 We have no way of deriving a contradiction from $[[empty |- b a :
@@
 In dependent types, the type
-$[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
+$[[A]]$ can take the form $[[(\ x . x ) Bool]]$. To assign meaning to
 this type, we need to first reduce it to $[[Bool]]$. However, we
@@
 of $[[Bool]]$ in our logical relation does not always correspond to
-well-typed or even closed terms. For example, we have $[[if false var 0
+well-typed or even closed terms. For example, we have $[[if false (#
+Void true #)
 true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
@@
 are an instance. \Rref{I-Red} enables us to reduce our types in order
-to assign meanings. Recall the type expression $[[(\ var 0)
+to assign meanings. Recall the type expression $[[(\ x . x )
 Bool]]$. \Rref{I-Red} says to know that
-$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
-show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
-Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
+$[[Interp I i  (\ x . x) Bool S ]]$ for some $[[S]]$, it suffices to
+show that $[[Interp I i Bool S]]$ since $[[(\ x . x) Bool =>
+Bool]]$. The derivation that $[[Interp I i (\ x . x) Bool { a | a =>+
   true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
@@
 Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
-, S0 ) | a in S , Interp I i B { a } S0 }]]$.
+, S0 ) | a in S , Interp I i B { a /x } S0 }]]$.
 \end{proof}
@@
 \rref{I-PiAlt} says that
-the function type $[[Pi A B]]$ has an interpretation if its input
+the function type $[[Pi x : A . B]]$ has an interpretation if its input
 type $[[A]]$ can be
 interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
-type $[[B {a}]]$, obtained by substituting $[[a]]$ into the output type
+type $[[B {a / x}]]$, obtained by substituting $[[a]]$ into the output type
 $[[B]]$, has some semantic interpretation. In its conclusion, we say
-that a term $[[b]]$ belongs to an interpretation of $[[Pi A B]]$ if
+that a term $[[b]]$ belongs to an interpretation of $[[Pi x : A . B]]$ if
 for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
 application form $[[b a]]$ belongs to all possible interpretations of
-$[[B {a}]]$ (the pre-condition ensures at least one interpretation
-exists for each $[[B {a}]]$ where $[[a in S]]$).
+$[[B {a  / x}]]$ (the pre-condition ensures at least one interpretation
+exists for each $[[B {a / x}]]$ where $[[a in S]]$).
 
@@
 determine the relation $[[R in S * PowerSet STm]]$ to be the relation
-$[[{ (a , S0 ) | a in S , Interp I  i B { a } S0 }]]$. This result is
+$[[{ (a , S0 ) | a in S , Interp I  i B { a /x } S0 }]]$. This result is
 formally shown through the improved inversion lemma for function types
@@
 relation directly using \rref{I-PiAlt} since the occurrence of
-$[[Interp I i B {a} S0]]$ in its conclusion not only violates the
+$[[Interp I i B {a/x} S0]]$ in its conclusion not only violates the
 syntactic strict positivity constraint required in proof assistants,
@@
 inversion principles for our logical relation. Given $[[Interp I i A
-S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi A0
-B0]]$, the inversion lemma allows us to say something about the set
+S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi x
+: A0 .B0]]$, the inversion lemma allows us to say something about the set
 $[[S]]$. Its proof is simple, but we sketch out the case for
@@
   \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
-  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S1]]$, then there exists $[[S]],[[R]]$ such that:
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp
+    I i Pi x : A . B S1]]$, then there exists $[[S]],[[R]]$ such that:
     \begin{itemize}
@@
     \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
+    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
+      i B { a /x } S0 #)]]$
     \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
@@
   function type.
-  We start by inducting over the derivation of $[[Interp I i Pi A B S]]$. There
+  We start by inducting over the derivation of $[[Interp I i Pi x : A . B S]]$. There
   are only two possible cases we need to consider.
@@
   \item[\Rref{I-Pi}:] Immediate.
-  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S1]]$.
+  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi x : A . B S1]]$.
     We know that there exists some $[[A0]]$ and
-    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
-    A0 B0 S1]]$. From the
+    $[[B0]]$ such that $[[Pi x : A . B => Pi x : A0 . B0]]$ and $[[Interp I i Pi
+    x : A0 . B0 S1]]$. From the
     induction hypothesis, there exists $[[S]]$ and $[[R]]$ such that :
@@
     \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B0 { a } S0 #)]]$
+    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
+      i B0 { a / x } S0 #)]]$
     \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
     \end{itemize}
-    By inverting the derivation of $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
-    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
+    By inverting the derivation of $[[Pi x : A . B => Pi x : A0 . B0]]$, we derive $[[A => A0]]$ and
+    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a /x} => B0 {a/x} ]]$ for all
     $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
@@
     \item $[[Interp I i A S ]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
+    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
+      i B { a /x } S0 #)]]$
     \end{itemize}
@@
   \label{lemma:piinvalt}
-  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
+  Suppose $[[Interp I i Pi x : A . B S]]$, then there exists some $[[S0]]$
   such that the following constraints hold:
@@
   \item $[[Interp I i A S0]]$
-  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
+  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a
+    / x}
     S1 #) #)]]$
   \item $[[S = { b | forall a, (# a in S0 , forall
-    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
+    S1, (# Interp I i B {a / x} S1,  b a in S1 #) #) }]]$
   \end{itemize}
@@
 there exists some $S$ such that $[[InterpR i A S]]$ and $[[a in S]]$ ?}
-% \begin{figure}[h]
-% \[
-% \begin{array}{lcl}
-
-%       [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
-%                      } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
-%       [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |= G]]
-%    \text{ then there exists some } [[j]] \text{
-%                        and } [[S]] \text{ such that } [[InterpR j A {rho} S]]
-%        \text{ and } [[a {rho} in S]] \\
-%       [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
-% \end{array}
-% \]
-%   \caption{Semantic Typing for \lang}
-%   \label{fig:semtyping}
-% \end{figure}x
-
-% The logical relation we define in Figure~\ref{fig:logrel} does not
-% include cases for variables. Likewise, for the base types such as
-% boolean and equality, the output set $[[S]]$ contains only terms that
-% evaluate to closed terms.
 
@@
 Define $[[rho |= G]]$ when
-\[ \forall n\ i\ S, \text{ if }[[n < |G|]]\text{ and
-                     } [[InterpR i (G n < up Suc n > ) { rho } S ]] \text{, then } [[rho n in S]] \]
+\[ \forall [[x]], [[A]], [[i]], \text { and } [[S]], \text{ if }[[x :
+  A in G]]\text{ and
+                     } [[InterpR i A { rho } S ]] \text{, then } [[rho x in S]] \]
 \end{definition}
@@
 The $[[rho |= G]]$ notation denotes the semantic well-formedness of a
-substitution $[[rho]]$ with respect to a context $[[G]]$. Intuitively, for any
-variable $[[n]]$ in the context (thus the restriction $[[n < | G |]]$),
-$[[rho n]]$ is a term that inhabits all possible interpretations of the type
-$[[G n]]$ from the context.
-
-Since we are in a dependently typed system, there
-may be variables in $[[G n]]$, too. The type $[[(G n < up Suc n > ) {
-  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
-operation $[[up Suc n]]$ can be viewed as an artifact of the de Bruijn
-representation to lift the type $[[G n]]$ to the right scope before
-substitution can be carried out.
+substitution $[[rho]]$ with respect to a context $[[G]]$. For every
+variable $[[x]]$ with its associated type $[[A]]$ in the context,
+$[[rho x]]$ is a term that inhabits all possible interpretations of the type
+$[[A {rho}]]$.
 
@@
 G]]$, we only need show that there exists some $[[i]]$ and $[[S]]$
-such that $[[InterpR i (G n < up Suc n > ) { rho } S]]$ and $[[rho n
+such that $[[InterpR i A { rho } S]]$ and $[[rho x
 in S]]$ since the logical relation is a partial function (Lemma~\ref{lemma:logreldeter}). This idea is encapsulated in the second of the two structural
@@
   If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
-  $[[rho .: a |= G ++ A]]$.
+  $[[rho .: x -> a |= G ++ x : A]]$.
 \end{lemma}
@@
 \begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
-Define $[[|= G]]$ when
-\[
-   \forall [[n < |G|]],\text{ there exists some } [[i]] \text{ such that } [[drop Suc n G |= G n : Set i]] \]
+Define $[[|= G]]$ inductively as follows.
+\drules[SCtx]{$[[|= G]]$}{Semantic context well-formedness}{Empty, Cons}
+% \[
+%    \forall x : A in G,\text{ there exists some } i \text{ such that } drop Suc n G |= G n : Set i \]
 \end{definition}
 
-In this definition, we use the notation $[[drop n G]]$ to denote
-the typing context obtained by dropping the last $[[n]]$ elements of
-$[[G]]$. When $[[G]]$ has fewer than $[[i]]$ elements, $[[drop n G]]$ returns
-the empty list. Intuitively, $[[|= G]]$ says that all
-types in the context must correspond to some set according to the logical
-relation. The truncation to the context $[[G]]$ is due to the telescopic
-scoping of the types in the context. An alternative approach would be to define $[[|= G]]$
-inductively, similar to $[[|- G]]$, but we find it easier to use this definition and
-recover the structural rules as lemmas.
-\begin{lemma}[Semantic context well-formedness empty]
-  \label{lemma:semwffempty}
-  $[[|= G]]$ holds when $[[G]]$ is empty.
-\end{lemma}
-\begin{lemma}[Semantic context well-formedness cons]
-  \label{lemma:semwffcons}
-  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
-\end{lemma}
+% In this definition, we use the notation $drop n G$ to denote
+% the typing context obtained by dropping the last $n$ elements of
+% $G$. When $G$ has fewer than $i$ elements, $drop n G$ returns
+% the empty list. Intuitively, $|= G$ says that all
+% types in the context must correspond to some set according to the logical
+% relation. The truncation to the context $G$ is due to the telescopic
+% scoping of the types in the context. An alternative approach would be to define $|= G$
+% inductively, similar to $|- G$, but we find it easier to use this definition and
+% recover the structural rules as lemmas.
+% \begin{lemma}[Semantic context well-formedness empty]
+%   \label{lemma:semwffempty}
+%   $|= G$ holds when $G$ is empty.
+% \end{lemma}
+% \begin{lemma}[Semantic context well-formedness cons]
+%   \label{lemma:semwffcons}
+%   If $|= G$ and $G |= A : Set i$, then $|= G ++ A$.
+% \end{lemma}
 
@@
 % structural rules:
-% \scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
-% Ctx-empty?}
-% \begin{proof}
-%   By the definition of semantic context well-formedness, the goal is
-%   to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i   (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
-%   proven by case analysis on whether $[[i]]$ is zero.
-% \end{proof}
 
@@
   \label{lemma:stvar}
-  If $[[|= G]]$ and $[[n < |G|]]$, then $[[G |= var n : G n < up Suc n  >]]$.
+  If $[[|= G]]$ and $[[x : A in G]]$, then $[[G |= x : A]]$.
 \end{lemma}
@@
   \begin{itemize}
-  \item $[[InterpR i G n < up Suc n > { rho } S]]$
-  \item $[[rho n in S]]$
+  \item $[[InterpR i A { rho } S]]$
+  \item $[[rho x in S]]$
   \end{itemize}
-  From the definition of $[[|= G]]$, we know that there exists some
-  $[[i]]$ such that $[[drop Suc n G |= G n : Set i]]$. This is not
-  immediately useful since $[[rho |= G]]$, not the truncated
-  context. We claim, without giving the proof, that from $[[drop Suc n
-  G |= G n  : Set i]]$, we can derive $[[G |= G n {up Suc n} : Set
-  i]]$. A rigorous justification would require us to formally state
-  and show renaming/weakening lemma for semantic typing. We instead
-  provide an informal justification since the proof of renaming
-  is uninteresting and only requires unfolding definitions, but its
-  statement introduces unnecessary line noise related to shifting.
-
-  Intuitively, a substitution $[[rho]]$
-  that closes over a context should always induce a truncated
-  substitution for each truncated context. As a result, if
-  some term $[[b]]$ is semantically well-typed under some truncated
-  context $[[G]]$, then it must also be semantically well-typed under
-  the context $[[G]]$ up to shifting.
-%
-  Knowing that $[[G |= G n {up Suc n} : Set i]]$ holds, we can then
-  apply Lemma~\ref{lemma:setinv} to show that
-  $[[InterpR i G n < up Suc n > { rho } S]]$. It now
-  suffices to show $[[rho n in S]]$, but that is immediate from the
-  definition of $[[rho |= G]]$.
+  \yl{todo}
+%   From the definition of $|= G$, we know that there exists some
+%   $i$ such that $drop Suc n G |= G n : Set i$. This is not
+%   immediately useful since $rho |= G$, not the truncated
+%   context. We claim, without giving the proof, that from $drop Suc n
+%   G |= G n  : Set i$, we can derive $G |= G n {up Suc n} : Set
+%   i$. A rigorous justification would require us to formally state
+%   and show renaming/weakening lemma for semantic typing. We instead
+%   provide an informal justification since the proof of renaming
+%   is uninteresting and only requires unfolding definitions, but its
+%   statement introduces unnecessary line noise related to shifting.
+
+%   Intuitively, a substitution $rho$
+%   that closes over a context should always induce a truncated
+%   substitution for each truncated context. As a result, if
+%   some term $b$ is semantically well-typed under some truncated
+%   context $G$, then it must also be semantically well-typed under
+%   the context $G$ up to shifting.
+% %
+%   Knowing that $G |= G n {up Suc n} : Set i$ holds, we can then
+%   apply Lemma~\ref{lemma:setinv} to show that
+%   $InterpR i G n < up Suc n > { rho } S$. It now
+%   suffices to show $rho n in S$, but that is immediate from the
+%   definition of $rho |= G$.
 \end{proof}
@@
 % following condition holds.
-% \[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
-%   [[D ren xi  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]
-
-% \begin{lemma}[Truncate is valid]
-%   \label{lemma:validtruncate}
-%   If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i   G]]$ to $[[G]]$.
-% \end{lemma}
-% \begin{proof}
-%   By the definition of a valid renaming, we must show that given $[[j   < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
-%   \begin{itemize}
-%   \item $[[ren up i j]] = i + j  < [[| G | ]]$
-%   \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
-%   \end{itemize}
-%   The first bullet point is immediate from the fact that $[[j]] <
-%   [[|G|]] - [[i]]$. The second bullet follows from unfolding the
-%   definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
-%   $[[i]] + [[j]] < [[|G|]]$.
-% \end{proof}
-
-% \begin{lemma}[Renaming for $[[rho |= G]]$]
-%   \label{lemma:renamingval}
-%   If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
-%   $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
-% \scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
-% \end{lemma}
-% \begin{proof}
-%   Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
-%   that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i   < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.
-
-%   By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
-% have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho } ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$.
-From $[[rho |= D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho } S]]$ and $[[rho ren xi i in S]]$.
-% \end{proof}
-
-% \begin{lemma}[Renaming for $[[G |= a : A]]$]
-%   \label{lemma:semrenaming}
-%   If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
-%   $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
-% \end{lemma}
-% \begin{proof}
-%   Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
-% \end{proof}
 
@@
   \label{lemma:stpi}
-  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
-  A B : Set i]]$.
+  If $[[G |= A : Set i]]$ and $[[G ++ x  : A |= B : Set i]]$, then $[[G |= Pi
+  x : A . B : Set i]]$.
 \end{lemma}
@@
   conclusion, it now suffices to show that given $[[rho |= G]]$, there
-  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
+  exists some $[[S]]$ such that $[[InterpR i Pi x : A{rho} . B{rho .:
+    x -> x} S]]$.
   From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
   there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
-From $[[G ++ A |= B : Set i]]$, we know that there must
-exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
+From $[[G ++ x : A |= B : Set i]]$, we know that there must
+exists $[[S]]$ such that $[[InterpR i B {rho .: x -> a} S]]$ for every $[[a
 in S0]]$. The conclusion immediately follows from the admissible \rref{I-PiAlt}.
@@
   \label{lemma:stabs}
-  If $[[G |= Pi A B : Set i]]$ and $[[G ++ A |= b : B]]$, then $[[G |=
-  \ A b : Pi A B]]$.
+  If $[[G |= Pi x : A . B : Set i]]$ and $[[G ++ x : A |= b : B]]$, then $[[G |=
+  \ x . b : Pi x : A . B]]$.
 \end{lemma}
 \begin{proof}
-  By unfolding the definition of $[[G |= \ A b : Pi A B]]$, we need to
+  By unfolding the definition of $[[G |= \ x . b : Pi x : A . B]]$, we need to
   show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
-  $[[S]]$ such that $[[InterpR i Pi A {rho} B {up rho} S]]$ and $[[\
-  A{rho} b{up rho} in S]]$.
-
-  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
+  $[[S]]$ such that $[[InterpR i Pi x :  A {rho} . B {rho .: x -> x}
+  S]]$ and $[[\ x .
+   b{rho .: x -> x} in S]]$.
+
+  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi x : A . B : Set
   i]]$, there exists some set $[[S]]$ such that
-  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
-  $[[\A{rho} b{up rho} in S
+  $[[InterpR i Pi x : A {rho} . B {rho .: x -> x} S]]$. It now suffices to show that
+  $[[\ x . b{rho .: x -> x} in S
   ]]$. By Lemma~\ref{lemma:piinvalt}, the alternative inversion
@@
   \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
-    {rho .: a}
+    {rho .: x -> a}
     S1 #) #)]]$
   \item $[[S = { b | forall a, (# a in S0 , forall
-      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
+      S1, (# InterpR i B {rho .: x -> a} S1,  b a in S1 #) #) }]]$
   \end{itemize}
-  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
+  To show that $[[\ x . b{rho .: x -> x} in S]]$, we need to prove
   that given $[[a in S0]]$,
-  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
+  $[[Interp I i B {rho .: x -> a} S1]]$, we have  $[[( \ x . b{rho .: x
+    -> x} )
   a in S1]]$.
   By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
-  under expansion. Since $[[( \A{rho} b{up rho} )
-  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
-  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
+  under expansion. Since $[[( \ x . b{rho .: x -> x} )
+  a => b {rho .: x -> a}]]$, it suffices to show that
+  $[[b {rho .: x -> a} in S1]]$, which is immediate from $[[G ++ x : A |= b :
   B]]$ and the fact that the logical relation is deterministic and
@@
   \label{lemma:stapp}
-  If $[[G |= b : Pi A B]]$ and $[[G |= a : A]]$, then $[[G |= b a : B {a}]]$.
+  If $[[G |= b : Pi x : A . B]]$ and $[[G |= a : A]]$, then $[[G |= b
+  a : B {a / x}]]$.
 \end{lemma}
@@
 $[[i]]$ and $[[S1]]$
-such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a} {rho}
-S1]]$, or equivalently, $[[InterpR i B {rho .: a {rho}} S1]]$ since
-$[[B {a}{rho}]] = [[B {rho .: a {rho}}]]$. By the premise $[[G |= b :
-Pi A B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
+such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a / x} {rho}
+S1]]$, or equivalently, $[[InterpR i B {rho .: x -> a {rho}} S1]]$ since
+$[[B {a / x}{rho}]] = [[B {rho .: x -> a {rho}}]]$. By the premise $[[G |= b :
+Pi x : A . B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
 there exists some $[[i]]$ and $[[S0]]$ such that:
@@
   \item $[[forall a0, (# a0 in S0 implies (# exists S1 , InterpR i B
-    {rho .: a0}
+    {rho .: x ->  a0}
     S1 #) #)]]$
   \item $[[forall a0, (# a0 in S0 , forall
-      S1, (# InterpR i B {rho .: a0} S1,  b {rho} a0 in S1 #) #)]]$
+      S1, (# InterpR i B {rho .: x -> a0} S1,  b {rho} a0 in S1 #) #)]]$
   \end{itemize}
@@
       \beta\text{-}\mathit{normal\ terms} &
-      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
+      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi x : f . f]]\
                     |\ [[f ~ f : f]]\\
-            & & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
+            & & |   & [[\ x . f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
     \end{array}
@@
   \label{lemma:extwn}
-  If $[[wn (a var i)]]$, then $[[wn a]]$.
+  If $[[wn (a x)]]$, then $[[wn a]]$.
 \end{lemma}
@@
   By induction over the length of the reduction sequence in $[[wn (a
-  var i)]]$. The conclusion follows from Lemmas~\ref{lemma:parantirenaming} and
+  x)]]$. The conclusion follows from Lemmas~\ref{lemma:parantirenaming} and
   \ref{lemma:parnenf}.
@@
   \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
-  A]]$. To derive $[[wn (Pi A B)]]$, it remains to show $[[wn B]]$. We
-  use the outermost induction hypothesis to show that $[[var 0]]$
-  semantically inhabits $[[A]]$, from which we derive $[[wn (B {var 0})]]$
+  A]]$. To derive $[[wn (Pi x : A . B)]]$, it remains to show $[[wn B]]$. We
+  use the outermost induction hypothesis to show that $[[x]]$
+  semantically inhabits $[[A]]$, from which we derive $[[wn (B {x / x})]]$
   and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
 \end{proof}
-
-% \begin{lemma}[Adequacy: $CR_1$ for type]
-%   \label{lemma:cr1ty}
-%   If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
-% \end{lemma}
-% \begin{proof}
-%   Immediately from \rref{I-Ne} and \rref{I-Red}.
-% \end{proof}
-
-% \begin{lemma}[Adequacy: $CR$ for interpreted sets]
-%   \label{lemma:crel}
-%   If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
-% \end{lemma}
-% \begin{proof}
-%   By induction over the derivation of $[[Interp I i A S]]$. The base
-%   cases are all immediate.
-
-%   In the function case ($[[Pi A B]]$), we use the induction
-%   hypothesis to show that variables, which are special cases of
-%   neutral terms, semantically inhabit the input type $[[A]]$. We apply the
-%   function to an arbitrary variable and we can then use
-%   Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.
-
-%   To show
-%   the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
-%   such that $[[wne b]]$,
-%   and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
-%   us a term that semantically inhabits $[[B {a}]]$. From the induction
-%   hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
-%   holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
-%   from the induction hypothesis, since every term that evaluates to
-%   some neutral form semantically inhabits $[[B {a}]]$. Done.
-% \end{proof}
-
-% \begin{lemma}[Adequacy: $CR_2$ for type]
-%   \label{lemma:cr2ty}
-%   If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
-% \end{lemma}
-% \begin{proof}
-%   By induction over the derivation of $[[Interp I i A S]]$. The
-%   function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
-% \end{proof}
-
-% \begin{lemma}[Adequacy: $CR_2$ for types (Rec)]
-%   \label{lemma:cr2tyrec}
-%   If $[[InterpR i A S]]$, then $[[wn A]]$.
-% \end{lemma}
-% \begin{proof}
-%   By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
-% \end{proof}
-
-% \begin{lemma}[Adequacy: $CR$ for terms (Rec)]
-%   \label{lemma:crelrec}
-%   If $[[InterpR i A S]]$, then $[[S]] \in CR$.
-% \end{lemma}
-% \begin{proof}
-%   After unfolding the definition of $[[InterpR i A S]]$, trivial by
-%   Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
-% \end{proof}
 
@@
 confluence when the lambda term contains type annotations. A simple
-counterexample is $[[\- B ((\- A a) {up 1} var 0)]]$; depending on
+counterexample is $[[\- y : B . ((\- x : A . a) y)]]$ where $[[y notin
+fv (\- x : A . a)]]$; depending on
 whether \rref{P-AbsEta} is performed on the whole term or
 \rref{P-AppAbs} is used on the inner $\beta$ redex, we end up with the
-terms $[[\- B a]]$ or $[[\- A a]]$, where $[[A]]$ and $[[B]]$ are not
+terms $[[\- x : B . a]]$ (after $\alpha$-conversion) or $[[\- x : A . a]]$, where $[[A]]$ and $[[B]]$ are not
 necessarily syntactically equal terms. \citet{choudhury:ddc} resolve
@@
 not satisfy even the weaker substitution property
-(Lemma~\ref{lemma:parsubst}). % That is, given a term $[[A]]$
-% and $a \leadsto^{lo} b$, we do not necessarily have $[[A {a}]]
-% \leadsto^{lo} [[A {b}]]$ since the variable might appear to the
-% right of some redex.
+(Lemma~\ref{lemma:parsubst}).
 With this alternative formulation, we would have to prove how the

### 2024-02-27 paper/paper.tex
note about name of identity substitution

@@
 \yl{new overview starts here}
-Here is an overview of our contribution.
+More concretely, our paper makes the following contributions. 
 \begin{itemize}
@@
 substitutions, which are mappings from de Bruijn indices to terms.  One
-example of a substitution is the function $\ottkw{id}_{tm}$, the identity
+example of a substitution is the function $\ottkw{id}_{tm}$,\scw{Can we just 
+call this \ottkw{id} now?} the identity
 substitution that maps each index to the corresponding term variable.  % The

### 2024-02-27 paper/paper.tex
comment about nominal

@@
 variable $[[x]]$ to itself, protecting it from being replaced.
+\scw{This substitution operation is pretty strange when it comes to the 
+BVC. What variables are we allowed to use when we get to a bound variable $x$ in this function? I'd prefer if we instead had a side condition that it must be 
+the case that $\rho(x) = x$ already, then we use BVC to pick such an $x$. But then we need to know that such an $x$ exists, and that isn't necessarily true.}
+\scw{Maybe we should handwave here, and note that we aren't covering all of the 
+details of the treatment of variable binding in the text.
+If readers want to understand that part, they should look at the Coq code, which uses de Bruijn indices anyways.
+Our goal is to convey the understanding of the Coq proof.
+}
 

### 2024-02-27 paper/paper.tex
compress typing figure

@@
 example of a substitution is the function $\ottkw{id}_{tm}$,\scw{Can we just 
-call this \ottkw{id} now?} the identity
+call this $\ottkw{id}$ now?} the identity
 substitution that maps each index to the corresponding term variable.  % The
@@
 \yl{Explain why iterated substitution is not good enough}
+\scw{resolved below, right?}
 
@@
 \drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
-  Eq, Refl, J, If, Bool, True, False, Void}
+ Void, Bool, True, False}
+\[ \drule[width=5in]{T-If}\]
+\[ \drule[width=3in]{T-Eq} \qquad \drule{T-Refl} \]
+\[ \drule[width=4in]{T-J} \]
 \end{minipage}

### 2024-02-27 paper/paper.tex
use = in statistics figure

@@
   want to (or can) reuse.}
+\scw{I'm not sure that this is the best place for this discussion.}
 Our development relies on off-the-shelf tools...
@@
     % Library (Autosubst 2)  & 491 & - \\
-    Syntactic typing (specification) &  68 & - & - \\
-    Renaming & 46 & -  & - \\
-    Untyped reduction & 349 & - & - \\
+    Syntactic typing (specification) &  68 & = & = \\
+    Renaming & 46 & =  & = \\
+    Untyped reduction & 349 & = & = \\
     Neutral and normal forms & N/A & 269 & N/A \\
@@
 syntactic and typing specification, they differ in only the categories
-that are relevant to the logical relation. We use - as a marker that
+that are relevant to the logical relation. We use = as a marker that
 the line count is the same.

### 2024-02-27 paper/paper.tex
make header of fig 10 shorter

@@
 %\newcommand\dheader[1]{\rotatebox{60}{{#1}}}
-\newcommand\header[1]{\rotatebox{60}{{#1}}}
+\newcommand\header[1]{\rotatebox{0}{{#1}}}
 
@@
     \hline
-      & \header{Universes} & \header{Inductives} & \header{Conversion}
-      & \header{Large Elim} & Main results \\
+%      & \header{Universes} & \header{Inductives} & \header{Conversion}
+%      & \header{Large Elim} & Main results \\
+      & \header{Univ} & \header{Ind} & \header{Conv}
+      & \header{Large E} & Main results \\
+
     \hline
@@
     \\
-  Universes: & Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
-  Inductives: & Identity types (Id), Natural numbers (Nat), W-types \\
-  Conversion:& Untyped (U), Typed (T),  Extensional (E) \\
-  Large Eliminations:& included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
+  \underline{Univ}erses: & Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
+  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), W-types \\
+  \underline{Conv}ersion:& Untyped (U), Typed (T),  Extensional (E) \\
+  \underline{Large E}liminations:& Included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
   \\
@@
 
+\scw{Should we add a column for whether their logical relation is unary or binary?} 
 Figure~\ref{fig:featurematrix} presents several mechanized proofs
@@
 Each of these proofs is significantly larger than than our development; but they
-also prove more results about different object languages.
+also prove more results about different object languages.\scw{Can we mention the 
+line counts in the text, when they are known?}
 The table provides a comparison between the various features of their
@@
 logical relation can be defined recursively for well-formed types using
-a size metric, which has been explored in~\citet{liu2023dependently}.
+a size metric, which has been used in~\citet{liu2023dependently}.
+Their development is around 8,000 lines of Coq nonblank, noncomment code.
 
@@
 from~\citet{decagda} to the decidability type checking of a
-bidirectional type system.
+bidirectional type system. \scw{How big are these proofs?}
 

### 2024-02-27 paper/paper.tex
more futzing with figure, we have room for more columns if we want
intro

@@
 \section{Introduction}
-In this paper, we present a \emph{short} and \emph{mechanized} proof of
+This paper presents a \emph{short} and \emph{mechanized} proof of
 logical consistency for \lang{}, a dependent type theory with a full
@@
 
-Our goal with this work is to demonstrate how the proof technique of
-\emph{syntactic logical relations} can be applied to reason about dependent
-type theories.  Logical relations are a powerful proof technique for reasoning
-about lambda calculi and their operational semantics. They have been used to
-show diverse properties such as strong
-normalization~\citep{girard1989proofs,
+Our goal with this work is to demonstrate the application of the proof
+technique of \emph{syntactic logical relations} to dependent type theories.
+Logical relations are a powerful proof technique, and have been used to show
+diverse properties such as strong normalization~\citep{girard1989proofs,
   geuvers1994short}, contextual equivalence~\citep{constable1986implementing},
-representation independence~\citep{pitts1998existential}, noninterference~\citep{bowman2015noninterference}, compiler correctness~\citep{benton2009biorthogonality,perconti2014compiler}, and the
-decidability of conversion algorithms~\citep{harper2005equivalence,Abel12,abel2013normalization}. \scw{% add citations?
-  emphasize that we are working
-with syntactic logical relations.}
-
-Tutorial material on logical relations~\citep{skorstengaard2019introduction,
-  harpertait, harperkripke, pierce2002types,
-  pierce2004advanced,harper2016practical} is primarily focused on systems
-with simple or polymorphic types. In that context, syntactic logical relations
-can be defined as simple recursive functions over the structure of types, or
-(in the case of recursive types) defined over the evaluation steps of the
-computation. However, neither of these techniques can be used to define a
-logical relation in the context of a predicative dependent type theory, so a
-young researcher might be excused for thinking that proofs that use logical
-relations are not applicable for such languages.
-
-But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs for dependent type theories}
-Recent authors have developed tour-de-force
+representation independence~\citep{pitts1998existential},
+noninterference~\citep{bowman2015noninterference}, compiler
+correctness~\citep{benton2009biorthogonality,perconti2014compiler}, and the
+decidability of conversion
+algorithms~\citep{harper2005equivalence,Abel12,abel2013normalization}.
+
+However, tutorial material on syntactic logical
+relations~\citep{skorstengaard2019introduction, harpertait, harperkripke,
+  pierce2002types, pierce2004advanced,harper2016practical} is primarily
+focused on systems with simple or polymorphic types. In that context,
+syntactic logical relations can be defined as simple recursive functions over
+the structure of types, or (in the case of recursive types) defined over the
+evaluation steps of the computation. Yet, neither of these techniques can be
+used to define a logical relation in the context of a predicative dependent
+type theory, so a novice researcher might be excused for thinking that proofs
+that use logical relations are not applicable for such languages.
+
+But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs
+  for dependent type theories} Recent authors have developed tour-de-force
 mechanizations for the metatheory of modern proof
@@
 reasoning, our development is accompanied by a short mechanized proof script,
-of less than 1000 lines of code, developed using the Coq/Rocq proof
+of less than 1000 lines of code, developed using the Coq proof
 assistant~\cite{coq}.
@@
 We have streamlined our proof through a number of means: the careful selection
-of the features that we include in the object type system and the results
-that we prove about it, in addition to the judicious use of automation.  Our
+of the features that we include in the object type system and the results that
+we prove about it, in addition to the judicious use of automation.  Our
 language is small, but includes enough to be illustrative. For example, we
 eschew inductive datatypes or W-types, but we do include propositional
-equality and booleans to capture the challenges presented by indexed
-types and dependent pattern matching. We do not show the decidability of type
-checking, nor do we develop a PER semantics, but we prove logical consistency, which states that empty types are not inhabited in an empty context, and demonstrate how our
-consistency proof can be extended (at a moderate cost of 500 lines of code) to
-show the existence of $\beta\eta$-normal forms for well-typed open \emph{and}
-closed terms. We include a full predicative universe hierarchy and
-type-level computation to demonstrate the logical strength of the
-approach. % \scw{Say that we don't have impredicativity, but neither does anyone
-  % else?}\yl{Nobody else but Barras. I'll find the citation}
+equality and booleans to capture the challenges presented by indexed types and
+dependent pattern matching. We do not show the decidability of type checking,
+nor do we develop a PER semantics, but we prove logical consistency, which
+states that empty types are not inhabited in an empty context, and demonstrate
+how our consistency proof can be extended (at a moderate cost of 500 lines of
+code) to show the existence of $\beta\eta$-normal forms for well-typed open
+\emph{and} closed terms. We include a full predicative universe hierarchy and
+type-level computation to demonstrate the logical strength of the approach.
+
 %%
@@
 
-The result of our work is a development that an interested researcher can
+More concretely, our paper makes the following contributions. 
+\begin{itemize}
+\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
+  theory of interest. A key design choice that impacts our proofs is the use
+  of an untyped conversion rule, inspired by Pure Type
+  Systems~\citep{barendregt1991introduction}, and specified through parallel
+  reduction~\citep{takahashi-parallel-reduction,
+    barendregt:lambda-calculi-with-types}.
+\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
+  \lang{}, the property of interest, to motivate a logical relation. We define
+  the relation first inductively and then later prove that it is a partial
+  function. Based on this definition, we define semantic typing and prove the
+  fundamental theorem, from which consistency follows as a corollary
+  (Section~\ref{sec:logrelproof}). Thanks to the design of \lang{}, our proof
+  showcases the special treatment required to model many of the most common
+  features of dependent type theories, thus making our proof applicable to a
+  broad range of type systems.
+\item We strengthen our logical relation to prove the existence of
+  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
+  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
+  modifications made to our initial logical relation are small and closely
+  mirror the necessary extensions for a simply-typed language. We use this
+  part to show that once we have established the base techniques, we can port
+  ideas from proofs about simpler languages to the dependently typed setting.
+\item We mechanize all our proofs using the Coq proof assistant, with 963
+  lines of code for the consistency proof and a moderate increase to 1357
+  lines of code for the normalization proof.  We discuss the specifics related
+  to our choice of Coq as our metatheory, including our use of off-the-shelf
+  semantic engineering infrastructure and automation tools, in
+  Section~\ref{sec:logrelmech}.  Our proof scripts, with comments, are
+  available to reviewers as supplementary materials.
+%  We use Coq's custom notations to ease the
+%  effort of matching the pen and paper representation to our formalism.
+  \yl{I want to say something about the benefits of using existing libraries
+    such as Autosubst 2 and CoqHammer rather than rolling your own
+    undocumented libraries that nobody want to (or can) reuse.}\scw{Save that
+    for the mechanization section}
+% Our development relies on off-the-shelf tools...
+\item We compare our work to existing proofs by logical relations and other
+  proof techniques for proving consistency and normalization.  We provide an
+  overview of this prior work (Section~\ref{sec:relatedwork}) and also give an
+  in-depth explanation on how various design decisions affect the size of our
+  proof and its extensibility to additional features
+  (Section~\ref{sec:discuss}).
+\end{itemize}
+
+The result of our work is an artifact that an interested researcher can
 navigate and understand. We accompany this mechanized proof with an informal
-description, presented here using set-theoretic notation and terminology so that
-the material is accessible to readers with a general mathematical background.
-That said, our explanations do not stray too far away from our proof scripts.
-We link each lemma directly to its counterpart in the proof script,
-anticipating that readers may wish to reference and validate our formalism.
-The typeset proofs purposefully follow the
-mechanized proofs closely while avoiding, as much as possible, artifacts specific to Coq.
+description, presented here using set-theoretic notation and terminology so
+that the material is accessible to readers with a general mathematical
+background.  That said, our explanations do not stray too far away from our
+proof scripts.  We link each lemma directly to its counterpart in the proof
+script, anticipating that readers may wish to reference and validate our
+formalism.  The typeset proofs purposefully follow the mechanized proofs
+closely while avoiding, as much as possible, artifacts specific to Coq.
 
@@
 the strength of the logic in which they are expressed. An informal proof that
-attempts to be agnostic or ambiguous about the underlying
-metatheory % may struggle to even
-% convey the validity of their definition and
-requires substantial
-effort from the reader to even understand whether it is theoretically
-possible to encode such proofs in a given ambient logic, such as those found
-in modern proof assistants.
-
-% The close correspondence between a typeset proof and a mechanized
-% proof comes at the potential risk that the artifacts specific to
-% theorem provers may become visible in the presented proof.
-%However, for proofs by logical relation for
-%dependent types, we find our approach advantageous for a few reasons.
-Furthermore, while a proof assistant may reject perfectly valid definitions,
-it helps us keep our definition precise and unambiguous. A precise definition
-is crucial, as the reasoning principles we are allowed to use are directly tied
-to the definition, which is sometimes obscured by concise notations that
-diverge too far from a formal proof.
+attempts to be agnostic or ambiguous about the underlying metatheory requires
+substantial effort from the reader to even understand whether it is
+theoretically possible to encode such proofs in a given ambient logic, such as
+those found in modern proof assistants.
 
@@
 consistent with our mechanization.
-
-\yl{new overview starts here}
-More concretely, our paper makes the following contributions. 
-\begin{itemize}
-\item In Section~\ref{sec:spec},
-we present the type system of \lang{},
-a dependently typed system with an untyped conversion rule in the
-style of Pure Type Systems~\citep{barendregt1991introduction}. The
-\lang{} language supports many common features of a dependent
-type theory: a countable universe hierarchy, a boolean base
-type, an intensional identity type, and support for large elimination.
-\item In Section~\ref{sec:logreldep}, we formulate logical consistency for \lang{}, the property of
-interest, to motivate a logical relation, and proceed to define it as
-an inductive relation which we later prove to be a partial
-function. Based on the definition of the logical relation, we define
-a notion of semantic typing and prove the fundamental theorem, from
-which logical consistency follows as a corollary
-(Section~\ref{sec:logrelproof}). Thanks to the good coverage of
-features in our object language, our proof showcases the
-special treatment required to model those features, thus making
-our proof applicable to a broad range of type systems.
-\item We strengthen our logical relation to prove the existence of
-  $\beta$-normal forms (Section~\ref{sec:extension}) and
-  $\beta\eta$-normal forms (Section~\ref{sec:betaeta}) for well-typed
-  open \emph{and} closed terms. The modifications made to the logical
-  relation are small and closely mirror an analogous extension for a
-  simply typed language. We use these extensions to demonstrate the idea
-  that once we have established the base techniques, we can port ideas
-  that have been studied in the context of a simpler system to the
-  dependently typed setting.
-\item We mechanize all our proofs using the Coq proof assistant, with
-  963 lines of code for the consistency proof and a moderate increase
-  to 1357 lines of code for the normalization proof.
-  We discuss the specifics
-  related to our choice of Coq as our metatheory, including our use of
-  off-the-shelf semantic engineering infrastructure and automation
-  tools, in Section~\ref{sec:logrelmech}.
-  Our proof scripts, with comments, are available to reviewers as
-  supplementary materials. We use Coq's custom notations to ease the
-  effort of matching the pen and paper representation to our formalism.
- \yl{I want to say something
-  about the benefits of using existing libraries such as Autosubst 2
-  and CoqHammer rather than rolling your own undocumented libraries that nobody
-  want to (or can) reuse.}
-\scw{I'm not sure that this is the best place for this discussion.}
-Our development relies on off-the-shelf tools...
-\item We discuss in Section~\ref{sec:relatedwork} how our proof
-relates to existing proofs by logical relations and other proof
-techniques for proving consistency and normalization. In
-Section~\ref{sec:discuss}, we give an in-depth explanation on how
-various design decisions (e.g. the choice of untyped equality) affect
-the size of our proof and its extensibility to features such as
-type-directed conversion.
-\end{itemize}
-We hope our success at creating a short and mechanized proof for a
-relatively feature-complete language will encourage future researchers
-to leverage the tool of logical relations in mechanized
-proofs for dependent types.
-
-\yl{new overview ends here}
-
-The key contributions of each section of our paper are as follows.
-\scw{I'm still not happy with this overview / contributions list. }
-
-In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type theory of
-interest, reminiscent of Martin-Löf style predicative type systems. A key
-design choice that impacts our proofs is the use of an untyped conversion
-rule, inspired by Pure Type Systems~\citep{barendregt1991introduction}, and
-specified by parallel
-reduction~\citep{takahashi-parallel-reduction, barendregt:lambda-calculi-with-types} % to normal form~\citet{barendregt}
-.
-Because we believe that a good coverage of features makes our proof applicable
-to a broad range of type systems, \lang{} incorporates some of the most common
-features of dependent type theory, including an intensional identity type, a
-boolean type, large elimination, dependent elimination, and a full universe
-hierarchy. While our use of untyped conversion is nonstandard for MLTT, we
-discuss how our proof can be adjusted for systems with a typed, judgmental
-equality.
-
-In Section~\ref{sec:logreldep}, we formulate the consistency property and
-motivate where a proof by logical relations is required. The most important
-contributions of this section are the definition of the logical relation as an
-inductively defined relation and demonstrating how to reason about this
-definition. We also show that the logical relation respects evaluation
-(referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
-prove that the logical relation is a partial function: that is, each type can
-have at most one unique interpretation.
-
-In Section~\ref{sec:logrelproof}, we show that our definitions can be the
-subject of well-established reasoning techniques for logical relations. In
-particular, we define a notion of semantic typing in terms of our logical
-relation and prove the fundamental theorem, which states that syntactic typing
-implies semantic typing. Consistency follows from this result as a corollary.
-
-In Section~\ref{sec:extension}, we extend the definition of the logical relation
-to include terms with free variables, and prove that every well-typed term
-has a $\beta$-normal form. After laying the foundation for handling type-level
-computation in Section~\ref{sec:logreldep}, we show how this extension requires mechanical
-changes that directly mirror an analogous extension for a simply typed
-language. We use this section to demonstrate the idea that once we have
-established the base techniques, we can port ideas that have been studied in
-the context of a simpler system to the dependently typed setting.  We believe
-it is possible to extend our logical relation to a Kripke-style and relational
-model in a similar fashion.
-
-In Section~\ref{sec:betaeta}, we show how to incorporate the function $\eta$ law
-into the conversion rule for \lang{}.
-
-In Section~\ref{sec:logrelmech}, we discuss the details of our
-mechanization, including our use of existing libraries such as
-Autosubst 2~\citep{autosubst2} for handling bindings and
-CoqHammer~\citep{czajka2018hammer} for general-purpose automation.
-
-In Section~\ref{sec:relatedwork}, we discuss how our proof relates to
-existing proofs by logical relations and other proof techniques for
-proving consistency and normalization.
-
-In Section~\ref{sec:discuss}, we explain the various factors that contribute
-to our short proof.
-
-We hope our success at creating a short and mechanized proof for a
-relatively feature-complete language will encourage future researchers
-to leverage the tool of logical relations in mechanized
-proofs for dependent types.
-
-\scw{TODO: this introduction is still not done. At this point, I think the
-  first part has gotten a little long. We could say what we mean more
-  concisely. We also need to work on the overview at the end of this section
-  --- it need to be more descriptive about contributions and less of a table
-  of contents.}
-
-% % For
-% % example, \citet{geuvers1994short} requires impredicativity to encode
-% % Calculus of Constructions, an impredicative dependent type theory, but
-% % also relies on the classification property to define the logical
-% % relation over well-formed types and kinds. Their simple
-% % presentation of the logical relation as an inductively defined
-% % function over types, reminiscent of logical relations for simply typed
-% % languages,
 
@@
 \section{Logical Relation}
+\label{sec:logreldep}
+
 % \scw{This section pre-supposes that we want to define a logical relation,
@@
 \subsection{Definition of the Logical Relation}
-\label{sec:logreldep}
 \begin{figure}[h]
@@
 % \begin{minipage}{0.8\textwidth}
-  \begin{tabular}{| l |  l  | l | l | l | l| }
+  \begin{tabular}{| l |  c  | l | c | c | l | }
     \hline
@@
       & \header{Univ} & \header{Ind} & \header{Conv}
-      & \header{Large E} & Main results \\
+      & \header{L E} & Main results \\
 
@@
     $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & Consistency \\
+    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & Consistency \\
     NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & Correctness of NBE \\
-
-    % \hline
     $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} & Decidability of conversion \\
     % \hline
-    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
     % \hline
-    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & Consistency \\
+    MLTT-\'a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
+    % \hline
     \hline
@@
     \\
-  \underline{Univ}erses: & Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
+  \underline{Univ}erses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
   \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), W-types \\
   \underline{Conv}ersion:& Untyped (U), Typed (T),  Extensional (E) \\
-  \underline{Large E}liminations:& Included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
+  \underline{L}arge \underline{E}liminations:& Included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
   \\
   $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
+  Core Nuprl &\citet{anand2014towards} \\
+  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
   NBE-in-Coq& \citet{nbeincoq} \\
-  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
-  MLTT-a-la-Coq &\citet{martin-lof-a-la-coq} \\
-  Core Nuprl &\citet{anand2014towards} \\
+  MLTT-\'a-la-Coq &\citet{martin-lof-a-la-coq} \\
   \end{tabular}

### 2024-02-28 paper/paper.tex
a few more edits in the intro

@@
 \begin{abstract}
-Proof by logical relation is a powerful technique that has been used
+Proof by logical relations is a powerful technique that has been used
 to derive metatheoretic properties of type systems, such as
@@
 But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs
-  for dependent type theories} Recent authors have developed tour-de-force
+  for dependent type theories?} Recent authors have developed tour-de-force
 mechanizations for the metatheory of modern proof
@@
 400,000 lines of code. As a result, their use of logical relations are
-difficult to isolate from their surrounding contexts, and thus, not accessible
+difficult to isolate from their surrounding contexts and inaccessible
 to casual readers.
@@
 proof scripts.  We link each lemma directly to its counterpart in the proof
-script, anticipating that readers may wish to reference and validate our
-formalism.  The typeset proofs purposefully follow the mechanized proofs
-closely while avoiding, as much as possible, artifacts specific to Coq.
+script, anticipating that readers may wish to see how these results may be
+expressed and verified in a proof assistant.  The typeset proofs purposefully
+follow the mechanized proofs closely while avoiding, as much as possible,
+artifacts specific to Coq.
 

### 2024-02-28 paper/paper.tex
Revise de Bruijn-related text in Section 4 (to st-var)

@@
 $[[rho x]]$ is a term that inhabits all possible interpretations of the type
-$[[A {rho}]]$.
-
-Finally, despite the for all quantification over the interpreted set
-$[[S]]$ and the universe level $[[i]]$ in the definition of $[[rho |=
-G]]$, we only need show that there exists some $[[i]]$ and $[[S]]$
-such that $[[InterpR i A { rho } S]]$ and $[[rho x
-in S]]$ since the logical relation is a partial function (Lemma~\ref{lemma:logreldeter}). This idea is encapsulated in the second of the two structural
-law about $[[rho |= G]]$.
+$[[A {rho}]]$. The $\forall$ quantifier in its definition might look excessive since we
+know from Lemma~\ref{lemma:logreldeter} that each type can have at most
+one interpretation. However, since $[[rho |= G]]$ mostly appears in
+the position of a hypothesis, the $\forall$ statement is easy to
+instantiate and makes our proofs slightly easier. The few cases where we need to prove $[[rho |= G]]$ are
+handled by the following two structural properties, the second of
+which depends on Lemma~\ref{lemma:logreldeter}.
+% Finally, despite the for all quantification over the interpreted set
+% $[[S]]$ and the universe level $[[i]]$ in the definition of $[[rho |= G]]$, we only need show that there exists some $[[i]]$ and $[[S]]$
+% such that $[[InterpR i A { rho } S]]$ and $[[rho x in S]]$ since the logical relation is a partial function (Lemma~\ref{lemma:logreldeter}).
+This idea is encapsulated in the second of the two structural
+% law about $[[rho |= G]]$.
 \begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
@@
 \begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
-Define $[[|= G]]$ inductively as follows.
-\drules[SCtx]{$[[|= G]]$}{Semantic context well-formedness}{Empty, Cons}
-% \[
-%    \forall x : A in G,\text{ there exists some } i \text{ such that } drop Suc n G |= G n : Set i \]
+Define $[[|= G]]$ as follows.
+\[ \forall [[x : A in G]],\text{ there exists some } i \text{ such that } [[G |= A : Set i]] \]
 \end{definition}
@@
   \end{itemize}
-  \yl{todo}
-%   From the definition of $|= G$, we know that there exists some
-%   $i$ such that $drop Suc n G |= G n : Set i$. This is not
-%   immediately useful since $rho |= G$, not the truncated
-%   context. We claim, without giving the proof, that from $drop Suc n
-%   G |= G n  : Set i$, we can derive $G |= G n {up Suc n} : Set
-%   i$. A rigorous justification would require us to formally state
-%   and show renaming/weakening lemma for semantic typing. We instead
-%   provide an informal justification since the proof of renaming
-%   is uninteresting and only requires unfolding definitions, but its
-%   statement introduces unnecessary line noise related to shifting.
-
-%   Intuitively, a substitution $rho$
-%   that closes over a context should always induce a truncated
-%   substitution for each truncated context. As a result, if
-%   some term $b$ is semantically well-typed under some truncated
-%   context $G$, then it must also be semantically well-typed under
-%   the context $G$ up to shifting.
-% %
-%   Knowing that $G |= G n {up Suc n} : Set i$ holds, we can then
-%   apply Lemma~\ref{lemma:setinv} to show that
-%   $InterpR i G n < up Suc n > { rho } S$. It now
-%   suffices to show $rho n in S$, but that is immediate from the
-%   definition of $rho |= G$.
+  By the definition of semantic context well-formedness, we deduce
+  from $[[|= G]]$ and $[[x : A in G]]$ that there exists some universe
+  level $[[i]]$ such that $[[G |= A : Set i]]$. By
+  the equivalence from Lemma~\ref{lemma:setinv}, there exists $[[S]]$
+  such that $[[InterpR i A {rho} S]]$. However, by the definition of
+  $[[rho |= G]]$, we know that $[[rho x in S]]$, which is exactly what
+  we need for the conclusion.
 \end{proof}

### 2024-02-28 paper/paper.tex
add freshness condition to Ctx-Cons. remove fig 2. adjust description of simultanous substitution. more description of typing rules.

--- 2024-02-28 paper/paper.tex
@@
 
-In this section, we present the dependent type theory whose logical
-consistency will be proven in Section~\ref{sec:logrelproof}. For conciseness,
-we refer to this system as \lang.
+In this section, we present the dependent type theory, \lang{}, whose logical
+consistency will be proven in Section~\ref{sec:logrelproof}.
 
@@
 type theory, terms and types are collapsed into the same syntactic
-category. The type $[[Set i]]$ represent universe types.%  and $[[x]]$
+category. The type $[[Set i]]$ represent universe types, annotated by 
+the universe level, a natural number $i$.%  and $[[x]]$
 % represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
@@
 % conditions.
-\yl{need to fix the text}
-As a result, variables do not appear in binding forms, such as in
-abstractions $[[ \ x . a ]]$ and dependent function types $[[Pi x : A . B]]$.  This latter
-form binds a new variable of type $[[A]]$ that can appear in the codomain of
-the type $[[B]]$.  We use the notation $[[A -> B]]$ when the output type
-$[[B]]$ is not dependent on the input variable.
-We include in \lang{} the intensional identity type
-$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
-$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
-$[[b]]$, and $[[c]]$ is the term whose type is to be casted.
-Finally, \lang{} includes booleans, with standard syntax.
-\scw{Example of
- what you can use identity types for? Or an example of a program that uses
-J?}
+Abstractions $[[ \ x . a ]]$ and dependent function types $[[Pi x : A . B]]$
+are binding forms for the variable $x$ in the body of the function and
+codomain of the function type.  \footnote{In the exposition in this paper, we
+  adopt the Barendregt Variable Convention and assume that binding forms are
+  equal up to $\alpha$-conversion.  In some places we will be deliberately
+  informal about the treatment of variables; our mechanized proofs use de
+  Bruijn representation.}  We use the notation $[[A -> B]]$ when the output
+type $[[B]]$ is not dependent on the input variable.  We include in \lang{}
+the intensional identity type $[[a ~ b : A]]$ whose proofs can be eliminated
+by the J-eliminator $[[J c a b p]]$, where $[[p]]$ is an equality proof
+between $[[a]]$ and $[[b]]$, and $[[c]]$ is the term whose type is to be
+casted.  Finally, \lang{} includes booleans, with standard syntax.
+\scw{Example of what you can use identity types for? Or an example of a
+  program that uses J?}
 
@@
 % of a few representative cases of substitution.
-\begin{figure}[ht]
-  \begin{equation*}
-    \begin{split}
-      \begin{array}{lll}
-        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
-        [[idtm x]] & := & [[x]] \\ \\
-        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho
-        .: x -> a]]) } \\
-        ([[rho .: x -> a]])([[x]]) & := & [[a]] \\
-        ([[rho .: x -> a]])([[y]]) & := & [[rho y]] \text { if  } y
-                                          \neq x\\ \\
-      \end{array}
-    \end{split}
-    \qquad \qquad
-    \begin{split}
-      \begin{array}{lll}
-        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
-        [[x { rho }  ]] & := &  [[rho x]] \\
-        [[(Pi x : A . B) { rho }]] & := & [[Pi x : A { rho } . B { rho .: x -> x }]] \\
-        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
-        [[(\ x . a) { rho }]] & := & [[\ x . (a { rho .: x -> x })]] \\
-        \ldots
-      \end{array}
-    \end{split}
-  \end{equation*}
-  \caption{Auxiliary Functions over Syntax}
-  \label{fig:auxdef}
-\end{figure}
+%\begin{figure}[ht]
+%  \begin{equation*}
+%    \begin{split}
+%      \begin{array}{lll}
+%        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
+%        [[idtm x]] & := & [[x]] \\ \\
+%        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: x -> a]]) } \\
+%        ([[rho .: x -> a]])([[x]]) & := & [[a]] \\
+%        ([[rho .: x -> a]])([[y]]) & := & [[rho y]] \text { if  } y
+%                                          \neq x\\ \\
+%      \end{array}
+%    \end{split}
+%    \qquad \qquad
+%    \begin{split}
+%      \begin{array}{lll}
+%        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
+%        [[x { rho }  ]] & := &  [[rho x]] \\
+%        [[(Pi x : A . B) { rho }]] & := & [[Pi x : A { rho } . B { rho .: x -> x }]] \\
+%        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
+%        [[(\ x . a) { rho }]] & := & [[\ x . (a { rho .: x -> x })]] \\
+%        \ldots
+%      \end{array}
+%    \end{split}
+%  \end{equation*}
+%  \caption{Auxiliary Functions over Syntax}
+%  \label{fig:auxdef}
+% \end{figure}
 
@@
 
-\yl{Need to fix the de bruijn text}
-We adapt from \citet{autosubst2} notation for simultaneous renaming,
-substitution, and other auxiliary definitions used in our paper, summarized in
-Figure~\ref{fig:auxdef}. We need these definitions to specify the reduction
-and typing relations for \lang{}. The metavariable $[[rho]]$ represents
-substitutions, which are mappings from de Bruijn indices to terms.  One
-example of a substitution is the function $\ottkw{id}_{tm}$,\scw{Can we just 
-call this $\ottkw{id}$ now?} the identity
-substitution that maps each index to the corresponding term variable.  % The
-% $...$ substitution is more general and adds $[[n]]$ to the index
-% before injecting it as a term.
-The extension operation, $[[(rho .: x -> a)]]$,
-allows us to update the substitution $[[rho]]$ to map the variable
+% We adapt from \citet{autosubst2} notation for simultaneous renaming,
+% substitution, and other auxiliary definitions used in our paper, summarized in
+% Figure~\ref{fig:auxdef}. 
+Our reduction a typing relations are defined in terms of \emph{simultaneous
+  substitutions}, $[[rho]]$, which are mappings from variables to terms.  We
+use $[[idtm]]$ as the identity substitution.  The extension operation,
+$[[(rho .: x -> a)]]$, updates the substitution $[[rho]]$ to map the variable
 $[[x]]$ to $[[a]]$ rather than $[[rho x]]$.
 
-The substitution operator, which takes the form $[[a {rho}]]$,
-traverses the syntax of $[[a]]$ and replaces each variable $[[x]]$
-with the term $[[rho x]]$. When traversing under binders (e.g. in the
-$[[(\ x . a) { rho }]]$ case), we extend $[[rho]]$ to map the bound
-variable $[[x]]$ to itself, protecting it from being replaced.
-\scw{This substitution operation is pretty strange when it comes to the 
-BVC. What variables are we allowed to use when we get to a bound variable $x$ in this function? I'd prefer if we instead had a side condition that it must be 
-the case that $\rho(x) = x$ already, then we use BVC to pick such an $x$. But then we need to know that such an $x$ exists, and that isn't necessarily true.}
-\scw{Maybe we should handwave here, and note that we aren't covering all of the 
-details of the treatment of variable binding in the text.
-If readers want to understand that part, they should look at the Coq code, which uses de Bruijn indices anyways.
-Our goal is to convey the understanding of the Coq proof.
-}
+The substitution operator, which takes the form $[[a {rho}]]$, traverses the
+syntax of $[[a]]$ and replaces each variable $[[x]]$ with the term
+$[[rho x]]$. When traversing under binders (e.g. in the
+$[[(\ x . a) { rho }]]$ case), it must be the case that $[[rho]]$ maps the
+bound variable to itself and that the bound variable does not appear freely in
+the application of the substitution to any other variable.
+%
+\scw{Maybe we should handwave more here, and note that we aren't covering all
+  of the details of the treatment of variable binding in the text. If readers
+  want to understand that part, they should look at the Coq code, which uses
+  de Bruijn indices anyways. Our goal is to convey the understanding of the
+  Coq proof. }
 
@@
 
-\yl{Explain why iterated substitution is not good enough}
-\scw{resolved below, right?}
-
-In our proofs, we find it
-more convenient to formulate simultaneous substitution directly rather than
-recovering it from single substitution.  In particular, this shows up in the
-definition of semantic typing in Section~\ref{sec:logrelproof}, which relies
-on simultaneous substitution.
+When reasoning about logical relations, we find it more convenient to
+formulate simultaneous substitution directly rather than recovering it from
+single substitution.  In particular, this shows up in the definition of
+semantic typing in Section~\ref{sec:logrelproof}, which relies on simultaneous
+substitution.
 
@@
 terms to type check and have the same type. The use of an untyped relation for
-type conversion is reminiscent of Barendregt's Pure Type
-Systems~\citet{barendregt1991introduction} and makes our formulation different
-from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental
-equality takes the form $\Gamma \vdash a \equiv b : A$, from which one can
-usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
-reduction. By working with an untyped judgement, we can establish its
+conversion is similar to Barendregt's Pure Type
+Systems~\cite{barendregt1991introduction} and differs
+from MLTT~\citep{Martin-Lof-1973}, where the judgmental
+equality takes the form $\Gamma \vdash a \equiv b : A$.
+% from which one can
+% usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
+% reduction. 
+By working with an untyped judgement, we can establish its
 properties independently from the type system and the logical relation, using
 well-established syntactic approaches.
-
-\citet{siles2012pure} shows the equivalence of Barendregt's Pure Type
-System~\citep{barendregt1991introduction}, which employs untyped equality, and
+%
+\citet{siles2012pure} show the equivalence of Barendregt's Pure Type
+System, which employs untyped equality, and
 its variant that uses typed judgmental equality. This assures us that we do
@@
  $\eta$-equivalence of functions in Section~\ref{XXX} }
-Finally, we compare this
+We compare this
 definition with type-directed approaches to equality in
@@
 \begin{minipage}{0.9\textwidth}
-\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
+\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{}
+\[ \drule{Ctx-Empty} \qquad \drule[width=3in]{Ctx-Cons} \]
 \drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
@@
 
-\yl{Var, T-Pi, T-Abs are self-explanatory now?}
+\yl{Var, T-Pi, T-Abs are self-explanatory now?}  \scw{Need to actually mention
+  these rules} The typing rules of \lang{} are standard rules for dependent
+type theories.  The variable rule, \rref{T-Var}, uses the auxiliary relation
+$[[x : A in G]]$, that holds when a variable declaration is found in the
+typing context.  The typing of universes ensures that each one belongs to the
+next higher level.  \Rref{T-Pi} ensures predicative quantification by
+requiring that all parts of the type be typeable at the same universe
+level. \Rref{T-Abs} ensures that all functions have well-formed dependent
+types. In an application (\rref{T-App}) the argument is substituted for the
+variable in the result type.
+
 % In \rref{T-Var}, $G n$ is the partial function defined by the equations
@@
 The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
-matching.  The result type of this expression $[[A {a / x}]]$ is composed of some
+matching.  The result type of this expression, $[[A {a / x}]]$, is composed of some
 \emph{motive} $[[A]]$, a type where its single free variable has been replaced
@@
 variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
-witnessing the information gained through dependent pattern matching.  In the
-typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
-for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
-therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
-shifted to refer to the same binders.
+witnessing the information gained through dependent pattern matching. 
+% In the
+% typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
+% for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
+% therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
+% shifted to refer to the same binders.
 

### 2024-02-28 paper/paper.tex
Remove the x -> x substitutions in Sec 4

@@
 \end{definition}
-
-% In this definition, we use the notation $drop n G$ to denote
-% the typing context obtained by dropping the last $n$ elements of
-% $G$. When $G$ has fewer than $i$ elements, $drop n G$ returns
-% the empty list. Intuitively, $|= G$ says that all
-% types in the context must correspond to some set according to the logical
-% relation. The truncation to the context $G$ is due to the telescopic
-% scoping of the types in the context. An alternative approach would be to define $|= G$
-% inductively, similar to $|- G$, but we find it easier to use this definition and
-% recover the structural rules as lemmas.
-% \begin{lemma}[Semantic context well-formedness empty]
-%   \label{lemma:semwffempty}
-%   $|= G$ holds when $G$ is empty.
-% \end{lemma}
-% \begin{lemma}[Semantic context well-formedness cons]
-%   \label{lemma:semwffcons}
-%   If $|= G$ and $G |= A : Set i$, then $|= G ++ A$.
-% \end{lemma}
+Recall that $[[|- G]]$ is defined inductively in terms of the
+syntactic typing judgment. We take a different approach here with its
+semantic counterpart $[[|= G]]$. The definition of
+$[[|= G]]$ is not telescopic: with $[[|- G]]$, a variable appearing
+earlier in the context is well-scoped under a truncated context,
+whereas with $[[|= G]]$, the types are only required to be
+semantically well-formed with respect to the full context, regardless
+of their position in $[[G]]$. Our definition of $[[|= G]]$ could be
+strengthened, though the simpler definition is sufficient for showing the
+fundamental lemma.
+
+We can recover the structural rules for $[[|= G]]$ as lemmas.
+\begin{lemma}[Semantic context well-formedness empty]
+  \label{lemma:semwffempty}
+  $[[|= G]]$ holds when $[[G]]$ is empty.
+\end{lemma}
+\begin{lemma}[Semantic context well-formedness cons]
+  \label{lemma:semwffcons}
+  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ x : A]]$.
+\end{lemma}
 
@@
   conclusion, it now suffices to show that given $[[rho |= G]]$, there
-  exists some $[[S]]$ such that $[[InterpR i Pi x : A{rho} . B{rho .:
-    x -> x} S]]$.
+  exists some $[[S]]$ such that $[[InterpR i (Pi x : A . B){rho} S]]$.
   From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
@@
   show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
-  $[[S]]$ such that $[[InterpR i Pi x :  A {rho} . B {rho .: x -> x}
-  S]]$ and $[[\ x .
-   b{rho .: x -> x} in S]]$.
+  $[[S]]$ such that $[[InterpR i (Pi x : A . B){rho}
+  S]]$ and $[[(\ x . b) {rho} in S]]$.
 
@@
   i]]$, there exists some set $[[S]]$ such that
-  $[[InterpR i Pi x : A {rho} . B {rho .: x -> x} S]]$. It now suffices to show that
-  $[[\ x . b{rho .: x -> x} in S
+  $[[InterpR i (Pi x : A . B){rho} S]]$. It now suffices to show that
+  $[[(\ x . b){rho} in S
   ]]$. By Lemma~\ref{lemma:piinvalt}, the alternative inversion
@@
   \end{itemize}
-  To show that $[[\ x . b{rho .: x -> x} in S]]$, we need to prove
+  To show that $[[(\ x . b){rho} in S]]$, we need to prove
   that given $[[a in S0]]$,
-  $[[Interp I i B {rho .: x -> a} S1]]$, we have  $[[( \ x . b{rho .: x
-    -> x} )
+  $[[Interp I i B {rho .: x -> a} S1]]$, we have  $[[(\x . b ){rho}
   a in S1]]$.
   By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
-  under expansion. Since $[[( \ x . b{rho .: x -> x} )
+  under expansion. Since $[[( \ x . b ){rho}
   a => b {rho .: x -> a}]]$, it suffices to show that
@@
     S1 #) #)]]$
-  \item $[[forall a0, (# a0 in S0 , forall
-      S1, (# InterpR i B {rho .: x -> a0} S1,  b {rho} a0 in S1 #) #)]]$
+  \item $[[forall a0, (# a0 in S0 implies (# forall
+      S1, (# InterpR i B {rho .: x -> a0} S1,  b {rho} a0 in S1 #) #) #)]]$
   \end{itemize}
@@
 not normalizing.
-\scw{I don't see this precondition in the figure!}
+% \scw{I don't see this precondition in the figure!}\yl{resolved}
 Furthermore, the side condition $[[a <=> b]]$ is only

### 2024-02-28 paper/paper.tex
todo in sec 2
Comments

@@
 We can be specific and say that that the only reductions available for these terms are identity reductions and cite
-\footnote{\dotv{normalform.v}{nf\_refl}} }. Instead of the
+\footnote{\dotv{normalform.v}{nf\_refl}} }.\yl{I always think of
+normal form as the specific definition that says the relation can't
+step. We can refer to nfrefl but that requires some explanation about
+the definition of normal form and the complication that it doesn't
+hold in $\eta$ (maybe it's fine if we just don't mention it in the
+$\eta$ case)} Instead of the
 metavariables $e$ and $f$, we also
@@
 \scw{This lemma is only used to prove \texttt{wn\_antirenaming}, which is then
-used more generally. Maybe we should replace it with that?}
+used more generally. Maybe we should replace it with
+that?}\yl{Anti-renaming is mentioned once in Section 6 when we talk abou how confluence with $\eta$
+depends on anti-renaming for Par. I don't think that discussion in
+Section 6 is very important so I'm fine with replacing it  }
 \begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
@@
 contained within the interpretation.
-\scw{Also, cite where this form of proof originally
-appeared? This part is standard.}
 Therefore, we formally define when a set is a \emph{reducibility candidate}
@@
 
-\scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
-sentence elsewhere? I'd like to use the previous sentence as a conclusion for
-this section.}
-Finally, due to technical
-details, our mechanized proof about the existence of $\beta\eta$ form does not
-immediately induce a decision procedure for type conversion in
-\lang{}. We discuss this issue and its workaround in
-Section~\ref{sec:conversionalgo}.
+% \scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
+% sentence elsewhere? I'd like to use the previous sentence as a conclusion for
+% this section.}
+% Finally, due to technical
+% details, our mechanized proof about the existence of $\beta\eta$ form does not
+% immediately induce a decision procedure for type conversion in
+% \lang{}. We discuss this issue and its workaround in
+% Section~\ref{sec:conversionalgo}.
 \section{Mechanization}
@@
   to construct runtime relevant data.}.
-
 \section{Related Work}
@@
 using parallel reduction in \rref{I-Red} allows us to delay the
-factorization property until we need to extract a normalization
-algorithm from our fundamental theorem (Section~\ref{sec:conversionalgo}).
+factorization property until we need to justify the use leftmost-outermost
+reduction as a deterministic reduction strategy for normalization.
 

### 2024-02-28 paper/paper.tex
Revert I-Pi to use a function

@@
 \begin{proof}
-Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
-, S0 ) | a in S , Interp I i B { a /x } S0 }]]$.
+The precondition $[[forall a, (# a in S
+implies (# exists S0 , Interp I i B { a / x } S0 #) #)]]$ from
+\rref{I-PiAlt} immediately induces a function $[[F in S -> PowerSet
+STm]]$ such that $[[forall a, (# a in S implies Interp I i B { a / x }
+F a #)]]$, which is exactly what we need to apply \rref{I-Pi}.
 \end{proof}
-Compared to the \rref{I-Pi}, which contains the mysterious relation
-$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
+\Rref{I-PiAlt} directly captures the meaning of
 a well-behaved dependent function type. The precondition of
@@
 when combined with the functionality of the logical relation, uniquely
-determine the relation $[[R in S * PowerSet STm]]$ to be the relation
+determine the function $[[F in S -> PowerSet STm]]$ to be the
+functional relation
 $[[{ (a , S0 ) | a in S , Interp I  i B { a /x } S0 }]]$. This result is
@@
 syntactic strict positivity constraint required in proof assistants,
-but is genuinely non-monotone when we rewrite the inductive definition
-as an endofunction over the domain of relations.
+but is genuinely non-monotone when we treat the inductive definition
+as the fixed point of an endofunction over the domain of relations.
 Intuitively, the failure of monotonicity stems from the fact
@@
 grows, whereas the
-relation $[[R]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ and
-prevents the set of witnesses from growing. While it might
+function $[[F]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ as
+$[[F a]]$ for each $[[a in S]]$,
+thus preventing the set of witnesses from growing. While it might
 be possible to restrict the domain with additional constraints such as
@@
 slight disadvantage of \rref{I-Pi} is that we need to construct the
-function $[[R]]$ each time we apply it, though this is mitigated by
+function $[[F]]$ each time we apply it, though this is mitigated by
 the immediate admissibility of \rref{I-PiAlt} and the alternative $\Pi$ inversion
@@
   \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp
-    I i Pi x : A . B S1]]$, then there exists $[[S]],[[R]]$ such that:
+    I i Pi x : A . B S1]]$, then there exists $[[S]],[[F]]$ such that:
     \begin{itemize}
     \item $[[Interp I i A S ]]$
-    \item $[[R in S * PowerSet STm]]$
-    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
-      i B { a /x } S0 #)]]$
-    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
+    \item $[[S1 = { b | forall a, (# a in S implies b a in F a #) }]]$
     \end{itemize}
@@
     x : A0 . B0 S1]]$. From the
-    induction hypothesis, there exists $[[S]]$ and $[[R]]$ such that :
+    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
     \begin{itemize}
     \item $[[Interp I i A0 S ]]$
-    \item $[[R in S * PowerSet STm]]$
-    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
-      i B0 { a / x } S0 #)]]$
-    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
+    \item $[[F in S -> PowerSet STm]]$
+    \item $[[forall a, (# a in S implies Interp I i B0 { a / x } F a #)]]$
+    \item $[[S1 = { b | forall a, (# a in S implies b a in F a #) }]]$
     \end{itemize}
@@
     $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
-    $[[R]]$ additionally satisfies the following properties:
+    $[[F]]$ additionally satisfies the following properties:
     \begin{itemize}
     \item $[[Interp I i A S ]]$
-    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I
-      i B { a /x } S0 #)]]$
+    \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
     \end{itemize}
@@
 improved inversion lemma for function types whose statement is free of
-the relation $[[R]]$, analogous to the admissible rule \rref{I-PiAlt}.
+the relation $[[F]]$, analogous to the admissible rule \rref{I-PiAlt}.
 \begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
@@
 % relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
+In Coq, we
 

### 2024-02-28 paper/paper.tex
pass through section 7

--- 2024-02-28 paper/paper.tex
@@
 consistency will be proven in Section~\ref{sec:logrelproof}.
+
+\scw{Need to acknowledge the lack of type annotation on abstractions.}
 
@@
 $\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes, parallel
-reduction can not take any real reduction steps and can only step into a term
+reduction cannot take any real reduction steps and can only step into a term
 itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
@@
 fundamental theorem.
-\begin{corollary}[Existence of $\beta$-normal form\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
+\begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
   \label{corollary:exbetanf}
@@
   \begin{minipage}{0.9\textwidth}
-  \begin{tabular}{ l |  c  | c | c }
+  \begin{tabular}{ |l |  c  | c | c| }
+    \hline
     & Consistency & Normalization & Syntactic metatheory \\
     \hline
-    % Library (Autosubst 2)  & 491 & - \\
-    Syntactic typing (specification) &  68 & = & = \\
-    Renaming & 46 & =  & = \\
-    Untyped reduction & 349 & = & = \\
-    Neutral and normal forms & N/A & 269 & N/A \\
-    Logical relation & 331 & 426 & N/A \\
-    Semantic typing and soundness & 169 & 199 & N/A \\
-    Syntactic soundness & N/A & N/A  & 660 \\
+    Syntactic typing         &  83 & =   & = \\
+    Untyped reduction        & 344 & =   & = \\
+    Neutral and normal forms & -   & 273 & - \\
+    Logical relation         & 338 & 430 & - \\
+    Semantic soundness       & 190 & 214 & - \\
+    Syntactic soundness      & -   & -   & 629 \\
     \hline
-    Total & 963 & 1357 & 1123 \\
+    Total                    & 955 & 1344 & 1056 \\
+    \hline
   \end{tabular}
   \end{minipage}
-  \caption{Statistics of the Coq Development}
+  \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
+left. The marker - indicates the file does not
+contribute to the total.}
   \label{fig:linecount}
@@
 
+To demonstrate the scale of our proof scripts, 
 Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
-code\footnote{calculated by the \texttt{cloc} tool, available from \url{https://github.com/AlDanial/cloc}.} for each file of our
-development, including the base consistency proof from
-Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the extension to
-normalization for open and closed terms from Section~\ref{sec:extension}.
-
-Because the normalization and the consistency development share the same
-syntactic and typing specification, they differ in only the categories
-that are relevant to the logical relation. We use = as a marker that
-the line count is the same.
-
-Autosubst 2 takes our syntax specification in higher-order abstract
-syntax and generates Coq syntax specification, renaming and
-substitution functions, and lemmas and tactics that allow reasoning
-about those functions. The auto-generated syntax file by Autosubst 2
-and library header files are omitted from the figure.
-
-Orthogonal to the development of our consistency and normalization
-proof, we have also proven the syntactic soundness of our system through subject
-reduction and progress. The syntactic soundness proof shares the
-definition of renaming with our semantic soundness proof, which is
-factored out as a separate file under the Renaming (common) category.
-\scw{Should add this line count as another column, for comparison. I've done so,
-but may need to update the numbers. }
-
-\subsection{Artifacts Specific to Coq}
-In this section, we discuss the Coq encoding of the definitions and proofs presented
-in Section~\ref{sec:logrelproof} and the artifacts of this encoding that are
-specific to Coq.
+code\footnote{calculated by the \texttt{cloc} tool, available from
+  \url{https://github.com/AlDanial/cloc}.} for each file of our development,
+including the base consistency proof from Section~\ref{sec:logreldep} and
+\ref{sec:logrelproof} and the extension to $\beta$-normalization for open and
+closed terms from Section~\ref{sec:extension}.  For comparison, we have also
+proven syntactic type soundness through subject
+reduction\footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
+progress\footnote{\dotv{syntactic\_soundness.v}{wt\_progress}}.
+
+The Autosubst 2 tool takes our 13 line syntax specification, written in
+higher-order abstract syntax, and generates the Coq syntax specification,
+renaming and substitution functions, and lemmas and tactics that allow
+reasoning about those functions. The auto-generated syntax file (291 LOC)
+and other Autosubst library files are are not included in the figure.
+
+\subsection{Axioms}
+Our Coq development assumes two axioms: functional extensionality 
+and propositional extensionality. The former is also required by 
+the Autosubst 2 libraries. Both axioms are known to be consistent 
+with Coq's metatheory.
+
+Furthermore, these axioms also bridge the gap between our mechanization 
+and our informal proofs.
+
+For example, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$ are
+equal, it suffices to show the extensional property that
+$\forall x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
+occasionally in our presented proofs.
+
+In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm
+  -> Prop}, a predicate over \lang{} terms.  In axiom-free Coq, predicates do
+not come with the extensionality property. Given two predicates $P$ and $Q$,
+we cannot conclude that $P = Q$ when given a proof of
+$\forall x, P(x) \iff Q(x)$. But this is exactly the statement of the 
+axiom of functional extensionality.
+
+\subsection{Encoding the logical relation in Coq}
+We next discuss specific details of the Coq encoding of the logical relation
+presented in Section~\ref{sec:logreldep}.
+
+First, this definition requires the use of Coq's impredicatve \texttt{Prop}
+sort. In particular, the relation $[[R]]$ in \rref{I-Pi} requires the use of
+impredicativity because $[[R]]$ must later be instantiated into something
+defined in terms of the logical relation itself (e.g. in
+Lemma~\ref{lemma:piintroalt}).
 
@@
 
-In our proof presented in Section~\ref{sec:logreldep}, the relation
-$[[R]]$ in \rref{I-Pi} requires Coq's impredicative \texttt{Prop} sort
-since $[[R]]$ can later be instantiated into something defined in
-terms of the logical relation itself (e.g. in
-Lemma~\ref{lemma:piintroalt}). \scw{I'm having trouble connecting the rule in
-the figure with rule \texttt{InterpExtFun} in the coq development. Would it make
-sense to include the definition of ProdSpace? What is the Coq type of R?}
-\yl{I think I might just inline the definition of ProdSpace and then
-  rename the variables from the Coq code to match the ones from the paper}
-
 In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
-\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S}
-has type \texttt {tm -> Prop}.  However, if desired, we could consistently
-replace the use of \texttt{Prop} with Coq's predicative sort
-\texttt{Type}. This variation of the definition could be used to define the
-interpretation for any \emph{finite} number of universes. The use of
+\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.
+
+However, if desired, we could consistently replace the use of \texttt{Prop}
+with Coq's predicative sort \texttt{Type} in the definition of
+$[[Interp I i A S]]$. This alternative definition could be part of
+the interpretation for any \emph{finite} number of universes. The use of
 \texttt{Type} becomes troublesome only when we attempt to define
@@
 Figure~\ref{fig:logrelrec} that recursively calls itself at smaller universe
-levels. Therefore, the one feature of \lang{} that truly requires impredicativity
-is its countable universe hierarchy. \scw{I'm not sure the last part of this sentence
-adds much: ``Although, lemmas such as the
-admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only simply
-our mechanization, but also give better intuition on what our logical relation
-means.'' Yes, we need impredicativity to define these lemmas, but without impredicativity
-our logical relation means something else.}
+levels. Therefore, the one feature of \lang{} that truly requires
+impredicativity is its countable universe hierarchy. \scw{I'm not sure the
+  last part of this sentence adds much: ``Although, lemmas such as the
+  admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only
+  simply our mechanization, but also give better intuition on what our logical
+  relation means.'' Yes, we need impredicativity to define these lemmas, but
+  without impredicativity our logical relation means something else.}
 
@@
 
-Finally, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$
-are equal, it suffices to show the extensional property that $\forall
-x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
-occasionally in our presented proofs.
-In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm ->
-Prop}, a predicate over \lang{} terms.
-In axiom-free Coq, predicates do not come with the extensionality
-property. Given two predicates $P$ and $Q$, we cannot conclude that
-$P = Q$ when given a proof of $\forall
-x, P(x) \iff Q(x)$. While
-predicate extensionality is not necessary in our development, we
-assume it as an axiom since it is known to be consistent with Coq's
-metatheory and helps bridge the gap between our mechanization and the
-proofs we present.
-
 % In Coq, there is a distinction between computable functions and
@@
 \label{sec:automation}
-Our Coq mechanization heavily uses automation, though instead of
-defining custom tactics, we rely mostly on off-the-shelf tools such as
+Our Coq mechanization heavily uses automation, supported by the tools
 Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.
 
-We use the Autosubst 2 framework to specify our syntax in HOAS and
-produce Coq syntax files in de Bruijn representation. Additionally,
-Autosubst 2 provides a powerful tactic \texttt{asimpl} that
-can be used to prove the equivalence of two terms constructed using
-the primitive operators provided by the framework. This greatly
-simplifies the reasoning about substitution in our development as
-almost all substitution related properties about the syntax are
-immediately discharged by \texttt{asimpl}.
+We use the Autosubst 2 framework to produce Coq syntax files based on a de
+Bruijn representation of variable binding and capture-avoiding substitution.
+In addition to these generated definitions, Autosubst 2 provides a powerful
+tactic \texttt{asimpl} that can be used to prove the equivalence of two terms
+constructed using the primitive operators provided by the framework. This
+tactic simplifies the reasoning about substitution as many
+substitution-related properties about syntax are immediately discharged by
+\texttt{asimpl}.
 
@@
 The automation provided by CoqHammer not only gives us a proof that is shorter
-and more resilient to changes, but also gives useful documentation for readers
-who wish to understand the mechanized proof. Although automation performs
-extensive search, it does not use lemmas or invert derivations that are not
-specified in the \texttt{use} or \texttt{inv} flag.
+and more resilient to changes, but also provides useful documentation for
+readers who wish to understand the mechanized proof. Although automation
+performs extensive search, it does not use lemmas or invert derivations that
+are not specified in the \texttt{use} or \texttt{inv} flags.
 
@@
 The soundness of the algorithm is immediate. The completeness of the
-algorithm is justified the confluence property of the untyped
+algorithm is justified by the confluence property of the untyped
 reduction relation. Suppose $[[a <=> b]]$, $[[a =>+ f0]]$, $[[b =>+
@@
 However, % with a little more effort,
-it is possible to recover a
+it should be possible to recover a
 precise algorithm. The first step is to define a deterministic

### 2024-02-28 paper/paper.tex
remove old comments

@@
 
-\scw{This section needs to say:
-  \begin{itemize}
-\item definitional equality is untyped so that we can prove
-  properties about it independent of the type system
-\item definitional equality is based on parallel reduction so that we can take
-  advantage of the algorithmic structure later. This definition is easier to
-  invert because there are fewer derivation. That is important when reasoning
-  about our logical relation.
-\item can prove that parallel reduction is equivalent to other definitions of
-  equality using standard means (cite Barendregt's book).  (we go halfway
-  there by showing it is an equivalence relation). Can also prove equivalent
-  to typed relation. (cite Siles, also CoreSpec?)
-\item All of the proofs in this section are standard, well known, and use
-  techniques that are well-suited for proof assistants. In fact, de Bruijn's
-  paper that introduces de Bruijn indices was part of a mechanized confluence
-  proof for the untyped lambda calculus.
-\item extensions: forward reference to eta equivalence for functions in later
-  section. type-directed equivalence is future work
-  \end{itemize}
-}
+% \scw{This section needs to say:
+%  \begin{itemize}
+% \item definitional equality is untyped so that we can prove
+%  properties about it independent of the type system
+% \item definitional equality is based on parallel reduction so that we can take
+%  advantage of the algorithmic structure later. This definition is easier to
+%  invert because there are fewer derivation. That is important when reasoning
+%  about our logical relation.
+% \item can prove that parallel reduction is equivalent to other definitions of
+%  equality using standard means (cite Barendregt's book).  (we go halfway
+%  there by showing it is an equivalence relation). Can also prove equivalent
+%  to typed relation. (cite Siles, also CoreSpec?)
+% \item All of the proofs in this section are standard, well known, and use
+%  techniques that are well-suited for proof assistants. In fact, de Bruijn's
+%  paper that introduces de Bruijn indices was part of a mechanized confluence
+%  proof for the untyped lambda calculus.
+% \item extensions: forward reference to eta equivalence for functions in later
+%  section. type-directed equivalence is future work
+%  \end{itemize}
+%}
 
@@
 
-% \scw{motivate parallel reduction. Why are you telling us this now. Is this
-%   an explanation of definitional equality and its properties?}
 Before we specify the typing rules, we first specify the equational theory
@@
 \citet{takahashi-parallel-reduction}. However, this choice simplifies later
-proofs, as we discuss in Section~\ref{sec:discuss}. % \scw{Forward reference to further
-  % discussion, comparing with whnf reduction?} \yl{resolved?}
+proofs, as we discuss in Section~\ref{sec:discuss}.
 
@@
 
-\yl{Var, T-Pi, T-Abs are self-explanatory now?}  \scw{Need to actually mention
-  these rules} The typing rules of \lang{} are standard rules for dependent
+The typing rules of \lang{} are standard for dependent
 type theories.  The variable rule, \rref{T-Var}, uses the auxiliary relation
@@
   \begin{itemize}
-  \item Dependent function types (result type mentions argument value, so can't define
-    relation by recursion over type structure)
+  \item Large eliminations
   \item Definitional equality (not all types look like types)
-  \item Identity types(?)
   \end{itemize}
@@
   identity types in the right way.
-}
-\yl{I don't think identity types and dependent function types should
-  be listed as reasons why the logrel can't be a recursive
-  function. System DE has both dependent function types and identity
-  types but the logical fragment doesn't have type-level computation
-  so it can be defined structurally by specifying some kind of size metric
 }
@@
 \end{corollary}
-% \scw{Need a transition here}
-
-Since the definition of our logical relation is an inductive relation,
+
+Because the definition of our logical relation is an inductive relation,
 it is not immediately obvious why each type $[[A]]$ can only uniquely
@@
 \end{proof}
-
-% The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
-% counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
-% than inductively over the context. \scw{why is that the case?} It is easy to recover the same
-% structural rules:
 

### 2024-02-28 paper/paper.tex
Add discussion about i-picoq

@@
 
-First, this definition requires the use of Coq's impredicatve \texttt{Prop}
-sort. In particular, the relation $[[R]]$ in \rref{I-Pi} requires the use of
-impredicativity because $[[R]]$ must later be instantiated into something
-defined in terms of the logical relation itself (e.g. in
-Lemma~\ref{lemma:piintroalt}).
+% First, this definition requires the use of Coq's impredicatve \texttt{Prop}
+% sort. In particular, the function $[[F]]$ in \rref{I-Pi} requires the use of
+% impredicativity because $[[F]]$ must later be instantiated into something
+% defined in terms of the logical relation itself (e.g. in
+% Lemma~\ref{lemma:piintroalt}).
+In the Coq mechanized proof, we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. The definition of $[[Interp I i A S]]$ has type
+\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.
 
@@
 % relation itself (e.g. in the proof of Lemma~\ref{lemma:piintroalt}).
-In Coq, we
-
-In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
-\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.
 
@@
 levels. Therefore, the one feature of \lang{} that truly requires
-impredicativity is its countable universe hierarchy. \scw{I'm not sure the
-  last part of this sentence adds much: ``Although, lemmas such as the
-  admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only
-  simply our mechanization, but also give better intuition on what our logical
-  relation means.'' Yes, we need impredicativity to define these lemmas, but
-  without impredicativity our logical relation means something else.}
+impredicativity is its countable universe hierarchy.
 
@@
 
+\Rref{I-PiCoq} shows how \rref{I-Pi} is actually encoded in our
+mechanized proof.
+\begin{center}
+  \drule[]{I-PiCoq}
+\end{center}
+Compared to \rref{I-Pi}, \rref{I-PiCoq} replaces the function
+$[[F]]$ with a total relation $[[R]]$. The equivalence of these two
+rules follows from the fact that the logical relation is a
+partial function (Lemma~\ref{lemma:logreldeter}). In set-theoretic
+notation, \rref{I-Pi} is more readable. However, if we want to encode
+the same rule in Coq, we must encode $[[F]]$ as a relation (with type
+\texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the following functionality constraint.
+\begin{minted}{coq}
+              forall a S0 S1, F a S0 -> F a S1 -> S0 = S1.
+\end{minted}
+In comparison, \rref{I-PiCoq} does not require this side condition and
+results in a simpler definition.
+
+Finally, we note that we cannot ascribe $[[F]]$ the type \texttt{tm -> (tm ->
+  Prop)} since Coq requires functions of such type to be
+computable. While defining $[[F]]$ as a computable Coq function rather than a
+functional relation does result in a concise encoding of \rref{I-Pi},
+we will have trouble instantiating $[[F]]$ with the logical relation,
+which is defined as a relation that we prove to be functional, rather
+than a computable function.
+
 % In Coq, there is a distinction between computable functions and
@@
 % \rref{I-Pi} as follows in our Coq mechanization.
-% \begin{center}
-%   \drule[]{I-PiCoq}
-% \end{center}
 % It should be easy to verify that the preconditions of

### 2024-02-28 paper/paper.tex
Minor fixes and motivation for Sec 6

@@
 fully mechanized the consistency proof using the Coq proof assistant
-in under 1000 lines of code, with 500 lines of additional code for the
+in under 1000 lines of code, with 400 lines of additional code for the
 $\beta\eta$-normal form extension.
@@
   We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
-  natural numbers as term variables. The side condition $[[idtm |=
+  variables as terms. The side condition $[[idtm |=
   G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
@@
 
-\yl{commented out the following paragraph because I accidentally wrote
-the same thing twice. Will do a comparison and if this is better
-written I'll replace the part from the $\beta\eta$ section with it}
 % Due to the non-deterministic nature of parallel reduction, we need to take a
@@
 proof. And include pointers on where readers might refer to it.}
-We can easily incorporate the function $\eta$ law to the equational
+\yl{Added the related work at the beginning but I wonder if we should
+  just say that we want to match our expressiveness with existing systems}
+
+\citet{nbeincoq,decagda,martin-lof-a-la-coq} all
+include the $\eta$ law for functions in their equational theory and
+use relational models to justify its validity.
+In our system, we can easily incorporate the function $\eta$ law to the equational
 theory of \lang{} by adding the following parallel reduction rule.
@@
 \end{center}
-% \Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
-% equational theory since convertibility ($[[a <=> b]]$), the untyped relation
-% used for type conversion, is built on top of parallel reduction.
-We can recover the same confluence result about parallel reduction using the
+In this section, we show how we easily extend the existence of $\beta$-normal
+forms from Section~\ref{sec:extension} to the existence of
+$\beta\eta$-normal forms after we add the function $\eta$ law to our
+equational theory.
+
+First, we recover the same confluence result about parallel reduction using the
 standard techniques from \citet{barendregt:lambda-calculi-with-types,
@@
 (Lemma~\ref{lemma:pardiamond}). Another complication is that
-confluence\scw{diamond property?} and anti-renaming are now proven through
+the anti-renaming property and the diamond property for parallel reduction are now proven through
 induction on a size metric of lambda terms; \rref{P-AbsEta} reduces a term
@@
 
-Note that, after this extension, the specification of our logical relation does not require
-any updates. As before, $\ottkw{ne}$ and $\ottkw{nf}$
+Note that, after this extension, the specification of our logical
+relation does not require
+any updates. The proof of the fundamental theorem also remains
+identical since the complications introduced by $\eta$ are hidden
+behind the proofs of the diamond property and the anti-renaming property.
+As before, $\ottkw{ne}$ and $\ottkw{nf}$
 represent $\beta$-neutral and $\beta-$normal forms, and the
@@
 of dependent function types and thus are closely related to our
-mechanization. \citet{anand2014towards} further shows it is possible
+mechanization. \citet{anand2014towards} further show it is possible
 to encode a finite universe hierarchy without the use of
-either impredicativity or induction-recursion.\scw{How did the Nuprl
-proof include a full hierarchy? Did they use impredicativity?}
+either impredicativity or induction-recursion. Their encoding of a countable
+universe hierarchy still relies on impredicativity, similar to our
+development.
+% \scw{How did the Nuprl
+% proof include a full hierarchy? Did they use impredicativity?}
 % \yl{commented out because I think this paragraph is too subjective}

### 2024-02-28 paper/paper.tex
Add line count

@@
 developments. However, because these proofs show diverse results about real
-systems and algorithms, these developments range in size from 20,000 to
-400,000 lines of code. As a result, their use of logical relations are
+systems and algorithms, these developments range in size from 10,000 to
+300,000 lines of code. As a result, their use of logical relations are
 difficult to isolate from their surrounding contexts and inaccessible
@@
 logical relation is defined using the induction-recursion scheme,
-which is available in Agda but not in Coq.
+which is available in Agda but not in Coq. Their development involves
+around 10,000 lines of Agda code.
 \citet{martin-lof-a-la-coq} transports the logical relation
@@
 from~\citet{decagda} to the decidability type checking of a
-bidirectional type system. \scw{How big are these proofs?}
+bidirectional type system. Their development is larger and has around
+30,000 lines of Coq code.
+\scw{How big are these proofs?}
 
 \citet{anand2014towards} mechanize the metatheory of
-Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
+Nuprl~\citep{constable1986implementing} in Coq. This metatheory is an
 extensional type theory with features such as dependent functions,
-inductive types, and a full universe hierarchy. \scw{What do they prove?}
+inductive types, partial types, and a full universe hierarchy. They construct a PER
+model in Coq to show the logical consistency of their language. Their
+development has been further extended with features such as
+intersection types, union types, and quotient types. The extensive
+coverage of features result in a Coq development with around
+330,000 lines of code.
 \citet{nbeincoq}
@@
 only base type from the object language, large elimination is not
-supported despite the one predicative universe.
+supported despite the one predicative universe. Their development has
+around 20,000 lines of Coq code.
 Both \citet{anand2014towards} and \citet{nbeincoq} leverage the

### 2024-02-28 paper/paper.tex
More comments

@@
 But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs
-  for dependent type theories?} Recent authors have developed tour-de-force
+  for dependent type theories?}\yl{There is a long paragraph in Sec 8 about how
+  pnp proofs like Geuvers' can't be easily mechanized, but I'm not sure if we
+  can also argue that they are hard to read, too (I personally think that's the
+  case, but I don't know if the reviewers will agree)} Recent authors have developed tour-de-force
 mechanizations for the metatheory of modern proof
@@
 %  effort of matching the pen and paper representation to our formalism.
-  \yl{I want to say something about the benefits of using existing libraries
-    such as Autosubst 2 and CoqHammer rather than rolling your own
-    undocumented libraries that nobody want to (or can) reuse.}\scw{Save that
-    for the mechanization section}
+  % \yl{I want to say something about the benefits of using existing libraries
+  %   such as Autosubst 2 and CoqHammer rather than rolling your own
+  %   undocumented libraries that nobody want to (or can) reuse.}\scw{Save that
+  %   for the mechanization section}
 % Our development relies on off-the-shelf tools...
@@
   Coq proof. }
+\yl{I agree. Should mention upfront that the presentation is not
+  watertight and rigorous, though we have one in our mechanization}
 
@@
 
-\scw{Should we add a column for whether their logical relation is unary or binary?}
+\scw{Should we add a column for whether their logical relation is
+  unary or binary?}\yl{Technically a Kripke-style model has the
+  context as an argument to the logrel, too. By binary, we are really
+  talking about whether the interpreted set is binary or not. I think
+  it is tricky to explain what we actually mean. }
 Figure~\ref{fig:featurematrix} presents several mechanized proofs
@@
 Each of these proofs is significantly larger than than our development; but they
-also prove more results about different object languages.\scw{Can we mention the
-line counts in the text, when they are known?}
+also prove more results about different object languages.% \scw{Can we mention the
+% line counts in the text, when they are known?}
 The table provides a comparison between the various features of their
@@
 30,000 lines of Coq code.
-\scw{How big are these proofs?}
+% \scw{How big are these proofs?}
 
@@
 proofs need Kripke style because they are defining typed relations, not untyped
-relations. }
+relations. } \yl{I wonder if
+we can just assume some more technical knowledge from the readers in
+this section.}
 However, our early dependence on confluence

### 2024-02-28 paper/paper.tex
change meta-and from , to /\ on Jon's suggestion

@@
 
-But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs
-  for dependent type theories?}\yl{There is a long paragraph in Sec 8 about how
-  pnp proofs like Geuvers' can't be easily mechanized, but I'm not sure if we
-  can also argue that they are hard to read, too (I personally think that's the
-  case, but I don't know if the reviewers will agree)} Recent authors have developed tour-de-force
+But this is not the case. 
+% \scw{Add a sentence about pen-and-paper LR proofs
+%  for dependent type theories?}\yl{There is a long paragraph in Sec 8 about how
+%  pnp proofs like Geuvers' can't be easily mechanized, but I'm not sure if we
+%  can also argue that they are hard to read, too (I personally think that's the
+%  case, but I don't know if the reviewers will agree)} 
+Recent authors have developed tour-de-force
 mechanizations for the metatheory of modern proof
@@
 systems and algorithms, these developments range in size from 10,000 to
-300,000 lines of code. As a result, their use of logical relations are
-difficult to isolate from their surrounding contexts and inaccessible
+300,000 lines of code. As a result, their uses of logical relations are
+difficult to isolate from the surrounding contexts and inaccessible
 to casual readers.
@@
   available to reviewers as supplementary materials.
-%  We use Coq's custom notations to ease the
-%  effort of matching the pen and paper representation to our formalism.
-  % \yl{I want to say something about the benefits of using existing libraries
-  %   such as Autosubst 2 and CoqHammer rather than rolling your own
-  %   undocumented libraries that nobody want to (or can) reuse.}\scw{Save that
-  %   for the mechanization section}
-% Our development relies on off-the-shelf tools...
 \item We compare our work to existing proofs by logical relations and other
@@
 The result of our work is an artifact that an interested researcher can
-navigate and understand. We accompany this mechanized proof with an informal
-description, presented here using set-theoretic notation and terminology so
-that the material is accessible to readers with a general mathematical
+navigate and understand. We accompany this short mechanized proof with an
+informal description, presented here using set-theoretic notation and
+terminology so that it is accessible to readers with a general mathematical
 background.  That said, our explanations do not stray too far away from our
@@
 
-Not only does this connection aid readers that wish to, like us, adopt proof
-assistants for their day-to-day use, we also find that this precision is
-important for conveying the proof technique itself. Unlike properties that are
-derivable through syntactic means, proofs by logical relation make demands on
-the strength of the logic in which they are expressed. An informal proof that
-attempts to be agnostic or ambiguous about the underlying metatheory requires
-substantial effort from the reader to even understand whether it is
-theoretically possible to encode such proofs in a given ambient logic, such as
-those found in modern proof assistants.
-
-Finally, our approach does not result in a verbose proof on paper, thanks to
-the already short mechanization. While we do make conservative simplifications
-to our typeset proof, such as avoiding the clunky syntax for well-founded
-recursion in Coq, we keep the overall structure of our typeset proofs
-consistent with our mechanization.
-
-% However, we make the
-% typeset definitions as close as possible to our mechanization so the
-% definitions and the proofs can be easily encoded in a proof
-% assistant.
-% We find the close correspondence
-% between our typeset presentation and the Coq development advantageous
-% for a couple of reasons. First, compared to other proof techniques,
-% such as the use of preservation and progress to derive syntactic
-% soundness, defining a logical relation for a dependently typed
-% language is demanding on the strength of the metatheory.
-
-% % Depending on its application, we care about certain metatheoretic
-% % properties about a type system. As a programming language, we may care
-% % about type soundness, which states that a well-typed never gets stuck
-% % during evaluation.
-% When a dependently type system is used as a program logic where terms
-% encode proofs, we want our type system to be logically consistent,
-% meaning that the empty type is not inhabited.
-
-% The consistency proofs of various dependently typed systems, including Martin-Lof's
-% type theory and the Calculus of Constructions, have long been
-% available in the literature. In particular, recent works such as \citet{nbeincoq},
-% \citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize the
-% correctness of the NbE algorithm, the decidability of type
-% conversion, decidability of type
-% checking respectively for dependently typed systems. From these
-% properties, consistency can be derived as a corollary.
-
-% The underlying technique of the forementioned
-% works is proof by logical relation,
-% which involves interpreting types as reducibility predicates,
-% representing sets of terms satisfying certain properties with respect
-% to the reduction relation. While the proof technique and the
-% consistency result for dependent types are both well-established,
-% there is a lack of rigorous and accessible material that shows
-% how proof by logical relation can be applied to dependently typed
-% systems.
-% \scw{You are choosing to focus on mechanized logical relations proofs for
-% dependent type theories, not consistency proofs for dependent type theories.
-% Why? (And your title shuld match}
-
-% Introductory materials about logical relations or standard textbooks such as
-% \citet{skorstengaard2019introduction}, \citet{harper2016practical},
-% and \citet{pierce2002types}
-% talk about logical relations for simply or polymorphically typed
-% languages, proving results such as termination for closed terms and parametricity.
-% \scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
-% Does Aspinall and Hofmann's chapter of ATTAPL include a logical
-% relation?}
-% \yl{skorstengaard is from the OPLSS notes. Cited ATTAPL}
-% \yl{Don't have a copy of }
-% \scw{Seems like you should also mention POPLmark reloaded somehere too?}
-% \citet{pierce2004advanced}, \citet{harperkripke}, and \citet{abel2019poplmark} show how
-% a Kripke-style logical relation can be used to include scoping
-% information in the logical relation and
-% derive properties such as the existence of normal form or strong
-% normalization for open and closed terms in simply typed languages.
-% Overall, the introductory texts about logical relations
-% cover systems and properties with varying degrees of complexity, from
-% simply typed to polymorphicly typed, logical predicate to logical
-% equivalence, closed terms to open terms.
-
-% The glaring gap here is the
-% lack of fully dependently typed systems where computations may appear
-% at the type-level. It is far from obvious why proof by
-% logical relation is even applicable to dependent types, since the type
-% may very well-be a computation that is yet to be evaluated.
-% While it is assuring that proof by logical relations for dependent
-% types is available in mechanized forms, % the key to address the
-% % complexities of dependent types is obscured in
-% \citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
-% Kripke-style models that obscure the technique for addressing
-% type-level computation. The added complexivity is evident
-% from the size of their code base. All three developments involve 20
-% thousand to 30 thousand lines of Coq or Agda code.
-
-% The goal of this paper is to give a tutorial on proof by logical relation for
-% dependent types in a simple and digestable format. \scw{This sentence should come as
-% early as possible.} Dependently typed systems comes in many flavors and
-% vary greatly in their expressiveness.
-% In Section~\ref{sec:spec}, we introduce \lang{}, a dependently typed
-% language that is small but relatively complete in features \scw{list these features explicitly:
-% large eliminations, indexed types, others? \ldots} related to
-% dependently types, including large elimination, a full universe
-% hierarchy, an intensional identity type and a boolean base type, both
-% of which are equipped with dependent elimination forms (e.g. J eliminator
-% for the identity type). In particular, the inclusion of identity
-% types, which is absent from \citet{decagda}, \citet{nbeincoq},
-% helps demonstrate how indexed types are handled when defining the
-% logical relation.
-
-% % We choose boolean
-% % types over natural numbers as our base type for simplicity, but include an
-% % intensional identity type in our type system to show how indexed
-% % types are treated in a logical relational proof. \scw{Talk about
-% % identity type first.} Unlike
-% % \citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
-% % \citet{anand2014towards}, we include an infinite hierarchy of
-% % universes to not only support type-level computations, but also avoid
-% % the unnecessary code duplication pointed out by \citet{nbeincoq} when
-% % big and small types are treated non-uniformly.
-
-% In Section~\ref{sec:logrelproof}, we use the standard technique that
-% is commonly seen in ...
-
-% Our consistency proof is short and fully mechanized. The proof scripts
-% involve less than 1000 lines of manually written Coq code. In fact, what we find
-% encouraging is that among the 1000
-% lines of Coq code, 400 lines are related to the specification of the
-% type system, semantics, and properties related to untyped lambda
-% terms. The semantic type soundness proof through logical relation
-% takes almost the same amount of code as our syntactic type soundness
-% proof!
-% Thanks to the conciseness of
-% the proof, we are able to present it in detail in
-% Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
-% structure of our mechanization closely corresponds to the proof we
-% present in the text, enabling us to label at the footnote each lemma
-% directly to their counterpart in the proof script for the readers to
-% reference and validate.
-
-% The technique we use is most similar to the one from \citet{nbeincoq},
-% which leverages impredicativity to define the logical
-% relation as a partial function.
-% Rather than framing impredicativity as a
-% mechanism for encoding induction
-% recursion~\citep{induction-recursion-dybjer}, a scheme in which
-% semantic models (including logical relations) for dependent types can
-% be defined, we opt for a direct explanation through the informal
-% language of sets, where impredicativity manifests in the form of
-% second-order logical formulas and thus more intuitive to grasp
-% for readers with a general mathematical background.
-
-% A naive attempt at proving consistency through induction over the
-% typing derivation would fail since the inductive hypothesis is not
-% strong enough to derive the consistency result. Instead, one typically
-% relies on the technique referred to as proof by logical relation to
-% interpret types as reducibility predicates to strengthen the inductive
-% hypothesis.
-
-% This paper is specifically
-% about establishing logical consistency for a fully dependently typed
-% system with an infinite universe hierarchy and support for large
-% elimination .
-% The type system, presented in Section~\ref{sec:spec}, is
-% most similar to Martin-Lof's predicate type theory with the minor
-% difference that type conversion is based on untyped equality.
-
-% Rather,
-% our goal is to present the proof in a form that is digestable by a
-% working type theorist and can be more readily mechanized in a proof
-% assistant. Compared to existing efforts at mechanizing logical
-% consistency or stronger properties such as existence of normal
-% form~\citep{nbeincoq},
-% decidable type checking~\citep{decagda}, our work is minimal since it requires very
-% little scaffolding and therefore results in an extremely succinct
-% proof of under 1000 lines of manually written Coq code for a dependent
-% type theory that is reasonably complete in its features.
-
-% The key technique that underlies our consistency proof is proof by
-% logical relation. In
-% Section~\ref{sec:spec}, we present the dependent type theory of
-% interest. In Section~\ref{sec:logreldep}, we give the definition
-% of the logical relation for the dependent type theory. Rather than
-% presenting the logical relation as an inductive-recursive definition,
-% we use the more elementary concept of a partial function to capture
-% the interpretation of types. The alternative representation requires us
-% to show that the set of equations indeed defines a partial function;
-% that is, for each input, there should always be a unique
-% output.
-% From the interpretation function, we can define the semantic
-% typing judgment for the set of lambda terms.
-% In Section~\ref{sec:logrelproof}, we prove the fundamental theorem,
-% which states that syntactic typing implies semantic typing. Once the
-% fundamental theorem is established, logical consistency follows as a
-% trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
-% specifics related to the Coq mechanization of the proof described in earlier
-% section.
-% Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
-% existing literature related to logical consistency about dependent
-% type theory.
+Not only does this close connection aid readers that wish to, like us, adopt
+proof assistants for their day-to-day use, but we also find that this
+precision is important for conveying the proof technique itself. Unlike
+properties that are derivable through syntactic means, proofs by logical
+relations make demands on the strength of the metalogic in which they are
+expressed. An informal proof that attempts to be agnostic or ambiguous about
+the underlying metatheory requires substantial effort from the reader to
+understand whether it is definable in a given ambient logic.
+
+%% SCW: I think we said this already in the 2nd paragraph above.
+% Finally, our approach does not result in a verbose proof on paper, thanks to
+% the already short mechanization. While we do make a few simplifications
+% to our typeset proof, such as avoiding the clunky syntax for well-founded
+% recursion in Coq, we keep the overall structure of our typeset proofs
+% consistent with our mechanization.
 
@@
                   & \mbox{conditional expression} \\ \\
-% \mathit{Renaming}\\
-% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
-\mathit{Substitutions}\\
-[[rho]] & \in & [[SVar -> STm]] & \\ \\
-\mathit{Typing Contexts}\\
- [[G]]       & ::= & [[empty]]\ |\ [[G ++ x : A]] &  \\ \\
+\end{array}
+\]
+\[
+\begin{array}{lll}
+\mathit{Substitutions}      & \qquad\qquad & \mbox{\emph{Typing Contexts}}  \\
+[[rho]] \in [[SVar -> STm]] & & [[G]] ::= [[empty]]\ |\ [[G ++ x : A]] 
 \end{array}
@@
 type theory, terms and types are collapsed into the same syntactic
-category. The type $[[Set i]]$ represent universe types, annotated by
-the universe level, a natural number $i$.%  and $[[x]]$
-% represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
-% $[[n]]$ are all metavariables representing natural numbers, we always use
-% $[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
-% variables.
-% We use de Bruijn indices for our Coq development, but use nominal
-% representation in the paper because it is more amenable to automation
-% (Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
-% conditions.
-Abstractions $[[ \ x . a ]]$ and dependent function types $[[Pi x : A . B]]$
-are binding forms for the variable $x$ in the body of the function and
-codomain of the function type.  \footnote{In the exposition in this paper, we
-  adopt the Barendregt Variable Convention and assume that binding forms are
-  equal up to $\alpha$-conversion.  In some places we will be deliberately
-  informal about the treatment of variables; our mechanized proofs use de
-  Bruijn representation.}  We use the notation $[[A -> B]]$ when the output
-type $[[B]]$ is not dependent on the input variable.  We include in \lang{}
-the intensional identity type $[[a ~ b : A]]$ whose proofs can be eliminated
-by the J-eliminator $[[J c a b p]]$, where $[[p]]$ is an equality proof
-between $[[a]]$ and $[[b]]$, and $[[c]]$ is the term whose type is to be
-casted.  Finally, \lang{} includes booleans, with standard syntax.
-\scw{Example of what you can use identity types for? Or an example of a
-  program that uses J?}
-
-% We find de Bruijn
-% representation advantageous for specification since it leaves very
-% little ambiguity about the variable freeness side
-% conditions, making our proof more easily reproducible. Furthermore, as we discuss in
-% Section~\ref{sec:automation}, de Bruijn representation is much more
-% amenable to automated reasoning in proof assistants.
-
-% \scw{Need to explain the notations in the text. What are these operations?
-% What do readers need to understand about them? Help me understand
-% the figure} \yl{resolved}
-% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
-% indices work.} \yl{resolved?}
-% We omit most of the
-% definitions of renaming and substitution and only show the definition
-% of a few representative cases of substitution.
-%\begin{figure}[ht]
-%  \begin{equation*}
-%    \begin{split}
-%      \begin{array}{lll}
-%        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
-%        [[idtm x]] & := & [[x]] \\ \\
-%        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: x -> a]]) } \\
-%        ([[rho .: x -> a]])([[x]]) & := & [[a]] \\
-%        ([[rho .: x -> a]])([[y]]) & := & [[rho y]] \text { if  } y
-%                                          \neq x\\ \\
-%      \end{array}
-%    \end{split}
-%    \qquad \qquad
-%    \begin{split}
-%      \begin{array}{lll}
-%        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
-%        [[x { rho }  ]] & := &  [[rho x]] \\
-%        [[(Pi x : A . B) { rho }]] & := & [[Pi x : A { rho } . B { rho .: x -> x }]] \\
-%        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
-%        [[(\ x . a) { rho }]] & := & [[\ x . (a { rho .: x -> x })]] \\
-%        \ldots
-%      \end{array}
-%    \end{split}
-%  \end{equation*}
-%  \caption{Auxiliary Functions over Syntax}
-%  \label{fig:auxdef}
-% \end{figure}
-
-% Without providing
-% the full definition of the renaming and substitution functions, it is
-% impossible to tell the binding structure from the syntax
-% alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
-% with the de Bruijn depth of each term, though we note that the
-% syntax we work with is unscoped and the choice does matter when
-% we extend our logical relation to open terms in Section~\ref{sec:extension}.
-
-% Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
-% syntax, including renaming, substitution, and operations over
-% the typing context or substitution.
+category. The type $[[Set i]]$ represent a universe type, annotated by its
+universe level, a natural number $i$.  Abstractions $[[ \ x . a ]]$ and
+dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
+$x$ in the body of the function and codomain of the function type.
+\footnote{In the exposition in this paper, binding forms are equal up to
+  $\alpha$-conversion and we adopt the Barendregt Variable
+  Convention~\cite{barendregt:book}, which lets us assume that bound variables are
+  distinct.  In some places, we are informal about the treatment of variables
+  and substitution; our mechanized proofs make these notions precise by using
+  de Bruijn indices~\cite{debruijn:autosubst}. }  We use the notation
+$[[A -> B]]$ when the output type $[[B]]$ is not dependent on the input
+variable.  We include in \lang{} the intensional identity type $[[a ~ b : A]]$
+whose proofs can be eliminated by the J-eliminator $[[J c a b p]]$, where
+$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[c]]$ is the
+term whose type is to be casted.  Finally, \lang{} includes booleans, with
+standard syntax.  \scw{Example of what you can use identity types for? Or an
+  example of a program that uses J?}
 
@@
 input is $[[true]]$ or $[[false]]$.
-
-% We adapt from \citet{autosubst2} notation for simultaneous renaming,
-% substitution, and other auxiliary definitions used in our paper, summarized in
-% Figure~\ref{fig:auxdef}.
+\scw{This sentence seems out of place.}
+
 Our reduction a typing relations are defined in terms of \emph{simultaneous
@@
 Before we specify the typing rules, we first specify the equational theory
-used in its conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). This
-relation is often called \textit{definitional equality} in dependent type theories
-because it defines the equivalence that the syntactic type system works
-up to.
-
-For \lang{}, we use an untyped equational theory, defined in an algorithmic
-style using the notion of parallel reduction, written $[[a => b]]$, and shown
-in Figure~\ref{fig:par}. (For brevity, reflexivity and congruence rules are
-omitted from this figure). The notation $[[a =>+ b]]$ indicates the transitive
-closure of parallel reduction.
+used in the conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). The
+equivalence relation used in this rule is often referred to
+\textit{definitional equality} in dependent type theories because it defines
+the equivalence that the syntactic type system works up to.
+
+In \lang{}, we use a relation called \emph{convertibility} for definitional
+equality. Two terms are convertible, if they reduce to a common form. The
+reduction that we use is called \emph{parallel reduction}, written
+$[[a => b]]$. The notation $[[a =>+ b]]$ indicates the transitive closure of
+parallel reduction.
 
@@
 
-Convertibility is an equivalence relation. We sketch the sequence of lemmas
-required to derive this property below and omit the details. The key step in
-proving transitivity, is showing that parallel reduction satisfies the diamond
-property.  Our technique for proving the properties about parallel reduction
-is based on \citet{takahashi-parallel-reduction}. A modern exposition of the
-same technique can be found in \citet{plfa22.08}.
-
+The definition of the parallel reduction relation, appears in
+Figure~\ref{fig:par}. (For brevity, the reflexivity and congruence rules of
+this relation are omitted from this figure).
+
+We prove, through standard
+techniques~\citet{takahashi-parallel-reduction,plfa22.08}, the following
+properties of parallel reduction.
 \begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
@@
 \end{lemma}
+
+Convertibility is an equivalence relation. The key step in
+proving transitivity is showing the diamond property for parallel reduction.
+
 \begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
@@
 \end{lemma}
-% From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
-% \ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
-% equivalence relation.
-
-Our definition of convertibility for \lang{} is unusual in that it is directly
-defined via parallel reduction, instead of using the related notion of
-$\beta$-equivalence~\cite{barendregt:pts,cic}. This choice does not change the
-language definition; a detailed argument of the equivalence between
-$[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
-\citet{barendregt:lambda-calculi-with-types} and
+
+The convertibility relation that we use for conversion in \lang{} is unusual
+in that it is directly defined via parallel reduction, instead of using the
+related notion of $\beta$-equivalence~\cite{barendregt:pts,cic}. This choice
+does not change the language definition; a detailed argument of the
+equivalence between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found
+in \citet{barendregt:lambda-calculi-with-types} and
 \citet{takahashi-parallel-reduction}. However, this choice simplifies later
@@
 conversion is similar to Barendregt's Pure Type
-Systems~\cite{barendregt1991introduction} and differs
+Systems~\cite{barendregt1991introduction,barendregt:pts} and differs
 from MLTT~\citep{Martin-Lof-1973}, where the judgmental
@@
   wrong message that we extended Siles' work with eta:
- Furthermore, discuss how this definition can be extended with
+ Furthermore, we discuss how this definition can be extended with
  $\eta$-equivalence of functions in Section~\ref{XXX} }
@@
 
-% In \rref{T-Var}, $G n$ is the partial function defined by the equations
-% $ (G ++ A) 0 := A$ and $ (G ++ A) Suc n  := G n$.  The
-% precondition $n < | G | $, where $ | G | $ represents the length of
-% the context $G$, ensures that $G n$ has a defined output. The type
-% $G n$ needs to be shifted by $Suc n$ since the types stored in the
-% context $G$ are scoped differently. In the typing judgment
-% $G |- a : A$, we want both $a$ and $A$ to have the same
-% scope. However, the closer a type is to the head of $G$, the smaller de
-% Bruijn it has since there is less variables in the context it can refer to
-% (see \rref{Ctx-Cons}). Shifting $G n$ by $Suc n$ weakens the type so
-% it is scoped consistently with the full context $G$ rather than a
-% truncated version of $G$.
-
-% Although the use of de Bruijn indices complicates the variable rule, the rules
-% that involve binding, such as \rref{T-Pi,T-Abs}, are straightforward.
-
 \Rref{T-Conv} uses the convertibility relation from earlier
@@
 
+\scw{Cassia did not follow our discussion of dependent pattern matching. We need to expand}
 The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
@@
 branch. As a result, the type system communicates the information gained from
-the % run-time \yl{commented out run-time per Jonathan's comment}
-test to each of the branches of the expression.
+the test to each of the branches of the expression.
 
@@
 $p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
-is the body of the elimination form. In this rule, $[[B]]$ is the motive and has two free variables.  The substitution for these
+is the body of the elimination form. In this rule, $[[B]]$ is the motive and
+has two free variables.  When checking $[[c]]$, the substitution for these
 variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
 witnessing the information gained through dependent pattern matching.
-% In the
-% typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
-% for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
-% therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
-% shifted to refer to the same binders.
-
-% As a result, there is no loss of generality from
-
-% It is trivial to embed a system with judgmental
-% equality to a system with untyped equality by erasing typing
-% information. As a result, it is easy to port our consistency result from our
-% type system to a variant with typed judgmental equality
-% as long as the typed system does not include $\eta$ laws
-% that would require type annotations (e.g. the $\eta$-law for unit
-% types).
-% and we discuss in Section~\ref{sec:extension} how we can extend
-% our proof to handle those rules.
-% \scw{Maybe move more of this paragraph and the next to the discussion/related
-%   work section, and add a forward pointer here.}
-% \scw{Just say that like Coq, the definition of the system uses untyped
-% equality}
 
@@
 case, we are given $[[empty |- b : Pi x : A . B]]$ and $[[empty |- a : A]]$, and
-the equality that $[[B {a / x} = Bool]]$. Our goal is to show that
+the equality that $[[B {a / x} = Void]]$. Our goal is to show that
 $[[empty |- b a : Void]]$ is not possible. However, note that there is
 nothing we know of $[[b]]$ or $[[a]]$ from the induction hypothesis
-because neither $[[Pi x : A . B]]$ nor $[[A]]$ is equal to $[[Bool]]$.
+because neither $[[Pi x : A . B]]$ nor $[[A]]$ is equal to $[[Void]]$.
 We have no way of deriving a contradiction from $[[empty |- b a :
@@
 functional relation
-$[[{ (a , S0 ) | a in S , Interp I  i B { a /x } S0 }]]$. This result is
+$[[{ (a , S0 ) | a in S /\ Interp I  i B { a /x } S0 }]]$. This result is
 formally shown through the improved inversion lemma for function types
@@
   \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
-  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl /\ a <=> b  }]]$.
   \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp
@@
     S1 #) #)]]$
-  \item $[[S = { b | forall a, (# a in S0 , forall
-    S1, (# Interp I i B {a / x} S1,  b a in S1 #) #) }]]$
+  \item $[[S = { b | forall a, (# a in S0 implies forall
+    S1, (# Interp I i B {a / x} S1 /\  b a in S1 #) #) }]]$
   \end{itemize}
@@
     S1 #) #)]]$
-  \item $[[S = { b | forall a, (# a in S0 , forall
-      S1, (# InterpR i B {rho .: x -> a} S1,  b a in S1 #) #) }]]$
+  \item $[[S = { b | forall a, (# a in S0 implies forall
+      S1, (# InterpR i B {rho .: x -> a} S1 /\  b a in S1 #) #) }]]$
   \end{itemize}
@@
   \item $[[forall a0, (# a0 in S0 implies (# forall
-      S1, (# InterpR i B {rho .: x -> a0} S1,  b {rho} a0 in S1 #) #) #)]]$
+      S1, (# InterpR i B {rho .: x -> a0} S1 /\  b {rho} a0 in S1 #) #) #)]]$
   \end{itemize}

### 2024-02-28 paper/paper.tex
add Sigma types to Fig 9

@@
 \end{figure}
-The definition of in Figure~\ref{fig:logrelrec} explains how the $[[j
+The definition in Figure~\ref{fig:logrelrec} explains how the $[[j
 < i]]$ constraint in \rref{I-Set} makes our system predicative; the
@@
 that a term $[[b]]$ belongs to an interpretation of $[[Pi x : A . B]]$ if
-for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
+for all $[[a in S]]$, where $[[S]]$ is an interpretation of $[[A]]$, the
 application form $[[b a]]$ belongs to all possible interpretations of
@@
 In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
-these two rules are equivalent in the sense that every derivation of
+these two rules are equivalent in the sense that every derivation
 involving \rref{I-Pi} can be systematically replaced by
@@
 \subsection{Properties about the Logical Relation}
-In this section, we prove four important facts about
+In the rest of this section, we prove four important facts about
 our logical relation: irrelevance (Lemma~\ref{lemma:logrelcoherence}),
@@
 it is not immediately obvious why each type $[[A]]$ can only uniquely
-corresponds to one set $[[S]]$. The following lemma shows that our
+correspond to one set $[[S]]$. The following lemma shows that our
 logical relation is indeed functional.
@@
 improved inversion lemma for function types whose statement is free of
-the relation $[[F]]$, analogous to the admissible rule \rref{I-PiAlt}.
+the relation $[[F]]$, analogous to the admissible \rref{I-PiAlt}.
 \begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
@@
   depend on anything}
+\scw{Jon didn't understand this proof}
 \begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
@@
   all $[[j < i]]$. However, by the definition of $[[I]]$, we know that
-  $[[A in I j]]$ if only if there exists some $[[S]]$ such that
+  $[[A in I j]]$ if and only if there exists some $[[S]]$ such that
   $[[Interp I i A S]]$. By \rref{I-Red}, we must also have $[[B in I
@@
 which depends on Lemma~\ref{lemma:logreldeter}.
-% Finally, despite the for all quantification over the interpreted set
-% $[[S]]$ and the universe level $[[i]]$ in the definition of $[[rho |= G]]$, we only need show that there exists some $[[i]]$ and $[[S]]$
-% such that $[[InterpR i A { rho } S]]$ and $[[rho x in S]]$ since the logical relation is a partial function (Lemma~\ref{lemma:logreldeter}).
-This idea is encapsulated in the second of the two structural
-% law about $[[rho |= G]]$.
+%
 \begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
@@
 From $[[G ++ x : A |= B : Set i]]$, we know that there must
-exists $[[S]]$ such that $[[InterpR i B {rho .: x -> a} S]]$ for every $[[a
+exist $[[S]]$ such that $[[InterpR i B {rho .: x -> a} S]]$ for every $[[a
 in S0]]$. The conclusion immediately follows from the admissible \rref{I-PiAlt}.
@@
   Proof by mutual induction over the derivation of $[[G |- a :
-  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
+  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follow
   from Lemmas~\ref{lemma:semwffempty} and \ref{lemma:semwffcons}. The semantic typing rules
@@
   case of the inversion property (Lemma~\ref{lemma:interpinv}), we
-  know that $[[S]]$ must be the empty set and therefore contradicts
+  know that $[[S]]$ must be the empty set, contradicting
   the assumption that $[[a in S]]$.
@@
 hold in $\eta$ (maybe it's fine if we just don't mention it in the
-$\eta$ case)} Instead of the
+$\eta$ case)}\scw{A terminal form is a syntactic characterization of terms that don't step 
+according to a particular relation. A normal form is a terminal form of a normalizing relation.}
+ Instead of the
 metavariables $e$ and $f$, we also
@@
 
+\scw{Cassia didn't understand this part}
 The rule for equality type $[[a ~ b : A]]$ is augmented with the precondition that
@@
 not normalizing.
-% \scw{I don't see this precondition in the figure!}\yl{resolved}
 Furthermore, the side condition $[[a <=> b]]$ is only
@@
   [[{A | exists S , InterpR i A S}]]$.
-  We then proceed by structural over the derivation of $[[Interp I i A
+  We then proceed by structural induction over the derivation of $[[Interp I i A
   S]]$. The only interesting cases are \rref{I-Pi} and \rref{I-Set}.
@@
 once we have laid the foundational technique for handling dependent types in
-our logical relation, the extension to open terms mostly boil down to
+our logical relation, the extension to open terms mostly boils down to
 properties that can be independently derived from the logical relation through
@@
 \label{sec:betaeta}
-\scw{Need an introduction to this section. Why do we care about this new
-rule? What does this section demonstrate?}
-\scw{The way we describe this
-section should more clearly indicate that we have a mechanized version of this
-proof. And include pointers on where readers might refer to it.}
-\yl{Added the related work at the beginning but I wonder if we should
-  just say that we want to match our expressiveness with existing systems}
-
-\citet{nbeincoq,decagda,martin-lof-a-la-coq} all
+\citet{nbeincoq,decagda,martin-lof-a-la-coq}
 include the $\eta$ law for functions in their equational theory and
@@
 forms from Section~\ref{sec:extension} to the existence of
-$\beta\eta$-normal forms after we add the function $\eta$ law to our
-equational theory.
+$\beta\eta$-normal forms after this addition.
 
@@
 numbers less than $[[i]]$, the universe level. In Coq, we only require
-$[[I]]$ to be function with the set of natural numbers as its domain.
+$[[I]]$ to be a function with the set of natural numbers as its domain.
 In the Coq encoding of $[[InterpR i A S]]$ (Figure~\ref{fig:logrelrec}), we define $[[I]]$ as follows.
@@
     NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & Correctness of NBE \\
-    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} & Decidability of conversion \\
-    % \hline
-    % \hline
-    MLTT-\'a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
-    % \hline
+    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & {\boxedsymbols ✓} & Decidability of conversion \\
+    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & {\boxedsymbols ✓} & Decidibility of type checking \\
     \hline
-    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
-    % CoreSpec & \\
-    % PCUIC & \\
-    %   & \lang{} & \citet{anand2014towards} &
-    %                                          \citet{barras2012semantical}
-    % & \citet{martin-lof-a-la-coq}  \\
-    % \hline
-    % Universe & Countable & Countable & Impredicative + Countable & One\\
-    % % \hline
-    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
-    % \hline
-    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
-    % \hline
-    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
   \end{tabular}
@@
   \underline{Univ}erses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
-  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), W-types \\
+  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), $\Sigma$-types ($\Sigma$), W-types \\
   \underline{Conv}ersion:& Untyped (U), Typed (T),  Extensional (E) \\
@@
   \caption{Feature matrix for dependently typed languages with
-    mechanized proofs by logical relation}
+    mechanized logical relations}
   \label{fig:featurematrix}

### 2024-02-28 paper/paper.tex
add column to figure for arity of the relation

@@
 % \begin{minipage}{0.8\textwidth}
-  \begin{tabular}{| l |  c  | l | c | c | l | }
+  \begin{tabular}{| l |  c  | l | c | c | c | l | }
     \hline
@@
 %      & \header{Large Elim} & Main results \\
-      & \header{Univ} & \header{Ind} & \header{Conv}
-      & \header{L E} & Main results \\
+      & \header{U} & \header{Ind} & \header{C} 
+      & \header{L E} & \header{A} & Main results \\
 
     \hline
-    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓}
+    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} & 1 
     & Consistency and normalization \\
     % \hline
-    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & Consistency \\
-    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & Consistency \\
-    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & Correctness of NBE \\
-    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & {\boxedsymbols ✓} & Decidability of conversion \\
-    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & {\boxedsymbols ✓} & Decidibility of type checking \\
+    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & 1 & Consistency \\
+    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & 2 & Consistency \\
+    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & ?? & Correctness of NBE \\
+    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & {\boxedsymbols ✓} & 2 & Decidability of conversion \\
+    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & {\boxedsymbols ✓} & ?? & Decidibility of type checking \\
     \hline
@@
     \\
-  \underline{Univ}erses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
+  \underline{U}niverses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
   \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), $\Sigma$-types ($\Sigma$), W-types \\
-  \underline{Conv}ersion:& Untyped (U), Typed (T),  Extensional (E) \\
+  \underline{C}onversion:& Untyped (U), Typed (T),  Extensional (E) \\
   \underline{L}arge \underline{E}liminations:& Included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
+  \underline{A}rity of interpretation: & Sets of terms (1), Relations between terms (2) \\
   \\
@@
 \end{figure}
+\scw{Extra columns in Fig 10:  Predicate vs. Relational interpretation (we can explain this) / Typed vs. Untyped interpretation}
 
@@
   it is tricky to explain what we actually mean. }
+\scw{I don't think it is that trick}
 Figure~\ref{fig:featurematrix} presents several mechanized proofs
@@
 either impredicativity or induction-recursion. Their encoding of a countable
-universe hierarchy still relies on impredicativity, similar to our
+universe hierarchy relies on impredicativity, similar to our
 development.

### 2024-02-28 paper/paper.tex
last batch of typos

@@
 conversion rule for a dependently typed language with one predicative
-universe level and typed judgmental equality with function
+universe level and a typed judgmental equality that includes the function
 $\eta$ law. They
@@
 development.
-% \scw{How did the Nuprl
-% proof include a full hierarchy? Did they use impredicativity?}
-% \yl{commented out because I think this paragraph is too subjective}
-% However, instead of explaining the inductive
-% logical relation as a convoluted workaround to the lack of
-% induction-recursion in Coq, we give a self-contained explanation of
-% our logical relation.
 
@@
 $CC^\omega$, a system that incorporates a predicative universe on top
-of the impredicative sort in Calculus of Constructions. We choose to
+of the impredicative sort in the Calculus of Constructions. We choose to
 focus on a syntactic term model so we do not have to take the extra step
@@
 only involve properties that are derivable through syntactic means. For
-example, \citet{coqcoqcorrect2019} proves the correctness of a type checker
-for the Polymorphic, Cumulative Calculus of Inductive Constructions (PCUIC),
+example, \citet{coqcoqcorrect2019} prove the correctness of a type checker
+for the Predicative, Cumulative Calculus of Inductive Constructions (PCUIC),
 Coq's core calculus, assuming the strong normalization property of the object
@@
 dependent Haskell, and prove the syntactic type soundness of the type
-system. Compared to properties such as consistency, normalization, and
-parametricity, the most difficult part for syntactic type soundness is the
-consistency of definitional (not propositional) equality. In their setting,
-the result follows from the confluence of parallel reduction.
+system. Because System D includes nontermination, they proved
+the consistency of definitional equality from the confluence of 
+parallel reduction.
 
@@
 languages. However, what remains unanswered is what makes our development
-significantly shorter compared to others. Are we proving simpler results, for
+significantly shorter compared to others. Are we proving simpler results for
 smaller languages, are we making more use of automation, or is our proof
@@
 justify consistency. Compared to the systems from
-\citet{decagda,nbeincoq, martin-lof-a-la-coq}, the conversion rule for
+\citet{decagda, nbeincoq, martin-lof-a-la-coq}, the conversion rule for
 \lang{} is untyped and therefore we do not
@@
 procedure for type
-conversion~\citet{pierce2004advanced}. \citet{nbeincoq, decagda}, on
+conversion~\cite{pierce2004advanced}. \citet{nbeincoq, decagda}, on
 the other hand, need
@@
 diamond property, which is easy to derive syntactically.
-\citet{decagda,nbeincoq} employ a deterministic weak
-head reduction relation. The use of a deterministic reduction relation
-makes the functionality of a logical relation trivial to prove, but
-the downside is that weak head reduction (denoted by $\leadsto^{wh}$)
-or leftmost-outermost reduction (denoted by $\leadsto^{lo}$) does
-not satisfy even the weaker substitution property
-(Lemma~\ref{lemma:parsubst}).
-With this alternative formulation, we would have to prove how the
-deterministic reduction relates to a full non-deterministic reduction
-relation to prove the fundamental theorem. This would amount
-to proving the factorization theorem of the deterministic reduction
-relation with respect to full non-deterministic reduction. Instead,
-using parallel reduction in \rref{I-Red} allows us to delay the
-factorization property until we need to justify the use leftmost-outermost
-reduction as a deterministic reduction strategy for normalization.
+
+In contrast, \citet{decagda} and \citet{nbeincoq} employ a deterministic weak
+head reduction relation. The use of a deterministic reduction relation makes
+the functionality of a logical relation trivial to prove, but the downside is
+that weak head reduction (denoted by $\leadsto^{wh}$) or leftmost-outermost
+reduction (denoted by $\leadsto^{lo}$) does not satisfy even the weaker
+substitution property (Lemma~\ref{lemma:parsubst}).  With this alternative
+formulation, we would have to prove how the deterministic reduction relates to
+a full non-deterministic reduction relation to prove the fundamental
+theorem. This would amount to proving the factorization theorem of the
+deterministic reduction relation with respect to full non-deterministic
+reduction. Instead, using parallel reduction in \rref{I-Red} allows us to
+delay the factorization property until we need to justify the use
+leftmost-outermost reduction as a deterministic reduction strategy for
+normalization.
 
@@
 Fortunately, there are syntactic workarounds for the $\Pi$-injectivity
-problem that allow us to recover confluence property independently
+problem that allow us to recover the confluence property independently
 from the logical relation. \citet{siles2012pure} generalize the
@@
 \label{sec:conclusion}
-In this work, we show a short and mechanized proof of consistency
-through proof by logical relation for a fully dependently typed
-language with a full universe hierarchy, an intensional identity type,
-and large elimination. We show the extensibility of our approach by
-proving the existence of $\beta\eta$-normal forms with only small
-and mechanical changes to our proof development. Our Coq mechanization
-leverages existing Coq libraries for reasoning about metatheory and
-general-purpose automation, allowing us to significantly reduce the
-verbosity typically associated with mechanized proofs and recover a
-declarative proof style that is usually only available in pen and
-paper. Our development shows that proof by logical relation for
-dependent types should be more accessible since it does not always
+In this work, we present a short and mechanized proof of consistency through
+proof by logical relations for a dependently typed language with a full
+universe hierarchy, large eliminations, an intensional identity type, and
+dependent eliminators. We show the extensibility of our approach by proving
+the existence of $\beta\eta$-normal forms with only small and mechanical
+changes to our proof development. Our Coq mechanization leverages existing Coq
+libraries for reasoning about metatheory and for general purpose automation,
+allowing us to significantly reduce the verbosity typically associated with
+mechanized proofs. The result is a declarative proof style that is usually
+only available on pen and paper. Our development shows that proofs by logical
+relations for dependent types are accessible and do not
 require months of effort to implement. We hope our proof can inspire
-researchers to more actively mechanize results such as consistency,
-normalization for dependent type theory.
+researchers to more frequently mechanize results, such as consistency and
+normalization, for their dependent type theories.
 

### 2024-02-28 paper/paper.tex
Add unicode brackets
Move ott dependencies into ott preamble
Fill in the missing entries of Fig. 9

@@
   ,anonymous=\ifanonymous true\else false\fi]{acmart}
-\usepackage{fontspec}
-\newfontfamily\boxedsymbols{DejaVu Sans}
 \usepackage{ottalt}
@@
 \usepackage{xspace}
-\usepackage{tcolorbox}
 \usepackage[para]{footmisc}
-\definecolor{lightgray}{gray}{0.85}
 \newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
@@
     Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & 2 & Consistency \\
-    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & ?? & Correctness of NBE \\
+    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & 2 & Correctness of NBE \\
     $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & {\boxedsymbols ✓} & 2 & Decidability of conversion \\
-    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & {\boxedsymbols ✓} & ?? & Decidibility of type checking \\
+    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & {\boxedsymbols ✓} & 2 & Decidibility of type checking \\
     \hline

### 2024-02-28 paper/paper.tex
missing refs

--- 2024-02-28 paper/paper.tex
@@
 
-\commentstrue         %% toggle comments here
+\commentsfalse        %% toggle comments here
 \extendedfalse
@@
   $\alpha$-conversion and we adopt the Barendregt Variable
-  Convention~\cite{barendregt:book}, which lets us assume that bound variables are
-  distinct.  In some places, we are informal about the treatment of variables
-  and substitution; our mechanized proofs make these notions precise by using
-  de Bruijn indices~\cite{debruijn:autosubst}. }  We use the notation
-$[[A -> B]]$ when the output type $[[B]]$ is not dependent on the input
-variable.  We include in \lang{} the intensional identity type $[[a ~ b : A]]$
-whose proofs can be eliminated by the J-eliminator $[[J c a b p]]$, where
-$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[c]]$ is the
-term whose type is to be casted.  Finally, \lang{} includes booleans, with
-standard syntax.  \scw{Example of what you can use identity types for? Or an
-  example of a program that uses J?}
+  Convention~\cite{barendregt:lambda-calculus}, which lets us assume that
+  bound variables are distinct.  In some places, we are informal about the
+  treatment of variables and substitution; our mechanized proofs make these
+  notions precise by using de Bruijn indices~\cite{debruijn1994automath}. }
+We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
+on the input variable.  We include in \lang{} the intensional identity type
+$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
+$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
+$[[b]]$, and $[[c]]$ is the term whose type is to be casted.  Finally, \lang{}
+includes booleans, with standard syntax.  \scw{Example of what you can use
+  identity types for? Or an example of a program that uses J?}
 
@@
 says that an equality type $[[a ~ b : A]]$ corresponds to the
-set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
-otherwise corresponds to the empty set. Side conditions like $[[a <=>
+set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ also holds and
+otherwise corresponds to the empty set. Conditions like $[[a <=>
 b]]$ are typically required for indexed types, of which equality types
@@
 of \lang{}.
-
+%
 \begin{figure}[h]
@@
 \end{figure}
-
+%
 The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
@@
 not normalizing.
-Furthermore, the side condition $[[a <=> b]]$ is only
-required/available when the equality proof reduces to $[[refl]]$. If the proof
-term reduces to a neutral term, then there is nothing we need to show/can learn
+Furthermore, the condition $[[a <=> b]]$ is only
+required when the equality proof reduces to $[[refl]]$. If the proof
+term reduces to a neutral term, then there is nothing we need to show
 about the relationship between $[[a]]$ and $[[b]]$.
@@
 \begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
-  Let $[[S]]$ be a set of lambda terms. We say that $[[S]] \in CR$
+  Let $[[S]]$ be a set of terms. We say that $[[S]] \in CR$
   if and only if conditions $CR_1$ and $CR_2$ hold.
@@
   The function case requires Lemmas~\ref{lemma:extwn} and
-  \ref{lemma:wnewn} we have shown earlier.
-
-  The \rref{I-Set} case is the most involved. We need to show that
+  \ref{lemma:wnewn}, which we have shown earlier.
+
+  The \rref{I-Set} case is the most interesting. We must show that
   for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
@@
 \end{proof}
+
+\begin{corollary}[Canonicity]
+If $[[G |- a : bool]]$ then $[[a =>+ true]]$ or $[[a =>+ false]]$. 
+\end{corollary}
 

### 2024-02-28 paper/paper.tex
add canonicity corollary to the paper

@@
 
+Our soundness theorem also tells us something about closed terms of type 
+$[[Bool]]$; they either reduce to $[[true]]$ or $[[false]]$.
+
+\begin{corollary}[Canonicity\footnote{\dotv{soundness.v}{canonicity}}]
+If $[[empty |- b : Bool]]$, then either $[[b =>+ true]]$ or $[[b =>+ false]]$.
+\end{corollary}
+\begin{proof}
+  The proof is similar to above, except that we use the $[[Bool]]$ case of the
+  inversion property.
+\end{proof}
+
 \section{Existence of $\beta$-normal forms}
@@
 Furthermore, if we know that an application of a term to a variable
-has a normal form, then we know the term must have a normal form.
+has a normal form, then we know that the term must have a normal form.
 \begin{lemma}[Wn extensionality\footnote{\dotv{normalform.v}{ext\_wn}}]
@@
 \end{proof}
-
-\begin{corollary}[Canonicity]
-If $[[G |- a : bool]]$ then $[[a =>+ true]]$ or $[[a =>+ false]]$. 
-\end{corollary}
 

### 2024-02-28 paper/paper.tex
add future work to conclusion

@@
 
-\section{Conclusion}
+\section{Future work and Conclusions}
 \label{sec:conclusion}
@@
 the existence of $\beta\eta$-normal forms with only small and mechanical
-changes to our proof development. Our Coq mechanization leverages existing Coq
-libraries for reasoning about metatheory and for general purpose automation,
-allowing us to significantly reduce the verbosity typically associated with
-mechanized proofs. The result is a declarative proof style that is usually
-only available on pen and paper. Our development shows that proofs by logical
-relations for dependent types are accessible and do not
-require months of effort to implement. We hope our proof can inspire
-researchers to more frequently mechanize results, such as consistency and
-normalization, for their dependent type theories.
+changes to our proof development. Related work gives us confidence that we
+could extend our logical relation to include features such as full inductive
+datatypes, irrelevant arguments, and type-directed conversion; though it is
+not clear how much of the brevity of the development we will be able to
+maintain.  Furthermore, we hope that mechanized logical relations proofs will
+eventually grow to include other features found in dependent type theories,
+such as impredicative universes, universe polymorphism, and cumulativity.
+
+Our Coq mechanization leverages existing Coq libraries for reasoning about
+metatheory and for general purpose automation, allowing us to significantly
+reduce the verbosity typically associated with mechanized proofs. The result
+is a declarative proof style that is usually only available on pen and
+paper. Our development shows that proofs by logical relations for dependent
+types are accessible and do not require months of effort to implement.
+We hope our proof can inspire researchers to more frequently mechanize
+results, such as consistency and normalization, for their dependent type
+theories. By design, the goals of our work are modest.
 

### 2024-02-28 paper/paper.tex
Typos up until Sec 4

@@
 fully mechanized the consistency proof using the Coq proof assistant
-in under 1000 lines of code, with 400 lines of additional code for the
+in under 1000 lines of code, with 500 lines of additional code for the
 $\beta\eta$-normal form extension.
@@
 
-Our reduction a typing relations are defined in terms of \emph{simultaneous
+Our reduction and typing relations are defined in terms of \emph{simultaneous
   substitutions}, $[[rho]]$, which are mappings from variables to terms.  We
@@
 in that it is directly defined via parallel reduction, instead of using the
-related notion of $\beta$-equivalence~\cite{barendregt:pts,cic}. This choice
+related notion of $\beta$-equivalence~\citep{barendregt1991introduction,coquand1990:cic}. This choice
 does not change the language definition; a detailed argument of the
@@
 conversion is similar to Barendregt's Pure Type
-Systems~\cite{barendregt1991introduction,barendregt:pts} and differs
+Systems~\cite{barendregt1991introduction} and differs
 from MLTT~\citep{Martin-Lof-1973}, where the judgmental
@@
 evaluate to $[[true]]$ or $[[false]]$. However, the characterization
-of $[[Bool]]$ in our logical relation does not always correspond to
-well-typed or even closed terms. For example, we have $[[if false (#
+of $[[Void]]$ and $[[Bool]]$ in our logical relation does not always correspond to
+well-typed or even closed terms. For example, the term $[[if false (#
 Void true #)
@@
 \begin{proof}
-  Immediate from Lemma~\ref{lemma:logreldeter} and
+  Immediate from Lemmas~\ref{lemma:logreldeter} and
   \ref{lemma:logrelcumulativity}.
@@
   Figure~\ref{fig:logrelrec}, we unfold $[[InterpR i A S]]$ by one
-  step into $[[ Interp I i A S  ]]$ where $[[I i]] := [[{A | exists S
-    , InterpR i A S}]]$.
+  step into $[[ Interp I i A S  ]]$ where $[[I j]] := [[{A | exists S
+    , InterpR j A S}]]$.
   We then proceed by induction over the derivation of $[[Interp I i A S]]$.
@@
   $[[A in I j]]$ if and only if there exists some $[[S]]$ such that
-  $[[Interp I i A S]]$. By \rref{I-Red}, we must also have $[[B in I
+  $[[Interp I j A S]]$. By \rref{I-Red}, we must also have $[[B in I
   j]]$ for all $[[B => A]]$.
@@
   pierce2002types}. In our setting, attempting to show that
-$[[empty |- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ will fail in
+$[[empty |- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ through induction over the derivation of $[[empty |- a : A]]$ will fail in
 \rref{T-Abs}, where the induction hypothesis is not helpful since the body of
@@
   \item $[[S = { b | forall a, (# a in S0 implies forall
-      S1, (# InterpR i B {rho .: x -> a} S1 /\  b a in S1 #) #) }]]$
+      S1, (# InterpR i B {rho .: x -> a} S1 implies  b a in S1 #) #) }]]$
   \end{itemize}
@@
   \item $[[forall a0, (# a0 in S0 implies (# forall
-      S1, (# InterpR i B {rho .: x -> a0} S1 /\  b {rho} a0 in S1 #) #) #)]]$
+      S1, (# InterpR i B {rho .: x -> a0} S1 implies  b {rho} a0 in S1 #) #) #)]]$
   \end{itemize}

### 2024-02-28 paper/paper.tex
Typo fixes

@@
 lemmas about reduction. First, a renaming $\xi$ is a generalization of
-weakening when working with de Bruijn indices. It consistently renumbers the
-variables that appears in terms. If a renamed term has been reduced, we can
+weakening when working with simultaneous substitutions. It consistently maps the
+variables that appears in terms to other variables. If a renamed term has been reduced, we can
 always recover the result of the reduction without the renaming.
@@
 itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
-sufficient.  In fact, all the properties we have shown in
+sufficient.
+
+All the properties we have shown in
 Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the fundamental
@@
   that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
-  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
+  We pick the $[[rho]]$ to be the identity substitution $[[idtm]]$, which injects
   variables as terms. The side condition $[[idtm |=
@@
 \ref{sec:logrelproof} and the extension to $\beta$-normalization for open and
-closed terms from Section~\ref{sec:extension}.  For comparison, we have also
+closed terms from Section~\ref{sec:extension}.
+
+The development with the $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises 1573
+lines of non-blank, non-comment lines of code. We choose not to
+include it in the chart, since the slight difference in lemma
+dependencies makes the comparison in each category less
+informative. However, we note that compared to the
+$\beta$-normalization extension, the $\beta\eta$ extension has the
+same line count in the definition of the logical relation and the
+semantic soundness proof. The difference only appears in files related
+to untyped reduction and normal forms.
+
+For comparison, we have also
 proven syntactic type soundness through subject
@@
 % Lemma~\ref{lemma:piintroalt}).
-In the Coq mechanized proof, we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. The definition of $[[Interp I i A S]]$ has type
+In the Coq mechanized proof, % we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. 
+the definition of $[[Interp I i A S]]$ has type
 \texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.
@@
 readers who wish to understand the mechanized proof. Although automation
-performs extensive search, it does not use lemmas or invert derivations that
+performs extensive search, we can configure it to not use lemmas or invert derivations that
 are not specified in the \texttt{use} or \texttt{inv} flags.
@@
   \texttt{Prop} data type with a single constructor, to be eliminated
-  to construct runtime relevant data.}.
+  to construct runtime relevant data.}. We leave the proof of the
+factorization property and the extraction of a conversion algorithm as
+part of our future work.
 \section{Related Work}

### 2024-02-28 paper/paper.tex
revision to conclusion

@@
 the existence of $\beta\eta$-normal forms with only small and mechanical
-changes to our proof development. Related work gives us confidence that we
-could extend our logical relation to include features such as full inductive
-datatypes, irrelevant arguments, and type-directed conversion; though it is
-not clear how much of the brevity of the development we will be able to
-maintain.  Furthermore, we hope that mechanized logical relations proofs will
-eventually grow to include other features found in dependent type theories,
-such as impredicative universes, universe polymorphism, and cumulativity.
-
-Our Coq mechanization leverages existing Coq libraries for reasoning about
-metatheory and for general purpose automation, allowing us to significantly
-reduce the verbosity typically associated with mechanized proofs. The result
-is a declarative proof style that is usually only available on pen and
-paper. Our development shows that proofs by logical relations for dependent
-types are accessible and do not require months of effort to implement.
-We hope our proof can inspire researchers to more frequently mechanize
-results, such as consistency and normalization, for their dependent type
-theories. By design, the goals of our work are modest.
+changes to our proof development.  Our Coq mechanization leverages existing
+Coq libraries for reasoning about metatheory and for general purpose
+automation, allowing us to significantly reduce the verbosity typically
+associated with mechanized proofs. The result is a declarative proof style
+that is usually only available on pen and paper.
+
+Related work gives us confidence that we could extend our logical relation to
+include features such as full inductive datatypes, irrelevant arguments, and
+type-directed conversion; however, it is not clear how much of the brevity of
+this development can be maintained.  Furthermore, we hope that mechanized
+logical relations proofs will eventually grow to include other features found
+in dependent type theories, such as impredicative universes, universe
+polymorphism, and cumulativity.  Regardless, our development shows that proofs
+by logical relations for dependent types are accessible and do not require
+months of effort to implement.  We hope our proof can inspire researchers to
+more frequently mechanize results, such as consistency and normalization, for
+their dependent type theories.
 

### 2024-02-28 paper/paper.tex
Change name to lambda pi and typo

--- 2024-02-28 paper/paper.tex
@@
 \newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
-\newcommand{\lang}{$\lambda^{U\mathbb{B}\ottkw{Id}}$\xspace}
+\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
 \inputott{rules}
@@
 the existing logical relation proofs for systems with typed
-judgmental. We leave that as part of our future work.
+judgmental equality. We leave that as part of our future work.
 % Therefore, we believe that even in a

### 2024-02-28 paper/paper.tex
Remove the algo part

--- 2024-02-28 paper/paper.tex
@@
 
+\ifextended
 \subsection{Extraction of a conversion algorithm}
@@
 part of our future work.
+\fi
 \section{Related Work}

### 2024-02-28 paper/paper.tex
Move the example about L.E

--- 2024-02-28 paper/paper.tex
@@
 We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
-on the input variable.  We include in \lang{} the intensional identity type
+on the input variable. For simplicity, we omit the type annotations in the abstraction
+forms. We discuss how the inclusion of type annotations can affect our
+our development in Section~\ref{sec:betaeta}, where we extend our
+consistency result to the existence of $\beta\eta$-normal forms.
+We include in \lang{} the intensional identity type
 $[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
@@
   identity types for? Or an example of a program that uses J?}
-
-The \lang language is expressive enough to support large
-eliminations, the ability to compute a type using a term as input. For
-example, the function $[[\ x . if x Bool Void]]$ returns
-either $[[Bool]]$ or $[[Void]]$ depending on whether its
-input is $[[true]]$ or $[[false]]$.
-\scw{This sentence seems out of place.}
 
@@
 
+The universe hierarchy and the boolean base type gives \lang{} the
+ability to compute a type using a term as input, a feature commonly referred to
+as large elimination. For
+example, we can type check the function $[[\ x . if x Bool Void]]$,
+which returns either $[[Bool]]$ or $[[Void]]$ depending on whether its
+input is $[[true]]$ or $[[false]]$.
 % Working with a system with untyped equality has the huge benefit that

### 2024-02-28 paper/paper.tex
Minor
Add paragraph about cumulativity and imrpedicativity

@@
 consistency proof for a dependently typed language that contains a
-rich set of features, including a full universe
+rich set of features, including a countable universe
 hierarchy, booleans, and an intensional identity type. We show that
@@
 parallel reduction.
+
+Compared to the systems we describe in this section, the most notable
+features we are missing include cumulativity and impredicativity. Our
+semantic model already satisfies the cumulativity property
+(Lemma~\ref{lemma:logrelcumulativity}), though we need to extend our
+convertibility relation into a subtyping relation in the specification
+of our syntactic typing rules. Impredicativity, on the other hand, is
+known to be difficult to model when the impredicative sort is at the
+bottom of a predicative universe hierarchy; in this scenario, the
+erasure technique from \citet{geuvers1994short} is no longer
+applicable\citep{abel2013normalization}. Whether we can find a
+similarly short and simple treatment for impredicativity remains an
+open question, which we hope to explore in the future.
 

### 2024-02-28 paper/paper.tex
25 pages

--- 2024-02-28 paper/paper.tex
@@
 \begin{figure}[h]
-\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
+\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
+\[ \drule[width=5in]{I-Pi} \]
 \caption{Logical relation for \lang}
@@
 sets with the following signatures.
-\begin{equation*}
-  \begin{split}
-    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
-    [[S]] &\in [[PowerSet STm]]
-  \end{split}
-\end{equation*}
+\[
+    [[I]] \in [[ { j | j < i  } ->  PowerSet STm ]] \qquad\qquad\qquad
+    [[S]] \in [[PowerSet STm]]
+\]
 The notation $[[PowerSet STm]]$ denotes the powerset of the set of
@@
 To tie the knot and obtain an interpretation of all universe levels,
-we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
+we define in Definition~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
 using the well-foundedness of $<$ relation on natural
@@
 
-\begin{figure}[h]
+\begin{definition}[Logical relation for all universe levels]
+\label{fig:logrelrec}
 \begin{equation*}
@@
 \end{equation*}
-\caption{Logical relation for all universe levels}
-\label{fig:logrelrec}
-\end{figure}
-The definition in Figure~\ref{fig:logrelrec} explains how the $[[j
+\end{definition}
+Definition~\ref{fig:logrelrec} explains how the $[[j
 < i]]$ constraint in \rref{I-Set} makes our system predicative; the
@@
 
-By unfolding the definition in Figure~\ref{fig:logrelrec}, we
+By unfolding Definition~\ref{fig:logrelrec}, we
 show that the same introduction rules for $[[Interp I i A S]]$ are
@@
 \begin{proof}
-  By the definition of $[[InterpR i A S]]$ from
-  Figure~\ref{fig:logrelrec}, we unfold $[[InterpR i A S]]$ by one
+  By the definition of $[[InterpR i A S]]$ 
+  (Definition~\ref{fig:logrelrec}), we unfold $[[InterpR i A S]]$ by one
   step into $[[ Interp I i A S  ]]$ where $[[I j]] := [[{A | exists S
@@
 
-  By its definition in Figure~\ref{fig:logrelrec}, we have the
+  By Definition~\ref{fig:logrelrec}, we have the
   equality $[[InterpR i A S]] = [[Interp I i A S]]$  where $[[I i]] :=
@@
 
-Furthermore, these axioms also bridge the gap between our mechanization
-and our informal proofs.
-
-For example, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$ are
-equal, it suffices to show the extensional property that
+These axioms bridge the gap between our mechanization and our informal
+proofs. For example, in set theory, to show that two sets $[[S0]]$ and
+$[[S1]]$ are equal, it suffices to show the extensional property that
 $\forall x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
 occasionally in our presented proofs.
-
-In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm
+%
+However, in Coq, sets of terms ($[[PowerSet STm]]$) are encoded as the 
+type \texttt{tm
   -> Prop}, a predicate over \lang{} terms.  In axiom-free Coq, predicates do
@@
 we cannot conclude that $P = Q$ when given a proof of
-$\forall x, P(x) \iff Q(x)$. But this is exactly the statement of the
-axiom of functional extensionality.
+$\forall x, P(x) \iff Q(x)$. But this is exactly the statement of 
+functional extensionality.
 
@@
 \texttt{Type} becomes troublesome only when we attempt to define
-$[[InterpR i A S]]$, the top-level logical relation defined in
-Figure~\ref{fig:logrelrec} that recursively calls itself at smaller universe
+$[[InterpR i A S]]$, the top-level logical relation 
+(Definition~\ref{fig:logrelrec}) that recursively calls itself at smaller universe
 levels. Therefore, the one feature of \lang{} that truly requires
@@
 $[[I]]$ to be a function with the set of natural numbers as its domain.
-In the Coq encoding of $[[InterpR i A S]]$ (Figure~\ref{fig:logrelrec}), we define $[[I]]$ as follows.
+In the Coq encoding of $[[InterpR i A S]]$ (Definition~\ref{fig:logrelrec}), we define $[[I]] \in [[SNat -> PowerSet STm]]$ as follows.
 \begin{equation*}
   \begin{split}
-    [[I]]   &\in [[SNat -> PowerSet STm]] \\
     [[I j]] &=
@@
 \geq i$ case is junk data that does not affect the result of the logical
-relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Figure~\ref{fig:logrelrec}.
+relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Definition~\ref{fig:logrelrec}.
 
@@
 \begin{center}
-  \drule[]{I-PiCoq}
+  \drule[width=5in]{I-PiCoq}
 \end{center}
@@
 First, the metatheoretic properties that we prove are indeed
-simpler. Compared to Core Nuprl~\citep{anand2014towards}, our system
+simpler. Compared to Core Nuprl, our system
 lacks extensionality, which would require a relational model to
-justify consistency. Compared to the systems from
-\citet{decagda, nbeincoq, martin-lof-a-la-coq}, the conversion rule for
-\lang{} is untyped and therefore we do not
+justify consistency. Because the conversion rule for
+\lang{} is untyped, we do not
 need a Kripke-style relational model to prove $\Pi$-injectivity
-among other properties. Furthermore, we prove the existence of normal
+among other properties, unlike systems with typed conversion. Furthermore, we prove the existence of normal
 forms, which induces a simple \emph{normalize-and-compare}
@@
 
-\section{Future work and Conclusions}
+\section{Conclusion}
 \label{sec:conclusion}
-In this work, we present a short and mechanized proof of consistency through
-proof by logical relations for a dependently typed language with a full
-universe hierarchy, large eliminations, an intensional identity type, and
-dependent eliminators.  We show the extensibility of our approach by proving
-the existence of $\beta\eta$-normal forms with only small and mechanical
-changes to our proof development.  Our Coq mechanization leverages existing
-Coq libraries for reasoning about metatheory and for general purpose
-automation, allowing us to significantly reduce the verbosity typically
-associated with mechanized proofs. The result is a declarative proof style
-that is usually only available on pen and paper.
+In this work, we present a short and mechanized proof by logical relations for
+a dependently typed language with a full universe hierarchy, large
+eliminations, an intensional identity type, and dependent eliminators.  We
+show the extensibility of our approach by proving the existence of
+$\beta\eta$-normal forms with only small and mechanical changes to our proof
+development.  Our Coq mechanization leverages existing Coq libraries for
+reasoning about metatheory and for general purpose automation, allowing us to
+significantly reduce the verbosity typically associated with mechanized
+proofs. The result is a declarative proof style that rivals pen and paper.
 

### 2024-02-28 paper/paper.tex
more shrinking.  not there yet

@@
 
-To demonstrate the scale of our proof scripts,
-Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
-code\footnote{calculated by the \texttt{cloc} tool, available from
+To demonstrate the scale of our proof scripts, Figure~\ref{fig:linecount}
+shows the number of non-blank, non-comment lines of code\footnote{calculated
+  by the \texttt{cloc} tool, available from
   \url{https://github.com/AlDanial/cloc}.} for each file of our development,
 including the base consistency proof from Section~\ref{sec:logreldep} and
-\ref{sec:logrelproof} and the extension to $\beta$-normalization for open and
-closed terms from Section~\ref{sec:extension}.
-
-The development with the $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises 1573
-lines of non-blank, non-comment lines of code. We choose not to
-include it in the chart, since the slight difference in lemma
-dependencies makes the comparison in each category less
-informative. However, we note that compared to the
-$\beta$-normalization extension, the $\beta\eta$ extension has the
-same line count in the definition of the logical relation and the
-semantic soundness proof. The difference only appears in files related
-to untyped reduction and normal forms.
-
-For comparison, we have also
-proven syntactic type soundness through subject
-reduction\footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
+\ref{sec:logrelproof} and the extension to $\beta$-normalization from
+Section~\ref{sec:extension}.  For comparison, we have also proven syntactic
+type soundness through preservation
+\footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
 progress\footnote{\dotv{syntactic\_soundness.v}{wt\_progress}}.
+
+The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
+1573 lines of non-blank, non-comment lines of code. We choose not to include
+it in the chart, because of slight differences in lemma dependencies for
+untyped reduction and normal forms that make the comparison less
+informative. However, when compared to the $\beta$-normalization extension,
+the $\beta\eta$ extension has the same line count in the definition of the
+logical relation and the semantic soundness proof.
 
@@
 reasoning about those functions. The auto-generated syntax file (291 LOC)
-and other Autosubst library files are are not included in the figure.
-
-\subsection{Axioms}
+and other Autosubst library files are also not included in the figure.
+
+\paragraph{Axioms}
 Our Coq development assumes two axioms: functional extensionality
@@
 with Coq's metatheory.
-
+%
 These axioms bridge the gap between our mechanization and our informal
@@
 
-\subsection{Encoding the logical relation in Coq}
+\paragraph{Encoding the logical relation in Coq}
 We next discuss specific details of the Coq encoding of the logical relation
@@
 $[[I]]$ to be a function with the set of natural numbers as its domain.
-In the Coq encoding of $[[InterpR i A S]]$ (Definition~\ref{fig:logrelrec}), we define $[[I]] \in [[SNat -> PowerSet STm]]$ as follows.
+In the Coq encoding of $[[InterpR i A S]]$, we define $[[I]] \in [[SNat -> PowerSet STm]]$ as follows.
 \begin{equation*}
@@
 
-\subsection{Automation}
+\paragraph{Automation}
 \label{sec:automation}
@@
 In the most general sense, a logical relation can be viewed as a
-practical technique that uses a type-indexed relation to assign
-meanings to types and strengthen the induction hypothesis for the
+practical technique that uses a type-indexed relation to 
+strengthen the induction hypothesis for the
 property of interest. The original idea of this technique can be
@@
 The same idea is explained in \citet{girard1989proofs} and extended to
-prove the strong normalization property of System F.
-
-Tait's method has been successfully applied to dependently typed
+prove strong normalization of System F.
+%
+Tait's method has also been successfully applied to dependently typed
 languages to prove strong normalization \cite{Martin-Lof-1973,luo1990extended,geuvers1994short, barendregt:lambda-calculi-with-types}.
@@
 % \begin{minipage}{0.8\textwidth}
-  \begin{tabular}{| l |  c  | l | c | c | c | l | }
+  \begin{tabular}{| l |  c  | l | c | c | c | l | l | }
     \hline
@@
 a size metric, which has been used in~\citet{liu2023dependently}.
-Their development is around 8,000 lines of Coq nonblank, noncomment code.
+Their development is around 8,000 lines of nonblank, noncomment code.
 
@@
 from~\citet{decagda} to the decidability type checking of a
-bidirectional type system. Their development is larger and has around
+bidirectional type system. Their development has around
 30,000 lines of Coq code.
-% \scw{How big are these proofs?}
 
@@
 intersection types, union types, and quotient types. The extensive
-coverage of features result in a Coq development with around
+coverage of features results in a Coq development with around
 330,000 lines of code.
@@
 
-Compared to the systems we describe in this section, the most notable
-features we are missing include cumulativity and impredicativity. Our
-semantic model already satisfies the cumulativity property
-(Lemma~\ref{lemma:logrelcumulativity}), though we need to extend our
-convertibility relation into a subtyping relation in the specification
-of our syntactic typing rules. Impredicativity, on the other hand, is
-known to be difficult to model when the impredicative sort is at the
-bottom of a predicative universe hierarchy; in this scenario, the
-erasure technique from \citet{geuvers1994short} is no longer
-applicable\citep{abel2013normalization}. Whether we can find a
-similarly short and simple treatment for impredicativity remains an
-open question, which we hope to explore in the future.
+Compared to the systems described here, the most notable features we are
+missing are cumulativity and impredicativity. Our semantic model already
+satisfies the cumulativity property (Lemma~\ref{lemma:logrelcumulativity}),
+but we need to extend our convertibility relation into a subtyping relation in
+our syntactic typing rules. Impredicativity, on the other hand, is known to be
+difficult to model when the impredicative sort is at the bottom of a
+predicative universe hierarchy; in this scenario, the erasure technique from
+\citet{geuvers1994short} is not
+applicable\citep{abel2013normalization}. Whether there is a similarly short
+and simple treatment for impredicativity remains an open question.
 
@@
 
-The short consistency proof we present achieves the goal of
+Our short consistency proof achieves the goal of
 demonstrating the technique of proof by logical relation for dependently typed
 languages. However, what remains unanswered is what makes our development
-significantly shorter compared to others. Are we proving simpler results for
-smaller languages, are we making more use of automation, or is our proof
+significantly shorter. Are we proving simpler results for
+smaller languages, or making more use of automation, or is our proof
 technique genuinely more efficient?

### 2024-02-28 paper/paper.tex
Back to 25 pages

@@
 \begin{proof}
-  Immediate by lexicographical induction over the length of the reduction sequences in
+  Immediate by induction over the length of the reduction sequences in
   $[[wne a]]$ and $[[wn b]]$.
@@
   $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
-  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
+  conclude that $[[wn a]]$ and $[[wn A]]$.
 \end{proof}
@@
 the same rule in Coq, we must encode $[[F]]$ as a relation (with type
-\texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the following functionality constraint.
-\begin{minted}{coq}
-              forall a S0 S1, F a S0 -> F a S1 -> S0 = S1.
-\end{minted}
+\texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the functionality constraint:
+\mintinline{coq}{forall a S0 S1, F a S0 -> F a S1 -> S0 = S1}.
 In comparison, \rref{I-PiCoq} does not require this side condition and
@@
 In contrast, \citet{decagda} and \citet{nbeincoq} employ a deterministic weak
-head reduction relation. The use of a deterministic reduction relation makes
-the functionality of a logical relation trivial to prove, but the downside is
-that weak head reduction (denoted by $\leadsto^{wh}$) or leftmost-outermost
-reduction (denoted by $\leadsto^{lo}$) does not satisfy even the weaker
-substitution property (Lemma~\ref{lemma:parsubst}).  With this alternative
-formulation, we would have to prove how the deterministic reduction relates to
-a full non-deterministic reduction relation to prove the fundamental
-theorem. This would amount to proving the factorization theorem of the
-deterministic reduction relation with respect to full non-deterministic
-reduction. Instead, using parallel reduction in \rref{I-Red} allows us to
-delay the factorization property until we need to justify the use
-leftmost-outermost reduction as a deterministic reduction strategy for
-normalization.
+head reduction relation. % The use of a
+A deterministic reduction relation makes
+the functionality of a logical relation trivial to prove, but fails to
+satisfy the substitution property (Lemma~\ref{lemma:parsubst}), an
+issue that has been observed by
+\citet{casinghino:combining-proofs-programs}. If we had chosen to work
+with a deterministic reduction relation, we would likely need
+results such as the factorization theorem~\citep{takahashi-parallel-reduction,factorization-essentially} in our development before we
+can prove the fundamental theorem, leading to a more complicated proof.
+% With
+% this alternative formulation, we would have to prove how the deterministic reduction relates to
+% a full non-deterministic reduction relation to prove the fundamental
+% theorem. This would amount to proving the factorization theorem of the
+% deterministic reduction relation with respect to full non-deterministic
+% reduction. Instead, using parallel reduction in \rref{I-Red} allows us to
+% delay the factorization property until we need to justify the use
+% leftmost-outermost reduction as a deterministic reduction strategy for
+% normalization.
 

### 2024-02-28 paper/paper.tex
Fix linums

@@
     Logical relation         & 338 & 430 & - \\
-    Semantic soundness       & 190 & 214 & - \\
+    Semantic soundness       & 192 & 211 & - \\
     Syntactic soundness      & -   & -   & 629 \\
     \hline
-    Total                    & 955 & 1344 & 1056 \\
+    Total                    & 957 & 1341 & 1056 \\
     \hline
@@
 shows the number of non-blank, non-comment lines of code\footnote{calculated
-  by the \texttt{cloc} tool, available from
-  \url{https://github.com/AlDanial/cloc}.} for each file of our development,
+  by the \texttt{tokei} tool, available from
+  \url{https://github.com/XAMPPRocky/tokei}.} for each file of our development,
 including the base consistency proof from Section~\ref{sec:logreldep} and
@@
 The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
-1573 lines of non-blank, non-comment lines of code. We choose not to include
+1568 lines of non-blank, non-comment lines of code. We choose not to include
 it in the chart, because of slight differences in lemma dependencies for

### 2024-02-28 beta/count-beta.sh

@@ -1,0 +1,2 @@
+#!/usr/bin/env sh
+tokei -f semtypingopen.v soundnessopen.v typing.v join.v normalform.v

### 2024-02-28 beta/count-consistency.sh

@@ -1,0 +1,2 @@
+#!/usr/bin/env sh
+tokei -f semtyping.v soundness.v typing.v join.v

### 2024-02-28 eta/count-eta.sh

@@ -1,0 +1,2 @@
+#!/usr/bin/env sh
+tokei -f semtyping.v soundness.v typing.v join.v normalform.v

### 2024-02-28 paper/paper.tex
Minor
Add missing footnotes

@@
 We can recover the structural rules for $[[|= G]]$ as lemmas.
-\begin{lemma}[Semantic context well-formedness empty]
+\begin{lemma}[Semantic context well-formedness empty\footnote{\dotv{soundness.v}{SemWff\_nil}}]
   \label{lemma:semwffempty}
@@
 \end{lemma}
-\begin{lemma}[Semantic context well-formedness cons]
+\begin{lemma}[Semantic context well-formedness cons\footnote{\dotv{soundness.v}{SemWff\_cons}}]
   \label{lemma:semwffcons}
@@
 \begin{minted}{coq}
-Lemma par_triangle a : forall b, (a ⇒ b) -> (b ⇒ tstar a).
+Lemma Par_triangle a : forall b, (a ⇒ b) -> (b ⇒ tstar a).
 Proof.
-  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par. 
+  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,Par_cong ctrs:Par.
 Qed.

### 2024-02-29 paper/paper.tex
pass through 3 and 3.1

@@
 natural numbers strictly less than the parameter $[[i]]$, which
-represents the current universe level.  In \rref{I-Set}, function
+represents the current universe level.  In \rref{I-Set}, the function
 $[[I]]$ is used to define the meaning of
 universes that are strictly smaller than the current level $[[i]]$. The
-restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
-be predicative and we will explain the reason shortly.
-
-To tie the knot and obtain an interpretation of all universe levels,
-we define in Definition~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
-using the well-foundedness of $<$ relation on natural
-numbers (recall that
-the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
-numbers strictly less than $[[i]]$ as its input).
-The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
+restriction $[[j < i]]$ in \rref{I-Set} ensures the predicativity of the system.
+
+We tie the knot and obtain an interpretation for all universe levels
+below. The judgment $[[InterpR i A S]]$ reads that the type $[[A]]$ is a
 level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
 $[[S]]$. 
-
 \begin{definition}[Logical relation for all universe levels]
 \label{fig:logrelrec}
+Define $[[InterpR i A S]]$ recursively through the well-foundedness of the $<$
+relation on natural numbers.
 \begin{equation*}
@@
 interpretation of the $[[i]]_{th}$ universe is only dependent on
-universes strictly lower than $[[i]]$, which have been defined earlier.
-Removing the ordering constraint would result in a
-system where one can encode Girard's
-paradox~\citep{girard-thesis}. Our proof would no longer work with the
-restriction removed since the definition of $[[InterpR i A S]]$ would
-not be well-founded when $[[Interp I i A S]]$ calls $[[I]]$ on
+universes strictly lower than $[[i]]$, which have already been defined.
+This restriction ensures that that the relation is well-defined: without 
+it the definition of $[[InterpR i A S]]$ would
+not be well-founded; $[[Interp I i A S]]$ would call $[[I]]$ on
 universe levels greater than or equal to $[[i]]$, which are yet to be defined.
-
-By unfolding Definition~\ref{fig:logrelrec}, we
+%% SCW: removing the constraing wouldn't result in any system at all
+% Removing the ordering constraint would result in a
+% system where one can encode Girard's
+% paradox~\citep{girard-thesis}.
+
+By unfolding Definition~\ref{fig:logrelrec}, we can
 show that the same introduction rules for $[[Interp I i A S]]$ are
-admissible for $[[InterpR i A S]]$, by
-instantiating $[[I]]$ with $[[I i]] := [[{A | exists S , InterpR i A
-  S}]]$, the same function $[[I]]$ used in the definition of $[[InterpR i A
-S]]$. We give as example the following rules.
+admissible for $[[InterpR i A S]]$. For example, 
+we can prove the following derived rules:
 \begin{center}
@@
 the more explicit definition not only because it is directly definable
-in existing proof assistants where inductive definitions must appear
-at the top level, but also because it makes clear the induction
+in proof assistants that lack induction-recursion,
+% where inductive definitions must appear
+%at the top level,
+but also because it makes clear the induction
 principle we are allowed to use when reasoning about $[[InterpR i A
@@
 
-For the majority of the properties we are about to prove in this section, we
-do not need any information about the parameterized function $[[I]]$.
-Each property about $[[InterpR i A S]]$ follows as a corollary of
-a property about $[[Interp I i A S]]$ with no or few assumptions imposed on
-$[[I]]$. As a result, we usually state our lemmas in terms of
-$[[Interp I i A S]]$ without duplicating them in terms of $[[InterpR i
-A S]]$.
-
-% To derive consistency, it suffices to restrict $[[S]]$ to the set of
-% terms that reduce to closed terms.
+We next take a closer look at the inductive relation $[[Interp I i A S]]$, defined 
+in Figure~\ref{fig:logrel}.
 \Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
@@
 example, the $[[Void]]$ type should not have any inhabitants under the
-empty context, where as the $[[Bool]]$ type corresponds to terms that
-evaluate to $[[true]]$ or $[[false]]$. However, the characterization
-of $[[Void]]$ and $[[Bool]]$ in our logical relation does not always correspond to
+empty context, where as the $[[Bool]]$ type only contains terms that
+reduce to $[[true]]$ or $[[false]]$. Note that the characterization
+of $[[Bool]]$ (and other inhabited types) in our logical relation does not always correspond to
 well-typed or even closed terms. For example, the term $[[if false (#
@@
 says that an equality type $[[a ~ b : A]]$ corresponds to the
-set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ also holds and
+set of terms that reduce to $[[refl]]$ when $[[a <=> b]]$ also holds and
 otherwise corresponds to the empty set. Conditions like $[[a <=>
 b]]$ are typically required for indexed types, of which equality types
-are an instance. \Rref{I-Red} enables us to reduce our types in order
+are an instance. \Rref{I-Red} enables us to reduce types in order
 to assign meanings. Recall the type expression $[[(\ x . x )
-Bool]]$. \Rref{I-Red} says to know that
+Bool]]$. \Rref{I-Red} says that to know that
 $[[Interp I i  (\ x . x) Bool S ]]$ for some $[[S]]$, it suffices to
@@
 
-\Rref{I-Pi} is the most complicated rule in our logical
-relation. Instead of explaining it directly, we consider the following
-simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
+\Rref{I-Pi} is the most complex rule in our logical
+relation. Instead of explaining it directly, we first consider the following
+simplified version, \rref{I-PiAlt}, that follows directly from \rref{I-Pi}.
 \begin{center}
@@
 \end{center}
-\begin{lemma}[I-PiAlt admissibility]
+\Rref{I-PiAlt} directly captures the meaning of
+a well-behaved dependent function type. The precondition of
+the rule says that
+the function type $[[Pi x : A . B]]$ has an interpretation if its input
+type $[[A]]$ can be
+interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
+type $[[B {a / x}]]$, obtained by substituting $[[a]]$ into the output type
+$[[B]]$, has some semantic interpretation $[[S0]]$. In its conclusion, the 
+interpretation of $[[Pi x : A . B]]$ is the set of terms $[[b]]$, such that
+for all $[[a in S]]$, where $[[S]]$ is an interpretation of $[[A]]$, the
+application form $[[b a]]$ belongs to all possible interpretations of
+$[[B {a  / x}]]$ (the pre-condition ensures at least one interpretation
+exists for each $[[B {a / x}]]$ where $[[a in S]]$).
+
+\scw{Some researchers distinguish between derivable rules (those that 
+are simple corollaries), and admissible rules (those that must be proven 
+via induction.)}
+\begin{lemma}[I-PiAlt derivability]
   \label{lemma:piintroalt}
-  \Rref{I-PiAlt} is admissible.
+  \Rref{I-PiAlt} is derivable from \rref{I-Pi}.
 \end{lemma}
@@
 \end{proof} 
-\Rref{I-PiAlt} directly captures the meaning of
-a well-behaved dependent function type. The precondition of
-\rref{I-PiAlt} says that
-the function type $[[Pi x : A . B]]$ has an interpretation if its input
-type $[[A]]$ can be
-interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
-type $[[B {a / x}]]$, obtained by substituting $[[a]]$ into the output type
-$[[B]]$, has some semantic interpretation. In its conclusion, we say
-that a term $[[b]]$ belongs to an interpretation of $[[Pi x : A . B]]$ if
-for all $[[a in S]]$, where $[[S]]$ is an interpretation of $[[A]]$, the
-application form $[[b a]]$ belongs to all possible interpretations of
-$[[B {a  / x}]]$ (the pre-condition ensures at least one interpretation
-exists for each $[[B {a / x}]]$ where $[[a in S]]$).
-
 In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
@@
 inductive relation with \rref{I-PiAlt}, we opt for our current
-formulation of \rref{I-Pi} so we immediately obtain a
-well-defined inductive relation and a usable induction principle. The
+\rref{I-Pi} that immediately produces a
+well-defined inductive relation and usable induction principle. The
 slight disadvantage of \rref{I-Pi} is that we need to construct the
 function $[[F]]$ each time we apply it, though this is mitigated by
-the immediate admissibility of \rref{I-PiAlt} and the alternative $\Pi$ inversion
-principle (Lemma~\ref{lemma:piinvalt}), which we derive after proving functionality.
+the derivability of \rref{I-PiAlt} and the alternative $\Pi$ inversion
+principle (Lemma~\ref{lemma:piinvalt}).
 
@@
 property (Lemma~\ref{lemma:logrelbackclos}).
+For the majority of the properties that we prove in this section, we
+do not need any information about the parameterized function $[[I]]$.
+Each property about $[[InterpR i A S]]$ follows as a corollary of
+a property about $[[Interp I i A S]]$ with no or few assumptions imposed on
+$[[I]]$. As a result, we usually state our lemmas in terms of
+$[[Interp I i A S]]$ without duplicating them in terms of $[[InterpR i
+A S]]$.
 

### 2024-02-29 paper/paper.tex
edits in 3.2

@@
 
-\scw{Some researchers distinguish between derivable rules (those that 
-are simple corollaries), and admissible rules (those that must be proven 
-via induction.)}
 \begin{lemma}[I-PiAlt derivability]
@@
 \subsection{Properties about the Logical Relation}
-In the rest of this section, we prove four important facts about
-our logical relation: irrelevance (Lemma~\ref{lemma:logrelcoherence}),
-functionality (Lemma~\ref{lemma:logreldeter}), cumulativity
-(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure
-property (Lemma~\ref{lemma:logrelbackclos}).
-For the majority of the properties that we prove in this section, we
-do not need any information about the parameterized function $[[I]]$.
-Each property about $[[InterpR i A S]]$ follows as a corollary of
-a property about $[[Interp I i A S]]$ with no or few assumptions imposed on
-$[[I]]$. As a result, we usually state our lemmas in terms of
-$[[Interp I i A S]]$ without duplicating them in terms of $[[InterpR i
-A S]]$.
+In the rest of this section, we develop the theory of our logical relation
+with the goal of showing four key facts: irrelevance
+(Lemma~\ref{lemma:logrelcoherence}), functionality
+(Lemma~\ref{lemma:logreldeter}), cumulativity
+(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure property
+(Lemma~\ref{lemma:logrelbackclos}).  For the majority of the properties that
+we prove in this section, we do not need any information about the
+parameterized function $[[I]]$.  Each property about $[[InterpR i A S]]$
+follows as a corollary of a property about $[[Interp I i A S]]$ with no or few
+assumptions imposed on $[[I]]$. As a result, we usually state our lemmas in
+terms of $[[Interp I i A S]]$ without duplicating them in terms of
+$[[InterpR i A S]]$.
 
@@
 preservation holds in the usual forward direction too.
-\begin{lemma}[Preservation of the logical relation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
+\begin{lemma}[Forward preservation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
   \label{lemma:interppreservation}
@@
 \begin{proof}
-  By the definition of $[[InterpR i A S]]$ 
-  (Definition~\ref{fig:logrelrec}), we unfold $[[InterpR i A S]]$ by one
+  By the definition of $[[InterpR i A S]]$, we unfold $[[InterpR i A S]]$ by one
   step into $[[ Interp I i A S  ]]$ where $[[I j]] := [[{A | exists S

### 2024-02-29 paper/paper.tex
Minor

@@
   ideas from proofs about simpler languages to the dependently typed setting.
-\item We mechanize all our proofs using the Coq proof assistant, with 963
-  lines of code for the consistency proof and a moderate increase to 1357
+\item We mechanize all our proofs using the Coq proof assistant, with 957
+  lines of code for the consistency proof and a moderate increase to 1568
   lines of code for the normalization proof.  We discuss the specifics related
@@
 on the input variable. For simplicity, we omit the type annotations in the abstraction
-forms. We discuss how the inclusion of type annotations can affect our
+forms. We discuss how the inclusion of type annotations can affect
 our development in Section~\ref{sec:betaeta}, where we extend our

### 2024-10-10 paper/icfp.tex
Try to update for JFP

--- 2024-02-29 paper/paper.tex
@@
 information into the logical relation avoids extra bookkeeping and the
-need for a Kripke-style logical relation.
+need for a Kripke-style logical relation when we extend our logical
+relation to prove the existence of $\beta$-normal forms (Section~\ref{sec:extension}).
 \yl{Not sure what to cite from Derek Dreyer. I know his blog post
@@
 functional relation
-$[[{ (a , S0 ) | a in S /\ Interp I  i B { a /x } S0 }]]$. This result is
+$[[{ (a , S0 ) | a in S implies Interp I  i B { a /x } S0 }]]$. This result is
 formally shown through the improved inversion lemma for function types
@@
     $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
-    $[[F]]$ additionally satisfies the following properties:
+    $[[F]]$ additionally satisfy the following properties.
     \begin{itemize}
@@
 
-  The only interesting case is \Rref{I-Red}. Given that
+  The only interesting case is \rref{I-Red}. Given that
   $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
@@
   premise $[[Interp I i A S0]]$.
-  All cases that are not \rref{I-Red} follow immediately from the
-  Lemma~\ref{lemma:interpinv}.
+  All cases that are not \rref{I-Red} follow immediately from
+  Lemma~\ref{lemma:interpinv}, the inversion properties.
 
@@
 improved inversion lemma for function types whose statement is free of
-the relation $[[F]]$, analogous to the admissible \rref{I-PiAlt}.
+the relation $[[F]]$, analogous to the derivable \rref{I-PiAlt}.
 \begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
@@
   \item $[[S = { b | forall a, (# a in S0 implies forall
-    S1, (# Interp I i B {a / x} S1 /\  b a in S1 #) #) }]]$
+    S1, (# Interp I i B {a / x} S1 implies  b a in S1 #) #) }]]$
   \end{itemize}
@@
 reductions can be applied. This result can be used to show that our type
-conversion relation is decidable, as computation of the normal form will
-always terminate.
+conversion relation is decidable.% , as computation of the normal form will
+% always terminate.\yl{dependent on reduction strategy}
 
@@
        \beta\text{-}\mathit{neutral\ terms} &
-      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
+      [[e]] & ::= & [[x]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                     f]] \\ \\
@@
 
-The formulation of the valuation and semantic well-typedness from
-Section~\ref{sec:logrelproof} and the fundamental lemma remains
+The formulation of semantic well-typedness and the fundamental lemma from
+Section~\ref{sec:logrelproof} remains
 unchanged.
@@
 open terms~\citep{harperkripke} in the simply typed lambda calculus.
-Indeed, a mechanization of the latter proof appears in ~\citet{abel2019poplmark}.
-In this setting, as above, adequacy must be proven before the
-fundamental theorem so we can handle elimination rules such as
-\rref{T-If} and \rref{T-App} where the scrutinee is a neutral term. % In
+Indeed, a mechanization of normalization generalized to open terms
+appears in ~\citet{abel2019poplmark}.  In this setting, as above,
+adequacy must be proven before the
+fundamental theorem so they can handle elimination rules such as
+\rref{T-App} where the scrutinee is a neutral term. % In
 % \citet{abel2019poplmark}, a variant of Lemma~\ref{lemma:extwn} is used
@@
 As before, $\ottkw{ne}$ and $\ottkw{nf}$
-represent $\beta$-neutral and $\beta-$normal forms, and the
+represent $\beta$-neutral and $\beta$-normal forms, and the
 fundamental lemma shows us that every well-typed term has a
@@
 $\forall x, P(x) \iff Q(x)$. But this is exactly the statement of 
-functional extensionality.
+predicate extensionality, an immediate corollary from functional
+extensionality and propositional extensionality.
 
@@
 
-Finally, we note that we cannot ascribe $[[F]]$ the type \texttt{tm -> (tm ->
+We note that we cannot ascribe $[[F]]$ the type \texttt{tm -> (tm ->
   Prop)} since Coq requires functions of such type to be
@@
 of the derivations of parallel reduction. The option
-\texttt{use:Par\_refl,par\_cong} allows the automation tactic to use the
+\texttt{use:Par\_refl,Par\_cong} allows the automation tactic to use the
 reflexivity and congruence properties of parallel reduction as lemmas.
@@
 \citet{geuvers1994short} is not
-applicable\citep{abel2013normalization}. Whether there is a similarly short
+applicable~\citep{abel2013normalization}. Whether there is a similarly short
 and simple treatment for impredicativity remains an open question.

### 2024-10-11 paper/paper.tex
Fixed some font/symbol stuff

@@
 %%
-\documentclass[acmsmall,screen=true,
-\ifpublic review=false\else,review=true\fi
-  ,anonymous=\ifanonymous true\else false\fi]{acmart}
+\documentclass[\ifpublic nolinenum\else\fi]{jfp}
+
+\let\Bbbk\relax
 \usepackage{ottalt}
-\usepackage{minted}
+\usepackage[outputdir=latex.out]{minted}
 \usepackage{xspace}
+\usepackage{stmaryrd}
+\usepackage{bbding}
+\usepackage{amsthm,amssymb}
+\usepackage{dejavu}
 \usepackage[para]{footmisc}
@@
 
-%% to allow unicode characters in minted code
-\usepackage{newunicodechar}
-\newunicodechar{⇒}{$\Rightarrow$}
-
+\let\draftnote\relax % this is already defined in JFP
 \usepackage{draft}
@@
 
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+
+\DeclareTextCommand{\nobreakspace}{TU}{\leavevmode\nobreak\ }
+
 %%
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}
+\title[T]{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}
 
@@
 \author{Yiyun Liu}
-\orcid{0009-0006-8717-2498}
-\affiliation{
-  \institution{University of Pennsylvania}
-  \city{Philadelphia}
-  \country{USA}
-}
-\email{liuyiyun@seas.upenn.edu}
+% \orcid{0009-0006-8717-2498}
+\affiliation{University of Pennsylvania (\email{liuyiyun@seas.upenn.edu})}
 
 \author{Stephanie Weirich}
-\orcid{0000-0002-6756-9168}
-\affiliation{
-  \institution{University of Pennsylvania}
-  \city{Philadelphia}
-  \country{USA}
-}
-\email{sweirich@seas.upenn.edu}
+% \orcid{0000-0002-6756-9168}
+\affiliation{University of Pennsylvania (\email{sweirich@seas.upenn.edu})}
+
+\author{Jonathan Chan}
+\affiliation{University of Pennsylvania (\email{jcxz@seas.upenn.edu})}
 
@@
 consistency and parametricity. While there exists a
-plethora of introductory materials about logical relation in the
-context of simply typed or polymorphic lambda calculus, a streamlined
+plethora of introductory materials about logical relations in the
+context of simply typed or polymorphic lambda calculi, a streamlined
 presentation of proof by logical relation for a dependently typed language
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
 \[
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
 \begin{minipage}{0.9\textwidth}
@@
 \subsection{Definition of the Logical Relation}
-\begin{figure}[h]
+\begin{figure}[H]
 \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
@@
 
-\section{Existence of $\beta$-normal forms}
+\section{Existence of \texorpdfstring{$\beta$}{beta}-normal forms}
 \label{sec:extension}
@@
 %
-\begin{figure}[h]
+\begin{figure}[H]
   \[
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
@@
 
-\section{Existence of $\beta\eta$-normal forms}
+\section{Existence of \texorpdfstring{$\beta\eta$-}{beta-eta--}normal forms}
 \label{sec:betaeta}
@@
 \label{sec:logrelmech}
-\begin{figure}[h]
+\begin{figure}[H]
   \begin{minipage}{0.9\textwidth}
@@
 reduction.
-\begin{minted}{coq}
-Lemma Par_triangle a : forall b, (a ⇒ b) -> (b ⇒ tstar a).
+\begin{minted}[escapeinside=||,mathescape=true]{coq}
+Lemma Par_triangle a : forall b, (a |$\Rightarrow$| b) -> (b |$\Rightarrow$| tstar a).
 Proof.
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
 
@@
     \hline
-    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} & 1 
+    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & $\checkmark$ & 1 
     & Consistency and normalization \\
     % \hline
-    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & 1 & Consistency \\
-    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & 2 & Consistency \\
-    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & 2 & Correctness of NBE \\
-    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & {\boxedsymbols ✓} & 2 & Decidability of conversion \\
-    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & {\boxedsymbols ✓} & 2 & Decidibility of type checking \\
+    $\lambda^\theta$ & 0 & Id, Nat & U & $\XSolidBrush$ & 1 & Consistency \\
+    Core Nuprl & $\mathbb{N}$ & W-Types & E & $\checkmark$ & 2 & Consistency \\
+    NBE-in-Coq & 1 & Nat & T & $\XSolidBrush$ & 2 & Correctness of NBE \\
+    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidability of conversion \\
+    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidibility of type checking \\
     \hline
@@
   \underline{C}onversion:& Untyped (U), Typed (T),  Extensional (E) \\
-  \underline{L}arge \underline{E}liminations:& Included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
+  \underline{L}arge \underline{E}liminations:& Included ($\checkmark$), not included ($\XSolidBrush$) \\
   \underline{A}rity of interpretation: & Sets of terms (1), Relations between terms (2) \\

### 2024-10-11 paper/paper.tex
It compiles!!

@@
 %%
-\documentclass[\ifpublic nolinenum\else\fi]{jfp}
+\documentclass[\ifpublic nolinenum\else\fi,online,OA]{jfp}
 
@@
 
+\journaltitle{JFP}
+\cpr{The Author(s),}
+\doival{10.1017/}
+\totalpg{\pageref{lastpage}}
+\jnlDoiYr{2025}
+\lefttitle{Y. Liu et al.}
+\righttitle{Logical Relations for Type Theory}
+
 %%
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title[T]{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}
+\title{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}
 
@@
 %% used to denote shared contribution to the research.
+\begin{authgrp}
 \author{Yiyun Liu}
 % \orcid{0009-0006-8717-2498}
-\affiliation{University of Pennsylvania (\email{liuyiyun@seas.upenn.edu})}
+\affiliation{University of Pennsylvania (\email{\elink{liuyiyun@seas.upenn.edu}})}
 
@@
 % \orcid{0000-0002-6756-9168}
-\affiliation{University of Pennsylvania (\email{sweirich@seas.upenn.edu})}
+\affiliation{University of Pennsylvania (\email{\elink{sweirich@seas.upenn.edu}})}
 
 \author{Jonathan Chan}
-\affiliation{University of Pennsylvania (\email{jcxz@seas.upenn.edu})}
+\affiliation{University of Pennsylvania (\email{\elink{jcxz@seas.upenn.edu}})}
+\end{authgrp}
 
@@
 %%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Logical Relation, Dependent Types, Logical Consistency, Coq}
-
-%%
 %% This command processes the author and affiliation and title
 %% information and builds the first part of the formatted document.
-\maketitle
+\maketitle[T]
 

### 2024-11-01 paper/paper.tex
Clean up comments
Minor
Pass through introduction

@@
 
-%%
-%% This is file `sample-acmsmall.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `acmsmall')
-%%
-%% IMPORTANT NOTICE:
-%%
-%% For the copyright see the source file.
-%%
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-acmsmall.tex.
-%%
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%%
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
 \documentclass[\ifpublic nolinenum\else\fi,online,OA]{jfp}
 
-\let\Bbbk\relax
-\usepackage{ottalt}
-\usepackage[outputdir=latex.out]{minted}
-\usepackage{xspace}
-\usepackage{stmaryrd}
-\usepackage{bbding}
+\let\Bbbk\relax % amsmath tries to redefine this
 \usepackage{amsthm,amssymb}
-\usepackage{dejavu}
-\usepackage[para]{footmisc}
-\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
-\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
-\inputott{rules}
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
 \let\draftnote\relax % this is already defined in JFP
 \usepackage{draft}
+\usepackage[outputdir=latex.out]{minted} % latexrun puts everything in latex.out
+\usepackage{xspace,stmaryrd,bbding}
+\usepackage[para]{footmisc}
+\usepackage{ottalt}
+\inputott{rules}
 
@@
 \newnote{yl}{purple} % Yiyun Liu
+\newnote{jc}{darkgreen}
 \else
@@
 \newcommand{\yl}[1]{}
+\newcommand{\jc}[1]{}
 \fi
+
+\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
+\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
 
@@
 
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
 \citestyle{acmauthoryear}
 
-%%
-%% end of the preamble, start of the body of the document source.
 \begin{document}
@@
 \journaltitle{JFP}
-\cpr{The Author(s),}
+\volno{35}
+\edno{0}
+\cpr{Cambridge University Press}
 \doival{10.1017/}
@@
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
 \title{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
 \begin{authgrp}
@@
 
+% \author{Jonathan Chan}
+% \orcid{0000-0003-0830-3180}
+% \affiliation{University of Pennsylvania (\email{\elink{jcxz@seas.upenn.edu}})}
+
 \author{Stephanie Weirich}
@@
 \affiliation{University of Pennsylvania (\email{\elink{sweirich@seas.upenn.edu}})}
-
-\author{Jonathan Chan}
-\affiliation{University of Pennsylvania (\email{\elink{jcxz@seas.upenn.edu}})}
 \end{authgrp}
 
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-% \renewcommand{\shortauthors}{Trovato et al.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
@@
 
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
 \maketitle[T]
@@
 This paper presents a \emph{short} and \emph{mechanized} proof of
-logical consistency for \lang{}, a dependent type theory with a full
-predicative universe hierarchy, large eliminations, an intensional identity
+logical consistency for \lang{}, a dependent type theory with a
+predicative universe hierarchy, large elimination, an intensional identity
 type, a boolean base type, and dependent elimination forms.
@@
   pierce2002types, pierce2004advanced,harper2016practical} is primarily
-focused on systems with simple or polymorphic types. In that context,
+focused on simple or polymorphic types. In that context,
 syntactic logical relations can be defined as simple recursive functions over
-the structure of types, or (in the case of recursive types) defined over the
-evaluation steps of the computation. Yet, neither of these techniques can be
+the structure of types, or in the case of recursive types, defined over the
+evaluation steps of the computation. Yet neither of these techniques can be
 used to define a logical relation in the context of a predicative dependent
-type theory, so a novice researcher might be excused for thinking that proofs
-that use logical relations are not applicable for such languages.
+type theory, so a novice researcher may have the impression
+that logical relations are not applicable to dependent types.
 
@@
 assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
-have relied on logical relations defined as part of their
+rely on logical relations as part of their
 developments. However, because these proofs show diverse results about real
@@
 
-Thus, our paper provides a gentle and accessible introduction to a powerful
-technique for dependent type theories. To promote the use of machine-assisted
-reasoning, our development is accompanied by a short mechanized proof script,
-of less than 1000 lines of code, developed using the Coq proof
-assistant~\cite{coq}.
-
+Our paper instead provides a gentle and accessible introduction to a powerful
+technique for dependent type theories.
+It is accompanied by a short mechanized proof development
+of fewer than 1000 lines of code, developed using the Coq proof
+assistant~\citep{coq}.
 We have streamlined our proof through a number of means: the careful selection
 of the features that we include in the object type system and the results that
-we prove about it, in addition to the judicious use of automation.  Our
-language is small, but includes enough to be illustrative. For example, we
-eschew inductive datatypes or W-types, but we do include propositional
+we prove about it, in addition to the judicious use of automation.
+
+Our language is small, but includes enough to be illustrative. For example, we
+eschew inductive datatypes and W types, but we do include propositional
 equality and booleans to capture the challenges presented by indexed types and
@@
 nor do we develop a PER semantics, but we prove logical consistency, which
-states that empty types are not inhabited in an empty context, and demonstrate
-how our consistency proof can be extended (at a moderate cost of 500 lines of
-code) to show the existence of $\beta\eta$-normal forms for well-typed open
-\emph{and} closed terms. We include a full predicative universe hierarchy and
-type-level computation to demonstrate the logical strength of the approach.
+states that empty types are not inhabited in an empty context,
+and extend our consistency proof to show the existence of $\beta\eta$-normal
+forms for well-typed closed \emph{and} open terms,
+at a moderate cost of 600 lines of code.
+We include a full predicative universe hierarchy with large elimination
+to demonstrate the logical strength of the approach.
+% JC: I changed "type-level computation" to "large elimination"
+% since I think it's usually the latter that gets referred to
+% when talking about logical strength and expressivity.
 
@@
 
-More concretely, our paper makes the following contributions.
+Concretely, our paper makes the following contributions.
 \begin{itemize}
@@
 \item In Section~\ref{sec:logreldep}, we formulate logical consistency for
-  \lang{}, the property of interest, to motivate a logical relation. We define
-  the relation first inductively and then later prove that it is a partial
-  function. Based on this definition, we define semantic typing and prove the
+  \lang{}, the property of interest, to motivate a logical relation. We
+  define the relation inductively, prove its functionality,
+  define semantic typing in terms of the relation, and prove the
   fundamental theorem, from which consistency follows as a corollary
-  (Section~\ref{sec:logrelproof}). Thanks to the design of \lang{}, our proof
-  showcases the special treatment required to model many of the most common
+  (Section~\ref{sec:logrelproof}). Our proof
+  showcases the special treatment required to model many common
   features of dependent type theories, thus making our proof applicable to a
@@
   modifications made to our initial logical relation are small and closely
-  mirror the necessary extensions for a simply-typed language. We use this
-  part to show that once we have established the base techniques, we can port
+  mirror the necessary extensions for a simply typed language. This shows
+  that once we have established the base techniques, we can port
   ideas from proofs about simpler languages to the dependently typed setting.
-\item We mechanize all our proofs using the Coq proof assistant, with 957
-  lines of code for the consistency proof and a moderate increase to 1568
-  lines of code for the normalization proof.  We discuss the specifics related
-  to our choice of Coq as our metatheory, including our use of off-the-shelf
+\item We mechanize all our proofs in Coq, with 983
+  lines of code for the consistency proof with $\beta$ rules and a moderate
+  increase to 1596 lines of code for the normalization proof with $\beta\eta$
+  rules. We discuss our choice of using Coq as our metatheory,
+  including the use of off-the-shelf
   semantic engineering infrastructure and automation tools, in
-  Section~\ref{sec:logrelmech}.  Our proof scripts, with comments, are
-  available to reviewers as supplementary materials.
+  Section~\ref{sec:logrelmech}. Our proof development is available as
+  supplementary material.
 \item We compare our work to existing proofs by logical relations and other
   proof techniques for proving consistency and normalization.  We provide an
-  overview of this prior work (Section~\ref{sec:relatedwork}) and also give an
-  in-depth explanation on how various design decisions affect the size of our
+  overview of prior work (Section~\ref{sec:relatedwork}) and explain
+  how various design decisions affect the size of our
   proof and its extensibility to additional features
@@
 
-The result of our work is an artifact that an interested researcher can
-navigate and understand. We accompany this short mechanized proof with an
-informal description, presented here using set-theoretic notation and
-terminology so that it is accessible to readers with a general mathematical
-background.  That said, our explanations do not stray too far away from our
-proof scripts.  We link each lemma directly to its counterpart in the proof
-script, anticipating that readers may wish to see how these results may be
-expressed and verified in a proof assistant.  The typeset proofs purposefully
-follow the mechanized proofs closely while avoiding, as much as possible,
-artifacts specific to Coq.
-
-Not only does this close connection aid readers that wish to, like us, adopt
-proof assistants for their day-to-day use, but we also find that this
-precision is important for conveying the proof technique itself. Unlike
-properties that are derivable through syntactic means, proofs by logical
-relations make demands on the strength of the metalogic in which they are
-expressed. An informal proof that attempts to be agnostic or ambiguous about
-the underlying metatheory requires substantial effort from the reader to
-understand whether it is definable in a given ambient logic.
+The result of our work is an artifact targeted towards researchers
+familiar with the syntax of dependent types, as well as logical relations
+for simple or polymorphic types, and who wish to understand using logical
+relations \emph{for} dependent types.
+The short mechanized proof is accompanied here by a pen-and-paper description
+using set-theoretic notation and terminology so that it is accessible to
+readers with a general mathematical background. This pen-and-paper proof
+purposefully follows the mechanized proof closely while avoiding Coq-specific
+details as much as possible, and lemmas are linked directly to their
+counterparts in the mechanization.
+
+Not only does this close connection aid readers who wish to adopt proof
+assistants for mechanizing metatheory, this precision is also important for
+conveying the proof technique itself. Unlike properties derivable through
+purely syntactic means, proofs by logical relations make demands on the
+strength of the metatheory in which they are expressed. An informal proof that
+attempts to be agnostic or ambiguous about the underlying metatheory requires
+substantial effort from the reader to understand whether it is definable in a
+given ambient logic.
 
@@
 
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
 \bibliographystyle{ACM-Reference-Format}
@@
 
-%%
-%% If your work has an appendix, this is the place to put it.
+\label{lastpage}
 
 \end{document}
-\endinput
-%%
-%% End of file `sample-acmsmall.tex'.

### 2024-11-01 paper/paper.tex
Remove endpoints from J eliminator in paper

@@
                   & \mbox{function types, abstractions, applications} \\
-            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
+            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c p]]
                   & \mbox{equality types, reflexivity proof, J eliminator} \\
@@
 $[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
-$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
+$[[J c p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
 $[[b]]$, and $[[c]]$ is the term whose type is to be casted.  Finally, \lang{}
@@
        \beta\text{-}\mathit{neutral\ terms} &
-      [[e]] & ::= & [[x]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
+      [[e]] & ::= & [[x]]\ |\ [[e f]]\ |\ [[J e f]]\ |\ [[if e f
                     f]] \\ \\

### 2024-11-05 paper/paper.tex
Tighten intro paragraphs in Section 2; add an absurd eliminator (still need to add to the proofs!)

@@
 \usepackage[outputdir=latex.out]{minted} % latexrun puts everything in latex.out
-\usepackage{xspace,stmaryrd,bbding}
+\usepackage{xspace,stmaryrd,mathtools,bbding}
 \usepackage[para]{footmisc}
@@
 
-\begin{figure}[H]
-\[
-\begin{array}{lcll}
-% \mathit{Natural\ numbers}\\
-% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
-\mathit{Terms}\\
-[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[x]]\  |\ [[Void]]
-                  & \mbox{universes, variables, empty type} \\
-            & |   & [[Pi x : A . B]]\ |\ [[\ x . a]]\ |\ [[a b]]
-                  & \mbox{function types, abstractions, applications} \\
-            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c p]]
-                  & \mbox{equality types, reflexivity proof, J eliminator} \\
-            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
-                  & \mbox{boolean type, true, false} \\
-            & |   & [[if a b0 b1]]
-                  & \mbox{conditional expression} \\ \\
-\end{array}
-\]
-\[
-\begin{array}{lll}
-\mathit{Substitutions}      & \qquad\qquad & \mbox{\emph{Typing Contexts}}  \\
-[[rho]] \in [[SVar -> STm]] & & [[G]] ::= [[empty]]\ |\ [[G ++ x : A]] 
-\end{array}
-\]
+\begin{figure}[h]
+\begin{align*}
+[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & \Coloneqq && \textit{Terms} \\
+            & \mid [[Set i]] \mid [[x]] \mid [[Void]] \mid [[absurd b]]
+              && \textrm{universes, variables, empty type, explosion} \\
+            & \mid [[Pi x : A . B]] \mid [[\ x . a]] \mid [[a b]]
+              && \textrm{function types, abstractions, applications} \\
+            & \mid [[a ~ b : A ]] \mid [[refl]] \mid [[J c p]]
+              && \textrm{equality types, reflexivity proof, J eliminator} \\
+            & \mid [[Bool]] \mid [[true]] \mid [[false]] \mid [[if a b0 b1]]
+              && \textrm{boolean type, true, false, conditionals} \\
+            % & \mid [[if a b0 b1]]
+            %   && \textrm{conditional expression} \\
+[[G]] & \Coloneqq [[empty]] \mid [[G ++ x : A]] && \textit{Typing contexts} \\
+[[rho]] & \in [[SVar -> STm]] && \textit{Substitutions}
+\end{align*}
 \caption{Syntax of \lang}
@@
 
-In this section, we present the dependent type theory, \lang{}, whose logical
+In this section, we present the dependent type theory \lang{}, whose logical
 consistency will be proven in Section~\ref{sec:logrelproof}.
-
-\scw{Need to acknowledge the lack of type annotation on abstractions.}
 
@@
 type theory, terms and types are collapsed into the same syntactic
-category. The type $[[Set i]]$ represent a universe type, annotated by its
-universe level, a natural number $i$.  Abstractions $[[ \ x . a ]]$ and
+category. We use a Curry-style (extrinsic) syntax, where terms are not
+annotated by types, and only terms which participate in reduction are part of
+introduction and elimination forms.
+
+The type $[[Set i]]$ represents a universe at level $i$,
+a natural number. Abstractions $[[ \ x . a ]]$ and
 dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
-$x$ in the body of the function and codomain of the function type.
+$x$ in the body of the function and codomain of the function type.%
 \footnote{In the exposition in this paper, binding forms are equal up to
@@
   treatment of variables and substitution; our mechanized proofs make these
-  notions precise by using de Bruijn indices~\cite{debruijn1994automath}. }
+  notions precise by using de Bruijn indices~\citep{debruijn1994automath}. }
 We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
-on the input variable. For simplicity, we omit the type annotations in the abstraction
-forms. We discuss how the inclusion of type annotations can affect
+on the input variable. Type annotations in abstraction forms are omitted, and
+we discuss how the inclusion of type annotations can affect
 our development in Section~\ref{sec:betaeta}, where we extend our
 consistency result to the existence of $\beta\eta$-normal forms.
-We include in \lang{} the intensional identity type
-$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
-$[[J c p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
-$[[b]]$, and $[[c]]$ is the term whose type is to be casted.  Finally, \lang{}
-includes booleans, with standard syntax.  \scw{Example of what you can use
-  identity types for? Or an example of a program that uses J?}
+Intensional identity types $[[a ~ b : A]]$, or equality types, represent an
+equality between terms $[[a]]$ and $[[b]]$ of the same type $[[A]]$, whose
+canonical inhabitant is the reflexive proof $[[refl]]$. Equality proofs $[[p]]$
+can be eliminated by the J eliminator $[[J c p]]$, which casts the term $[[c]]$
+from one end of the equality to the other. Finally, we have booleans along with
+the standard boolean values and conditional expressions.
 
 Our reduction and typing relations are defined in terms of \emph{simultaneous
-  substitutions}, $[[rho]]$, which are mappings from variables to terms.  We
-use $[[idtm]]$ as the identity substitution.  The extension operation,
-$[[(rho .: x -> a)]]$, updates the substitution $[[rho]]$ to map the variable
+substitutions} $[[rho]]$, which are mappings from variables to terms. We
+use $[[idtm]]$ as the identity substitution, and the extension operator
+$[[rho .: x -> a]]$ updates the substitution $[[rho]]$ to map the variable
 $[[x]]$ to $[[a]]$ rather than $[[rho x]]$.
-
-The substitution operator, which takes the form $[[a {rho}]]$, traverses the
-syntax of $[[a]]$ and replaces each variable $[[x]]$ with the term
-$[[rho x]]$. When traversing under binders (e.g. in the
-$[[(\ x . a) { rho }]]$ case), it must be the case that $[[rho]]$ maps the
-bound variable to itself and that the bound variable does not appear freely in
-the application of the substitution to any other variable.
+% When traversing under binders (e.g. in the
+% $[[(\ x . a) { rho }]]$ case), it must be the case that $[[rho]]$ maps the
+% bound variable to itself and that the bound variable does not appear freely in
+% the application of the substitution to any other variable.
 %
@@
   watertight and rigorous, though we have one in our mechanization}
-
-The substitution operator is referred to as simultaneous substitution as it
-substitutes for all variables at once. It is possible to recover single
-substitution by composing the extension operator and the identity
-substitution: $[[a { b / x }]] := [[a { idtm .: x -> b }]]$.
-
-When reasoning about logical relations, we find it more convenient to
-formulate simultaneous substitution directly rather than recovering it from
-single substitution.  In particular, this shows up in the definition of
-semantic typing in Section~\ref{sec:logrelproof}, which relies on simultaneous
-substitution.
+The substitution operator $[[a {rho}]]$ replaces every free variable $[[x]]$ of
+$[[a]]$ by $[[rho x]]$ simultaneously. A single substitution corresponds to
+composing extension and identity:
+$[[a { b / x }]] \coloneqq [[a { idtm .: x -> b }]]$.
+
+When reasoning about logical relations, it is more convenient to formulate
+definitions in terms of simultaneous substitutions than single substitutions,
+such as semantic typing in Section~\ref{sec:logrelproof}.
+\jc{I've rewritten this paragraph to be more succinct,
+but I'm still not certain it's necessary at all.
+We're not focussing on the merits of simultaneous substitution over
+other representations of substitution and binding, after all.}
 
@@
 \begin{figure}[H]
-\begin{minipage}{0.9\textwidth}
 \drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
-% \drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
-% \drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
-\end{minipage}
-\caption{Parallel reduction ($\beta$-rules only) }
+\caption{Parallel reduction ($\beta$ rules only) }
 \label{fig:par}

### 2024-11-06 paper/paper.tex
Rewording a bit of Section 2

@@
 
-\commentsfalse        %% toggle comments here
+\commentstrue         %% toggle comments here
 \extendedfalse
-\anonymoustrue
+\anonymousfalse
 
@@
 \newnote{yl}{purple} % Yiyun Liu
-\newnote{jc}{darkgreen}
+\newnote{jc}{red} % Jonathan Chan
 \else
@@
 
-\begin{figure}[H]
-\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
-\caption{Parallel reduction ($\beta$ rules only) }
-\label{fig:par}
-\end{figure}
-
 Before we specify the typing rules, we first specify the equational theory
-used in the conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). The
-equivalence relation used in this rule is often referred to
-\textit{definitional equality} in dependent type theories because it defines
-the equivalence that the syntactic type system works up to.
-
-In \lang{}, we use a relation called \emph{convertibility} for definitional
-equality. Two terms are convertible, if they reduce to a common form. The
-reduction that we use is called \emph{parallel reduction}, written
-$[[a => b]]$. The notation $[[a =>+ b]]$ indicates the transitive closure of
-parallel reduction.
+used in the conversion \rref{T-Conv} in Figure~\ref{fig:typing},
+known as \textit{definitional equality} in dependent type theories because it
+defines the equivalence that the syntactic type system works up to. 
+The definitional equality we use is \emph{convertibility}:
+two terms are convertible if they reduce to a common form.
+The reduction that we use is \emph{parallel reduction}, written
+$[[a => b]]$, with $[[a =>+ b]]$ indicating its reflexive, transitive closure.
+The parallel reduction relation is defined in Figure~\ref{fig:par},
+which omits reflexivity and congruence rules for brevity.
 
@@
 
-The definition of the parallel reduction relation, appears in
-Figure~\ref{fig:par}. (For brevity, the reflexivity and congruence rules of
-this relation are omitted from this figure).
+\begin{figure}
+\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
+\caption{Parallel reduction ($\beta$ rules only) }
+\label{fig:par}
+\end{figure}
 
 We prove, through standard
-techniques~\citet{takahashi-parallel-reduction,plfa22.08}, the following
+techniques~\citep{takahashi-parallel-reduction,plfa22.08}, the following
 properties of parallel reduction.
-\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
+
+\begin{lemma}[Reflexivity (parallel reduction)\footnote{\dotv{join.v}{Par\_refl}}]
   \label{lemma:parrefl}
@@
 \end{lemma}
-\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
+\begin{lemma}[Congruence (p.r.)\footnote{\dotv{join.v}{par\_cong}}]
   \label{lemma:parcong}
@@
 \end{lemma}
-\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
+\begin{corollary}[Substitution (p.r.)\footnote{\dotv{join.v}{par\_subst}}]
   \label{lemma:parsubst}
@@
 \end{corollary}
-\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
+\begin{lemma}[Diamond property\footnote{\dotv{join.v}{par\_confluent}}]
   \label{lemma:pardiamond}
@@
 
-\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
+\begin{lemma}[Reflexivity (convertibility)\footnote{\dotv{join.v}{Coherent\_reflexive}}]
   \label{lemma:coherencerefl}
@@
 \end{lemma}
-\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
+\begin{lemma}[Symmetry (convertibility)\footnote{\dotv{join.v}{Coherent\_symmetric}}]
   \label{lemma:coherencesym}
@@
 \end{lemma}
-\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
+\begin{lemma}[Transitivity (convertibility)\footnote{\dotv{join.v}{Coherent\_transitive}}]
   \label{lemma:coherencetrans}
@@
 
-The convertibility relation that we use for conversion in \lang{} is unusual
+The convertibility relation that we use for definitional equality is unusual
 in that it is directly defined via parallel reduction, instead of using the
-related notion of $\beta$-equivalence~\citep{barendregt1991introduction,coquand1990:cic}. This choice
-does not change the language definition; a detailed argument of the
+related notion of $\beta$-equivalence~\citep{barendregt1991introduction,coquand1990:cic}.
+This choice does not change the language definition; a detailed argument of the
 equivalence between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found
@@
 
-Our definition of equality is untyped: the judgement does not require the two
+Our definitional equality is untyped: the judgment does not require the two
 terms to type check and have the same type. The use of an untyped relation for
@@
 Systems~\cite{barendregt1991introduction} and differs
-from MLTT~\citep{Martin-Lof-1973}, where the judgmental
+from MLTT~\citep{Martin-Lof-1973}, where definitional
 equality takes the form $\Gamma \vdash a \equiv b : A$.
@@
 % reduction.
-By working with an untyped judgement, we can establish its
-properties independently from the type system and the logical relation, using
+By working with an untyped judgment, we can establish its
+properties independently of the type system and the logical relation, using
 well-established syntactic approaches.
@@
 System, which employs untyped equality, and
-its variant that uses typed judgmental equality. This assures us that we do
+its variant that uses typed equality. This assures us that we do
 not lose generality working with a system with untyped
 conversion.
+\jc{I'm not too fond of this last sentence because we \emph{do} lose
+the ability to add typed $\eta$ equivalences.}
 \yl{removed the eta equivalence sentence because it might give the
@@
 
-\begin{figure}[H]
-\begin{minipage}{0.9\textwidth}
-\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{}
-\[ \drule{Ctx-Empty} \qquad \drule[width=3in]{Ctx-Cons} \]
-\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
- Void, Bool, True, False}
-\[ \drule[width=5in]{T-If}\]
-\[ \drule[width=3in]{T-Eq} \qquad \drule{T-Refl} \]
-\[ \drule[width=4in]{T-J} \]
-\end{minipage}
-\caption{Syntactic typing for \lang}
+\begin{figure}
+\begin{mathpar}
+  \fbox{$[[ |- G]]$} \hfill \textit{(Context well-formedness)} \\
+  \drule{Ctx-Empty} \drule[width=3in]{Ctx-Cons}
+\end{mathpar}
+\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Void, Absurd, Bool, True, False}
+\begin{mathpar}
+  \drule[width=3in]{T-If} \drule[width=2in]{T-Eq} \\
+  \drule{T-Refl} \drule[width=4in]{T-J}
+\end{mathpar}
+\caption{Syntactic typing}
 \label{fig:typing}
@@
 
-Figure~\ref{fig:typing} gives the full typing rules for \lang{}. The premises
-wrapped in \colorbox{lightgray}{gray} boxes can be shown to be admissible
-syntactically, though some of them are required to strengthen the inductive
+Figure~\ref{fig:typing} gives the complete typing rules. The premises
+highlighted in \colorbox{lightgray}{gray} can be shown to be admissible
+syntactically, though they are required to strengthen the inductive
 hypothesis of the fundamental theorem.
 
-The typing rules of \lang{} are standard for dependent
+These rules are standard for dependent
 type theories.  The variable rule, \rref{T-Var}, uses the auxiliary relation
-$[[x : A in G]]$, that holds when a variable declaration is found in the
-typing context.  The typing of universes ensures that each one belongs to the
+$[[x : A in G]]$ that holds when a variable declaration is found in the
+typing context. \Rref{T-Set} ensures that each universe belongs to the
 next higher level. \Rref{T-Pi} ensures predicative quantification by
-requiring that all parts of the type be typeable at the same universe
+requiring that the domain and codomain types be typeable at the same universe
 level. \Rref{T-Abs} ensures that all functions have well-formed dependent
-types. In an application (\rref{T-App}) the argument is substituted for the
+types. In the application \rref{T-App}, the argument is substituted for the
 variable in the result type.
-
 \Rref{T-Conv} uses the convertibility relation from earlier
@@
 \scw{Cassia did not follow our discussion of dependent pattern matching. We need to expand}
-The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
-matching.  The result type of this expression, $[[A {a / x}]]$, is composed of some
-\emph{motive} $[[A]]$, a type where its single free variable has been replaced
-with the condition of the if expression. When typing the true branch, this
-substitution replaces the variable by $[[true]]$, and similarly for the false
-branch. As a result, the type system communicates the information gained from
-the test to each of the branches of the expression.
-
-A similar sort of dependent pattern matching occurs when eliminating identity
-types. Such types are checked for well-formedness with \rref{T-Eq} and
-introduced by \rref{T-Refl}. In \rref{T-J}, the elimination form, the subterm
-$p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
-is the body of the elimination form. In this rule, $[[B]]$ is the motive and
-has two free variables.  When checking $[[c]]$, the substitution for these
-variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
-witnessing the information gained through dependent pattern matching.
-
-The universe hierarchy and the boolean base type gives \lang{} the
-ability to compute a type using a term as input, a feature commonly referred to
-as large elimination. For
-example, we can type check the function $[[\ x . if x Bool Void]]$,
-which returns either $[[Bool]]$ or $[[Void]]$ depending on whether its
-input is $[[true]]$ or $[[false]]$.
+The elimination form for booleans in \rref{T-If} demonstrates dependent pattern
+matching: the type of the expression \emph{depends} on the term being matched.
+This result type involves a \emph{motive} $[[A]]$ abstracted over a boolean.
+For the overall expression, $[[x]]$ is filled in by the match target $[[a]]$,
+while in the $[[true]]$ (resp. $[[false]]$) branch, $[[x]]$ is filled in by
+$[[true]]$ (resp. $[[false]]$). Informally, the type of each branch is more
+precise than those in a nondependent match expression, since it knows which
+branch it is in. An alternative perspective is that to prove $[[A {a/x}]]$, it
+suffices to show that it holds for the two canonical cases that $[[a]]$ is
+$[[true]]$ or is $[[false]]$.
+
+% The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
+% matching.  The result type of this expression, $[[A {a / x}]]$, is composed of some
+% \emph{motive} $[[A]]$, a type where its single free variable has been replaced
+% with the condition of the if expression. When typing the true branch, this
+% substitution replaces the variable by $[[true]]$, and similarly for the false
+% branch. As a result, the type system communicates the information gained from
+% the test to each of the branches of the expression.
+
+Dependent pattern matching similarly occurs when eliminating equality types.
+Their well-formedness in \rref{T-Eq} enforces that the endpoints of the
+equality have the same type, and there is a single canonical proof introduced
+by \rref{T-Refl}. The elimination form in \rref{T-J} takes a proof of an
+equality $[[p]]$ between $[[a]]$ and $[[b]]$, as well as an elimination body
+$[[c]]$. Here, the motive $[[B]]$ is abstracted over the right endpoint $[[x]]$
+and an equality $[[y]]$ between it and the fixed left endpoint $[[a]]$. The
+eliminator states that to prove $[[B{b, p / x, y}]]$, it suffices to prove the
+canonical case that $[[p]]$ is $[[refl]]$ (and consequently that $[[b]]$ is
+$[[a]]$), which is witnessed by $[[c]]$.
+
+% In \rref{T-J}, the elimination form, the subterm
+% $p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
+% is the body of the elimination form. In this rule, $[[B]]$ is the motive and
+% has two free variables.  When checking $[[c]]$, the substitution for these
+% variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
+% witnessing the information gained through dependent pattern matching.
+
+The universe hierarchy and the boolean base type give the
+ability to compute a type using a term as input, referred to as
+\emph{large elimination}. For example, the well-typed function
+$[[\ x . if x Bool Void]]$ returns either $[[Bool]]$ or $[[Void]]$
+depending on whether its input is $[[true]]$ or $[[false]]$.
 % Working with a system with untyped equality has the huge benefit that

### 2024-12-02 paper/paper.tex
Edits up to partway through 3.1

@@
 Our reduction and typing relations are defined in terms of \emph{simultaneous
-substitutions} $[[rho]]$, which are mappings from variables to terms. We
-use $[[idtm]]$ as the identity substitution, and the extension operator
+substitutions} $[[rho]]$, which are mappings from variables to terms. Simultaneous
+substitutions are more convenient to reason about than single substitutions,
+especially when dealing with semantic typing in Section~\ref{sec:logrelproof}.
+We use $[[idtm]]$ as the identity substitution, and the extension operator
 $[[rho .: x -> a]]$ updates the substitution $[[rho]]$ to map the variable
-$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.
+$[[x]]$ to $[[a]]$ rather than to $[[rho x]]$.
 % When traversing under binders (e.g. in the
@@
   watertight and rigorous, though we have one in our mechanization}
-The substitution operator $[[a {rho}]]$ replaces every free variable $[[x]]$ of
+The substitution operation $[[a {rho}]]$ replaces every free variable $[[x]]$ of
 $[[a]]$ by $[[rho x]]$ simultaneously. A single substitution corresponds to
@@
 $[[a { b / x }]] \coloneqq [[a { idtm .: x -> b }]]$.
-
-When reasoning about logical relations, it is more convenient to formulate
-definitions in terms of simultaneous substitutions than single substitutions,
-such as semantic typing in Section~\ref{sec:logrelproof}.
-\jc{I've rewritten this paragraph to be more succinct,
-but I'm still not certain it's necessary at all.
-We're not focussing on the merits of simultaneous substitution over
-other representations of substitution and binding, after all.}
 
@@
 
-Before we define our logical relation, we first formally specify the
-consistency property that we want to prove.
+Our ultimate goal is to prove the following consistency property for our type theory.
+
 \begin{theorem}[Logical Consistency]
@@
 The property can be formulated in a simply typed language, where
-$[[Void]]$ is similarly defined as a type that has no term. A related
-property, referred to as the termination property (for closed terms),
+$[[Void]]$ is a type with no constructors. A related
+property, the \emph{termination property} (for closed terms),
 is commonly used in introductory materials such as
@@
 
-A naive attempt to proving Theorem~\ref{theorem:consistency} by
-induction on the derivation $[[empty |- a : Void]]$ would succeed at
+A na\"ive attempt at proving Theorem~\ref{theorem:consistency} by
+induction on the derivation $[[empty |- a : Void]]$ would succeed in
 almost all cases except for \rref{T-App}. In the application
@@
 the equality that $[[B {a / x} = Void]]$. Our goal is to show that
-$[[empty |- b a : Void]]$ is not possible. However, note that there is
-nothing we know of $[[b]]$ or $[[a]]$ from the induction hypothesis
-because neither $[[Pi x : A . B]]$ nor $[[A]]$ is equal to $[[Void]]$.
-We have no way of deriving a contradiction from $[[empty |- b a :
+$[[empty |- b a : Void]]$ is not possible. Unfortunately, there is
+nothing we know about $[[b]]$ or $[[a]]$ from the induction hypothesis
+because neither $[[Pi x : A . B]]$ nor $[[A]]$ are equal to $[[Void]]$,
+so we have no way of deriving a contradiction from $[[empty |- b a :
 Void]]$. The takeaway from this failed attempt is that, in order to
@@
 
-The complexity of applying proof by logical relation to dependent types stems
-from the fact that the logical relation is much harder to define. In
-simply typed languages, the logical relation is defined as a recursive
-function over the type $[[A]]$.
-In dependent types, the type
-$[[A]]$ can take the form $[[(\ x . x ) Bool]]$. To assign meaning to
-this type, we need to first reduce it to $[[Bool]]$. However, we
-cannot write a function that performs the reduction because we do not
-know the termination of well-typed terms a priori. As a result, we
-define the logical relation as an inductively defined relation,
-reminiscent of how we specify the reduction graph of a partial
-function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.
+\jc{Minor comment but ``proof by logical relation'' sounds grammatically strange to me,
+but I'm not sure whether ``proof by logical relations'' or
+``proof by a logical relation'' would be any more correct...}
+
+The challenge behind applying proofs by logical relation to dependent types
+stems from the difficulty in defining the logical relation itself. In simply
+typed languages, the logical relation is a recursive function over the type
+$[[A]]$. In dependently typed languages, the type $[[A]]$ can take the form
+$[[(\ x . x ) Bool]]$, for example. To assign meaning to this type, we need to
+first reduce it to $[[Bool]]$. However, we cannot write a function that
+performs the reduction because we do not know the termination of well-typed
+terms a priori. As a result, we define the logical relation as an inductively
+defined relation, reminiscent of how we specify the reduction graph of a
+partial function.
+We later recover functionality of the relation in Lemma~\ref{lemma:logreldeter}.
 
 \subsection{Definition of the Logical Relation}
-\begin{figure}[H]
+
+\begin{figure}
 \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
@@
 \end{figure}
+
 The logical relation for \lang{}, which takes the form $[[Interp I i A
-S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
+S]]$, is defined as an inductively generated relation in Figure~\ref{fig:logrel}.
 % The logical relation takes the form $[[Interp I i A S]]$.
-Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
-numbers respectively, as introduced earlier in
-Figure~\ref{fig:syntax}.
+The metavariables $[[A]]$ and $[[i]]$ are terms and naturals, respectively.
 % \scw{Many introductory texts define the relation as a recursive function
@@
 The metavariables $[[I]]$ and $[[S]]$ are
-sets with the following signatures.
+sets with the following signatures,
+using $[[PowerSet STm]]$ to denote the powerset of the set of terms.
 \[
@@
 \]
-The notation $[[PowerSet STm]]$ denotes the powerset of the set of
-\lang{} terms.
-The function $[[I]]$ is a family of sets of terms indexed by
-natural numbers strictly less than the parameter $[[i]]$, which
-represents the current universe level.  In \rref{I-Set}, the function
-$[[I]]$ is used to define the meaning of
+The function $[[I]]$ is a family of sets of terms indexed by natural numbers
+strictly less than $[[i]]$, which represents the current universe level.
+In \rref{I-Set}, the function $[[I]]$ is used to define the meaning of
 universes that are strictly smaller than the current level $[[i]]$. The
-restriction $[[j < i]]$ in \rref{I-Set} ensures the predicativity of the system.
-
-We tie the knot and obtain an interpretation for all universe levels
-below. The judgment $[[InterpR i A S]]$ reads that the type $[[A]]$ is a
-level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
-$[[S]]$. 
+restriction $[[j < i]]$ in \rref{I-Set} ensures predicativity of the system.
+
+Predicativity allows us to tie the knot and obtain an interpretation for all
+universe levels as the judgment $[[InterpR i A S]]$, which says that the type
+$[[A]]$ is a type at universe level $[[i]]$, \emph{semantically} inhabited by
+terms from the set $[[S]]$.
+
 \begin{definition}[Logical relation for all universe levels]
 \label{fig:logrelrec}
-Define $[[InterpR i A S]]$ recursively through the well-foundedness of the $<$
+$[[InterpR i A S]]$ is defined recursively by well-foundedness of the $<$
 relation on natural numbers.
@@
 \end{definition}
-Definition~\ref{fig:logrelrec} explains how the $[[j
-< i]]$ constraint in \rref{I-Set} makes our system predicative; the
-interpretation of the $[[i]]_{th}$ universe is only dependent on
-universes strictly lower than $[[i]]$, which have already been defined.
-This restriction ensures that that the relation is well-defined: without 
-it the definition of $[[InterpR i A S]]$ would
-not be well-founded; $[[Interp I i A S]]$ would call $[[I]]$ on
-universe levels greater than or equal to $[[i]]$, which are yet to be defined.
+
+Our system is predicative because the interpretation of the $[[i]]$th universe
+is dependent only on universes strictly below $[[i]]$,
+which have already been defined.
+This restriction ensures that the relation is well defined; otherwise, the
+definition of $[[InterpR i A S]]$ would not be well founded, and
+$[[Interp I i A S]]$ could call $[[I]]$ on universe levels greater than
+or equal to $[[i]]$, which are yet to be defined.
 %% SCW: removing the constraing wouldn't result in any system at all
@@
 the more explicit definition not only because it is directly definable
-in proof assistants that lack induction-recursion,
-% where inductive definitions must appear
-%at the top level,
+in proof assistants that lack induction--recursion,
 but also because it makes clear the induction
-principle we are allowed to use when reasoning about $[[InterpR i A
-S]]$.
-
-% The paragraph below might be useful but I don't know how to phrase
-% it well
+principle we are allowed to use when reasoning about $[[InterpR i A S]]$.
 
@@
 
-We next take a closer look at the inductive relation $[[Interp I i A S]]$, defined 
-in Figure~\ref{fig:logrel}.
-\Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
+We next take a closer look at the remaining rules for the inductive relation
+$[[Interp I i A S]]$ in Figure~\ref{fig:logrel}.
+\Rref{I-Void, I-Bool} capture terms that \emph{behave} like the inhabitants
 of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For

### 2024-12-03 paper/paper.tex
More edits and one giant comment

@@
 \Rref{I-Void, I-Bool} capture terms that \emph{behave} like the inhabitants
-of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
-example, the $[[Void]]$ type should not have any inhabitants under the
-empty context, where as the $[[Bool]]$ type only contains terms that
-reduce to $[[true]]$ or $[[false]]$. Note that the characterization
-of $[[Bool]]$ (and other inhabited types) in our logical relation does not always correspond to
-well-typed or even closed terms. For example, the term $[[if false (#
-Void true #)
-true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
-  true \/ a =>+ false }]]$ since it evaluates to $[[true]]$.
+of the $[[Void]]$ and $[[Bool]]$ types under an empty context. In particular,
+the $[[Void]]$ type has no inhabitants, while the $[[Bool]]$ type only contains
+terms that reduce to $[[true]]$ or $[[false]]$. Note that the characterization
+of $[[Bool]]$ (and other inhabited types) in our logical relation does not
+always correspond to well-typed or even closed terms. For example, the term
+$[[if false (# Void true #) true]]$ is ill typed under the empty context,
+but still belongs to the set $[[{ a | a =>+ true \/ a =>+ false }]]$ since it
+evaluates to $[[true]]$.
 The independence of syntactic typing in our logical relation allows
@@
 need for a Kripke-style logical relation when we extend our logical
-relation to prove the existence of $\beta$-normal forms (Section~\ref{sec:extension}).
+relation to prove the existence of $\beta$-normal forms in Section~\ref{sec:extension}.
 \yl{Not sure what to cite from Derek Dreyer. I know his blog post
@@
 of the rust papers?}
-
-% unsurprising; when considering only closed terms, the empty type
-% should not be inhabited and therefore corresponds to the empty set,
-% whereas the boolean type is semantically inhabited by terms that
-% evaluate to the boolean values $[[true]]$ or $[[false]]$.
-\Rref{I-Eq}
-says that an equality type $[[a ~ b : A]]$ corresponds to the
+\jc{Should there be an explanation of what Kripke-style logical relations are here?}
+
+\Rref{I-Eq} states that an equality type $[[a ~ b : A]]$ corresponds to the
 set of terms that reduce to $[[refl]]$ when $[[a <=> b]]$ also holds and
-otherwise corresponds to the empty set. Conditions like $[[a <=>
-b]]$ are typically required for indexed types, of which equality types
+otherwise corresponds to the empty set. Conditions like $[[a <=> b]]$ are
+typically required for indexed types, of which equality types
 are an instance. \Rref{I-Red} enables us to reduce types in order
-to assign meanings. Recall the type expression $[[(\ x . x )
-Bool]]$. \Rref{I-Red} says that to know that
-$[[Interp I i  (\ x . x) Bool S ]]$ for some $[[S]]$, it suffices to
-show that $[[Interp I i Bool S]]$ since $[[(\ x . x) Bool =>
-Bool]]$. The derivation that $[[Interp I i (\ x . x) Bool { a | a =>+
-  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
-and \rref{I-Bool}.
-
-\Rref{I-Pi} is the most complex rule in our logical
-relation. Instead of explaining it directly, we first consider the following
-simplified version, \rref{I-PiAlt}, that follows directly from \rref{I-Pi}.
+to assign meanings. Recalling expression $[[(\ x . x ) Bool]]$, \rref{I-Red}
+says that to know that $[[Interp I i  (\ x . x) Bool S ]]$ for some $[[S]]$,
+it suffices to show that $[[Interp I i Bool S]]$, since
+$[[(\ x . x) Bool => Bool]]$.
+The derivation for $[[Interp I i (\ x . x) Bool { a | a =>+ true \/ a =>+ false }]]$
+therefore follows by composing \rref{I-Red} and \rref{I-Bool}.
+
+\Rref{I-Pi} is the most complex rule in our logical relation.
+Consider first the following simplified \rref{I-PiAlt}, which follows from \rref{I-Pi}.
+
 \begin{center}
@@
 \end{center}
+
 \Rref{I-PiAlt} directly captures the meaning of
-a well-behaved dependent function type. The precondition of
-the rule says that
-the function type $[[Pi x : A . B]]$ has an interpretation if its input
-type $[[A]]$ can be
-interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
-type $[[B {a / x}]]$, obtained by substituting $[[a]]$ into the output type
-$[[B]]$, has some semantic interpretation $[[S0]]$. In its conclusion, the 
-interpretation of $[[Pi x : A . B]]$ is the set of terms $[[b]]$, such that
-for all $[[a in S]]$, where $[[S]]$ is an interpretation of $[[A]]$, the
-application form $[[b a]]$ belongs to all possible interpretations of
-$[[B {a  / x}]]$ (the pre-condition ensures at least one interpretation
-exists for each $[[B {a / x}]]$ where $[[a in S]]$).
+a well-behaved dependent function type. The preconditions of
+the rule says that the function type $[[Pi x : A . B]]$ has an interpretation
+if its input type $[[A]]$ can be interpreted as some set $[[S]]$, and for all
+terms $[[a in S]]$, the type $[[B {a / x}]]$, obtained by substituting $[[a]]$
+into the output type $[[B]]$, has some semantic interpretation $[[S0]]$. In its
+conclusion, the  interpretation of $[[Pi x : A . B]]$ is the set of terms
+$[[b]]$ such that for all $[[a in S]]$, where $[[S]]$ is an interpretation of
+$[[A]]$, the application form $[[b a]]$ belongs to all possible interpretations
+of $[[B {a  / x}]]$. The second precondition ensures at least one interpretation
+exists for each $[[B {a / x}]]$ where $[[a in S]]$.
+
+\jc{I think at this point it's unclear why \rref{I-PiAlt} is shaped the way it is.
+I found myself asking: If in the second precondition we've established the existence of $[[S0]]$,
+why do we need to quantify over all interpretations $[[S0]]$?
+In fact, instead of the second precondition,
+why can't split it into two and say that there exists some family of sets
+$S_0 : \forall a, a \in S \to [[PowerSet STm]]$
+such that $\forall a, (m : a \in S) \to [[Interp I i B { a / x } (# S0 a m #)]]$,
+then conclude that the intepretation of the function type yields a set over $[[b]]$
+such that $\forall a, (m : a \in S) \to [[b in S0 a m]]$?
+As far as I can tell, the interpretation would be well formed,
+but proving the fundamental theorem would then require something like indefinite description,
+since the induction hypothesis would yield $\forall a, a \in S \to \exists S_0, \dots$,
+the mere existence of such a family of sets,
+while the constructor requires the actual family of sets itself.
+This is subtle but maybe warrants explaining.}
 

### 2024-12-03 paper/paper.tex
Rewrote some chunks of 3.1

@@
 \Rref{I-Pi} is the most complex rule in our logical relation.
-Consider first the following simplified \rref{I-PiAlt}, which follows from \rref{I-Pi}.
+It states that, to build the interpretation of a dependent function type,
+we first require the interpretation $[[S]]$ of its domain $[[A]]$.
+Because the codomain depends on the function input,
+rather than a single interpretation, we require an interpretation $[[F a]]$
+of the codomain $[[B {a / x}]]$ for each $[[a in S]]$.
+The interpretation of the overall function type is the set of terms $[[b]]$
+such that for every term $[[a]]$ in the interpretation $[[S]]$ of $[[A]]$,
+$[[b a]]$ is in the interpretation $[[F a]]$ of $[[B {a / x}]]$.
+
+However, this form of the rule is less convenient to work with, since it
+requires constructing the sets $[[F]]$ separately from the proof that they are
+indeed interpretations of $[[B]]$. A more convenient form is the following
+\rref{I-PiAlt}, which combines the existence of interpretations of the codomain
+with their proofs.
 
@@
 
-\Rref{I-PiAlt} directly captures the meaning of
-a well-behaved dependent function type. The preconditions of
-the rule says that the function type $[[Pi x : A . B]]$ has an interpretation
-if its input type $[[A]]$ can be interpreted as some set $[[S]]$, and for all
-terms $[[a in S]]$, the type $[[B {a / x}]]$, obtained by substituting $[[a]]$
-into the output type $[[B]]$, has some semantic interpretation $[[S0]]$. In its
-conclusion, the  interpretation of $[[Pi x : A . B]]$ is the set of terms
-$[[b]]$ such that for all $[[a in S]]$, where $[[S]]$ is an interpretation of
-$[[A]]$, the application form $[[b a]]$ belongs to all possible interpretations
-of $[[B {a  / x}]]$. The second precondition ensures at least one interpretation
-exists for each $[[B {a / x}]]$ where $[[a in S]]$.
-
-\jc{I think at this point it's unclear why \rref{I-PiAlt} is shaped the way it is.
-I found myself asking: If in the second precondition we've established the existence of $[[S0]]$,
-why do we need to quantify over all interpretations $[[S0]]$?
-In fact, instead of the second precondition,
-why can't split it into two and say that there exists some family of sets
-$S_0 : \forall a, a \in S \to [[PowerSet STm]]$
-such that $\forall a, (m : a \in S) \to [[Interp I i B { a / x } (# S0 a m #)]]$,
-then conclude that the intepretation of the function type yields a set over $[[b]]$
-such that $\forall a, (m : a \in S) \to [[b in S0 a m]]$?
-As far as I can tell, the interpretation would be well formed,
-but proving the fundamental theorem would then require something like indefinite description,
-since the induction hypothesis would yield $\forall a, a \in S \to \exists S_0, \dots$,
-the mere existence of such a family of sets,
-while the constructor requires the actual family of sets itself.
-This is subtle but maybe warrants explaining.}
-
-\begin{lemma}[I-PiAlt derivability]
+The second precondition now states that for every $[[a in S]]$,
+there must exist some interpretation $[[S0]]$ of $[[B {a / x}]]$.
+The interpretation of the overall function type then states that $[[b a]]$
+must be in every interpretation $[[S0]]$ of $[[B {a / x}]]$.
+Later, Lemma~\ref{lemma:logreldeter} proves functionality of the interpretation
+of types, which guarantees that there must only be one such interpretation.
+This alternate rule for the interpretation of function types follows from \rref{I-Pi}.
+
+\begin{lemma}[I-PiAlt derivability\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
   \label{lemma:piintroalt}
@@
 \end{lemma}
+
 \begin{proof}
-The precondition $[[forall a, (# a in S
-implies (# exists S0 , Interp I i B { a / x } S0 #) #)]]$ from
-\rref{I-PiAlt} immediately induces a function $[[F in S -> PowerSet
-STm]]$ such that $[[forall a, (# a in S implies Interp I i B { a / x }
-F a #)]]$, which is exactly what we need to apply \rref{I-Pi}.
+  The precondition
+  $[[forall a, (# a in S implies (# exists S0 , Interp I i B { a / x } S0 #) #)]]$
+  from \rref{I-PiAlt} immediately induces a function
+  $[[F in S -> PowerSet STm]]$ such that
+  $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$,
+  which is exactly what we need to apply \rref{I-Pi}.
 \end{proof}
-In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
-these two rules are equivalent in the sense that every derivation
-involving \rref{I-Pi} can be systematically replaced by
-\rref{I-PiAlt}. This equivalence follows directly from the fact
-that the logical relation is a partial function, a result we will show
-in Lemma~\ref{lemma:logreldeter}. The preconditions of \rref{I-Pi},
-when combined with the functionality of the logical relation, uniquely
-determine the function $[[F in S -> PowerSet STm]]$ to be the
-functional relation
-$[[{ (a , S0 ) | a in S implies Interp I  i B { a /x } S0 }]]$. This result is
-formally shown through the improved inversion lemma for function types
-(Lemma~\ref{lemma:piinvalt}).
-
-Unfortunately, we cannot define the function case of our logical
-relation directly using \rref{I-PiAlt} since the occurrence of
-$[[Interp I i B {a/x} S0]]$ in its conclusion not only violates the
-syntactic strict positivity constraint required in proof assistants,
+
+While \rref{I-PiAlt} is an instantiation of \rref{I-Pi}, by functionality,
+the two rules are equivalent in the sense that every derivation involving
+\rref{I-Pi} can be systematically replaced by \rref{I-PiAlt}.
+Furthermore, functionality uniquely determines the function
+$[[F in S -> PowerSet STm]]$ to be the functional relation
+$[[{ (a , S0 ) | a in S implies Interp I i B { a /x } S0 }]]$.
+This result is shown by Lemma~\ref{lemma:piinvalt},
+the corresponding inversion lemma for \rref{I-PiAlt}.
+
+Unfortunately, we cannot directly define the interpretation of function types
+by \rref{I-PiAlt}, since the occurrence of $[[Interp I i B {a/x} S0]]$ in its
+conclusion not only violates the syntactic strict positivity constraint on
+inductive definitions required by proof assistants,
 but is genuinely non-monotone when we treat the inductive definition
 as the fixed point of an endofunction over the domain of relations.
-Intuitively, the failure of monotonicity stems from the fact
-that the witness picked in the precondition is not necessarily the
-same witness being referred to in the post condition as the relation
-grows, whereas the
-function $[[F]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ as
-$[[F a]]$ for each $[[a in S]]$,
-thus preventing the set of witnesses from growing. While it might
+\jc{What?}
+Intuitively, the failure of monotonicity stems from the fact that the witness
+picked in the precondition is not necessarily the same witness being referred
+to in the postcondition as the relation grows, \jc{Why?} whereas the function $[[F]]$
+in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ as $[[F a]]$ for each
+$[[a in S]]$, thus preventing the set of witnesses from growing. While it might
 be possible to restrict the domain with additional constraints such as
@@
 \rref{I-Pi} that immediately produces a
-well-defined inductive relation and usable induction principle. The
-slight disadvantage of \rref{I-Pi} is that we need to construct the
-function $[[F]]$ each time we apply it, though this is mitigated by
-the derivability of \rref{I-PiAlt} and the alternative $\Pi$ inversion
-principle (Lemma~\ref{lemma:piinvalt}).
+well-defined inductive relation and usable induction principle.
+Then by deriving \rref{I-PiAlt} and its inversion lemma,
+we avoid needing to manipulate the function $[[F]]$ directly.
 

### 2024-12-03 paper/paper.tex
Finished edits for Section 3

@@
 \begin{align*}
-[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & \Coloneqq && \textit{Terms} \\
+[[a]],[[b]],[[c]],[[p]],[[A]],[[B]],[[C]] & \Coloneqq && \textit{Terms} \\
             & \mid [[Set i]] \mid [[x]] \mid [[Void]] \mid [[absurd b]]
@@
 
-\subsection{Properties about the Logical Relation}
+\subsection{Properties of the Logical Relation}
 In the rest of this section, we develop the theory of our logical relation
@@
 follows as a corollary of a property about $[[Interp I i A S]]$ with no or few
-assumptions imposed on $[[I]]$. As a result, we usually state our lemmas in
+assumptions imposed on $[[I]]$. As a result, we state most of our lemmas in
 terms of $[[Interp I i A S]]$ without duplicating them in terms of
@@
 
-First, we prove a family of simple properties, which we refer to as
-inversion principles for our logical relation. Given $[[Interp I i A
-S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi x
-: A0 .B0]]$, the inversion lemma allows us to say something about the set
-$[[S]]$. Its proof is simple, but we sketch out the case for
+First, we prove inversion principles for our logical relation. Given
+$[[Interp I i A S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or
+$[[Pi x : A . B]]$, the inversion lemma allows us to say something about the set
+$[[S]]$. The proofs are simple, but we sketch out the case for
 functions to help readers confirm their understanding of \rref{I-Pi}.
-\begin{lemma}[Inversion of the logical relation]
+
+\begin{lemma}[Inversion (logical relation)]
   \label{lemma:interpinv}\leavevmode
@@
   \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl /\ a <=> b  }]]$.
-  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp
-    I i Pi x : A . B S1]]$, then there exists $[[S]],[[F]]$ such that:
+  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
+    If $[[Interp I i Pi x : A . B S1]]$, then there exist
+    $[[S in PowerSet STm]]$, $[[F in S -> PowerSet STm]]$ such that:
     \begin{itemize}
     \item $[[Interp I i A S ]]$
-    \item $[[F in S -> PowerSet STm]]$
     \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
@@
 \end{lemma}
+
 \begin{proof}
-  As mentioned earlier, we only show the inversion property for the
-  function type.
-  We start by inducting over the derivation of $[[Interp I i Pi x : A . B S]]$. There
-  are only two possible cases we need to consider.
+  We show only the inversion property for the function type.
+  We start by induction on the derivation of $[[Interp I i Pi x : A . B S1]]$.
+  There are only two possible cases we need to consider.
   \begin{description}
   \item[\Rref{I-Pi}:] Immediate.
-  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi x : A . B S1]]$.
-    We know that there exists some $[[A0]]$ and
-    $[[B0]]$ such that $[[Pi x : A . B => Pi x : A0 . B0]]$ and $[[Interp I i Pi
-    x : A0 . B0 S1]]$. From the
-    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
+  \item[\Rref{I-Red}:]
+    We are given that $[[Pi x : A . B => C]]$ and that $[[Interp I i C S1]]$.
+    By inversion on parallel reduction, we have that $[[C]] = [[Pi x : A0 . B0]]$
+    for some $[[A0]], [[B0]]$, and that $[[A => A0]]$ and $[[B => B0]]$.
+    From the induction hypothesis on $[[Interp I i (# Pi x : A0 . B0 #) S1]]$,
+    there exist $[[S in PowerSet STm]]$ and $[[F in S -> PowerSet STm]]$ such that:
     \begin{itemize}
       \item $[[Interp I i A0 S ]]$
-    \item $[[F in S -> PowerSet STm]]$
       \item $[[forall a, (# a in S implies Interp I i B0 { a / x } F a #)]]$
@@
     \end{itemize}
-    By inverting the derivation of $[[Pi x : A . B => Pi x : A0 . B0]]$, we derive $[[A => A0]]$ and
-    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a /x} => B0 {a/x} ]]$ for all
-    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
-    $[[F]]$ additionally satisfy the following properties.
+    By Lemma~\ref{lemma:parsubst}, we have $[[B {a /x} => B0 {a/x}]]$ for all
+    $[[a]]$. As a result, $[[S]]$ and $[[F]]$ satisfy the following properties
+    by \rref{I-Red} on $[[A => A0]]$ and $[[B {a /x} => B0 {a/x}]]$, respectively:
     \begin{itemize}
-    \item $[[Interp I i A S ]]$
+      \item $[[Interp I i A S]]$
       \item $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$
@@
   \end{description}
+  \vspace{-2\baselineskip}
 \end{proof}
@@
 $[[Interp I i B S]]$ also holds. The following property shows that
-preservation holds in the usual forward direction too.
-\begin{lemma}[Forward preservation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
+preservation holds in the usual forward direction as well.
+
+\begin{lemma}[Forward preservation (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
   \label{lemma:interppreservation}
@@
 \end{lemma}
+
 \begin{proof}
-  We carry out the proof by induction over the derivation of $[[Interp
-  I i A S]]$.
+  We carry out the proof by induction on the derivation of
+  $[[Interp I i A S]]$.
 
@@
   B0 S]]$. By \rref{I-Red} and $[[B1 => B]]$, we conclude that
-  $[[Interp I i B S]]$.
-
-  The remaining cases all fall from induction hypotheses and basic
-  properties about convertibility and parallel reduction we have
-  established in Section~\ref{sec:spec}.
+  $[[Interp I i B1 S]]$.
+
+  The remaining cases all fall from the induction hypotheses and the basic
+  properties of convertibility and parallel reduction established in
+  Section~\ref{sec:spec}.
 \end{proof}
-From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
-derive the following corollary that two convertible types can always interpret
+
+From forward preservation and \rref{I-Red}, we can easily
+derive the following corollary that two convertible types always interpret
 into the same set. We adopt the terminology from \citet{martin-lof-a-la-coq}
-and refer to this property as irrelevance.
-\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
+and refer to this property as \emph{irrelevance}.
+
+\begin{corollary}[Irrelevance (l.r.)\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
   \label{lemma:logrelcoherence}
@@
 
-Because the definition of our logical relation is an inductive relation,
-it is not immediately obvious why each type $[[A]]$ can only uniquely
-correspond to one set $[[S]]$. The following lemma shows that our
-logical relation is indeed functional.
-\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
+Because the definition of our logical relation is an inductive relation, it is
+not immediately obvious that each type $[[A]]$ interprets to a unique set $[[S]]$.
+The following lemma shows that our logical relation is indeed functional.
+
+\begin{lemma}[Functionality (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
   \label{lemma:logreldeter}
@@
 \end{lemma}
+
 \begin{proof}
-  The proof proceeds by induction over the derivation of the first
+  The proof proceeds by induction on the derivation of the first
   premise $[[Interp I i A S0]]$.
@@
 
-  For \rref{I-Red}, we are given that there exists some $[[B]]$ such
-  that $[[A => B]]$ and $[[Interp I i B S0]]$. Our goal is to show
-  that given $[[Interp I i A S1]]$ for some $[[S1]]$, we have $[[S0 =
-  S1]]$. By the preservation property
-  (Lemma~\ref{lemma:interppreservation}),
-  we know that $[[Interp I i B S1]]$ since $[[A => B]]$. The statement
-  $[[S0 = S1]]$ then immediately follows from the induction hypothesis.
+  In \rref{I-Red}, there exists some $[[B]]$ such that $[[A => B]]$ and
+  $[[Interp I i B S0]]$. Our goal is to show that given $[[Interp I i A S1]]$
+  for some $[[S1]]$, we have $[[S0 = S1]]$. By forward preservation
+  (Lemma~\ref{lemma:interppreservation}) and $[[A => B]]$,
+  we know that $[[Interp I i B S1]]$.
+  Then $[[S0 = S1]]$ immediately follows from the induction hypothesis.
 \end{proof}
 
-Lemma~\ref{lemma:logreldeter} enables us to show the following
-improved inversion lemma for function types whose statement is free of
-the relation $[[F]]$, analogous to the derivable \rref{I-PiAlt}.
-\begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
-  \label{lemma:piinvalt}
-  Suppose $[[Interp I i Pi x : A . B S]]$, then there exists some $[[S0]]$
-  such that the following constraints hold:
+Functionality allows us to prove an inversion lemma for the derivable
+\rref{I-PiAlt}, which does not mention the function $[[F]]$ found in \rref{I-Pi}.
+
+\begin{lemma}[Alternate inversion of function types (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
+  \label{lemma:piinvalt}\leavevmode
+  If $[[Interp I i Pi x : A . B S]]$, then there exists some $[[S0]]$
+  such that:
   \begin{itemize}
     \item $[[Interp I i A S0]]$
-  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a
-    / x}
-    S1 #) #)]]$
-  \item $[[S = { b | forall a, (# a in S0 implies forall
-    S1, (# Interp I i B {a / x} S1 implies  b a in S1 #) #) }]]$
+    \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a / x} S1 #) #)]]$
+    \item $[[S = { b | forall a, (# a in S0 implies forall S1, (# Interp I i B {a / x} S1 implies  b a in S1 #) #) }]]$
   \end{itemize}
 \end{lemma}
+
 \begin{proof}
@@
 
-The next lemma shows that our logical relation satisfies
-cumulativity. That is, if a type has an interpretation at a lower
-universe level, then we can obtain the same interpretation at a higher
-universe level.
-\begin{lemma}[Logical relation cumulativity\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
+The next lemma shows that the logical relation is cumulative:
+if a type has an interpretation at a lower universe level,
+then we can obtain the same interpretation at a higher universe level.
+
+\begin{lemma}[Cumulativity (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
   \label{lemma:logrelcumulativity}
-  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
-\end{lemma}
+  If $[[Interp I i A S]]$ and $[[i < j]]$, then $[[Interp I j A S]]$.
+\end{lemma}
+
 \begin{proof}
-  Trivial by structural induction over the derivation of $[[Interp I
-  i0 A S]]$.
+  Trivial by structural induction over the derivation of $[[Interp I i A S]]$.
 \end{proof}
-Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
-implicitly assume that $[[I]]$ is defined on the set of natural
-numbers less than $[[i1]]$.
-
-\begin{corollary}[Logical relation is functional with different levels\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
+
+Note that in the statement of cumulativity, we implicitly assume that $[[I]]$
+is defined on naturals strictly less than $[[j]]$. Then by cumulativity and
+trichotomy of the order on naturals (i.e. for any $i, j$, either $i < j$ or
+$i = j$ or $i > j$), we show that functionality holds even when considering
+interpretations of type at different universe levels.
+
+\begin{corollary}[Level-irrelevant functionality (l.r.)\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
   \label{lemma:logreldeterhet}
@@
 \end{corollary}
+
 \begin{proof}
@@
 
-\begin{definition}[Sets closed under expansion]
-  We say that a set of terms $[[S]]$ is closed under expansion if
-  given $[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
-\end{definition}
 The final property we want to show is that the output set $[[S]]$ from
@@
 in terms of $[[InterpR i A S]]$ rather than $[[Interp I i A S]]$
-because we need to know something about $[[I]]$ for this property to
-hold in the \rref{I-Set} case.
-
-\yl{Can lift this lemma somewhere earlier in the section since it doesn't really
-  depend on anything}
-\scw{Jon didn't understand this proof}
-\begin{lemma}[Interpreted sets are closed under expansion\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
+because we need the actual instantiation of $I$ to prove the \rref{I-Set} case.
+
+\begin{definition}[Closure under expansion]
+  We say that a set of terms $[[S]]$ is closed under expansion if
+  given $[[a in S]]$, then $[[b in S]]$ for all $[[b => a]]$.
+\end{definition}
+
+\begin{lemma}[Backward closure of interpretations (l.r.)\footnote{\dotv{semtyping.v}{InterpUnivN\_back\_clos}}]
   \label{lemma:logrelbackclos}
-  If $[[InterpR i A S]]$, then the set $[[S]]$ is closed under expansion.
-\end{lemma}
+  If $[[InterpR i A S]]$, then $[[S]]$ is closed under expansion.
+\end{lemma}
+
 \begin{proof}
-  By the definition of $[[InterpR i A S]]$, we unfold $[[InterpR i A S]]$ by one
-  step into $[[ Interp I i A S  ]]$ where $[[I j]] := [[{A | exists S
-    , InterpR j A S}]]$.
-  We then proceed by induction over the derivation of $[[Interp I i A S]]$.
-
-  All cases are trivial except for the \rref{I-Set} case, where we
-  want to show that the set $[[I j]]$ is closed under expansion for
-  all $[[j < i]]$. However, by the definition of $[[I]]$, we know that
-  $[[A in I j]]$ if and only if there exists some $[[S]]$ such that
-  $[[Interp I j A S]]$. By \rref{I-Red}, we must also have $[[B in I
-  j]]$ for all $[[B => A]]$.
+  By the definition of $[[InterpR i A S]]$, we unfold $[[InterpR i A S]]$ into
+  $[[Interp I i A S]]$ where $[[I j]] := [[{A | exists S, InterpR j A S}]]$ for
+  $[[j < i]]$. We then proceed by induction on the derivation of
+  $[[Interp I i A S]]$.
+
+  All cases are trivial except for the \rref{I-Set} case, where the goal is to
+  show that $[[I j]]$ is closed under expansion for all $[[j < i]]$;
+  that is, if $[[B => A]]$ and $[[A in I j]]$, then $[[B in I j]]$.
+  By the definition of $[[I]]$, this is equivalent to showing that if
+  $[[B => A]]$ and $[[Interp I j A S]]$ for some $[[S]]$,
+  then there exists some $[[S0]]$ such that $[[Interp I j B S0]]$.
+  Letting $[[S0]]$ be $[[S]]$, by \rref{I-Red},
+  we have that $[[Interp I j B S]]$ as required.
 \end{proof}
@@
 \label{sec:logrelproof}
+
 \scw{Would it make sense to define the notation $a \in [\![A]\!]^i$ when

### 2024-12-03 paper/paper.tex
Supress an orphan

@@
 
-The next lemma shows that the logical relation is cumulative:
+The next lemma shows cumulativity of the logical relation:
 if a type has an interpretation at a lower universe level,
-then we can obtain the same interpretation at a higher universe level.
+then we obtain the same interpretation at a higher level.
 

### 2024-12-04 paper/paper.tex
Finished edits for Section 4

@@
     $[[S in PowerSet STm]]$, $[[F in S -> PowerSet STm]]$ such that:
+    \vspace{-0.5\baselineskip}
     \begin{itemize}
@@
 
+An immediate consequence of the inversion principle for the interpretation
+of universes is that if a type is in such an interpretation, then that type
+itself has an interpretation.
+
+\begin{corollary}\hspace{-0.5em}\footnote{\dotv{semtyping.v}{InterpUnivN\_Univ\_inv'}} \label{lemma:interpinvSet}
+  If $[[InterpR i Set j S]]$ and $[[A in S]]$,
+  then there exists some set $[[S0]]$ such that $[[InterpR j A S0]]$ and $[[j < i]]$.
+\end{corollary}
+
 \Rref{I-Red} bakes into the logical relation the backward preservation
@@
 there exists some $S$ such that $[[InterpR i A S]]$ and $[[a in S]]$ ?}
+\jc{I like this notation too, $[[a in InterpR i A]]$}
 
 In this section, we show that all closed, well-typed terms are contained
-within their type-indexed sets. In other words, $[[empty |- a : A]]$ implies
-$[[InterpR i A S]]$ and $[[a in S]]$.  This result gives us consistency
-because we know that $[[InterpR i Void S]]$ is defined, and that $[[S]]$ must
-be the empty set. Therefore, if there were some closed, well-typed term of type
-$[[Void]]$, it would need to be a member of the empty set, a contradiction.
+in the interpretation of their types. In other words, $[[empty |- a : A]]$
+implies that there exists a set $[[S]]$ such that $[[InterpR i A S]]$ and
+$[[a in S]]$ hold; we write $[[a in InterpR i A]]$ as shorthand. This result
+gives us consistency because $[[InterpR i Void emptyset]]$ holds, so if there
+were a closed, well-typed term of type $[[Void]]$, it would be a member of the
+empty set, which is a contradiction.
 
 To prove this result, we define a notion of semantic typing based on the
-logical relation we have defined in Section~\ref{sec:logreldep} and prove the
-fundamental lemma, which states that syntactic typing implies semantic typing.
-Semantic typing extends our logical relation from being a (type-indexed)
-family of predicates on closed terms, to a type-indexed family of predicates
-on open terms.
-
-The necessity of semantic typing as an extra layer of
-definition on top of the logic relation can be understood in simply typed
-languages~\citep{skorstengaard2019introduction, harpertait,
-  pierce2002types}. In our setting, attempting to show that
-$[[empty |- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ through induction over the derivation of $[[empty |- a : A]]$ will fail in
-\rref{T-Abs}, where the induction hypothesis is not helpful since the body of
-the lambda term is typed under a non-empty context. Through the definition of
-semantic typing, we can state a strengthened property that is actually
-provable.
-
-\begin{definition}[Semantic well-formed substitution\footnote{\dotv{soundness.v}{$\rho$\_ok}}]
-Define $[[rho |= G]]$ when
-\[ \forall [[x]], [[A]], [[i]], \text { and } [[S]], \text{ if }[[x :
-  A in G]]\text{ and
-                     } [[InterpR i A { rho } S ]] \text{, then } [[rho x in S]] \]
+logical relation and prove the fundamental lemma, which states that syntactic
+typing implies semantic typing. Semantic typing extends the logical relation
+from being a type-indexed family of predicates on closed terms to a
+type-indexed family of predicates on open terms.
+
+The necessity of semantic typing as an extra layer of definitions on top of the
+logic relation is understood in the simply typed 
+setting~\citep{skorstengaard2019introduction, harpertait,pierce2002types}.
+In our setting,
+\jc{Is this trying to say that our setting is different from the simply typed setting,
+or the same as the simply typed setting? Are the definitions to follow on semantic typing
+taken from the works cited above, or are they our own modifications?}  
+attempting to show that $[[empty |- a : A]]$ implies $[[a in InterpR i A]]$ by
+induction on the derivation of $[[empty |- a : A]]$ will fail for \rref{T-Abs},
+where the induction hypothesis is not helpful since the body of the lambda term
+is typed under a nonempty context. Using the definition of semantic typing, we
+can state a strengthened property that is actually provable.
+
+\begin{definition}[Semantically well formed substitutions\footnote{\dotv{soundness.v}{$\rho$\_ok}}]
+  We say that a substitution $[[rho]]$ is semantically well formed with respect
+  to the context $[[G]]$, written as $[[rho |= G]]$, when
+  \[ \forall [[x : A in G]], [[i]], [[S]],
+  [[InterpR i A { rho } S ]] \text{ implies } [[rho x in S]]. \]
 \end{definition}
@@
 
-The $[[rho |= G]]$ notation denotes the semantic well-formedness of a
-substitution $[[rho]]$ with respect to a context $[[G]]$. For every
-variable $[[x]]$ with its associated type $[[A]]$ in the context,
-$[[rho x]]$ is a term that inhabits all possible interpretations of the type
-$[[A {rho}]]$. The $\forall$ quantifier in its definition might look excessive since we
-know from Lemma~\ref{lemma:logreldeter} that each type can have at most
-one interpretation. However, since $[[rho |= G]]$ mostly appears in
-the position of a hypothesis, the $\forall$ statement is easy to
-instantiate and makes our proofs slightly easier. The few cases where we need to prove $[[rho |= G]]$ are
-handled by the following two structural properties, the second of
-which depends on Lemma~\ref{lemma:logreldeter}.
-%
-\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
+$[[rho |= G]]$ states that for every variable $[[x]]$ and its type $[[A]]$ in the $[[G]]$,
+$[[rho x]]$ is a term that inhabits every interpretation of $[[A {rho}]]$.
+While functionality of the logical relation tells us each type has at most one
+interpretation, quantifying over all possible interpretations $[[S]]$ makes the
+proofs slightly simpler. Since $[[rho |= G]]$ typically appears as a hypothesis,
+$[[S]]$ is easy to instantiate. The following structural properties handle the
+few cases where we need to prove $[[rho |= G]]$ as a goal; the second property
+depends on functionality.
+
+\begin{lemma}[Well-formed $[[rho]]$ w.r.t. empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
   \label{lemma:rhowfempty}
-  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
-\end{lemma}
-\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\rho$\_ok\_cons}}]
-  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
+  $[[rho |= empty]]$ holds for any $[[rho]]$.
+\end{lemma}
+
+\begin{lemma}[Well-formed $[[rho]]$ w.r.t. cons\footnote{\dotv{soundness.v}{$\rho$\_ok\_cons}}]
+  \label{lemma:rhowfcons}
+  If $[[a in InterpR i A]]$ and $[[rho |= G]]$, then
   $[[rho .: x -> a |= G ++ x : A]]$.
@@
 
-We next define semantic well-typedness.
+Semantic well-typedness is then defined using well formed substitutions
+to handle the contexts of open terms.
 
 \begin{definition}[Semantic typing\footnote{\dotv{soundness.v}{SemWt}}]
-Define $[[G |= a : A]]$  when
-\[ \forall [[rho]], \text{ if }[[rho |=
-                       G]]\text{ then there exists some } [[j]] \text{
-                       and } [[S]] \text{ such that } [[InterpR j A
-                       {rho} S]] \text{ and } [[a {rho} in S]]  \]
+  We say that $[[a]]$ is semantically typed as $[[A]]$ under the context $[[G]]$,
+  written $[[G |= a : A]]$, when
+  \[ \forall [[rho |= G]], \exists [[j]] \text{ such that } [[a {rho} in InterpR j A {rho}]]. \]
 \end{definition}
 
-This definition says the term $[[a]]$ can be semantically typed $[[A]]$ under
-the context $[[G]]$ if for all substitutions $[[rho]]$ such that
-$[[rho |= G]]$, the type $[[A { rho }]]$ can be interpreted as the set
-$[[S]]$, and $[[a { rho } in S]]$. Our definition of semantic well-typedness
-is standard, though dependent types add a small twist that we apply the
-$[[rho]]$ to $[[A]]$ and require that $[[A { rho }]]$ has some interpretation.
-
-Finally, we define semantic well-formedness for contexts, analogous to the
+Semantic typing $[[G |= a : A]]$ says that for all well-formed substitutions $[[rho |= G]]$,
+$[[a {rho}]]$ is in the interpretation of $[[A { rho }]]$ at some universe level.
+This definition is standard, though dependency of types on terms requires
+also applying the substitution $[[rho]]$ to the type $[[A]]$, and that
+$[[A { rho }]]$ have an interpretation.
+Finally, we define semantic well-formedness of contexts, analogous to the
 relation $[[|-G]]$.
@@
 \begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
-Define $[[|= G]]$ as follows.
-\[ \forall [[x : A in G]],\text{ there exists some } i \text{ such that } [[G |= A : Set i]] \]
+  We say that the context $[[G]]$ is semantically well formed, written $[[|= G]]$, when
+  \[ \forall [[x : A in G]], \exists i \text{ such that } [[G |= A : Set i]]. \]
 \end{definition}
+
 Recall that $[[|- G]]$ is defined inductively in terms of the
-syntactic typing judgment. We take a different approach here with its
+syntactic typing judgment. We take a different approach here for its
 semantic counterpart $[[|= G]]$. The definition of
-$[[|= G]]$ is not telescopic: with $[[|- G]]$, a variable appearing
+$[[|= G]]$ is not telescopic: for $[[|- G]]$, a variable appearing
 earlier in the context is well-scoped under a truncated context,
-whereas with $[[|= G]]$, the types are only required to be
-semantically well-formed with respect to the full context, regardless
-of their position in $[[G]]$. Our definition of $[[|= G]]$ could be
-strengthened, though the simpler definition is sufficient for showing the
-fundamental lemma.
-
+whereas for $[[|= G]]$, the types are only required to be
+semantically well-formed with respect to the full context, regardless of their
+position in $[[G]]$. While the definition of $[[|= G]]$ could be strengthened,
+the simpler definition is sufficient for showing the fundamental lemma.
 We can recover the structural rules for $[[|= G]]$ as lemmas.
-\begin{lemma}[Semantic context well-formedness empty\footnote{\dotv{soundness.v}{SemWff\_nil}}]
+
+\begin{lemma}[Empty context well-formedness\footnote{\dotv{soundness.v}{SemWff\_nil}}]
   \label{lemma:semwffempty}
-  $[[|= G]]$ holds when $[[G]]$ is empty.
-\end{lemma}
-\begin{lemma}[Semantic context well-formedness cons\footnote{\dotv{soundness.v}{SemWff\_cons}}]
+  $[[|= empty]]$ holds.
+\end{lemma}
+\begin{lemma}[Cons context well-formedness\footnote{\dotv{soundness.v}{SemWff\_cons}}]
   \label{lemma:semwffcons}
@@
 
-The following lemma makes the statement $[[G |= A : Set i]]$ easier to
+The following lemma makes statements of the form $[[G |= A : Set i]]$ easier to
 work with.
+
 \begin{lemma}[Set Inversion\footnote{\dotv{soundness.v}{SemWt\_Univ}}]
   \label{lemma:setinv}
-  The following two statements are equivalent:
+  The following are equivalent:
   \begin{itemize}
-  \item $[[G |= A : Set i]]$
-  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
-    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
+  \item $[[G |= A : Set i]]$;
+  \item $\forall [[rho |= G]]$, there exists a set $[[S]]$ such that
+    $[[InterpR i (A {rho}) S]]$.
   \end{itemize}
 \end{lemma}
+
 \begin{proof}
-  The forward direction is immediate by
-  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
+  The forward direction is immediate by Lemma~\ref{lemma:interpinvSet}.
+  We now consider the backward direction
   and show that $[[G |= A : Set i]]$ given the second bullet.
 
-  Suppose $[[rho |= G]]$, then we know that there exists some $[[S]]$
+  Suppose $[[rho |= G]]$. We know that there exists some $[[S]]$
   such that $[[InterpR i (A {rho}) S]]$. By the definition of semantic
@@
   Pick $[[Suc i]]$ for $[[j]]$ and $[[ { A | exists S , InterpR i A S }
-  ]]$ for $[[S0]]$ and it is trivial to verify the conditions hold.
+  ]]$ for $[[S0]]$; it is trivial to verify the conditions hold.
 \end{proof}
@@
 Next, we show some non-trivial cases of the fundamental theorem as
-top-level lemmas. For example, we can define the semantic analogue to the
-syntactic typing rule for variables (\rref{T-Var}).
+top-level lemmas, namely for \rref{T-Var}, \rref{T-Set}, \rref{T-Pi},
+\rref{T-Abs}, and \rref{T-App}.
+
 \begin{lemma}[ST-Var]
@@
   need to show that there exists some $[[i]]$ and $[[S]]$ such that
-  \begin{itemize}
-  \item $[[InterpR i A { rho } S]]$
-  \item $[[rho x in S]]$
-  \end{itemize}
+  $[[InterpR i A { rho } S]]$ and $[[rho x in S]]$ hold.
   By the definition of semantic context well-formedness, we deduce
@@
   level $[[i]]$ such that $[[G |= A : Set i]]$. By
-  the equivalence from Lemma~\ref{lemma:setinv}, there exists $[[S]]$
-  such that $[[InterpR i A {rho} S]]$. However, by the definition of
-  $[[rho |= G]]$, we know that $[[rho x in S]]$, which is exactly what
-  we need for the conclusion.
+  the equivalence from Lemma~\ref{lemma:setinv}, there exists an $[[S]]$
+  such that $[[InterpR i A {rho} S]]$. Finally, by the definition of
+  $[[rho |= G]]$, we know that $[[rho x in S]]$.
 \end{proof}
@@
   By unfolding the definition of $[[G |= \ x . b : Pi x : A . B]]$, we need to
-  show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
+  show that given some $[[rho |= G]]$, there exist some $[[i]]$ and
   $[[S]]$ such that $[[InterpR i (Pi x : A . B){rho}
@@
   $[[(\ x . b){rho} in S
-  ]]$. By Lemma~\ref{lemma:piinvalt}, the alternative inversion
+  ]]$. By Lemma~\ref{lemma:piinvalt}, the alternate inversion
   principle for \rref{I-Pi}, there exists some $[[S0]]$ such
-  that all following conditions hold:
+  that the following hold:
   \begin{itemize}
-  \item $[[InterpR i A{rho} S0]]$
+  \item $[[InterpR i A{rho} S0]]$;
   \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
     {rho .: x -> a}
-    S1 #) #)]]$
+    S1 #) #)]]$; and
   \item $[[S = { b | forall a, (# a in S0 implies forall
-      S1, (# InterpR i B {rho .: x -> a} S1 implies  b a in S1 #) #) }]]$
+      S1, (# InterpR i B {rho .: x -> a} S1 implies  b a in S1 #) #) }]]$.
   \end{itemize}
   To show that $[[(\ x . b){rho} in S]]$, we need to prove
-  that given $[[a in S0]]$,
+  that given $[[a in S0]]$ and
   $[[Interp I i B {rho .: x -> a} S1]]$, we have $[[(\x . b ){rho}
@@
   a => b {rho .: x -> a}]]$, it suffices to show that
-  $[[b {rho .: x -> a} in S1]]$, which is immediate from $[[G ++ x : A |= b :
-  B]]$ and the fact that the logical relation is deterministic and
-  cumulative (Lemma~\ref{lemma:logreldeterhet}).
+  $[[b {rho .: x -> a} in S1]]$.
+  By $[[G ++ x : A |= b : B]]$, Lemma~\ref{lemma:rhowfcons}, and $[[a in S0]]$,
+  we know that there exist some $[[j]], [[S2]]$ such that
+  $[[InterpR j B {rho .: x -> a} S2]]$ and $[[b {rho .: x -> a} in S2]]$.
+  Finally, by level-irrelevant functionality of the logical relation
+  (Lemma~\ref{lemma:logreldeterhet}), we have that $[[S1 = S2]]$
+  and thus $[[b {rho .: x -> a} in S1]]$.
 \end{proof}
@@
 such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a / x} {rho}
-S1]]$, or equivalently, $[[InterpR i B {rho .: x -> a {rho}} S1]]$ since
+S1]]$, or equivalently $[[InterpR i B {rho .: x -> a {rho}} S1]]$, since
 $[[B {a / x}{rho}]] = [[B {rho .: x -> a {rho}}]]$. By the premise $[[G |= b :
 Pi x : A . B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
-there exists some $[[i]]$ and $[[S0]]$ such that:
+there exists some $[[i]]$ and $[[S0]]$ such that the following hold:
   \begin{itemize}
-  \item $[[InterpR i A{rho} S0]]$
+  \item $[[InterpR i A{rho} S0]]$;
   \item $[[forall a0, (# a0 in S0 implies (# exists S1 , InterpR i B
     {rho .: x ->  a0}
-    S1 #) #)]]$
+    S1 #) #)]]$; and
   \item $[[forall a0, (# a0 in S0 implies (# forall
-      S1, (# InterpR i B {rho .: x -> a0} S1 implies  b {rho} a0 in S1 #) #) #)]]$
+      S1, (# InterpR i B {rho .: x -> a0} S1 implies  b {rho} a0 in S1 #) #) #)]]$.
   \end{itemize}
@@
 \begin{proof}
-  Proof by mutual induction over the derivation of $[[G |- a :
-  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follow
-  from Lemmas~\ref{lemma:semwffempty} and \ref{lemma:semwffcons}. The semantic typing rules
+  By mutual induction over the derivation of $[[G |- a : A]]$ and $[[|- G]]$.
+  The cases related to context well-formedness immediately follow
+  from Lemmas~\ref{lemma:semwffempty} and \ref{lemma:semwffcons}.
+  The semantic typing rules
   (Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp})
-  can be used to discharge their syntactic counterparts
-  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
-  cases not covered by the lemmas are similar to the ones already
-  shown.
+  are used to discharge their syntactic counterparts
+  (e.g. Lemma~\ref{lemma:stabs} for \rref{T-Abs}). The remaining
+  cases not covered by those lemmas are similar to the ones already shown.
 \end{proof}
 
-Recall the logical consistency property
-(Theorem~\ref{theorem:consistency}), which states that the judgment
-$[[empty |- a : Void ]]$ is not derivable. We now give a proof of the
-property using the fundamental lemma.
+We now give a proof of logical consistency (Theorem~\ref{theorem:consistency}),
+which states that the judgment $[[empty |- a : Void]]$ is not derivable,
+using the fundamental theorem.
 
 \begin{proof}
-  Suppose $[[empty |- a : Void]]$ is derivable, then by the
-  fundamental lemma, we have $[[empty |= a : Void]]$, which states
-  that for all $[[rho |= empty]]$, and for all $[[j]]$, $[[S]]$ such
-  that $[[InterpR j Void S]]$, we have $[[a {rho} in S]]$. By
-  Lemma~\ref{lemma:rhowfempty}, any $[[rho]]$ we pick trivially
-  satisfies $[[rho |= G]]$. For convenience, we pick $[[rho]]$ as
-  $[[idtm]]$, though any $[[rho]]$ would work since
-  $[[empty |- a : Void]]$ ensures there is no free variable in
-  $[[a]]$. We have $[[a {idtm}]] = [[a in S]]$. By the $[[Void]]$
-  case of the inversion property (Lemma~\ref{lemma:interpinv}), we
-  know that $[[S]]$ must be the empty set, contradicting
-  the assumption that $[[a in S]]$.
+  Suppose $[[empty |- a : Void]]$ is derivable. By the
+  fundamental theorem, we have $[[empty |= a : Void]]$, which states
+  that for all $[[rho |= empty]]$, $[[j]]$, $[[S]]$ such
+  that $[[InterpR j Void S]]$, we have $[[a {rho} in S]]$.
+  By Lemma~\ref{lemma:rhowfempty}, any $[[rho]]$ we pick trivially
+  satisfies $[[rho |= G]]$; for convenience, we pick $[[idtm]]$,
+  so that we have $[[a in S]]$.
+  However, by the $[[Void]]$ case of the inversion property
+  (Lemma~\ref{lemma:interpinv}), $[[S]]$ must be the empty set,
+  which is a contradiction.
 \end{proof}
 
-Our soundness theorem also tells us something about closed terms of type 
-$[[Bool]]$; they either reduce to $[[true]]$ or $[[false]]$.
+The fundamental theorem also tells us that closed terms of type 
+$[[Bool]]$ reduce to either $[[true]]$ or $[[false]]$.
 
@@
 \begin{proof}
-  The proof is similar to above, except that we use the $[[Bool]]$ case of the
-  inversion property.
+  Similar to above, using instead the $[[Bool]]$ case of the inversion property.
 \end{proof}
@@
 %
-\begin{figure}[H]
+\begin{figure}
   \[

### 2024-12-05 paper/paper.tex
Edits for Section 5

@@
 consistency will be proven in Section~\ref{sec:logrelproof}.
-
-The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
+Its syntax is given in Figure~\ref{fig:syntax}. As a dependent
 type theory, terms and types are collapsed into the same syntactic
@@
 \begin{figure}
-\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
+\drules[I]{$[[Interp I i A S]]$}{Logical relation}{Set, Void, Bool, Eq, Red}
 \[ \drule[width=5in]{I-Pi} \]
-\caption{Logical relation for \lang}
+\caption{Logical relation for \lang{}}
 \label{fig:logrel}
@@
 
-The logical relation for \lang{}, which takes the form $[[Interp I i A
-S]]$, is defined as an inductively generated relation in Figure~\ref{fig:logrel}.
+The logical relation\footnote{\dotv{semtyping.v}{InterpExt}} for \lang{},
+which takes the form $[[Interp I i A S]]$,
+is defined as an inductively generated relation in Figure~\ref{fig:logrel}.
 % The logical relation takes the form $[[Interp I i A S]]$.
@@
 \label{sec:extension}
-In this section, we show how the logical relation from
-Section~\ref{sec:logreldep} can be extended to show the existence of $\beta$
-normal forms for (open and closed) well-typed terms.  In other words, we prove
-that it is possible to repeatedly use the parallel reduction relation to
-reduce any term to its unique normal form, where no further (non-identity)
-reductions can be applied. This result can be used to show that our type
-conversion relation is decidable.% , as computation of the normal form will
-% always terminate.\yl{dependent on reduction strategy}
-
-The goal of this section is also to demonstrate that our logical relations
-proof technique can be extended to reason about the reduction properties of
-open terms, not just the reduction of terms after closing substitutions.
-Reasoning about open terms is particularly important for dependently-typed
-languages because type checking involves working with open terms.
+
+In this section, we extend the logical relation in Figure~\ref{fig:logrel}
+to show the existence of $\beta$-normal forms for both open and closed
+well-typed terms. More precisely, we prove it possible to repeatedly apply
+parallel reduction to reduce a term to its unique normal form. Consequently,
+this shows that the convertibility relation is decidable.
+
+This extension of the logical relation demonstrates that our proof technique
+can also be used to reason about the reduction properties of open terms, not
+just of terms after closing substitutions. Reasoning about open terms is
+especially important for dependently typed languages, as type checking involves
+working with open terms.
 \scw{
@@
 }
-While this extension employs well-known techniques, it continues to be short and
-demonstrates the robustness of our initial framework.
-
-We begin this part with a description of the $\beta$-normal forms
-of \lang{}.
-%
+While the extension use well-known techniques,
+\jc{Does this need a citation?}
+it remains short and demonstrates the robustness of our initial framework.
+
 \begin{figure}
@@
 \end{figure}
-%
-The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
-neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here?
+
+Figure~\ref{fig:nenf} gives the $\beta$-neutral forms $[[e]]$ and
+$\beta$-normal forms $[[f]]$ of \lang{}. Neutral terms consist of variables and
+elimination forms that do not reduce any further, while normal terms consist of
+neutral terms and other introduction forms. We use the judgment forms $[[ne a]]$
+and $[[nf a]]$ to indicate that there exists a term $[[e]]$ (resp. $[[f]]$)
+such that $[[a = e]]$ (resp. $[[a = f]]$). Additionally, we define predicates
+to describe terms that weak normalization: reduction to neutral or normal forms
+by parallel reduction.
+
+% The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
+% neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here?
 We can be specific and say that that the only reductions available for these terms are identity reductions and cite
-\footnote{\dotv{normalform.v}{nf\_refl}} }.\yl{I always think of
-normal form as the specific definition that says the relation can't
-step. We can refer to nfrefl but that requires some explanation about
-the definition of normal form and the complication that it doesn't
-hold in $\eta$ (maybe it's fine if we just don't mention it in the
-$\eta$ case)}\scw{A terminal form is a syntactic characterization of terms that don't step 
-according to a particular relation. A normal form is a terminal form of a normalizing relation.}
- Instead of the
-metavariables $e$ and $f$, we also
-use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
-$[[e]]$ or $[[f]]$ such that $[[a = e]]$ or $[[a = f]]$.
-
-The predicates $[[wne a]]$ and $[[wn
-a]]$ describe terms that can evaluate into $\beta$-neutral or
-$\beta$-normal form through parallel reduction and are defined as
-follows.
-\[
-\begin{array}{llcl}
-\mbox{\emph{weakly normalizes to a neutral form}} & [[wne a]] &\iff& \exists [[e]], [[a =>+ e]] \\
-\mbox{\emph{weakly normalizes to a normal form}}  & [[wn a]]  &\iff& \exists [[f]], [[a =>+ f]] \\
-\end{array}
-\]
-
-\begin{figure}[H]
-  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
-  \caption{Extended logical relation (new and changed rules) }
+% \footnote{\dotv{normalform.v}{nf\_refl}} }.\yl{I always think of
+% normal form as the specific definition that says the relation can't
+% step. We can refer to nfrefl but that requires some explanation about
+% the definition of normal form and the complication that it doesn't
+% hold in $\eta$ (maybe it's fine if we just don't mention it in the
+% $\eta$ case)}\scw{A terminal form is a syntactic characterization of terms that don't step 
+% according to a particular relation. A normal form is a terminal form of a normalizing relation.}
+
+\begin{definition}[Weak normalization]
+  We define weak normalization of a term $[[a]]$ to a neutral form, written
+  $[[wne a]]$, or to a normal form, written $[[wn a]]$, as the following:
+  \begin{align*}
+    [[wne a]] &\iff \exists [[e]], [[a =>+ e]] \\
+    [[wn a]]  &\iff \exists [[f]], [[a =>+ f]]
+  \end{align*}
+\end{definition}
+
+\begin{figure}
+  \drules[I]{$[[Interp I i A S]]$}{Extended Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
+  \caption{Extended logical relation for \lang{} (new and changed rules)}
   \label{fig:logrelopen}
@@
 
-The updated logical relation is shown in
-Figure~\ref{fig:logrelopen}.~\footnote{\dotv{semtypingopen.v}{InterpExt}}
-There is one new rule in this figure, \rref{I-Ne}. In a non-empty context, a
-type itself may evaluate to a neutral term and in turn can only be inhabited
-by neutral terms.  Otherwise, the rest of the rules in this figure are updates
-to the analogous rules in Figure~\ref{fig:logrel}. Note that, we omit the
-rules for the function and universe cases because they are identical to the
-original version.
-
-% The idea of a term blocked from $\beta$ reduction is
-% captured by $[[e]]$, the set of neutral terms. We augment the
-% interpretation of booleans and the empty types with terms that
-% evaluate to neutral terms. Furthermore, types that are neutral terms
-% can also be assigned a meaning since they may be inhabited by other
-% neutral terms.
-The changes to \rref{I-Bool} and \rref{I-Void}
-follow the same pattern: an open term of type $[[Bool]]$ does not necessarily
-reduce to $[[true]]$ or $[[false]]$, but may reduce to a variable, or more
-generally, a neutral term. Likewise, while the $[[Void]]$ type remains
-uninhabited under an empty context, it may be inhabited when there is a
-variable in the context that has type $[[Void]]$ or that can be eliminated to
-type $[[Void]]$.
-
-\scw{Cassia didn't understand this part}
-The rule for equality type $[[a ~ b : A]]$ is augmented with the precondition that
-$[[a]]$, $[[b]]$, and $[[A]]$ are all normal forms because
-otherwise our model would include equality types that are themselves
-not normalizing.
-Furthermore, the condition $[[a <=> b]]$ is only
-required when the equality proof reduces to $[[refl]]$. If the proof
-term reduces to a neutral term, then there is nothing we need to show
-about the relationship between $[[a]]$ and $[[b]]$.
+The modifications to the logical relation\footnote{\dotv{semtypingopen.v}{InterpExt}}
+are given in Figure~\ref{fig:logrelopen}. The new \rref{I-Ne} asserts that
+only terms that reduce to neutral forms inhabit the interpretation of neutral
+types. The remaining rules are updates to existing rules; we omit those for
+function types and universes as they are unchanged.
+
+The changes in \rref{I-BoolNew} and \rref{I-VoidNew} follow the same pattern. An open
+term of type $[[Bool]]$ may reduce to $[[true]]$ or $[[false]]$ as before, but
+may also reduce to a neutral term, such as a variable. Likewise, while in an
+empty context the interpretation of $[[Void]]$ remains empty, if the context
+is not empty, then it may contain, for instance, a variable in the context of
+type $[[Void]]$ or that can be eliminated into type $[[Void]]$.
+
+In \rref{I-EqNew}, we add preconditions asserting that $[[a]]$, $[[b]]$, and
+$[[A]]$ are normal so that we only model normalizing equality types.
+Furthermore, we only require $[[a <=> b]]$ when the equality proof reduces to
+$[[refl]]$. If the proof term reduces to a neutral term, there is nothing we
+can assert about the relationship between $[[a]]$ and $[[b]]$.
 

### 2024-12-05 paper/paper.tex
idk what to do about antirenaming

@@
 Because we are working with open terms, we need a few additional syntactic
-lemmas about reduction. First, a renaming $\xi$ is a generalization of
+lemmas about reduction. First, a renaming $[[xi]]$ is a generalization of
 weakening when working with simultaneous substitutions. It consistently maps the
@@
 always recover the result of the reduction without the renaming.
-%
+
 \scw{This lemma is only used to prove \texttt{wn\_antirenaming}, which is then
-used more generally. Maybe we should replace it with
-that?}\yl{Anti-renaming is mentioned once in Section 6 when we talk abou how confluence with $\eta$
+used more generally. Maybe we should replace it with that?}
+\yl{Anti-renaming is mentioned once in Section 6 when we talk about how confluence with $\eta$
 depends on anti-renaming for Par. I don't think that discussion in
-Section 6 is very important so I'm fine with replacing it  }
+Section 6 is very important so I'm fine with replacing it}
+
 \begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]

### 2024-12-09 paper/paper.tex
Add a little more foreshadowing in Section 3

@@
 Our system is predicative because the interpretation of the $[[i]]$th universe
-is dependent only on universes strictly below $[[i]]$,
-which have already been defined.
+is dependent only on universes strictly below $[[i]]$.
+% which have already been defined.
 This restriction ensures that the relation is well defined; otherwise, the
@@
 \subsection{Properties of the Logical Relation}
+
 In the rest of this section, we develop the theory of our logical relation
-with the goal of showing four key facts: irrelevance
+with the goal of showing four key properties: irrelevance
 (Lemma~\ref{lemma:logrelcoherence}), functionality
@@
 (Lemma~\ref{lemma:logrelcumulativity}), and the backward closure property
-(Lemma~\ref{lemma:logrelbackclos}).  For the majority of the properties that
-we prove in this section, we do not need any information about the
-parameterized function $[[I]]$.  Each property about $[[InterpR i A S]]$
-follows as a corollary of a property about $[[Interp I i A S]]$ with no or few
-assumptions imposed on $[[I]]$. As a result, we state most of our lemmas in
-terms of $[[Interp I i A S]]$ without duplicating them in terms of
-$[[InterpR i A S]]$.
+(Lemma~\ref{lemma:logrelbackclos}).
+These properties semantically correspond to ones that hold for syntactic typing,
+and allow us to prove that syntactic typing implies semantic typing, which we
+define in the next section.
+
+In particular, irrelevance states that convertible
+types have the same interpretation, which is used to interpret \rref{T-Conv}.
+Functionality with cumulativity shows that all interpretations of a type are
+the same across all universe levels, which allows us to treat a type that
+appears twice in a judgment uniformly, such as the function domain type in
+\rref{T-App}. Finally, backward closure ensures that anything that reduces to
+a term in the interpretation of a type is itself in the interpretation, which
+we use to handle $\beta$-reductions in the cases for introduction forms.
+
+For the majority of the properties that we prove in this section, we need no
+information about the parameterized function $[[I]]$.  Each property about
+$[[InterpR i A S]]$ follows as a corollary of a property about
+$[[Interp I i A S]]$ with no or few assumptions imposed on $[[I]]$. As a result,
+we state most of our lemmas in terms of $[[Interp I i A S]]$ without duplicating
+them in terms of $[[InterpR i A S]]$.
 

### 2024-12-09 paper/paper.tex
Reorganizing Section 5 up to Adequacy

@@
 \begin{figure}
-\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
+\drules[P]{$[[a => b]]$}{Parallel reduction}{AppAbs, IfTrue, IfFalse, JRefl}
 \caption{Parallel reduction ($\beta$ rules only) }
@@
 \begin{figure}
-  \[
-    \begin{array}{llcl}
-       \beta\text{-}\mathit{neutral\ terms} &
-      [[e]] & ::= & [[x]]\ |\ [[e f]]\ |\ [[J e f]]\ |\ [[if e f
-                    f]] \\ \\
-      \beta\text{-}\mathit{normal\ terms} &
-      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi x : f . f]]\
-                    |\ [[f ~ f : f]]\\
-            & & |   & [[\ x . f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
-    \end{array}
-  \]
-  \caption{$\beta$-neutral and normal forms}
+  \begin{align*}
+    [[e]] & ::= [[x]] \mid [[e f]] \mid [[J e f]] \mid [[if e f f]] \mid [[absurd e]]
+    && \textit{Neutral terms} \\
+    [[f]] & ::= [[e]] \mid [[Set i]] \mid [[Pi x : f . f]] \mid [[Bool]] \mid [[f ~ f : f]] \mid [[Void]] \mid [[\ x . f]] \mid [[true]] \mid [[false]] \mid [[refl]]
+    && \textit{Normal terms}
+  \end{align*}
+  \caption{$\beta$-neutral and $\beta$-normal forms}
   \label{fig:nenf}
@@
 
-\begin{figure}
-  \drules[I]{$[[Interp I i A S]]$}{Extended Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
-  \caption{Extended logical relation for \lang{} (new and changed rules)}
-  \label{fig:logrelopen}
-\end{figure}
+Weakly normalizing terms are composed of other weakly normalizing terms
+following the structure of neutral and normal forms. For example, an
+application normalizes to neutral form if the function normalizes to neutral
+form and the argument to normal form. Similar lemmas hold for
+function types\footnote{\dotv{normalform.v}{wn\_pi}},
+abstractions\footnote{\dotv{normalform.v}{wn\_abs}},
+conditionals\footnote{\dotv{normalform.v}{wne\_if}},
+and absurdity\footnote{\dotv{normalform.v}{wne\_absurd}}.
+
+\begin{lemma}[Application (wne)\footnote{\dotv{normalform.v}{wne\_app}}]
+  \label{lemma:wnewn}
+  If $[[wne a]]$ and $[[wn b]]$, then $[[wne (a b)]]$.
+\end{lemma}
+
+\begin{proof}
+  By induction over the length of reduction sequences in
+  $[[wne a]]$ and $[[wn b]]$.
+\end{proof}
 
@@
 function types and universes as they are unchanged.
+
+\begin{figure}
+  \drules[I]{$[[Interp I i A S]]$}{Extended logical relation}{Ne, VoidNew, BoolNew, EqNew}
+  \caption{Extended logical relation for \lang{} (new and changed rules)}
+  \label{fig:logrelopen}
+\end{figure}
 
@@
 
-Because we are working with open terms, we need a few additional syntactic
-lemmas about reduction. First, a renaming $[[xi]]$ is a generalization of
-weakening when working with simultaneous substitutions. It consistently maps the
-variables that appears in terms to other variables. If a renamed term has been reduced, we can
-always recover the result of the reduction without the renaming.
+Because we are working with open terms, we need additional lemmas for parallel
+reduction. First, it preserves neutral and normal forms.
+
+\begin{lemma}[Preservation of neutral and normal forms (p.r.)\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
+  \label{lemma:parnenf}
+  If $[[a => b]]$, then $[[ne a]]$ implies $[[ne b]]$,
+  and $[[nf a]]$ implies $[[nf b]]$.
+\end{lemma}
+
+Since neutral and normal forms have no redexes, if $[[a]]$ is neutral or normal,
+parallel reduction never $\beta$-reduces, so $[[a = b]]$ additionally holds,
+but preservation of neutral and normal forms is sufficient for our proof.
+
+Next, parallel reduction also satisfies an antirenaming property. A
+\emph{renaming} $[[xi]]$ is a special case of substitution where all variables
+are replaced by other variables. In this special case, if a renamed term
+reduces, then we can separate the reduction from the renaming, asserting that
+renaming never introduces new redexes.
+
+\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
+  \label{lemma:parantirenaming}
+  If $[[a < xi > => b0]]$, then there exists some $[[b]]$ such that
+  $[[b < xi > = b0]]$ and $[[a => b]]$.
+\end{lemma}
 
@@
 
-\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
-  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
-exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
-\end{lemma}
-
-We can show that parallel reduction preserves $\beta$-normal and
-neutral forms.
-\begin{lemma}[Par preserves $\beta$-neutral and normal forms\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
-  \label{lemma:parnenf}
-  If $[[a => b]]$, then
-  \begin{itemize}
-  \item $[[ne a]]$ implies $[[ne b]]$
-  \item $[[nf a]]$ implies $[[nf b]]$
-  \end{itemize}
-\end{lemma}
-Lemma~\ref{lemma:parnenf} could have been strengthened to say that if
-$[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
-$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes, parallel
-reduction cannot take any real reduction steps and can only step into a term
-itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
-sufficient.
+Together, these lemmas prove an inversion property for weakly normalizing forms:
+if an application of a term to a variable is weakly normalizing, then the term
+must also be weakly normalizing.
+
+\begin{lemma}\hspace{-0.5em}\footnote{\dotv{normalform.v}{ext\_wn}}
+  \label{lemma:extwn}
+  If $[[wn (a x)]]$, then $[[wn a]]$.
+\end{lemma}
+\begin{proof}
+  By induction over the length of the reduction sequence in $[[wn (a x)]]$.
+  When the reduction is a $\beta$-reduction, $[[a]]$ is a function and the
+  application reduces to a renaming by $[[x]]$, so the conclusion follows
+  from Lemma~\ref{lemma:parantirenaming}.
+\end{proof}
 
@@
 lemma can be proven in the same order, where the new cases due to \rref{I-Ne}
-and the augmentation of neutral terms to \rref{I-Void, I-Eq, I-Bool} can be
-immediately discharged by Lemma~\ref{lemma:parnenf}.
-
-Furthermore, Lemma~\ref{lemma:parnenf}, in its current weaker form, would
-still hold after we extend our equational theory with the function $\eta$
-rule, where parallel reduction can take $\eta$ steps but still preserves
-$\beta$-normal form.
-
-We also need to know that the $[[wne a]]$ and $[[wn a]]$ relations can be
-justified compositionally. For example, an application has a neutral form when
-the function has a neutral form and the argument has a normal form.
-
-\begin{lemma}[Wne application\footnote{\dotv{normalform.v}{wne\_app}}]
-  \label{lemma:wnewn}
-  If $[[wne a]]$ and $[[wn b]]$, then $[[wne (a b)]]$.
-\end{lemma}
-\begin{proof}
-  Immediate by induction over the length of the reduction sequences in
-  $[[wne a]]$ and $[[wn b]]$.
-\end{proof}
-
-Furthermore, if we know that an application of a term to a variable
-has a normal form, then we know that the term must have a normal form.
-\begin{lemma}[Wn extensionality\footnote{\dotv{normalform.v}{ext\_wn}}]
-  \label{lemma:extwn}
-  If $[[wn (a x)]]$, then $[[wn a]]$.
-\end{lemma}
-\begin{proof}
-  By induction over the length of the reduction sequence in $[[wn (a
-  x)]]$. The conclusion follows from Lemmas~\ref{lemma:parantirenaming} and
-  \ref{lemma:parnenf}.
-\end{proof}
+and the modifications to \rref{I-Void, I-Eq, I-Bool} can be discharged by
+Lemma~\ref{lemma:parnenf}.
+
+% \jc{Don't mention this right now; delay to $\eta$ section.}
+% Furthermore, Lemma~\ref{lemma:parnenf}, in its current weaker form, would
+% still hold after we extend our equational theory with the function $\eta$
+% rule, where parallel reduction can take $\eta$ steps but still preserves
+% $\beta$-normal form.
 
@@
   A]]$. To derive $[[wn (Pi x : A . B)]]$, it remains to show $[[wn B]]$. We
-  use the outermost induction hypothesis to show that $[[x]]$
-  semantically inhabits $[[A]]$, from which we derive $[[wn (B {x / x})]]$
+  use the outermost induction hypothesis to show that an arbitrary $[[y]]$
+  semantically inhabits $[[A]]$, from which we derive $[[wn (B {y / x})]]$
   and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).

### 2024-12-09 paper/paper.tex
Add related work about gluing

@@
 
-\paragraph{Axioms}
+\subsection{Axioms}
 Our Coq development assumes two axioms: functional extensionality
@@
 
-\paragraph{Encoding the logical relation in Coq}
+\subsection{Encoding the logical relation in Coq}
 We next discuss specific details of the Coq encoding of the logical relation
@@
 
-\paragraph{Automation}
+\subsection{Automation}
 \label{sec:automation}
@@
 precise and thus more directly encodable in proof assistants.
-Our logical relation resembles their definition of a
-semantic universe hierarchy, although we close our relation
-under expansion with respect to parallel reduction rather
-than weak-head reduction. \scw{Why is this important?} \yl{if you are
-  referring to the sentence about weak-head reduction, it can
-  be deleted because it's discussed in Section 9 already}
-Furthermore, \citet{Abel12} and
-\citet{abel2008betaeta} use their semantic universe hierarchy as a
-measure to define Kripke-style logical relations, from which they
-derive the correctness of their conversion algorithms. In our work, we
-use the semantic universe hierarchy directly in our definition of
-semantic typing because it is sufficient for our purposes
-(consistency and normalization).
+% Our logical relation resembles their definition of a
+% semantic universe hierarchy.
+Unlike \lang, their systems support type-directed $\eta$ laws. These
+type-directed equational rules complicate the proof significantly
+since the confluence of full reduction is not available until the
+fundamental theorem is proven. Therefore,
+instead of a logical predicate, a relational model with
+syntactic type annotations is needed to remove the early reliance on
+confluence, despite the fact that the underlying type theory is
+intensional.
+
+\citet{coquand:canonicity,coquand:prop} use a \emph{proof-relevant} logical
+relation to prove canonicity and normalization for intensional
+dependent type theories. % Intuitively, proof-relevant logical relations
+% can be viewed as the set theoretic augmented with information
+% about syntactic terms so the fundamental theorem can give us more
+Compared
+to the proof-irrelevant logical predicate we present, Coquand's proof
+has the benefit that it does not rely on the confluence of reduction,
+and can be easily extended to more $\eta$ laws without relying on a
+complex relational model. Instead of working with concrete syntax,
+Coquand works directly with an algebraic representation of dependent
+type theory called category with families (CwF)~\citep{hofmann:cwf},
+though the idea of proof-relevant logical relations for dependent
+types can be expressed independently of category
+theory~\citep{barras2012semantical}.
+
+More abstractly, the proof technique by Coquand can be viewed as an
+instance of a categorical construction called Artin Gluing
+~\citep{kaposi:gluing}. Different metatheoretic
+properties, such as normalization, parametricity, and
+$\Pi$-injectivity can be obtained by gluing the initial CwF with the
+corresponding CwF model. \citet{sterling:tait} introduces the notion of
+\emph{Synthetic Tait Computability}, which allows further
+simplification of such proofs by working in the internal language
+formed by gluing.
+However, proof-relevant logical relations are more demanding on the
+expressive power of the metatheory. For example, the mechanization of
+NbE by \citet{altenkirch:normalisation} relies on Quotient Inductive
+Inductive Types (QIIT)~\citep{kaposi:qiit}.
+
+% To support type-directed $\eta$ laws, \citet{Abel12,abel2008betaeta}
+% use a partial equivalence relation to model their type system. The 
+% prove the injectivity of $\Pi$ types cannot be derived 
+% Furthermore, \citet{Abel12} and
+% \citet{abel2008betaeta} use their semantic universe hierarchy as a
+% measure to define Kripke-style logical relations, from which they
+% derive the correctness of their conversion algorithms. In our work, we
+% use the semantic universe hierarchy directly in our definition of
+% semantic typing because it is sufficient for our purposes
+% (consistency and normalization).
 

### 2024-12-09 paper/paper.tex
Rewording proof of adequacy and normalization

@@
 \newcommand{\lang}{$\lambda^{\Pi}$\xspace}
+\newcommand{\CR}{\mathit{CR}}
 
@@
 
-\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
+\begin{lemma}[Par antirenaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
   \label{lemma:parantirenaming}
@@
 Before we can prove the fundamental theorem and derive the normalization
-property as its corollary, we need to additionally formulate and prove an
-\emph{adequacy property} about the logical relation.  This property, that the
-interpretation of each type is a \emph{reducibility candidate}, allows us to
-conclude that every term in each interpretation has a normal form. In the
-previous section, we only needed a property of the interpretation of the
-$[[Void]]$ type. However, for this section, we need to know something about
-the interpretation of every type.
-
-Furthermore, to prove this adequacy property, we need to strengthen it to also
-give us more information about neutral terms as we proceed by induction. In
-particular, we need to know that all terms that reduce to neutral forms are
-contained within the interpretation.
-Therefore, we formally define when a set is a \emph{reducibility candidate}
-(shortened as $CR$) as follows. Our definition of $CR$ is inspired by
-\citet{girard1989proofs}, but not identical since we only care about
-weak normalization.
-\begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
-  Let $[[S]]$ be a set of terms. We say that $[[S]] \in CR$
-  if and only if conditions $CR_1$ and $CR_2$ hold.
+property as its corollary, we need to formulate and prove an \emph{adequacy}
+property of the logical relation, which states that the interpretation of each
+type is a \emph{reducibility candidate} ($\CR$), which characterizes sets of
+weakly normalizing terms. In contrast to proving consistency, which only
+requires knowing that the interpretation of $[[Void]]$ is empty, adequacy
+allows us to conclude that all terms in \emph{every} interpretation have a
+normal form.
+
+To prove adequacy, we need to strengthen it to with more information about
+neutral terms as we proceed by induction. In particular, we need to know that
+all terms that reduce to neutral forms are contained within the interpretation.
+Therefore, we define reducibility candidates as follows, inspired by
+\citet{girard1989proofs}, but modified for weak normalization only. Adequacy
+then states that interpretations are in $\CR$.
+
+\begin{definition}[Reducibility Candidates ($\CR$)\footnote{\dotv{semtypingopen.v}{CR}}]
+  A set of terms $[[S]]$ is in $\CR$
+  if and only if conditions $\CR_1$ and $\CR_2$ hold.
   \begin{itemize}
-  \item $[[S]] \in CR_1 \iff\ [[forall a, (#  wne a implies a in S #)]]$
-  \item $[[S]] \in CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$
+  \item $[[S]] \in \CR_1 \iff\ [[forall a, (# wne a implies a in S #)]]$
+  \item $[[S]] \in \CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$
   \end{itemize}
@@
 
-We now state and prove the adequacy lemma.
 \begin{lemma}[Adequacy\footnote{\dotv{semtypingopen.v}{adequacy}}]
   \label{lemma:adequacy}
-  If $[[InterpR i A S]]$, then we have $[[S]] \in CR$.
-\end{lemma}
+  If $[[InterpR i A S]]$, then $[[S]] \in \CR$.
+\end{lemma}
+
 \begin{proof}
@@
   induction hypothesis that for all $[[j < i]]$, $[[InterpR j A S]]$
-  implies $[[S]] \in CR$. Our goal is to show $[[InterpR i A S]]$
-  implies $[[S]] \in CR$.
-
-  By Definition~\ref{fig:logrelrec}, we have the
-  equality $[[InterpR i A S]] = [[Interp I i A S]]$  where $[[I i]] :=
-  [[{A | exists S , InterpR i A S}]]$.
-  We then proceed by structural induction over the derivation of $[[Interp I i A
-  S]]$. The only interesting cases are \rref{I-Pi} and \rref{I-Set}.
-  The function case requires Lemmas~\ref{lemma:extwn} and
-  \ref{lemma:wnewn}, which we have shown earlier.
+  implies $[[S]] \in \CR$. Our goal is to show that $[[InterpR i A S]]$
+  implies $[[S]] \in \CR$.
+
+  By Definition~\ref{fig:logrelrec}, we have $[[Interp I i A S]]$,
+  where $[[I j]] := [[{A | exists S , InterpR j A S}]]$ for $[[j < i]]$.
+  We then proceed by structural induction on the derivation of
+  $[[Interp I i A S]]$. The interesting cases are \rref{I-Pi} and \rref{I-Set}.
+  For \rref{I-Pi}, proving $S \in \CR_1$ requires Lemma~\ref{lemma:wnewn}. When
+  proving $S \in \CR_2$, by the structural induction hypotheses, we have that
+  $[[wn (b a)]]$ for any $[[wne a]]$, while the goal is to show that $[[wn b]]$.
+  We pick an arbitrary $[[x]]$ and conclude by applying Lemma~\ref{lemma:extwn}
+  to $[[b x]]$.
 
   The \rref{I-Set} case is the most interesting. We must show that
-  for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
-  \in CR$. We immediately know that $[[{A | exists S, InterpR j A S
-  }]] \in CR_1$ by \rref{I-Ne}. It remains to show that $[[{A | exists S, InterpR j A S
-  }]] \in CR_2$, or equivalently, for all $[[A]]$, $[[InterpR j A S]]$
-  implies $[[wn A]]$. Suppose $[[InterpR j A S]]$ for an arbitrary
-  $[[A]]$. We have $[[InterpR j A S]]=[[Interp I j A S]]$ where $[[I]]$
-  has the same definition from earlier but its domain restricted to
-  numbers less than $[[j]]$. We perform another induction on
-  the derivation of $[[Interp I j A S]]$. All cases are trivial except for the case for
-  \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
-  A]]$. To derive $[[wn (Pi x : A . B)]]$, it remains to show $[[wn B]]$. We
-  use the outermost induction hypothesis to show that an arbitrary $[[y]]$
-  semantically inhabits $[[A]]$, from which we derive $[[wn (B {y / x})]]$
-  and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
+  for all $[[j < i]]$, $[[{A | exists S, InterpR j A S}]] \in \CR$.
+  We immediately know that $[[{A | exists S, InterpR j A S}]] \in \CR_1$
+  by \rref{I-Ne}. It remains to show that
+  $[[{A | exists S, InterpR j A S}]] \in \CR_2$, or equivalently that
+  $[[Interp I j A S]]$ implies $[[wn A]]$, where $[[I]]$ is defined as above
+  over $k < j$.
+
+  We assume $[[Interp I j A S]]$ and proceed once more by structural induction
+  on its derivation. All cases are trivial except for \rref{I-Pi}. The induction
+  hypothesis immediately gives $[[wn A]]$. To derive $[[wn (Pi x : A . B)]]$,
+  it remains to show $[[wn B]]$. We use the outermost induction hypothesis on
+  $[[Interp I j A S]]$ to show that an arbitrary $[[y]]$ semantically inhabits
+  $[[A]]$ by $\CR_1$, followed by the structural induction hypothesis to derive
+  $[[wn (B {y / x})]]$. We conclude $[[wn B]]$ by applying antirenaming
+  (Lemma~\ref{lemma:parantirenaming}).
 \end{proof}
@@
 The formulation of semantic well-typedness and the fundamental lemma from
-Section~\ref{sec:logrelproof} remains
-unchanged.
-The proof of the fundamental lemma\footnote{\dotv{soundnessopen.v}{soundness}} is still carried out by induction
-over the typing derivation, where the additional neutral term related
-cases are handled by Lemma~\ref{lemma:adequacy}, the adequacy property.
-
-The normalization property then follows as a corollary of the
-fundamental theorem.
+Section~\ref{sec:logrelproof} remains unchanged. The proof of the fundamental
+lemma\footnote{\dotv{soundnessopen.v}{soundness}} is still carried out by
+induction over the typing derivation, where the additional neutral term related
+cases are handled by adequacy. The normalization property then follows as a
+corollary of the fundamental theorem.
+
 \begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
@@
   that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
-  We pick the $[[rho]]$ to be the identity substitution $[[idtm]]$, which injects
-  variables as terms. The side condition $[[idtm |=
-  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
-  including variables, semantically inhabit any $[[S0]]$ where
-  $[[S0]]$ is the interpretation of some type. With our choice of
-  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
-  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
-  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
-  conclude that $[[wn a]]$ and $[[wn A]]$.
+  We pick the $[[rho]]$ to be the identity substitution $[[idtm]]$, and the
+  condition $[[idtm |= G]]$ is satisfied by adequacy, which says that variables,
+  as neutral terms, semantically inhabit the interpretations of the types in
+  $[[G]]$. Performing the identity substitutions, we then know that there exist
+  $[[i]]$ and $[[S]]$ such that $[[InterpR i A S]]$ and $[[a in S]]$. Finally,
+  by adequacy again, we conclude that $[[wn a]]$ and $[[wn A]]$.
 \end{proof}

### 2024-12-09 paper/paper.tex
Finished editing Section 5

@@
 open terms~\citep{harperkripke} in the simply typed lambda calculus.
-Indeed, a mechanization of normalization generalized to open terms
-appears in ~\citet{abel2019poplmark}.  In this setting, as above,
+Indeed, \citet{abel2019poplmark} mechanize normalization generalized
+to open terms. In this setting, as above,
 adequacy must be proven before the
-fundamental theorem so they can handle elimination rules such as
-\rref{T-App} where the scrutinee is a neutral term. % In
+fundamental theorem so that they can handle elimination rules such as
+\rref{T-App} where the function is a neutral term. % In
 % \citet{abel2019poplmark}, a variant of Lemma~\ref{lemma:extwn} is used
@@
 %
-Dependent types make the adequacy proof slightly more complicated because we
+Dependent types make the adequacy proof more complicated because we
 also need to know that every \emph{type} has a normal form, not just
@@
 
-Overall, despite the dependently typed setting, it is in fact reassuring that
+Overall, despite the dependently typed setting, it is reassuring that
 once we have laid the foundational technique for handling dependent types in
-our logical relation, the extension to open terms mostly boils down to
-properties that can be independently derived from the logical relation through
+our logical relation, the extension to open terms boils down to
+properties that can be derived independently from the logical relation through
 syntactic means.
@@
 \label{sec:betaeta}
+
 \citet{nbeincoq,decagda,martin-lof-a-la-coq}

### 2024-12-10 paper/paper.tex
Add extension section (WIP)

@@
 
-\begin{lemma}[Adequacy\footnote{\dotv{semtypingopen.v}{adequacy}}]
+\begin{lemma}[Adequacy]
   \label{lemma:adequacy}
+  \footnote{\dotv{semtypingopen.v}{adequacy}}
   If $[[InterpR i A S]]$, then $[[S]] \in \CR$.
@@
 represent $\beta$-neutral and $\beta$-normal forms, and the
-fundamental lemma shows us that every well-typed term has a
+fundamental theorem shows us that every well-typed term has a
 $\beta$-normal form. However, in the presence of the $\eta$ reduction
@@
 \fi
+
+\section{Extensions}
+In this section, we discuss how the normalization
+proof of \lang can be extended to incorporate features including
+natural numbers, dependent pairs, cumulative
+universes with subtyping, and $\eta$ law for functions. All these
+extensions have been implemented in our development. More advanced
+features such as type-directed $\eta$ laws are out of the scope of
+this paper, and we instead point the readers to existing work for
+techniques that can be readily adapted to our minimal development.
+
+\subsection{Natural Numbers}
+\begin{figure}
+  \drules[T]{$[[G |- a : A]]$}{Typing for Naturals}{Nat, Zero, Suc, Ind}
+  \drules[I]{$[[Interp I i A S]]$}{Logical Relation for Naturals}{Nat}
+  \drules[N]{$[[NatOk a]]$}{Natural values}{Zero, Neu, Suc}
+\caption{Rules for natural numbers}
+\label{fig:nat}
+\end{figure}
+
+We can replace the boolean type with the natural number type as our
+observable type. By introducing an infinite data type to \lang,
+we significantly increase its expressiveness.
+
+The syntactic and semantic rules for natural numbers can be found in
+Figure~\ref{fig:nat}. \Rref*{T-Zero, T-Suc} allow us to construct natural
+numbers. \Rref*{T-Ind} allows us to eliminate natural numbers
+through the induction principle where $[[a]]$ is the base case,
+$[[b]]$ the inductive case, and $[[c]]$ the natural number being
+eliminated.
+
+\Rref{I-Nat} gives us the semantic
+interpretation of natural number; a term $[[a]]$ semantically inhabits
+$[[Nat]]$ if $[[a]]$ reduces to some term $[[b]]$ such that $[[NatOk
+b]]$ holds.
+The $\vdash_\mathbb{N}$ judgment inductively characterizes the
+normal form of (potentially open) terms of type $[[Nat]]$ and is
+particularly useful in the \rref*{T-Ind} case of the fundamental
+theorem. Let $[[rho |= G]]$, the induction hypothesis $[[G |= c :
+Nat]]$ tells us that $[[c { rho } =>+ b ]]$ for $[[NatOk b]]$. The proof
+then follows by induction over the derivation of $[[NatOk b]]$. When
+$[[b]]$ is neutral, we conclude the proof with
+\nameref{lemma:adequacy}. When $[[b]]$ is $[[zero]]$, we conclude the
+proof with the outer induction hypothesis $[[G |= a : A {zero / x}]]$.
+When $[[b]]$ is a successor, we conclude the proof with the outer
+induction hypothesis $[[G ++ x : Nat ++ y : A |= b : A { succ x / x
+}]]$.
+
+\subsection{Cumulativity}
+\label{sec:cumulativity}
+The semantic model we build in \cref{sec:logreldep} satisfies
+\nameref{lemma:logrelcumulativity}.
+
 \section{Related Work}

### 2024-12-11 paper/paper.tex
Lots of extra stuff about encoding the logical relation

@@
 % Section~\ref{sec:conversionalgo}.
+
 \section{Mechanization}
 \label{sec:logrelmech}
-\begin{figure}[H]
-  \begin{minipage}{0.9\textwidth}
+
+\begin{figure}[h]
   \begin{tabular}{ |l |  c  | c | c| }
@@
   \end{tabular}
-  \end{minipage}
   \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
@@
 
-\subsection{Encoding the logical relation in Coq}
-We next discuss specific details of the Coq encoding of the logical relation
-presented in Section~\ref{sec:logreldep}.
+\subsection{Encoding the logical relation}
+
+As discussed, the logical relations $[[Interp I i A S]]$ in
+Figures~\ref{fig:logrel} and \ref{fig:logrelopen} are encoded as inductive
+definitions indexed by sets of terms representing the semantic inhabitants of
+$[[A]]$. In the Coq mechanization, we represent sets of terms as predicates
+over terms asserting whether the term is in the set, so $[[S]]$ has type
+\texttt{tm -> Prop}, and $[[I]]$ has type \texttt{forall j, j < i -> tm -> Prop},
+representing the set of types in universes strictly smaller than $[[i]]$.
+Because the overall inductive definition represents a functional relation
+between a type at a universe level and its interpretation as a set of terms,
+it too lives in \texttt{Prop}.
+
+Due to the limitations of eliminating inductives in \texttt{Prop}, the encoding
+of the rule for function types differs from the presentation in \rref{I-Pi},
+reproduced below.
+%
+\begin{mathpar}
+  \drule[width=\textwidth]{I-Pi}
+\end{mathpar}
+
+While the first and last premises are propositions, $[[S]]$ and $[[F]]$
+represent sets of terms, which are not propositions. While this is not an issue
+when proving properties of the logical relation, which never require projecting
+out these sets, it is an issue when proving the fundamental theorem, and in
+particular proving the function case in Lemma~\ref{lemma:stpi}. By the definition
+of semantic typing, the induction hypothesis (after some instantiation) gives
+%
+$$\forall [[a in S]], \exists [[i]], [[S0]] \text{ such that }
+  [[InterpR i B {a / x} S0]] \text{ and } [[b a in S0]].$$
+
+We would like to define $[[F]]$ as the function that projects out $[[S1]]$ from
+the induction hypothesis, but the existential quantification being a proposition
+prevents us from doing so without the axiom of choice. To avoid relying on
+classical reasoning, we instead encode $[[F]]$ as a functional relation
+$[[R in S * PowerSet STm]]$ such that
+$\forall [[a]], \exists [[S0]], [[( a , S0 ) in R]]$ holds. The actual encoding
+is given as \rref{I-PiCoq} below.
+%
+\begin{mathpar}
+  \drule[width=\textwidth]{I-PiCoq}
+\end{mathpar}
+
+From this rule, we are still able to derive \rref{I-PiAlt} as before. The
+benefit of working with the derived rule is that it does not involve $[[R]]$,
+which is uniquely determined to be
+$[[{ (a , S0 ) | a in S implies Interp I i B { a /x } S0 }]]$ as before.
+
+An alternative to these various encodings in Coq is to use
+\emph{induction--recursion} (IR)~\citep{IR}, which is available in Agda~\citep{agda}.
+IR permits mutually defining an inductive definition along with a recursive function
+on the inductive. We can change our perspective on the logical relation and
+view it as two separate pieces: an inductive interpretation $\llbracket A \rrbracket^{i}$
+of types, and a recursive interpretation $[[a in InterpR i A]]$ of types as sets of terms.
+The rule for function types and its interpretation would be defined as follows.
+%
+\begin{mathpar}
+  \drule[width=\textwidth]{IIR-Pi} \\
+  [[b in InterpIR i Pi x : A . B]] \coloneqq
+    [[forall a, (# a in InterpIR i A implies b a in InterpIR i B {a / x} #)]]
+\end{mathpar}
+
+Because the inductive definition no longer needs to quantify over all sets of terms,
+we can work entirely within Agda's lowest \texttt{Set} universe without worrying
+about impredicativity or universe polymorphism. However, to define the above,
+we still need an intermediate definition involving the interpretation $[[I]]$
+of types at $[[j < i]]$ and tie the knot as before. While the same properties
+are provable, the inductive--recursive logical relation changes the structure
+and the order of these proofs; for instance, a weaker form of functionality
+is provable before and is required to prove forward preservation.
+
+Further details of the IR encoding can be found in our Agda mechanization of
+the consistency of \lang. Due to the lack of automation support, proofs about
+syntactic properties of terms and proofs using well-founded induction on levels
+are written manually. Even so, the total number of nonblack, noncomment lines
+of code remains small at below 1200 lines, and functional extensionality is
+still the only axiom used.
 
@@
 % Lemma~\ref{lemma:piintroalt}).
-In the Coq mechanized proof, % we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. 
-the definition of $[[Interp I i A S]]$ has type
-\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.
+% In the Coq mechanized proof,
+% we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. 
+% the definition of $[[Interp I i A S]]$ has type
+% \texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.
 
@@
 
-However, if desired, we could consistently replace the use of \texttt{Prop}
-with Coq's predicative sort \texttt{Type} in the definition of
-$[[Interp I i A S]]$. This alternative definition could be part of
-the interpretation for any \emph{finite} number of universes. The use of
-\texttt{Type} becomes troublesome only when we attempt to define
-$[[InterpR i A S]]$, the top-level logical relation 
-(Definition~\ref{fig:logrelrec}) that recursively calls itself at smaller universe
-levels. Therefore, the one feature of \lang{} that truly requires
-impredicativity is its countable universe hierarchy.
-
-The definition of $[[Interp I i A S]]$ has an almost one-to-one
-correspondence to the Coq definition. The main difference is the
-specification of $[[I]]$. In
-Section~\ref{sec:logreldep}, we define $[[I]]$ as a function over
-numbers less than $[[i]]$, the universe level. In Coq, we only require
-$[[I]]$ to be a function with the set of natural numbers as its domain.
-In the Coq encoding of $[[InterpR i A S]]$, we define $[[I]] \in [[SNat -> PowerSet STm]]$ as follows.
-\begin{equation*}
-  \begin{split}
-    [[I j]] &=
-     \begin{cases}
-      \ [[{A | exists S , InterpR j A S}]] & \text{when } j < i \\
-      \ [[emptyset]] & \text{otherwise}
-    \end{cases}
-  \end{split}
-\end{equation*}
-Since $[[I]]$ is only applied to numbers strictly less than $[[i]]$ in
-\rref{I-Set}, we can retroactively show that the set we return in the $j
-\geq i$ case is junk data that does not affect the result of the logical
-relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Definition~\ref{fig:logrelrec}.
-
-\Rref{I-PiCoq} shows how \rref{I-Pi} is actually encoded in our
-mechanized proof.
-\begin{center}
-  \drule[width=5in]{I-PiCoq}
-\end{center}
-Compared to \rref{I-Pi}, \rref{I-PiCoq} replaces the function
-$[[F]]$ with a total relation $[[R]]$. The equivalence of these two
-rules follows from the fact that the logical relation is a
-partial function (Lemma~\ref{lemma:logreldeter}). In set-theoretic
-notation, \rref{I-Pi} is more readable. However, if we want to encode
-the same rule in Coq, we must encode $[[F]]$ as a relation (with type
-\texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the functionality constraint:
-\mintinline{coq}{forall a S0 S1, F a S0 -> F a S1 -> S0 = S1}.
-In comparison, \rref{I-PiCoq} does not require this side condition and
-results in a simpler definition.
-
-We note that we cannot ascribe $[[F]]$ the type \texttt{tm -> (tm ->
-  Prop)} since Coq requires functions of such type to be
-computable. While defining $[[F]]$ as a computable Coq function rather than a
-functional relation does result in a concise encoding of \rref{I-Pi},
-we will have trouble instantiating $[[F]]$ with the logical relation,
-which is defined as a relation that we prove to be functional, rather
-than a computable function.
+% However, if desired, we could consistently replace the use of \texttt{Prop}
+% with Coq's predicative sort \texttt{Type} in the definition of
+% $[[Interp I i A S]]$. This alternative definition could be part of
+% the interpretation for any \emph{finite} number of universes. The use of
+% \texttt{Type} becomes troublesome only when we attempt to define
+% $[[InterpR i A S]]$, the top-level logical relation 
+% (Definition~\ref{fig:logrelrec}) that recursively calls itself at smaller universe
+% levels. Therefore, the one feature of \lang{} that truly requires
+% impredicativity is its countable universe hierarchy.
+
+% The definition of $[[Interp I i A S]]$ has an almost one-to-one
+% correspondence to the Coq definition. The main difference is the
+% specification of $[[I]]$. In
+% Section~\ref{sec:logreldep}, we define $[[I]]$ as a function over
+% numbers less than $[[i]]$, the universe level. In Coq, we only require
+% $[[I]]$ to be a function with the set of natural numbers as its domain.
+% In the Coq encoding of $[[InterpR i A S]]$, we define $[[I]] \in [[SNat -> PowerSet STm]]$ as follows.
+% \begin{equation*}
+%   \begin{split}
+%     [[I j]] &=
+%      \begin{cases}
+%       \ [[{A | exists S , InterpR j A S}]] & \text{when } j < i \\
+%       \ [[emptyset]] & \text{otherwise}
+%     \end{cases}
+%   \end{split}
+% \end{equation*}
+% Since $[[I]]$ is only applied to numbers strictly less than $[[i]]$ in
+% \rref{I-Set}, we can retroactively show that the set we return in the $j
+% \geq i$ case is junk data that does not affect the result of the logical
+% relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Definition~\ref{fig:logrelrec}.
+
+% \Rref{I-PiCoq} shows how \rref{I-Pi} is actually encoded in our
+% mechanized proof.
+% \begin{center}
+%   \drule[width=5in]{I-PiCoq}
+% \end{center}
+% Compared to \rref{I-Pi}, \rref{I-PiCoq} replaces the function
+% $[[F]]$ with a total relation $[[R]]$. The equivalence of these two
+% rules follows from the fact that the logical relation is a
+% partial function (Lemma~\ref{lemma:logreldeter}). In set-theoretic
+% notation, \rref{I-Pi} is more readable. However, if we want to encode
+% the same rule in Coq, we must encode $[[F]]$ as a relation (with type
+% \texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the functionality constraint:
+% \mintinline{coq}{forall a S0 S1, F a S0 -> F a S1 -> S0 = S1}.
+% In comparison, \rref{I-PiCoq} does not require this side condition and
+% results in a simpler definition.
+
+% We note that we cannot ascribe $[[F]]$ the type \texttt{tm -> (tm ->
+%   Prop)} since Coq requires functions of such type to be
+% computable. While defining $[[F]]$ as a computable Coq function rather than a
+% functional relation does result in a concise encoding of \rref{I-Pi},
+% we will have trouble instantiating $[[F]]$ with the logical relation,
+% which is defined as a relation that we prove to be functional, rather
+% than a computable function.
 
@@
 
-\subsection{Automation}
+\subsection{Automation in Coq}
 \label{sec:automation}
+
 Our Coq mechanization heavily uses automation, supported by the tools

### 2024-12-13 paper/paper.tex
Update numbers

@@
 \begin{figure}[h]
-  \begin{tabular}{ |l |  c  | c | c| }
+  \centering
+  \begin{tabular}{ | l |  c  | c | c | c | }
     \hline
-    & Consistency & Normalization & Syntactic metatheory \\
+    & Consistency & Normalization & Safety\\
     \hline
-    Syntactic typing         &  83 & =   & = \\
-    Untyped reduction        & 344 & =   & = \\
-    Neutral and normal forms & -   & 273 & - \\
-    Logical relation         & 338 & 430 & - \\
-    Semantic soundness       & 192 & 211 & - \\
-    Syntactic soundness      & -   & -   & 629 \\
+    Libraries            &  15 & $=\joinrel=$ & $=\joinrel=$ \\
+    Syntactic typing     &  87 & $=\joinrel=$ & $=\joinrel=$ \\
+    Untyped reduction    & 345 & $=\joinrel=$ & $=\joinrel=$ \\
+    Neutral/normal forms & --- & 287  & ---  \\
+    Logical relation     & 295 & 385  & ---  \\
+    Semantic soundness   & 195 & 215  & ---  \\
+    Syntactic soundness  & --- & ---  & 655  \\
     \hline
-    Total                    & 957 & 1341 & 1056 \\
+    Total                & 934 & 1331 & 1099 \\
     \hline
   \end{tabular}
-  \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
-left. The marker - indicates the file does not
-contribute to the total.}
+  \caption{Nonblank, noncomment lines of code of the Coq development.
+    The $=\joinrel=$ marker indicates that the line count is the same as the column to the left.
+    The --- marker indicates the file does not contribute to the total.}
   \label{fig:linecount}
@@
 To demonstrate the scale of our proof scripts, Figure~\ref{fig:linecount}
-shows the number of non-blank, non-comment lines of code\footnote{calculated
-  by the \texttt{tokei} tool, available from
-  \url{https://github.com/XAMPPRocky/tokei}.} for each file of our development,
-including the base consistency proof from Section~\ref{sec:logreldep} and
-\ref{sec:logrelproof} and the extension to $\beta$-normalization from
+shows the number of non-blank, non-comment lines of code%
+  \footnote{calculated by \texttt{tokei}:
+    \url{https://github.com/XAMPPRocky/tokei} \hfill}
+for each file of our development,
+including the base consistency proof from Sections~\ref{sec:logreldep} and
+\ref{sec:logrelproof}, along with the extension to $\beta$-normalization from
 Section~\ref{sec:extension}. For comparison, we have also proven syntactic
-type soundness through preservation
+type safety through preservation
 \footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
@@
 The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
-1568 lines of non-blank, non-comment lines of code. We choose not to include
+1549 lines of non-blank, non-comment lines of code. We choose not to include
 it in the chart, because of slight differences in lemma dependencies for
@@
 
-The Autosubst 2 tool takes our 13 line syntax specification, written in
+The Autosubst 2 tool takes our 16-line syntax specification, written in
 higher-order abstract syntax, and generates the Coq syntax specification,
 renaming and substitution functions, and lemmas and tactics that allow
-reasoning about those functions. The auto-generated syntax file (291 LOC)
-and other Autosubst library files are also not included in the figure.
+reasoning about those functions. The autogenerated syntax file and other
+Autosubst library files (516 LOC total) are not included in the figure.
 
 \subsection{Axioms}
+
 Our Coq development assumes two axioms: functional extensionality
@@
 $[[S1]]$ are equal, it suffices to show the extensional property that
-$\forall x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
-occasionally in our presented proofs.
-%
-However, in Coq, sets of terms ($[[PowerSet STm]]$) are encoded as the 
-type \texttt{tm
-  -> Prop}, a predicate over \lang{} terms.  In axiom-free Coq, predicates do
-not come with the extensionality property. Given two predicates $P$ and $Q$,
-we cannot conclude that $P = Q$ when given a proof of
-$\forall x, P(x) \iff Q(x)$. But this is exactly the statement of 
-predicate extensionality, an immediate corollary from functional
-extensionality and propositional extensionality.
+$\forall x, x \in [[S0]] \iff x \in [[S1]]$. We use this property in a few
+proofs, notably for inversion (Lemma~\ref{lemma:interpinv}) and functionality
+(Lemma~\ref{lemma:logreldeter}), where we deal with equalities of
+the interpretations, which are sets of terms.
+
+In the Coq mechanization, sets of terms $[[PowerSet STm]]$ are encoded as the
+predicates over terms \texttt{tm -> Prop} asserting whether the term is in the
+set. We want to be able to show that two predicates $P$ and $Q$ are equal when
+$\forall x, P(x) \iff Q(x)$ holds, but this predicate extensionality property
+does not hold in axiom-free Coq, and requires both functional and propositional
+extensionality.
 
@@
 definitions indexed by sets of terms representing the semantic inhabitants of
-$[[A]]$. In the Coq mechanization, we represent sets of terms as predicates
-over terms asserting whether the term is in the set, so $[[S]]$ has type
-\texttt{tm -> Prop}, and $[[I]]$ has type \texttt{forall j, j < i -> tm -> Prop},
-representing the set of types in universes strictly smaller than $[[i]]$.
-Because the overall inductive definition represents a functional relation
-between a type at a universe level and its interpretation as a set of terms,
-it too lives in \texttt{Prop}.
+$[[A]]$. $[[S]]$ therefore has type \texttt{tm -> Prop}, while $[[I]]$ has type
+\texttt{forall j, j < i -> tm -> Prop}, representing a set of types in
+universes strictly smaller than $[[i]]$. Because the overall inductive
+definition represents a functional relation between a type at a universe level
+and its interpretation as a set of terms, it too lives in \texttt{Prop}.
 
@@
 are written manually. Even so, the total number of nonblack, noncomment lines
-of code remains small at below 1200 lines, and functional extensionality is
+of code remains small at below 1300 lines, and functional extensionality is
 still the only axiom used.
@@
 language. \citet{weirich:systemd} define System D, a core calculus of
-dependent Haskell, and prove the syntactic type soundness of the type
+dependent Haskell, and prove the syntactic type safety of the type
 system. Because System D includes nontermination, they proved

### 2024-12-13 beta/count-beta.sh

@@ -1,2 +1,2 @@
 #!/usr/bin/env sh
-tokei -f semtypingopen.v soundnessopen.v typing.v join.v normalform.v
+tokei -f imports.v semtypingopen.v soundnessopen.v typing.v join.v normalform.v

### 2024-12-13 beta/count-consistency.sh

@@ -1,2 +1,2 @@
 #!/usr/bin/env sh
-tokei -f semtyping.v soundness.v typing.v join.v
+tokei -f imports.v semtyping.v soundness.v typing.v join.v

### 2024-12-13 beta/count-safety.sh

@@ -1,0 +1,2 @@
+#!/usr/bin/env sh
+tokei -f imports.v typing.v join.v syntactic_soundness.v

### 2024-12-13 paper/paper.tex
Minor edits through Section 8

@@
 
+\jc{Delete this section}
+
 \citet{nbeincoq,decagda,martin-lof-a-la-coq}
@@
 
-The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
-1549 lines of non-blank, non-comment lines of code. We choose not to include
-it in the chart, because of slight differences in lemma dependencies for
-untyped reduction and normal forms that make the comparison less
-informative. However, when compared to the $\beta$-normalization extension,
-the $\beta\eta$ extension has the same line count in the definition of the
-logical relation and the semantic soundness proof.
+% The $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises
+% 1549 lines of non-blank, non-comment lines of code. We choose not to include
+% it in the chart, because of slight differences in lemma dependencies for
+% untyped reduction and normal forms that make the comparison less
+% informative. However, when compared to the $\beta$-normalization extension,
+% the $\beta\eta$ extension has the same line count in the definition of the
+% logical relation and the semantic soundness proof.
 
@@
 In addition to these generated definitions, Autosubst 2 provides a powerful
-tactic \texttt{asimpl} that can be used to prove the equivalence of two terms
+\texttt{asimpl} tactic used to prove the equivalence of two terms
 constructed using the primitive operators provided by the framework. This
@@
 that if $[[a => b]]$, then $[[b]]\Rightarrow [[a]]^*$, where $[[a]]^*$
-is the Takahashi translation~\citep{takahashi-parallel-reduction}
+is the Takahashi translation~\citep{takahashi-parallel-reduction},
 which roughly corresponds to simultaneous reduction of the redexes in
@@
 
-The flag \texttt{lq:on} tunes CoqHammer's search algorithm.  While this flag
-appears arcane, when developing our proof scripts we never specify this option
-manually. Instead, we first invoke the \texttt{best} tactic provided by
+The flag \texttt{lq:on} tunes CoqHammer's search algorithm, which is never
+specified manually. Instead, we first invoke the \texttt{best} tactic provided by
 CoqHammer, specifying only the \texttt{inv}, \texttt{ctrs}, and lemmas that we
@@
 readers who wish to understand the mechanized proof. Although automation
-performs extensive search, we can configure it to not use lemmas or invert derivations that
-are not specified in the \texttt{use} or \texttt{inv} flags.
+performs extensive search, we can configure it to only use lemmas or invert
+derivations specified by the \texttt{use} or \texttt{inv} flags.
+
+\jc{Just a digression but I find that CoqHammer automated proofs are \emph{not}
+good documentation compared to how we might have otherwise written the proofs
+manually. While the flags indicate what lemmas and inversions \emph{may} be used,
+the \texttt{best} tactic does not delete unused ones, so there's no way of knowing
+which ones actually \emph{are} used, and the automation hides away \emph{how} they
+are used. Automation makes proofs resilient to change when things go right, but
+when things go wrong, it becomes much harder to discern what lemmas I need to add
+when I don't even know where the existing ones are used.}
 
@@
 \section{Extensions}
+
 In this section, we discuss how the normalization
@@
 natural numbers, dependent pairs, cumulative
-universes with subtyping, and $\eta$ law for functions. All these
+universes with subtyping, and the $\eta$ law for functions. These
 extensions have been implemented in our development. More advanced
@@
 
+\jc{We may want to briefly mention the LOC for each addition?}
+
 \subsection{Natural Numbers}
 \begin{figure}
-  \drules[T]{$[[G |- a : A]]$}{Typing for Naturals}{Nat, Zero, Suc, Ind}
-  \drules[I]{$[[Interp I i A S]]$}{Logical Relation for Naturals}{Nat}
+  \drules[T]{$[[G |- a : A]]$}{Typing (naturals)}{Nat, Zero, Suc}
+  \vspace{-\baselineskip}
+  \begin{mathpar}
+    \drule[width=\textwidth]{T-Ind}
+  \end{mathpar}
+  \drules[I]{$[[Interp I i A S]]$}{Logical relation (naturals)}{Nat}
   \drules[N]{$[[NatOk a]]$}{Natural values}{Zero, Neu, Suc}
@@
 
-\Rref{I-Nat} gives us the semantic
-interpretation of natural number; a term $[[a]]$ semantically inhabits
-$[[Nat]]$ if $[[a]]$ reduces to some term $[[b]]$ such that $[[NatOk
-b]]$ holds.
-The $\vdash_\mathbb{N}$ judgment inductively characterizes the
+\Rref{I-Nat} gives us the semantic interpretation of natural number:
+a term $[[a]]$ semantically inhabits $[[Nat]]$ if $[[a]]$ reduces to
+some term $[[b]]$ such that $[[NatOk b]]$ holds.
+The $[[NatOk b]]$ judgment inductively characterizes the
 normal form of (potentially open) terms of type $[[Nat]]$ and is
@@
 \nameref{lemma:logrelcumulativity}.
+
+\subsection{\texorpdfstring{$\beta\eta$}{beta-eta}-normal forms}
 
@@
     $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidability of conversion \\
-    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidibility of type checking \\
+    MLTT-\`a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidibility of type checking \\
     \hline
@@
 In this work, we present a short and mechanized proof by logical relations for
-a dependently typed language with a full universe hierarchy, large
-eliminations, an intensional identity type, and dependent eliminators.  We
-show the extensibility of our approach by proving the existence of
-$\beta\eta$-normal forms with only small and mechanical changes to our proof
-development.  Our Coq mechanization leverages existing Coq libraries for
-reasoning about metatheory and for general purpose automation, allowing us to
-significantly reduce the verbosity typically associated with mechanized
-proofs. The result is a declarative proof style that rivals pen and paper.
+a dependently typed language with a predicative universe hierarchy, large
+elimination, an intensional identity type, and dependent eliminators.
+From the fundamental theorem for the logical relation, we prove consistency
+and canonicity, then show the extensibility of our approach by proving the
+existence of $\beta$($\eta$)-normal forms and adding natural numbers and
+cumulativity to the language. These extensions only require small and mechanical
+changes to our proof development. Our Coq mechanization leverages existing Coq
+libraries for reasoning about metatheory and for general purpose automation,
+allowing us to significantly reduce the verbosity typically associated with
+mechanized proofs. The result is a declarative proof style that rivals pen and paper.
 
@@
 
-% Type soundness can be proven through a syntactic
-% approach~\citep{syntacticsoundness} as a corollary of two properties:
-% progress and preservation. % The syntactic type soundness proof
-% % varies in complexity depending on the underlying type
-% % system. For example, a type system that tracks information flow would
-% % require additional structural rules related to security levels. In
-% % this paper, we focus on one specific type of complexity: the
-% In Figure~\ref{fig:stlcsoundness}, we summarize the structure of the
-% syntactic type soundness proof for the simply typed lambda
-% calculus. Each lemma can be proven by structural induction over the
-% typing derivation, while using the previous established results as
-% lemmas for specific cases that do not immediately follow from the
-% induction hypothesis. If we make our language more complex by adding
-% full dependent type support, the overall structure remains almost
-% identical.
-
-% NbE in Coq
-
 \bibliographystyle{ACM-Reference-Format}

### 2024-12-16 paper/paper.tex
Edits of Sections 9.1, 9.2

@@
 \usepackage[outputdir=latex.out]{minted} % latexrun puts everything in latex.out
-\usepackage{xspace,stmaryrd,mathtools,bbding}
+\usepackage{xspace,stmaryrd,mathtools,pifont}
 \usepackage[para]{footmisc}
@@
 \newcommand{\CR}{\mathit{CR}}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
 
@@
 rich set of features, including a countable universe
-hierarchy, booleans, and an intensional identity type. We show that
+hierarchy, booleans, and a propositional equality type. We show that
 the logical relation can be easily extended to prove the existence of
@@
 logical consistency for \lang{}, a dependent type theory with a
-predicative universe hierarchy, large elimination, an intensional identity
+predicative universe hierarchy, large elimination, a propositional equality
 type, a boolean base type, and dependent elimination forms.
@@
 consistency result to the existence of $\beta\eta$-normal forms.
-Intensional identity types $[[a ~ b : A]]$, or equality types, represent an
+Propositional equality types $[[a ~ b : A]]$, represent an
 equality between terms $[[a]]$ and $[[b]]$ of the same type $[[A]]$, whose
@@
   all quantified things are types. And we might need that information to interpret, say,
-  identity types in the right way.
+  equality types in the right way.
 }
@@
 \subsection{Logical relations for dependent types}
-In the most general sense, a logical relation can be viewed as a
+In the most general sense, a logical relation is a
 practical technique that uses a type-indexed relation to 
 strengthen the induction hypothesis for the
-property of interest. The original idea of this technique can be
-traced back to
-\citet{tait1967:reducibility}. This proof maps
+property of interest. This technique originates from
+\citet{tait1967:reducibility}, who maps
 types to sets of terms satisfying certain properties related to reduction.
-The same idea is explained in \citet{girard1989proofs} and extended to
+The same idea is explained by \citet{girard1989proofs} and extended to
 prove strong normalization of System F.
-%
 Tait's method has also been successfully applied to dependently typed
-languages to prove strong normalization \cite{Martin-Lof-1973,luo1990extended,geuvers1994short, barendregt:lambda-calculi-with-types}.
+languages to prove strong normalization \citep{Martin-Lof-1973,luo1990extended,geuvers1994short, barendregt:lambda-calculi-with-types}.
 
@@
 can be challenging to adapt to a theorem prover since many details
-are hidden behind concise notations.
+are hidden behind concise notation.
 For example, \citet{geuvers1994short} presents the interpretation for types as
 an inductively defined total function over the set of syntactically
-well-formed types. \scw{The issue is that Geuver's metalogic is (untyped) set
-theory, but Coq and Agda use type theory.}
+well-formed types.
 In untyped set theory, it makes sense to define the logical relation as a
-simply-typed function that takes a type and returns some set; however
-in constructive type theory, the metalogic of Coq and Agda, the
-interpretation function must be a dependently-typed function whose
-return type depends on the derivation of the well-typedness of its
-input. The well-typedness derivation and the proof of the
-classification theorem\scw{what is the classification theorem}\yl{It
-  says all well-formed terms A (either G |- A : .. or G |- .. : A)
-then A is either an object, a type, or a kind. The classification is
-used to guide erasure}
-are examined in the body
-of the interpretation function to decide whether an argument of an
-application should be erased during interpretation. As a result, this
-definition causes difficulties for modern proof assistants.
-Due to the
-impredicativity of the object language, \citet{geuvers1994short}'s
-proof cannot be encoded in Agda, which has a predicative
-metatheory. Due to the use of proof-relevant derivations, even in
-Coq, a proof assistant that supports impredicativity, one would need
-to constantly juggle between the impredicative but irrelevant sort
-\texttt{Prop} sort and the predicative but relevant sort
-\texttt{Type}.
-
-More recent work such as \citet{Abel12}
-and \citet{abel2008betaeta} make their definitions more explicit and
-precise and thus more directly encodable in proof assistants.
-% Our logical relation resembles their definition of a
-% semantic universe hierarchy.
-Unlike \lang, their systems support type-directed $\eta$ laws. These
-type-directed equational rules complicate the proof significantly
-since the confluence of full reduction is not available until the
-fundamental theorem is proven. Therefore,
-instead of a logical predicate, a relational model with
-syntactic type annotations is needed to remove the early reliance on
-confluence, despite the fact that the underlying type theory is
-intensional.
-
-\citet{coquand:canonicity,coquand:prop} use a \emph{proof-relevant} logical
+function that takes a type and returns a set; however,
+in constructive type theory, the metalogic of Coq and Agda, the return type of
+the interpretation function depends on the derivation of the well-typedness of
+its input. The body of the interpretation function uses the well-typedness
+derivation, along with a proof classifying whether the input is a kind, a type,
+or a term, to determine whether the argument of an application should be erased
+during interpretation.
+
+As a result, Geuvers' interpretation causes difficulties for modern proof
+assistants. Due to the impredicativity of the object language, the proof
+cannot be encoded in Agda, whose metatheory is predicative. However, even
+though Coq supports impredicativity, the proof relevance of the well-typedness
+derivations would require juggling between the impredicative but irrelevant
+\texttt{Prop} sort and the predicative but relevant \texttt{Type} sort.
+
+More recent work such by \citet{Abel12} and \citet{abel2008betaeta} makes the
+definitions more explicit and precise and thus more directly encodable in proof
+assistants. Unlike \lang, their systems support type-directed $\eta$ laws.
+These type-directed equational rules complicate the proof significantly
+since the confluence of full reduction is not available until after the
+fundamental theorem is proven. Therefore, instead of a logical predicate,
+they need a relational model with syntactic type annotations to remove the early
+reliance on confluence, in spite of the object type theory being intensional.
+
+\citet{coquand:canonicity,coquand:prop} uses a \emph{proof-relevant} logical
 relation to prove canonicity and normalization for intensional
-dependent type theories. % Intuitively, proof-relevant logical relations
-% can be viewed as the set theoretic augmented with information
-% about syntactic terms so the fundamental theorem can give us more
-Compared
-to the proof-irrelevant logical predicate we present, Coquand's proof
-has the benefit that it does not rely on the confluence of reduction,
+dependent type theories. In contrast to the proof-irrelevant logical predicate
+we present, Coquand's proof does not rely on the confluence of reduction,
 and can be easily extended to more $\eta$ laws without relying on a
@@
 Coquand works directly with an algebraic representation of dependent
-type theory called category with families (CwF)~\citep{hofmann:cwf},
+type theory using categories with families (CwF)~\citep{hofmann:cwf},
 though the idea of proof-relevant logical relations for dependent
@@
 
-More abstractly, the proof technique by Coquand can be viewed as an
-instance of a categorical construction called Artin Gluing
-~\citep{kaposi:gluing}. Different metatheoretic
-properties, such as normalization, parametricity, and
-$\Pi$-injectivity can be obtained by gluing the initial CwF with the
+More abstractly, Coquand's proof technique is an instance of a categorical
+construction called Artin gluing~\citep{kaposi:gluing}. Different
+metatheoretic properties, such as normalization, parametricity, and
+$\Pi$-injectivity, can be obtained by gluing the initial CwF with the
 corresponding CwF model. \citet{sterling:tait} introduces the notion of
-\emph{Synthetic Tait Computability}, which allows further
+synthetic Tait computability, which allows further
 simplification of such proofs by working in the internal language
@@
 expressive power of the metatheory. For example, the mechanization of
-NbE by \citet{altenkirch:normalisation} relies on Quotient Inductive
-Inductive Types (QIIT)~\citep{kaposi:qiit}.
+NbE by \citet{altenkirch:normalisation} relies on quotient
+inductive--inductive types (QIIT)~\citep{kaposi:qiit}.
 
@@
 
-%\newcommand\dheader[1]{\rotatebox{60}{{#1}}}
-\newcommand\header[1]{\rotatebox{0}{{#1}}}
-
-\begin{figure}[H]
-
-% \begin{minipage}{0.8\textwidth}
-  \begin{tabular}{| l |  c  | l | c | c | c | l | l | }
+\begin{figure}
+  \begin{tabular}{| l |  c  | l | c | c | c | l | l | r | }
     \hline
+%     \newcommand\header[1]{#1}
 %     & \header{Universes} & \header{Inductives} & \header{Conversion}
 %     & \header{Large Elim} & Main results \\
-      & \header{U} & \header{Ind} & \header{C} 
-      & \header{L E} & \header{A} &  Main results \\
-
+      & U & Ind & C & LE & A & Main results & LOC \\
     \hline
-    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & $\checkmark$ & 1 
-    & Consistency and normalization \\
-    % \hline
-    $\lambda^\theta$ & 0 & Id, Nat & U & $\XSolidBrush$ & 1 & Consistency \\
-    Core Nuprl & $\mathbb{N}$ & W-Types & E & $\checkmark$ & 2 & Consistency \\
-    NBE-in-Coq & 1 & Nat & T & $\XSolidBrush$ & 2 & Correctness of NBE \\
-    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidability of conversion \\
-    MLTT-\`a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidibility of type checking \\
+    \lang{} (this work) & $\mathbb{N}$ & Eq, Bool & U & \cmark & 1 
+    & Consistency, normalization & 1331 \\
+    $\lambda^\theta$ & 0 & Eq, Nat & U & \xmark & 1 & Consistency & \textasciitilde{}8k\\
+    Core Nuprl & $\mathbb{N}$ & W types & E & \cmark & 2 & Consistency & \textasciitilde{}330k \\
+    NBE-in-Coq & 1 & Nat & T & \xmark & 2 & Correctness of NbE & \textasciitilde{}20k \\
+    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & \cmark & 2 & Decidability: conversion & \textasciitilde{}10k \\
+    MLTT-\`a-la-Coq & 1 & Eq, Nat, $\Sigma$ & T & \cmark & 2 & Decidibility: type checking & \textasciitilde{}30k \\
     \hline
@@
 
-  \begin{tabular}{ll}
-    \\
-  \underline{U}niverses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
-  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), $\Sigma$-types ($\Sigma$), W-types \\
-  \underline{C}onversion:& Untyped (U), Typed (T),  Extensional (E) \\
-  \underline{L}arge \underline{E}liminations:& Included ($\checkmark$), not included ($\XSolidBrush$) \\
-  \underline{A}rity of interpretation: & Sets of terms (1), Relations between terms (2) \\
+  \begin{tabular}{ll} \\
+  \underline{U}niverses: & Countable ($\mathbb{N}$), zero (0), one (1) \\
+  \underline{Ind}uctives: & Equality types (Eq), naturals (Nat), dependent pairs ($\Sigma$), W types \\
+  \underline{C}onversion: & Untyped (U), typed (T),  extensional (E) \\
+  \underline{L}arge \underline{E}limination: & Included (\cmark), not included (\xmark) \\
+  \underline{A}rity of interpretation: & Sets of terms (1), relations between terms (2) \\
   \\
@@
   NBE-in-Coq& \citet{nbeincoq} \\
-  MLTT-\'a-la-Coq &\citet{martin-lof-a-la-coq} \\
+  MLTT-\`a-la-Coq &\citet{martin-lof-a-la-coq} \\
   \end{tabular}
@@
 
-\scw{Should we add a column for whether their logical relation is
-  unary or binary?}\yl{Technically a Kripke-style model has the
-  context as an argument to the logrel, too. By binary, we are really
-  talking about whether the interpreted set is binary or not. I think
-  it is tricky to explain what we actually mean. }
-\scw{I don't think it is that trick}
 Figure~\ref{fig:featurematrix} presents several mechanized proofs
-that feature logical-relations arguments for dependently-typed languages.
-Each of these proofs is significantly larger than than our development; but they
-also prove more results about different object languages.% \scw{Can we mention the
-% line counts in the text, when they are known?}
+that feature logical relations for dependently-typed languages.
+Each of these proofs is significantly larger than our development, but they
+also prove more results about different object languages.
 The table provides a comparison between the various features of their
@@
 impactful to the definition of the logical relation.
-
-\citet{casinghino:combining-proofs-programs} introduce $\lambda^\theta$, a dependently typed
-programming language that uses modality to distinguish between logical
-proofs and programs. % The programmatic fragment
+For a basic comparison of development size, we include the approximate
+nonblank, noncomment lines of proof code for each project.
+
+\citet{casinghino:combining-proofs-programs} introduce $\lambda^\theta$,
+a dependently typed programming language that uses modalities to distinguish
+logical proofs from programs.
+% The programmatic fragment
 % includes recursive data types and supports general recursion at the
@@
 The consistency proof of $\lambda^\theta$'s logical fragment has been
-mechanized in Coq through a step-indexed logical relation;
+mechanized in Coq using a step-indexed logical relation;
 step-indexing is required to model the programmatic fragment, which
@@
 a size metric, which has been used in~\citet{liu2023dependently}.
-Their development is around 8,000 lines of nonblank, noncomment code.
 
@@
 properties to facilitate the reuse of their definition. The
-logical relation is defined using the induction-recursion scheme,
+logical relation is defined using induction--recursion,
 which is available in Agda but not in Coq. Their development involves
 around 10,000 lines of Agda code.
-\citet{martin-lof-a-la-coq} transports the logical relation
-from \citet{decagda} in the predicative fragment of Coq and further
-extends the decidability of type conversion result
-from~\citet{decagda} to the decidability type checking of a
-bidirectional type system. Their development has around
-30,000 lines of Coq code.
+\citet{martin-lof-a-la-coq} adapts the logical relation
+by \citet{decagda} to the predicative fragment of Coq and further
+extends the decidability of type conversion result to the decidability of
+type checking a bidirectional type system.
 
 \citet{anand2014towards} mechanize the metatheory of
-Nuprl~\citep{constable1986implementing} in Coq. This metatheory is an
-extensional type theory with features such as dependent functions,
-inductive types, partial types, and a full universe hierarchy. They construct a PER
-model in Coq to show the logical consistency of their language. Their
-development has been further extended with features such as
-intersection types, union types, and quotient types. The extensive
-coverage of features results in a Coq development with around
+Nuprl~\citep{constable1986implementing} in Coq. The object theory is an
+extensional type theory with dependent functions, inductive types,
+partial types, and a full universe hierarchy. They construct a PER
+model in Coq to show the logical consistency of the language. Their
+development has been further extended with intersection types,
+union types, and quotient types, resulting in a Coq development of around
 330,000 lines of code.
-\citet{nbeincoq}
-mechanize the normalization-by-evaluation algorithm in Coq for a
+\citet{nbeincoq} mechanize a normalization-by-evaluation algorithm in Coq for a
 dependently typed language with one predicative universe, similar to
 \citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
-type system has no elimination form for natural numbers, the
-only base type from the object language, large elimination is not
-supported despite the one predicative universe. Their development has
-around 20,000 lines of Coq code.
-Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
-impredicative \texttt{Prop} sort of Coq to define the interpretation
+type system has no elimination form for natural numbers, the only base type,
+large elimination is not supported.
+Both \citet{anand2014towards} and \citet{nbeincoq} leverage
+Coq's impredicative \texttt{Prop} sort to define the interpretation
 of dependent function types and thus are closely related to our
-mechanization. \citet{anand2014towards} further show it is possible
-to encode a finite universe hierarchy without the use of
-either impredicativity or induction-recursion. Their encoding of a countable
+mechanization. \citet{anand2014towards} further show it possible
+to encode a finite universe hierarchy using neither impredicativity
+nor induction--recursion. Their encoding of a countable
 universe hierarchy relies on impredicativity, similar to our
 development.
+\jc{Wait I'm confused, do they use impredicativity or not?}
 
@@
 a dependently typed language with a predicative universe hierarchy, large
-elimination, an intensional identity type, and dependent eliminators.
+elimination, a propositional equality type, and dependent eliminators.
 From the fundamental theorem for the logical relation, we prove consistency

### 2024-12-16 paper/paper.tex
Added notes to Section 9

@@
 
+\jc{Missing from the table:
+\begin{itemize}
+  \item Altenkirch and Kaposi (2016) using QIITs, mentioned above (Agda);
+    mechanization might be part of Kaposi's thesis (2016)
+  \item prior work Chapman (2009), Type Theory should Eat Itself (Agda);
+    possibly also his thesis (2008)
+  \item Kipling by McBride (2010), mixing deep and shallow embedding
+    using host conversion (Agda) to avoid the whole QIIT business
+  \item Kaposi et al. (2019), Shallow Embedding of Type Theory (Agda),
+    although at this point I'm not sure if this counts as having a syntax
+\end{itemize}
+}
+
 Figure~\ref{fig:featurematrix} presents several mechanized proofs
@@
 
-\citet{barras2010sets, Wang2013SemanticsOI} assign
+\Citet{barras2010sets} and \citet{Wang2013SemanticsOI} assign
 set-theoretic semantics to dependent type theory in Coq. Unlike the
-previous efforts, which primarily focus on predicative
-type theory and more direct reducibility models,
-\citet{barras2010sets, Wang2013SemanticsOI} tackle extensions of
-$CC^\omega$, a system that incorporates a predicative universe on top
-of the impredicative sort in the Calculus of Constructions. We choose to
-focus on a syntactic term model so we do not have to take the extra step
-of mechanizing mathematical objects such as sets and domains.
+previous efforts, which focus on predicative type theory and direct
+reducibility models, they tackle extensions of
+CC$^\omega$, which extends the Calculus of Constructions with predicative
+universes on top of the impredicative sort. We choose to
+focus on a syntactic term model to avoid mechanizing mathematical objects
+such as sets and domains.
 
@@
 example, \citet{coqcoqcorrect2019} prove the correctness of a type checker
-for the Predicative, Cumulative Calculus of Inductive Constructions (PCUIC),
-Coq's core calculus, assuming the strong normalization property of the object
-language. \citet{weirich:systemd} define System D, a core calculus of
+for the Predicative Calculus of Cumulative Inductive Constructions (pCuIC),
+Coq's core calculus, assuming strong normalization.
+\citet{weirich:systemd} define System D, a core calculus of
 dependent Haskell, and prove the syntactic type safety of the type
-system. Because System D includes nontermination, they proved
-the consistency of definitional equality from the confluence of 
+system. Because System D includes nontermination, they prove
+consistency of definitional equality from the confluence of 
 parallel reduction.
-
-Compared to the systems described here, the most notable features we are
-missing are cumulativity and impredicativity. Our semantic model already
-satisfies the cumulativity property (Lemma~\ref{lemma:logrelcumulativity}),
-but we need to extend our convertibility relation into a subtyping relation in
-our syntactic typing rules. Impredicativity, on the other hand, is known to be
+\jc{Why specifically mention System D? Surely there are other more notable
+(in reviewers' eyes) dependent languages for which type safety have been proven?}
+
+Compared to the systems described here, the most notable feature we are
+missing is impredicativity, which is known to be
 difficult to model when the impredicative sort is at the bottom of a
-predicative universe hierarchy; in this scenario, the erasure technique from
-\citet{geuvers1994short} is not
+predicative universe hierarchy.
+\jc{This makes it sound like it's easy when it's at the top of a predicative
+universe hierarchy, which I don't think is what you meant; did you mean when
+impredicativity is combined with some sort of large elimination of base types/
+type-level computation?}
+In this scenario, the erasure technique from \citet{geuvers1994short} is not
 applicable~\citep{abel2013normalization}. Whether there is a similarly short

### 2024-12-18 paper/paper.tex
Moved a comment from 9.2 to 9.1
Added a paragraph about shallow embeddings

@@
 inductive--inductive types (QIIT)~\citep{kaposi:qiit}.
+\jc{I don't understand how the QIIT model relates to gluing,
+or what using QIIT has to do with proof relevance of the logical relation.}
+
+So far, the works mentioned interpret types as sets or relations on deep
+representations of terms. In contrast, shallow embeddings interpret types as
+sets of \emph{meta-level} terms in some type theory. The logical relation
+interprets the object language into the meta language, and its soundness holds
+by virtue of the interpretation function being well typed and definable.
+Examples include work by \citet{kipling}, who mixes deep and shallow
+representations, and work by \citet{shallow}, who directly implement the
+signature of a CwF as Agda terms. The key property of shallow embeddings is
+that object-level conversion is represented as meta-level equality, which has
+the benefit of delegating handling conversion to the type checker, but with the
+drawback that object conversion must coincide with meta equality. Conversion
+cannot be customized to convert more terms than dictated by their
+interpretations.
 
@@
 
-\jc{Missing from the table:
-\begin{itemize}
-  \item Altenkirch and Kaposi (2016) using QIITs, mentioned above (Agda);
-    mechanization might be part of Kaposi's thesis (2016)
-  \item prior work Chapman (2009), Type Theory should Eat Itself (Agda);
-    possibly also his thesis (2008)
-  \item Kipling by McBride (2010), mixing deep and shallow embedding
-    using host conversion (Agda) to avoid the whole QIIT business
-  \item Kaposi et al. (2019), Shallow Embedding of Type Theory (Agda),
-    although at this point I'm not sure if this counts as having a syntax
-\end{itemize}
-}
-
 Figure~\ref{fig:featurematrix} presents several mechanized proofs
-that feature logical relations for dependently-typed languages.
+that feature logical relations for dependently typed languages.
 Each of these proofs is significantly larger than our development, but they
@@
 logical relation is defined using induction--recursion,
-which is available in Agda but not in Coq. Their development involves
-around 10,000 lines of Agda code.
+which is available in Agda but not in Coq.
 \citet{martin-lof-a-la-coq} adapts the logical relation
@@
 development has been further extended with intersection types,
-union types, and quotient types, resulting in a Coq development of around
-330,000 lines of code.
+union types, and quotient types.
 \citet{nbeincoq} mechanize a normalization-by-evaluation algorithm in Coq for a

### 2024-12-18 paper/paper.tex
Uniformly write 'proof by logical relation' (without the s) for now

@@
 \begin{abstract}
-Proof by logical relations is a powerful technique that has been used
+Proof by logical relation is a powerful technique that has been used
 to derive metatheoretic properties of type systems, such as
@@
   supplementary material.
-\item We compare our work to existing proofs by logical relations and other
+\item We compare our work to existing proofs by logical relation and other
   proof techniques for proving consistency and normalization.  We provide an
@@
 conveying the proof technique itself. Unlike properties derivable through
-purely syntactic means, proofs by logical relations make demands on the
+purely syntactic means, proofs by logical relation make demands on the
 strength of the metatheory in which they are expressed. An informal proof that
@@
 \label{sec:conclusion}
-In this work, we present a short and mechanized proof by logical relations for
+In this work, we present a short and mechanized proof by logical relation for
 a dependently typed language with a predicative universe hierarchy, large

### 2024-12-18 paper/paper.tex
Comments on discussion

@@
 need a Kripke-style relational model to prove $\Pi$-injectivity
-among other properties, unlike systems with typed conversion. Furthermore, we prove the existence of normal
+(among other properties), unlike systems with typed conversion.
+\jc{Why do we need Pi-injectivity? Why would a Kripke-style model be required
+to prove it for typed conversion? What is a Kripke-style model?}
+Furthermore, we prove the existence of normal
 forms, which induces a simple \emph{normalize-and-compare}
 procedure for type
-conversion~\cite{pierce2004advanced}. \citet{nbeincoq, decagda}, on
+conversion~\cite{pierce2004advanced}. \Citet{nbeincoq} and \citet{decagda}, on
 the other hand, need
 to show how their algorithmic conversion procedure is sound and complete
-with respect to their respective declarative equational theory.
+with respect to their respective declarative equational theories.
 \scw{I'm getting confused by this paragraph. Does this reorganization sense:
@@
 reduction relation, to close over our semantic interpretation of types
-and terms. Parallel reduction is non-deterministic, but it satisfies
+and terms. Parallel reduction is nondeterministic, but it satisfies
 useful structural properties such as congruence
@@
 (Lemma~\ref{lemma:pardiamond}). We pay the price of using a
-non-deterministic reduction relation when we want to prove that our
+nondeterministic reduction relation when we want to prove that our
 logical relation is a partial function; because of \rref{I-Red}, we
@@
 with a deterministic reduction relation, we would likely need
-results such as the factorization theorem~\citep{takahashi-parallel-reduction,factorization-essentially} in our development before we
-can prove the fundamental theorem, leading to a more complicated proof.
+results such as the factorization
+theorem~\citep{takahashi-parallel-reduction,factorization-essentially}
+in our development before we can prove the fundamental theorem,
+leading to a more complicated proof.
+
 % With
@@
 before the fundamental theorem is established can be alarming.
+\jc{The rest of this paragraph is about confluence and Pi-injectivity,
+not untyped conversion Kripke-style, so maybe the first sentence of the
+paragraph shouldn't start with talking about them?}
 In a system with type-directed reduction,
@@
 from the logical relation. \citet{siles2012pure} generalize the
-notion of Type Parallel One Step Reduction from \citet{adams2006pure}
+notion of typed parallel one-step reduction by \citet{adams2006pure}
 to syntactically prove $\Pi$-injectivity for arbitrary Pure Type

### 2024-12-18 paper/paper.tex
Remove some handled comments

@@
 conversion.
-\jc{I'm not too fond of this last sentence because we \emph{do} lose
-the ability to add typed $\eta$ equivalences.}
-\yl{removed the eta equivalence sentence because it might give the
-  wrong message that we extended Siles' work with eta:
- Furthermore, we discuss how this definition can be extended with
- $\eta$-equivalence of functions in Section~\ref{XXX} }
 We compare this
@@
 
-\scw{Cassia did not follow our discussion of dependent pattern matching. We need to expand}
 The elimination form for booleans in \rref{T-If} demonstrates dependent pattern
@@
 
-\scw{Would it make sense to define the notation $a \in [\![A]\!]^i$ when
-there exists some $S$ such that $[[InterpR i A S]]$ and $[[a in S]]$ ?}
-\jc{I like this notation too, $[[a in InterpR i A]]$}
-
 In this section, we show that all closed, well-typed terms are contained
@@
 \end{lemma}
-
-\scw{This lemma is only used to prove \texttt{wn\_antirenaming}, which is then
-used more generally. Maybe we should replace it with  that?}
-\yl{Anti-renaming is mentioned once in Section 6 when we talk about how confluence with $\eta$
-depends on anti-renaming for Par. I don't think that discussion in
-Section 6 is very important so I'm fine with replacing it}
 
